### 2010-12-01 CacheProof.tex
cache proof

\documentclass{article}

\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}

\begin{lemma}
Child-state is at x due to parent receiving x-resp and child is at y due to child receiving y-resp $\Rightarrow$ x = y.
\label{CRecvPRecvNetwork}
\end{lemma}

\begin{proof}
Proof by contradiction, assuming x $\neq$ y.

\begin{figure}
\centering
%\includegraphics{CRecvPRecvNetwork.1}
\end{figure}

Child has sent x-resp before receiving y-resp ($\because$ y-resp is the last message causing child transition).

Parent has sent y-resp before receiving x-resp ($\because$ x-resp is the last message causing child-state transition).

$\Rightarrow$ The corresponding y-req should have been sent by the child before sending x-resp ($\because$ FIFO ordering between child to parent req, resp).

$\Rightarrow$ x-resp is not voluntary.

$\Rightarrow$ x-req is sent by parent before sending y-resp ($\because$ x-resp is sent by child, and hence x-req is received by child before receving y-resp and FIFO order ordering between parent to child messages).

But y-resp will not be sent by parent if there is a pending request.

\begin{figure}
\centering
%\includegraphics{CRecvPRecvNetwork.2}
\end{figure}

$\Rightarrow \Leftarrow$
\end{proof}

\begin{lemma}
Child-state is at x due to parent sending x-resp and child is at y due to child sending y-resp, corresponding x-resp was sent by parent before receiving y-resp and y-resp was sent by child before receiving x-resp $\Rightarrow$ x = y.
\label{InsaneNetwork}
\end{lemma}

\begin{proof}
Proof by contradiction, assuming x $\neq$ y.

\begin{figure}
\centering
%\includegraphics{InsaneNetwork.1}
\end{figure}

x-resp has not reached child ($\because$ y-resp is the last message sent by child causing child transition, and child is still in y).

y-resp has not reached parent ($\because$ x-resp is the last message sent by parent causing child-state transition, and child-state is still in x).

Corresponding x-req was received by parent before receiving y-resp.

$\Rightarrow$ x-req was sent by child before sending y-resp ($\because$ FIFO ordering between child to parent req, resp).

$\Rightarrow$ y-resp is not voluntary.

$\Rightarrow$ Corresponding y-req was sent by parent before sending x-resp ($\because$ FIFO ordering between parent to child messages).

$\Rightarrow$ x-resp was sent by parent before receiving y-resp for y-req sent by parent which is not possible ($\because$ parent does not respond when there is a pending request).

\begin{figure}
\centering
%\includegraphics{InsaneNetwork.2}
\end{figure}

$\Rightarrow \Leftarrow$

\end{proof}

\begin{lemma}
Child-state is at x due to parent sending x-resp and child is at y due to child receiving y-resp $\Rightarrow$ y $\le$ x.
\label{CRecvPSend<=}
\end{lemma}

\begin{proof}
Proof by contradiction, assuming x $<$ y.

\begin{figure}
\centering
%\includegraphics{CRecvPSend<=.1}
\end{figure}

Parent sends x-resp only after y-resp ($\because$ x-resp is the last message causing child-state transition).

Child can send corresponding x-req for the x-resp from parent only after child receives y-resp ($\because$ only one pending request is allowed). But y-resp is the last message causing child transition, so child is still at y.

$\Rightarrow$ x-req can not be sent by child since x $<$ y.

$\Rightarrow \Leftarrow$

\end{proof}

\begin{lemma}
Child-state is at x due to parent receiving x-resp and child is at y due to child sending y-resp $\Rightarrow$ y $\le$ x.
\label{CSendPRecv<=}
\end{lemma}

\begin{proof}
Proof by contradiction, assuming x $<$ y.

\begin{figure}
\centering
%\includegraphics{CSendPRecv<=.1}
\end{figure}

Child sends y-resp only after x-resp ($\because$ y-resp is the last message causing child transition).

Child must be in z $>$ y to send y-resp.

$\Rightarrow$ x $<$ z ($\because$ x $<$ y).

$\Rightarrow$ Child must have sent z-req after sending x-resp and transitioning to x, and eventually got back z-resp.

$\Rightarrow$ Parent must have received z-req and sent z-resp, while child-state transitions to z. But this is not possible as x-resp was the last message causing child state transition.

$\Rightarrow \Leftarrow$

\end{proof}

\begin{lemma}
Child-state is at x due to parent sending x-resp and child is at y due to child sending y-resp $\Rightarrow$ y $\le$ x.
\label{CSendPSend<=}
\end{lemma}

\begin{proof}

Proof by contradiction, assuming x $<$ y.

\begin{enumerate}
\item Parent sends x-resp before receiving y-resp and x-resp reaches child after child sends y-resp - Not possible by Lemma \ref{InsaneNetwork}.
\item Parent sends x-resp before receiving y-resp and x-resp reaches child before child sends y-resp.

\begin{figure}
\centering
%\includegraphics{CSendPSend<=.1}
\end{figure}

Child must be in z $>$ y to send y-resp.

$\Rightarrow$ x $<$ z ($\because$ x $<$ y and y $<$ z)

$\Rightarrow$ After child received x-resp, somehow it must reach higher state z. This is possible only if parent sends z-resp, thereby child-state transitioning to z. But x-resp was the last message causing child-state transition.

$\Rightarrow \Leftarrow$

\item Parent receives y-resp before parent sends x-resp.

  \begin{enumerate}
  \item Child sends x-req corresponding to x-resp after sending y-resp.

  \begin{figure}
  \centering
  %\includegraphics{CSendPSend<=.2}
  \end{figure}

  Child is at y $>$ x as y-resp is the last message causing transition for child. Since x $<$ y, child wont send x-req.

  $\Rightarrow \Leftarrow$

  \item Child sends x-req corresponding to x-resp before sending y-resp.

  \begin{figure}
  \centering
  %\includegraphics{CSendPSend<=.3}
  \end{figure}

  Child is at z $<$ x when x-req is sent and z' $>$ y when y-resp is sent.

  $\Rightarrow$ z $<$ z' ($\because$ x $<$ y and z' $>$ y).

  $\Rightarrow$ z'-req must be sent by child before x-resp is received (for x-req). But this is not possible as only one pending request is allowed.

  $\Rightarrow \Leftarrow$

  \end{enumerate}

\end{enumerate}

\end{proof}

\begin{lemma}
Child-state is at x due to parent sending x-resp and child is at y due to child receiving y-resp and parent just received z-req $\Rightarrow$ y = x.
\label{CRecvPSendReq}
\end{lemma}

\begin{proof}

Proof by contradiction, assuming x $\neq$ y.

\begin{figure}
\centering
%\includegraphics{CRecvPSendReq.1}
\end{figure}

Parent sends x-resp only after y-resp ($\because$ x-resp is the last message causing child-state transition).

Child sends corresponding x-req only after receiving y-resp ($\because$ only one pending request is allowed).

$\Rightarrow$ Child can not send z-req before receiving x-resp ($\because$ only one pending request is allowed and x-resp has not reached child yet).

$\Rightarrow \Leftarrow$

\end{proof}

\begin{lemma}
Child-state is at x due to parent receiving x-resp and child is at y due to child sending y-resp and parent just received z-req $\Rightarrow$ y = x.
\label{CSendPRecvReq}
\end{lemma}

\begin{proof}

Proof by contradiction, assuming x $\neq$ y.

\begin{figure}
\centering
%\includegraphics{CSendPRecvReq.1}
\end{figure}

Child sends y-resp only after sending x-resp ($\because$ y-resp is the last message causing child transition).

Parent receives z-req before receiving y-resp ($\because$ x-resp is the last message causing child-state transition).

$\Rightarrow$ Child sent z-req before sending y-resp ($\because$ FIFO ordering between child to parent req, resp).

$\Rightarrow$ y-resp is not voluntary ($\because$ child has pending z-req).

$\Rightarrow$ Parent does not receive z-req till it receives y-resp for the corresponding y-req that parent sent.

$\Rightarrow \Leftarrow$

\end{proof}

\begin{lemma}
Child-state is at x due to parent sending x-resp and child is at y due to child sending y-resp and parent just received z-req $\Rightarrow$ y = x.
\label{CSendPSendReq}
\end{lemma}

\begin{proof}

Proof by contradiction, assuming x $\neq$ y.

\begin{enumerate}
\item Parent sends x-resp before receiving y-resp and x-resp reaches child after child sends y-resp - Not possible because of Lemma \ref{InsaneNetwork}.
\item Parent sends x-resp before receiving y-resp and x-resp reaches child before child sends y-resp.

\begin{figure}
\centering
%\includegraphics{CSendPSend.1}
\end{figure}

y-resp can not be sent voluntarily by child as z-req is pending.

$\Rightarrow$ There is a pending y-req from parent.

$\Rightarrow$ Parent wont receive z-req ($\because$ only one pending request is allowed)

$\Rightarrow \Leftarrow$

\item Parent receives y-resp before parent sends x-resp.

\begin{figure}
\centering
%\includegraphics{CSendPSend.2}
\end{figure}

$\Rightarrow$ There is a pending x-req from child. This is not possible since there are two pending requests from child as child has not received x-resp at this moment.

$\Rightarrow \Leftarrow$

\end{enumerate}

\end{proof}

\begin{lemma}
Child-state is at x due to parent sending x-resp and child was at y before sending z-resp due to child receiving y-resp and parent just received z-resp $\Rightarrow$ y = x.
\label{CRecvPSendResp}
\end{lemma}

\begin{proof}

Proof by contradiction, assuming x $\neq$ y.

\begin{figure}
\centering
%\includegraphics{CRecvPSendResp.1}
\end{figure}

Parent sends x-resp only after y-resp ($\because$ x-resp is the last message causing child-state transition).

Child sends corresponding x-req only after receiving y-resp ($\because$ only one pending request is allowed).

$\Rightarrow$ z-resp is not voluntary.

$\Rightarrow$ Corresponding z-req must be received by child before receiving x-resp ($\because$ x-resp has not yet been received by child when it sent z-resp).

$\Rightarrow$ x-resp is sent by parent between sending z-req ($\because$ FIFO ordering between parent to child messages) and receiving z-resp, which is not possible.

$\Rightarrow \Leftarrow$

\end{proof}

\begin{lemma}
Child-state is at x due to parent receiving x-resp and child is at y before sending z-resp due to child sending y-resp and parent just received z-resp $\Rightarrow$ y = x.
\label{CSendPRecvResp}
\end{lemma}

\begin{proof}

Parent must receive y-resp before receiving z-resp ($\because$ FIFO ordering between child to parent resps). This means that x-resp is not the last message inducing child-state transition.

$\Rightarrow \Leftarrow$

\end{proof}

\begin{lemma}
Child-state is at x due to parent sending x-resp and child is at y before sending z-resp due to child sending y-resp and parent just received z-resp $\Rightarrow$ y = x.
\label{CSendPSendResp}
\end{lemma}

\begin{proof}

Proof by contradiction, assuming x $\neq$ y.

\begin{enumerate}
\item Parent sends x-resp before receiving y-resp.

z-resp reaches parent.

\begin{figure}
\centering
%\includegraphics{CSendPSendResp.1}
\end{figure}

$\Rightarrow$ y-resp reaches parent ($\because$ FIFO ordering between child to parent resps), after x-resp is sent which is not possible.

$\Rightarrow \Leftarrow$

\item Parent receives y-resp before parent sends x-resp.

\begin{figure}
\centering
%\includegraphics{CSendPSendResp.2}
\end{figure}

There is a corresponding x-req that parent receives when sending x-resp. Parent receives x-req before z-resp is received.

$\Rightarrow$ x-req was sent by child before z-resp was sent ($\because$ FIFO ordering between child to parent req, resp).

$\Rightarrow$ z-resp is not voluntary.

$\Rightarrow$ Corresponding z-req was received by child before x-resp was received ($\because$ x-resp was not yet received).

$\Rightarrow$ x-resp was sent by parent between sending z-req ($\because$ FIFO ordering between parent to child messages) and z-resp. This is not possible.

$\Rightarrow \Leftarrow$

\end{enumerate}

\end{proof}

\begin{theorem}
If a request is made for $X$ to transition to $y$, then $X$ will eventually reach $y$ if
\begin{itemize}
\item In every level there is atleast one buffer from parent to child, one buffer for child to parent req, each exclusive to that level and one overall child to parent resp buffer, i.e., each level has a separate virtual channel or network for each of parent-to-child-mesg, child-to-parent-req and one overall child-to-parent-resp for all levels.
\item A request is not left to starve if there exists a local state transition involving that request.
\end{itemize}
\end{theorem}

\begin{proof}
Assume initially that there are infinite buffers.

A leaf can always process any request from its parent and send appropriate response. It doesn't generate any new requests to wait on.

A root can always process any request from its children and send appropriate response. It doesn't generate any new requests to wait on.

Any response sent to a node will be consumed by the node.

Assume for induction that any node whose distance from the leaf less than n eventually generates appropriate response on a request. If n is 1, hypothesis holds (the node is the leaf itself). Assume simultaneosly that any node whose distance from the root less than m eventually generates appropriate response on a request. If m is 1, hypothesis holds (the node is the root itself).

Consider a node whose distance from leaf is n and distance from root is m.

Consider a request to this node from its parent. If it ever gets ``processed'', it can at most generate requests to its children, which will send back appropriate responses (due to induction hypothesis) leading to a good state for the node, and this node will respond its parent. If there is a pending request from child which is getting ``processed'', it can either generate requests to its children, which will send back appropriate responses (due to induction hypothesis) or it can generate request to its parent. But the node can respond to a request from its parent even when it has its own request to parent. So the request from parent will eventually be fully processed leading to an appropriate response.

Consider a request to this node from the child. If it only generates requests to children, then they will send back appropriate responses (by induction hypothesis). Otherwise, it can generate request to its parent. But this will also generate appropriate response by induction hypothesis.

Thus any request will generate an appropriate response if we had infinite buffers.

Let there be just one buffer for parent-to-child, one buffer for child-to-parent-req for each level and one buffer for child-to-parent-resp. The responses are always consumed, so just one buffer overall is suffficient. At the leaf level, any request to a leaf will be processed. So one buffer is sufficient from parent to leaf at that level. Similarly, any request to the root will be processed. So one buffer is sufficient from child to root at that level. (This is true as long as the arbiter avoids starvation). Thus, by induction, one can prove.
\end{proof}

\begin{theorem}
At a moment, child-state is x and child is y $\Rightarrow$ y $\le$ x.
\label{<=}
\end{theorem}

\begin{proof}
Initially this is true (all states are invalid).

If only child-state has changed, it can only go up from invalid.

If only child has changed to x, then it can change only due to a x-resp from parent ($\because$ invalid $<$ x). This means child-state also changes. So this is not possible.

If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSend<=}, \ref{CSendPRecv<=} and \ref{CSendPSend<=}, examines all scenarios.
\end{proof}

\begin{lemma}
When parent receives z-req, child-state is x and child is y $\Rightarrow$ y = x.
\label{Req=}
\end{lemma}

\begin{proof}
Initially this is true (all states are invalid).

If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSendReq}, \ref{CSendPRecvReq} and \ref{CSendPSendReq}, examines all scenarios.
\end{proof}

\begin{lemma}
When parent receives z-resp, child-state is x and child is y $\Rightarrow$ y = x.
\label{Resp=}
\end{lemma}

\begin{proof}
Initially this is true (all states are invalid).

If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSendResp}, \ref{CSendPRecvResp} and \ref{CSendPSendResp}, examines all scenarios.
\end{proof}

\end{document}

### 2011-10-04 old/CacheProof.tex
retaining old stuff

--- 2010-12-01 CacheProof.tex
@@
 
-\begin{theorem}
-If a request is made for $X$ to transition to $y$, then $X$ will eventually reach $y$ if
-\begin{itemize}
-\item In every level there is atleast one buffer from parent to child, one buffer for child to parent req, each exclusive to that level and one overall child to parent resp buffer, i.e., each level has a separate virtual channel or network for each of parent-to-child-mesg, child-to-parent-req and one overall child-to-parent-resp for all levels.
-\item A request is not left to starve if there exists a local state transition involving that request.
-\end{itemize}
-\end{theorem}
-
-\begin{proof}
-Assume initially that there are infinite buffers.
-
-A leaf can always process any request from its parent and send appropriate response. It doesn't generate any new requests to wait on.
-
-A root can always process any request from its children and send appropriate response. It doesn't generate any new requests to wait on.
-
-Any response sent to a node will be consumed by the node.
-
-Assume for induction that any node whose distance from the leaf less than n eventually generates appropriate response on a request.
-If n is 1, hypothesis holds (the node is the leaf itself).
-Assume simultaneosly that any node whose distance from the root less than m eventually generates appropriate response on a request.
-If m is 1, hypothesis holds (the node is the root itself).
-
-Consider a node whose distance from leaf is n and distance from root is m.
-
-Consider a request to this node from its parent.
-If it ever gets ``processed'', it can at most generate requests to its children, which will send back appropriate responses (due to induction hypothesis) leading to a good state for the node, and this node will respond its parent.
-If there is a pending request from child which is getting ``processed'', it can either generate requests to its children, which will send back appropriate responses (due to induction hypothesis) or it can generate request to its parent.
-But the node can respond to a request from its parent even when it has its own request to parent.
-So the request from parent will eventually be fully processed leading to an appropriate response.
-
-Consider a request to this node from the child.
-If it only generates requests to children, then they will send back appropriate responses (by induction hypothesis).
-Otherwise, it can generate request to its parent.
-But this will also generate appropriate response by induction hypothesis.
-
-Thus any request will generate an appropriate response if we had infinite buffers.
-
-Let there be just one buffer for parent-to-child, one buffer for child-to-parent-req for each level and one buffer for child-to-parent-resp.
-The responses are always consumed, so just one buffer overall is suffficient.
-At the leaf level, any request to a leaf will be processed.
-So one buffer is sufficient from parent to leaf at that level.
-Similarly, any request to the root will be processed.
-So one buffer is sufficient from child to root at that level.
-(This is true as long as the arbiter avoids starvation).
-Thus, by induction, one can prove.
-\end{proof}
-
-\begin{theorem}
+\begin{lemma}
 At a moment, child-state is x and child is y $\Rightarrow$ y $\le$ x.
 \label{<=}
-\end{theorem}
+\end{lemma}
 
-\begin{proof}
-Initially this is true (all states are invalid).
-
-If only child-state has changed, it can only go up from invalid.
-
-If only child has changed to x, then it can change only due to a x-resp from parent ($\because$ invalid $<$ x). This means child-state also changes. So this is not possible.
-
-If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSend<=}, \ref{CSendPRecv<=} and \ref{CSendPSend<=}, examines all scenarios.
-\end{proof}
+%\begin{proof}
+%Initially this is true (all states are invalid).
+%
+%If only child-state has changed, it can only go up from invalid.
+%
+%If only child has changed to x, then it can change only due to a x-resp from parent ($\because$ invalid $<$ x). This means child-state also changes. So this is not possible.
+%
+%If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSend<=}, \ref{CSendPRecv<=} and \ref{CSendPSend<=}, examines all scenarios.
+%\end{proof}
 
@@
 
-\begin{proof}
-Initially this is true (all states are invalid).
-
-If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSendReq}, \ref{CSendPRecvReq} and \ref{CSendPSendReq}, examines all scenarios.
-\end{proof}
+%\begin{proof}
+%Initially this is true (all states are invalid).
+%
+%If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSendReq}, \ref{CSendPRecvReq} and \ref{CSendPSendReq}, examines all scenarios.
+%\end{proof}
 
@@
 
-\begin{proof}
-Initially this is true (all states are invalid).
-
-If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSendResp}, \ref{CSendPRecvResp} and \ref{CSendPSendResp}, examines all scenarios.
-\end{proof}
+%\begin{proof}
+%Initially this is true (all states are invalid).
+%
+%If both child and child-state have changed, Lemmas \ref{CRecvPRecvNetwork}, \ref{CRecvPSendResp}, \ref{CSendPRecvResp} and \ref{CSendPSendResp}, examines all scenarios.
+%\end{proof}
 

### 2012-10-27 impl/Connect.h
stuff

@@ -26,6 +26,11 @@ public:
   }
 
   void cycle() {
+//    printf("%p nextCReq: %d ", this, nextCReq);
+//    for(Child i = 0; i < childs; i++) {
+//      printf("%d", csReqToP[i]->empty());
+//    }
+//    printf("\n");
     if(!pReqFromC->full()) {
       for(Child i = 0; i < childs; i++) {
         Child c = (nextCReq + i)%childs;

### 2012-10-27 impl/Debug.h

@@ -1,6 +1,6 @@
 #pragma once
 
-#ifdef DEBUG
+#ifdef Debug
 #define printSendRespToP(lineAddr, trigger, index, to)\
   printf("%p SendRespToP: %llx %c %d.%d %d->%d\n",\
          this, lineAddr, trigger == Forced? 'F': 'V', index.set, index.way, cache.st[index.set][index.way], to);

### 2012-10-27 impl/System_2_2.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 2;
+U8 levels = 1;
+Child childs[] = {4};
+MshrPtr mshrs[] = {64};
+Way ways[] = {8, 16};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {5};
+Latency dataLats[] = {10};
+Latency memLat = 40;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-27 impl/System_2_1_1.cpp
stuff

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 2;
+U8 levels = 2;
+Child childs[] = {2, 2};
+MshrPtr mshrs[] = {32, 64};
+Way ways[] = {8, 16, 16};
+U8 setSzs[] = {6, 8, 10};
+Latency tagLats[] = {5, 10};
+Latency dataLats[] = {5, 10};
+Latency memLat = 40;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-27 impl/CacheTypes.h
stuff

@@ -91,8 +91,8 @@ public:
     to = _to;
   }
   ~toCs() {
-    delete children;
-    delete from;
+    delete[] children;
+    delete[] from;
   }
 } ToCs;
 

### 2012-10-27 impl/Fifo.h

@@ -10,7 +10,7 @@ using namespace std;
 
 typedef class fifo {
 private:
-  queue<void*, list<void*> > q;
+  queue<void*> q;
   U8 size;
   U8 numElems;
   bool enqEn;

### 2012-10-28 impl/CacheTypes.h
stuff

@@ -17,17 +17,23 @@ typedef struct {
   Way way;
 } Index;
 
+typedef enum {Req, Resp, FwdReq, FwdResp} FromPType;
+
 typedef class fromP {
 public:
-  bool isReq;
+  FromPType type;
   Index index;
   LineAddr lineAddr;
   St from;
   St to;
 
-  fromP(bool _isReq, Index _index, LineAddr _lineAddr, St _from, St _to) {
-    isReq = _isReq; index = _index; lineAddr = _lineAddr; from = _from; to = _to;
-  }
+  bool isFwdReq;
+  bool changeState;
+  Child fwdC;
+  St fwdTo;
+
+  fromP(FromPType _type, Index _index, LineAddr _lineAddr, St _from, St _to):
+       type(_type), index(_index), lineAddr(_lineAddr), from(_from), to(_to), isFwdReq(false) {}
   ~fromP() {}
 } FromP;
 
@@ -65,25 +71,25 @@ typedef class toCs {
 public:
   Child childs;
   bool* children;
-  bool isReq;
+  FromPType type;
   Index index;
   LineAddr lineAddr;
   St* from;
   St to;
 
-  toCs(Child _childs, bool* _children, bool _isReq, Index _index,
+  toCs(Child _childs, bool* _children, FromPType _type, Index _index,
        LineAddr _lineAddr, St* _from, St _to) {
-    childs = _childs; children = _children; isReq = _isReq;
+    childs = _childs; children = _children; type = _type;
     index = _index; lineAddr = _lineAddr; from = _from; to = _to;
   }
-  toCs(Child _childs, Child c, bool _isReq, Index _index,
+  toCs(Child _childs, Child c, FromPType _type, Index _index,
        LineAddr _lineAddr, St _from, St _to) {
     childs = _childs;
     children = new bool[childs];
     for(Child i = 0; i < childs; i++)
       children[i] = false;
     children[c] = true;
-    isReq = _isReq;
+    type = _type;
     index = _index;
     lineAddr = _lineAddr;
     from = new St[childs];
@@ -145,7 +151,7 @@ public:
 
 FromP* toCs2fromP(ToCs* toCs, Child c) {
   St from = toCs->from == NULL? 0: toCs->from[c];
-  FromP* ret = new FromP(toCs->isReq, toCs->index, toCs->lineAddr, from, toCs->to);
+  FromP* ret = new FromP(toCs->type, toCs->index, toCs->lineAddr, from, toCs->to);
   return ret;
 }
 

### 2012-10-28 impl/DataFeeder.h

@@ -59,6 +59,8 @@ public:
           count--;
         return;
       }
+      if(to == 2)
+        to = 3;
       feed(to);
     }
   }

### 2012-10-28 impl/Debug.h

@@ -3,11 +3,11 @@
 #ifdef Debug
 #define printSendRespToP(lineAddr, trigger, index, to)\
   printf("%p SendRespToP: %llx %c %d.%d %d->%d\n",\
-         this, lineAddr, trigger == Forced? 'F': 'V', index.set, index.way, cache.st[index.set][index.way], to);
+         this, lineAddr, trigger == Forced? 'F': 'V', index.set, index.way, cache.st[index.set][index.way], to)
 
 #define printSendRespToC(c, lineAddr, index, to)\
   printf("%p SendRespToC: %d %llx %d.%d %d->%d\n",\
-         this, c, lineAddr, index.set, index.way, cache.cstates[index.set][index.way][c], to);
+         this, c, lineAddr, index.set, index.way, cache.cstates[index.set][index.way][c], to)
 
 #define printSendReqToCs(cs, lineAddr, index, to){\
   printf("%p SendReqToCs: ", this);\
@@ -19,39 +19,44 @@
 
 #define printSendReqToP(lineAddr, index, to)\
   printf("%p SendReqToP: %llx %d.%d %d->%d\n",\
-         this, lineAddr, index.set, index.way, cache.st[index.set][index.way], to);
+         this, lineAddr, index.set, index.way, cache.st[index.set][index.way], to)
+
+#define printMemorySendResp(lineAddr, cIndex)\
+  printf("%p MemorySendResp: %llx %d.%x\n",\
+         this, lineAddr, cIndex.set, cIndex.way);
 
 #define printHandleRespFromC(c, lineAddr, trigger, index, to)\
-  printf("%p  HandleRespFromC: %d %llx %c %d.%d %d\n",\
-         this, c, lineAddr, trigger == Forced? 'F': 'V', index.set, index.way, to);
+  printf("%p HandleRespFromC: %d %llx %c %d.%d %d\n",\
+         this, c, lineAddr, trigger == Forced? 'F': 'V', index.set, index.way, to)
 
 #define printHandleRespFromP(lineAddr, index, to)\
-  printf("%p  HandleRespFromP: %llx %d.%d %d\n",\
-         this, lineAddr, index.set, index.way, to);
+  printf("%p HandleRespFromP: %llx %d.%d %d\n",\
+         this, lineAddr, index.set, index.way, to)
 
-#define printHandleReqFromC(c, lineAddr, to)\
-  printf("%p  HandleReqFromC: %d %llx %d\n",\
[…]

### 2012-10-28 impl/InternalCtrl.h

@@ -45,12 +45,13 @@ private:
   Mshr* mshr;
 
   St compat(St to) {
-    if(to == 2)
+    assert(to != 2);
+    if(to == 3)
       return 0;
     if(to == 1)
       return 1;
     if(to == 0)
-      return 2;
+      return 3;
   }
 
   bool isCompat(Child c, Index index, St to) {
@@ -81,7 +82,7 @@ private:
 
   void sendRespToC(Index& index, St to, Child c, Index& cIndex, LineAddr lineAddr, Latency lat) {
     printSendRespToC(c, lineAddr, index, to);
-    ToCs* resp = new ToCs(childs, c, false, cIndex, lineAddr, cache.cstates[index.set][index.way][c], to);
+    ToCs* resp = new ToCs(childs, c, Resp, cIndex, lineAddr, cache.cstates[index.set][index.way][c], to);
     latToCs = cache.cstates[index.set][index.way][c] == 0? dataLat: lat;
     cache.cstates[index.set][index.way][c] = to;
     toCs.enq(resp);
@@ -95,7 +96,7 @@ private:
     printSendReqToCs(highChildren, lineAddr, index, to);
     cache.cReq[index.set][index.way] = true;
     cache.waitCs[index.set][index.way] = to;
-    ToCs* req = new ToCs(childs, highChildren, true, index, lineAddr, NULL, to);
+    ToCs* req = new ToCs(childs, highChildren, Req, index, lineAddr, NULL, to);
     toCs.enq(req);
     latToCs = tagLat;
   }
@@ -148,7 +149,7 @@ private:
       } else if(m.isReplacing) {
         if(!isCHigher(index, 0)) {
           cache.cReq[index.set][index.way] = false;
-          sendRespToP(index, (St)0, Voluntary, index, msg->lineAddr, msg->trigger == Forced? (Latency)1: tagLat);
+          sendRespToP(index, (St)0, Voluntary, index, msg->lineAddr, msg->trigger == Forced? 1: tagLat);
           resetLine(index, m.lineAddr);
           sendReqToP(index, m.lineAddr, m.to);
           m.isReplacing = false;
@@ -173,7 +174,7 @@ private:
     if(fromP.empty())
       return false;
     FromP* msg = (FromP*) fromP.first();
-    if(msg->isReq)
+    if(msg->type != Resp)
        return false;
     printHandleRespFromP(msg->lineAddr, msg->index, msg->to);
     Index index = msg->index;
@@ -194,8 +195,8 @@ private:
[…]

### 2012-10-28 impl/L1Normal.h

@@ -44,9 +44,9 @@ private:
     if(fromP.empty())
       return false;
     FromP* msg = (FromP*) fromP.first();
-    printHandleRespFromP(msg->lineAddr, msg->index, msg->to);
-    if(msg->isReq)
+    if(msg->type != Resp)
        return false;
+    printHandleRespFromP(msg->lineAddr, msg->index, msg->to);
     Index index = msg->index;
     cache.st[index.set][index.way] = msg->to;
     cache.pReq[index.set][index.way] = false;
@@ -119,10 +119,10 @@ private:
     if(fromP.empty())
       return false;
     FromP* msg = (FromP*) fromP.first();
-    if(!msg->isReq)
+    if(msg->type != Req)
        return false;
-    printHandleReqFromP(msg->lineAddr, msg->to);
     bool present = cache.isPresent(msg->lineAddr);
+    printHandleReqFromP(msg->lineAddr, msg->to, present);
     if(!present) {
       fromP.deq();
       return true;

### 2012-10-28 impl/Memory.h

@@ -2,6 +2,7 @@
 
 #include "Fifo.h"
 #include "CacheTypes.h"
+#include "Debug.h"
 
 typedef class memory {
 private:
@@ -13,18 +14,20 @@ private:
 
   Fifo respFromP;
 
-  void sendCResp(Index& cIndex, LineAddr lineAddr) {
-    FromP* resp = new FromP(false, cIndex, lineAddr, 0, 2);
+  void sendRespToC(Index& cIndex, LineAddr lineAddr) {
+    FromP* resp = new FromP(Resp, cIndex, lineAddr, 0, 3);
     latRespFromP = latency;
     respFromP.enq(resp);
+    printMemorySendResp(lineAddr, cIndex);
   }
 
-  bool handleCReq() {
+  bool handleReqFromC() {
     if(reqToP->empty())
       return false;
     ReqToP* msg = (ReqToP*) reqToP->first();
-    sendCResp(msg->index, msg->lineAddr);
+    sendRespToC(msg->index, msg->lineAddr);
     reqToP->deq();
+    printHandleReqFromC(0, msg->lineAddr, msg->to, true);
     delete msg;
     return true;
   }
@@ -55,7 +58,7 @@ public:
       respToP->deq();
       delete msg;
     } else {
-      handleCReq();
+      handleReqFromC();
     }
   }
 

### 2012-10-28 impl/System_4_8.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 4;
+U8 levels = 1;
+Child childs[] = {8};
+MshrPtr mshrs[] = {8};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-28 impl/CacheTypes.h
stuff

@@ -17,7 +17,7 @@ typedef struct {
   Way way;
 } Index;
 
-typedef enum {Req, Resp, FwdReq, FwdResp} FromPType;
+typedef enum {Req, Resp, FwdReq, FwdResp, PAck} FromPType;
 
 typedef class fromP {
 public:
@@ -27,13 +27,15 @@ public:
   St from;
   St to;
 
-  bool isFwdReq;
-  bool changeState;
   Child fwdC;
   St fwdTo;
 
   fromP(FromPType _type, Index _index, LineAddr _lineAddr, St _from, St _to):
-       type(_type), index(_index), lineAddr(_lineAddr), from(_from), to(_to), isFwdReq(false) {}
+        type(_type), index(_index), lineAddr(_lineAddr), from(_from), to(_to) {}
+  fromP(FromPType _type, Index _index, LineAddr _lineAddr, St _to, Child _fwdC, St _fwdTo):
+        type(_type), index(_index), lineAddr(_lineAddr), to(_to), fwdC(_fwdC), fwdTo(_fwdTo) {}
+  fromP(FromPType _type, Index _index, LineAddr _lineAddr):
+        type(_type), index(_index), lineAddr(_lineAddr) {}
   ~fromP() {}
 } FromP;
 
@@ -44,13 +46,12 @@ public:
   St from;
   St to;
 
-  reqToP(Index _index, LineAddr _lineAddr, St _from, St _to) {
-    index = _index; lineAddr = _lineAddr; from = _from; to = _to;
-  }
+  reqToP(Index _index, LineAddr _lineAddr, St _from, St _to):
+         index(_index), lineAddr(_lineAddr), from(_from), to(_to) {}
   ~reqToP() {}
 } ReqToP;
 
-typedef enum {Forced, Voluntary} Trigger;
+typedef enum {Forced, Voluntary, CAck} Trigger;
 
 typedef class respToP {
 public:
@@ -60,9 +61,10 @@ public:
   St to;
   bool dirty;
 
-  respToP(Trigger _trigger, Index _index, LineAddr _lineAddr, St _to, bool _dirty) {
-    trigger = _trigger; index = _index; lineAddr = _lineAddr; to = _to; dirty = _dirty;
-  }
+  respToP(Trigger _trigger, Index _index, LineAddr _lineAddr, St _to, bool _dirty):
[…]

### 2012-10-28 impl/Connect.h

@@ -26,11 +26,6 @@ public:
   }
 
   void cycle() {
-//    printf("%p nextCReq: %d ", this, nextCReq);
-//    for(Child i = 0; i < childs; i++) {
-//      printf("%d", csReqToP[i]->empty());
-//    }
-//    printf("\n");
     if(!pReqFromC->full()) {
       for(Child i = 0; i < childs; i++) {
         Child c = (nextCReq + i)%childs;

### 2012-10-28 impl/CacheTypes.h
stuff

@@ -30,12 +30,9 @@ public:
   Child fwdC;
   St fwdTo;
 
-  fromP(FromPType _type, Index _index, LineAddr _lineAddr, St _from, St _to):
-        type(_type), index(_index), lineAddr(_lineAddr), from(_from), to(_to) {}
-  fromP(FromPType _type, Index _index, LineAddr _lineAddr, St _to, Child _fwdC, St _fwdTo):
-        type(_type), index(_index), lineAddr(_lineAddr), to(_to), fwdC(_fwdC), fwdTo(_fwdTo) {}
-  fromP(FromPType _type, Index _index, LineAddr _lineAddr):
-        type(_type), index(_index), lineAddr(_lineAddr) {}
+  fromP(FromPType _type, Index _index, LineAddr _lineAddr, St _from, St _to,
+        Child _fwdC, St _fwdTo): type(_type), index(_index), lineAddr(_lineAddr),
+        from(_from), to(_to), fwdC(_fwdC), fwdTo(_fwdTo) {}
   ~fromP() {}
 } FromP;
 
@@ -78,15 +75,17 @@ public:
   LineAddr lineAddr;
   St* from;
   St to;
+  Child fwdC;
+  St fwdTo;
 
   toCs(Child _childs, bool* _children, FromPType _type, Index _index,
-       LineAddr _lineAddr, St* _from, St _to):
+       LineAddr _lineAddr, St* _from, St _to, Child _fwdC, St _fwdTo):
        childs(_childs), children(_children), type(_type), index(_index),
-       lineAddr(_lineAddr), from(_from), to(_to) {}
+       lineAddr(_lineAddr), from(_from), to(_to), fwdC(_fwdC), fwdTo(_fwdTo) {}
   toCs(Child _childs, Child c, FromPType _type, Index _index,
-       LineAddr _lineAddr, St _from, St _to):
+       LineAddr _lineAddr, St _from, St _to, Child _fwdC, St _fwdTo):
        childs(_childs), type(_type), index(_index), lineAddr(_lineAddr),
-       to(_to) {
+       to(_to), fwdC(_fwdC), fwdTo(_fwdTo) {
     children = new bool[childs];
     for(Child i = 0; i < childs; i++)
       children[i] = false;
@@ -147,7 +146,7 @@ public:
 
 FromP* toCs2fromP(ToCs* toCs, Child c) {
   St from = toCs->from == NULL? 0: toCs->from[c];
[…]

### 2012-10-28 impl/InternalCtrl.h

@@ -82,7 +82,7 @@ private:
 
   void sendRespToC(Index& index, St to, Child c, Index& cIndex, LineAddr lineAddr, Latency lat) {
     printSendRespToC(c, lineAddr, index, to);
-    ToCs* resp = new ToCs(childs, c, Resp, cIndex, lineAddr, cache.cstates[index.set][index.way][c], to);
+    ToCs* resp = new ToCs(childs, c, Resp, cIndex, lineAddr, cache.cstates[index.set][index.way][c], to, 0, 0);
     latToCs = cache.cstates[index.set][index.way][c] == 0? dataLat: lat;
     cache.cstates[index.set][index.way][c] = to;
     toCs.enq(resp);
@@ -96,7 +96,7 @@ private:
     printSendReqToCs(highChildren, lineAddr, index, to);
     cache.cReq[index.set][index.way] = true;
     cache.waitCs[index.set][index.way] = to;
-    ToCs* req = new ToCs(childs, highChildren, Req, index, lineAddr, NULL, to);
+    ToCs* req = new ToCs(childs, highChildren, Req, index, lineAddr, NULL, to, 0, 0);
     toCs.enq(req);
     latToCs = tagLat;
   }

### 2012-10-28 impl/Memory.h

@@ -15,7 +15,7 @@ private:
   Fifo respFromP;
 
   void sendRespToC(Index& cIndex, LineAddr lineAddr) {
-    FromP* resp = new FromP(Resp, cIndex, lineAddr, 0, 3);
+    FromP* resp = new FromP(Resp, cIndex, lineAddr, 0, 3, 0, 0);
     latRespFromP = latency;
     respFromP.enq(resp);
     printMemorySendResp(lineAddr, cIndex);

### 2012-10-28 impl/CacheTypes.h
stuff

@@ -57,12 +57,11 @@ public:
   LineAddr lineAddr;
   St to;
   bool dirty;
+  bool fwd;
+  Child c;
 
-  respToP(Trigger _trigger, Index _index, LineAddr _lineAddr, St _to, bool _dirty):
-          trigger(_trigger), index(_index), lineAddr(_lineAddr), to(_to), dirty(_dirty) {}
-  respToP(Trigger _trigger, Index _index, LineAddr _lineAddr):
-          trigger(_trigger), index(_index), lineAddr(_lineAddr) {}
-
+  respToP(Trigger _trigger, Index _index, LineAddr _lineAddr, St _to, bool _dirty, bool _fwd, Child _c):
+          trigger(_trigger), index(_index), lineAddr(_lineAddr), to(_to), dirty(_dirty), fwd(_fwd), c(_c) {}
   ~respToP() {}
 } RespToP;
 
@@ -156,6 +155,7 @@ ReqFromC* reqToP2reqFromC(ReqToP* reqToP, Child c) {
 }
 
 RespFromC* respToP2respFromC(RespToP* respToP, Child c) {
-  RespFromC* ret = new RespFromC(c, respToP->trigger, respToP->index, respToP->lineAddr, respToP->to, respToP->dirty);
+  RespFromC* ret = new RespFromC(respToP->fwd? respToP->c: c, respToP->trigger,
+                                 respToP->index, respToP->lineAddr, respToP->to, respToP->dirty);
   return ret;
 }

### 2012-10-28 impl/InternalCtrl.h

@@ -73,7 +73,7 @@ private:
   void sendRespToP(Index& index, St to, Trigger trigger, Index& pIndex, LineAddr lineAddr, Latency lat) {
     printSendRespToP(lineAddr, trigger, index, to);
     cache.st[index.set][index.way] = to;
-    RespToP* resp = new RespToP(trigger, pIndex, lineAddr, to, cache.dirty[index.set][index.way]);
+    RespToP* resp = new RespToP(trigger, pIndex, lineAddr, to, cache.dirty[index.set][index.way], false, 0);
     respToP.enq(resp);
     latPResp = cache.dirty[index.set][index.way]? dataLat : lat;
     if(to == 0)

### 2012-10-28 impl/L1Normal.h

@@ -17,7 +17,7 @@ private:
   void sendRespToP(Index& index, St to, Trigger trigger, Index& pIndex, LineAddr lineAddr) {
     printSendRespToP(lineAddr, trigger, index, to);
     cache.st[index.set][index.way] = to;
-    RespToP* resp = new RespToP(trigger, pIndex, lineAddr, to, cache.dirty[index.set][index.way]);
+    RespToP* resp = new RespToP(trigger, pIndex, lineAddr, to, cache.dirty[index.set][index.way], false, 0);
     respToP.enq(resp);
     if(to == 0)
       cache.replaceRem(index);

### 2012-10-29 impl/SystemNormal.h
Stuff

@@ -154,16 +154,22 @@ public:
     printf("Cycle count: %llu\n", cycCount);
     for(U8 i = 0; i < levels; i++) {
       for(ThreadId j = 0; j < numCtrls[i]; j++) {
-        printf("Level %u  Cache# %u: hits %llu PermMiss: %llu InclusiveMiss: %llu PresenceMiss: %llu\n",
+        printf("Level %u  Cache# %u: hits: %llu PermMiss: %llu InclusiveMiss: %llu PresenceMiss: %llu\n",
                 i, j, ctrls[i][j]->hit, ctrls[i][j]->noPermMiss, ctrls[i][j]->inclusiveMiss, ctrls[i][j]->notPresentMiss);
+        printf("             : reqToCC: %llu respToCC: %llu respToCDataC: %llu reqToPC: %llu respToPC: %llu respToPDataC: %llu\n",
+                ctrls[i][j]->reqToCC, ctrls[i][j]->respToCC, ctrls[i][j]->respToCDataC, ctrls[i][j]->reqToPC, ctrls[i][j]->respToPC, ctrls[i][j]->respToPDataC);
       }
     }
 
     for(ThreadId j = 0; j < cores; j++) {
       printf("L1 Inst Cache# %u: hits %llu PermMiss: %llu InclusiveMiss: %llu PresenceMiss: %llu\n",
               2*j, l1s[2*j]->hit, l1s[2*j]->noPermMiss, l1s[2*j]->inclusiveMiss, l1s[2*j]->notPresentMiss);
+      printf("             : reqToPC: %llu respToPC: %llu respToPDataC: %llu\n",
+              l1s[2*j]->reqToPC, l1s[2*j]->respToPC, l1s[2*j]->respToPDataC);
       printf("L1 Data Cache# %u: hits %llu PermMiss: %llu InclusiveMiss: %llu PresenceMiss: %llu\n",
               2*j+1, l1s[2*j+1]->hit, l1s[2*j+1]->noPermMiss, l1s[2*j+1]->inclusiveMiss, l1s[2*j+1]->notPresentMiss);
+      printf("             : reqToPC: %llu respToPC: %llu respToPDataC: %llu\n",
+              l1s[2*j+1]->reqToPC, l1s[2*j+1]->respToPC, l1s[2*j+1]->respToPDataC);
     }
   }
 

### 2012-10-29 impl/SystemNormal_1024_2048.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 1024;
+U8 levels = 1;
+Child childs[] = {2048};
+MshrPtr mshrs[] = {64};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-29 impl/SystemNormal_128_256.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 128;
+U8 levels = 1;
+Child childs[] = {256};
+MshrPtr mshrs[] = {64};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-29 impl/SystemNormal_16_32.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 16;
+U8 levels = 1;
+Child childs[] = {32};
+MshrPtr mshrs[] = {32};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-29 impl/SystemNormal_256_512.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 256;
+U8 levels = 1;
+Child childs[] = {512};
+MshrPtr mshrs[] = {64};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-29 impl/SystemNormal_32_64.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 32;
+U8 levels = 1;
+Child childs[] = {64};
+MshrPtr mshrs[] = {64};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-29 impl/SystemNormal_512_1024.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 512;
+U8 levels = 1;
+Child childs[] = {1024};
+MshrPtr mshrs[] = {64};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-29 impl/SystemNormal_64_128.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 64;
+U8 levels = 1;
+Child childs[] = {128};
+MshrPtr mshrs[] = {64};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-29 impl/SystemNormal_8_16.cpp

@@ -1,0 +1,20 @@
+#include "SystemNormal.h"
+
+using namespace std;
+
+ThreadId cores = 8;
+U8 levels = 1;
+Child childs[] = {16};
+MshrPtr mshrs[] = {16};
+Way ways[] = {8, 8};
+U8 setSzs[] = {6, 8};
+Latency tagLats[] = {4};
+Latency dataLats[] = {10};
+Latency memLat = 50;
+
+int main() {
+
+  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
+  sys.run();
+  sys.display();
+}

### 2012-10-29 impl/CacheTypes.h
stuff

@@ -48,7 +48,7 @@ public:
   ~reqToP() {}
 } ReqToP;
 
-typedef enum {Forced, Voluntary, CAck} Trigger;
+typedef enum {Forced, Voluntary, CAck, CNack} Trigger;
 
 typedef class respToP {
 public:
@@ -119,9 +119,10 @@ public:
   LineAddr lineAddr;
   St to;
   bool dirty;
+  bool fwd;
 
-  respFromC(Child _c, Trigger _trigger, Index _index, LineAddr _lineAddr, St _to, bool _dirty):
-            c(_c), trigger(_trigger), index(_index), lineAddr(_lineAddr), to(_to), dirty(_dirty) {}
+  respFromC(Child _c, Trigger _trigger, Index _index, LineAddr _lineAddr, St _to, bool _dirty, bool _fwd):
+            c(_c), trigger(_trigger), index(_index), lineAddr(_lineAddr), to(_to), dirty(_dirty), fwd(_fwd) {}
 
   ~respFromC() {}
 } RespFromC;
@@ -156,6 +157,6 @@ ReqFromC* reqToP2reqFromC(ReqToP* reqToP, Child c) {
 
 RespFromC* respToP2respFromC(RespToP* respToP, Child c) {
   RespFromC* ret = new RespFromC(respToP->fwd? respToP->c: c, respToP->trigger,
-                                 respToP->index, respToP->lineAddr, respToP->to, respToP->dirty);
+                                 respToP->index, respToP->lineAddr, respToP->to, respToP->dirty, respToP->fwd);
   return ret;
 }

### 2012-10-29 impl/Debug.h

@@ -1,5 +1,7 @@
 #pragma once
 
+#include <cassert>
+
 #ifdef Debug
 #define printSendRespToP(lineAddr, trigger, index, to)\
   printf("%p SendRespToP: %llx %c %d.%d %d->%d\n",\
@@ -47,6 +49,8 @@
 
 #define printCycle printf("\nCYCLE %lld\n", cycCount)
 
+#define Assert(c) assert(c)
+
 #else
 #define printSendRespToP(lineAddr, trigger, index, to)
 #define printSendRespToC(c, lineAddr, index, to)
@@ -59,4 +63,5 @@
 #define printHandleReqFromP(lineAddr, to, present)
 #define printHandleReqFromCore(lineAddr, to)
 #define printCycle
+#define Assert(c)
 #endif

### 2012-10-29 impl/Fifo.h

@@ -1,10 +1,8 @@
 #pragma once
 
 #include "Types.h"
+#include "Debug.h"
 #include <queue>
-
-#include <cassert>
-#include <list>
 
 using namespace std;
 
@@ -26,12 +24,12 @@ public:
     return numElems == 0;
   }
   void enq(void* x) {
-    assert (numElems < size);
+    Assert (numElems < size);
     q.push(x);
     enqEn = true;
   }
   void deq() {
-    assert (numElems > 0);
+    Assert (numElems > 0);
     q.pop();
     deqEn = true;
   }

### 2012-10-29 impl/InternalCtrl.h

@@ -29,6 +29,7 @@ public:
 typedef class internalCtrl {
 private:
   U8 setSz;
+  Way ways;
   Child childs;
   Cache cache;
   FreeList mshrFl;
@@ -45,7 +46,7 @@ private:
   Mshr* mshr;
 
   St compat(St to) {
-    assert(to != 2);
+    Assert(to != 2);
     if(to == 3)
       return 0;
     if(to == 1)
@@ -75,7 +76,13 @@ private:
     cache.st[index.set][index.way] = to;
     RespToP* resp = new RespToP(trigger, pIndex, lineAddr, to, cache.dirty[index.set][index.way], false, 0);
     respToP.enq(resp);
-    latPResp = cache.dirty[index.set][index.way]? dataLat : lat;
+    if(cache.dirty[index.set][index.way]) {
+      latPResp = dataLat;
+      respToPDataC++;
+    } else {
+      latPResp = lat;
+      respToPC++;
+    }
     if(to == 0)
       cache.replaceRem(index);
   }
@@ -83,7 +90,13 @@ private:
   void sendRespToC(Index& index, St to, Child c, Index& cIndex, LineAddr lineAddr, Latency lat) {
     printSendRespToC(c, lineAddr, index, to);
     ToCs* resp = new ToCs(childs, c, Resp, cIndex, lineAddr, cache.cstates[index.set][index.way][c], to, 0, 0);
-    latToCs = cache.cstates[index.set][index.way][c] == 0? dataLat: lat;
+    if(cache.cstates[index.set][index.way][c] == 0) {
+      latToCs = dataLat;
+      respToCDataC++;
+    } else {
+      latToCs = lat;
+      respToCC++;
+    }
     cache.cstates[index.set][index.way][c] = to;
     toCs.enq(resp);
     cache.replaceUpd(index);
@@ -91,8 +104,14 @@ private:
 
   void sendReqToCs(Index& index, LineAddr lineAddr, St to, bool skip, Child skipChild) {
     bool* highChildren = new bool[childs];
-    for(Child i = 0; i < childs; i++)
-       highChildren[i] = (!skip || i != skipChild) && cache.cstates[index.set][index.way][i] > to;
+    for(Child i = 0; i < childs; i++) {
+      if((!skip || i != skipChild) && cache.cstates[index.set][index.way][i] > to) {
+        highChildren[i] = true;
+        reqToCC++;
+      }
+      else
[…]

### 2012-10-29 impl/L1Normal.h

@@ -19,6 +19,10 @@ private:
     cache.st[index.set][index.way] = to;
     RespToP* resp = new RespToP(trigger, pIndex, lineAddr, to, cache.dirty[index.set][index.way], false, 0);
     respToP.enq(resp);
+    if(cache.dirty[index.set][index.way])
+      respToPDataC++;
+    else
+      respToPC++;
     if(to == 0)
       cache.replaceRem(index);
   }
@@ -28,6 +32,7 @@ private:
     processing = true;
     cache.pReq[index.set][index.way] = true;
     cache.waitS[index.set][index.way] = to;
+    reqToPC++;
     ReqToP* req = new ReqToP(index, lineAddr, cache.st[index.set][index.way], to);
     reqToP.enq(req);
   }
@@ -100,6 +105,8 @@ private:
         reqFromCore.deq();
         cache.replaceUpd(index);
         hit++;
+        if(msg->to == 3)
+           cache.dirty[index.set][index.way] = true;
         delete msg;
         return true;
       }
@@ -144,13 +151,15 @@ public:
   Counter notPresentMiss;
   Counter noPermMiss;
   Counter inclusiveMiss;
+  Counter reqToPC, respToPC, respToPDataC;
 
   l1Normal(Way ways, U8 _setSz):
           setSz(_setSz), cache(ways, setSz, 0),
           fromP(2), reqFromCore(2),
           reqToP(2), respToP(2),
           priority(C), processing(false),
-          hit(0), notPresentMiss(0), noPermMiss(0), inclusiveMiss(0) {}
+          hit(0), notPresentMiss(0), noPermMiss(0), inclusiveMiss(0),
+          reqToPC(0), respToPC(0), respToPDataC(0) {}
 
   ~l1Normal() {}
 

### 2012-10-29 impl/System_1_1.cpp

@@ -1,20 +1,0 @@
-#include "SystemNormal.h"
-
-using namespace std;
-
-ThreadId cores = 1;
-U8 levels = 1;
-Child childs[] = {2};
-MshrPtr mshrs[] = {64};
-Way ways[] = {8, 16};
-U8 setSzs[] = {6, 8};
-Latency tagLats[] = {5};
-Latency dataLats[] = {10};
-Latency memLat = 40;
-
-int main() {
-
-  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
-  sys.run();
-  sys.display();
-}

### 2012-10-29 impl/System_2_1_1.cpp

@@ -1,20 +1,0 @@
-#include "SystemNormal.h"
-
-using namespace std;
-
-ThreadId cores = 2;
-U8 levels = 2;
-Child childs[] = {2, 2};
-MshrPtr mshrs[] = {32, 64};
-Way ways[] = {8, 16, 16};
-U8 setSzs[] = {6, 8, 10};
-Latency tagLats[] = {5, 10};
-Latency dataLats[] = {5, 10};
-Latency memLat = 40;
-
-int main() {
-
-  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
-  sys.run();
-  sys.display();
-}

### 2012-10-29 impl/System_2_2.cpp

@@ -1,20 +1,0 @@
-#include "SystemNormal.h"
-
-using namespace std;
-
-ThreadId cores = 2;
-U8 levels = 1;
-Child childs[] = {4};
-MshrPtr mshrs[] = {64};
-Way ways[] = {8, 16};
-U8 setSzs[] = {6, 8};
-Latency tagLats[] = {5};
-Latency dataLats[] = {10};
-Latency memLat = 40;
-
-int main() {
-
-  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
-  sys.run();
-  sys.display();
-}

### 2012-10-29 impl/System_4_8.cpp

@@ -1,20 +1,0 @@
-#include "SystemNormal.h"
-
-using namespace std;
-
-ThreadId cores = 4;
-U8 levels = 1;
-Child childs[] = {8};
-MshrPtr mshrs[] = {8};
-Way ways[] = {8, 8};
-U8 setSzs[] = {6, 8};
-Latency tagLats[] = {4};
-Latency dataLats[] = {10};
-Latency memLat = 50;
-
-int main() {
-
-  SystemNormal sys(cores, levels, childs, mshrs, ways, setSzs, tagLats, dataLats, memLat);
-  sys.run();
-  sys.display();
-}

### 2012-10-29 impl/run.sh
stuff

@@ -1,0 +1,10 @@
+#!/bin/sh
+
+dirs="BT CG DC FT IS LU MG SP"
+export OMP_NUM_THREADS=$1
+for dir in $dirs; do
+  cd $dir
+  ../a.out > $2 &
+  cd ..
+done
+

### 2012-10-30 impl/CacheTypes.h
stuff

@@ -145,8 +145,7 @@ public:
 } ReqFromCore;
 
 FromP* toCs2fromP(ToCs* toCs, Child c) {
-  St from = toCs->from == NULL? 0: toCs->from[c];
-  FromP* ret = new FromP(toCs->type, toCs->index, toCs->lineAddr, from, toCs->to, toCs->fwdC, toCs->fwdTo);
+  FromP* ret = new FromP(toCs->type, toCs->index, toCs->lineAddr, toCs->from[c], toCs->to, toCs->fwdC, toCs->fwdTo);
   return ret;
 }
 

### 2012-10-30 impl/L1M.h

@@ -1,0 +1,195 @@
+#pragma once
+
+#include "Cache.h"
+#include "Fifo.h"
+#include "CacheTypes.h"
+#include "Debug.h"
+
+typedef class l1M {
+private:
+  U8 setSz;
+  Cache cache;
+  Fifo fromP, reqFromCore, reqToP, respToP;
+  Fifo respFromS, respToS;
+
+  bool processing;
+  Who priority;
+
+  void sendRespToP(Index& index, St to, Trigger trigger, Index& pIndex, LineAddr lineAddr) {
+    printSendRespToP(lineAddr, trigger, index, to);
+    cache.st[index.set][index.way] = to;
+    RespToP* resp = new RespToP(trigger, pIndex, lineAddr, to, cache.dirty[index.set][index.way], false, 0);
+    respToP.enq(resp);
+    if(cache.dirty[index.set][index.way])
+      respToPDataC++;
+    else
+      respToPC++;
+    if(to == 0)
+      cache.replaceRem(index);
+  }
+
+  void sendReqToP(Index& index, LineAddr lineAddr, St to) {
+    printSendReqToP(lineAddr, index, to);
+    processing = true;
+    cache.pReq[index.set][index.way] = true;
+    cache.waitS[index.set][index.way] = to;
+    reqToPC++;
+    ReqToP* req = new ReqToP(index, lineAddr, cache.st[index.set][index.way], to);
+    reqToP.enq(req);
+  }
+
+  void resetLine(Index& index, LineAddr lineAddr) {
+    cache.pReq[index.set][index.way] = false;
+    cache.cReq[index.set][index.way] = false;
+    cache.tag[index.set][index.way] = lineAddr >> setSz;
+    cache.st[index.set][index.way] = 0;
+    cache.dirty[index.set][index.way] = false;
+  }
+
+  bool handleRespFromP() {
+    if(fromP.empty())
+      return false;
+    FromP* msg = (FromP*) fromP.first();
+    if(msg->type != Resp)
+       return false;
+    printHandleRespFromP(msg->lineAddr, msg->index, msg->to);
+    Index index = msg->index;
+    cache.st[index.set][index.way] = msg->to;
+    cache.pReq[index.set][index.way] = false;
+    fromP.deq();
+    delete msg;
+    cache.replaceUpd(index);
+    ReqFromCore* m = (ReqFromCore*) reqFromCore.first();
+    reqFromCore.deq();
+    processing = false;
+    delete m;
+    return true;
+  }
+
[…]

### 2012-10-30 impl/L2M.h

@@ -8,7 +8,7 @@
 
 #include "Debug.h"
 
-typedef class internalCtrl {
+typedef class l2M {
 private:
   U8 setSz;
   Way ways;
@@ -84,14 +84,17 @@ private:
     cache.replaceUpd(index);
   }
 
-  void sendReqToCs(Index& index, LineAddr lineAddr, St to, bool skip, Child skipChild) {
+  void sendReqToCs(Index& index, LineAddr lineAddr, St to, bool skip, Child skipChild, St childTo) {
     bool* highChildren = new bool[childs];
+    St highState = 0;
     St* cstates = new St[childs];
     for(Child i = 0; i < childs; i++) {
       cstates[i] = cache.cstates[index.set][index.way][i];
       if((!skip || i != skipChild) && cstates[i] > to) {
         highChildren[i] = true;
         reqToCC++;
+        if(highState < cstates[i])
+          highState = cstates[i];
       }
       else
         highChildren[i] = false;
@@ -99,7 +102,8 @@ private:
     printSendReqToCs(highChildren, lineAddr, index, to);
     cache.cReq[index.set][index.way] = true;
     cache.waitCs[index.set][index.way] = to;
-    ToCs* req = new ToCs(childs, highChildren, Req, index, lineAddr, cstates, to, 0, 0);
+    ToCs* req = new ToCs(childs, highChildren, skipChild && highState >= 2? FwdReq: Req,
+                         index, lineAddr, cstates, to, skipChild, childTo);
     toCs.enq(req);
     latToCs = tagLat;
   }
@@ -163,7 +167,11 @@ private:
         if(isCompat(m.c, index, m.to)) {
           cache.cReq[index.set][index.way] = false;
           if(!cache.pReq[index.set][index.way]) {
-            sendRespToC(index, m.to, m.c, m.index, m.lineAddr, msg->trigger == Forced? 1: tagLat);
+            if(msg->fwd) {
+              cache.cstates[index.set][index.way][m.c] = m.to;
+            } else {
+              sendRespToC(index, m.to, m.c, m.index, m.lineAddr, msg->trigger == Forced? 1: tagLat);
+            }
             mshrFl.free(mshrPtr);
           }
         }
@@ -240,7 +248,7 @@ private:
         return true;
       } else {
[…]

### 2012-10-30 impl/Mshr.h

@@ -1,0 +1,21 @@
+#include "CacheTypes.h"
+
+typedef class mshr {
+public:
+  Who who;
+  Child c;
+  Index index;
+  St to;
+  bool isReplacing;
+  LineAddr lineAddr;
+  bool isPrev;
+  MshrPtr prevMshr;
+
+  mshr() {}
+  mshr(Who _who, Child _c, Index _index, St _to,
+       bool _isReplacing, LineAddr _lineAddr, bool _isPrev, MshrPtr _prevMshr) :
+       who(_who), c(_c), index(_index), to(_to), isReplacing(_isReplacing),
+       lineAddr(_lineAddr), isPrev(_isPrev), prevMshr(_prevMshr) {}
+  ~mshr() {}
+} Mshr;
+

### 2012-10-30 impl/SystemNormal_1024_2048.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 1024;
 U8 levels = 1;
 Child childs[] = {2048};
 MshrPtr mshrs[] = {64};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 impl/SystemNormal_128_256.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 128;
 U8 levels = 1;
 Child childs[] = {256};
 MshrPtr mshrs[] = {64};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 impl/SystemNormal_16_32.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 16;
 U8 levels = 1;
 Child childs[] = {32};
 MshrPtr mshrs[] = {32};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 impl/SystemNormal_256_512.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 256;
 U8 levels = 1;
 Child childs[] = {512};
 MshrPtr mshrs[] = {64};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 impl/SystemNormal_32_64.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 32;
 U8 levels = 1;
 Child childs[] = {64};
 MshrPtr mshrs[] = {64};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 impl/SystemNormal_4_8.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 4;
 U8 levels = 1;
 Child childs[] = {8};
 MshrPtr mshrs[] = {8};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 impl/SystemNormal_512_1024.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 512;
 U8 levels = 1;
 Child childs[] = {1024};
 MshrPtr mshrs[] = {64};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 impl/SystemNormal_64_128.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 64;
 U8 levels = 1;
 Child childs[] = {128};
 MshrPtr mshrs[] = {64};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 impl/SystemNormal_8_16.cpp

@@ -6,7 +6,7 @@ ThreadId cores = 8;
 U8 levels = 1;
 Child childs[] = {16};
 MshrPtr mshrs[] = {16};
-Way ways[] = {8, 8};
+Way ways[] = {8, 16};
 U8 setSzs[] = {6, 8};
 Latency tagLats[] = {4};
 Latency dataLats[] = {10};

### 2012-10-30 paper/isca40format/Body.tex
added class files

\section{Introduction}

This document provides the formatting instructions for submissions to the 40th
Annual International Symposium on Computer Architecture,
2013. In an effort to respect the efforts of reviewers and in
the interest of fairness to all prospective authors, we request that all
submissions to ISCA-40 follow the formatting and submission rules detailed
below.  Submissions that violate these instructions may not be
reviewed, at the discretion of the program chair, in order to maintain a review
process that is fair to all potential authors.  This is a generous format,
with plenty of space -- there should be no need to tweak it in any
significant way.

An example submission (formatted using the ISCA-40 submission format) that
contains the submission and formatting guidelines can be downloaded from
the conference website at {\em http://isca2013.eew.technion.ac.il}.

All questions regarding paper formatting and submission should be directed to
the program chair.

\section{Preparation Instructions}

\subsection{Paper Formatting}

All submissions should contain a maximum of 12 pages of single-spaced
two-column text. If you are using \LaTeX~\cite{lamport94} 
to typeset your paper, then we 
strongly suggest
that you use the template available at
http://isca2013.eew.technion.ac.il -- this
document was prepared with that template.  
If you are using a different
software package to typeset your paper, then please adhere to the guidelines
given in Table~\ref{table:formatting}.

\begin{table}[h!]
  \centering
  \begin{tabular}{|l|l|}
    \hline
    \textbf{Field} & \textbf{Value}\\
    \hline
    \hline
    Page limit & 12 pages\\
    \hline
    Paper size & US Letter 8.5in $\times$ 11in\\
    \hline
    Top margin & 1in\\
    \hline
    Bottom margin & 1in\\
    \hline
    Left margin & 0.75in\\
    \hline
    Right margin & 0.75in\\
    \hline
    Separation between columns & 0.25in\\
    \hline
    Body font & 10pt\\
    \hline
    Abstract font & 10pt, italicized\\
    \hline
    Section heading font & 12pt, bold\\
    \hline
    Subsection heading font & 10pt, bold\\
    \hline
    Caption font & 9pt, bold\\
    \hline
    References & 8pt\\
    \hline
  \end{tabular}
  \caption{Formatting guidelines for submission. }
  \label{table:formatting}
\end{table}

\textbf{Please ensure that you include page numbers with your
submission}. This makes it easier for the reviewers to refer to
different parts of your paper when they provide comments.

Also, please ensure that your submission has a banner at the top of 
the title page, similar to this one, which contains the submission
number and the notice of confidentiality.  If using the template,
just replace XXX in the template with the submission number
you receive from the submission website.

If you use bibtex, please note that the references.bib file provided in the
template example includes some format-specific incantations at the top of the
file.  If you substitute your own bib file, you will probably want to include
these incantations at the top of it.

\subsection{Content}

\noindent\textbf{\sout{Author List.}} All submissions are double
blind. Therefore, please do not include any author names in the
submission. You must also ensure that the metadata included in the
PDF does not give away the authors. If you are improving upon your
prior work, refer to your prior work in the third person and include
a full citation for the work in the bibliography. For example, if
you are building on {\em your own} prior work in the papers \cite{nicepaper,nicepaper2}, 
you would say something like "While the authors of \cite{nicepaper,nicepaper2} did X and Y,
this paper does X, Y and Z, and is therefore much better."  Do NOT omit or
anonymize references for blind review.

\noindent\textbf{Figures and Tables.} Ensure that the figures and
tables are legible.  Please also ensure that you refer to your
figures in the main text. The proceedings will be printed in
gray-scale, and many reviewers print the papers in
gray-scale. Therefore, if you must use colors for your figures, ensure
that the different colors are highly distinguishable in gray-scale.
If a figure is not easily understandable in gray-scale, then assume
it will not be understood by the reviewers.  In many cases, it
is better to just prepare your documents without color.

\noindent\textbf{Main Body.} Avoid bad page or column breaks in
your main text, i.e., last line of a paragraph at the top of a
column or first line of a paragraph at the end of a column. If you
begin a new section or sub-section near the end of a column,
ensure that you have at least 2 lines of body text on the same
column.

\section{Submission Instructions}

\subsection{Paper Authors}

Declare all the authors of the paper upfront. Addition/removal of authors once
the paper is accepted will have to be approved by the program chair.  The
paper selection process is carefully run in a way that maximizes fairness
by seeking to eliminate all conflicts of interest.  Late changes to author 
lists can invalidate that process.

\subsection{Declaring Conflicts of Interest}

The authors must register all their conflicts into the paper submission site.
Conflicts are needed to resolve assignment of reviewers. 
Please get the conflicts
right.  You have a week between the registration of the paper and final
submission -- there is no need to do the conflicts in a rush at the last
second. If a paper is found to
have an undeclared conflict that causes a problem OR if a paper is found to declare false conflicts in order to abuse or ``game'' the review system, the paper may be rejected.

Please declare a conflict of interest (COI) if any of the following exist for any author of a paper:

\begin{enumerate}
\item Your Ph.D. advisor and Ph.D. students forever.
\item Family relations by blood or marriage forever (if they might be potential reviewers).
\item Other past or current advisors
\item People with whom you collaborated in the last five years. Collaborators include:
\begin{itemize}
\item Co-authors on an accepted/rejected/pending research paper, 
\item Co-PIs on an accepted/rejected/pending grant, 
\item Those who are funders (decision-makers) regarding your research grants, and researchers whom you fund. 
\end{itemize}
You many exclude ``service'' collaborations like writing a CSTB report or serving on a program committee together.
\item People who shared your primary institution in the last five years. Note that if either you or they have moved, there could be several institutions to consider.
\end{enumerate}

There may be others with whom you know a conflict of interest exists.  However,
you will need to justify this conflict of interest.  Please be reasonable.  For
example, just because a reviewer works on similar topics as the paper you are
submitting is on, you cannot declare a conflict of interest with them. The PC Chair may contact co-authors to explain COIs whose origin is not clear.

You will have to declare all conflicts with PC members as well as non-PC members 
with whom you
have a conflict of interest.  When in doubt, contact the program chair.

\subsection{Concurrent Submissions and Resubmissions of Already Published Papers}

By submitting a manuscript to ISCA-40, the authors guarantee that the
manuscript has not been previously published or accepted for publication in a
substantially similar form in any conference or journal. The authors also
guarantee that no paper which contains significant overlap with the
contributions of the submitted paper is under review to any other conference or
journal or workshop, or will be submitted to one of them during the ISCA-40
review period. Violation of any of these conditions will lead to rejection.

Extended versions of papers accepted to IEEE Computer Architecture Letters can
be submitted to ISCA-19.  If you are in doubt, contact the program chair.

Finally, we also note that the ACM Plagiarism Policy ({\em http://www.acm.org/publications/policies/plagiarism\_policy}) covers a range of ethical issues concerning the misrepresentation of other works or one's own work.

### 2012-11-04 paper/isca40format/Introduction.tex
stuff before changes

\section{Introduction} With the increasing number of cores, bus-based
cache-coherence protocols have started being replaced by message-passing
directory-based protocols. Bus-based protocols allow only one cache-line
transaction to happen at a time, locking the bus till the transaction is
complete. This serializes memory accesses across each processor, even if they
are for different cache lines. On the other hand, message-passing,
directory-based protocols can handle multiple independent transactions in
parallel, thus scaling better with increasing core counts.

In a message-passing directory-based protocol, the coherence state changes are
orchestrated by passing messages in the form of upgrade or downgrade requests
and responses between the caches via the directory. Unlike bus-based protocols
which lock the bus till all the required state transitions are done thereby
ensuring atomicity of state transitions, in a message-passing protocol, requests
and responses are split leading to transient states. This creates scenarios not
present in bus-based protocols, which can lead to coherence violations or
deadlocks, if the protocol is not designed or implemented carefully.

For instance, consider a scenario in which a cache having read permission for
a line gets a store request from its core and hence requires write permissions
for that line. The cache would send an upgrade request to the directory
controller and may eventually get a response from the directory controller. The
response would signify that the permissions in the other caches \emph{were}
compatible with this cache having write permission for the line, \ie the other
caches neither \emph{had} read nor write permissions for that line. But the messages
take time to propagate. Thus, as emphasized, the response message is based on
the old permissions that the other caches had, not the permissions that the
other caches have at the moment the original cache gets the response. If an
other cache gets a read or write permission for that line during the time the
messages are in flight, coherence would be violated.

Consider a different scenario in which a cache, already having read permission
for a line, gets a store request, and hence sends an upgrade request for write
permission to the directory controller. Instead of getting a response, the cache
may receive a request from the directory controller asking this cache to give up
the read permission for the line \ie an invalidate request. This could happen if
the directory controller received a request for write permission from another
cache earlier. It is not easy to determine the correct behavior of the cache
receiving the invalidate request -- whether it should ignore the invalidate
request and wait till it receives a response from the directory controller for
its request, or if it should invalidate its line. If it ignores the invalidate
request, is there any guarantee that it would receive a response for its request
for write permission, or would it deadlock? As we will be showing later, this
could create a deadlock.

Consider yet another scenario in which a cache receives an invalidate request
from the directory controller for a line that is not present in the cache. Is
such a request even possible? The directory does not have access to the
up-to-date state of the cache, and so can potentially make this request. How
should a cache handle this request; should it drop such a request, or respond
saying that it does not have the line? Again, reasoning about these questions is
very difficult, as one has to take into account all the possible transitions
that can happen in the entire system.

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a message-passing directory-based setting.
Even for a 2-level cache hierarchy, a private L1 cache for each core and a
shared L2 cache, designing and implementing the protocol is very hard. This
problem is greatly exacerbated in a multi-level cache hierarchy, which is
becoming the norm because of the increasing core counts.

In this paper, we present a systematic design methodology to design hierarchical
messaging-passing, directory-based inclusive cache coherence protocols. We
restrict our methodology to inclusive protocols as we want our protocols to
scale to multiple levels of hierarchy. A non-inclusive or exclusive cache
coherence protocol requires probing every level of the hierarchy starting from
the root (LLC) to the leaf (L1 cache) to process any request for permission. We
believe that this would not scale as the number of cores and the number of
cache levels increase.

We present a set of transitions that the caches and directories must follow on
sending or receiving a request or a response. We have formally proved that
these behaviors will ensure the following properties: a) the directory's notion
of a cache's state is ``conservative'' in that a cache is never going to have
more permissions that what the directory thinks the cache has, and b) any
request will eventually get a corresponding response. It should be noted that
the formal proof was written using the Coq theorem prover \cite{}, and the
discussion of the formal proof is beyond the scope of this paper.

The flexibility of our methodology to design multiple cache coherence protocols
comes from the fact that we do not assign any meaning or semantics to the value
of a coherence state. All the transitions that we present are in the form of
upgrades or downgrades from state $x$ to state $y$, without giving concrete
values for $x$ or $y$. We can thus use our methodology to design multiple
hierarchical cache coherence protocols: MSI, MESI, MOSI, MOESI. We will show
how each of these protocols obey the set of transitions that we require for our
methodology, proving the correctness of all these protocols.

%\paragraph{Paper organization:} Section \ref{sec:setup} gives the setup for our
%methodology. It gives the state transitions of the cache and directory
%controllers on sending or receiving messages. Section \ref{sec:properties} gives
%the two properties that are guaranteed if the state transitions that we present
%are obeyed. We present the intuition for proving these properties starting from
%the state transitions that we present in \ref{sec:setup}. In Section
%\ref{sec:protocols}, we show how these properties can be used to implement a
%hierarchical MSI, MESI, MOSI and MOESI protocols. We also demonstrate how the
%correctness and deadlock freedom of each of the protocols follows straightaway
%from our methodology with no additional verification effort. Our claim is that
%our state transitions and the guaranteed properties can be used to formally
%prove any message-passing directory-based hierarchical inclusive cache coherence
%protocols by appropriate mapping. In Section \ref{sec:performance}, we show that
%these formally verified family of protocols require no extra state bits, nor
%extra message transfers compared to other published protocols. In Section
%\ref{sec:related}, we present some of the related work and finally in Section
%\ref{sec:conclusion}, we present the directions we would like to take this work
%in the future and conclude.

%We carried out the proof using the Coq theorem prover \cite{}. Unlike
%model-checking based techniques, the theorem prover merely checks the
%correctness of a supplied mathematical proof, thereby avoiding the state
%explosion problem. The proof is not restricted to a fixed number of cores,
%number of caches levels or number of caches in each level. Instead the proof can
%be parameterized for these quantities, thus ensuring that the same proof holds
%for any configuration, without any additional computational complexity. The
%downside of a theorem prover is that it shifts the onus of supplying a correct
%mathematical proof to the user, instead of searching through the state space to
%expose violations of specifications \ala model checking.

%$S$ and $I$ in an MSI protocol
%The state transitions that we require does not enforce a particular cache
%coherence protocol. It merely dictates how the states change on sending or
%receiving a response message, or after handling a request message. To be more
%concrete, consider a simple example. A directory controller receives a Write
%permission request for a line from a cache. If the directory thinks that some
%other cache has a read or write permission for that line, then a correct cache
%coherence protocol requires that the other cache invalidates that line before
%the directory sends a response to the first cache giving it Write permission.
%But this ``policy'' is not enforced by the state transitions that we present.
%Instead, if the directory decides to give Write permission to the original cache
%when it thinks another cache still has that line, then our state transitions
%will permit it. The system will no longer be coherent, but it will not violate
%the two properties that our state transitions guarantee (the directory would now
%think that the original cache has Write permission and some other caches also
%have some permissions for that line). This enables us to implement a wide range
%of correct cache coherence protocols with minimal proof burden.
%
%For example, it is straight forward to implement MSI or MESI protocols making
%use of the properties guaranteed in our methodology. If the directory controller
%receives a Write permission request from a cache for a line, then it sends
%requests to every cache which has that line. Our second property of getting a
%response for every request ensures that these caches would respond eventually,
%making their local states for that line Invalid. Our first property guarantees
%that when the directory controller thinks the state of a cache for a line is
%Invalid, then the actual states in the caches at that time should not be greater
%than Invalid. Thus, once all the other caches have responded, the directory
%controller can respond to the original cache. This way, the two properties can
%be used to implement MSI or MESI in a straightforward manner. We will present
%how this is done in greater detail in the paper.
%
%A more complicated protocol with cache intervention (like MOSI and MOESI)
%can also be implemented using the two properties that we guarantee, but it
%requires an additional step of mapping the protocol to our transitions (and
%proving that the mapping is correct). We will present this in greater detail in
%the paper.
%
%Finally, we will show that these formally verified family of protocols require
%no extra state bits, nor extra message transfers compared to other published
%protocols. In fact, our claim is that our state transitions and the two
%guaranteed properties can be used to formally prove any message-passing
%directory-based hierarchical inclusive cache coherence protocols by
%appropriately mapping the protocol to our transitions, as shown in the
%MOSI/MOESI mapping.

### 2012-11-04 paper/isca40format/Protocols.tex
stuff before changes

\section{Description of protocols}
\label{sec:protocols}

This has to expand/rewritten, and must relate to our methodology.

\subsection{MSI}
This is the simple protocol. M implies read and write permission, S implies
read permission and I no permission. Whenever a request for M comes from a
cache to the directory, then the directory must ensure that no other cache is in
S or M state. If other caches are in S or M state, the directory sends a
downgrade request to each of these caches to go to I state. Once all the
directory's notion of the other caches' states is I, the directory can send an M
response to the original cache. Similarly, whenever a request for S comes from a
cache to a directory, then the directory must ensure that no othe cache is in M
state. If so, it sends a downgrade request to S state and once every cache is in
S state or I state, the directory sends an S response to the original cache.
Data is transferred from the directory to the cache whenever the directory
responds to a cache in I states. Data is transferred from the cache to the
directory whenever the cache responds transitioning from M to S or I state.

In a hierarchical setting, the directory itself serves as a cache for the higher
level directories, with a notion of state associated with the directory. In this
scenario, when a directory gets a request from its caches, it first ensures that
its own state is not lower than what is requested. If it is lower, it sends a
request to the parent  for upgrade (in addition to sending requests to the other
caches for downgrading)

\subsection{MESI}
This is the same as the MSI protocol except that the caches have a dirty bit.
Thus, data transfer from a cache to a directory (or from a lower level directory
to a higher level directory) is orchestrated only when the bit is dirty. In
fact, all protocols can have a MESI variant in which the dirty bit is kept track
of.

\subsection{MESI, M forwarding}
This involves a cache in M state transferring data to another cache requesting M
or S. The directory, on receiving a request from a cache, checks to see if any
other cache is in M state. If so, it sends a request to the other cache to a)
downgrade its state appropriately (S or I) and b) to send a response to the
first cache with data. Finally the first cache sends a response to the directory
saying that a) the other cache has downgraded and b) the original cache has
received the data. If no other cache was in M state, the directory responds to
the first cache's request like in simple MESI.

\subsection{MOESI, only M or O forwarding}
This involves an extra O state by the cache. A cache in O state is the owner of
the line, and it handles forwarding the data if any other cache requests for S
or M. A cache in O state may not have the line coherent with the memory.
Whenever a cache requests the directory for M state, if another cache has the
line in M or O state (and all other caches are in I state), then the directory
requests the other cache to a) downgrade to I state and b) forward the data to
the first cache. Upon receiving the data, the first cache moves to M state and
sends a response to the directory indicating that a) it has upgraded to M state
and b) the other cache has downgraded to I state.  Whenever a cache requests the
directory for S state, if another cache has the line in M or O state, then the
directory requests the other cache to a) downgrade to S state and b) forward the
data to the first cache. Upon receiving the data, the first cache moves to O
states and sends a response to the directory indicating that b) it has upgrade
to O state and b) the other cache has downgraded to S state. If no other cache
was in M or O state, the directory responds to the first cache's request like in
simple MESI.

\subsection{MOESI, M, O or S forwarding}
This is a variation of the MOESI protocol with the following extra transition.
If a directory receives an M request from a cache, and if one cache is in O or M
state and other caches are in S state, then the directory sends an a request to
every cache to a) downgrade to I, and b) forward the I response to the first
cache.  Simultaneously, the directory sends the count of the number of caches
sending the forward response to the first cache. Once all the forwarded
responses arrive at the first cache, it responds to the directory saying a) all
other caches have downgraded to I and b) it has upgraded to M.

### 2012-11-04 paper/isca40format/Setup.tex
stuff before changes

\section{Abstract directory-based protocol}
\label{sec:setup}

We first present the state
In an directory-based inclusive cache coherence protocol, the directory keeps
information about the states (of the lines) of all the caches that it services.
The directory essentially has a copy of the 
The directory essentially has a copy of the states of the cache lines of each
cache that it services, but using a more compact encoding. For instance, if a
cache has write permission for a line, no other cache would have any permission
-- in this case, storing just the index of the cache having the write permission is enough,
instead of keeping a replica of the state of the line in each cache.  one
cache can have write permission for a cache line, so it's not necessary to 
The directory can be thought of as having a copy of the states of each cache
line in each state.

### 2012-11-06 paper/isca40format/AbstractProtocol.tex
stuff reasonable

\section{Abstract directory-based protocol}

\label{sec:abstract}

\newcommand{\One}[2]{\text{#1$\langle#2\rangle$}}
\newcommand{\two}[3]{\text{#1$\langle#2,#3\rangle$}}
\newcommand{\three}[4]{\text{#1$\langle#2,#3,#4\rangle$}}
\newcommand{\four}[5]{\text{#1$\langle#2,#3,#4,#5\rangle$}}
\newcommand{\True}{\texttt{True}}
\newcommand{\False}{\texttt{False}}
\newcommand{\assigns}{:=}
\newcommand{\AndAnd}{\;\&\&\;}
\newcommand{\OrOr}{\;||\;}

We will give the abstract directory-based protocol for a hierarchical system of
caches. The caches are assumed to be organized in the form of a tree. The leaves
of this tree represent the L1 caches, with each L1 cache connected to exactly
one core. The root of the tree is a single Last-level cache (LLC), connected to
the main memory.

The notation we adopt is primarily for conciseness of description while allowing
maximum flexibility in the implementation. An realistic efficient implementation
would forego some of the flexibility, and we will show such implementations
later.

A node keeps information about the state of each cache line in the node as well
as in each of its children, \ie the directory. Figure~\ref{table:states} shows
the precise information that a node keeps for each cache line in the node. The
notation we adopt is primarily for conciseness of description. For instance, the
directory need not contain the exact copy of each of the child's coherence
state, for each cache line, instead the encoding can be more efficient. We will
show efficient implementations later.

\begin{figure}
\begin{tabular}{|l|p{.64\columnwidth}|}
\hline
isPresent(a) & Cache line for address $a$ is present\\
\hline
data(a) & Cache line data for address $a$\\
\hline
state(a) & Coherence state for address $a$\\
\hline
dir(a, c) & Directory's copy of the coherence state for address $a$ in child
$c$\\
\hline
isCWait(c, a) & Check if the node waiting for a response from child $c$
for address $a$\\
\hline
cWait(cm a) & The coherence state to which the node has requested child $c$ to
go to, for address $a$\\
\hline
isPWait(a) & Check if the node waiting for a response from its parent for
address $a$\\
\hline
replacingAddr(a) & The address of the line that is going to replace the line
corresponding to address $a$\\
\hline
\end{tabular}
\caption{Information that the node keeps for a cache line}
\label{table:states}
\end{figure}

Figure~\ref{table:messages} shows the incoming and outgoing messages to and from
the node, respectively.

\begin{figure}
\begin{tabular}{|l|p{.6\columnwidth}|}
\hline
\multicolumn{2}{|c|}{Incoming messages}\\
\hline
\three{ReqFromC}{c}{a}{x} & Request from child $c$ upgrade $c$'s coherence
state for address $a$ to $x$\\
\two{ReqFromP}{a}{x} & Request from parent to downgrade node's coherence
state for address $a$ to $x$\\
\three{RespFromC}{c}{a}{x} & Message from child $c$ denoting that $c$'s
coherence state for address $a$ has changed to $x$\\
\three{DataFromC}{c}{a}{d} & Data $d$ of the cache line received from $c$ for
address $a$\\
\two{RespFromP}{a}{x} & Response from parent giving permission to
upgrade the node's coherence state for address $a$ to $x$\\
\two{DataFromP}{a}{d} & Data $d$ of the cache line received from parent for
address $a$\\
\hline
\multicolumn{2}{|c|}{Outgoing messages}\\
\hline
\three{ReqToC}{c}{a}{x} & Request to child $c$ to downgrade $c$'s
coherence state for address $a$ to $x$\\
\two{ReqToP}{a}{x} & Request to parent to upgrade node's coherence
state for address $a$ to $x$\\
\three{RespToC}{c}{a}{x} & Response to child $c$ granting permission to
upgrade $c'$ coherence state for address $a$ to $x$\\
\three{DataToC}{c}{a}{d} & Data $d$ of the cache line sent to the child
$c$ for address $a$\\
\two{RespToP}{a}{x} & Message to parent denoting that the node's
coherence state for address $a$ has changed to $x$\\
\two{DataToP}{c}{d} & Data $d$ of the cache line sent to the parent
for address $a$\\
\hline
\multicolumn{2}{|c|}{Eviction messages}\\
\hline
\One{EvictReq}{a} & Request to evict cache line for address $a$ (generated by
the node itself)\\
\hline
\end{tabular}
\caption{Incoming, outgoing and eviction messages}
\label{table:messages}
\end{figure}

Each message coming from a particular child or the parent is independent of the
messages coming from another source, in addition to being independent of the
messages for a different address. The node can process messages coming from
different sources, and for different addresses in any order. \emph{ A response
or a request should not be handled before handling an earlier response for the
same address from the same source}. Even though our abstract protocol requires
this ordering, we later show that is easy to enforce these even if the
underlying network doesn't enforce point-to-point FIFO ordering or priorities
between different kinds of messages.

Evicting a cache line for replacement can also be treated as a request message,
albeit generated internally by the node. Such a request would be generated if a
cache line requested by a child is currently not present in the cache, and some
other cache line has to be evicted to make room for the requested cache line.

To allow maximum flexibility of implementation, the node is given separate
access queues for each kind of message (request of response) from each source
and for each address. An implementation of a protocol using our methodology does
not have to keep a separate access queue for each source and each address, but
it must adhere to the following rules: \emph{\emph{a)} a request from any child
should not block a request from the parent, and \emph{b)} no request should
block any response}. As we show later, these rules are necessary to avoid
deadlocks. This flexibility allows one to design a wide range of blocking
policy.

The values of the coherence state (and hence the directory state) belong to a
totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
defined between every pair of elements in that set. If a line is not present,
then the state of the line is the least value in this totally ordered set, for
comparison purposes. Let us call this least value as $Inv$, a shorthand for
Invalid. Typically, a line in state $Inv$ signifies no permissions.

We now describe the procedures for a node to handle requests and responses. Note
that a message handler would be triggered only if the corresponding message is
present.  Similarly, if the required resources for handling a message are not
present (for example, if the handling of the request from a child requires
sending further requests to its children or the parent, and the send buffers are
full), then the handler for request from a child would not be triggered.

Figure~\ref{alg:handlePReq} specifies how to handle a node's incoming request
from the parent for downgrading the coherence state of address $a$ to $x$.

Function $hasDataXfer(x, y)$ specifies whether data has to be transferred
from a node to its parent when the node downgrades the coherence state of some
line from $x$ to $y$.  This function is used at the node to determine if the
child has transferred data or not on receiving a response from the child, and
also to determine if the node has to send data to the parent, for processing a
request from the parent. In a typical coherence protocol, the data has to be
only transferred by a node to its parent if the cache line has been modified.

\floatstyle{boxed} 
\restylefloat{figure}

\begin{figure}
\begin{algorithmic}
\Function{handlePReq}{\two{ReqFromP}{a}{x}}
\If {$!isPresent(a)$}
  \State remove$(\two{ReqFromP}{a}{x})$
\ElsIf {$\forall c, !isCWait(c, a)$}
  \If {$state(a) \le x$}
     \State remove$(\two{ReqFromP}{a}{x})$
  \ElsIf {$\forall c, dir(c, a) \le x$}
     \State remove$(\two{ReqFromP}{a}{x})$
     \If {hasDataXfer$(state(a), x)$}
        \State send$(\two{DataToP}{a}{data(a)})$
     \EndIf
     \State $state(a) \assigns x$
     \State send$(\two{RespToP}{a}{x})$
  \Else
     \ForAll{$c$}
        \If {$dir(c, a) > x$}
           \State $isCWait(c, a) \assigns \True$
           \State $cWait(c, a) \assigns x$
           \State send$(\three{ReqToC}{c}{a}{x})$
        \EndIf
     \EndFor
  \EndIf
\EndIf
\EndFunction
\end{algorithmic}
\caption{Handle request from parent}
\label{alg:handlePReq}
\end{figure}

Figure~\ref{alg:handlePResp} specifies how to handle a node's incoming response
from the parent for upgrading the coherence state of address $a$ to $x$.

\begin{figure}
\begin{algorithmic}
\Function{handlePResp}{\two{RespFromP}{a}{x}}
  \State remove$(\two{RespFromP}{a}{x})$
  \If {$state(a) \neq Inv$}
     \State remove$(\two{DataFromP}{a}{d})$
     \State $data(a) = d$
  \EndIf
  \State $state(a) \assigns x$
  \State $isPWait(a) \assigns \False$ 
\EndFunction
\end{algorithmic}
\caption{Handle response from parent}
\label{alg:handlePResp}
\end{figure}

Figure~\ref{alg:handleCResp} specifies how to handle a node's incoming response
from a child for downgrading the child's coherence state for address $a$ to $x$.

\begin{figure}
\begin{algorithmic}
\Function{handleCResp}{\three{RespFromC}{c}{a}{x}}
  \State remove$(\two{RespFromC}{c}{a}{x})$
  \If {hasDataXfer$(dir(c,a), x)$}
     \State remove$(\two{DataFromC}{c}{a}{d})$
     \State $data(a) = d$
  \EndIf
  \State $dir(c,a) \assigns x$
  \If {$isCWait(c, a) \AndAnd cWait(c, a) >= x$}
    \State $isCWait(c, a) \assigns \False$
  \EndIf
\EndFunction
\end{algorithmic}
\caption{Handle response from child}
\label{alg:handleCResp}
\end{figure}

Figure~\ref{alg:handleCReq} specifies how to handle a node's incoming request
from a child for upgrading the child's coherence state for address $a$ to $x$.

Function \emph{compatible(x)} gives the highest state that a line in another
cache can be in without violating coherency if the line is in one cache in state
$x$. For example, the highest state that is compatible with a state having
read-write permissions is the state having no permissions (\ie $Inv$).

Function \emph{getReplace(a)} gives the address of the line that can be replaced
in order to accommodate the line in address $a$. However, if a line is waiting
for a response (from either a child or the parent), then that line can not be
replaced (ensured by the check for $isCWait$ and $isPWait$) in the pseudo-code.

\begin{figure}
\begin{algorithmic}
\Function{handleCReq}{\three{ReqFromC}{c}{a}{x}}
  \If {$isPresent(a)$}
     \If {$!isPWait(a)\ \AndAnd \forall c, !isCWait(c, a)$}
        \If {$state(a) \ge x \AndAnd$ \\
              \hspace{2.5cm} $\forall c' \neq c, dir(c', a) \le \text{compatible}(x)$}
           \State remove$(\three{ReqFromC}{c}{a}{x})$
           \If {$dir(c,a) \neq Inv)$}
              \State send$(\three{DataToC}{c}{a}{data(a)})$
           \EndIf
           \State $dir(c, a) \assigns x$
           \State send$(\three{RespToC}{c}{a}{x})$
        \Else
           \If {$state(a) < x$}
              \State $isPWait(a) \assigns \True$
              \State send$(\two{ReqToP}{a}{x})$
           \EndIf
           \ForAll {$c' \neq c$}
              \If {$dir(c', a) > \text{compatible}(x)$}
                 \State $isCWait(c', a) \assigns \True$
                 \State $cWait(c', a) \assigns \text{compatible}(x)$
                 \State send$(\three{ReqToC}{c'}{a}{\text{compatible}(x)})$
              \EndIf
           \EndFor
        \EndIf
     \EndIf
  \Else
     \State \textbf{let} $a' = \text{getReplace}(a)$
     \If {$!isPWait(a') \AndAnd !isCWait(a')$}
        \State $replacingAddr(a') \assigns a$
        \State send$(\One{EvictReq}{a'})$
     \EndIf
  \EndIf
\EndFunction
\end{algorithmic}
\caption{Handle request from child}
\label{alg:handleCReq}
\end{figure}

Figure~\ref{alg:handleEvict} specifies how to handle an eviction message (in
order to replace the current line in address $a$ with the line for some other
address, which is given by $replacingAddr(a)$).
\begin{figure}
\begin{algorithmic}
\Function{handleEvict}{\One{EvictReq}{a}}
   \If {$!isPresent(a) \OrOr$\\
        \hspace{3cm} $\forall c, dir(c, a) = Inv$}
      \State remove$(\One{EvictReq}{a})$
      \If {$state(a) \neq Inv$}
         \State send$(\two{RespToP}{a}{Inv})$
         \If {hasDataXfer$(state(a), Inv)$}
            \State send$(\two{RespToP}{a}{data(a)})$
         \EndIf
      \EndIf
      \State $isPresent(a) \assigns \False$
      \State $isPresent(replacingAddr(a)) \assigns \True$
   \Else
      \If {$\forall c, !isCWait(c, a)$}
         \ForAll{$c$}
            \If {$dir(c, a) > Inv$}
               \State $isCWait(c, a) \assigns \True$
               \State $cWait(c, a) \assigns Inv$
               \State send$(\three{ReqToC}{c}{a}{Inv})$
            \EndIf
         \EndFor
      \EndIf
   \EndIf
\EndFunction
\end{algorithmic}
\caption{Evict a line in address $a$}
\label{alg:handleEvict}
\end{figure}

\floatstyle{plain} 
\restylefloat{figure}

Note that the handlers continue to be triggered till the corresponding message
has been removed.

In order to avoid starvation, once a response is available, the handler for that
response must be triggered eventually.

Once a node starts handling a request from the parent, it can ignore all other
requests (from the children, or eviction requests) to the same address, till the
node has removed the request from the parent.

Similarly, once a node starts handling a request from a child, it can ignore
other requests from other children to the same address, till it has removed the
request from the child. However, it can not ignore requests from the parent to
the same address, in order to avoid deadlocks.

%The following are the messages transferred to and from the node
%For the abstract protocol, we consider a hierarchical system of caches. The
%caches are organized in the form of a tree. The leaves of this tree are the L1
%caches connected to each core, and the root of the tree is the LLC, connected to
%the main memory. Each node (leaf and non-leaf) contains several cache lines.
%For each cache line in a node, there is a tag and the coherence state of the
%line. Non-leaf nodes contain the directory of each line in addition to the
%coherence state. Each node can communicate only with its parent or children
%nodes.
%
%The directory of a cache line in a node has a copy of the coherence state of
%that cache line in each child-node. Note that the directory can use a clever bit
%encoding in order to reduce the number of bits used, but the information content
%of the directory remains the same. So, for the purposes of designing and
%verification of a protocol, our assumption about the directory state containing
%a copy of each cache state for each line is valid; only the bit counts in the
%implementation varies.
%
%The abstract protocol defines the behavior of each node, \ie the transitions of
%the directory as well as the coherence state on sending or receiving a message.
%The transitions are based on local information, \ie a node can access or modify
%only the states present in the node.
%
%The values of the coherence state (and hence the directory state) belong to a
%totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
%defined between every pair of elements in that set.
%
%We will use the following notation to specify the information associated with a
%particular cache line for address $a$.
%
%\begin{itemize}
%
%\item $data[a]$ denotes the data for address $a$.
%
%\item $state[a]$ denotes the coherence state of address $a$.
%
%\item $dir[a](c)$ denotes the directory's copy of the coherence state of address
%$a$ in cache $c$.
%
%\item $tag[a]$ denotes the tag of address $a$.
%
%\item $isCWait[a]$ denotes if the node is waiting for a response from any of its
%child regarding address $a$.
%
%\item $cWait[a]$ denotes the request that a node has sent its children for
%address $a$ and hence waiting for a response.
%
%\item $isPWait[a]$ denotes if the node is waiting for a response from its parent
%regarding address $a$.
%
%\item $pWait[a]$ denotes the request that a node has sent its parent for address
%$a$ and hence waiting for a response.
%
%\end{itemize}
%
%Each node sends requests and responses to its children and its parent to
%orchestrate changing of coherence states. Requests are sent to a node's parent
%for upgrading the coherence state of a line in that node, and to the node's
%children to downgrade the coherence states of the corresponding children.
%Responses received from a parent signify that the directory in the parent has
%been upgraded appropriately for that cache. Responses received from a child
%signify that the coherence state of the child has been downgraded appropriately.
%Each message coming from a particular child or the parent is independent of the
%messages coming from another source, in addition to being independent of the
%messages for a different address. The node can process messages coming from
%different sources, and for different addresses in any order. From the same
%source and for the same address, we assume that neither a response not a request
%can not overtake an earlier response
%
%Even though our abstract protocol requires these message orderings, we later
%show that is easy to enforce these even if the underlying network doesn't
%enforce point-to-point FIFO ordering or priorities between different kinds of
%messages.
%
%To allow maximum flexibility of implementation, the node has separate access
%queues for each kind of message (request of response) from each source and for
%each address. An implementation of a protocol using our methodology does not
%have to keep a separate access queue for each source and each address, but it
%has to adhere to the following rules: a) a request from any child should not
%cause head-of-line blocking of a request from a parent, and b) no request should
%cause head-of-line blocking of any response. As we show later, these rules are
%necessary to avoid deadlocks. This flexibility allows one to use a wide range of
%blocking policy.
%
%The following notation will be used for the requests and responses sent and
%received:
%
%\begin{itemize}
%
%\item $p.sendReq(a, to)$ denotes sending a request to the parent
%in order to upgrade the coherence state of address $a$ in the node to state
%$to$.
%
%\item $c.sendReq(a, to)$ denotes sending a request to child $c$ in order to
%downgrade the coherence state of address $a$ in the $c$ to state $to$.
%
%\item $p.sendResp(a, to, d)$ denotes sending a response message to the parent
%signifying that the coherence state of the node for address $a$ has changed to
%$to$. $d$ denotes the cache line sent (if necessary) by the node to the parent.
%
%\item $c.sendResp(a, to, d)$ denotes sending a response message to child $c$
%signifying that the directory state of the node for address $a$ and child $c$
%has changed to $to$. $d$ denotes the cache line sent (if necessary) by the node
%to child $c$.
%
%\item $p.peekReq[a]$ denotes the request from the parent for address $a$. The
%message contains the value of the coherence state for address $a$ that this node
%should downgrade to.
%
%\item $c.peekReq[a]$ denotes the request from child $c$ for address $a$. The
%message contains the value of the coherence state for address $a$ that child $c$
%wants to upgrade to.
%
%\item $p.deqReq[a]$ denotes finishing up of the processing of the request from
%the parent for address $a$ and hence removing the request.
%
%\item $c.deqReq[a]$ denotes finishing up of the processing of the request from
%child $c$ for address $a$ and hence removing the request.
%
%\item $p.peekResp[a]$ denotes the response from the parent for address $a$. The
%message contains a pair $\langle to, d\rangle$, where $to$ denotes the value of
%the coherence state for address $a$ that this node should upgrade to, and $d$
%represents the cache line data sent by the parent (if necessary)
%
%\item $c.peekResp[a]$ denotes the response from child $c$ for address $a$. The
%message contains a pair $\langle to, d\rangle$, where $to$ denotes the value of
%the coherence state for address $a$ that child $c$ has downgraded to, and $d$
%represents the cache line data sent by the child (if necessary)
%
%\item $p.deqReq[a]$ denotes finishing up of the processing of the response from
%the parent for address $a$ and hence removing the response.
%
%\item $c.deqReq[a]$ denotes finishing up of the processing of the response from
%child $c$ for address $a$ and hence removing the response.
%
%\end{itemize}
%

### 2012-11-06 paper/isca40format/Introduction.tex
stuff reasonable

@@
-\section{Introduction} With the increasing number of cores, bus-based
-cache-coherence protocols have started being replaced by message-passing
-directory-based protocols. Bus-based protocols allow only one cache-line
-transaction to happen at a time, locking the bus till the transaction is
-complete. This serializes memory accesses across each processor, even if they
-are for different cache lines. On the other hand, message-passing,
-directory-based protocols can handle multiple independent transactions in
-parallel, thus scaling better with increasing core counts.
+\section{Introduction}
 
-In a message-passing directory-based protocol, the coherence state changes are
-orchestrated by passing messages in the form of upgrade or downgrade requests
-and responses between the caches via the directory. Unlike bus-based protocols
-which lock the bus till all the required state transitions are done thereby
-ensuring atomicity of state transitions, in a message-passing protocol, requests
-and responses are split leading to transient states. This creates scenarios not
-present in bus-based protocols, which can lead to coherence violations or
-deadlocks, if the protocol is not designed or implemented carefully.
+Large shared-memory multicores are likely to use cache-coherence protocols
+which are directory based, hierarchical, and implemented using message passing.
+Even though such protocols have been around for two decades (see, for example,
+DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
+such protocols remains a challenge. It is difficult to convince oneself that
+all corner cases have been handled adequately without a formal proof. And yet
+formal proofs of real implementations are practically non-existent. One can not
+settle for conservative choices in cache-coherence protocol designs because
+performance (\eg the number of messages needed to maintain coherence, the
+buffer requirements, etc) is as important as correctness. Consequently proving
+a simple protocol for a small number of caches, under unbounded buffer
+assumption is of limited use in gaining confidence in the correctness of a real
+protocol.
 
-For instance, consider a scenario in which a cache having read permission for
-a line gets a store request from its core and hence requires write permissions
-for that line. The cache would send an upgrade request to the directory
-controller and may eventually get a response from the directory controller. The
-response would signify that the permissions in the other caches \emph{were}
-compatible with this cache having write permission for the line, \ie the other
-caches neither \emph{had} read nor write permissions for that line. But the messages
-take time to propagate. Thus, as emphasized, the response message is based on
-the old permissions that the other caches had, not the permissions that the
-other caches have at the moment the original cache gets the response. If an
-other cache gets a read or write permission for that line during the time the
+In this paper we will consider systems where caches form a tree hierarchy and
+caches are \emph{inclusive}, that is, if a cache contains address $a$ then its
+parent is guaranteed to contain address $a$, though the parent may have stale
+data for $a$. Every cache or node in the system has an associated directory
+which contains information about its children caches. The coherence state
+changes are orchestrated by passing messages in the form of \emph{upgrade} or
+\emph{downgrade} requests and responses between the caches via the directory.
+Unlike bus-based protocols which lock the bus till all the required state
+transitions are done thereby ensuring atomicity of state transitions, in a
+message-passing protocol, requests and responses are split leading to transient
+states. This creates scenarios not present in bus-based protocols, which can
+lead to coherence violations or deadlocks, if the protocol is not designed or
+implemented carefully.
+
+Several factors contribute to the difficulty of designing protocols for such
+distributed, hierarchical and message-passing systems:
+
+\begin{enumerate}
+
+\item Lots of concurrent events must be dealt with without a central authority
+to impose a sequence on events;
+
+\item The state on which transitions are based is physically distributed in
+space and cannot be accessed atomically;
+
+\item Physical network may block the transmission of a message because of lack
+of communication buffers or head-of-the-line blocking;
+
+\item For each message type, determining which cache and directory states are
+possible and therefore must be dealt with is hard. An incomplete set of
+transition rules may cause the protocol to deadlock.
+
+\end{enumerate}
+
+As examples of problems in designing correct protocols, we describe several
+concrete and familiar scenarios from two-level systems (a private L1 cache for
+each processor and a shared L2 cache):
+
+\noindent \emph{Example 1:} A cache with a read permission for a line gets a
+store request from its processor and hence, requires a write permission for
+that line. The cache would send an upgrade request to the directory controller
+and may eventually get a response from the directory controller. The response
+would signify that the states in the other caches \emph{were} compatible with
+this cache having write permission for the line. That is, the other caches
+neither \emph{had} read nor write permissions for that line. But the messages
+take time to propagate and the response message is necessarily based on the old
+state bits (permissions) that the other caches had, not the permissions that
+the other caches have at the moment the original cache gets the response. If
+another cache gets a read or write permission for that line during the time the
 messages are in flight, coherence would be violated.
 
-Consider a different scenario in which a cache, already having read permission
-for a line, gets a store request, and hence sends an upgrade request for write
-permission to the directory controller. Instead of getting a response, the cache
-may receive a request from the directory controller asking this cache to give up
-the read permission for the line \ie an invalidate request. This could happen if
-the directory controller received a request for write permission from another
-cache earlier. It is not easy to determine the correct behavior of the cache
+\emph{Example 2:} A cache, already having read permission for a line, gets a
+store request, and sends an upgrade request for write permission to the
+directory controller. Instead of getting a response, the cache may receive a
+request from the directory controller asking this cache to give up the read
+permission for the line \ie an invalidate request. This could happen if the
+directory controller received a request for write permission from another cache
+earlier. It is not easy to determine the correct behavior of the cache
 receiving the invalidate request -- whether it should ignore the invalidate
@@
 its request, or if it should invalidate its line. If it ignores the invalidate
-request, is there any guarantee that it would receive a response for its request
-for write permission, or would it deadlock? As we will be showing later, this
-could create a deadlock.
+request, is there any guarantee that it would receive a response for its
+request for write permission, or would it deadlock?
 
-Consider yet another scenario in which a cache receives an invalidate request
-from the directory controller for a line that is not present in the cache. Is
-such a request even possible? The directory does not have access to the
-up-to-date state of the cache, and so can potentially make this request. How
-should a cache handle this request; should it drop such a request, or respond
-saying that it does not have the line? Again, reasoning about these questions is
-very difficult, as one has to take into account all the possible transitions
-that can happen in the entire system.
+\emph{Example 3:} A cache receives an invalidate request from the directory
+controller for a line that is not present in the cache. Is such a request even
+possible? The directory does not have access to the up-to-date state of the
+cache, and so can potentially make this request. How should a cache handle this
+request; should it drop such a request, or respond saying that it does not have
+the line?
 
@@
 correct cache coherence protocol in a message-passing directory-based setting.
-Even for a 2-level cache hierarchy, a private L1 cache for each core and a
-shared L2 cache, designing and implementing the protocol is very hard. This
-problem is greatly exacerbated in a multi-level cache hierarchy, which is
-becoming the norm because of the increasing core counts.
+Even for a 2-level cache hierarchy, designing and implementing the protocol is
+very hard. This problem is greatly exacerbated in a multi-level cache
+hierarchy, which is becoming the norm because of the increasing core counts.
 
-In this paper, we present a systematic design methodology to design hierarchical
-messaging-passing, directory-based inclusive cache coherence protocols. We
-restrict our methodology to inclusive protocols as we want our protocols to
-scale to multiple levels of hierarchy. A non-inclusive or exclusive cache
-coherence protocol requires probing every level of the hierarchy starting from
-the root (LLC) to the leaf (L1 cache) to process any request for permission. We
-believe that this would not scale as the number of cores and the number of
-cache levels increase.
+In this paper, we present a substantially easier method to design
+cache-coherence protocols for such systems. The method essentially breaks down
+the design problem in two parts: a part that is an abstract protocol shared by
+a number of concrete protocols, and another part that is specific to the
+protocol to be designed and implemented. The abstract protocol is parameterized
+by the number of cache levels and the number of caches in each level of the
+hierarchy as well as by a predicate which involves a local ``compatibility of
+state'' test. We illustrate our methodology by designing multiple
+hierarchical cache-coherence protocols, notably MSI, MESI, MOSI, MOESI, both in
+the presence as well as in the absence of point-to-point FIFO networks.
 
-We present a set of transitions that the caches and directories must follow on
-sending or receiving a request or a response. We have formally proved that
-these behaviors will ensure the following properties: a) the directory's notion
-of a cache's state is ``conservative'' in that a cache is never going to have
-more permissions that what the directory thinks the cache has, and b) any
-request will eventually get a corresponding response. It should be noted that
-the formal proof was written using the Coq theorem prover \cite{}, and the
-discussion of the formal proof is beyond the scope of this paper.
+Our method substantially reduces the design and proof burden for new protocols
+because of the common abstract protocol. The abstract protocol is a set of
+transitions that the caches and directories follow to send or receive a request
+or a response. We have formally proved that these transitions ensure the
+following properties: a) the directory's notion of a cache's state is
+``conservative'' in that a cache is never going to have more permissions that
+what the directory thinks the cache has; and b) any request will eventually get
+a corresponding response. A formal proof of the abstract protocol was written
+using the Coq theorem prover \cite{}, however, a discussion of the formal proof
+is beyond the scope of this paper.
 
-The flexibility of our methodology to design multiple cache coherence protocols
-comes from the fact that we do not assign any meaning or semantics to the value
-of a coherence state. All the transitions that we present are in the form of
-upgrades or downgrades from state $x$ to state $y$, without giving concrete
-values for $x$ or $y$. We can thus use our methodology to design multiple
-hierarchical cache coherence protocols: MSI, MESI, MOSI, MOESI. We will show
-how each of these protocols obey the set of transitions that we require for our
-methodology, proving the correctness of all these protocols.
+It should be noted that the flexibility of our methodology to design multiple
+cache coherence protocols comes from the fact that we do not assign any meaning
+to the value of a coherence state beyond upgrades or downgrades. Thus, how or
+why a particular cache state implies that a particular address may not be
+present in any other cache stems from protocol specific ``compatibility of
+state'' check.
 
-%\paragraph{Paper organization:} Section \ref{sec:setup} gives the setup for our
-%methodology. It gives the state transitions of the cache and directory
-%controllers on sending or receiving messages. Section \ref{sec:properties} gives
-%the two properties that are guaranteed if the state transitions that we present
-%are obeyed. We present the intuition for proving these properties starting from
-%the state transitions that we present in \ref{sec:setup}. In Section
-%\ref{sec:protocols}, we show how these properties can be used to implement a
-%hierarchical MSI, MESI, MOSI and MOESI protocols. We also demonstrate how the
-%correctness and deadlock freedom of each of the protocols follows straightaway
-%from our methodology with no additional verification effort. Our claim is that
-%our state transitions and the guaranteed properties can be used to formally
-%prove any message-passing directory-based hierarchical inclusive cache coherence
-%protocols by appropriate mapping. In Section \ref{sec:performance}, we show that
-%these formally verified family of protocols require no extra state bits, nor
-%extra message transfers compared to other published protocols. In Section
-%\ref{sec:related}, we present some of the related work and finally in Section
-%\ref{sec:conclusion}, we present the directions we would like to take this work
-%in the future and conclude.
+The main contributions of this paper are: 1. A new abstract protocol that
+satisfies two important invariants for distributed and hierarchal
+message-passing protocols; 2. Instantiation of the abstract protocol to
+implement four well-known protocols (MSI, MESI, MOSI, MOESI) and their
+variations and showing why their correctness is guaranteed as long as the
+abstract protocol is correct, \ie it provably implements the two invariants;
+and 3. Our method does not introduce any overhead over direct implementation of
+these protocols.  We think that this new method of designing protocols
+substantially reduces the required effort, and that our abstract protocol is
+flexible enough to implement (while ensuring correctness of) most if not all
+inclusive message-passing directory-based protocols.
 
-%We carried out the proof using the Coq theorem prover \cite{}. Unlike
-%model-checking based techniques, the theorem prover merely checks the
-%correctness of a supplied mathematical proof, thereby avoiding the state
-%explosion problem. The proof is not restricted to a fixed number of cores,
-%number of caches levels or number of caches in each level. Instead the proof can
-%be parameterized for these quantities, thus ensuring that the same proof holds
-%for any configuration, without any additional computational complexity. The
-%downside of a theorem prover is that it shifts the onus of supplying a correct
-%mathematical proof to the user, instead of searching through the state space to
-%expose violations of specifications \ala model checking.
-
-%$S$ and $I$ in an MSI protocol
-%The state transitions that we require does not enforce a particular cache
-%coherence protocol. It merely dictates how the states change on sending or
-%receiving a response message, or after handling a request message. To be more
-%concrete, consider a simple example. A directory controller receives a Write
-%permission request for a line from a cache. If the directory thinks that some
-%other cache has a read or write permission for that line, then a correct cache
-%coherence protocol requires that the other cache invalidates that line before
-%the directory sends a response to the first cache giving it Write permission.
-%But this ``policy'' is not enforced by the state transitions that we present.
-%Instead, if the directory decides to give Write permission to the original cache
-%when it thinks another cache still has that line, then our state transitions
-%will permit it. The system will no longer be coherent, but it will not violate
-%the two properties that our state transitions guarantee (the directory would now
-%think that the original cache has Write permission and some other caches also
-%have some permissions for that line). This enables us to implement a wide range
-%of correct cache coherence protocols with minimal proof burden.
-%
-%For example, it is straight forward to implement MSI or MESI protocols making
-%use of the properties guaranteed in our methodology. If the directory controller
-%receives a Write permission request from a cache for a line, then it sends
-%requests to every cache which has that line. Our second property of getting a
-%response for every request ensures that these caches would respond eventually,
-%making their local states for that line Invalid. Our first property guarantees
-%that when the directory controller thinks the state of a cache for a line is
-%Invalid, then the actual states in the caches at that time should not be greater
-%than Invalid. Thus, once all the other caches have responded, the directory
-%controller can respond to the original cache. This way, the two properties can
-%be used to implement MSI or MESI in a straightforward manner. We will present
-%how this is done in greater detail in the paper.
-%
-%A more complicated protocol with cache intervention (like MOSI and MOESI)
-%can also be implemented using the two properties that we guarantee, but it
-%requires an additional step of mapping the protocol to our transitions (and
-%proving that the mapping is correct). We will present this in greater detail in
-%the paper.
-%
-%Finally, we will show that these formally verified family of protocols require
-%no extra state bits, nor extra message transfers compared to other published
-%protocols. In fact, our claim is that our state transitions and the two
-%guaranteed properties can be used to formally prove any message-passing
-%directory-based hierarchical inclusive cache coherence protocols by
-%appropriately mapping the protocol to our transitions, as shown in the
-%MOSI/MOESI mapping.
+\paragraph{Paper organization:} Section \ref{sec:abstract} describes the abstract
+protocol; Section \ref{sec:properties} gives the two properties that are
+guaranteed by the abstract protocol and provides some intuition for proving
+these properties. We also discuss the communication buffer requirements for the
+abstract protocol and show that the requirement is only proportional to the
+number of levels and not to the number of processors. In Section
+\ref{sec:protocols}, we show how the abstract protocol can be used to implement
+MSI, MESI, MOSI and MOESI protocols and their variants. In Section
+\ref{sec:performance}, we show that these formally verified family of protocols
+require no extra state bits, nor extra message transfers compared to other
+published implementations. In Section \ref{sec:related}, we present some of the
+related work and finally in Section \ref{sec:conclusion}, we present some
+directions for the future work.

### 2012-11-07 paper/isca40format/AbstractProtocol.tex
with trigger

@@
 \newcommand{\OrOr}{\;||\;}
+\newcommand{\Inv}{\text{$Inv$}}
+
+\definecolor{trig-color}{gray}{.6}
 
@@
 The notation we adopt is primarily for conciseness of description while allowing
-maximum flexibility in the implementation. An realistic efficient implementation
-would forego some of the flexibility, and we will show such implementations
-later.
+maximum flexibility in the implementation.
 
@@
 \hline
-isPresent(a) & Cache line for address $a$ is present\\
+isPresent(a) & Check if cache line for address $a$ is present\\
 \hline
@@
 \hline
-cWait(cm a) & The coherence state to which the node has requested child $c$ to
+cWait(c, a) & The coherence state to which the node has requested child $c$ to
 go to, for address $a$\\
@@
 \hline
+whoTrigger(a) & The source of the request which made the node wait for a
+response. It is $None, Parent, Child \text{ or } Evict$\\
+\hline
+childTrigger(a) & If $whoTrigger(a)$ is $Child$, the index of the child which
+is the source of the request\\
+\hline
 \end{tabular}
@@
 
+The values of the coherence state (and hence the directory state) belong to a
+totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
+defined between every pair of elements in that set and that relation is
+transitive. For instance, in an MSI protocol, the values of the coherence state
+are from the set $\{M, S, I\}$ where $I < S$, $S < M$ and $I < M$. Our abstract
+protocol does not assign any semantics or meaning to the values of the coherence
+states. In a concrete cache coherence protocol, the values of the states are
+associated with permissions for reading and writing, and the state with fewer
+permissions is $<$ the state with more permissions. We use \Inv to denote the
+least element in the set containing the values of the coherence state.
+Typically, this denotes that the node has no permissions for reading or writing
+that line.
+
+In order to increase the coherence state of a cache line, a node has to send an
+\emph{upgrade request} to its parent and may eventually get an \emph{upgrade
+response} from the parent. In order to decrease the coherence state of a cache
+line for one of the node's children, the node has to send a \emph{downgrade
+request} to that child, and may eventually get a \emph{downgrade response} from
+that child.
+
 Figure~\ref{table:messages} shows the incoming and outgoing messages to and from
-the node, respectively.
+the node, respectively, and requests for eviction. An eviction request for an
+address is generated if a child has made a request for a different cache line
+whose address is not currently present in the cache.
 
@@
 
+\floatstyle{boxed} 
+\restylefloat{figure}
+
+\begin{figure}
+\begin{minipage}{\columnwidth}
+\begin{enumerate}
+\item A request or a response should not be handled before handling an earlier
+response for the same address from the same source
+
+\item A request from a child for an address should not block or prevent the
+handling of a request from the parent for the same address
+
+\item A request for an address should not block or prevent the handling of a
+response from any source for any address
+\end{enumerate}
+\end{minipage}
+\caption{Message ordering, and blocking restrictions}
+\label{list:messageOrder}
+\end{figure}
+
+\floatstyle{plain} 
+\restylefloat{figure}
+
 Each message coming from a particular child or the parent is independent of the
@@
 messages for a different address. The node can process messages coming from
-different sources, and for different addresses in any order. \emph{ A response
-or a request should not be handled before handling an earlier response for the
-same address from the same source}. Even though our abstract protocol requires
-this ordering, we later show that is easy to enforce these even if the
-underlying network doesn't enforce point-to-point FIFO ordering or priorities
-between different kinds of messages.
-
-Evicting a cache line for replacement can also be treated as a request message,
-albeit generated internally by the node. Such a request would be generated if a
-cache line requested by a child is currently not present in the cache, and some
-other cache line has to be evicted to make room for the requested cache line.
+different sources, and for different addresses in any order, subject to the
+ordering requirements in Figure~\ref{list:messageOrder}. We later show that is
+easy to enforce this ordering even if the underlying network doesn't enforce
+point-to-point FIFO ordering or priorities between different kinds of messages.
 
@@
 not have to keep a separate access queue for each source and each address, but
-it must adhere to the following rules: \emph{\emph{a)} a request from any child
-should not block a request from the parent, and \emph{b)} no request should
-block any response}. As we show later, these rules are necessary to avoid
-deadlocks. This flexibility allows one to design a wide range of blocking
-policy.
+it must adhere to the blocking restrictions in Figure~\ref{list:messageOrder}.
+These restrictions are necessary to avoid deadlocks. The flexibility of the
+abstract protocol allows one to design a wide range of blocking policies.
 
-The values of the coherence state (and hence the directory state) belong to a
-totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
-defined between every pair of elements in that set. If a line is not present,
-then the state of the line is the least value in this totally ordered set, for
-comparison purposes. Let us call this least value as $Inv$, a shorthand for
-Invalid. Typically, a line in state $Inv$ signifies no permissions.
+Our abstract protocol also depends on the designer correctly defining the three
+functions shown in Figure~\ref{table:functions}. These functions abstract away
+the coherence aspect of the cache coherence protocol. Note that $maxCompat$
+function can be derived from $isCompat$ function, if $isCompat$ is defined
+correctly. Figure~\ref{fig:msiFunc} shows the functions defined for MSI (--
+denotes undefined).
+
+\begin{figure}
+\begin{tabular}{|c|p{.63\columnwidth}|}
+\hline
+isCompat(x, y) & Checks if two different caches can have coherence states
+$x$ and $y$, respectively, for the same address\\
+\hline
+maxCompat(x) & Returns a coherence state value $y$ such that, $\forall z \le y,
+isCompat(x, z)$ and $\forall w > y, !isCompat(x, w)$\\
+\hline
+dataXfer(x, y) & Checks if cache line data has to be transferred by a cache to its
+parent on downgrading from $x$ to $y$\\
+\hline
+\end{tabular}
+\caption{Functions to abstract away the coherence aspect}
+\label{table:functions}
+\end{figure}
+
+\begin{figure}
+\centering
+\begin{subfigure}{.45\columnwidth}
+\centering
+\begin{tabular}{|c||c|c|c|}
+\hline
+& I & S & M\\
+\hline\hline
+I & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\hline
+S & $\checkmark$ & $\checkmark$ & $\times$\\
+\hline
+M & $\checkmark$ & $\times$ & $\times$\\
+\hline
+\end{tabular}
+\label{table:isCompat}
+\caption{$isCompat$ for MSI}
+\end{subfigure}
+\quad
+\begin{subfigure}{.45\columnwidth}
+\centering
+\begin{tabular}{|c||c|c|c|}
+\hline
+\backslashbox{$x$}{$y$} & I & S & M\\
+\hline\hline
+I & -- & -- & --\\
+\hline
+S & $\times$ & -- & --\\
+\hline
+M & $\checkmark$ & $\checkmark$ & --\\
+\hline
+\end{tabular}
+\label{table:dataXfer}
+\caption{$dataXfer(x, y)$ for MSI}
+\end{subfigure}
+\caption{Coherence abstraction for MSI}
+\label{fig:msiFunc}
+\end{figure}
 
@@
 that a message handler would be triggered only if the corresponding message is
-present.  Similarly, if the required resources for handling a message are not
-present (for example, if the handling of the request from a child requires
-sending further requests to its children or the parent, and the send buffers are
-full), then the handler for request from a child would not be triggered.
+present. If the required resources for handling a message are not present (for
+example, if the handling of the request from a child requires sending further
+requests to its children or the parent, and the send buffers are full), then the
+handler for request from a child would not be triggered.
 
 Figure~\ref{alg:handlePReq} specifies how to handle a node's incoming request
-from the parent for downgrading the coherence state of address $a$ to $x$.
+from the parent for downgrading the coherence state of address $a$ to $x$. Note
+that a request from the parent can be handled successfully only if the node is
+not already waiting for a response from a child.
 
-Function $hasDataXfer(x, y)$ specifies whether data has to be transferred
-from a node to its parent when the node downgrades the coherence state of some
-line from $x$ to $y$.  This function is used at the node to determine if the
-child has transferred data or not on receiving a response from the child, and
-also to determine if the node has to send data to the parent, for processing a
-request from the parent. In a typical coherence protocol, the data has to be
-only transferred by a node to its parent if the cache line has been modified.
+Figure~\ref{alg:handleCReq} specifies how to handle a node's incoming request
+from a child for upgrading the child's coherence state for address $a$ to $x$.
+Note that a request from a child can be handled successfully only if the node is
+not already waiting for a response from the parent or from any child.
+
+Figure~\ref{alg:handlePResp} specifies how to handle a node's incoming response
+from the parent for upgrading the coherence state of address $a$ to $x$.
+
+Figure~\ref{alg:handleCResp} specifies how to handle a node's incoming response
+from a child for downgrading the child's coherence state for address $a$ to $x$.
+
+Figure~\ref{alg:handleEvict} specifies how to handle an eviction message (in
+order to replace the current line in address $a$ with the line for some other
+address, which is given by $replacingAddr(a)$).
 
@@
 \begin{algorithmic}
-\Function{handlePReq}{\two{ReqFromP}{a}{x}}
+\Rule{handlePReq}{\two{ReqFromP}{a}{x}}
 \If {$!isPresent(a)$}
@@
      \State remove$(\two{ReqFromP}{a}{x})$
-     \If {hasDataXfer$(state(a), x)$}
+     \If {dataXfer$(state(a), x)$}
         \State send$(\two{DataToP}{a}{data(a)})$
@@
      \EndFor
+     \State \textcolor{trig-color}{$whoTrigger(a) \assigns Parent$}
   \EndIf
 \EndIf
-\EndFunction
+\EndRule
 \end{algorithmic}
@@
 
-Figure~\ref{alg:handlePResp} specifies how to handle a node's incoming response
-from the parent for upgrading the coherence state of address $a$ to $x$.
-
 \begin{figure}
 \begin{algorithmic}
-\Function{handlePResp}{\two{RespFromP}{a}{x}}
-  \State remove$(\two{RespFromP}{a}{x})$
-  \If {$state(a) \neq Inv$}
-     \State remove$(\two{DataFromP}{a}{d})$
-     \State $data(a) = d$
-  \EndIf
-  \State $state(a) \assigns x$
-  \State $isPWait(a) \assigns \False$ 
-\EndFunction
-\end{algorithmic}
-\caption{Handle response from parent}
-\label{alg:handlePResp}
-\end{figure}
-
-Figure~\ref{alg:handleCResp} specifies how to handle a node's incoming response
-from a child for downgrading the child's coherence state for address $a$ to $x$.
-
-\begin{figure}
-\begin{algorithmic}
-\Function{handleCResp}{\three{RespFromC}{c}{a}{x}}
-  \State remove$(\two{RespFromC}{c}{a}{x})$
-  \If {hasDataXfer$(dir(c,a), x)$}
-     \State remove$(\two{DataFromC}{c}{a}{d})$
-     \State $data(a) = d$
-  \EndIf
-  \State $dir(c,a) \assigns x$
-  \If {$isCWait(c, a) \AndAnd cWait(c, a) >= x$}
-    \State $isCWait(c, a) \assigns \False$
-  \EndIf
-\EndFunction
-\end{algorithmic}
-\caption{Handle response from child}
-\label{alg:handleCResp}
-\end{figure}
-
-Figure~\ref{alg:handleCReq} specifies how to handle a node's incoming request
-from a child for upgrading the child's coherence state for address $a$ to $x$.
-
-Function \emph{compatible(x)} gives the highest state that a line in another
-cache can be in without violating coherency if the line is in one cache in state
-$x$. For example, the highest state that is compatible with a state having
-read-write permissions is the state having no permissions (\ie $Inv$).
-
-Function \emph{getReplace(a)} gives the address of the line that can be replaced
-in order to accommodate the line in address $a$. However, if a line is waiting
-for a response (from either a child or the parent), then that line can not be
-replaced (ensured by the check for $isCWait$ and $isPWait$) in the pseudo-code.
-
-\begin{figure}
-\begin{algorithmic}
-\Function{handleCReq}{\three{ReqFromC}{c}{a}{x}}
+\Rule{handleCReq}{\three{ReqFromC}{c}{a}{x}}
   \If {$isPresent(a)$}
@@
         \If {$state(a) \ge x \AndAnd$ \\
-              \hspace{2.5cm} $\forall c' \neq c, dir(c', a) \le \text{compatible}(x)$}
+              \hspace{2.5cm} $\forall c' \neq c, dir(c', a) \le \text{maxCompat}(x)$}
            \State remove$(\three{ReqFromC}{c}{a}{x})$
-           \If {$dir(c,a) \neq Inv)$}
+           \If {$dir(c,a) \neq \Inv)$}
               \State send$(\three{DataToC}{c}{a}{data(a)})$
@@
            \ForAll {$c' \neq c$}
-              \If {$dir(c', a) > \text{compatible}(x)$}
+              \If {$dir(c', a) > \text{maxCompat}(x)$}
                  \State $isCWait(c', a) \assigns \True$
-                 \State $cWait(c', a) \assigns \text{compatible}(x)$
-                 \State send$(\three{ReqToC}{c'}{a}{\text{compatible}(x)})$
+                 \State $cWait(c', a) \assigns \text{maxCompat}(x)$
+                 \State send$(\three{ReqToC}{c'}{a}{\text{maxCompat}(x)})$
               \EndIf
            \EndFor
+           \State \textcolor{trig-color}{$whoTrigger(a) \assigns Child$}
+           \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
         \EndIf
@@
   \Else
-     \State \textbf{let} $a' = \text{getReplace}(a)$
+     \State \textbf{let} $a' \assigns \text{getReplace}(a)$
      \If {$!isPWait(a') \AndAnd !isCWait(a')$}
@@
         \State send$(\One{EvictReq}{a'})$
+        \State \textcolor{trig-color}{$whoTrigger(a) \assigns Evict$}
+        \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
      \EndIf
   \EndIf
-\EndFunction
+\EndRule
 \end{algorithmic}
@@
 
-Figure~\ref{alg:handleEvict} specifies how to handle an eviction message (in
-order to replace the current line in address $a$ with the line for some other
-address, which is given by $replacingAddr(a)$).
 \begin{figure}
 \begin{algorithmic}
-\Function{handleEvict}{\One{EvictReq}{a}}
-   \If {$!isPresent(a) \OrOr$\\
-        \hspace{3cm} $\forall c, dir(c, a) = Inv$}
+\Rule{handleEvict}{\One{EvictReq}{a}}
+   \If {$!isPresent(a) \OrOr \forall c, dir(c, a) = \Inv$}
       \State remove$(\One{EvictReq}{a})$
-      \If {$state(a) \neq Inv$}
-         \State send$(\two{RespToP}{a}{Inv})$
-         \If {hasDataXfer$(state(a), Inv)$}
+      \If {$state(a) \neq \Inv$}
+         \State send$(\two{RespToP}{a}{\Inv})$
+         \If {dataXfer$(state(a), \Inv)$}
             \State send$(\two{RespToP}{a}{data(a)})$
@@
       \State $isPresent(a) \assigns \False$
-      \State $isPresent(replacingAddr(a)) \assigns \True$
+      \State allocLine$(replacingAddr(a))$
+      \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
    \Else
@@
          \ForAll{$c$}
-            \If {$dir(c, a) > Inv$}
+            \If {$dir(c, a) > \Inv$}
                \State $isCWait(c, a) \assigns \True$
-               \State $cWait(c, a) \assigns Inv$
-               \State send$(\three{ReqToC}{c}{a}{Inv})$
+               \State $cWait(c, a) \assigns \Inv$
+               \State send$(\three{ReqToC}{c}{a}{\Inv})$
             \EndIf
@@
    \EndIf
-\EndFunction
+\EndRule
 \end{algorithmic}
@@
 
+\begin{figure}
+\begin{algorithmic}
+\Rule{handlePResp}{\two{RespFromP}{a}{x}}
+  \State remove$(\two{RespFromP}{a}{x})$
+  \If {$state(a) \neq \Inv$}
+     \State remove$(\two{DataFromP}{a}{d})$
+     \State $data(a) = d$
+  \EndIf
+  \State $state(a) \assigns x$
+  \State $isPWait(a) \assigns \False$ 
+  \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
+\EndRule
+\end{algorithmic}
+\caption{Handle response from parent}
+\label{alg:handlePResp}
+\end{figure}
+
+\begin{figure}
+\begin{algorithmic}
+\Rule{handleCResp}{\three{RespFromC}{c}{a}{x}}
+  \State remove$(\two{RespFromC}{c}{a}{x})$
+  \If {dataXfer$(dir(c,a), x)$}
+     \State remove$(\two{DataFromC}{c}{a}{d})$
+     \State $data(a) = d$
+  \EndIf
+  \State $dir(c,a) \assigns x$
+  \If {$isCWait(c, a) \AndAnd cWait(c, a) >= x$}
+    \State $isCWait(c, a) \assigns \False$
+  \EndIf
+  {\color{trig-color}
+  \If {$whoTrigger(a) = Parent$}
+     \State \textbf{trigger} handlePReq
+  \ElsIf {$whoTrigger(a) = Evict$}
+     \State \textbf{trigger} handleEvict
+  \ElsIf {$whoTrigger(a) = Child$}
+     \State \textbf{trigger} handleCReq($childTrigger(a)$)
+  \EndIf}
+\EndRule
+\end{algorithmic}
+\caption{Handle response from child}
+\label{alg:handleCResp}
+\end{figure}
+
+Function \emph{getReplace(a)} gives the address of the line that can be replaced
+in order to accommodate the line in address $a$. Note that if the line selected
+for eviction is waiting for a response from the parent or from any child, then
+it can not be replaced.
+
+Figure~\ref{alg:allocLine} shows the initiation that has to happen for
+allocating a new cache line for an address.
+
+\begin{figure}
+\begin{algorithmic}
+\Procedure{allocLine}{$a$}
+  \State $isPresent(a) \assigns \True$
+  \State \textcolor{trig-color}{$whoTrigger(a) \assigns None$}
+  \State $state(a) \assigns \Inv$
+  \ForAll {$c$}
+     \State $dir(c, a) \assigns \Inv$
+  \EndFor
+  \State $isCWait(a) \assigns \False$
+  \State $isPWait(a) \assigns \False$
+\EndProcedure
+\end{algorithmic}
+\caption{Allocate a new cache line for address $a$}
+\label{alg:allocLine}
+\end{figure}
+
 \floatstyle{plain} 
@@
 
-Note that the handlers continue to be triggered till the corresponding message
-has been removed.
-
-In order to avoid starvation, once a response is available, the handler for that
+Each handler will be triggered repeatedly till the corresponding message has
+been removed. In order to avoid starvation, the following two conditions must be
+maintained:
+\begin{enumerate}
+\item Once a response is available from any source, the handler for that
 response must be triggered eventually.
-
-Once a node starts handling a request from the parent, it can ignore all other
-requests (from the children, or eviction requests) to the same address, till the
-node has removed the request from the parent.
-
-Similarly, once a node starts handling a request from a child, it can ignore
-other requests from other children to the same address, till it has removed the
-request from the child. However, it can not ignore requests from the parent to
-the same address, in order to avoid deadlocks.
+\item After handling a response (or an evict request) the corresponding handler
+must trigger the original handler that requested the received response (or made
+an eviction request). The $whoTrigger(a)$ keeps track of the original handler
+for each address $a$. In case of a $Child$ trigger, the actual child
+corresponding to the original handler is kept track of in $childTrigger(a)$, for
+address $a$, in which case $handleCReq$ is triggered for the outstanding request
+from that child. In the figures, the code corresponding to this is grayed out
+\end{enumerate}
 

### 2012-11-07 paper/isca40format/Introduction.tex
with trigger

@@
 Large shared-memory multicores are likely to use cache-coherence protocols
-which are directory based, hierarchical, and implemented using message passing.
+which are directory-based, hierarchical, and implemented using message passing.
 Even though such protocols have been around for two decades (see, for example,
@@
 all corner cases have been handled adequately without a formal proof. And yet
-formal proofs of real implementations are practically non-existent. One can not
+formal proofs of real implementations are practically non-existent. One can't
 settle for conservative choices in cache-coherence protocol designs because
@@
 \item For each message type, determining which cache and directory states are
-possible and therefore must be dealt with is hard. An incomplete set of
+possible (and therefore must be dealt) with is difficult. An incomplete set of
 transition rules may cause the protocol to deadlock.
@@
 These examples give a glimpse of the complexity of designing and implementing a
-correct cache coherence protocol in a message-passing directory-based setting.
+correct cache coherence protocol in a message-passing, directory-based setting.
 Even for a 2-level cache hierarchy, designing and implementing the protocol is
-very hard. This problem is greatly exacerbated in a multi-level cache
+quite difficult. This problem is greatly exacerbated in a multi-level cache
 hierarchy, which is becoming the norm because of the increasing core counts.
@@
 a corresponding response. A formal proof of the abstract protocol was written
-using the Coq theorem prover \cite{}, however, a discussion of the formal proof
-is beyond the scope of this paper.
+using the Coq theorem prover \cite{}, however, a presentation of the formal
+proof is beyond the scope of this paper.
 
@@
 flexible enough to implement (while ensuring correctness of) most if not all
-inclusive message-passing directory-based protocols.
+inclusive, message-passing, directory-based protocols.
 

### 2012-11-07 paper/isca40format/AbstractProtocol.tex
will be removing evict req

@@
 
-\definecolor{trig-color}{gray}{.6}
+\definecolor{trig-color}{gray}{.4}
 
@@
 \hline
-replacingAddr(a) & The address of the line that is going to replace the line
-corresponding to address $a$\\
-\hline
 whoTrigger(a) & The source of the request which made the node wait for a
 response. It is $None, Parent, Child \text{ or } Evict$\\
+\hline
+replacingAddr(a) & If $whoTrigger(a)$ is $Evict$, the address replacing the line
+of address $a$\\
 \hline
@@
   \If {$isPresent(a)$}
-     \If {$!isPWait(a)\ \AndAnd \forall c, !isCWait(c, a)$}
+     \If {$!isPWait(a) \AndAnd \forall c, !isCWait(c, a)$}
         \If {$state(a) \ge x \AndAnd$ \\
-              \hspace{2.5cm} $\forall c' \neq c, dir(c', a) \le \text{maxCompat}(x)$}
+              \hspace{2.2cm} $\forall c' \neq c, dir(c', a) \le \text{maxCompat}(x)$}
            \State remove$(\three{ReqFromC}{c}{a}{x})$
@@
      \State \textbf{let} $a' \assigns \text{getReplace}(a)$
-     \If {$!isPWait(a') \AndAnd !isCWait(a')$}
-        \State $replacingAddr(a') \assigns a$
+     \If {$!isPWait(a') \AndAnd \forall c', !isCWait(c', a')$}
         \State send$(\One{EvictReq}{a'})$
@@
         \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
+        \State $replacingAddr(a') \assigns a$
      \EndIf
@@
          \If {dataXfer$(state(a), \Inv)$}
-            \State send$(\two{RespToP}{a}{data(a)})$
+            \State send$(\two{DataToP}{a}{data(a)})$
          \EndIf
@@
      \State $dir(c, a) \assigns \Inv$
+     \State $isCWait(c, a) \assigns \False$
   \EndFor
-  \State $isCWait(a) \assigns \False$
   \State $isPWait(a) \assigns \False$
@@
 
-\floatstyle{plain} 
-\restylefloat{figure}
-
 Each handler will be triggered repeatedly till the corresponding message has
@@
 maintained:
+
 \begin{enumerate}
+
 \item Once a response is available from any source, the handler for that
 response must be triggered eventually.
+
 \item After handling a response (or an evict request) the corresponding handler
@@
 address $a$, in which case $handleCReq$ is triggered for the outstanding request
-from that child. In the figures, the code corresponding to this is grayed out
+from that child. In the figures, the pseudo-code corresponding to this is grayed out
+
 \end{enumerate}
+
+The L1 caches, \ie the leaf nodes, do not have any children. Instead, they
+handle requests from cores, which resembles handling requests from children. A
+store request for a word is essentially a request by the core for write
+permissions for the corresponding cache line, and a load request for a word is a
+request for read permissions. These permissions can be mapped to the values of
+the coherence state.
+
+Because of the absence of children any request from the parent or an eviction
+request can be handled without producing more requests. As a result, we do not
+need a separate handler for eviction requests. We also do not need to store
+$whoTrigger$, since the only trigger that produces more requests, and hence has
+to wait for response is a request from the core. We omit the pseudo-code for
+handling a request from the core because it is very similar to an instance of
+Figure~\ref{alg:handleCReq} with no children.
+
+%Figure~\ref{alg:handleCoreReq} shows the handler for a request from the core,
+%which is very similar to a combination of \texttt{handleCReq} and
+%\texttt{handleEvict}. The code shows that the cache line is sent back to the
+%core for any request, which is not necessary for a store request in an actual
+%implementation.
+%
+%\begin{figure}
+%\begin{algorithmic}
+%\Rule{handleCoreReq}{\three{ReqFromCore}{a}{x}}
+%  \If {$isPresent(a)$}
+%     \If {$!isPWait(a)$}
+%        \If {$state(a) \ge x$}
+%           \State remove$(\three{ReqFromCore}{a}{x})$
+%           \State send$(\three{RespToCore}{a}{x}{data(a)})$
+%        \ElsIf {$state(a) < x$}
+%              \State $isPWait(a) \assigns \True$
+%              \State send$(\two{ReqToP}{a}{x})$
+%        \EndIf
+%     \EndIf
+%  \Else
+%     \State \textbf{let} $a' \assigns \text{getReplace}(a)$
+%     \If {$!isPWait(a')$}
+%        \If {$state(a') \neq Inv$}
+%           \State send($\two{RespToP}{a'}{Inv}$)
+%           \If {dataXfer($state(a'), Inv$)}
+%              \State send($\two{DataToP}{a'}{Inv}$)
+%           \EndIf
+%        \EndIf
+%        \State $isPresesnt(a') \assigns \False$
+%        \State allocLine($a$)
+%        \State $isPWait(a) \assigns \True$
+%        \State send$(\two{ReqToP}{a}{x})$
+%     \EndIf
+%  \EndIf
+%\EndRule
+%\end{algorithmic}
+%\caption{Handle request from core}
+%\label{alg:handleCoreReq}
+%\end{figure}
+
+\floatstyle{plain} 
+\restylefloat{figure}
 

### 2012-11-07 paper/isca40format/Introduction.tex
will be removing evict req

@@
 
-\emph{Example 2:} A cache, already having read permission for a line, gets a
-store request, and sends an upgrade request for write permission to the
+\noindent \emph{Example 2:} A cache, already having read permission for a line,
+gets a store request, and sends an upgrade request for write permission to the
 directory controller. Instead of getting a response, the cache may receive a
@@
 
-\emph{Example 3:} A cache receives an invalidate request from the directory
-controller for a line that is not present in the cache. Is such a request even
-possible? The directory does not have access to the up-to-date state of the
-cache, and so can potentially make this request. How should a cache handle this
-request; should it drop such a request, or respond saying that it does not have
-the line?
+\noindent \emph{Example 3:} A cache receives an invalidate request from the
+directory controller for a line that is not present in the cache. Is such a
+request even possible? The directory does not have access to the up-to-date
+state of the cache, and so can potentially make this request. How should a
+cache handle this request; should it drop such a request, or respond saying
+that it does not have the line?
 

### 2012-11-09 paper/isca40format/Invariants.tex
will be removing evict req

\section{Invariants}
\label{sec:invariants}

\newcommand{\threeAngle}[4]{\text{$#1\langle#2,#3,#4\rangle$}}

\newcommand{\Req}[3]{\threeAngle{Req}{#1}{#2}{#3}}
\newcommand{\Resp}[3]{\threeAngle{Resp}{#1}{#2}{#3}}
\newcommand{\Inv}{\text{$Inv$\;}}

The values of the coherence state (and hence the directory state) belong to a
totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
defined between any two set elements and this relation is transitive. If an
address is not present, then the coherence state corresponding to that address
is the lowest value in the set (which we denote by \Inv. For instance, in an
MSI protocol, the values of the coherence state are from the set $\{M, S, I\}$
where $I < S$, $S < M$ and $I < M$; if an address is not present, the state
corresponding to that address is $I$. Our abstract protocol does not assign any
semantics or meaning to the values of the coherence states. In a concrete cache
coherence protocol, the values of the states are associated with permissions
for reading and writing, and the state with fewer permissions is $<$ the state
with more permissions.

\begin{figure}\centering
\begin{tabularx}{\linewidth}{|l|X|}
\hline
$state[a]$ & Coherence state for address $a$\\
\hline
$dir[a][c]$ & Directory's version of the coherence state for address $a$ in child
$c$\\
\hline
\end{tabularx}
\caption{Information stored for each cache line}
\label{table:storage}
\end{figure}

\begin{figure}\centering
\begin{tabularx}{\linewidth}{|l|X|}
\hline
$isPresent(a)$ & Is line for address $a$ present in this node?\\
\hline
$isWait(n, a)$ & Is there a pending response from node $n$ (either parent $P$
or a child) for address $a$?\\
\hline
$isReplacing(a)$ & Is address $a$ currently being evicted?\\
\hline
$isEnterPending(a)$ & Is there any (child) request for address $a$ waiting to get
a cache line entry in the cache?\\
\hline
$isHandling(n, a)$ & Is there any request for address $a$ from node $n$ (either
parent $P$ or a child) that has started to be handled?\\
\hline
%$replacingAddr(a)$ & The address that is going to replace the line
%corresponding to address $a$\\
%\hline
%$cWaitState(c, a)$ & The state to which the node has requested its child $c$ to
%downgrade to and hence waiting for a response\\
%\hline
\end{tabularx}
\caption{Information that can be computed for each address (typically from the
MSHR)}
\label{table:functions}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|l|l|X|}
\hline
\multicolumn{3}{|c|}{\Req{n}{a}{x}}\\
\hline
\multirow{4}{*}{$n = P$} & Incoming & Request from parent for changing my
                                      $state[a]$ to $x$\\
\cline{2-3}
                         & Outgoing & Request to parent for permission to change
                                      my $state[a]$ to $x$\\
\hline
\multirow{4}{*}{$n \neq P$} & Incoming & Request from child $n$ for permission
                                         to change $n$'s $state[a]$ to $x$\\
\cline{2-3}
                            & Outgoing & Request to child $n$ to change $n$'s
                                         $state[a]$ to $x$\\
\hline
\multicolumn{3}{|c|}{\Resp{n}{a}{x}}\\
\hline
\multirow{4}{*}{$n = P$} & Incoming & Response from parent granting permission
                                      to change my $state[a]$ to $x$\\
\cline{2-3}
                         & Outgoing & Response to parent indicating that my
                                      $state[a]$ has changed to $x$\\
\hline
\multirow{4}{*}{$n \neq P$} & Incoming & Response from child indicating that
                                         $n$'s $state[a]$ has changed to $x$\\
\cline{2-3}
                            & Outgoing & Response to child $n$ granting permission
                                         to change $n$'s $state[a]$ to $x$\\
\hline
\end{tabularx}
\caption{Incoming and outgoing messages}
\label{table:messages}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
A response for address $a$ can be sent only if the line for $a$ is present in
the current node\\
\hline
\multicolumn{1}{|c|}{Response to child $c$ for address $a$}\\
\hline
\begin{enumerate}
\item A response to child $c$ for address $a$ can be sent only if a
request from child $c$ for $a$ is being handled
\item Response \Resp{c}{a}{x} can be sent only if $x > dir[a][c]$, $\forall c',
c' \neq c \rightarrow isCompatible(x, dir[a][c'])$ and  $x \le state[a]$
\end{enumerate}\\
\hline
\multicolumn{1}{|c|}{Response to parent $P$ for address $a$}\\
\hline
\begin{enumerate}
\item If there is a pending response from the parent for address $a$, a
response to the parent for $a$ can be sent only if a request from
the parent for $a$ is being handled
\item If there is no pending response from the parent for address $a$, a
response to the parent for $a$ can be sent any time
\item Response \Resp{P}{a}{x} can be sent only if $x < state[a]$ and $\forall
c, x \ge dir[a][c]$
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Sending a response: Preconditions}
\label{sendRespPre}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
A response for address $a$ can be sent only if the line for $a$ is present in
the current node\\
\hline
\multicolumn{1}{|c|}{Response to child $c$ for address $a$}\\
\hline
\begin{enumerate}
\item[] On sending \Resp{c}{a}{x}, $dir[a][c]$ changes to $x$
\end{enumerate}\\
\hline
\multicolumn{1}{|c|}{Response to parent $P$ for address $a$}\\
\hline
\begin{enumerate}
\item[] On sending \Resp{P}{a}{x}, $state[a]$ changes to $x$
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Sending a response: Effects}
\label{sendRespEff}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\multicolumn{1}{|c|}{On receiving a response from child $c$ for address $a$}\\
\hline
\begin{enumerate}
\item $dir[a][c]$ changes to $x$
\item Response \Resp{c}{a}{x} is removed
\end{enumerate}\\
\hline
\multicolumn{1}{|c|}{On receiving a response from parent $c$ for address $a$}\\
\hline
\begin{enumerate}
\item $state[a]$ changes to $x$
\item Response \Resp{P}{a}{x} is removed
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Receiving a response: Effects}
\label{recvResp}
\end{figure}

\floatstyle{boxed}
\restylefloat{figure}

\begin{figure}
\begin{enumerate}
\item $state[a]$ can change only on sending or receiving a response from the
parent
\item $dir[a][c]$ can change only on sending or receiving a response from child
$c$
\end{enumerate}
\caption{State and directory changes: Preconditions}
\label{stateChange}
\end{figure}

\floatstyle{plain}
\restylefloat{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\multicolumn{1}{|c|}{Request to child $c$ for address $a$}\\
\hline
\begin{enumerate}
\item A request can be sent to child $c$ for address $a$ only if there is no
pending response from child $c$ for $a$
\item Request \Req{c}{a}{x} can be sent only if $x < dir[a][c]$
\end{enumerate}\\
\hline
\multicolumn{1}{|c|}{Request to parent $P$ for address $a$}\\
\hline
\begin{enumerate}
\item A request can be sent to the parent for address $a$ only if
  \begin{enumerate}
  \item there is no pending response from the parent for address $a$, and
  \item line in address $a$ is not being replaced
  \end{enumerate}
\item Request \Req{P}{a}{x} can be sent only if $x > state[a]$
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Sending a request: Preconditions}
\label{sendReq}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is being handled}\\
\hline
\begin{itemize}
\item[] If $x \le state[a] \wedge \forall c', c' \neq c \rightarrow isCompatible(x, dir[a][c])$\newline
 then the node must send a response \Resp{c}{a}{x}
to $c$
\end{itemize}\\
\hline
\multicolumn{1}{|c|}{Address $a$ is marked as being replaced}\\
\hline
\begin{enumerate}
\item[] If $\forall c, dir[a][c] \le \Inv$, then the node must send a response
\Resp{P}{a}{\Inv} to the parent
\end{enumerate}\\
\hline
\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is being
handled}\\
\hline
\begin{enumerate}
\item[] If $\forall c, dir[a][c] \le x$, then the node must send a response
\Resp{P}{a}{x} to the parent
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Forced responses: Effects}
\label{respForReq}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is received}\\
\hline
\begin{enumerate}
\item If $a$ is present in the node, handling of request \Req{c}{a}{x} can be
started only if
  \begin{enumerate}
  \item no request \Req{n}{a}{x} from any node $n$ for address $a$ is being
  handled
  \item address $a$ is not currently being replaced
  \end{enumerate}
\item If $a$ is not present in the node, handling of request \Req{c}{a}{x} can be
started only if there is a line with address $a'$ such that
  \begin{enumerate}
  \item no request \Req{n}{a'}{x} from any node $n$ for address $a'$ is being
  handled
  \item address $a'$ is not currently being evicted
  \end{enumerate}
\Req{n}{a'}{x} has started
\end{enumerate}\\
\hline
\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is received}\\
\hline
\begin{enumerate}
\item[] Handling of request \Req{P}{a}{x} can be started only if there are no
pending responses from any child for address $a$
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Start handling a request: Preconditions}
\label{respForReqPre}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is received}\\
\hline
\begin{enumerate}
\item Start handling the request \Req{c}{a}{x}
\item If $a$ is not present in the node, get a line with address $a'$ such that
  \begin{enumerate}
  \item no request \Req{n}{a'}{x} from any node $n$ for address $a'$ is being
  handled
  \item address $a'$ is not currently being evicted
  \end{enumerate}
Mark $a'$ as being evicted
\end{enumerate}\\
\hline
\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is received}\\
\hline
\begin{enumerate}
\item[] Start handling the request \Req{P}{a}{x}
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Start handling a request: Effects}
\label{respForReqEff}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is being handled}\\
\hline
\begin{enumerate}
\item[] If $x \le dir[a][c]$, remove the request
\end{enumerate}\\
\hline
\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is being
handled}\\
\hline
\begin{enumerate}
\item[] If $x \ge state[a]$, remove the request
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Finish handling a request: Effects}
\label{finishHandle}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
Request \Req{c}{a}{x} from child $c$ is being handled, and address $a'$ is
replaced by $a$\\
\hline
\begin{enumerate}
\item If $state[a'] <= \Inv$, replace address of the line for $a'$ as $a$,
$state[a']$ as \Inv and $\forall c, dir(c, a)$ as \Inv
\item Unmark $a'$ as being evicted
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Finish replacing line: Effects}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\begin{enumerate}
\item[] If the state or dir of an address (either the requested address, or the
evicted address in case of a replacement) for some node $n$ is not satisfactory,
and there are no pending responses from $n$, send a request to make the state
or dir satisfactory
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Handling of a request: Effects}
\label{handling}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\begin{enumerate}
\item[] The effects during sending a response or sending a request is atomic,
\ie if there is no space available in the sending buffer, then the whole action
is retried later
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Order of receiving messages}
\label{atomicity}
\end{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|X|}
\hline
\begin{enumerate}
\item A request from a child should not block (\ie prevent start of the
handling of) a request from the parent
\item Requests for an address $a$ should not block (\ie prevent reception of)
responses for address $a$, even from a different source
\item A response for an address $a$ from a child $c$ should not be received
before a response from the same $c$ for the same $a$ sent earlier has been
received and removed
\item The handling of a request for an address $a$ from a source $n$ should not
be started before a response from the same $n$ for the same $a$ sent earlier
has been received and removed
\end{enumerate}\\
\hline
\end{tabularx}
\caption{Order of receiving messages}
\label{mesgOrder}
\end{figure}

### 2012-11-10 paper/isca40format/InvariantsHopeful.tex
will be removing evict req

--- 2012-11-09 paper/isca40format/Invariants.tex
@@
 \newcommand{\Resp}[3]{\threeAngle{Resp}{#1}{#2}{#3}}
-\newcommand{\Inv}{\text{$Inv$\;}}
+\newcommand{\Inv}{\text{$Inv$}{}}
+\newcommand{\Dirty}{\text{$Dirty$}{}}
+\newcommand{\True}{\text{$True$}{}}
+\newcommand{\False}{\text{$False$}{}}
 
@@
 address is not present, then the coherence state corresponding to that address
-is the lowest value in the set (which we denote by \Inv. For instance, in an
+is the lowest value in the set (which we denote by \Inv). For instance, in an
 MSI protocol, the values of the coherence state are from the set $\{M, S, I\}$
@@
 coherence protocol, the values of the states are associated with permissions
-for reading and writing, and the state with fewer permissions is $<$ the state
-with more permissions.
+for reading and writing, and the state with fewer permissions is less than the
+state with more permissions.
 
@@
 \hline
-$isReplacing(a)$ & Is address $a$ currently being evicted?\\
+$isEvicting(a)$ & Is address $a$ currently being evicted?\\
 \hline
-$isEnterPending(a)$ & Is there any (child) request for address $a$ waiting to get
-a cache line entry in the cache?\\
-\hline
+%$isEnterPending(a)$ & Is there any (child) request for address $a$ waiting to get
+%a cache line entry in the cache?\\
+%\hline
 $isHandling(n, a)$ & Is there any request for address $a$ from node $n$ (either
-parent $P$ or a child) that has started to be handled?\\
+parent $P$ or a child) that is being handled?\\
 \hline
@@
 \begin{figure}
-\begin{tabularx}{\linewidth}{|l|l|X|}
+\begin{tabularx}{\linewidth}{|cX|}
 \hline
-\multicolumn{3}{|c|}{\Req{n}{a}{x}}\\
+\multicolumn{2}{|c|}{\Req{n}{a}{x}, \Resp{n}{a}{x}}\\
 \hline
-\multirow{4}{*}{$n = P$} & Incoming & Request from parent for changing my
-                                      $state[a]$ to $x$\\
-\cline{2-3}
-                         & Outgoing & Request to parent for permission to change
-                                      my $state[a]$ to $x$\\
-\hline
-\multirow{4}{*}{$n \neq P$} & Incoming & Request from child $n$ for permission
-                                         to change $n$'s $state[a]$ to $x$\\
-\cline{2-3}
-                            & Outgoing & Request to child $n$ to change $n$'s
-                                         $state[a]$ to $x$\\
-\hline
-\multicolumn{3}{|c|}{\Resp{n}{a}{x}}\\
-\hline
-\multirow{4}{*}{$n = P$} & Incoming & Response from parent granting permission
-                                      to change my $state[a]$ to $x$\\
-\cline{2-3}
-                         & Outgoing & Response to parent indicating that my
-                                      $state[a]$ has changed to $x$\\
-\hline
-\multirow{4}{*}{$n \neq P$} & Incoming & Response from child indicating that
-                                         $n$'s $state[a]$ has changed to $x$\\
-\cline{2-3}
-                            & Outgoing & Response to child $n$ granting permission
-                                         to change $n$'s $state[a]$ to $x$\\
+$n:$ & Source or destination node ID (parent $P$, or a child)\\
+$a:$ & Address of the line corresponding to the request\\
+$x:$ & Value from the set of coherence states\\
 \hline
 \end{tabularx}
-\caption{Incoming and outgoing messages}
-\label{table:messages}
+\caption{Incoming and outgoing message}
+\label{message}
 \end{figure}
+
+%\begin{figure}
+%\begin{tabularx}{\linewidth}{|l|l|X|}
+%\hline
+%\multicolumn{3}{|c|}{\Req{n}{a}{x}}\\
+%\hline
+%\multirow{4}{*}{$n = P$} & Incoming & Request from parent for changing my
+%                                      $state[a]$ to $x$\\
+%\cline{2-3}
+%                         & Outgoing & Request to parent for permission to change
+%                                      my $state[a]$ to $x$\\
+%\hline
+%\multirow{4}{*}{$n \neq P$} & Incoming & Request from child $n$ for permission
+%                                         to change $n$'s $state[a]$ to $x$\\
+%\cline{2-3}
+%                            & Outgoing & Request to child $n$ to change $n$'s
+%                                         $state[a]$ to $x$\\
+%\hline
+%\multicolumn{3}{|c|}{\Resp{n}{a}{x}}\\
+%\hline
+%\multirow{4}{*}{$n = P$} & Incoming & Response from parent granting permission
+%                                      to change my $state[a]$ to $x$\\
+%\cline{2-3}
+%                         & Outgoing & Response to parent indicating that my
+%                                      $state[a]$ has changed to $x$\\
+%\hline
+%\multirow{4}{*}{$n \neq P$} & Incoming & Response from child indicating that
+%                                         $n$'s $state[a]$ has changed to $x$\\
+%\cline{2-3}
+%                            & Outgoing & Response to child $n$ granting permission
+%                                         to change $n$'s $state[a]$ to $x$\\
+%\hline
+%\end{tabularx}
+%\caption{Incoming and outgoing messages}
+%\label{table:messages}
+%\end{figure}
 
@@
 \hline
-A response for address $a$ can be sent only if the line for $a$ is present in
-the current node\\
+\begin{enumerate}
+\item[] A response for address $a$ can be sent only if the line for $a$ is present in
+the current node, \ie $isPresent(a)$
+\end{enumerate}\\
 \hline
@@
 \item A response to child $c$ for address $a$ can be sent only if a
-request from child $c$ for $a$ is being handled
-\item Response \Resp{c}{a}{x} can be sent only if $x > dir[a][c]$, $\forall c',
-c' \neq c \rightarrow isCompatible(x, dir[a][c'])$ and  $x \le state[a]$
+request from child $c$ for $a$ is being handled, \ie $isHandling(c, a)$
+\item Response \Resp{c}{a}{x} can be sent only if $x > dir[a][c]$
+% $\forall c', c' \neq c \rightarrow isCompatible(x, dir[a][c'])$ and  $x \le state[a]$
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item If there is a pending response from the parent for address $a$, a
-response to the parent for $a$ can be sent only if a request from
-the parent for $a$ is being handled
+\item If there is a pending response from the parent for address $a$, \ie $\neg
+isWait(P, a)$, a response to the parent for $a$ can be sent only if a request
+from the parent for $a$ is being handled, \ie $isHandling(P, a)$
 \item If there is no pending response from the parent for address $a$, a
 response to the parent for $a$ can be sent any time
-\item Response \Resp{P}{a}{x} can be sent only if $x < state[a]$ and $\forall
-c, x \ge dir[a][c]$
+\item Response \Resp{P}{a}{x} can be sent only if $x < state[a]$
+% and $\forall c, x \ge dir[a][c]$
 \end{enumerate}\\
@@
 A response for address $a$ can be sent only if the line for $a$ is present in
-the current node\\
+the current node, \ie $isPresent(a)$\\
 \hline
@@
 \end{tabularx}
-\caption{Sending a response: Effects}
+\caption{Sending a response: Side effects}
 \label{sendRespEff}
@@
 \item Response \Resp{c}{a}{x} is removed
+\item $isWait(c, a)$ is set \False
 \end{enumerate}\\
@@
 \item Response \Resp{P}{a}{x} is removed
+\item $isWait(P, a)$ is set \False
 \end{enumerate}\\
@@
 \item A request can be sent to child $c$ for address $a$ only if there is no
-pending response from child $c$ for $a$
+pending response from child $c$ for $a$, \ie $isWait(c, a)$
 \item Request \Req{c}{a}{x} can be sent only if $x < dir[a][c]$
@@
   \begin{enumerate}
-  \item there is no pending response from the parent for address $a$, and
-  \item line in address $a$ is not being replaced
+  \item there is no pending response from the parent for address $a$, \ie
+  $isWait(P, a)$, and
+  \item line in address $a$ is not being evicted, \ie $isEvicting(a)$
   \end{enumerate}
@@
 \hline
-\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is being handled}\\
-\hline
-\begin{itemize}
-\item[] If $x \le state[a] \wedge \forall c', c' \neq c \rightarrow isCompatible(x, dir[a][c])$\newline
- then the node must send a response \Resp{c}{a}{x}
-to $c$
-\end{itemize}\\
-\hline
-\multicolumn{1}{|c|}{Address $a$ is marked as being replaced}\\
+\multicolumn{1}{|X|}{\centering Request \Req{c}{a}{x} from child $c$ is being handled, \ie
+$isHandling(c, a)$}\\
 \hline
 \begin{enumerate}
-\item[] If $\forall c, dir[a][c] \le \Inv$, then the node must send a response
-\Resp{P}{a}{\Inv} to the parent
+\item[] If $x \le state[a] \wedge \forall c'\neq c, isCompatible(x, dir[a][c])$,
+ then a response \Resp{c}{a}{x} is sent to $c$ if $x > dir[a][c]$
+%\item If $x = \Inv$ then data for address $a$ is transferred to $c$
 \end{enumerate}\\
 \hline
-\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is being
-handled}\\
+\multicolumn{1}{|c|}{Address $a$ is marked as being evicted, \ie $isEviting(a)$}\\
 \hline
 \begin{enumerate}
-\item[] If $\forall c, dir[a][c] \le x$, then the node must send a response
-\Resp{P}{a}{x} to the parent
+\item[] If $\forall c, dir[a][c] \le \Inv$, a response \Resp{P}{a}{\Inv} is sent
+to the parent if $\Inv < state[a]$
+%\item If $state[a] \in \Dirty$ then data for address $a$ is transferred to $c$
+\end{enumerate}\\
+\hline
+\multicolumn{1}{|X|}{\centering Request \Req{P}{a}{x} from parent $P$ is being
+handled, \ie $isHandling(P, a)$}\\
+\hline
+\begin{enumerate}
+\item[] If $\forall c, dir[a][c] \le x$, a response \Resp{P}{a}{x} is sent to the
+parent if $x < state[a]$
+%\item If $state[a] \in \Dirty$ then data for address $a$ is transferred to the parent
 \end{enumerate}\\
@@
 \item If $a$ is present in the node, handling of request \Req{c}{a}{x} can be
-started only if
+started, \ie $isHandling(c, a)$ can be set \True only if
   \begin{enumerate}
   \item no request \Req{n}{a}{x} from any node $n$ for address $a$ is being
-  handled
-  \item address $a$ is not currently being replaced
+  handled, \ie $\forall n, \neg isHandling(n, a)$
+  \item address $a$ is not currently being evicted, \ie $\neg isEvicting(a)$
   \end{enumerate}
 \item If $a$ is not present in the node, handling of request \Req{c}{a}{x} can be
-started only if there is a line with address $a'$ such that
+started \ie $isHandling(c, a$ can be set \True only if there is a line with
+address $a'$ such that
   \begin{enumerate}
   \item no request \Req{n}{a'}{x} from any node $n$ for address $a'$ is being
-  handled
-  \item address $a'$ is not currently being evicted
+  handled, \ie $\neg isHandling(n, a')$
+  \item address $a'$ is not currently being evicted, \ie $\neg isEvicting(a')$
   \end{enumerate}
-\Req{n}{a'}{x} has started
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item[] Handling of request \Req{P}{a}{x} can be started only if there are no
-pending responses from any child for address $a$
+\item[] Handling of request \Req{P}{a}{x} can be started \ie $isHandling(P, a)$
+can be set \True, only if there are no pending responses from any child for
+address $a$, \ie $\forall c \in Children, \neg isWait(c, a)$
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item Start handling the request \Req{c}{a}{x}
+\item Handling of the request \Req{c}{a}{x} has started, \ie $isHandling(c, a)$
+is set \True
 \item If $a$ is not present in the node, get a line with address $a'$ such that
@@
   \item no request \Req{n}{a'}{x} from any node $n$ for address $a'$ is being
-  handled
-  \item address $a'$ is not currently being evicted
+  handled, \ie $\forall n, \neg isHandling(n, a')$
+  \item address $a'$ is not currently being evicted, \ie $\neg isEvicting(a')$
   \end{enumerate}
-Mark $a'$ as being evicted
+$a'$ is marked as being evicted, \ie $isEvicting(a')$ is set \True
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item[] Start handling the request \Req{P}{a}{x}
+\item[] Handling of the request \Req{P}{a}{x} is started, \ie $isHandling(P, a)$ is set \True
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item[] If $x \le dir[a][c]$, remove the request
+\item If $x \le dir[a][c]$, the request is removed
+\item $isHandling(c, a)$ is set \False
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item[] If $x \ge state[a]$, remove the request
+\item If $x \ge state[a]$, the request is removed
+\item $isHandling(P, a)$ is set \False
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item If $state[a'] <= \Inv$, replace address of the line for $a'$ as $a$,
+\item If $state[a'] <= \Inv$, address $a'$ for the line is replaced by $a$,
 $state[a']$ as \Inv and $\forall c, dir(c, a)$ as \Inv
-\item Unmark $a'$ as being evicted
+\item $a'$ is no longer being evicted, \ie set $isEvicted(a')$ to \False
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item[] If the state or dir of an address (either the requested address, or the
-evicted address in case of a replacement) for some node $n$ is not satisfactory,
-and there are no pending responses from $n$, send a request to make the state
-or dir satisfactory
+\item If a request that is being handled requires $state[a]$ to be $x$ while
+currently $state[a] < x$ and there is no pending response from the parent for
+$a$, \ie $\neg isWait(P, a)$, a request \Req{P}{a}{x} is sent to the parent and
+$isWait(P, a)$ is set \True
+\item If a request that is being handled requires $dir[a][c]$ for some $c$ to
+be $x$ while currently $dir[a][c] > x$ and there is no pending response from
+child $c$, \ie $\neg isWait(c, a)$, a request \Req{c}{a}{x} is sent to $c$ and
+$isWait(c, a)$ is set
+\True
 \end{enumerate}\\
@@
 \end{tabularx}
-\caption{Handling of a request: Effects}
+\caption{Handling of an incoming request: Effects}
 \label{handling}
@@
 \begin{enumerate}
-\item[] The effects during sending a response or sending a request is atomic,
-\ie if there is no space available in the sending buffer, then the whole action
-is retried later
+\item The effects during sending a response or sending a request is atomic,
+\ie if there is no space available in the sending buffer, then the action is
+aborted without causing any effect and retried later
+\item The effect of abort can be achieved by checking if the required resources
+are available before starting the action 
 \end{enumerate}\\
@@
 \end{tabularx}
-\caption{Order of receiving messages}
+\caption{Atomicity requirements}
 \label{atomicity}

### 2012-11-14 paper/isca40format/Atomic.tex
stuff

\newcommand{\fourAngle}[5]{\text{$#1\langle#2,#3,#4,#5\rangle$}}
\newcommand{\ReqCore}[4]{\fourAngle{Req}{#1}{#2}{#3}{#4}}

\subsection{Atomic MSI protocol for 2-levels, no replacement evictions}

XXXXXX: This subsection is probably not going to go into the paper, since it's
too obvious. The definitions have been repeated in the later subsections.

We start with an atomic MSI protocol, with two levels of the memory hierarchy --
the L1 caches and the main memory. In this atomic protocol, the memory
controller can read and write the coherence states as well as the data of each
address present in the L1 caches. As usual, coherence state for an address in
the cache denotes the permissions that the cache has for the address: $M = \{R,
W\}, S = \{R\}, I = \{\}$.

Let us assume that the caches have enough capacity so that no line gets evicted
for the sake of replacement. Lines may however get evicted to enforce the
single-writer property, \ie only one L1 cache can modify a cache line at any
time. Figure~\ref{alg:msi-easy} shows the pseudo-code for the behaviors of the
memory controller and each of the caches to implement the MSI protocol, using
the notations given in Figure~\ref{fig:notation}.

\begin{figure}
\begin{figure}[H]
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\State Pick a request \ReqCore{c}{a}{x}{d} from core to cache $c$ \newline
$s.t.$ $state[c][a] < x$;
\ForAll {$c' \neq c,\; s.t.\; toCompatible(x) < state[c'][a]$}
  \If {$state[c'][a] = M$}
    \State $memory[a] \gets data[c'][a]$;
  \EndIf
  \State $state[c'][a] \gets toCompatible(x)$;
\EndFor
\If {$state[c][a] = I$}
  \State $data[c][a] \gets memory[a]$;
\EndIf
\State $state[c][a] \gets x$;
\end{algorithmic}
\end{boxedminipage}
\subcaption{Behavior of memory controller}
\label{alg:mem}
\end{figure}
\begin{figure}[H]
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\State Pick a request \ReqCore{c}{a}{x}{d} from core to cache $c$ \newline
$s.t.$ $state[c][a] \not< x$;
\If {$state[c][a] = M$}
  \State $update(c, a, d)$;
\ElsIf {$state[c][a] = S$}
  \State $respond(c, a, d)$;
\EndIf
\State remove \ReqCore{c}{a}{x}{d}
\end{algorithmic}
\end{boxedminipage}
\subcaption{Behavior of cache $c$}
\end{figure}
\caption{Atomic 2-level MSI protocol: memory controller has instantaneous
access to all states and data of all L1 caches and data in the memory}
\label{alg:msi-easy}
\end{figure}

\begin{figure}
\begin{minipage}{\linewidth}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$toCompatible(s)$ & If the state of one cache is $s$ for address $a$, the
highest state of another cache for the same $a$
$\newline
toCompatible(M) = I,\newline
toCompatible(S) = S,\newline
toCompatible(I) = M
$\\
\hline
$x < y$ & Does state $x$ have fewer permissions that state $y$?
$I < M, I < S, S < M$\\
\hline
$\ReqCore{c}{a}{x}{d}$ & The incoming request for cache $c$ from a core\\
\multicolumn{1}{|r}{$a$} & the address of the cache line\\
\multicolumn{1}{|r}{$x$} & the state of the cache line the request is expecting
(\ie the minimum permissions required)\\
\multicolumn{1}{|r}{$d$} & information about the word offset, the byte enables
and the data (only in the case of store)\\
\hline
$update(c, a, d)$ & Action denoting update of a cache line for address $a$ in
cache $c$ with the incoming request's information $d$\\
\hline
$respond(c, a, d)$ & Action denoting cache $c$ responding to a read request for
address $a$ with the incoming request's information $d$\\
\hline
\end{tabularx}
\subcaption{Abstraction functions}
\end{minipage}

\begin{minipage}{\linewidth}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for
address $a$\\
$data[c][a]$ & data line of cache $c$ for address $a$\\
$memory[a]$ & data line of memory for address $a$\\
\hline
\end{tabularx}
\subcaption{Values stored by a cache for an address}
\end{minipage}
\caption{Notations used in Figure~\ref{alg:msi-easy}}
\label{fig:notation}
\end{figure}

XXXXXX: End of obvious subsection

### 2012-11-14 paper/isca40format/Fifo.tex
stuff

\section{Virtual channel requirements and ordering}
\label{sec:orderingBlocking}

\newcommand{\fourAngle}[6]{\text{$#1\langle#2\rightarrow#3,#4,#5\rightarrow#6\rangle$}}
\newcommand{\threeAngle}[5]{\text{$#1\langle#2\rightarrow#3,#4,#5\rangle$}}

\newcommand{\Req}[5]{\fourAngle{Req}{#1}{#2}{#3}{#4}{#5}}
\newcommand{\Resp}[5]{\fourAngle{Resp}{#1}{#2}{#3}{#4}{#5}}
\newcommand{\RespOld}[4]{\threeAngle{Resp}{#1}{#2}{#3}{#4}}

We already gave the ordering requirements between messages (Figure
\ref{ordering}), and which messages should not be blocked by other
messages (Figure \ref{blocking}). These are the minimum constraints to
ensure correctness and deadlock-freedom of the hierarchical coherence
protocols. Any practical implementation will restrict the ordering between
messages further, but they must ensure that the messages that should not be
blocked according to Figure \ref{blocking} are not blocked. In this
section, we will discuss several such implementations. Note that the
constraints in Figures \ref{ordering} are on messages for the same
address; messages for different addresses can be processed in any order. This
permits implementation of non-blocking caches using MSHRs.

\subsection{Virtual channels to prevent inadmissible blocking}

Since responses should not be blocked by requests (Constraint \ref{blockresp}),
the requests and responses should be in separate virtual channels. Since a
request from a child of a cache node should not block a request from the parent
of that node (Constraint \ref{blockreq}), the requests from the children to a
node and requests from the parent to the node should use separate virtual
channels.

Between any two levels in the hierarchy, it is enough to have just one virtual
channel for requests going from the higher to the lower level and one for
requests going from the lower to the higher level. %For example, in the
%hierarchy shown in Figure \ref{fig:vc}, only 6 virtual channels are required.
In general, if the memory system has $N$ levels, only $2N$ virtual channels are
required for requests, irrespective of the number of nodes in each level, the
number of children in each node, \etc. The formal proof for this requirement is
beyond the scope of this paper, but here's the outline. As long as a request,
which is still not removed from a virtual channel, does not produce more
requests that use the same virtual channel, the system will not deadlock. In
the memory hierarchy, a request from level $x$ to level $y$ does not produce
another request from level $x$ to $y$. So one virtual channel between every two
levels, in each direction, is sufficient to avoid network-induced deadlocks.

Since a response does not produce more messages and can always be dequeued from
the network, just having one (separate) virtual channel, $RespChannel$, for
responses in the entire system is sufficient to avoid deadlocks. This is
irrespective of the number of levels in the memory hierarchy, the number of
caches in each level, \etc.

\subsection{Guaranteeing ordering requirements using point-to-point FIFO
ordering in the network}

If the network provides FIFO ordering for messages sent between the same nodes
using the same virtual channel, and if all responses between two nodes use the
same virtual channel for each direction always, Constraint \ref{orderresp} is
guaranteed. As noted above, since one virtual channel is sufficient to carry
responses for the entire system, a point-to-point FIFO ordering for that
virtual channel is sufficient.

However, in order to guarantee Constraint \ref{orderreq}, FIFO ordering must be
guaranteed even for messages sent on different virtual channels between the
same two nodes (since requests and responses should go in at least two separate
virtual channels). Guaranteeing point-to-point FIFO ordering for messages on
different virtual channels is expensive.

\subsection{Guaranteeing ordering requirements without a point-to-point FIFO
ordering in the network}
\label{sec:nofifoorder}

Instead of relying on the network to provide a point-to-point FIFO ordering for
messages on different virtual channels, Constraint \ref{orderreq} can be
guaranteed by the protocol.

Request messages now have the format \Req{src}{dst}{addr}{fromState}{toState}.
A downgrade request for an address from a parent to a child contains the
directory state of the parent for the address just before the request was sent,
in addition to value of the state to downgrade to. Similarly an upgrade request
for an address from child to its parent should contain the state of the child
for the address just before the request was sent, in addition to the value of
the state to upgrade to.

When a child $c$ receives a request \Req{P}{c}{a}{x}{y} from the parent, if
$state[c][a] < x$, then handling of the request is postponed by the child till
it receives a response \RespOld{P}{c}{a}{z} from the parent, where $z \ge x$.
The formal proof is beyond the scope of the paper; the intuition here is that
\RespOld{P}{c}{a}{z} must have been sent by the parent before sending
\Req{P}{c}{a}{x}{y}. Similarly, when a parent receives request
\Req{c}{P}{a}{x}{y} from child $c$, if $dir[c][a] > x$, then handling of the
request is postponed by the parent till it receives a response
\RespOld{c}{P}{a}{z} from $c$, where $z \le x$.

Response messages sent from a child to its parent can also avoid relying on the
point-to-point FIFO ordering in the network using a similar approach. Response
messages from the child to its parent will now have the format
\Req{src}{P}{addr}{fromState}{toState}. A response \Resp{c}{P}{a}{x}{y} is not
handled by the parent if $dir[c][a] > x$ until it receives another response
\Resp{c}{P}{a}{w}{z} where $z \le x$.

One issue with this for responses is that since only one virtual channel
$RespChannel$ is used for responses throughout the system, a response sent
later for an address from a child to its parent can block a response for the
same address from the same child to its parent sent earlier. This violates
Constraint \ref{orderresp}. If separate virtual channels are used for each type
of response sent by children, classified by $toState$ \ie the value of the
coherence state that a child sending the response has downgraded to, Constraint
\ref{orderresp} will be satisfied. Thus we have one virtual channel,
$RespChannel_{toState}$, for each $toState$. The responses sent by a parent to
its children can use any of the $RespChannel_{toState}$ virtual channels, since
these responses have no ordering requirements, and can share the virtual
channel used for responses sent from children to their parents. Thus, for an
MSI protocol, just two virtual channels are required ($RespChannel_I$ and
$RespChannel_S$).

The most common scenario is when a child sends responses \emph{voluntarily}
only for evictions because of replacement; in all other cases, it only sends
responses on receiving requests from the parent. Since the parent does not send
a second request for the same addres before receiving the response for the
first request, every response from the child, except a response to downgrade to
state $I$ because of eviction, will be received by the parent in FIFO order.
Thus just 2 virtual channels, one for responses to downgrade to $I$ and one for
responses to downgrade to any other state is sufficient, even for protocols
other than MSI.

The implementation described above avoids the need for sending \emph{ack}
messages to guarantee the required ordering, thus reducing the number of
messages as well as unblocking a cache line (for handling further requests)
without waiting for acks.

### 2012-11-14 paper/isca40format/Invariants.tex
stuff

@@
 
+%\subsection{Desired goals}
+%\begin{enumerate}
+%\item Permissions can be Nothing, Read (only), Read and Write, \ie $\{\}, {R}, {R, W}$
+%\item To upgrade = to gain more permissions
+%\item To downgrade = to lose permissions
+%\item If the child cache wants to gain some permissions for an address, it can
+%do so only after the parent directory has authorized it
+%\end{enumerate}
+%
+%\begin{enumerate}
+%\item Permissions that a child cache has for an address is a subset of the
+%permissions that the parent directory believes the child has for that address
+%\item If the parent directory wants a child cache to lose some permissions for
+%an address, there is a message that the directory can send notifying the child
+%of the directory's intentions (this message is called a downgrade request)
+%\item If the cache wants to gain some permissions for an address, there is a
+%message that the child can send notifying the directory of the child's
+%intentions (this message is called an upgrade request)
+%\item If a node wants to send a message to some node, then it can send such the
+%message eventually (\ie it can never be blocked from sending the message
+%forever)
+%\item If the parent directory has sent a downgrade request for an address $a$
+%to a child cache $c$, the directory will eventually get a message from $c$
+%notifying the directory that the cache has given up the required permissions
+%for address $a$ in $c$
+%\item If the child cache has sent an upgrade request for an address $a$ to
+%parent directory, the child will eventually get a message from the directory
+%authorizing the upgrade
+%\end{enumerate}
+%
 \newcommand{\threeAngle}[4]{\text{$#1\langle#2,#3,#4\rangle$}}
@@
 \newcommand{\Resp}[3]{\threeAngle{Resp}{#1}{#2}{#3}}
-\newcommand{\Inv}{\text{$Inv$\;}}
+\newcommand{\Inv}{\text{$Inv$} }
+\newcommand{\Dirty}{\text{$Dirty$} }
+\newcommand{\True}{\text{$True$} }
+\newcommand{\False}{\text{$False$} }
+
+The caches are assumed to be organized in the form of a tree
+(Figure~\ref{fig:hierarchy}). The leaves of this tree represent the L1 caches,
+with each L1 cache connected to exactly one core. The root of the tree is a
+single Last-level cache (LLC), connected to the main memory. A (cache) node can
+communicate only with its parents or with its children via a request or
+response message whose format is in Figure~\ref{message}.
+
+\begin{figure}
+%\centering
+\begin{minipage}{5.5cm}
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.5]{hierarchy}
+\caption{Cache hierarchy}
+\label{fig:hierarchy}
+\end{figure}
+\end{minipage}
+\begin{minipage}{3cm}
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.53]{network}
+\caption{One possible network organization}
+\label{fig:network}
+\end{figure}
+\end{minipage}
+\end{figure}
+
+\begin{figure}
+\begin{tabularx}{\linewidth}{|cX|}
+\hline
+\multicolumn{2}{|c|}{\Req{n}{a}{x}, \Resp{n}{a}{x}}\\
+\hline
+$n:$ & Source or destination node ID (parent $P$, or a child)\\
+$a:$ & Address of the line corresponding to the request\\
+$x:$ & Value from the set of coherence states\\
+\hline
+\end{tabularx}
+\caption{Incoming and outgoing message}
+\label{message}
+\end{figure}
+
+The restrictions of which messages can not block each other, and the order in
+which messages should be received and handled is given in
+Figure~\ref{receiveOrder}.
+
+One way to implement the ordering restrictions is to have a network which
+guarantees FIFO ordering between any two nodes. Having one virtual channel for
+the messages from the parent and another for the messages from the children
+enforces the blocking restriction between requests from the parent and the
+children. Among messages from the parent (or the children), having two virtual
+channels, one for requests and one for responses, with responses having higher
+priority than requests enforces the blocking restrictions and the ordering
+requirements between requests and responses. Such an organization is shown in
+Figure~\ref{fig:network}.
+
+Note that the organization in Figure~\ref{fig:network} requires finishing
+processing of a request from a child before starting the processing of the
+request from another child, or for a different address. It makes the node
+behave more like a blocking cache, whose performance may be low. The
+restrictions given in Figure~\ref{receiveOrder} however do not impose this
+organization. One can process requests to different addresses in any order,
+using MSHRs.
+
+The ordering requirements of Figure~\ref{receiveOrder} can be enforced using
+prioritized virtual channels which guarantee FIFO ordering between messages
+sent from the same source to the same destination (as discussed above).
+However, as we show later, the ordering requirements can be enforced even when
+the underlying network does not guarantee FIFO ordering between two nodes.
+
+\begin{figure}
+\begin{tabularx}{\linewidth}{|X|}
+\hline
+\multicolumn{1}{|c|}{Blocking restrictions}\\
+\hline
+\begin{enumerate}
+\item A request from a child should not block (\ie prevent start of the
+handling of) a request from the parent
+\item Requests for an address $a$ should not block (\ie prevent reception of)
+responses for address $a$, even from a different source
+\end{enumerate}\\
+\hline
+\multicolumn{1}{|c|}{Ordering requirements}\\
+\hline
+\begin{enumerate}
+\item A response for an address $a$ from a child $c$ should not be received
+before a response from the same $c$ and the same $a$ sent earlier has been
+received and removed
+\item The handling of a request for an address $a$ from a source $n$ should not
+be started before a response from the same $n$ for the same $a$ sent earlier
+has been received and removed
+\end{enumerate}\\
+\hline
+\end{tabularx}
+\caption{Blocking restrictions and order of receiving messages}
+\label{receiveOrder}
+\end{figure}
 
@@
 
-\begin{figure}
-\begin{tabularx}{\linewidth}{|l|l|X|}
-\hline
-\multicolumn{3}{|c|}{\Req{n}{a}{x}}\\
-\hline
-\multirow{4}{*}{$n = P$} & Incoming & Request from parent for changing my
-                                      $state[a]$ to $x$\\
-\cline{2-3}
-                         & Outgoing & Request to parent for permission to change
-                                      my $state[a]$ to $x$\\
-\hline
-\multirow{4}{*}{$n \neq P$} & Incoming & Request from child $n$ for permission
-                                         to change $n$'s $state[a]$ to $x$\\
-\cline{2-3}
-                            & Outgoing & Request to child $n$ to change $n$'s
-                                         $state[a]$ to $x$\\
-\hline
-\multicolumn{3}{|c|}{\Resp{n}{a}{x}}\\
-\hline
-\multirow{4}{*}{$n = P$} & Incoming & Response from parent granting permission
-                                      to change my $state[a]$ to $x$\\
-\cline{2-3}
-                         & Outgoing & Response to parent indicating that my
-                                      $state[a]$ has changed to $x$\\
-\hline
-\multirow{4}{*}{$n \neq P$} & Incoming & Response from child indicating that
-                                         $n$'s $state[a]$ has changed to $x$\\
-\cline{2-3}
-                            & Outgoing & Response to child $n$ granting permission
-                                         to change $n$'s $state[a]$ to $x$\\
-\hline
-\end{tabularx}
-\caption{Incoming and outgoing messages}
-\label{table:messages}
-\end{figure}
+%\begin{figure}
+%\begin{tabularx}{\linewidth}{|l|l|X|}
+%\hline
+%\multicolumn{3}{|c|}{\Req{n}{a}{x}}\\
+%\hline
+%\multirow{4}{*}{$n = P$} & Incoming & Request from parent for changing my
+%                                      $state[a]$ to $x$\\
+%\cline{2-3}
+%                         & Outgoing & Request to parent for permission to change
+%                                      my $state[a]$ to $x$\\
+%\hline
+%\multirow{4}{*}{$n \neq P$} & Incoming & Request from child $n$ for permission
+%                                         to change $n$'s $state[a]$ to $x$\\
+%\cline{2-3}
+%                            & Outgoing & Request to child $n$ to change $n$'s
+%                                         $state[a]$ to $x$\\
+%\hline
+%\multicolumn{3}{|c|}{\Resp{n}{a}{x}}\\
+%\hline
+%\multirow{4}{*}{$n = P$} & Incoming & Response from parent granting permission
+%                                      to change my $state[a]$ to $x$\\
+%\cline{2-3}
+%                         & Outgoing & Response to parent indicating that my
+%                                      $state[a]$ has changed to $x$\\
+%\hline
+%\multirow{4}{*}{$n \neq P$} & Incoming & Response from child indicating that
+%                                         $n$'s $state[a]$ has changed to $x$\\
+%\cline{2-3}
+%                            & Outgoing & Response to child $n$ granting permission
+%                                         to change $n$'s $state[a]$ to $x$\\
+%\hline
+%\end{tabularx}
+%\caption{Incoming and outgoing messages}
+%\label{table:messages}
+%\end{figure}
 
@@
 \hline
-A response for address $a$ can be sent only if the line for $a$ is present in
-the current node\\
+\begin{enumerate}
+\item[] A response for address $a$ can be sent only if the line for $a$ is present in
+the current node, \ie $isPresent(a)$
+\end{enumerate}\\
 \hline
-\multicolumn{1}{|c|}{Response to child $c$ for address $a$}\\
+\multicolumn{1}{|c|}{Response to child $c$ for address $a$, \Resp{c}}\\
 \hline
@@
 \begin{enumerate}
-\item If there is a pending response from the parent for address $a$, a
-response to the parent for $a$ can be sent only if a request from
-the parent for $a$ is being handled
-\item If there is no pending response from the parent for address $a$, a
-response to the parent for $a$ can be sent any time
-\item Response \Resp{P}{a}{x} can be sent only if $x < state[a]$ and $\forall
-c, x \ge dir[a][c]$
+\item If there is a pending response from the parent for address $a$, \ie $
+isWait(P, a)$, a response to the parent for $a$ can be sent only if a request
+from the parent for $a$ is being handled, \ie $isHandling(P, a)$
+\item If there is no pending response from the parent for address $a$, \ie
+$\neg isWait(P, a)$, a response to the parent for $a$ can be sent any time
+\item Response \Resp{P}{a}{x} can be sent only if $x < state[a]$
+% and $\forall c, x \ge dir[a][c]$
 \end{enumerate}\\
@@
 \hline
-A response for address $a$ can be sent only if the line for $a$ is present in
-the current node\\
+\begin{enumerate}
+\item[] A response for address $a$ can be sent only if the line for $a$ is
+present in the current node, \ie $isPresent(a)$
+\end{enumerate}\\
 \hline
@@
 \hline
-\multicolumn{1}{|c|}{On receiving a response from child $c$ for address $a$}\\
+\multicolumn{1}{|c|}{On receiving a response \Resp{c}{a}{x}}\\
 \hline
@@
 \item Response \Resp{c}{a}{x} is removed
+\item $isWait(c, a)$ is set \False
 \end{enumerate}\\
 \hline
-\multicolumn{1}{|c|}{On receiving a response from parent $c$ for address $a$}\\
+\multicolumn{1}{|c|}{On receiving a response \Resp{P}{a}{x}}\\
 \hline
@@
 \item A request can be sent to child $c$ for address $a$ only if there is no
-pending response from child $c$ for $a$
+pending response from child $c$ for $a$, \ie $\neg isWait(c, a)$
 \item Request \Req{c}{a}{x} can be sent only if $x < dir[a][c]$
@@
   \begin{enumerate}
-  \item there is no pending response from the parent for address $a$, and
-  \item line in address $a$ is not being replaced
+  \item there is no pending response from the parent for address $a$, \ie
+  $\neg isWait(P, a)$, and
+  \item line in address $a$ is not being evicted, \ie $isEvicting(a)$
   \end{enumerate}
@@
 \begin{enumerate}
-\item If $a$ is present in the node, handling of request \Req{c}{a}{x} can be
-started only if
+\item If address $a$ is present in the node, \ie $isPresent(a)$, handling of
+request \Req{c}{a}{x} can be started, \ie $isHandling(c, a)$ can be set \True
+only if
   \begin{enumerate}
   \item no request \Req{n}{a}{x} from any node $n$ for address $a$ is being
-  handled
-  \item address $a$ is not currently being replaced
+  handled, \ie $\forall n, \neg isHandling(n, a)$
+  \item address $a$ is not currently being evicted, \ie $\neg isEvicting(a)$
   \end{enumerate}
-\item If $a$ is not present in the node, handling of request \Req{c}{a}{x} can be
-started only if there is a line with address $a'$ such that
+\item If $a$ is not present in the node, \ie $\neg isPresent(a)$, handling of
+request \Req{c}{a}{x} can be started \ie $isHandling(c, a)$ can be set \True
+only if there is a line with address $a'$ such that
   \begin{enumerate}
   \item no request \Req{n}{a'}{x} from any node $n$ for address $a'$ is being
-  handled
-  \item address $a'$ is not currently being evicted
+  handled, \ie $\neg isHandling(n, a')$
+  \item address $a'$ is not currently being evicted, \ie $\neg isEvicting(a')$
   \end{enumerate}
-\Req{n}{a'}{x} has started
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item Start handling the request \Req{c}{a}{x}
-\item If $a$ is not present in the node, get a line with address $a'$ such that
+\item Handling of the request \Req{c}{a}{x} has started, \ie $isHandling(c, a)$
+is set \True
+\item If $a$ is not present in the node, \ie $\neg isPresent(a)$, a line with
+address $a'$ is found such that
   \begin{enumerate}
   \item no request \Req{n}{a'}{x} from any node $n$ for address $a'$ is being
-  handled
-  \item address $a'$ is not currently being evicted
+  handled, \ie $\forall n, \neg isHandling(n, a')$
+  \item address $a'$ is not currently being evicted, \ie $\neg isEvicting(a')$
   \end{enumerate}
-Mark $a'$ as being evicted
+$a'$ is marked as being evicted, \ie $isEvicting(a')$ is set \True
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item If $state[a'] <= \Inv$, replace address of the line for $a'$ as $a$,
-$state[a']$ as \Inv and $\forall c, dir(c, a)$ as \Inv
-\item Unmark $a'$ as being evicted
+\item If $state[a'] <= \Inv$, address $a'$ for the line is replaced by $a$, \ie
+$isPresent(a')$ is set \False and $isPresent(a)$ is set \True
+\item $state[a']$ is set \Inv and $\forall c, dir(c, a)$ is set \Inv
+\item $a'$ is no longer being evicted, \ie set $isEvicted(a')$ to \False
 \end{enumerate}\\
@@
 \begin{figure}
+\centering
+\begin{subfigure}{\linewidth}
+\centering
+\includegraphics[scale=0.5]{reqFromP}
+\caption{Stages in handling a request from parent}
+\label{fig:reqFromP}
+\end{subfigure}
+
+\begin{subfigure}{\linewidth}
+\centering
+\includegraphics[scale=0.5]{reqFromC}
+\caption{Stages in handling a request from a child}
+\label{fig:reqFromC}
+\end{subfigure}
+\end{figure}
+
+\begin{figure}
 \begin{tabularx}{\linewidth}{|X|}
@@
 \end{tabularx}
-\caption{Order of receiving messages}
+\caption{Atomicity requirements}
 \label{atomicity}
 \end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\begin{enumerate}
-\item A request from a child should not block (\ie prevent start of the
-handling of) a request from the parent
-\item Requests for an address $a$ should not block (\ie prevent reception of)
-responses for address $a$, even from a different source
-\item A response for an address $a$ from a child $c$ should not be received
-before a response from the same $c$ for the same $a$ sent earlier has been
-received and removed
-\item The handling of a request for an address $a$ from a source $n$ should not
-be started before a response from the same $n$ for the same $a$ sent earlier
-has been received and removed
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Order of receiving messages}
-\label{mesgOrder}
-\end{figure}

### 2012-11-14 paper/isca40format/Msi.tex
stuff

\section{Informal description of the proof}
\label{sec:informal-proof}

\newcommand{\msg}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
\newcommand{\Req}[4]{\msg{Req}{#1}{#2}{#3}{#4}}
\newcommand{\Resp}[4]{\msg{Resp}{#1}{#2}{#3}{#4}}
\newcommand{\Data}[4]{\msg{Data}{#1}{#2}{#3}{#4}}
\newcommand{\Mesg}[4]{\msg{Mesg}{#1}{#2}{#3}{#4}}
\newcommand{\FirstInv}{\emph{FirstInv}}
%\input{Atomic}

Let us walk through our design methodology for our message-passing hierarchical
directory based protocols.

\subsection{Message-passing directory-based MSI protocol for 2-levels, no
replacement evictions}
For the sake of explanation, let us start with an MSI protocol, with just two
levels of the memory hierarchy -- the L1 caches and the main memory. We will
eventually build realistic protocols with various optimizations.

Let us assume that the caches have enough capacity so that no line gets evicted
for the sake of replacement. Lines may however get evicted to enforce the
single-writer property for cache coherence, \ie when one L1 cache has
permission to modify the data for an address, no other cache has read or write
permission for that address.

An L1 cache stores the information given in Figure \ref{table:lineinfo}. The
coherence state for an address denotes the permission that a cache has for the
address: $M = \{R, W\}, S = \{R\}, I = \{\}$. State $I$ is akin to the cache not
having the data corresponding to $a$, since it can neither read or write that
cache line.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for
address $a$\\
$data[c][a]$ & data in cache $c$ for address $a$\\
\hline
\end{tabularx}
\caption{Values stored by a cache for an address}
\label{table:lineinfo}
\end{figure}

In order to serve a request from the processor, if a cache has to \emph{upgrade}
its state for address $a$, \ie change from $x$ to $y$ where $x > y$
(Figure \ref{table:fns}), it can do so only after ensuring that the new state is
\emph{compatible} with the states of other caches for that address, \ie other
caches are in a state less than or equal to $toCompatible(y)$
(Figure \ref{table:fns}). This is needed to ensure the single-writer property of
the MSI protocol -- if a cache's state is $M$ for an address $a$, every other cache
must be in $I$ for address $a$. But, the caveat is that the L1 cache can access
only its own state and data, as opposed to that of other caches.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$toCompatible(s)$ & If the state of one cache is $s$ for address $a$, the
highest state of another cache for the same $a$
$\newline
toCompatible(M) = I,\newline
toCompatible(S) = S,\newline
toCompatible(I) = M
$\\
\hline
$x > y$ & Denotes if state $x$ has more permissions than state $y$.
$M > S, M > I, S > I$. Relations $\le, >, \ge$ are also defined appropriately\\
\hline
\end{tabularx}
\caption{Abstraction functions}
\label{table:fns}
\end{figure}

The memory controller has access to the data from the memory but not to the
state and data of the L1 caches. Instead, it only has access to a directory
which has a version of the coherence states of the caches for each address. The
version of the state in the directory may not be up to date, \ie may not be the
same as the one in the caches, all the time. Figure \ref{table:dirinfo} shows
the information stored in the memory and directory. We will use the word memory
and directory interchangeably for the 2-level protocol. The caches can not
communicate with each other but can communicate with the directory via messages.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$dir[c][a]$ & directory's version of the coherence state of cache $c$ for address $a$\\
$memory[a]$ & data in the memory address $a$\\
\hline
\end{tabularx}
\caption{Values stored by a directory and memory for an address}
\label{table:dirinfo}
\end{figure}

Let's us assume that we have a system which guarantees the invariants given
in Figure \ref{FirstInv}.
\floatstyle{boxed}
\restylefloat{figure}

\begin{figure}
\begin{enumerate}
\item The cache can upgrade $state[c][a]$ from $x$ to $y$ where $y > x$, \ie
\emph{downgrade} its directory state only after it receives the appropriate
authorization from the directory for address $a$. \label{cwaitresp}
\item If the cache wants to upgrade $state[c][a]$ to $x$, there is a
message \Req{c}{Dir}{a}{x} which cache $c$ can send to the directory requesting
the upgrade authorization, and the cache will eventually be able to send
the request. \label{csendreq}
\item If the directory receives an upgrade request \Req{c}{Dir}{a}{x} from cache
$c$ for address $a$, then just before receiving the request, $dir[c][a] < x$.
\label{drecvreq}
\item The directory can change $dir[c][a]$ from $x$ to $y$ where $y < x$, \ie
\emph{downgrade} its directory state only after it receives the appropriate
authorization from cache $c$ for address $a$. \label{dwaitresp}
\item If the directory wants to downgrade $dir[c][a]$ to $x$, there is a
message \Req{Dir}{c}{a}{x} which the directory can send to cache $c$ requesting
the downgrade authorization, and the directory will eventually be able to send
the request. \label{dsendreq}
\item If the directory has sent a downgrade request for an address $a$ to cache
$c$, it will eventually get the downgrade authorization response for address $a$
from cache $c$ to downgrade to $x$, where $x$ is less than or equal to the state
to which the directory sent a request to downgrade to. \label{drecvresp}
\item If $dir[c][a] = M$ just before receiving a downgrade authorization from
cache $c$ for address $a$, then the cache would have sent the data along with
the authorization. \label{drecvdata}
\item If the directory wants to send an authorization message to upgrade the
state of address $a$ in cache $c$ to $x$ (\Resp{Dir}{c}{a}{x}), it will
eventually be able to send it. \label{dsendresp}
\item The directory can send the data from the memory along with the upgrade
authorization message if it wants to. \label{dsenddata}
\item The directory's version of the cache's state is always conservative, \ie
$\forall c, a,\; dir[c][a] \ge state[c][a]$. \label{conservative}
\end{enumerate}
\caption{Initial set of invariants \FirstInv}
\label{FirstInv}
\end{figure}

\floatstyle{plain}
\restylefloat{figure}

We will walk through how the MSI coherence protocol (with no replacement
evictions) can be implemented using the above invariants and argue why it is a
correct implementation.

The directory's role is to ensure the single-writer property and that the last
updated data is being read. In order to implement the cache coherence protocols
using the invariants of Figure \ref{FirstInv}, the directory's ensures that at
all times, if for some $c, dir[c][a] = M$, the for every other $c', dir[c'][a] =
I$.

If a cache gets a request from the core which can not be handled because of lack
of permissions, it (eventually) sends an upgrade request to the directory.
\FirstInv.\ref{csendreq} ensures that such a request can eventually be sent.

If the directory receives a request from cache $c$ to upgrade the state of
address $a$ to $x$, because of \FirstInv.\ref{drecvreq}, $dir[c][a] < x$ and
since $cache[c][a] \le dir[c][a]$ because of \FirstInv.\ref{conservative},
$cache[c][a] < x$ and hence the directory must send an upgrade authorization
response. The directory sends a request \Req{Dir}{c'}{a}{toCompatible(x)} to
every cache $c'$ for which $dir[c'][a] > toCompatible(x)$. Such a request can be
sent eventually because of \FirstInv.\ref{dsendreq}.

Because of \FirstInv.\ref{drecvresp}, the directory will eventually receive
responses from each of the requested caches $c'$
(\Resp{c'}{Dir}{a}{toCompatible(x)}) authorizing the downgrade of $dir[c'][a]$
to $toCompatible(x)$. If $dir[c'][a] = M$ just before receiving the
authorization, then by \FirstInv.\ref{drecvdata}, the directory would also
receive data from $c'$. Because the directory maintains the single-writer
invariant, every other cache $c''$ would have $dir[c''][a] = I$ just before
receiving the authorization. This implies for every other cache $c''$,
$state[c''][a] = I$ by the \FirstInv.\ref{conservative} invariant, and hence
$c'$ has the most up-to-date copy for address $a$. Thus the memory is updated
with the received up-to-date copy for address $a$.

Once all the authorizations have been received, for every cache $c'$,
$dir[c'][a] \le toCompatible(x)$ (other than the one that made the request $c$).
Since the \FirstInv.\ref{conservative} invariant ensures that $cache[c'][a] \le
dir[c'][a]$, we have $cache[c'][a] \le toCompatible(x)$, ensuring the
single-writer property (because $toCompatible(M) = I$). The directory can now
change $dir[c][a]$ to $x$ and send \Resp{Dir}{c}{a}{x} authorizing $c$ to
upgrade state of address $a$ to $x$ (\FirstInv.\ref{dsendresp} ensures that such
a message can be sent). Just before sending the authorization, if $dir[c][a] =
I$, it implies that $state[c][a] = I$ because of the
\FirstInv.\ref{conservative} invariant.  Thus the cache doesn't have data
corresponding to address $a$ and the directory has to supply the most up to date
version of the address (from memory, since the memory is up to date), which it
can because of \FirstInv.\ref{dsenddata}.

The \FirstInv\ invariants essentially makes it straight forward to design the
behavior of the directory . For the sake of argument, let
\FirstInv.\ref{drecvreq} not be guaranteed. Consider a scenario the cache $c$
which has sent a request \Req{c}{Dir}{a}{M} earlier but hasn't received the
upgrade authorization response yet. Meanwhile, $c$ sends another upgrade request
\Req{c}{Dir}{a}{S}. Let us assume that the directory processed the requests in
the order in which they were sent. Once the directory has finished processing
the first request, it changes its directory state $dir[c][a]$ to $M$. When it
receives the second request, it is not easy to decide if the directory has to
drop the request or to send another response authorizing the cache to change to
$S$.

The difficulty in message passing protocols comes from the fact that it is
difficult to guarantee the above invariants, even for a 2-level memory hierarchy
discussed above as shown in the examples below.

\subsubsection*{Example 1: Inability to send messages because of network
congestion:}
For instance, the ability of a cache or the directory to send messages
(\FirstInv.\ref{dsendreq}, \FirstInv.\ref{dsendresp}, \FirstInv.\ref{csendreq},
\etc) depends upon the availability of free buffers in the network; if the
network remains congested all the time, then it would not be possible to send the
messages, and the system would deadlock.

\subsubsection*{Example 2: Multiple upgrade requests from the cache for the same
address:}
If a cache is allowed to make multiple requests to upgrade the state of an
address (say to $M$ and then to $S$) before receiving the upgrade
authorizations, then it might violate \FirstInv.\ref{drecvreq} if the requests
were processed by the directory in the order they were received.

\subsubsection*{Example 3: On sending an upgrade request, the cache waits till
it gets an authorization:}
If the behavior of the caches were such that if a cache $c$ has sent an
upgrade request \Req{c}{Dir}{a}{x} to the directory, it keeps waiting for the
upgrade authorization response \Resp{Dir}{c}{a}{x} before processing any
requests from the directory for address $a$, then the cache coupled with our
simple directory implementation would deadlock. Consider a scenario with two
caches $X$ and $Y$, with both the caches in state $S$ for address $a$. Both the
caches got a store request from the processor and want to upgrade their states
to $M$, and hence send \Req{X}{Dir}{a}{M} and \Req{Y}{Dir}{a}{M}, respectively,
to the directory. Let us say the directory receives \Req{Y}{Dir}{a}{M} first,
and in order to process it, according to our implementation of the directory,
sends a downgrade request \Req{Dir}{X}{a}{I} to cache $X$. But the cache 
does not process any request from the directory for address $a$, according to
our assumption, and hence will never send the downgrade authorization
\Resp{X}{Dir}{a}{I} to the directory, violating \FirstInv.\ref{drecvresp}, and
hence causing a deadlock.

\subsubsection*{Example 4: At the directory, authorization responses are blocked
by upgrade requests:}
Consider a scenario where two caches $X$ and $Y$ are both in state $S$ for address
$a$. $Y$ got a store request from the processor and sends an upgrade request
\Req{Y}{Dir}{a}{M} to the directory. The directory receives that request, and
sends a downgrade request \Req{X}{Dir}{a}{I} to $X$. $X$ sends a downgrade
authorization \Resp{X}{Dir}{a}{I} to the directory on receiving the downgrade
request. But the directory can not receive the authorization because it is
blocked by other requests (including \Req{Y}{Dir}{a}{M}). This violates
\FirstInv.\ref{drecvresp} and hence creates a deadlock.

These examples show that if the \FirstInv\ invariants are violated, then it
either breaks cache coherence (by creating deadlocks) or it makes correctly
designing the directory very difficult.

Before we show how the \FirstInv\ invariants can be guaranteed (which requires
the description of the behavior of the L1 caches), let us consider a
slightly more complicated system where caches can evict lines for replacement.

\subsection{Message-passing directory-based MSI protocol for 2-levels, with
replacement evictions}
We again have just two levels of memory hierarchy -- the L1 caches and the main
memory. But the caches have finite capacity. If receives a request from the core
for an address $a$ and has no space to accommodate address $a$, it evicts
another address $a'$ to make room for $a$.

In the previous scenario, where the caches didn't have to evict for replacement,
a cache only had to send downgrade authorizations to the directory on receiving
a corresponding request. But in the new scenario, a cache can send a downgrade
notification to the directory on evicting an address. Such a notification is
useful in keeping the directory's version of the cache's state up to date, so we
will restrict the family of directory-based protocols that we discuss to those
where the caches send such a notification, voluntarily.

The directory handles this notification in the exact manner by which it handled
a downgrade authorization response from the cache, so the notification message
is the same as a downgrade authorization message from cache $c$ for authorizing
the downgrade of the $dir[c][a]$ to $I$ (\Resp{c}{Dir}{a}{I}).

We will now discuss a behavior of the L1 cache that would violate the
invariant that the directory's version of the states of the caches for each
address is conservative.

\subsubsection*{Example 5: Cache sends voluntary notifications
(\Resp{c}{Dir}{a}{I}) indiscriminately}
Consider a system with two caches $X$ and $Y$. Let $state[X][a] = S, state[Y][a]
= I$ and let $dir[X][a] = S, dir[Y][a] = I$. $X$ receives a store request from
the core, and hence sends an upgrade request \Req{X}{Dir}{a}{M} to the
directory. Say $X$ was a non-blocking cache, and hence receives another load
request for address $a'$ from the core before receiving an upgrade authorization
for address $a$. $X$ does not have space in the cache for $a'$, and it decides
to evict address $a$, and sends \Resp{X}{Dir}{a}{I} to the directory, changing
$state[c][a]$ to $I$.  Meanwhile, the directory receives the request
\Req{X}{Dir}{a}{M} and since $dir[Y][a] \le I$, it responds to $X$ with an
upgrade authorization \Resp{Dir}{X}{a}{M} (along with the data). Later, the
directory receives the notification \Resp{X}{Dir}{a}{I} from $X$, and downgrades
$dir[X][a]$ to $I$. Meanwhile, $X$ receives the upgrade authorization
\Resp{Dir}{X}{a}{M} and changes $state[X][a]$ to $M$. Now,
\FirstInv.\ref{conservative} is broken, since $state[X][a] = M$ and $dir[X][a] =
I$. Let $Y$ get a load request from its core, and hence sends a
\Req{Y}{Dir}{a}{S} to the directory. The directory on receiving the request,
sends an upgrade authorization to $Y$ because $dir[X][a] = I$. This violates the
single-writer property and hence coherence.

### 2012-11-14 paper/isca40format/Stuff.tex
stuff

\begin{enumerate}
\item A cache can upgrade its state for an address $a$ only after it gets an
appropriate authorization message for address $a$ from the directory.
\item If a cache wants to upgrade its state to $x$ for an address, it will
eventually be able to send a request message (\Req{c}{Dir}{a}{x}) to the directory
intimating its intentions. \label{csendreq}
\item If the directory wants to send an authorization message to upgrade the
state of address $a$ in cache $c$ to $x$ (\Resp{Dir}{c}{a}{x}), it will eventually
be able to send it. \label{dsendresp}
\item If the directory wants to send data from the memory for address $a$ to
cache $c$ (\Data{Dir}{c}{a}{d}), it will eventually be able to send it.
\label{dsenddata}
\item A directory can \emph{downgrade} its version of the state of a cache $c$
for an address $a$, \ie change $dir[c][a]$ from $x$ to $y$ where $x > y$, only
after getting the appropriate authorization for address $a$ from cache $c$.
\item If a directory wants to downgrade its version of the state for a cache $c$
and address $a$ to $x$ (\Req{Dir}{c}{a}{x}), it will eventually be able to send a
\emph{request} message to the cache intimating its intentions. \label{dsendreq}
\item If the directory has sent a downgrade request for an address $a$ to cache
$c$, it will eventually get an authorization message for the downgrade of
address $a$ from $c$ (\Resp{c}{Dir}{a}{x}). If $dir[c][a] = M$ just before
receiving the authorization, the directory will also receive data from $c$ along
with the authorization. \label{drecvresp}
\item If a cache wants to send its data for address $a$ to the memory
(\Data{c}{dir}{a}{d}), it will eventually be able to send it. \label{csenddata}
\item If a cache wants to downgrade its state to $x$ for an address, it will
eventually be able to send the notification for downgrade to the directory.
\item The directory's version of a cache's state for an address is conservative
at all times, \ie $\forall a, c,\; state[c][a] \le dir[c][a]$. \label{conservative}
\end{enumerate}

### 2012-11-14 paper/isca40format/Variations.tex
stuff

\section{Implementing variations of basic MSI protocol}
\label{sec:variations}

%\renewcommand{\fourAngle}[6]{\text{$#1\langle#2\rightarrow#3,#4,#5\rightarrow#6\rangle$}}

\newcommand{\FwdReq}[2]{\text{$FwdReq\langle#1,#2\rangle$}}
\newcommand{\FwdAck}{\text{$FwdAck$}}

%\renewcommand{\Req}[5]{\fourAngle{Req}{#1}{#2}{#3}{#4}{#5}}
%\renewcommand{\Resp}[5]{\fourAngle{Resp}{#1}{#2}{#3}{#4}{#5}}

In this section we describe variations of the basic MSI hierarchical
message-passing directory-based protocol and show how our framework can be used
to design and argue their correctness easily.
%The
%arguments are based on the formally proven Invariants of Figure
%\ref{invariants}. The ordering requirements in Figure \ref{ordering} are
%handled at the protocol level by sending the state at which the sender was in
%just before sending the message as discussed in Section \ref{sec:nofifoorder}

\subsection{MESI optimization}
We first describe the MESI variation of the message-passing MSI protocol for a
two-level hierarchy containing multiple L1 caches and directory shared by all
the L1 caches. On receiving a request from a cache $c$ to upgrade $c$'s state
for address $a$ to $S$, the directory sends permission to upgrade the state of
$c$ for address $a$ to ``exclusive'' ($E$), if the directory state of every
other cache for that line is $I$ (and because of our Invariant
\ref{conservative}, the coherence state of every other cache for that line is
also $I$). Once cache $c$ has upgraded to $E$, it can automatically upgrade to
$M$ on receiving a store request from the processor, without sending any
notification to the directory.  However, if some other cache $c'$ sends a
request to the directory for address $a$ to upgrade to state $S$, the directory
has to ensure that the cache $c$ has downgraded its state from $E$ to $S$. If
$c$ remains at $E$ without having upgraded to $M$, then it doesn't have to send
the data back to the memory, while downgrading to $S$.

From the directory's perspective, states $M$ and $E$ are equivalent -- the only
distinction is that the cache would not send back the data while downgrading the
state for an address from $E$. Thus MESI is simply an optimization over the MSI
protocol in which the cache maintains an extra dirty bit for each address, and
sends back the cache line data to the directory during a downgrade only if the
dirty bit is set (the dirty bit can not be set for an address if the
corresponding state is $S$). If the dirty bit is not set during a downgrade, the
cache sends back a message denoting ``data unmodified'' in the place of actual
data. In the hierarchical setting too, MESI is simply an optimization over MSI
in which a child sends data back to its parent during a downgrade only if the
dirty bit is set.

\subsection{Cache-intervention based protocols}
In the protocols discussed so far, caches can directly communicate only with
their immediate children or with its immediate parent. In cache-intervention
based protocols, communication and data transfer can also take place between
nodes other than the immediate children or the immediate parent. Our framework
can be easily extended to include cache-intervention based protocols where
sibling caches, \ie caches that share the same immediate parent, can communicate
with each other. We show 2 protocols with sibling cache-intervention: a) the MSI
protocol with intervention by a cache in the $M$ state and b) the MOSI protocol. 
Note that in both the protocols, the MESI optimization can be implemented
whenever a child has to transfer data to its parent.

We first describe the two protocols using a two-level hierarchy containing
multiple L1 caches and a directory shared by all the L1 caches, and later
describe the extension to a hierarchical system.

\subsubsection{Intervention by a cache in the $M$ state:}
A downgrade request sent from the directory to a cache $c$ for address $a$ can
sometimes be augmented with a forwarding request \FwdReq{fwdDest}{toState},
which directs the cache to send an upgrade response to $fwdDest$ to upgrade
$fwdDest$'s state for address $a$ to $toState$, on the directory's behalf. A
downgrade response from cache $c$ is also augmented with \FwdAck{}, which
acknowledges the directory that the requested forwarding has been fulfilled.
Note that normal downgrade requests from the directory without a forwarding
request and normal responses from the cache without \FwdAck{} are also sent.

When a directory receives a request from cache $c'$ for address $a$ to upgrade
to state $x$ ($x$ = $M$ or $S$), it determines if some other cache is in
state $M$ for address $a$. If no cache is in state $M$ for address $a$, then the
directory behaves in the same manner as MSI -- we will not discuss this here
again.  But if some cache $c$ is in state $M$ for address $a$, it sends a
request to cache $c$ to downgrade its state for address $a$ to $toCompatible(x)$
(either $I$ or $S$ for $x$ = $M$ or $S$, respectively) augmented with a
forwarding request \FwdReq{c'}{x}. This forwarding request is essentially a
response sent by the directory to cache $c'$ to upgrade the state of $c'$ for
address $a$ to $x$, but it is sent via the cache $c$. The directory waits for a
downgrade response along with a \FwdAck{} from $c$ before processing further
requests for $a$.

If cache $c$ receives the downgrade request from the directory along with the
forward request \FwdReq{c'}{x}, while it is still in state $M$ for address $a$,
$c$ will send a response to $c'$ authorizing the upgrade of $state[c'][a]$ to
$x$, along with the data associated with address $a$. It will also send a
downgrade response to the directory notifying that $c$ has downgraded state $a$
to $toCompatible(x)$ along with a \FwdAck{}; it doesn't transfer the data to the
directory/memory. However, if cache $c$ has already downgraded to a lower state
than $M$ for address $a$ (because of eviction due to replacement), then the
cache simply drops the request just like in the MSI protocol discussed earlier;
when $c$ downgraded its state from $M$ for address $a$, it must have sent a
downgrade response to the directory for address $a$ along with the data (subject
to MESI optimization), just like in the MSI protocol discussed earlier,
albeit without a \FwdAck{}.

If the directory receives a downgrade response for address $a$ from cache $c$
along with a \FwdAck{}, the directory knows that $c$ has sent a response to cache
$c'$ authorizing $c'$ to upgrade its state for $a$ to $x$. But if the directory
receives a downgrade response for address $a$ from cache $c$ without a \FwdAck{},
the directory knows that $c$ has not sent a response to cache $c'$. In this
case, the directory sends a response along with the data to $c'$ authorizing it
to upgrade its state for address $a$ to $x$. Figure \ref{msi-inter} shows the
order of messages transferred between the caches in the scenario when forwarding
happens as well the in the scenario when no forwarding happens.

$c'$ would eventually get the upgrade to $x$ response for address $a$ either
from the directory or the cache $c$ -- it can not distinguish between the two
cases.

The protocol given above satisfies all the constraints \ref{} to \ref{}, except,
potentially, Constraint \ref{orderreq}. Before the upgrade response for address
$a$ sent from $c$ to $c'$ has reached $c'$, $c'$ could have received a downgrade
request sent from the parent. This happens as follows (Figure
\ref{msi-fifo-needed}, with the numbers indicating the order in which the
messages are sent, for outgoing edges, and received for incoming edges). Cache
$c$ sends the upgrade response (and data) to cache $c'$, and its own downgrade
response along with the \FwdAck{} to the directory. The directory receives the
downgrade response from $c$. Soon afterwards, the directory wants to downgrade
the state of $a$ in $c'$ to $I$ (let's say, because some other cache wants to
upgrade its state for $a$ to $M$).  The directory sends a downgrade to $I$
request to $c'$ for address $a$. This downgrade request to $c'$ from the parent
reaches $c'$ before the (forwarded) upgrade response from $c$, violating
Constraint \ref{orderreq}. However, as we saw in Section \ref{sec:nofifoorder},
Constraint \ref{orderreq} can be satisfied by tagging the downgrade request with
the state the directory was in for cache $c'$ and address $a$ right before the
request to $c'$ was sent -- FIFO ordering is not required from the network. $c'$
will not process the downgrade request till it has received the upgrade response
from $c$.

Since the protocol satisfies all the constraints \ref{} to \ref{}, it satisfies
all the invariants in Figure \ref{invariants}. Thus the directory is guaranteed
to have a conservative version of the states of each of the caches for each
address by Invariant \ref{conservative}. When the directory sends a downgrade
request for address $a$ to $c$ tagged with a \FwdReq{c'}{x} (when $dir[c][a]$ is
$M$), the directory assumes that both $c$ and $c'$ are in state $M$. However, it
is guaranteed that $c'$ can not upgrade to $x$ unless $c$ has sent an upgrade
response to $c'$, which guarantees that $c$ would have downgraded to
$toCompatible(x)$. This ensures both the Single-writer Invariant
\ref{singleWriter} as well as Read-from-last-writer Invariant \ref{readLast}.

Note that in our protocol, \FwdAck is sent from the forwarding cache $c$ rather
than the final cache $c'$ which is the case in commonly used protocols
\cite{hammer, gems}. This unblocks the directory to serve further requests for
the same address as soon as \FwdAck is received from $c$, instead of waiting for
an acknowledgement from $c'$.

Thus we have mapped the complex protocol with intervention by a cache in $M$
state to a simpler MSI protocol and proved its correctness using our framework.

\subsubsection{MOSI protocol}

### 2012-11-14 paper/isca40format/Variations.tex
stuff

@@
 $a$ sent from $c$ to $c'$ has reached $c'$, $c'$ could have received a downgrade
-request sent from the parent. This happens as follows (Figure
+request sent from the directory. This happens as follows (Figure
 \ref{msi-fifo-needed}, with the numbers indicating the order in which the
@@
 upgrade its state for $a$ to $M$).  The directory sends a downgrade to $I$
-request to $c'$ for address $a$. This downgrade request to $c'$ from the parent
+request to $c'$ for address $a$. This downgrade request to $c'$ from the directory
 reaches $c'$ before the (forwarded) upgrade response from $c$, violating

### 2012-11-14 paper/isca40format/Discussion.tex
will be removing evict req

\section{Discussion}
\label{sec:discussion}

\subsection{Efficient bit encoding in directories}
The directory of a node store its version of the state for each of the node's
children. The directory encoding can be optimized using the Single-writer
Invariant \ref{singlewriter}.

For the MSI protocol, if a child is in $M$ state, every other child will be in
$I$ state, so just storing the index of the child in $M$ state is sufficient.
Otherwise, just a bit-vector, one bit for each child is sufficient to denote
which children are in $S$ state. Overall, for each line present in the node,
if a node has $n$ children, the directory requires $max(\lceil log_2(n)\rceil, n)
+ 1$ bits (the extra bit denotes if the value stored is the index, \ie a child
is in $M$ state or the bit-vector, \ie no child is in $M$ state). An additional
$2$ bits are needed to store the coherence state of the node with respect to its
parents. Even if MESI optimization is done, just $2$ bits for the coherence
state is enough, overall, to decide when the data of an address is dirty (just
like in normal MESI protocols). So MSI requires $max(\lceil log_2(n)\rceil, n)
+ 3$ bits per line. This is the same for the $M$-only cache intervention
protocol too.

In a MOSI protocol, a node's child can be in $O$ while multiple other caches are
still in $S$ state. Thus, for a node with $n$ children, the directory requires
$\lceil log_2(n)\rceil + n + 2$ bits per line. The additional two bits are
required to distinguish between 3 cases: a) if any cache is in $M$ state, in
which case only the index of the cache is relevant, or b) if any cache is in $O$
state, in which case both the index of the cache in $O$ state as well as the
bit-vector giving the list of caches in $S$ state is relevant, or c) every cache
is in a state $\le S$, in which case only the bit-vector is relevant. The
coherence state again requires $2$ bits as before, so overall, $\lceil log_2(n)
\rceil + 4$ bits are required per line.

\subsection{Can this framework be extended to sparse-map directories? To
non-inclusive or exclusive cache hierarchies?}
Our framework relies on full-map directories. If the directory sends a downgrade
request to a cache, the cache drops the request if the cache has already
downgraded its state Constraint \ref{drop-req}. This constraint has to be
changed to allow caches to respond notifying the directory that it has already
downgraded. We believe it is a straight-forward extension, but we haven't
formally proven if this works.

When caches are not inclusive, though the memory system is hierarchical with
respect to data tranfer, it not hierarchical with respect to coherence (as
every cache in every level has to be probed for downgrading). We believe that
such a system can be modeled within our framework as a two-level system, with
slightly different rules for compatibility between caches. For example, for
non-inclusive caches, two caches can have the same address in the $M$ state if
one cache is another's ancestor in terms of the data transfer hierarchy.

%\subsection{Are all the constraints \ref{} to \ref{} necessary to implement
%correct protocols?}
%As we discussed above, our framework requires full-map directory. Within the
%design space full-map directory protocols, we believe that all of these
%constraints are necessary to have a correct system.  To illustrate this, we will
%use the example of not imposing Constraint \ref{down-resp}, \ie a cache
%downgrades its state without sending a response to the directory.
%
%Consider the case of the simple MSI protocol. Let us assume cache $c$ downgraded
%its state from $M$ to $I$ without informing the directory. The directory might
%later send a downgrade to $I$ request to $c$, which will be dropped by $c$. Now
%the system would deadlock because the directory would not receive a response
%from $c$ ever.
%
%Consider the $M$-only cache-intervention based protocol. Here, the directory
%already sends an upgrade response (in the form of a forwarding message)

### 2012-11-14 paper/isca40format/Fifo.tex
will be removing evict req

@@
 
-We already gave the ordering requirements between messages (Figure
-\ref{ordering}), and which messages should not be blocked by other
-messages (Figure \ref{blocking}). These are the minimum constraints to
-ensure correctness and deadlock-freedom of the hierarchical coherence
-protocols. Any practical implementation will restrict the ordering between
-messages further, but they must ensure that the messages that should not be
-blocked according to Figure \ref{blocking} are not blocked. In this
-section, we will discuss several such implementations. Note that the
-constraints in Figures \ref{ordering} are on messages for the same
-address; messages for different addresses can be processed in any order. This
-permits implementation of non-blocking caches using MSHRs.
+%We already gave the ordering requirements between messages (Figure
+%\ref{ordering}), and which messages should not be blocked by other
+%messages (Figure \ref{blocking}). These are the minimum constraints to
+%ensure correctness and deadlock-freedom of the hierarchical coherence
+%protocols. Any practical implementation will restrict the ordering between
+%messages further, but they must ensure that the messages that should not be
+%blocked according to Figure \ref{blocking} are not blocked. In this
+%section, we will discuss several such implementations. Note that the
+%constraints in Figures \ref{ordering} are on messages for the same
+%address; messages for different addresses can be processed in any order. This
+%permits implementation of non-blocking caches using MSHRs.
+
+In this section, we show how the constraints of figures \ref{blocking} and
+\ref{ordering} can be implemented realistically.
+%We already gave the ordering requirements between messages (Figure
+%\ref{ordering}), and which messages should not be blocked by other
+%messages (Figure \ref{blocking}). These are the minimum constraints to
+%ensure correctness and deadlock-freedom of the hierarchical coherence
+%protocols. Any practical implementation will restrict the ordering between
+%messages further, but they must ensure that the messages that should not be
+%blocked according to Figure \ref{blocking} are not blocked. In this
+%section, we will discuss several such implementations. Note that the
+%constraints in Figures \ref{ordering} are on messages for the same
+%address; messages for different addresses can be processed in any order. This
+%permits implementation of non-blocking caches using MSHRs.
 
@@
 
-Request messages now have the format \Req{src}{dst}{addr}{fromState}{toState}.
 A downgrade request for an address from a parent to a child contains the
@@
 for the address just before the request was sent, in addition to the value of
-the state to upgrade to.
+the state to upgrade to. Request messages have a $fromState$ tagged to it, and
+thus have a new format \Req{src}{dst}{addr}{fromState}{toState}.
 
@@
 point-to-point FIFO ordering in the network using a similar approach. Response
-messages from the child to its parent will now have the format
-\Req{src}{P}{addr}{fromState}{toState}. A response \Resp{c}{P}{a}{x}{y} is not
-handled by the parent if $dir[c][a] > x$ until it receives another response
-\Resp{c}{P}{a}{w}{z} where $z \le x$.
+messages from the child to its parent will also have a $fromState$ field with
+the new format \Req{src}{P}{addr}{fromState}{toState}. A response
+\Resp{c}{P}{a}{x}{y} is not handled by the parent if $dir[c][a] > x$ until it
+receives another response \Resp{c}{P}{a}{w}{z} where $z \le x$.
 
@@
 messages as well as unblocking a cache line (for handling further requests)
-without waiting for acks.
+ithout waiting for acks.

### 2012-11-15 paper/isca40format/Msi.tex
stuff

@@
-\section{Informal description of the proof}
-\label{sec:informal-proof}
+\section{A two-level message-passing directory-based MSI protocol}
+\label{sec:msi}
+
+\newcommand{\fourAngle}[6]{\text{$#1\langle#2\rightarrow#3,#4,#5\rightarrow#6\rangle$}}
+\newcommand{\threeAngle}[5]{\text{$#1\langle#2\rightarrow#3,#4,#5\rangle$}}
+
+%\newcommand{\Req}[5]{\fourAngle{Req}{#1}{#2}{#3}{#4}{#5}}
+\newcommand{\Req}[4]{\threeAngle{Req}{#1}{#2}{#3}{#4}}
+\newcommand{\Resp}[4]{\threeAngle{Resp}{#1}{#2}{#3}{#4}}
 
 \newcommand{\msg}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
-\newcommand{\Req}[4]{\msg{Req}{#1}{#2}{#3}{#4}}
-\newcommand{\Resp}[4]{\msg{Resp}{#1}{#2}{#3}{#4}}
+%\newcommand{\Req}[5]{\msg{Req}{#1}{#2}{#3}{#4}{#5}}
+%\newcommand{\Resp}[5]{\msg{Resp}{#1}{#2}{#3}{#4}{#5}}
 \newcommand{\Data}[4]{\msg{Data}{#1}{#2}{#3}{#4}}
 \newcommand{\Mesg}[4]{\msg{Mesg}{#1}{#2}{#3}{#4}}
-\newcommand{\FirstInv}{\emph{FirstInv}}
+\newcommand{\DirInv}{\emph{DirInv}}
 %\input{Atomic}
 
-Let us walk through our design methodology for our message-passing hierarchical
-directory based protocols.
-
-\subsection{Message-passing directory-based MSI protocol for 2-levels, no
-replacement evictions}
-For the sake of explanation, let us start with an MSI protocol, with just two
-levels of the memory hierarchy -- the L1 caches and the main memory. We will
-eventually build realistic protocols with various optimizations.
-
-Let us assume that the caches have enough capacity so that no line gets evicted
-for the sake of replacement. Lines may however get evicted to enforce the
-single-writer property for cache coherence, \ie when one L1 cache has
-permission to modify the data for an address, no other cache has read or write
-permission for that address.
-
-An L1 cache stores the information given in Figure \ref{table:lineinfo}. The
-coherence state for an address denotes the permission that a cache has for the
-address: $M = \{R, W\}, S = \{R\}, I = \{\}$. State $I$ is akin to the cache not
-having the data corresponding to $a$, since it can neither read or write that
-cache line.
+Let us walk through our framework using the example of an MSI protocol.  We will
+start with a simple system for the sake of explanation.  The system has two
+levels of memory hierarchy -- private L1 caches for each core and a shared
+memory. The L1 caches do not have direct access to the memory, and the memory
+does not have a direct access to the actual states in the  L1 caches, or L1's
+data. Instead, the main memory has direct access to a directory associated with
+it which keeps a version of the L1 cache states for each address. The L1 caches
+do not have direct access to this directory. The directory's version of the L1
+cache states may differ from the actual L1 cache states. We use the terms memory
+and directory interchangeably.
 
@@
 \hline
-$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for
-address $a$\\
+$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for address $a$\\
 $data[c][a]$ & data in cache $c$ for address $a$\\
-\hline
-\end{tabularx}
-\caption{Values stored by a cache for an address}
+\hline \end{tabularx}
+\caption{Values stored by an L1 cache $c$ for an address $a$}
 \label{table:lineinfo}
@@
 
-In order to serve a request from the processor, if a cache has to \emph{upgrade}
-its state for address $a$, \ie change from $x$ to $y$ where $x > y$
-(Figure \ref{table:fns}), it can do so only after ensuring that the new state is
-\emph{compatible} with the states of other caches for that address, \ie other
-caches are in a state less than or equal to $toCompatible(y)$
-(Figure \ref{table:fns}). This is needed to ensure the single-writer property of
-the MSI protocol -- if a cache's state is $M$ for an address $a$, every other cache
-must be in $I$ for address $a$. But, the caveat is that the L1 cache can access
-only its own state and data, as opposed to that of other caches.
+\begin{wrapfigure}{r}{.4\linewidth}
+\centering
+\begin{tabular}{|c|ccc|}
+\hline
+& $M$ & $S$ & $I$\\
+\hline
+$M$ & $=$ & $>$ & $>$\\
+$S$ & $<$ & $=$ & $>$\\
+$I$ & $<$ & $<$ & $=$\\
+\hline
+\end{tabular}
+\caption{$<$ and other relations for MSI states}
+\label{msi<}
+\end{wrapfigure}
 
-\begin{figure}
-\begin{tabularx}{\linewidth}{|cX|}
-\hline
-$toCompatible(s)$ & If the state of one cache is $s$ for address $a$, the
-highest state of another cache for the same $a$
-$\newline
-toCompatible(M) = I,\newline
-toCompatible(S) = S,\newline
-toCompatible(I) = M
-$\\
-\hline
-$x > y$ & Denotes if state $x$ has more permissions than state $y$.
-$M > S, M > I, S > I$. Relations $\le, >, \ge$ are also defined appropriately\\
-\hline
-\end{tabularx}
-\caption{Abstraction functions}
-\label{table:fns}
-\end{figure}
-
-The memory controller has access to the data from the memory but not to the
-state and data of the L1 caches. Instead, it only has access to a directory
-which has a version of the coherence states of the caches for each address. The
-version of the state in the directory may not be up to date, \ie may not be the
-same as the one in the caches, all the time. Figure \ref{table:dirinfo} shows
-the information stored in the memory and directory. We will use the word memory
-and directory interchangeably for the 2-level protocol. The caches can not
-communicate with each other but can communicate with the directory via messages.
+Each L1 cache stores the information given in Figure \ref{table:lineinfo}. The
+coherence state for an address denotes the permission that a cache has for the
+address: $M = \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$ relation for
+coherence state values based on the permissions the states represent. This is
+shown in Figure \ref{msi<}. State $I$ is akin to the cache not having the data
+corresponding to $a$, since it can neither read or write the $a$'s data. Figure
+\ref{table:dirinfo} shows the information stored in the memory and directory.
 
@@
 
-Let's us assume that we have a system which guarantees the invariants given
-in Figure \ref{FirstInv}.
+The L1 caches and the directory communicate with each using via messages. Each
+L1 cache communicates with the memory directly via 4 buffers -- two incoming
+buffers from the memory, one each for requests and responses from the memory,
+and two outgoing buffers into the memory, one each for requests and responses to
+the memory. All these buffers converge into or diverge from 4 buffers of the
+memory -- two incoming buffers, one each for requests and responses from the
+children and two outgoing buffers, one each for requests and responses to the
+children (Figure \ref{fig:setup}). All these buffers have the FIFO property. Not
+only that, we will also assume that requests sent from a source will never reach
+its destination till an earlier response sent by the same source has been
+handled (and hence removed) by the destination. We will show in Section
+\ref{sec:fifo} about how these network requirements can be realistically
+implemented.
+
+\subsection{No replacement evictions}
+
+We will first design an MSI protocol in which the L1 caches have enough
+capacity (or rather associativity) so that no line gets evicted due to
+replacement. Lines may however get evicted to enforce the following invariant
+which is needed for maintaining cache coherence.
+
+\begin{inv}
+Single-writer: When a cache has permission to modify data for an address ($M$
+state), no other cache has read or write permission for that address. In other
+words, the other caches must be in $I$ state.
+\label{singleWriter}
+\end{inv}
+
+\begin{wrapfigure}{r}{.46\linewidth}
+\begin{subfigure}{.27\linewidth}
+\centering
+\begin{tabular}{|ccc|}
+\hline
+$M$ & $\rightarrow$ & $I$\\
+$S$ & $\rightarrow$ & $S$\\
+$I$ & $\rightarrow$ & $I$\\
+\hline
+\end{tabular}
+\caption*{$toCompatible$}
+\label{toCompat}
+\end{subfigure}
+~~~~~~~~~~~~~~
+\begin{subfigure}{.21\linewidth}
+\centering
+\begin{tabular}{|c|c|}
+\hline
+$M$ & $\checkmark$\\
+$S$ & $\times$\\
+$I$ & $\times$\\
+\hline
+\end{tabular}
+\caption*{$isModified$}
+\label{isModified}
+\end{subfigure}
+\caption{$toCompatible$ and $isModified$ mappings}
+\label{fig:msimap}
+\end{wrapfigure}
+
+Because of the Single-writer invariant, it makes sense to talk about
+compatibility of the states of two caches for an address. Two cache states $x$
+and $Y$ are \emph{compatible} if two caches having the same address in states
+$x$ and $y$ do not violate the Single-writer invariant. We define a mapping
+$toCompatible(x)$ which gives the highest state which is compatible with a given
+state $x$ (Figure \ref{fig:msimap}).
+
+In addition to Single-writer Invariant, maintaining cache coherence also
+requires the following invariant:
+
+\begin{inv}
+Read-from-last-writer: When a cache reads data for an address, it reads the last
+modified version of the data for the address. If the data for an address has not
+been modified, the cache reads the data from memory.
+\end{inv}
+
+Function $isModified(x)$ determines if data for an address state $x$ can
+potentially be modified or not (Figure \ref{fig:msimap}).
+
+If a cache $c$ receives a request for an address $a$ from the processor, it can
+service the request straightaway if it has enough permissions for address $a$.
+Otherwise, it has to get to a state for $a$, say $y$, which has enough
+permissions. In other words, it has to \emph{upgrade} its state to $y$. In order
+to maintain the Single-writer Invariant \ref{singleWriter}, cache $c$ has to
+ensure that the states of other caches are \emph{compatible} with $y$. But the
+caveat is that $c$ can access only its own state and data, as opposed to that of
+other caches (or the directory).
+
+Cache $c$ has to instead communicate with the directory to ensure that the
+states of other caches are compatible. $c$ must send a message
+\Req{c}{Dir}{a}{y} requesting the directory to tell if it can upgrade its state
+for address $a$ to $y$. We call this an \emph{upgrade request}. Let us now take
+for granted that such a message can be sent (and see how this in ensured later).
+
+\begin{inv}
+A cache will eventually be able to send an upgrade request.
+\label{csendreq}
+\end{inv}
+
+We will also assume that that a cache can upgrade its state to $y$ only after it
+receives a message from the directory \Resp{Dir}{c}{a}{y} permitting it to
+upgrade its state for address $a$ to $y$. We call such a message as an
+\emph{upgrade response}.
+
+\begin{inv}
+Cache $c$ can upgrade $state[c][a]$ from $x$ to $y$ where $y > x$ only after it
+receives an upgrade response \Resp{Dir}{c}{a}{y} from the directory.
+\label{cwaitresp}
+\end{inv}
+
+We introduce two other kinds of messages: \emph{downgrade request}
+\Req{Dir}{c'}{a}{y} sent from the directory to a cache $c'$ asking $c'$ to
+downgrade its state for address $a$ to $y$, and \emph{downgrade response}
+\Resp{c'}{Dir}{a}{y} sent from $c'$ to the directory informing the directory
+that $c'$ has changed its state $state[c'][a]$ from $x$ to $y$, where $x > y$.
+
+How can the behavior of the directory be designed (in terms of sending and
+receiving messages) to ensure that cache $c$ will get an upgrade response
+eventually. Let's say the invariants given in Figure \ref{DirInv} are
+somehow guaranteed. With these invariants, let us design the behavior of the
+directory.
+
 \floatstyle{boxed}
 \restylefloat{figure}
-
 \begin{figure}
-\begin{enumerate}
-\item The cache can upgrade $state[c][a]$ from $x$ to $y$ where $y > x$, \ie
-\emph{downgrade} its directory state only after it receives the appropriate
-authorization from the directory for address $a$. \label{cwaitresp}
-\item If the cache wants to upgrade $state[c][a]$ to $x$, there is a
-message \Req{c}{Dir}{a}{x} which cache $c$ can send to the directory requesting
-the upgrade authorization, and the cache will eventually be able to send
-the request. \label{csendreq}
-\item If the directory receives an upgrade request \Req{c}{Dir}{a}{x} from cache
-$c$ for address $a$, then just before receiving the request, $dir[c][a] < x$.
+\begin{inv}
+Just before receiving an upgrade request \Req{c}{Dir}{a}{y} from $c$ $dir[c][a]
+< y$.
 \label{drecvreq}
-\item The directory can change $dir[c][a]$ from $x$ to $y$ where $y < x$, \ie
-\emph{downgrade} its directory state only after it receives the appropriate
-authorization from cache $c$ for address $a$. \label{dwaitresp}
-\item If the directory wants to downgrade $dir[c][a]$ to $x$, there is a
-message \Req{Dir}{c}{a}{x} which the directory can send to cache $c$ requesting
-the downgrade authorization, and the directory will eventually be able to send
-the request. \label{dsendreq}
-\item If the directory has sent a downgrade request for an address $a$ to cache
-$c$, it will eventually get the downgrade authorization response for address $a$
-from cache $c$ to downgrade to $x$, where $x$ is less than or equal to the state
-to which the directory sent a request to downgrade to. \label{drecvresp}
-\item If $dir[c][a] = M$ just before receiving a downgrade authorization from
-cache $c$ for address $a$, then the cache would have sent the data along with
-the authorization. \label{drecvdata}
-\item If the directory wants to send an authorization message to upgrade the
-state of address $a$ in cache $c$ to $x$ (\Resp{Dir}{c}{a}{x}), it will
-eventually be able to send it. \label{dsendresp}
-\item The directory can send the data from the memory along with the upgrade
-authorization message if it wants to. \label{dsenddata}
-\item The directory's version of the cache's state is always conservative, \ie
-$\forall c, a,\; dir[c][a] \ge state[c][a]$. \label{conservative}
-\end{enumerate}
-\caption{Initial set of invariants \FirstInv}
-\label{FirstInv}
+\end{inv}
+\begin{inv}
+The directory will eventually be able to send any downgrade request to any cache
+$c$.
+\label{dsendreq}
+\end{inv}
+\begin{inv}
+The directory can change $dir[c][a]$ from $x$ to $y$ where $y < x$, \ie
+downgrade its directory state only after receiving a downgrade response
+\Resp{c}{Dir}{a}{y} from $c$.
+\label{dwaitresp}
+\end{inv}
+\begin{inv}
+If the directory has sent a downgrade request \Req{Dir}{c}{a}{y} to cache $c$,
+it will eventually get a downgrade response \Resp{Dir}{c}{a}{z} from cache $c$
+such that $z \le y$.
+\label{drecvresp}
+\end{inv}
+\begin{inv}
+Just before receiving a downgrade response from $c$, if $isModified(dir[c][a])$
+is true, then $c$ would have sent the data along with the downgrade response.
+\label{drecvdata}
+\end{inv}
+\begin{inv}
+The directory will eventually be able to send any upgrade response to any cache
+$c$. It can also send the data to cache $c$ along with the upgrade response.
+\label{dsendresp}
+\end{inv}
+\begin{inv}
+Conservative directory:
+\begin{spacing}{.1}
+\begin{equation*}
+\forall c, a,\; dir[c][a] \ge state[c][a]
+\end{equation*}
+\end{spacing}
+\label{conservative}
+\end{inv}
+\caption{Directory invariants}
+\label{DirInv}
 \end{figure}
-
 \floatstyle{plain}
@@
 
-We will walk through how the MSI coherence protocol (with no replacement
-evictions) can be implemented using the above invariants and argue why it is a
-correct implementation.
+%The directory ensures that the states of any two caches for an address are
+%compatible by the following behavior.
+%When the directory receives an upgrade request
+%\Req{c}{Dir}{a}{x}, Invariant \ref{drecvreq} guarantees that $dir[c][a] < x$ and
+%because of Invariant \ref{conservative}, $cache[c][a] le dir[c][a]$, resulting
+%in $cache[c][a] < x$. Unless the directory sends an upgrade response to $c$
+%permitting it to upgrade to $x$, $c$ will remain in state $cache[c][a]$ forever
+On receiving an upgrade request \Req{c}{Dir}{a}{x}, the directory sends a
+downgrade request \Req{Dir}{c'}{a}{toCompatible(x)} to every cache $c'$
+for which $dir[c'][a] > toCompatible(x)$. Such a request can be sent eventually
+because of Invariant \ref{dsendreq}. The incoming upgrade request is not
+removed and no further upgrades requests are processed till an appropriate
+upgrade response is sent back from the directory to $c$.
 
-The directory's role is to ensure the single-writer property and that the last
-updated data is being read. In order to implement the cache coherence protocols
-using the invariants of Figure \ref{FirstInv}, the directory's ensures that at
-all times, if for some $c, dir[c][a] = M$, the for every other $c', dir[c'][a] =
-I$.
+Because of Invariant \ref{drecvresp}, the directory will eventually receive
+downgrade responses \Resp{c'}{Dir}{a}{z} from each of the requested caches $c'$
+such that $z \le toCompatible(x)$ (since the directory made a downgrade request
+to go to $toCompatible(x)$). Just before receiving the downgrade response, if
+$isModified(dir[c'][a])$ is true (\ie $dir[c'][a] = M$), then by Invariant
+\ref{drecvdata}, the directory would also receive data from $c'$ and the memory
+for address $a$ is updated with the received data.
 
-If a cache gets a request from the core which can not be handled because of lack
-of permissions, it (eventually) sends an upgrade request to the directory.
-\FirstInv.\ref{csendreq} ensures that such a request can eventually be sent.
+Once all the downgrade responses have been received, since no upgrade responses
+have been sent to any cache meanwhile, we have $\forall c' \neq c, dir[c'][a] \le
+toCompatible(x)$. Because of Invariant \ref{conservative}, it is guaranteed that
+$\forall c' \neq, cache[c'][a] \le toCompatible(x)$. The directory now changes
+$dir[c][a]$ to $x$ and sends an upgrade response \Resp{Dir}{c}{a}{x} to $c$. If
+$dir[c][a] = I$ right before the upgrade response is sent, then it implies that
+$cache[c][a] = I$. So the data from the memory for address $a$ is also sent
+along with the upgrade response. Note that such an upgrade response, along with
+the data, can always be sent eventually by the directory because of Invariant
+\ref{dsendresp}. Finally, the upgrade response from $c$ is removed by the
+directory.
 
-If the directory receives a request from cache $c$ to upgrade the state of
-address $a$ to $x$, because of \FirstInv.\ref{drecvreq}, $dir[c][a] < x$ and
-since $cache[c][a] \le dir[c][a]$ because of \FirstInv.\ref{conservative},
-$cache[c][a] < x$ and hence the directory must send an upgrade authorization
-response. The directory sends a request \Req{Dir}{c'}{a}{toCompatible(x)} to
-every cache $c'$ for which $dir[c'][a] > toCompatible(x)$. Such a request can be
-sent eventually because of \FirstInv.\ref{dsendreq}.
+The above behavior for the directory ensures that the states of any two caches
+for every address are compatible, as can seen by induction.  This property is
+true initially because every cache is in state $I$ for every address. If this
+property is true before a request is started received, it will be true by the
+time the request is removed and an upgrade response is sent. This is because an
+upgrade response is sent only when every cache other than the one requesting is
+at a state $\le toCompatible(x)$, where $x$ is the state the requesting cache
+wants to go to, and $toCompatible(x)$ is the highest state that a cache can be
+in to be compatible with another cache in state $x$ for an address.
 
-Because of \FirstInv.\ref{drecvresp}, the directory will eventually receive
-responses from each of the requested caches $c'$
-(\Resp{c'}{Dir}{a}{toCompatible(x)}) authorizing the downgrade of $dir[c'][a]$
-to $toCompatible(x)$. If $dir[c'][a] = M$ just before receiving the
-authorization, then by \FirstInv.\ref{drecvdata}, the directory would also
-receive data from $c'$. Because the directory maintains the single-writer
-invariant, every other cache $c''$ would have $dir[c''][a] = I$ just before
-receiving the authorization. This implies for every other cache $c''$,
-$state[c''][a] = I$ by the \FirstInv.\ref{conservative} invariant, and hence
-$c'$ has the most up-to-date copy for address $a$. Thus the memory is updated
-with the received up-to-date copy for address $a$.
+Whenever the directory receives a downgrade response from a cache $c'$, just
+before receiving the response if $isModified(dir[c'][a])$ is true, \ie
+$dir[c'][a] = M$, $c'$ must have been the only cache with permission to modify
+address $a$. This is because the directory ensures that the states of any two
+caches for address $a$ is compatible at all times (as shown above). Thus, to
+ensure the Read-from-last-writer Invariant, the potentially modified data from
+$c'$ is written back from into the memory on receiving the downgrade response
+from $c'$ as seen in our design.
 
-Once all the authorizations have been received, for every cache $c'$,
-$dir[c'][a] \le toCompatible(x)$ (other than the one that made the request $c$).
-Since the \FirstInv.\ref{conservative} invariant ensures that $cache[c'][a] \le
-dir[c'][a]$, we have $cache[c'][a] \le toCompatible(x)$, ensuring the
-single-writer property (because $toCompatible(M) = I$). The directory can now
-change $dir[c][a]$ to $x$ and send \Resp{Dir}{c}{a}{x} authorizing $c$ to
-upgrade state of address $a$ to $x$ (\FirstInv.\ref{dsendresp} ensures that such
-a message can be sent). Just before sending the authorization, if $dir[c][a] =
-I$, it implies that $state[c][a] = I$ because of the
-\FirstInv.\ref{conservative} invariant.  Thus the cache doesn't have data
-corresponding to address $a$ and the directory has to supply the most up to date
-version of the address (from memory, since the memory is up to date), which it
-can because of \FirstInv.\ref{dsenddata}.
-
-The \FirstInv\ invariants essentially makes it straight forward to design the
-behavior of the directory . For the sake of argument, let
-\FirstInv.\ref{drecvreq} not be guaranteed. Consider a scenario the cache $c$
-which has sent a request \Req{c}{Dir}{a}{M} earlier but hasn't received the
-upgrade authorization response yet. Meanwhile, $c$ sends another upgrade request
-\Req{c}{Dir}{a}{S}. Let us assume that the directory processed the requests in
-the order in which they were sent. Once the directory has finished processing
-the first request, it changes its directory state $dir[c][a]$ to $M$. When it
-receives the second request, it is not easy to decide if the directory has to
-drop the request or to send another response authorizing the cache to change to
-$S$.
+The invariants \ref{drecvreq} to \ref{conservative} in Figure \ref{DirInv}
+essentially makes it straight forward to design the behavior of the directory.
+For the sake of argument, let us assume Invariant \ref{drecvreq} not be
+guaranteed. Consider a scenario in which the cache $c$ which has sent a request
+\Req{c}{Dir}{a}{M}. Before receiving an upgrade response for this request, $c$
+sends another upgrade request \Req{c}{Dir}{a}{S}. Let us assume that the
+directory processed the requests in the order in which they were sent. Once the
+directory has finished processing the first request, it changes its directory
+state $dir[c][a]$ to $M$. When it receives the second request, it is not easy to
+decide if the directory has to drop the request or to send another response
+authorizing the cache to change to $S$. But if Invariant \ref{drecvreq} is
+guaranteed, it's guaranteed that such a scenario will never arise.
 
@@
 difficult to guarantee the above invariants, even for a 2-level memory hierarchy
-discussed above as shown in the examples below.
+discussed above as shown in the following scenarios.
 
-\subsubsection*{Example 1: Inability to send messages because of network
-congestion:}
-For instance, the ability of a cache or the directory to send messages
-(\FirstInv.\ref{dsendreq}, \FirstInv.\ref{dsendresp}, \FirstInv.\ref{csendreq},
-\etc) depends upon the availability of free buffers in the network; if the
-network remains congested all the time, then it would not be possible to send the
-messages, and the system would deadlock.
+%\begin{scen}
+%\emph{Inability to send messages because of network congestion:}
+%The ability of a cache or the directory to send messages (Invariants
+%\ref{dsendreq}, \ref{dsendresp} and \ref{csendreq}) depends upon the
+%availability of free buffers in the network; if the network remains congested
+%all the time, then it would not be possible to send the messages, and the system
+%would deadlock.
+%\end{scen}
+%
+%\begin{scen}
+%\emph{Multiple upgrade requests from the cache for the same address:}
+%If a cache is allowed to make multiple requests to upgrade the state of an
+%address (say to $M$ and then to $S$) before receiving the upgrade responses,
+%then it might violate Invariant \ref{drecvreq} if the requests were processed by
+%the directory in the order they were received. This makes the decision on
+%whether the directory drops the second request or not hard to reason about, as
+%shown above.
+%\end{scen}
 
-\subsubsection*{Example 2: Multiple upgrade requests from the cache for the same
-address:}
-If a cache is allowed to make multiple requests to upgrade the state of an
-address (say to $M$ and then to $S$) before receiving the upgrade
-authorizations, then it might violate \FirstInv.\ref{drecvreq} if the requests
-were processed by the directory in the order they were received.
+\begin{scen}
+\emph{On sending an upgrade request, the cache stops processing downgrade
+requests it gets an upgrade response:}
+Consider a scenario with two caches $c$ and $c'$, with both the caches in state
+$S$ for address $a$. Both the caches get a store request from the processor and
+send \Req{c}{Dir}{a}{M} and \Req{c'}{Dir}{a}{M}, respectively, to the directory.
+Let's say \Req{c'}{Dir}{a}{M} arrives at the directory first. The directory
+sends a downgrade request to cache $c$, but since $c$ does not process any
+downgrade requests till it gets its upgrade response, it will never send back a
+downgrade response to the directory, violating Invariant \ref{drecvresp},
+resulting in a deadlock. Note that this deadlock would occur even if a cache is
+waiting for the upgrade response, and stops processing downgrade requests for
+any address, even if the address corresponding to the downgrade request is not
+the same as the one corresponding to the upgrade response.
+\end{scen}
 
-\subsubsection*{Example 3: On sending an upgrade request, the cache waits till
-it gets an authorization:}
-If the behavior of the caches were such that if a cache $c$ has sent an
-upgrade request \Req{c}{Dir}{a}{x} to the directory, it keeps waiting for the
-upgrade authorization response \Resp{Dir}{c}{a}{x} before processing any
-requests from the directory for address $a$, then the cache coupled with our
-simple directory implementation would deadlock. Consider a scenario with two
-caches $X$ and $Y$, with both the caches in state $S$ for address $a$. Both the
-caches got a store request from the processor and want to upgrade their states
-to $M$, and hence send \Req{X}{Dir}{a}{M} and \Req{Y}{Dir}{a}{M}, respectively,
-to the directory. Let us say the directory receives \Req{Y}{Dir}{a}{M} first,
-and in order to process it, according to our implementation of the directory,
-sends a downgrade request \Req{Dir}{X}{a}{I} to cache $X$. But the cache 
-does not process any request from the directory for address $a$, according to
-our assumption, and hence will never send the downgrade authorization
-\Resp{X}{Dir}{a}{I} to the directory, violating \FirstInv.\ref{drecvresp}, and
-hence causing a deadlock.
+\begin{scen}
+\emph{At the directory, downgrade responses from children are blocked by upgrade
+requests:}
+Consider a scenario where two caches $c$ and $c'$ are both in state $S$ for
+address $a$. $c'$ gets a store request from the processor and sends an upgrade
+request \Req{c'}{Dir}{a}{M} to the directory. The directory receives that
+request, and sends a downgrade request \Req{c}{Dir}{a}{I} to $c$. $c$ sends a
+downgrade authorization \Resp{c}{Dir}{a}{I} to the directory on receiving the
+downgrade request. But the directory can not receive the authorization because
+it is blocked by other requests (including \Req{c'}{Dir}{a}{M}). This violates
+Invariant \ref{drecvresp} and hence creates a deadlock.
+\end{scen}
 
-\subsubsection*{Example 4: At the directory, authorization responses are blocked
-by upgrade requests:}
-Consider a scenario where two caches $X$ and $Y$ are both in state $S$ for address
-$a$. $Y$ got a store request from the processor and sends an upgrade request
-\Req{Y}{Dir}{a}{M} to the directory. The directory receives that request, and
-sends a downgrade request \Req{X}{Dir}{a}{I} to $X$. $X$ sends a downgrade
-authorization \Resp{X}{Dir}{a}{I} to the directory on receiving the downgrade
-request. But the directory can not receive the authorization because it is
-blocked by other requests (including \Req{Y}{Dir}{a}{M}). This violates
-\FirstInv.\ref{drecvresp} and hence creates a deadlock.
+These scenarios show that if the invariants of Figure \ref{DirInv} are violated,
+then it either breaks cache coherence (by creating deadlocks) or it makes
+it hard to reason about the behavior of the directory.
 
-These examples show that if the \FirstInv\ invariants are violated, then it
-either breaks cache coherence (by creating deadlocks) or it makes correctly
-designing the directory very difficult.
+Before we show how the invariants of Figure \ref{DirInv} can be guaranteed, let
+us consider a slightly more complicated system where caches can evict lines for
+replacement.
 
-Before we show how the \FirstInv\ invariants can be guaranteed (which requires
-the description of the behavior of the L1 caches), let us consider a
-slightly more complicated system where caches can evict lines for replacement.
+\subsection{Replacement evictions are allowed}
+If a cache receives a request from the core for an address $a$ and has no space
+to accommodate address $a$, it evicts another address $a'$ to make room for $a$.
 
-\subsection{Message-passing directory-based MSI protocol for 2-levels, with
-replacement evictions}
-We again have just two levels of memory hierarchy -- the L1 caches and the main
-memory. But the caches have finite capacity. If receives a request from the core
-for an address $a$ and has no space to accommodate address $a$, it evicts
-another address $a'$ to make room for $a$.
+In the previous scenario, the caches didn't have to evict for replacement; a
+cache had to send downgrade responses to the directory only on receiving a
+corresponding request. But in the new scenario, a cache can send a downgrade
+notification voluntarily to the directory on evicting an address. Such a
+notification is useful in keeping the directory's version of the cache's state
+up to date. We will restrict the family of directory-based protocols that that
+we discuss to those in which the caches send such notifications.
 
-In the previous scenario, where the caches didn't have to evict for replacement,
-a cache only had to send downgrade authorizations to the directory on receiving
-a corresponding request. But in the new scenario, a cache can send a downgrade
-notification to the directory on evicting an address. Such a notification is
-useful in keeping the directory's version of the cache's state up to date, so we
-will restrict the family of directory-based protocols that we discuss to those
-where the caches send such a notification, voluntarily.
+Since the directory can not distinguish between the downgrade notification sent
+voluntarily and a downgrade response sent because of a downgrade request from
+the directory, we will use the same message to denote both.
 
-The directory handles this notification in the exact manner by which it handled
-a downgrade authorization response from the cache, so the notification message
-is the same as a downgrade authorization message from cache $c$ for authorizing
-the downgrade of the $dir[c][a]$ to $I$ (\Resp{c}{Dir}{a}{I}).
+We will first discuss an example of the behavior of the L1 cache that would
+violate Invariant \ref{conservative}.
 
-We will now discuss a behavior of the L1 cache that would violate the
-invariant that the directory's version of the states of the caches for each
-address is conservative.
-
-\subsubsection*{Example 5: Cache sends voluntary notifications
-(\Resp{c}{Dir}{a}{I}) indiscriminately}
-Consider a system with two caches $X$ and $Y$. Let $state[X][a] = S, state[Y][a]
-= I$ and let $dir[X][a] = S, dir[Y][a] = I$. $X$ receives a store request from
-the core, and hence sends an upgrade request \Req{X}{Dir}{a}{M} to the
-directory. Say $X$ was a non-blocking cache, and hence receives another load
-request for address $a'$ from the core before receiving an upgrade authorization
-for address $a$. $X$ does not have space in the cache for $a'$, and it decides
-to evict address $a$, and sends \Resp{X}{Dir}{a}{I} to the directory, changing
-$state[c][a]$ to $I$.  Meanwhile, the directory receives the request
-\Req{X}{Dir}{a}{M} and since $dir[Y][a] \le I$, it responds to $X$ with an
-upgrade authorization \Resp{Dir}{X}{a}{M} (along with the data). Later, the
-directory receives the notification \Resp{X}{Dir}{a}{I} from $X$, and downgrades
-$dir[X][a]$ to $I$. Meanwhile, $X$ receives the upgrade authorization
-\Resp{Dir}{X}{a}{M} and changes $state[X][a]$ to $M$. Now,
-\FirstInv.\ref{conservative} is broken, since $state[X][a] = M$ and $dir[X][a] =
-I$. Let $Y$ get a load request from its core, and hence sends a
-\Req{Y}{Dir}{a}{S} to the directory. The directory on receiving the request,
-sends an upgrade authorization to $Y$ because $dir[X][a] = I$. This violates the
-single-writer property and hence coherence.
+\begin{scen}
+\emph{Cache sends voluntary downgrade responses indiscriminately:}
+Consider a system with only two caches $c$ and $c'$ (Figure \ref{fig:cross})
+such that $state[c][a] = dir[c][a] = S$, and $state[c'][a] = dir[c'][a] = I$.
+Let $c$ be a non-blocking cache, so it can process requests from the core even
+if there are pending misses. $c$ receives a store request for address $a$ from
+the core and sends \Req{c}{Dir}{a}{M} to the directory.  Before it gets back an
+upgrade response from the directory, it receives another load request for
+address $a'$ from the core. Suppose the line for $a'$ is not present in $c$ and
+$c$ has no space for $a'$, it decides to evict address $a$ to make room for
+$a'$. It voluntarily sends a downgrade response \Resp{c}{Dir}{a}{I} to the
+directory. Meanwhile, the directory receives the upgrade request
+\Req{c}{Dir}{a}{M}, and since the directory state of the only other cache is $I$
+for address $a$, the directory sends back an upgrade response. The directory
+then receives the downgrade response \Resp{c}{Dir}{a}{I} and changes $dir[c][a]$
+to $I$. $c$ finally receives the upgrade response \Resp{Dir}{c}{a}{M} and
+upgrades $state[c][a]$ to $M$. This breaks Invariant \ref{conservative}, since
+$state[c][a] = M$ and $dir[c][a] = I$. If the directory gets an upgrade to $M$
+request for address $a$ from $c'$, it will assume that $c$ is in state $I$ for
+address $a$ and send an upgrade to $M$ response to $c'$. This will violate the
+Single-writer invariant, thereby violating coherence.
+\end{scen}

### 2012-11-18 paper/isca40format/Introduction.tex
will be removing evict req

@@
 which contains information about its children caches. The coherence state
-changes are orchestrated by passing messages in the form of \emph{upgrade} or
-\emph{downgrade} requests and responses between the caches via the directory.
-Unlike bus-based protocols which lock the bus till all the required state
-transitions are done thereby ensuring atomicity of state transitions, in a
-message-passing protocol, requests and responses are split leading to transient
-states. This creates scenarios not present in bus-based protocols, which can
-lead to coherence violations or deadlocks, if the protocol is not designed or
-implemented carefully.
+changes are orchestrated by passing requests and responses between the caches
+via the directory.  Unlike bus-based protocols which lock the bus till all the
+required state transitions are done thereby ensuring atomicity of state
+transitions, in a message-passing protocol, requests and responses are split
+leading to transient states. This creates scenarios not present in bus-based
+protocols, which can lead to coherence violations or deadlocks, if the protocol
+is not designed or implemented carefully.
 
@@
 \begin{enumerate}
-
 \item Lots of concurrent events must be dealt with without a central authority
 to impose a sequence on events;
-
 \item The state on which transitions are based is physically distributed in
 space and cannot be accessed atomically;
-
 \item Physical network may block the transmission of a message because of lack
 of communication buffers or head-of-the-line blocking; 
-
-\item For each message type, determining which cache and directory states are
-possible (and therefore must be dealt) with is difficult. An incomplete set of
-transition rules may cause the protocol to deadlock.
-
+\item The cache or directory may reach a state which can not handle a
+particular message. An incomplete set of transition rules may cause the
+protocol to deadlock.
 \end{enumerate}
@@
 concrete and familiar scenarios from two-level systems (a private L1 cache for
-each processor and a shared L2 cache):
+each processor and a memory with a directory):
 
-\noindent \emph{Example 1:} A cache with a read permission for a line gets a
-store request from its processor and hence, requires a write permission for
-that line. The cache would send an upgrade request to the directory controller
-and may eventually get a response from the directory controller. The response
-would signify that the states in the other caches \emph{were} compatible with
-this cache having write permission for the line. That is, the other caches
-neither \emph{had} read nor write permissions for that line. But the messages
-take time to propagate and the response message is necessarily based on the old
-state bits (permissions) that the other caches had, not the permissions that
-the other caches have at the moment the original cache gets the response. If
-another cache gets a read or write permission for that line during the time the
-messages are in flight, coherence would be violated.
+\noindent \emph{Example 1:} A cache in state $S$ for a line gets a store request
+from the processor and sends a request for changing the line's state to $M$ to
+the directory controller. The controller may eventually send a response
+permitting the cache to go to $M$ state for that line. The response would
+signify that the states in the other caches for that line \emph{were} $I$. But
+the messages take time to propagate and the response is necessarily based on the
+old states of other caches -- the current state of another cache may not be $I$,
+violating the Single-writer-at-a-time invariant and hence breaking cache
+coherence.
 
-\noindent \emph{Example 2:} A cache, already having read permission for a line,
-gets a store request, and sends an upgrade request for write permission to the
-directory controller. Instead of getting a response, the cache may receive a
-request from the directory controller asking this cache to give up the read
-permission for the line \ie an invalidate request. This could happen if the
-directory controller received a request for write permission from another cache
-earlier. It is not easy to determine the correct behavior of the cache
-receiving the invalidate request -- whether it should ignore the invalidate
-request and wait till it receives a response from the directory controller for
-its request, or if it should invalidate its line. If it ignores the invalidate
-request, is there any guarantee that it would receive a response for its
-request for write permission, or would it deadlock?
+\noindent \emph{Example 2:} A cache in state $S$ for a line gets a store request
+from the processor and sends a request for changing the line's state to $M$ to
+the directory controller. Instead of getting a response permitting it to go to
+$M$ state, the cache may receive a request from the directory controller asking
+this cache to go to $I$ state (invalidate request)-- this could have happened if
+the directory is serving some other cache's request to go to $M$ state for the
+same line. It is not easy to determine the correct behavior of the former cache
+on receiving the invalidate request -- whether it should ignore the invalidate
+request and wait for a response to its request or whether it should invalidate
+the line. Would ignoring the invalidate request lead to a deadlock?
 
@@
 These examples give a glimpse of the complexity of designing and implementing a
-correct cache coherence protocol in a message-passing, directory-based setting.
-Even for a 2-level cache hierarchy, designing and implementing the protocol is
-quite difficult. This problem is greatly exacerbated in a multi-level cache
-hierarchy, which is becoming the norm because of the increasing core counts.
+correct cache coherence protocol in a directory-based non-bus setting.  Even for
+a 2-level cache hierarchy, designing and implementing the protocol is quite
+difficult. This problem is greatly exacerbated in a multi-level cache hierarchy,
+which is becoming the norm because of the increasing core counts.
 
 In this paper, we present a substantially easier method to design
-cache-coherence protocols for such systems. The method essentially breaks down
-the design problem in two parts: a part that is an abstract protocol shared by
-a number of concrete protocols, and another part that is specific to the
-protocol to be designed and implemented. The abstract protocol is parameterized
-by the number of cache levels and the number of caches in each level of the
-hierarchy as well as by a predicate which involves a local ``compatibility of
-state'' test. We illustrate our methodology by designing multiple
-hierarchical cache-coherence protocols, notably MSI, MESI, MOSI, MOESI, both in
-the presence as well as in the absence of point-to-point FIFO networks.
+directory-based cache-coherence protocols for such systems. The method breaks
+down the design problem in two parts: a part that is shared by a number of
+concrete protocols which we call \glob{}s, and another part that is specific to
+the protocol to be designed and implemented, called \policy{}. \glob{}s provide
+guarantees such as every request will eventually get a response, the
+directory's notion of the cache states is ``conservative'' \ie a cache can not
+have more permissions for a line than what the directory thinks it has, \etc;
+\glob{} enables \policy{} to focus on the high level properties such as the
+set of actions a cache or a directory should perform to maintain the
+Single-writer-at-a-time and the Read-from-last-writer invariants.
+
+The \glob{}s can in turn be guaranteed using a set of (\policy{} independent)
+\local{}s which dictate local properties of a node such as when a request can be
+handled, what state changes happen on sending a response, properties about the network,
+\etc. We have formally proved that \local{}s ensure \glob{}s using the
+Coq Theorem Prover \cite{}, however, a presentation of the formal proof is
+beyond the scope of this paper. The \glob{}s and the \local{}s form the
+cornerstone of our framework. The separation of \glob{}s and \local{}s is akin
+to having an abstraction layer which hides low level details.
+
+The \glob{}s of our framework can be used to design a wide range of cache
+coherence protocols, notably MSI, MESI, MOSI, MOESI, as well as their
+hierarchical version. We will show how these are designed as \policy{}s,
+focusing on high-level details.
 

### 2012-11-18 paper/isca40format/Msi.tex
will be removing evict req

--- 2012-11-14 paper/isca40format/Msi.tex
@@
-\section{Informal description of the proof}
-\label{sec:informal-proof}
+\section{MSI protocol}
+\label{sec:msi}
 
-\newcommand{\msg}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
-\newcommand{\Req}[4]{\msg{Req}{#1}{#2}{#3}{#4}}
-\newcommand{\Resp}[4]{\msg{Resp}{#1}{#2}{#3}{#4}}
-\newcommand{\Data}[4]{\msg{Data}{#1}{#2}{#3}{#4}}
-\newcommand{\Mesg}[4]{\msg{Mesg}{#1}{#2}{#3}{#4}}
-\newcommand{\FirstInv}{\emph{FirstInv}}
-%\input{Atomic}
+\newcommand{\printall}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
+\newcommand{\Req}[4]{\printall{Req}{#1}{#2}{#3}{#4}}
+\newcommand{\Resp}[4]{\printall{Resp}{#1}{#2}{#3}{#4}}
+\newcommand{\Data}[5]{\printall{Data}{#1}{#2}{#3}{#4}{#5}}
 
-Let us walk through our design methodology for our message-passing hierarchical
-directory based protocols.
+\newcommand{\lett}{\textbf{let}}
+\newcommand{\send}{\textbf{send}}
+\newcommand{\receive}{\textbf{receive}}
+\newcommand{\pop}{\textbf{pop}}
+\newcommand{\remove}{\textbf{remove}}
+\newcommand{\assert}{\textbf{assert}}
+\newcommand{\call}{\textbf{call}}
+\newcommand{\dReq}{\textsc{handleDownReq}}
+\newcommand{\uReq}{\textsc{handleUpReq}}
+\newcommand{\dReqL}{\textsc{sendDownReq}}
+\newcommand{\uReqL}{\textsc{sendUpReq}}
+\newcommand{\dResp}{\textsc{sendDownResp}}
+\newcommand{\uResp}{\textsc{sendUpResp}}
 
-\subsection{Message-passing directory-based MSI protocol for 2-levels, no
-replacement evictions}
-For the sake of explanation, let us start with an MSI protocol, with just two
-levels of the memory hierarchy -- the L1 caches and the main memory. We will
-eventually build realistic protocols with various optimizations.
-
-Let us assume that the caches have enough capacity so that no line gets evicted
-for the sake of replacement. Lines may however get evicted to enforce the
-single-writer property for cache coherence, \ie when one L1 cache has
-permission to modify the data for an address, no other cache has read or write
-permission for that address.
-
-An L1 cache stores the information given in Figure \ref{table:lineinfo}. The
-coherence state for an address denotes the permission that a cache has for the
-address: $M = \{R, W\}, S = \{R\}, I = \{\}$. State $I$ is akin to the cache not
-having the data corresponding to $a$, since it can neither read or write that
-cache line.
+Let us walk through our framework using the example of a hierarchical MSI
+protocol.
 
@@
 \hline
-$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for
-address $a$\\
-$data[c][a]$ & data in cache $c$ for address $a$\\
+$c.state[a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for address $a$\\
+$c.data[a]$ & data in cache $c$ for address $a$\\
+%$c.dir[n][a]$ & $c$'s notion of the coherence state of $n$ for address $a$\\
 \hline
 \end{tabularx}
-\caption{Values stored by a cache for an address}
+\caption{Notations}
 \label{table:lineinfo}
@@
 
-In order to serve a request from the processor, if a cache has to \emph{upgrade}
-its state for address $a$, \ie change from $x$ to $y$ where $x > y$
-(Figure \ref{table:fns}), it can do so only after ensuring that the new state is
-\emph{compatible} with the states of other caches for that address, \ie other
-caches are in a state less than or equal to $toCompatible(y)$
-(Figure \ref{table:fns}). This is needed to ensure the single-writer property of
-the MSI protocol -- if a cache's state is $M$ for an address $a$, every other cache
-must be in $I$ for address $a$. But, the caveat is that the L1 cache can access
-only its own state and data, as opposed to that of other caches.
+\begin{wrapfigure}{r}{.4\linewidth}
+\centering
+\begin{tabular}{|c|ccc|}
+\hline
+& $M$ & $S$ & $I$\\
+\hline
+$M$ & $=$ & $>$ & $>$\\
+$S$ & $<$ & $=$ & $>$\\
+$I$ & $<$ & $<$ & $=$\\
+\hline
+\end{tabular}
+\caption{$<$ and other relations for MSI states}
+\label{msi<}
+\end{wrapfigure}
 
-\begin{figure}
-\begin{tabularx}{\linewidth}{|cX|}
+Each cache stores the information given in Figure \ref{table:lineinfo}. The
+coherence state for an address denotes the (usual) permission that a cache has
+for the address: $M = \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$
+relation for coherence state values based on the permissions the states
+represent. This is shown in Figure \ref{msi<}. State $I$ is akin to address $a$
+not present in the cache.
+
+\begin{wrapfigure}{r}{.46\linewidth}
+\begin{subfigure}{.27\linewidth}
+\centering
+\begin{tabular}{|ccc|}
 \hline
-$toCompatible(s)$ & If the state of one cache is $s$ for address $a$, the
-highest state of another cache for the same $a$
-$\newline
-toCompatible(M) = I,\newline
-toCompatible(S) = S,\newline
-toCompatible(I) = M
-$\\
+$M$ & $\rightarrow$ & $I$\\
+$S$ & $\rightarrow$ & $S$\\
+$I$ & $\rightarrow$ & $I$\\
 \hline
-$x > y$ & Denotes if state $x$ has more permissions than state $y$.
-$M > S, M > I, S > I$. Relations $\le, >, \ge$ are also defined appropriately\\
+\end{tabular}
+\caption{$toCompatible$}
+\label{toCompat}
+\end{subfigure}
+~~~~~~~~~~~~~~
+\begin{subfigure}{.21\linewidth}
+\centering
+\begin{tabular}{|c|c|}
 \hline
-\end{tabularx}
-\caption{Abstraction functions}
-\label{table:fns}
-\end{figure}
+$M$ & $\checkmark$\\
+$S$ & $\times$\\
+$I$ & $\times$\\
+\hline
+\end{tabular}
+\caption{$isModified$}
+\label{isModified}
+\end{subfigure}
+\caption{$toCompatible$ and $isModified$ mappings}
+\label{fig:msimap}
+\end{wrapfigure}
 
-The memory controller has access to the data from the memory but not to the
-state and data of the L1 caches. Instead, it only has access to a directory
-which has a version of the coherence states of the caches for each address. The
-version of the state in the directory may not be up to date, \ie may not be the
-same as the one in the caches, all the time. Figure \ref{table:dirinfo} shows
-the information stored in the memory and directory. We will use the word memory
-and directory interchangeably for the 2-level protocol. The caches can not
-communicate with each other but can communicate with the directory via messages.
+If a cache $c$ receives a request for an address $a$ from the processor, it can
+service the request straightaway if it has enough permissions for address $a$.
+Otherwise, it has to \emph{upgrade} to a state $x$, which has enough
+permissions. The cache can not upgrade its state arbitrarily, it has to ensure
+that such a state change does not violate the single-writer property (Invariant
+\label{singleWriter}), by \emph{downgrading}, \ie lowering the states of other
+caches appropriately. We define a $toCompatible(x)$ mapping (Figure
+~\ref{toCompat}), which specifies the highest state a cache can be, if another
+cache is in state $x$ for the same address. We will also define a
+$isModified(x)$ mapping to specify if an address in state $x$ in a cache can
+have data different from what's present in the memory.
 
-\begin{figure}
-\begin{tabularx}{\linewidth}{|cX|}
-\hline
-$dir[c][a]$ & directory's version of the coherence state of cache $c$ for address $a$\\
-$memory[a]$ & data in the memory address $a$\\
-\hline
-\end{tabularx}
-\caption{Values stored by a directory and memory for an address}
-\label{table:dirinfo}
-\end{figure}
-
-Let's us assume that we have a system which guarantees the invariants given
-in Figure \ref{FirstInv}.
 \floatstyle{boxed}
@@
 \begin{figure}
-\begin{enumerate}
-\item The cache can upgrade $state[c][a]$ from $x$ to $y$ where $y > x$, \ie
-\emph{downgrade} its directory state only after it receives the appropriate
-authorization from the directory for address $a$. \label{cwaitresp}
-\item If the cache wants to upgrade $state[c][a]$ to $x$, there is a
-message \Req{c}{Dir}{a}{x} which cache $c$ can send to the directory requesting
-the upgrade authorization, and the cache will eventually be able to send
-the request. \label{csendreq}
-\item If the directory receives an upgrade request \Req{c}{Dir}{a}{x} from cache
-$c$ for address $a$, then just before receiving the request, $dir[c][a] < x$.
-\label{drecvreq}
-\item The directory can change $dir[c][a]$ from $x$ to $y$ where $y < x$, \ie
-\emph{downgrade} its directory state only after it receives the appropriate
-authorization from cache $c$ for address $a$. \label{dwaitresp}
-\item If the directory wants to downgrade $dir[c][a]$ to $x$, there is a
-message \Req{Dir}{c}{a}{x} which the directory can send to cache $c$ requesting
-the downgrade authorization, and the directory will eventually be able to send
-the request. \label{dsendreq}
-\item If the directory has sent a downgrade request for an address $a$ to cache
-$c$, it will eventually get the downgrade authorization response for address $a$
-from cache $c$ to downgrade to $x$, where $x$ is less than or equal to the state
-to which the directory sent a request to downgrade to. \label{drecvresp}
-\item If $dir[c][a] = M$ just before receiving a downgrade authorization from
-cache $c$ for address $a$, then the cache would have sent the data along with
-the authorization. \label{drecvdata}
-\item If the directory wants to send an authorization message to upgrade the
-state of address $a$ in cache $c$ to $x$ (\Resp{Dir}{c}{a}{x}), it will
-eventually be able to send it. \label{dsendresp}
-\item The directory can send the data from the memory along with the upgrade
-authorization message if it wants to. \label{dsenddata}
-\item The directory's version of the cache's state is always conservative, \ie
-$\forall c, a,\; dir[c][a] \ge state[c][a]$. \label{conservative}
-\end{enumerate}
-\caption{Initial set of invariants \FirstInv}
-\label{FirstInv}
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uReq}{$c , p, a, x$}
+  \If {$a$ is not present in $p$}
+    \State Choose address $a'$ for eviction;
+    \ForAll {$c' \in p.children$}
+      \State \call{} \dReq($p , c', a', I$);
+    \EndFor
+    \State \call{} \dResp($p , p.parent, a', I$);
+    \State Replace $a'$ with $a$ (initializing $p.state[a]$ to $I$);
+  \EndIf
+  \ForAll {$c' \in p.children, s.t., c' \neq c$}
+    \State \dReq($p , c', a, toCompatible(x)$);
+  \EndFor
+  \State \uReqL($p , p.parent, a, x$);
+  \State \uResp($p , c, a, x$);
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Handling upgrade to $x$ request from $c$ to $p$ for address $a$}
+\label{uReq}
+\end{subfigure}
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\dReq}{$p , c, a, x$}
+  \ForAll {$c' \in c.children$}
+    \State \dReqL($c , c', a, x$);
+  \EndFor
+  \State \dResp($c , p, a, x$);
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Handling downgrade to $x$ request from $p$ to $c$ for address $a$}
+\label{dReq}
+\end{subfigure}
+
+\caption{MSI protocol}
+\label{msi}
 \end{figure}
 
-\floatstyle{plain}
-\restylefloat{figure}
+\begin{figure}
 
-We will walk through how the MSI coherence protocol (with no replacement
-evictions) can be implemented using the above invariants and argue why it is a
-correct implementation.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uReqL}{$c , p, a, x$}
+  \State \call{} \uReq($c, p, a, x$);
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending upgrade to $x$ request from $c$ to $p$ for address $a$}
+\label{uReq}
+\end{subfigure}
 
-The directory's role is to ensure the single-writer property and that the last
-updated data is being read. In order to implement the cache coherence protocols
-using the invariants of Figure \ref{FirstInv}, the directory's ensures that at
-all times, if for some $c, dir[c][a] = M$, the for every other $c', dir[c'][a] =
-I$.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\dReq}{$p , c, a, x$}
+  \State \call{} \dReq($p, c, a, x$);
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending downgrade to $x$ request from $p$ to $c$ for address $a$}
+\label{dReq}
+\end{subfigure}
 
-If a cache gets a request from the core which can not be handled because of lack
-of permissions, it (eventually) sends an upgrade request to the directory.
-\FirstInv.\ref{csendreq} ensures that such a request can eventually be sent.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uResp}{$p , c, a, x$}
+  \If {$c.state[a] < x$}
+    \If {$c.state[a] = I$}
+      \State $c.data[a] \gets p.data[a]$;
+    \EndIf
+    \State $c.state[a] \gets x$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending upgrade to $x$ response from $p$ to $c$ for address $a$}
+\label{uResp}
+\end{subfigure}
 
-If the directory receives a request from cache $c$ to upgrade the state of
-address $a$ to $x$, because of \FirstInv.\ref{drecvreq}, $dir[c][a] < x$ and
-since $cache[c][a] \le dir[c][a]$ because of \FirstInv.\ref{conservative},
-$cache[c][a] < x$ and hence the directory must send an upgrade authorization
-response. The directory sends a request \Req{Dir}{c'}{a}{toCompatible(x)} to
-every cache $c'$ for which $dir[c'][a] > toCompatible(x)$. Such a request can be
-sent eventually because of \FirstInv.\ref{dsendreq}.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\dResp}{$c , p, a, x$}
+  \If {$c.state[a] > x$}
+    \If {$isModified(c.state[a])$}
+      \State $p.data[a] \gets c.data[a]$;
+    \EndIf
+    \State $c.state[a] \gets x$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending downgrade to $x$ response from $c$ to $p$ for address $a$}
+\label{dResp}
+\end{subfigure}
 
-Because of \FirstInv.\ref{drecvresp}, the directory will eventually receive
-responses from each of the requested caches $c'$
-(\Resp{c'}{Dir}{a}{toCompatible(x)}) authorizing the downgrade of $dir[c'][a]$
-to $toCompatible(x)$. If $dir[c'][a] = M$ just before receiving the
-authorization, then by \FirstInv.\ref{drecvdata}, the directory would also
-receive data from $c'$. Because the directory maintains the single-writer
-invariant, every other cache $c''$ would have $dir[c''][a] = I$ just before
-receiving the authorization. This implies for every other cache $c''$,
-$state[c''][a] = I$ by the \FirstInv.\ref{conservative} invariant, and hence
-$c'$ has the most up-to-date copy for address $a$. Thus the memory is updated
-with the received up-to-date copy for address $a$.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uResp}{$p , core, a, x$}
+  \State $c.data[a] \gets p.data[a]$;
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending upgrade to $x$ response from $p$ to $core$ for address $a$}
+\label{uResp}
+\end{subfigure}
 
-Once all the authorizations have been received, for every cache $c'$,
-$dir[c'][a] \le toCompatible(x)$ (other than the one that made the request $c$).
-Since the \FirstInv.\ref{conservative} invariant ensures that $cache[c'][a] \le
-dir[c'][a]$, we have $cache[c'][a] \le toCompatible(x)$, ensuring the
-single-writer property (because $toCompatible(M) = I$). The directory can now
-change $dir[c][a]$ to $x$ and send \Resp{Dir}{c}{a}{x} authorizing $c$ to
-upgrade state of address $a$ to $x$ (\FirstInv.\ref{dsendresp} ensures that such
-a message can be sent). Just before sending the authorization, if $dir[c][a] =
-I$, it implies that $state[c][a] = I$ because of the
-\FirstInv.\ref{conservative} invariant.  Thus the cache doesn't have data
-corresponding to address $a$ and the directory has to supply the most up to date
-version of the address (from memory, since the memory is up to date), which it
-can because of \FirstInv.\ref{dsenddata}.
+\caption{Atomic MSI}
+\label{atomic}
+\end{figure}
 
-The \FirstInv\ invariants essentially makes it straight forward to design the
-behavior of the directory . For the sake of argument, let
-\FirstInv.\ref{drecvreq} not be guaranteed. Consider a scenario the cache $c$
-which has sent a request \Req{c}{Dir}{a}{M} earlier but hasn't received the
-upgrade authorization response yet. Meanwhile, $c$ sends another upgrade request
-\Req{c}{Dir}{a}{S}. Let us assume that the directory processed the requests in
-the order in which they were sent. Once the directory has finished processing
-the first request, it changes its directory state $dir[c][a]$ to $M$. When it
-receives the second request, it is not easy to decide if the directory has to
-drop the request or to send another response authorizing the cache to change to
-$S$.
+\begin{figure}
 
-The difficulty in message passing protocols comes from the fact that it is
-difficult to guarantee the above invariants, even for a 2-level memory hierarchy
-discussed above as shown in the examples below.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uReqL}{$c , p, a, x$}
+  \State \send{} \Req{c}{p}{a}{x};
+  \State \pop{} \Resp{p}{c}{a}{z};
+  \If {$c.state[a] = I$}
+    \State \pop{} \Data{p}{c}{a}{d};
+    \State $c.data[a] \gets d$;
+  \EndIf
+  \State $c.state[a] \gets z$;
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending upgrade to $x$ request from $c$ to $p$ for address $a$}
+\label{uReqLocal}
+\end{subfigure}
 
-\subsubsection*{Example 1: Inability to send messages because of network
-congestion:}
-For instance, the ability of a cache or the directory to send messages
-(\FirstInv.\ref{dsendreq}, \FirstInv.\ref{dsendresp}, \FirstInv.\ref{csendreq},
-\etc) depends upon the availability of free buffers in the network; if the
-network remains congested all the time, then it would not be possible to send the
-messages, and the system would deadlock.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\dReqL}{$p , c, a, x$}
+  \State \send{} \Req{p}{c}{a}{x};
+  \State \pop{} \Resp{c}{p}{a}{z};
+  \If {$isModified(p.dir[c][a])$}
+    \State \pop{} \Data{c}{p}{a}{d};
+    \State $p.data[a] \gets d$;
+  \EndIf
+  \State $p.dir[c][a] \gets z$;
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending downgrade to $x$ request from $p$ to $c$ for address $a$}
+\label{dResp}
+\end{subfigure}
 
-\subsubsection*{Example 2: Multiple upgrade requests from the cache for the same
-address:}
-If a cache is allowed to make multiple requests to upgrade the state of an
-address (say to $M$ and then to $S$) before receiving the upgrade
-authorizations, then it might violate \FirstInv.\ref{drecvreq} if the requests
-were processed by the directory in the order they were received.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uResp}{$p , c, a, x$}
+  \If {$p.dir[c][a] < x$}
+    \If {$p.dir[c][a] = I$}
+      \State \send{} \Data{p}{c}{a}{p.data[a]};
+    \EndIf
+    \State $p.dir[c][a] \gets x$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending upgrade to $x$ response from $p$ to $c$ for address $a$}
+\label{uResp}
+\end{subfigure}
 
-\subsubsection*{Example 3: On sending an upgrade request, the cache waits till
-it gets an authorization:}
-If the behavior of the caches were such that if a cache $c$ has sent an
-upgrade request \Req{c}{Dir}{a}{x} to the directory, it keeps waiting for the
-upgrade authorization response \Resp{Dir}{c}{a}{x} before processing any
-requests from the directory for address $a$, then the cache coupled with our
-simple directory implementation would deadlock. Consider a scenario with two
-caches $X$ and $Y$, with both the caches in state $S$ for address $a$. Both the
-caches got a store request from the processor and want to upgrade their states
-to $M$, and hence send \Req{X}{Dir}{a}{M} and \Req{Y}{Dir}{a}{M}, respectively,
-to the directory. Let us say the directory receives \Req{Y}{Dir}{a}{M} first,
-and in order to process it, according to our implementation of the directory,
-sends a downgrade request \Req{Dir}{X}{a}{I} to cache $X$. But the cache 
-does not process any request from the directory for address $a$, according to
-our assumption, and hence will never send the downgrade authorization
-\Resp{X}{Dir}{a}{I} to the directory, violating \FirstInv.\ref{drecvresp}, and
-hence causing a deadlock.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\dResp}{$c , p, a, x$}
+  \If {$c.state[a] > x$}
+    \If {$isModified(c.state[a])$}
+      \State \send{} \Data{c}{p}{a}{c.data[a]};
+    \EndIf
+    \State $c.state[a] \gets x$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending downgrade to $x$ response from $c$ to $p$ for address $a$}
+\label{dResp}
+\end{subfigure}
 
-\subsubsection*{Example 4: At the directory, authorization responses are blocked
-by upgrade requests:}
-Consider a scenario where two caches $X$ and $Y$ are both in state $S$ for address
-$a$. $Y$ got a store request from the processor and sends an upgrade request
-\Req{Y}{Dir}{a}{M} to the directory. The directory receives that request, and
-sends a downgrade request \Req{X}{Dir}{a}{I} to $X$. $X$ sends a downgrade
-authorization \Resp{X}{Dir}{a}{I} to the directory on receiving the downgrade
-request. But the directory can not receive the authorization because it is
-blocked by other requests (including \Req{Y}{Dir}{a}{M}). This violates
-\FirstInv.\ref{drecvresp} and hence creates a deadlock.
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uResp}{$p , core, a, x$}
+  \State \send{} \Data{p}{c}{a}{p.data[a]};
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Sending upgrade to $x$ response from $p$ to $core$ for address $a$}
+\label{uResp}
+\end{subfigure}
 
-These examples show that if the \FirstInv\ invariants are violated, then it
-either breaks cache coherence (by creating deadlocks) or it makes correctly
-designing the directory very difficult.
-
-Before we show how the \FirstInv\ invariants can be guaranteed (which requires
-the description of the behavior of the L1 caches), let us consider a
-slightly more complicated system where caches can evict lines for replacement.
-
-\subsection{Message-passing directory-based MSI protocol for 2-levels, with
-replacement evictions}
-We again have just two levels of memory hierarchy -- the L1 caches and the main
-memory. But the caches have finite capacity. If receives a request from the core
-for an address $a$ and has no space to accommodate address $a$, it evicts
-another address $a'$ to make room for $a$.
-
-In the previous scenario, where the caches didn't have to evict for replacement,
-a cache only had to send downgrade authorizations to the directory on receiving
-a corresponding request. But in the new scenario, a cache can send a downgrade
-notification to the directory on evicting an address. Such a notification is
-useful in keeping the directory's version of the cache's state up to date, so we
-will restrict the family of directory-based protocols that we discuss to those
-where the caches send such a notification, voluntarily.
-
-The directory handles this notification in the exact manner by which it handled
-a downgrade authorization response from the cache, so the notification message
-is the same as a downgrade authorization message from cache $c$ for authorizing
-the downgrade of the $dir[c][a]$ to $I$ (\Resp{c}{Dir}{a}{I}).
-
-We will now discuss a behavior of the L1 cache that would violate the
-invariant that the directory's version of the states of the caches for each
-address is conservative.
-
-\subsubsection*{Example 5: Cache sends voluntary notifications
-(\Resp{c}{Dir}{a}{I}) indiscriminately}
-Consider a system with two caches $X$ and $Y$. Let $state[X][a] = S, state[Y][a]
-= I$ and let $dir[X][a] = S, dir[Y][a] = I$. $X$ receives a store request from
-the core, and hence sends an upgrade request \Req{X}{Dir}{a}{M} to the
-directory. Say $X$ was a non-blocking cache, and hence receives another load
-request for address $a'$ from the core before receiving an upgrade authorization
-for address $a$. $X$ does not have space in the cache for $a'$, and it decides
-to evict address $a$, and sends \Resp{X}{Dir}{a}{I} to the directory, changing
-$state[c][a]$ to $I$.  Meanwhile, the directory receives the request
-\Req{X}{Dir}{a}{M} and since $dir[Y][a] \le I$, it responds to $X$ with an
-upgrade authorization \Resp{Dir}{X}{a}{M} (along with the data). Later, the
-directory receives the notification \Resp{X}{Dir}{a}{I} from $X$, and downgrades
-$dir[X][a]$ to $I$. Meanwhile, $X$ receives the upgrade authorization
-\Resp{Dir}{X}{a}{M} and changes $state[X][a]$ to $M$. Now,
-\FirstInv.\ref{conservative} is broken, since $state[X][a] = M$ and $dir[X][a] =
-I$. Let $Y$ get a load request from its core, and hence sends a
-\Req{Y}{Dir}{a}{S} to the directory. The directory on receiving the request,
-sends an upgrade authorization to $Y$ because $dir[X][a] = I$. This violates the
-single-writer property and hence coherence.
+\caption{Realistic MSI protocol}
+\label{realistic}
+\end{figure}

### 2012-11-18 paper/isca40format/MsiNew.tex
will be removing evict req

\section{A two-level message-passing directory-based MSI protocol}
\label{sec:msi}

\newcommand{\fourAngle}[6]{\text{$#1\langle#2\rightarrow#3,#4,#5\rightarrow#6\rangle$}}
\newcommand{\threeAngle}[5]{\text{$#1\langle#2\rightarrow#3,#4,#5\rangle$}}

%\newcommand{\Req}[5]{\fourAngle{Req}{#1}{#2}{#3}{#4}{#5}}
\newcommand{\Req}[4]{\threeAngle{Req}{#1}{#2}{#3}{#4}}
\newcommand{\Resp}[4]{\threeAngle{Resp}{#1}{#2}{#3}{#4}}

\newcommand{\msg}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
%\newcommand{\Req}[5]{\msg{Req}{#1}{#2}{#3}{#4}{#5}}
%\newcommand{\Resp}[5]{\msg{Resp}{#1}{#2}{#3}{#4}{#5}}
%\newcommand{\Data}[4]{\msg{Data}{#1}{#2}{#3}{#4}}
%\newcommand{\Mesg}[4]{\msg{Mesg}{#1}{#2}{#3}{#4}}
%\input{Atomic}

Let us walk through our framework using the example of an MSI protocol.  We will
start with a simple system for the sake of explanation.  The system has two
levels of memory hierarchy -- private L1 caches for each core and a shared
memory. The L1 caches do not have direct access to the memory, and the memory
does not have a direct access to the actual states in the  L1 caches, or L1's
data. Instead, the main memory has direct access to a directory associated with
it which keeps a version of the L1 cache states for each address. The L1 caches
do not have direct access to this directory. The directory's version of the L1
cache states may differ from the actual L1 cache states. We use the terms memory
and directory interchangeably.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for address $a$\\
$data[c][a]$ & data in cache $c$ for address $a$\\
\hline \end{tabularx}
\caption{Values stored by an L1 cache $c$ for an address $a$}
\label{table:lineinfo}
\end{figure}

\begin{wrapfigure}{r}{.4\linewidth}
\centering
\begin{tabular}{|c|ccc|}
\hline
& $M$ & $S$ & $I$\\
\hline
$M$ & $=$ & $>$ & $>$\\
$S$ & $<$ & $=$ & $>$\\
$I$ & $<$ & $<$ & $=$\\
\hline
\end{tabular}
\caption{$<$ and other relations for MSI states}
\label{msi<}
\end{wrapfigure}

Each L1 cache stores the information given in Figure \ref{table:lineinfo}. The
coherence state for an address denotes the permission that a cache has for the
address: $M = \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$ relation for
coherence state values based on the permissions the states represent. This is
shown in Figure \ref{msi<}. State $I$ is akin to the cache not having the data
corresponding to $a$, since it can neither read or write the $a$'s data. Figure
\ref{table:dirinfo} shows the information stored in the memory and directory.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$dir[c][a]$ & directory's version of the coherence state of cache $c$ for address $a$\\
$memory[a]$ & data in the memory address $a$\\
\hline
\end{tabularx}
\caption{Values stored by a directory and memory for an address}
\label{table:dirinfo}
\end{figure}

The L1 caches and the directory communicate with each using via messages. Each
L1 cache communicates with its core via 2 buffers -- incoming for requests and
outgoing for responses. Each L1 cache communicates with the memory directly via
4 buffers -- two incoming buffers from the memory, one each for requests and
responses from the memory, and two outgoing buffers into the memory, one each
for requests and responses to the memory. All these buffers converge into or
diverge from 4 buffers of the memory -- two incoming buffers, one each for
requests and responses from the caches and two outgoing buffers, one each for
requests and responses to the caches (Figure \ref{fig:setup}). All these
buffers have the FIFO property. Not only that, we will also assume that requests
sent from a source will never reach its destination till an earlier response
sent by the same source has been serviced (and hence removed) by the destination.
%We will show in Section \ref{sec:fifo} about how these network requirements can
%be realistically implemented.

\subsection{No evictions due to replacement}
\label{sec:msi-magic}

We will first design an MSI protocol in which the L1 caches have enough
capacity (or rather associativity) so that no line gets evicted due to
replacement. Lines may however get evicted to enforce the following invariant
which is needed for maintaining cache coherence.

%\begin{inv}
\textbf{Single-writer Invariant}: When a cache has permission to modify data for
an address ($M$ state), no other cache has read or write permission for that
address. In other words, the other caches must be in $I$ state.
\label{singleWriter}
%\end{inv}

\begin{wrapfigure}{r}{.46\linewidth}
\begin{subfigure}{.27\linewidth}
\centering
\begin{tabular}{|ccc|}
\hline
$M$ & $\rightarrow$ & $I$\\
$S$ & $\rightarrow$ & $S$\\
$I$ & $\rightarrow$ & $I$\\
\hline
\end{tabular}
\caption*{$toCompatible$}
\label{toCompat}
\end{subfigure}
~~~~~~~~~~~~~~
\begin{subfigure}{.21\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $\checkmark$\\
$S$ & $\times$\\
$I$ & $\times$\\
\hline
\end{tabular}
\caption*{$isModified$}
\label{isModified}
\end{subfigure}
\caption{$toCompatible$ and $isModified$ mappings}
\label{fig:msimap}
\end{wrapfigure}

Because of the Single-writer invariant, it makes sense to talk about
compatibility of the states of two caches for an address. Two cache states $x$
and $Y$ are \emph{compatible} if two caches having the same address in states
$x$ and $y$ do not violate the Single-writer invariant. We define a mapping
$toCompatible(x)$ which gives the highest state which is compatible with a given
state $x$ (Figure \ref{fig:msimap}).

In addition to Single-writer Invariant, maintaining cache coherence also
requires the following invariant:

%\begin{inv}
\textbf{Read-from-last-writer}: When a cache reads data for an address, it reads
the last modified version of the data for the address. If the data for an
address has not been modified, the cache reads the data from memory.
%\end{inv}

Function $isModified(x)$ determines if data for an address state $x$ can
potentially be modified or not (Figure \ref{fig:msimap}).

If a cache $c$ receives a request for an address $a$ from the processor, it can
service the request straightaway if it has enough permissions for address $a$.
Otherwise, it has to get to a state for $a$, say $y$, which has enough
permissions. In other words, it has to \emph{upgrade} its state to $y$. In order
to maintain the Single-writer Invariant, cache $c$ has to
ensure that the states of other caches are \emph{compatible} with $y$. But the
caveat is that $c$ can access only its own state and data, as opposed to that of
other caches (or the directory).

Cache $c$ has to instead communicate with the directory to ensure that the
states of other caches are compatible. $c$ must send a message
\Req{c}{Dir}{a}{y} requesting the directory to tell if it can upgrade its state
for address $a$ to $y$. We call this an \emph{upgrade request}. Let us now take
for granted that such a message can be sent (Invariant \ref{csendreq} in Figure
\ref{DirInv}).

We will also assume that that a cache can upgrade its state to $y$ only after
receiving and servicing a message from the directory \Resp{Dir}{c}{a}{y}
permitting it to upgrade its state for address $a$ to $y$ (Invariant
\ref{cwaitresp} in Figure \ref{DirInv}). We call such a message as an
\emph{upgrade response}.

We introduce two other kinds of messages: \emph{downgrade request}
\Req{Dir}{c'}{a}{y} sent from the directory to a cache $c'$ asking $c'$ to
downgrade its state for address $a$ to $y$, and \emph{downgrade response}
\Resp{c'}{Dir}{a}{y} sent from $c'$ to the directory informing the directory
that $c'$ has changed its state $state[c'][a]$ from $x$ to $y$, where $x > y$.

How can the behavior of the directory be designed (in terms of sending and
receiving messages) to ensure that cache $c$ will get an upgrade response
eventually. Let's say the invariants given of Figure \ref{DirInv} are
somehow guaranteed (we have already assumed that invariants \ref{csendreq} and
\ref{cwaitresp} are guaranteed). With these invariants, let us design the
behavior of the directory.

\floatstyle{boxed}
\restylefloat{figure}
\begin{figure}
\begin{inv}
A cache will eventually be able to send an upgrade request.
\label{csendreq}
\end{inv}
\begin{inv}
Cache $c$ can upgrade $state[c][a]$ from $x$ to $y$ where $y > x$ only after it
receives and services an upgrade response \Resp{Dir}{c}{a}{y} from the directory.
\label{cwaitresp}
\end{inv}
\begin{inv}
Just before starting to service an upgrade request \Req{c}{Dir}{a}{y} from $c$,
$dir[c][a] < y$.
\label{drecvreq}
\end{inv}
\begin{inv}
The directory will eventually be able to send any downgrade request to any cache
$c$.
\label{dsendreq}
\end{inv}
\begin{inv}
The directory can change $dir[c][a]$ from $x$ to $y$ where $y < x$, \ie
downgrade its directory state only after receiving and servicing a downgrade
response \Resp{c}{Dir}{a}{y} from $c$.
\label{dwaitresp}
\end{inv}
\begin{inv}
If the directory has sent a downgrade request \Req{Dir}{c}{a}{y} to cache $c$,
it will eventually get a downgrade response \Resp{Dir}{c}{a}{z} from cache $c$
such that $z \le y$.
\label{drecvresp}
\end{inv}
\begin{inv}
When a cache $c$ sends a downgrade response \Resp{c}{Dir}{a}{y} to the
directory and $state[c][a] = x$ just before sending the downgrade response, then
$dir[c][a] = x$ from the time the response was sent by $c$ to the time the
directory services the response.
\label{cknows}
\end{inv}
\begin{inv}
When a directory sends an upgrade response \Resp{Dir}{c}{a}{y} to cache $c$ and
$dir[c][a] = x$ just before sending the downgrade response, then $cache[c][a] = x$
from the time the response was sent by the directory to the time the $c$
services the response.
\label{dknows}
\end{inv}
%\begin{inv}
%Just before receiving a downgrade response from $c$, if $isModified(dir[c][a])$
%is true, then $c$ would have sent the data along with the downgrade response.
%\label{drecvdata}
%\end{inv}
\begin{inv}
The directory will eventually be able to send any upgrade response to any cache
$c$.%It can also send the data to cache $c$ along with the upgrade response.
\label{dsendresp}
\end{inv}
\begin{inv}
Conservative directory:
\begin{spacing}{.1}
\begin{equation*}
\forall c, a,\; dir[c][a] \ge state[c][a]
\end{equation*}
\end{spacing}
\label{conservative}
\end{inv}
\caption{Directory invariants}
\label{DirInv}
\end{figure}
\floatstyle{plain}
\restylefloat{figure}

%The directory ensures that the states of any two caches for an address are
%compatible by the following behavior.
%When the directory receives an upgrade request
%\Req{c}{Dir}{a}{x}, Invariant \ref{drecvreq} guarantees that $dir[c][a] < x$ and
%because of Invariant \ref{conservative}, $cache[c][a] le dir[c][a]$, resulting
%in $cache[c][a] < x$. Unless the directory sends an upgrade response to $c$
%permitting it to upgrade to $x$, $c$ will remain in state $cache[c][a]$ forever
On receiving an upgrade request \Req{c}{Dir}{a}{x}, the directory sends a
downgrade request \Req{Dir}{c'}{a}{toCompatible(x)} to every cache $c'$
for which $dir[c'][a] > toCompatible(x)$. Such a request can be sent eventually
because of Invariant \ref{dsendreq}. The incoming upgrade request is not
removed and no further upgrades requests are processed till an appropriate
upgrade response is sent back from the directory to $c$.

Because of Invariant \ref{drecvresp}, the directory will eventually receive
downgrade responses \Resp{c'}{Dir}{a}{z} from each of the requested caches $c'$
such that $z \le toCompatible(x)$ (since the directory made a downgrade request
to go to $toCompatible(x)$). Just before servicing the downgrade response, if
$isModified(dir[c'][a])$ is true (\ie $dir[c'][a] = M$), the directory be
 waits for data to be sent by $c'$. $c'$ knows that the directory's
version of its state for address $a$, at the time \Resp{c'}{Dir}{a}{z} is
serviced by the directory will be the same as the state of $c'$ right before it
sends the downgrade response (by Invariant \ref{cknows}), and hence would know
if the directory is expecting data by simply checking for
$isModified(state[c'][a]$ at the time of sending the downgrade response. So $c'$
can be implemented to send the data along with the response if required. If the
directory receives data, the memory for address $a$ is updated with the received
data.

Once all the downgrade responses have been received and serviced, since no
upgrade responses have been sent to any cache in the meanwhile, we have $\forall
c' \neq c, dir[c'][a] \le toCompatible(x)$. Because of Invariant
\ref{conservative}, it is guaranteed that $\forall c' \neq, cache[c'][a] \le
toCompatible(x)$. The directory now changes $dir[c][a]$ to $x$ and sends an
upgrade response \Resp{Dir}{c}{a}{x} to $c$. If $dir[c][a] = I$ right before
changing it to $x$, then by Invariant \ref{dknows}, $cache[c][a]$ is going to be
$I$ till cache $c$ services the upgrade response. So, in this case, the data
from the memory for address $a$ is also sent along with the upgrade response.
Note that such an upgrade response, can always be sent eventually by the
directory because of Invariant \ref{dsendresp}. Finally, the upgrade response
from $c$ has been serviced and hence removed by the directory.

The above behavior for the directory ensures that the states of any two caches
for every address are compatible, thus obeying Single-writer Invariant. This can
be proven by noting that whenever an upgrade to $x$ request from $c$ is being
serviced, every other cache is ensured to be in a state $\le toCompatible(x)$
before an upgrade response is eventually sent to $c$. Similarly, it can be
easily shown to obey Read-from-last-writer Invariant.

%The above behavior for the directory ensures that the states of any two caches
%for every address are compatible, as can seen by induction.  This property is
%true initially because every cache is in state $I$ for every address. If this
%property is true before the servicing of a request is started, it will be true by the
%time the request is serviced and an upgrade response is sent. This is because an
%upgrade response is sent only when every cache other than the one requesting is
%at a state $\le toCompatible(x)$, where $x$ is the state the requesting cache
%wants to go to, and $toCompatible(x)$ is the highest state that a cache can be
%in to be compatible with another cache in state $x$ for an address.

%Whenever the directory receives a downgrade response from a cache $c'$, just
%before servicing the response if $isModified(dir[c'][a])$ is true, \ie
%$dir[c'][a] = M$, $c'$ must have been the only cache with permission to modify
%address $a$. This is because the directory ensures that the states of any two
%caches for address $a$ is compatible at all times (as shown above). Thus, to
%ensure the Read-from-last-writer Invariant, the potentially modified data from
%$c'$ is written back from into the memory on receiving the downgrade response
%from $c'$ as seen in our design.

The invariants \ref{drecvreq} to \ref{conservative} in Figure \ref{DirInv}
essentially makes it straight forward to design the behavior of the directory.
Without these invariants, it will be hard to reason about the correctness of the
protocol as shown by the following examples.

For the sake of argument, let us assume Invariant \ref{drecvreq} not be
guaranteed. Consider a scenario in which the cache $c$ which has sent a request
\Req{c}{Dir}{a}{M}. Before servicing an upgrade response for this request, $c$
sends another upgrade request \Req{c}{Dir}{a}{S}. Once the directory has
finished processing the first request, it changes its directory state
$dir[c][a]$ to $M$. When it starts servicing the second request, it is not easy
to decide if the directory has to drop the request or to send an upgrade to $S$
response.

As another example, let us assume that Invariant \ref{dknows} does not hold.
Let's say the $dir[c][a] = S$ right before the directory sends an upgrade
response \Resp{Dir}{c}{a}{M} to cache $c$. The directory will not send the data
because in its version of the state of $c$ for address $a$, cache $c$ already
has the data. But if the directory was wrong, and at the time cache $c$ services
the upgrade response, if $state[c][a]$ is $I$, then cache $c$ will no longer
have the right data, breaking the Read-from-last-writer Invariant.

The difficulty in message passing protocols comes from the fact that it is
difficult to guarantee the above invariants, even for a 2-level memory hierarchy
discussed above. as shown in the following scenarios.

%These scenarios show that if the invariants of Figure \ref{DirInv} are violated,
%then it either breaks cache coherence (by creating deadlocks) or it makes
%it hard to reason about the behavior of the directory.

%\begin{scen}
%\emph{Inability to send messages because of network congestion:}
%The ability of a cache or the directory to send messages (Invariants
%\ref{dsendreq}, \ref{dsendresp} and \ref{csendreq}) depends upon the
%availability of free buffers in the network; if the network remains congested
%all the time, then it would not be possible to send the messages, and the system
%would deadlock.
%\end{scen}
%
%\begin{scen}
%\emph{Multiple upgrade requests from the cache for the same address:}
%If a cache is allowed to make multiple requests to upgrade the state of an
%address (say to $M$ and then to $S$) before receiving the upgrade responses,
%then it might violate Invariant \ref{drecvreq} if the requests were processed by
%the directory in the order they were received. This makes the decision on
%whether the directory drops the second request or not hard to reason about, as
%shown above.
%\end{scen}

\begin{scen}
\emph{On sending an upgrade request, the cache stops processing downgrade
requests it gets an upgrade response:}
Consider a scenario with two caches $c$ and $c'$, with both the caches in state
$S$ for address $a$. Both the caches get a store request from the processor and
send \Req{c}{Dir}{a}{M} and \Req{c'}{Dir}{a}{M}, respectively, to the directory.
Let's say \Req{c'}{Dir}{a}{M} arrives at the directory first. The directory
sends a downgrade request to cache $c$, but since $c$ does not process any
downgrade requests till it gets its upgrade response, it will never send back a
downgrade response to the directory, violating Invariant \ref{drecvresp},
resulting in a deadlock. Note that this deadlock would occur even if a cache
waiting for an upgrade response for an address, stops processing downgrade
requests for other addresses.
\label{noprocess}
\end{scen}

\begin{scen}
\emph{At the directory, downgrade responses from caches are blocked by upgrade
requests:}
Consider a scenario where two caches $c$ and $c'$ are both in state $S$ for
address $a$. $c'$ gets a store request from the processor and sends an upgrade
request \Req{c'}{Dir}{a}{M} to the directory. The directory receives that
request, and sends a downgrade request \Req{c}{Dir}{a}{I} to $c$. $c$ sends a
downgrade response \Resp{c}{Dir}{a}{I} to the directory on receiving the
downgrade request. But the directory can not service the downgrade response because
it is blocked by other requests (including \Req{c'}{Dir}{a}{M}). This violates
Invariant \ref{drecvresp} and hence creates a deadlock.
\label{scen:block}
\end{scen}

Let us consider a slightly more complicated system where caches can evict lines
for replacement before proceeding further.

\subsection{Evictions due to replacement are allowed}
If a cache starts servicing a request from the core for an address $a$ and has
no space to accommodate address $a$, it evicts another address $a'$ to make room
for $a$.

In the previous scenario, the caches didn't have to evict for replacement; a
cache had to send downgrade responses to the directory only on servicing a
corresponding request. But in the new scenario, a cache can send a downgrade
notification voluntarily to the directory on evicting an address. The directory
has to process this new notification: on servicing the downgrade to $y$
notification for an address $a$ from cache $c$, the directory changes
$dir[c][a]$ to $y$, keeping the directory's version of the cache's state
up-to-date. Since the directory can not distinguish between the downgrade
notification sent voluntarily and a downgrade response sent because of a
downgrade request from the directory, we will use the same message to denote
both.

%We will restrict the family of directory-based protocols that that
%we discuss to those in which the caches send such notifications.

We will first discuss an example of the behavior of the L1 cache that would
violate Invariant \ref{conservative}, if the cache is allowed to voluntarily
send a downgrade response due to eviction.

\begin{scen}
\emph{Cache sends voluntary downgrade responses indiscriminately:}
Consider a system with only two caches $c$ and $c'$ (Figure \ref{fig:cross})
such that $state[c][a] = dir[c][a] = S$, and $state[c'][a] = dir[c'][a] = I$.
Let $c$ be a non-blocking cache, so it can process requests from the core even
if there are pending misses. $c$ receives a store request for address $a$ from
the core and sends \Req{c}{Dir}{a}{M} to the directory. Before it gets back an
upgrade response from the directory, it receives another load request for
address $a'$ from the core. Suppose the line for $a'$ is not present in $c$ and
$c$ has no space for $a'$, it decides to evict address $a$ to make room for
$a'$. It voluntarily sends a downgrade response \Resp{c}{Dir}{a}{I} to the
directory. Meanwhile, the directory receives the upgrade request
\Req{c}{Dir}{a}{M}, and since the directory state of the only other cache is $I$
for address $a$, the directory sends back an upgrade response. The directory
then receives the downgrade response \Resp{c}{Dir}{a}{I} and changes $dir[c][a]$
to $I$. $c$ finally receives the upgrade response \Resp{Dir}{c}{a}{M} and
upgrades $state[c][a]$ to $M$. This breaks Invariant \ref{conservative}, since
$state[c][a] = M$ and $dir[c][a] = I$. If the directory gets an upgrade to $M$
request for address $a$ from $c'$, it will assume that $c$ is in state $I$ for
address $a$ and send an upgrade to $M$ response to $c'$. This will violate the
Single-writer invariant, thereby violating coherence.
\label{indisc}
\end{scen}

\subsection{Meta-rules that guarantee the invariants of Figure \ref{DirInv}}

We now show meta-rules for implementation of a message-passing directory-based
protocol in order to guarantee the invariants of Figure \ref{DirInv}; this has
been formally proven using the Coq theorem prover. We will present some
intuition behind the proof by showing how these constraints avoid scenarios
\ref{noprocess}, \ref{scen:block} and \ref{indisc}, though the formal proof is
beyond the scope of the paper. These meta-rules are amenable to direct
implementation because each of the meta-rules access only local states.

%We will present the intuition behind the proof by showing how
%these constraints avoid scenarios \ref{noprocess}, \ref{scen:block} and
%\ref{indisc}, though the formal proof is beyond the scope of the paper.

\floatstyle{boxed}
\restylefloat{figure}
\begin{figure}
\begin{cons}
Requests for an address $a$ should not block responses for address $a$, even
from a different source.
\end{cons}
\caption{Meta-rules for blocking}
\label{block}
\end{figure}
\begin{figure}
\begin{cons}
Responses to the directory for the same address $a$ from the same cache $c$
should be serviced by the directory in the order they were sent.
\end{cons}
\begin{cons}
Requests for an address $a$ from a source $n$ (cache or directory) should not be
serviced by a destination $m$ (directory or cache) before $m$ has serviced all the
responses for the same address $a$ from the same source $n$ sent earlier. In
other words, requests between caches and directory should not overtake responses
between caches and directory if they are sent from the same source to the same
destination for the same address.
\end{cons}
\caption{Meta-rules for ordering}
\label{order}
\end{figure}

Figures \ref{block} and \ref{order} gives the meta-rules about which messages
should not block each other, and the ordering requirements between messages. The
setup in Figure \ref{fig:setup} and the stated FIFO requirements guarantees
these meta-Rules. We will show how these meta-rules can be realistically
realized in Section \ref{sec:fifo}.

\begin{figure}
\begin{cons}
A response for address $a$ can be sent by a cache to the directory or by the
directory to the cache only if the line for $a$ is present in the sending node.
\end{cons}
\begin{cons}
A response from the directory to cache $c$ for address $a$ can be sent only if a
request from $c$ for $a$ is currently being serviced by the directory.
\ref{dreqhandle}
\end{cons}
\begin{cons}
If $dir[c][a] = x$, a directory can send a response \Resp{Dir}{c}{a}{y} only if
$y > x$ and $dir[c][a]$ changes from $x$ to $y$ on sending the response.
\end{cons}
\begin{cons}
If there is a pending response from the directory for an address $a$, a cache
$c$ can not voluntarily send a downgrade response for address $a$ to the
directory. But $c$ can send a downgrade response to the directory if it is
servicing a downgrade request for address $a$ from the directory.
\ref{creqhandle}
\end{cons}
\begin{cons}
If $state[c][a] = x$, a cache $c$ can send a response \Resp{c}{Dir}{a}{y} only if
$y < x$ and $state[c][a]$ changes from $x$ to $y$ on sending the response.
\end{cons}
\begin{cons}
On servicing a response \Resp{Dir}{c}{a}{x}, the directory changes $dir[c][a]$
to $x$.
\end{cons}
\begin{cons}
On servicing a response \Resp{c}{Dir}{a}{x}, cache $c$ changes $state[c][a]$
to $x$.
\end{cons}
\begin{cons}
$state[c][a]$ can change only if cache $c$ sends a response to the directory or
services a response from the directory, for address $a$.
\end{cons}
\begin{cons}
$dir[c][a]$ can change only if the directory sends a response to cache $c$ or
services a response from cache $c$, for address $a$.
\end{cons}
\caption{Sending and servicing responses}
\label{sendRespPre}
\end{figure}

Figure \ref{sendRespPre} shows the the meta-rules regarding when responses can
be sent, and the state changes accompanying the sending and servicing of a
response. In these meta-rules, ``being serviced'' means having started the
servicing of a request.  For example, if servicing a request requires more
requests to be sent, ``being serviced'' corresponds to having started sending
those requests.

Meta-Rule \ref{creqhandle} prevents Scenario \ref{indisc} from happening. A
cache $c$ should not send a voluntary downgrade response \Resp{c}{Dir}{a}{I} if
it is waiting for a pending response from the directory, which was being
violated in that scenario ($c$ evicted address $a$ to replace it with address
$a'$ while still waiting for an upgrade response for $a$ from the directory).

Meta-Rule \ref{dreqhandle} states that a directory can not voluntarily send an
upgrade response to a cache. This again leads to Invariant \ref{conservative}
being violated. The intuition behind having meta-rules \ref{creqhandle} and
\ref{dreqhandle} is that they prevent ``crossing'' of two response messages as
shown in Figure \ref{crossing}. This preserves Invariant \ref{conservative}.

\begin{figure}
\begin{cons}
A request \Req{Dir}{c}{a}{x} can be sent only if there is no pending response
from cache $c$ for address $a$, and $x < dir[c][a]$ just before the directory
sends the request.
\label{dirreq}
\end{cons}
\begin{cons}
A request \Req{c}{Dir}{a}{x} can be sent only if there is no pending response
from the directory for address $a$, and $x > state[c][a]$ just before the
directory sends the request.
\label{creq}
\end{cons}
\begin{cons}
Requests from the directory for an address $a$ should be serviced even if there
is a pending response from the directory for any address (either $a$ or other
addresses).
\label{cblock}
\end{cons}
\caption{Sending and servicing requests}
\label{sendReqPre}
\end{figure}

Figure \ref{sendReqPre} shows the meta-rules regarding when requests can be
sent. If Meta-Rule \ref{cblock} is obeyed, then Scenario \ref{noprocess} will
be avoided.

### 2012-11-18 paper/isca40format/Threads.tex
will be removing evict req

\section{Caches as a distributed system of multi-threaded execution engines}
\label{sec:threads}

The memory hierarchy in a processor can be specified as a collection of nodes
(Figure \ref{distNodes}). Each node has local state associated with it. The
state may contain the cache and the directory among other book-keeping
information. A node also has multiple input and multiple output channels
through which it can send and receive messages. Most of the channels are
connected to a network, which delivers messages between nodes
%; the message delivery order is left unspecified
. Some of the channels may not be connected to the network; they represent
inputs and outputs to the memory system, \viz the requests from and responses to
cores.

The behavior of each node can be best described as a system of suspendable local
threads, with a scheduler local to that node.  A thread can be in one of the two
states: \emph{executing} or \emph{suspended}. Once a thread finishes executing,
the scheduler does one of the following: a) picks one of suspended threads which
has become ready to execute due to changed conditions or b) creates a new thread
to handle an incoming message. An executing thread can read the local state
of the node, send messages to output channels, receive messages from input
channels or spawn more threads. The input and output channels are blocking, so
if a thread tries to send to a full output channel or receive from an empty
input channel, the thread would not be able to proceed. The thread can also
fail to proceed due to some condition in the local state of the node. In
all these cases, the thread gets suspended. It remains suspended till the
required conditions are met (for instance, the output channel becomes free, the
input channel receives a pertinent message, or other local state changes),
Once the required conditions for a suspended thread are met the scheduler may
pick it for execution. The execution is resumed from the exact point where it
was suspended. Once the thread finishes its execution, all the book-keeping
resources associated with that thread is freed up and reclaimed.

Incoming messages remain in the channel until they are explicitly removed by
the thread dealing with that message. If the scheduler sees an incoming message
that requires a new thread to be created for handling, and if the required
book-keeping resources are available, then a thread is created.
Different threads require different amounts of book-keeping resources. If the
required book-keeping resources are not available, the scheduler might pick
another suspended thread which is ready to execute, or if that's not possible,
spin-waits.  The threads and incoming messages have a memory address associated
with them. For an incoming message that does not require a new thread to be
created, its address is used by the scheduler to decide if a suspended thread
is ready to execute. Similarly, the address associated with a thread that has
just finished executing is used to determine if the local state of the node
associated with that address has changed so that a suspended thread associated
with that address can resume execution.

Note that the thread analogy we described above has nothing to do with software
threads. We are describing the behavior of a cache controller using the familiar
notion of suspendable threads. The threads essentially correspond to the various
requests being handled by a cache controller's Miss Status Handler Register
(MSHR).

Earlier, each thread is started,, because of an incoming message, with a
handler \textbf{procedure} particular to the type of the message. A procedure
groups together actions which include state updates ($s \gets value$),
\textbf{send} messages into output channels and \textbf{pop} messages from
input channels. \textbf{send} and \textbf{pop} are blocking calls, as mentioned
earlier. Procedures can \textbf{call} other procedures but they can not be
recursive.

### 2012-11-18 paper/isca40format/ArvindGood/Abstract.tex
stuff

We provide a framework for designing hierarchical directory-based
cache-coherence protocols. We reduce the cache-coherence design problem into two
parts: a) a common substrate shared between most directory-based cache coherence
protocols and b) the part specific to each cache-coherence protocol. The common
substrate includes a set of ``\glob{}s'' which provide the framework over which
protocol-specific parts, the ``\policy{}'' can be designed. \glob{}s can be
thought of as a high-level abstraction layer, which allows \policy{} to focus on
high-level properties, instead of worrying about corner cases. \glob{}s include
guarantees like every request from any node getting a response eventually, \etc.
\policy{} includes the actions the directory takes in order to preserve cache
coherence, \viz to preserve the single-writer-at-a-time invariant, and the
reader-reads-last-updated-value invariant. We have formally proven that a set of
(\policy{} independent) ``\local{}s'' guarantee \glob{}s using the Coq theorem
prover. The \local{}s dictate constraints and actions local to a cache or to the
directory as well as properties of the network.

Our framework is flexible and extensible to design multiple hierarchical cache
coherence protocols like MSI, MESI, MOSI, MOESI, \etc -- these protocols are
just different \policy{}s. Our framework has no extra overhead in terms of
state-bits or number of messages transmitted. In fact, the protocols designed
using our framework does not require acks for some of the scenarios where acks
are needed in commonly used protocols \cite{hammer, moesi}. We show that the
overall performance of the protocols designed using our framework is also better
in because of the lack of acks.

%implementations using our framework has no extra overhead in terms of
%state-bits or number of messages passed to service each request. In fact, we
%show that the protocols designed using our framework does not require acks for
%some of the scenarios in which commonly used protocols \cite{hammer, moesi} use
%acks. Our framework
%
%The part specific to each protocol, the ``\policy{}'', includes what actions a
%directory should take on receiving a request from its caches, whether to ask a
%cache to forward the data to another cache, \etc. We have formally proven that
%a set of easy-to-implement ``\local{}s'' guarantee the ``\glob{}s'' rules like
%every request by a cache or a directory will eventually get a response, We
%present a method to develop implementations of hierarchical, message-passing,
%directory-based, cache-coherence protocols. Our method is based on the cache
%system preserving two invariants: a) the directory's notion of the permissions
%for the caches it serves is ``conservative'', \ie the caches do not have more
%permission than what is assumed by the directory, and b) every request by a
%cache or a directory will eventually get a response. We specify an abstract
%protocol, that is, a set of meta rules for state transitions, which has been
%shown, using the Coq theorem prover, to preserve these invariants. We have also
%shown that this abstract protocol can be implemented using $2N + 1$ virtual
%channels where $N$ is the number of levels and each channel can have as few as
%one buffer. 
%
%In this paper, we use the abstract protocol to develop formally proven
%implementations of a number of well-known protocols like MSI, MESI, MOSI and
%MOESI, both in the presence as well as in the absence of point-to-point FIFO
%networks. All these protocols can be expressed in terms of an extra
%``compatibility of state'' check in the state transitions allowed by the
%abstract protocol. We finally show that the protocols we design using our
%methodology have no extra overhead (in terms of state bits or the number of
%messages passed to service each request) compared to the existing
%implementations of these protocols.

### 2012-11-18 paper/isca40format/ArvindGood/Init.tex
stuff

\section{Initialize}
\label{sec:init}

Cache coherence protocols maintain coherence by ensuring the following two
invariants:

\begin{inv}
\textbf{Single-writer Invariant}: When a cache has permission to modify data for
an address, no other cache has read or write permission for that
address.
\label{singleWriter}
\end{inv}

\begin{inv}
\textbf{Read-from-last-writer}: When a cache reads data for an address, it reads
the last modified version of the data for the address. If the data for an
address has not been modified, the cache reads the data from memory.
\label{lastRead}
\end{inv}

Because of the requirement to maintain the Single-writer property (Invariant
\ref{singleWriter}), it makes sense to talk about compatibility of the states of
two caches for an address. Two coherence states $x$ and $x$ are \emph{compatible}
if two caches having the same address in states $x$ and $x$ do not violate the
single-writer property.

### 2012-11-18 paper/isca40format/ArvindGood/Introduction.tex
stuff

--- 2012-11-18 paper/isca40format/Introduction.tex
@@
 concrete and familiar scenarios from two-level systems (a private L1 cache for
-each processor and a memory with a directory):
+each processor and a memory with a directory). States $M, S$ and $I$ have their
+usual meaning in MSI protocol ($M$ represents read and write permissions for an
+address, $S$, read-only permissions and $I$, no permission)
 

### 2012-11-18 paper/isca40format/stuff/AbstractProtocol.tex
stuff

--- 2012-11-07 paper/isca40format/AbstractProtocol.tex
@@
 \newcommand{\One}[2]{\text{#1$\langle#2\rangle$}}
+\newcommand{\qOne}[2]{\text{#1[$#2$]}}
+\newcommand{\qTwo}[3]{\text{#1[$#2$][$#3$}}
 \newcommand{\two}[3]{\text{#1$\langle#2,#3\rangle$}}
@@
 \newcommand{\OrOr}{\;||\;}
-\newcommand{\Inv}{\text{$Inv$}}
+\newcommand{\Inv}{\text{$Inv$}\;}
 
@@
 
-The notation we adopt is primarily for conciseness of description while allowing
-maximum flexibility in the implementation.
-
-A node keeps information about the state of each cache line in the node as well
-as in each of its children, \ie the directory. Figure~\ref{table:states} shows
-the precise information that a node keeps for each cache line in the node. The
-notation we adopt is primarily for conciseness of description. For instance, the
-directory need not contain the exact copy of each of the child's coherence
-state, for each cache line, instead the encoding can be more efficient. We will
-show efficient implementations later.
+A node keeps information about the coherence state of each cache line in the
+node as well as in each of its children, \ie the directory.
+Figure~\ref{table:states} shows the precise information that a node keeps for
+each cache line in the node. The notation we adopt is primarily for conciseness
+of description. For instance, the directory need not contain the exact copy of
+each of the child's coherence state for each cache line; instead the encoding
+can be more efficient. All the information (except $present, data, state$ and
+$dir$) corresponds to a request being handled for a given address; an
+implementation can store this information only for the requests being processed
+instead of for each address.
 
@@
 \hline
-isPresent(a) & Check if cache line for address $a$ is present\\
+isPresent[a] & Is cache line for address $a$ present?\\
 \hline
-data(a) & Cache line data for address $a$\\
+data[a] & Cache line data for address $a$\\
 \hline
-state(a) & Coherence state for address $a$\\
+state[a] & Coherence state for address $a$\\
 \hline
-dir(a, c) & Directory's copy of the coherence state for address $a$ in child
+dir[a][c] & Directory's copy of the coherence state for address $a$ in child
 $c$\\
 \hline
-isCWait(c, a) & Check if the node waiting for a response from child $c$
-for address $a$\\
+isCWait[a] & Is node waiting for a response from any child for address $a$?\\
 \hline
-cWait(c, a) & The coherence state to which the node has requested child $c$ to
-go to, for address $a$\\
+cWaitState[a] & The coherence state to which the node has requested the children to
+downgrade to, for address $a$\\
 \hline
-isPWait(a) & Check if the node waiting for a response from its parent for
-address $a$\\
+isPWait[a] & Is node waiting for a response from its parent for
+address $a$?\\
 \hline
-whoTrigger(a) & The source of the request which made the node wait for a
-response. It is $None, Parent, Child \text{ or } Evict$\\
+isPReq[a] & Is node handling a request from parent for addess $a$\\
 \hline
-replacingAddr(a) & If $whoTrigger(a)$ is $Evict$, the address replacing the line
-of address $a$\\
+isCReq[a] & Is node handling a request from a child for address $a$\\
 \hline
-childTrigger(a) & If $whoTrigger(a)$ is $Child$, the index of the child which
-is the source of the request\\
+whichCReq[a] & The child whose request the node is handling for address $a$\\
+\hline
+isEvict[a] & Is the cache line for address $a$ being evicted?\\
+\hline
+replacingAddr[a] & If the cache line for address $a$ is being evicted, the address replacing the line\\
 \hline
@@
 totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
-defined between every pair of elements in that set and that relation is
+defined between any two set elements and this relation is
 transitive. For instance, in an MSI protocol, the values of the coherence state
@@
 permissions is $<$ the state with more permissions. We use \Inv to denote the
-least element in the set containing the values of the coherence state.
+least element in the set containing the values of the coherence state of a line.
 Typically, this denotes that the node has no permissions for reading or writing
@@
 Figure~\ref{table:messages} shows the incoming and outgoing messages to and from
-the node, respectively, and requests for eviction. An eviction request for an
-address is generated if a child has made a request for a different cache line
-whose address is not currently present in the cache.
+the node, respectively.
 
@@
 \three{RespFromC}{c}{a}{x} & Message from child $c$ denoting that $c$'s
-coherence state for address $a$ has changed to $x$\\
-\three{DataFromC}{c}{a}{d} & Data $d$ of the cache line received from $c$ for
-address $a$\\
+coherence state for address $a$ has downgraded to $x$\\
+\three{DataFromC}{c}{a}{d} & Data $d$ of the cache line received from child $c$
+for address $a$\\
 \two{RespFromP}{a}{x} & Response from parent giving permission to
@@
 \two{RespToP}{a}{x} & Message to parent denoting that the node's
-coherence state for address $a$ has changed to $x$\\
+coherence state for address $a$ has downgraded to $x$\\
 \two{DataToP}{c}{d} & Data $d$ of the cache line sent to the parent
@@
 \hline
-\multicolumn{2}{|c|}{Eviction messages}\\
-\hline
-\One{EvictReq}{a} & Request to evict cache line for address $a$ (generated by
-the node itself)\\
-\hline
 \end{tabular}
-\caption{Incoming, outgoing and eviction messages}
+\caption{Incoming and outgoing messages}
 \label{table:messages}
 \end{figure}
+
+Each message coming from a particular child or the parent is independent of the
+messages coming from another source, in addition to being independent of the
+messages for a different address. The node can process messages coming from
+different sources, or corresponding to different addresses, in any order,
+subject to the ordering requirements in Figure~\ref{list:messageOrder}. The
+message ordering restrictions are necessary to avoid deadlocks. We later show
+that is easy to enforce this ordering even if the underlying network doesn't
+enforce point-to-point FIFO ordering or priorities between different kinds of
+messages.
 
@@
 \begin{enumerate}
-\item A request or a response should not be handled before handling an earlier
-response for the same address from the same source
+\item A request or a response should not be handled before handling a response
+sent earlier from the same source for the same address
 
@@
 
-Each message coming from a particular child or the parent is independent of the
-messages coming from another source, in addition to being independent of the
-messages for a different address. The node can process messages coming from
-different sources, and for different addresses in any order, subject to the
-ordering requirements in Figure~\ref{list:messageOrder}. We later show that is
-easy to enforce this ordering even if the underlying network doesn't enforce
-point-to-point FIFO ordering or priorities between different kinds of messages.
-
-To allow maximum flexibility of implementation, the node is given separate
-access queues for each kind of message (request of response) from each source
-and for each address. An implementation of a protocol using our methodology does
-not have to keep a separate access queue for each source and each address, but
-it must adhere to the blocking restrictions in Figure~\ref{list:messageOrder}.
-These restrictions are necessary to avoid deadlocks. The flexibility of the
-abstract protocol allows one to design a wide range of blocking policies.
+%To allow maximum flexibility of implementation, the node is given separate
+%access queues for each kind of message (request of response) from each source
+%and for each address (\ie \qTwo{ReqFromC}{c}{a}, \qOne{ReqFromPQ}{a},
+%\qTwo{RespFromCQ}{c}{a}, \qTwo{ReqToCQ}{c}{a}, \qOne{RespToPQ}{a}, \etc). An
+%implementation of a protocol using our methodology does not have to keep a
+%separate access queue for each source and each address, but it must adhere to
+%the blocking restrictions in Figure~\ref{list:messageOrder}.  These
+%restrictions are necessary to avoid deadlocks. The flexibility of the abstract
+%protocol allows one to design a wide range of blocking policies.
 
@@
 function can be derived from $isCompat$ function, if $isCompat$ is defined
-correctly. Figure~\ref{fig:msiFunc} shows the functions defined for MSI (--
+correctly. Figure~\ref{fig:msiFunc} shows the functions defined for MSI (``--''
 denotes undefined).
@@
 \hline
-isCompat(x, y) & Checks if two different caches can have coherence states
-$x$ and $y$, respectively, for the same address\\
+isCompat(x, y) & Returns if two different sibling caches can have coherence
+states $x$ and $y$, respectively, for the same address or not\\
 \hline
@@
 \hline
-dataXfer(x, y) & Checks if cache line data has to be transferred by a cache to its
-parent on downgrading from $x$ to $y$\\
+dataXfer(x, y) & Return if cache line data has to be transferred by a cache to its
+parent on downgrading ist coherence state from $x$ to $y$ or not\\
 \hline
@@
 
-We now describe the procedures for a node to handle requests and responses. Note
+We now describe the how a node handles requests and responses.
+
+Note
 that a message handler would be triggered only if the corresponding message is
 present. If the required resources for handling a message are not present (for
-example, if the handling of the request from a child requires sending further
-requests to its children or the parent, and the send buffers are full), then the
-handler for request from a child would not be triggered.
+example, if the handling of the request requires sending further requests to
+its children or the parent, and the corresponding send buffers are full), then
+the handler for that message would not be triggered.
 
-Figure~\ref{alg:handlePReq} specifies how to handle a node's incoming request
+Figure~\ref{alg:handlePReq} specifies how a node handles an incoming request
 from the parent for downgrading the coherence state of address $a$ to $x$. Note
 that a request from the parent can be handled successfully only if the node is
-not already waiting for a response from a child.
+not already waiting for a response from some child.
 
-Figure~\ref{alg:handleCReq} specifies how to handle a node's incoming request
+Figure~\ref{alg:handleCReq} specifies how a node handles an incoming request
 from a child for upgrading the child's coherence state for address $a$ to $x$.
-Note that a request from a child can be handled successfully only if the node is
-not already waiting for a response from the parent or from any child.
+Note that a request from a child can be handled successfully only if the node
+is not already waiting for a response from the parent or from some child.
 
-Figure~\ref{alg:handlePResp} specifies how to handle a node's incoming response
-from the parent for upgrading the coherence state of address $a$ to $x$.
+Figure~\ref{alg:handlePResp} specifies how a node handles an incoming response
+from the parent for permitting upgrade of the coherence state of address $a$ to
+$x$.
 
-Figure~\ref{alg:handleCResp} specifies how to handle a node's incoming response
-from a child for downgrading the child's coherence state for address $a$ to $x$.
+Figure~\ref{alg:handleCResp} specifies how a node handles an incoming response
+from a child notifying the downgrade of the child's coherence state for address
+$a$ to $x$.
 
-Figure~\ref{alg:handleEvict} specifies how to handle an eviction message (in
-order to replace the current line in address $a$ with the line for some other
-address, which is given by $replacingAddr(a)$).
+XXX Figure~\ref{alg:handleEvict} specifies how a cache line replacement is handled.
 
@@
 \begin{algorithmic}
-\Rule{handlePReq}{\two{ReqFromP}{a}{x}}
-\If {$!isPresent(a)$}
-  \State remove$(\two{ReqFromP}{a}{x})$
-\ElsIf {$\forall c, !isCWait(c, a)$}
+\Rule{handlePReq}{a}
+   \State let mesg \two{ReqFromP}{a}{x} exist in \qOne{ReqFromPQ}{a}
+   \If {$\forall c, !isCWait(c][a)$}
      \If {$state(a) \le x$}
         \State remove$(\two{ReqFromP}{a}{x})$
-  \ElsIf {$\forall c, dir(c, a) \le x$}
+     \State remove$(\two{ReqFromP}{a}{x})$
+     \ElsIf {$\forall c, dir(c][a) \le x$}
         \State remove$(\two{ReqFromP}{a}{x})$
@@
         \ForAll{$c$}
-        \If {$dir(c, a) > x$}
-           \State $isCWait(c, a) \assigns \True$
-           \State $cWait(c, a) \assigns x$
+           \If {$dir(c][a) > x$}
+              \State $isCWait(c][a) \assigns \True$
+              \State $cWait(c][a) \assigns x$
               \State send$(\three{ReqToC}{c}{a}{x})$
@@
 \begin{algorithmic}
-\Rule{handleCReq}{\three{ReqFromC}{c}{a}{x}}
+\Rule{handleCReq}{$c, a$}
+   \State let mesg \two{ReqFromC}{c}{a}{x} exist in \qTwo{ReqFromCQ}{c}{a}
    \If {$isPresent(a)$}
-     \If {$!isPWait(a) \AndAnd \forall c, !isCWait(c, a)$}
+      \If {$!isPWait(a) \AndAnd \forall c, !isCWait(c][a)$}
          \If {$state(a) \ge x \AndAnd$ \\
@@
             \EndIf
-           \State $dir(c, a) \assigns x$
+            \State $dir(c][a) \assigns x$
             \State send$(\three{RespToC}{c}{a}{x})$
@@
       \If {$!isPWait(a') \AndAnd \forall c', !isCWait(c', a')$}
-        \State send$(\One{EvictReq}{a'})$
          \State \textcolor{trig-color}{$whoTrigger(a) \assigns Evict$}
@@
          \State $replacingAddr(a') \assigns a$
+         \State \textcolor{trig-color}{\textbf{trigger} handleEvict($a$)}
       \EndIf
@@
 \begin{algorithmic}
-\Rule{handleEvict}{\One{EvictReq}{a}}
-   \If {$!isPresent(a) \OrOr \forall c, dir(c, a) = \Inv$}
-      \State remove$(\One{EvictReq}{a})$
+\Rule{handleEvict}{$a$}
+   \If {$!isPresent(a) \OrOr \forall c, dir(c][a) = \Inv$}
       \If {$state(a) \neq \Inv$}
@@
       \State allocLine$(replacingAddr(a))$
-      \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
+      \State \textcolor{trig-color}{$whoTrigger(a) \assigns None$}
+      \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a), a$)}
    \Else
-      \If {$\forall c, !isCWait(c, a)$}
+      \If {$\forall c, !isCWait(c][a)$}
          \ForAll{$c$}
-            \If {$dir(c, a) > \Inv$}
-               \State $isCWait(c, a) \assigns \True$
-               \State $cWait(c, a) \assigns \Inv$
+            \If {$dir(c][a) > \Inv$}
+               \State $isCWait(c][a) \assigns \True$
+               \State $cWait(c][a) \assigns \Inv$
                \State send$(\three{ReqToC}{c}{a}{\Inv})$
@@
 \begin{algorithmic}
-\Rule{handlePResp}{\two{RespFromP}{a}{x}}
+\Rule{handlePResp}{$a$}
+   \State let mesg \two{RespFromP}{a}{x} exist in \qOne{RespFromPQ}{a}
    \State remove$(\two{RespFromP}{a}{x})$
    \If {$state(a) \neq \Inv$}
+      \State let mesg \two{DataFromP}{a}{d} exist in \qOne{DataFromPQ}{a}
       \State remove$(\two{DataFromP}{a}{d})$
@@
    \State $isPWait(a) \assigns \False$ 
-  \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
+   \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a), a$)}
 \EndRule
@@
 \begin{algorithmic}
-\Rule{handleCResp}{\three{RespFromC}{c}{a}{x}}
-  \State remove$(\two{RespFromC}{c}{a}{x})$
+\Rule{handleCResp}{$c, a$}
+   \State let mesg \three{RespFromC}{c}{a}{x} exist in \qTwo{RespFromCQ}{c}{a}
+   \State remove$(\three{RespFromC}{c}{a}{x})$
    \If {dataXfer$(dir(c,a), x)$}
-     \State remove$(\two{DataFromC}{c}{a}{d})$
+      \State let mesg \three{DataFromC}{c}{a}{d} exist in \qTwo{DataFromPQ}{c}{a}
+      \State remove$(\three{DataFromC}{c}{a}{d})$
       \State $data(a) = d$
@@
    \State $dir(c,a) \assigns x$
-  \If {$isCWait(c, a) \AndAnd cWait(c, a) >= x$}
-    \State $isCWait(c, a) \assigns \False$
+   \If {$isCWait(c][a) \AndAnd cWait(c][a) >= x$}
+     \State $isCWait(c][a) \assigns \False$
    \EndIf
@@
    \If {$whoTrigger(a) = Parent$}
-     \State \textbf{trigger} handlePReq
+      \State \textbf{trigger} handlePReq($a$)
    \ElsIf {$whoTrigger(a) = Evict$}
-     \State \textbf{trigger} handleEvict
+      \State \textbf{trigger} handleEvict($(a$)
    \ElsIf {$whoTrigger(a) = Child$}
-     \State \textbf{trigger} handleCReq($childTrigger(a)$)
+      \State \textbf{trigger} handleCReq($childTrigger(a), a$)
    \EndIf}
@@
 
-Function \emph{getReplace(a)} gives the address of the line that can be replaced
-in order to accommodate the line in address $a$. Note that if the line selected
-for eviction is waiting for a response from the parent or from any child, then
-it can not be replaced.
+Function \emph{getReplace(a)}, used in \textsc{handleCReq} gives the address of
+the line that can be replaced in order to accommodate the line in address $a$.
+Note that if the line selected for eviction is waiting for a response from the
+parent or from any child, then it can not be replaced.
 
@@
   \ForAll {$c$}
-     \State $dir(c, a) \assigns \Inv$
-     \State $isCWait(c, a) \assigns \False$
+     \State $dir(c][a) \assigns \Inv$
+     \State $isCWait(c][a) \assigns \False$
   \EndFor
@@
 
-\item After handling a response (or an evict request) the corresponding handler
+\item After handling a response (or an eviction) the corresponding handler
 must trigger the original handler that requested the received response (or made
@@
 The L1 caches, \ie the leaf nodes, do not have any children. Instead, they
-handle requests from cores, which resembles handling requests from children. A
-store request for a word is essentially a request by the core for write
-permissions for the corresponding cache line, and a load request for a word is a
-request for read permissions. These permissions can be mapped to the values of
-the coherence state.
-
-Because of the absence of children any request from the parent or an eviction
-request can be handled without producing more requests. As a result, we do not
-need a separate handler for eviction requests. We also do not need to store
-$whoTrigger$, since the only trigger that produces more requests, and hence has
-to wait for response is a request from the core. We omit the pseudo-code for
-handling a request from the core because it is very similar to an instance of
-Figure~\ref{alg:handleCReq} with no children.
+handle requests from cores. A store word request is essentially a request by
+the core for write permissions for the corresponding cache line, and a load
+word request is a request for read permissions. These permissions can be mapped
+to the values of the coherence state. Because of the absence of children any
+request from the parent as well as any eviction can be handled without
+producing more requests. As a result, we do not need a separate handler for
+eviction requests. We also do not need to store $whoTrigger(a)$, since the only
+trigger that produces more requests, and hence has to wait for response is a
+request from the core. We omit the pseudo-code for handling a request from the
+core because it is very similar to an instance of \textsc{handleCReq} with no
+children.
 

### 2012-11-18 paper/isca40format/stuff/Invariants.tex
stuff

\section{Invariants}
\label{sec:invariants}

\floatstyle{boxed}
\restylefloat{figure}

The caches are assumed to be organized in the form of a tree
(Figure~\ref{fig:hierarchy}). The leaves of this tree represent the L1 caches,
with each L1 cache connected to exactly one core. The root of the tree is a
single Last-level cache (LLC), connected to the main memory. A (cache) node can
communicate only with its parents or with its children via a request or
response message whose format is in Figure~\ref{message}.

\begin{figure}
\begin{inv}
An incoming request message from a child should not block an incoming request
message from the parent, even for a different address\label{cReqNoBlockPReq}
\end{inv}
\begin{inv}
An incoming request messages for an address $a$ should not block an incoming
response message for address $a$, even from a different
source.\label{reqNoBlockReq}
\end{inv}
\caption{Blocking restrictions}
\label{blocking}
\end{figure}

\begin{figure}
\begin{inv}
An incoming response message for an address $a$ from a child $c$ should not be
received before another incoming response message from the same $c$ and the
same $a$ sent earlier has been received\label{cRespFifo}
\end{inv}
\begin{inv}
An incoming request message for an address $a$ from a source $n$ should not be
received before another incoming response message from the same $n$ and the
same $a$ sent earlier has been received\label{reqNoOvertakeResp}
\end{inv}
\caption{Ordering requirements}
\label{order}
\end{figure}

\begin{figure}
\begin{inv}
A response \Resp{p}{c}{a}{x}, where $p = c.parent$, can be sent from $p$ only
if
\begin{enumerate}
\item a thread that handling a request from $c$ to $p$ for
address $a$ is executing, and
\item $x > p.dir[c][a]$
\end{enumerate}
\label{cSendRespPre}
\end{inv}
\begin{inv}
If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$
\label{cSendRespPost}
\end{inv}
\begin{inv}
A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
if $x < c.state[a]$ \label{pSendRespPre1}
\end{inv}
\begin{inv}
If there is a thread in node $c$ that is waiting to receive a response from its
parent $p$ for an address $a$, then response \Resp{c}{p}{a}{x} can be sent only
on executing a thread that is handling a request from $p$ to $c$ for address
$a$. For instance, $a$ can not be evicted from $c$, if there is a thread
handling a request from $p$ to $c$ for address $a$.
\label{pSendRespPre2}
\end{inv}
\begin{inv}
If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$
\label{pSendRespPost}
\end{inv}
\caption{Invariants for sending responses}
\label{sendResp}
\end{figure}

\begin{figure}
\begin{inv}
On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
\gets x$
\label{pRecvResp}
\end{inv}
\begin{inv}
On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
\gets x$
\label{cRecvResp}
\end{inv}
\caption{Invariants for receiving responses}
\label{recvResp}
\end{figure}

\begin{figure}
\begin{inv}
$c.state[a]$ can change only on $c$ sending a response to or receiving a
response from $c.parent$
\label{cState}
\end{inv}
\begin{inv}
$p.dir[c][a]$ can change only on $p$ sending or receiving a response from
its child $c$
\label{pState}
\end{inv}
\caption{Invariants for state changes}
\label{stateChange}
\end{figure}

\begin{figure}
\begin{inv}
\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$
\label{pSendReqPre}
\end{inv}
\begin{inv}
\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$
\label{cSendReqPre}
\end{inv}
\caption{Invariants for sending a request}
\label{sendReq}
\end{figure}

\begin{figure}
\begin{inv}
$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{pHandleReq}
\end{inv}
\begin{inv}
$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from $p$, or
\item a thread in $p$ is handling a request for address
$a$ from one of $p$'s children $c'$, and that thread will not send any
(downgrade) request to any child for address $a$ anymore
\end{enumerate}
\label{cHandleReq}
\end{inv}
\caption{Invariants for receiving a request (used by the scheduler)}
\label{recvReq}
\end{figure}

\begin{figure}
\begin{inv}
Address $a$ present in node $c$ can not be chosen for eviction by a thread
executing in $c$ if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{evict}
\caption{Invariants for suspending a thread during choosing an address for eviction}
\end{inv}
\end{figure}

### 2012-11-18 paper/isca40format/stuff/InvariantsOld.tex
stuff

--- 2012-11-09 paper/isca40format/Invariants.tex
@@
 \newcommand{\Inv}{\text{$Inv$\;}}
+\newcommand{\Dirty}{\text{$Dirty$\;}}
 
@@
 \hline
-$isEnterPending(a)$ & Is there any (child) request for address $a$ waiting to get
-a cache line entry in the cache?\\
-\hline
+%$isEnterPending(a)$ & Is there any (child) request for address $a$ waiting to get
+%a cache line entry in the cache?\\
+%\hline
 $isHandling(n, a)$ & Is there any request for address $a$ from node $n$ (either
@@
 request from child $c$ for $a$ is being handled
-\item Response \Resp{c}{a}{x} can be sent only if $x > dir[a][c]$, $\forall c',
-c' \neq c \rightarrow isCompatible(x, dir[a][c'])$ and  $x \le state[a]$
+\item Response \Resp{c}{a}{x} can be sent only if $x > dir[a][c]$
+% $\forall c', c' \neq c \rightarrow isCompatible(x, dir[a][c'])$ and  $x \le state[a]$
 \end{enumerate}\\
@@
 response to the parent for $a$ can be sent any time
-\item Response \Resp{P}{a}{x} can be sent only if $x < state[a]$ and $\forall
-c, x \ge dir[a][c]$
+\item Response \Resp{P}{a}{x} can be sent only if $x < state[a]$
+% and $\forall c, x \ge dir[a][c]$
 \end{enumerate}\\
@@
 \hline
-\begin{itemize}
-\item[] If $x \le state[a] \wedge \forall c', c' \neq c \rightarrow isCompatible(x, dir[a][c])$\newline
- then the node must send a response \Resp{c}{a}{x}
-to $c$
-\end{itemize}\\
+\begin{enumerate}
+\item If $x \le state[a] \wedge \forall c'\neq c, isCompatible(x, dir[a][c])$,
+ then the node must send a response \Resp{c}{a}{x} to $c$ if $x > dir[a][c]$
+\item If $x = \Inv$ then data for address $a$ is transferred to $c$
+\end{enumerate}\\
 \hline
@@
 \begin{enumerate}
-\item[] If $\forall c, dir[a][c] \le \Inv$, then the node must send a response
-\Resp{P}{a}{\Inv} to the parent
+\item If $\forall c, dir[a][c] \le \Inv$, then the node must send a response
+\Resp{P}{a}{\Inv} to the parent if $\Inv < state[a]$
+\item If $state[a] \in \Dirty$ then data for address $a$ is transferred to $c$
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item[] If $\forall c, dir[a][c] \le x$, then the node must send a response
-\Resp{P}{a}{x} to the parent
+\item If $\forall c, dir[a][c] \le x$, then the node must send a response
+\Resp{P}{a}{x} to the parent if $x < state[a]$
+\item If $state[a] \in \Dirty$ then data for address $a$ is transferred to the
+parent
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item[] If the state or dir of an address (either the requested address, or the
-evicted address in case of a replacement) for some node $n$ is not satisfactory,
-and there are no pending responses from $n$, send a request to make the state
-or dir satisfactory
+\item If the request that is being handled for address $a$ requires $state[a]$
+to be $x$ while currently $state[a] < x$ and there is no pending response from
+the parent send \Req{P}{a}{x} to the parent
+\item If the request that is being handled for address $a$ requires $dir[a][c]$
+for some $c$ to be $x$ while currently $dir[a][c] > x$ and there is no pending
+response from child $c$ send \Req{c}{a}{x} to $c$
 \end{enumerate}\\
@@
 \begin{enumerate}
-\item[] The effects during sending a response or sending a request is atomic,
-\ie if there is no space available in the sending buffer, then the whole action
-is retried later
+\item The effects during sending a response or sending a request is atomic,
+\ie if there is no space available in the sending buffer, then the action is
+aborted without causing any effect and retried later
+\item The abort can be achieved by checking if the required resources are
+available before starting the action 
 \end{enumerate}\\

### 2012-11-18 paper/isca40format/stuff/InvariantsSomething.tex
stuff

--- 2012-11-14 paper/isca40format/Invariants.tex
@@
 \label{sec:invariants}
-
-%\subsection{Desired goals}
-%\begin{enumerate}
-%\item Permissions can be Nothing, Read (only), Read and Write, \ie $\{\}, {R}, {R, W}$
-%\item To upgrade = to gain more permissions
-%\item To downgrade = to lose permissions
-%\item If the child cache wants to gain some permissions for an address, it can
-%do so only after the parent directory has authorized it
-%\end{enumerate}
-%
-%\begin{enumerate}
-%\item Permissions that a child cache has for an address is a subset of the
-%permissions that the parent directory believes the child has for that address
-%\item If the parent directory wants a child cache to lose some permissions for
-%an address, there is a message that the directory can send notifying the child
-%of the directory's intentions (this message is called a downgrade request)
-%\item If the cache wants to gain some permissions for an address, there is a
-%message that the child can send notifying the directory of the child's
-%intentions (this message is called an upgrade request)
-%\item If a node wants to send a message to some node, then it can send such the
-%message eventually (\ie it can never be blocked from sending the message
-%forever)
-%\item If the parent directory has sent a downgrade request for an address $a$
-%to a child cache $c$, the directory will eventually get a message from $c$
-%notifying the directory that the cache has given up the required permissions
-%for address $a$ in $c$
-%\item If the child cache has sent an upgrade request for an address $a$ to
-%parent directory, the child will eventually get a message from the directory
-%authorizing the upgrade
-%\end{enumerate}
-%
-\newcommand{\threeAngle}[4]{\text{$#1\langle#2,#3,#4\rangle$}}
-
-\newcommand{\Req}[3]{\threeAngle{Req}{#1}{#2}{#3}}
-\newcommand{\Resp}[3]{\threeAngle{Resp}{#1}{#2}{#3}}
-\newcommand{\Inv}{\text{$Inv$} }
-\newcommand{\Dirty}{\text{$Dirty$} }
-\newcommand{\True}{\text{$True$} }
-\newcommand{\False}{\text{$False$} }
 
@@
 \begin{figure}
-%\centering
-\begin{minipage}{5.5cm}
-\begin{figure}[H]
-\centering
-\includegraphics[scale=0.5]{hierarchy}
-\caption{Cache hierarchy}
-\label{fig:hierarchy}
-\end{figure}
-\end{minipage}
-\begin{minipage}{3cm}
-\begin{figure}[H]
-\centering
-\includegraphics[scale=0.53]{network}
-\caption{One possible network organization}
-\label{fig:network}
-\end{figure}
-\end{minipage}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|cX|}
+\begin{tabularx}{\linewidth}{|X|}
 \hline
-\multicolumn{2}{|c|}{\Req{n}{a}{x}, \Resp{n}{a}{x}}\\
+\multicolumn{1}{|c|}{Blocking restrictions}\\
 \hline
-$n:$ & Source or destination node ID (parent $P$, or a child)\\
-$a:$ & Address of the line corresponding to the request\\
-$x:$ & Value from the set of coherence states\\
+\begin{enumerate}
+\item A request from a child should not block the handling of a request from
+the parent, even for a different address.
+\item Requests for an address $a$ should not block the handling of responses
+for address $a$, even from a different source.
+\end{enumerate}\\
+\hline
+\multicolumn{1}{|c|}{Ordering requirements}\\
+\hline
+\begin{enumerate}
+\item A response for an address $a$ from a child $c$ should not be received
+before a response from the same $c$ and the same $a$ sent earlier has been
+received.
+\item The handling of a request for an address $a$ from a source $n$ should not
+be started before a response from the same $n$ for the same $a$ sent earlier
+has been received.
+\end{enumerate}\\
 \hline
 \end{tabularx}
-\caption{Incoming and outgoing message}
-\label{message}
+\caption{Blocking restrictions and order of receiving messages}
+\label{receiveOrder}
 \end{figure}
@@
 the underlying network does not guarantee FIFO ordering between two nodes.
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\multicolumn{1}{|c|}{Blocking restrictions}\\
-\hline
-\begin{enumerate}
-\item A request from a child should not block (\ie prevent start of the
-handling of) a request from the parent
-\item Requests for an address $a$ should not block (\ie prevent reception of)
-responses for address $a$, even from a different source
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{Ordering requirements}\\
-\hline
-\begin{enumerate}
-\item A response for an address $a$ from a child $c$ should not be received
-before a response from the same $c$ and the same $a$ sent earlier has been
-received and removed
-\item The handling of a request for an address $a$ from a source $n$ should not
-be started before a response from the same $n$ for the same $a$ sent earlier
-has been received and removed
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Blocking restrictions and order of receiving messages}
-\label{receiveOrder}
-\end{figure}
 

### 2012-11-18 paper/isca40format/stuff/Threads.tex
stuff

--- 2012-11-18 paper/isca40format/Threads.tex
@@
 threads, with a scheduler local to that node. A thread can be in one of the two
-states: \emph{executing} or \emph{suspended}. Once a thread finishes executing,
-the scheduler does one of the following: a) picks one of suspended threads which
-has become ready to execute due to changed conditions or b) creates a new thread
-to handle an incoming message. An executing thread can read the local state
-of the node, send messages to output channels, receive messages from input
-channels or spawn more threads. The input and output channels are blocking, so
-if a thread tries to send to a full output channel or receive from an empty
-input channel, the thread would not be able to proceed. The thread can also
-fail to proceed due to some condition in the local state of the node. In
-all these cases, the thread gets suspended. It remains suspended till the
-required conditions are met (for instance, the output channel becomes free, the
-input channel receives a pertinent message, or other local state changes),
-Once the required conditions for a suspended thread are met the scheduler may
-pick it for execution. The execution is resumed from the exact point where it
-was suspended. Once the thread finishes its execution, all the book-keeping
-resources associated with that thread is freed up and reclaimed.
+states: \emph{executing} or \emph{suspended}. At a given time, either one
+thread or the scheduler will be executing. When the scheduler is executing,
+it does one of the following: a) picks one of suspended threads
+which has become ready to execute due to changed conditions or b) creates a new
+thread to handle an incoming message. An executing thread can read the local
+state of the node, send messages to output channels, receive messages from
+input channels, \etc% or spawn more threads . The input and output channels are
+blocking, so if a thread tries to send to a full output channel or receive from
+an empty input channel, the thread would not be able to proceed. The thread can
+also fail to proceed due to some condition in the local state of the node. In
+all these cases, the thread gets suspended, giving back control to the
+scheduler. It remains suspended till the required conditions are met (for
+instance, the output channel becomes free, the input channel receives a
+pertinent message, or other local state changes), Once the required conditions
+for a suspended thread are met the scheduler may pick it for execution. The
+execution is resumed from the exact point where it was suspended. Once the
+thread finishes its execution, all the book-keeping resources associated with
+that thread is freed up and reclaimed.
 
@@
 book-keeping resources are available, then a thread is created.
-Different threads require different amounts of book-keeping resources. If the
+%Different threads require different amounts of book-keeping resources.
+If the
 required book-keeping resources are not available, the scheduler might pick
@@
 
-Earlier, each thread is started,, because of an incoming message, with a
-handler \textbf{procedure} particular to the type of the message. A procedure
-groups together actions which include state updates ($s \gets value$),
-\textbf{send} messages into output channels and \textbf{pop} messages from
-input channels. \textbf{send} and \textbf{pop} are blocking calls, as mentioned
-earlier. Procedures can \textbf{call} other procedures but they can not be
-recursive.
+\begin{figure}
+\begin{tabularx}{\linewidth}{|l|X|}
+\hline
+\send{} $m$ & Send a message $m$ to the node's output channel. This action
+blocks till the channel has space to send.\\
+\receive{} $m$ & Receive and remove a message $m$ from the node's input channel.
+This action blocks till a message is received.\\
+$x \gets value$ & Local state $x$ of the node gets updates to $value$.\\
+\call{} $f(args)$ & Execute \textbf{procedure} $f$ with $args$ arguments.\\
+\start{} $f(args)$ & Start a new thread that executes \textbf{procedure} $f$
+with $args$ arguments. This action is performed only by the scheduler.\\
+\resume{} $t$ & Resume thread $t$ from where it stopped execution. This action
+is performed only by the scheduler\\
+\hline
+\end{tabularx}
+\caption{Actions used for describing a node's behavior}
+\label{actions}
+\end{figure}
+
+When a new message arrives in an input channel that requires creation of a
+thread, the scheduler first removes the message from the channel and then
+creates a thread that starts executing the \procedure{} specific to the type of
+the incoming message. Figure \ref{actions} show the notations we use to
+describe the actions performed by an executing thread. Procedures can \call{}
+other procedures but they can not be recursive. We usually denote the execution
+of actions \receive{} and \send{} using their verb forms, for instance,
+,``receiving'' or ``sending''. We also denote the thread created by the
+scheduler to handle an incoming message as the thread ``handling'' that
+message.

### 2012-11-18 paper/isca40format/stuff/Variations.tex
stuff

\section{Implementing variations of basic MSI protocol}
\label{sec:variations}

%\renewcommand{\fourAngle}[6]{\text{$#1\langle#2\rightarrow#3,#4,#5\rightarrow#6\rangle$}}

\newcommand{\FwdReq}[3]{\text{$FwdReq\langle#1,#2\rightarrow#3\rangle$}}
\newcommand{\FwdAck}{\text{$FwdAck$}}

%\renewcommand{\Req}[5]{\fourAngle{Req}{#1}{#2}{#3}{#4}{#5}}
%\renewcommand{\Resp}[5]{\fourAngle{Resp}{#1}{#2}{#3}{#4}{#5}}

In this section we describe variations of the basic MSI hierarchical
message-passing directory-based protocol and show how our framework can be used
to design and argue their correctness easily.
%The
%arguments are based on the formally proven Invariants of Figure
%\ref{invariants}. The ordering requirements in Figure \ref{ordering} are
%handled at the protocol level by sending the state at which the sender was in
%just before sending the message as discussed in Section \ref{sec:nofifoorder}

\subsection{MESI optimization}
We first describe the MESI variation of the message-passing MSI protocol for a
two-level hierarchy containing multiple L1 caches and directory shared by all
the L1 caches. On receiving a request from a cache $c$ to upgrade $c$'s state
for address $a$ to $S$, the directory sends permission to upgrade the state of
$c$ for address $a$ to ``exclusive'' ($E$), if the directory state of every
other cache for that line is $I$ (and because of our Invariant
\ref{conservative}, the coherence state of every other cache for that line is
also $I$). Once cache $c$ has upgraded to $E$, it can automatically upgrade to
$M$ on receiving a store request from the processor, without sending any
notification to the directory.  However, if some other cache $c'$ sends a
request to the directory for address $a$ to upgrade to state $S$, the directory
has to ensure that the cache $c$ has downgraded its state from $E$ to $S$. If
$c$ remains at $E$ without having upgraded to $M$, then it doesn't have to send
the data back to the memory, while downgrading to $S$.

From the directory's perspective, states $M$ and $E$ are equivalent -- the only
distinction is that the cache would not send back the data while downgrading the
state for an address from $E$. Thus MESI is simply an optimization over the MSI
protocol in which the cache maintains an extra dirty bit for each address, and
sends back the cache line data to the directory during a downgrade only if the
dirty bit is set (the dirty bit can not be set for an address if the
corresponding state is $S$). If the dirty bit is not set during a downgrade, the
cache sends back a message denoting ``data unmodified'' in the place of actual
data. In the hierarchical setting too, MESI is simply an optimization over MSI
in which a child sends data back to its parent during a downgrade only if the
dirty bit is set.

\subsection{Cache-intervention based protocols}
In the protocols discussed so far, caches can directly communicate only with
their immediate children or with its immediate parent. In cache-intervention
based protocols, communication and data transfer can also take place between
nodes other than the immediate children or the immediate parent. Our framework
can be easily extended to include cache-intervention based protocols where
sibling caches, \ie caches that share the same immediate parent, can communicate
with each other. We show 2 protocols with sibling cache-intervention: a) the MSI
protocol with intervention by a cache in the $M$ state and b) the MOSI/MOESI
protocol. Note that in both the protocols, the MESI optimization can be
implemented whenever a child has to transfer data to its parent -- the MOSI
protocol with MESI optimization is popularly called the MOESI protocol.

We first describe the two protocols using a two-level hierarchy containing
multiple L1 caches and a directory shared by all the L1 caches, and later
describe the extension to a hierarchical system.

\subsubsection{Intervention by a cache in the $M$ state:}
\label{sec:m-only}

In this protocol, a downgrade request sent from the directory to a cache $c$ for
address $a$ can sometimes be augmented with a forwarding request
\FwdReq{fwdDest}{fromState}{toState}, which directs the cache to send an upgrade
response to $fwdDest$ to upgrade $fwdDest$'s state for address $a$ from
$fromState$ to $toState$, on the directory's behalf. A downgrade response from
cache $c$ can also be augmented with \FwdAck{}, which acknowledges the directory
that the requested forwarding has been fulfilled.  Note that normal downgrade
requests from the directory without a forwarding request and normal responses
from the cache without \FwdAck{} are also sent.

Let us walk through the behavior of the system for handling a request from cache
$c'$ for address $a$ to upgrade to state $x$. Here $x = M$ or $x = S$ and let us
denote $toCompatible(x)$ by $y$. Checking if $dir[c][a] > y$ or if $state[c][a]
> y$ is akin to checking if $dir[c][a] = M$ or $state[c][a] = M$ in this
protocol since $y \le S$; using this (general) notation becomes convenient
later.

When a directory receives a request from cache $c'$ for address $a$ to upgrade
to state $x$, it determines if $\exists c, dir[c][a] > y$. If $\forall c,
dir[c][a] \le y$, then the directory behaves in the same manner as MSI -- the
directory responds to $c'$ authorizing it to upgrade to state $x$ for address
$a$, and sends data from memory if necessary. But if $\exists c, dir[c][a] > y$,
then it sends a request to cache $c$ to downgrade its state for address $a$ to
$toCompatible(x)$ augmented with a forwarding request \FwdReq{c'}{w}{x}, where
$w = dir[c'][a]$. This forwarding request is essentially a response sent by the
directory to cache $c'$ to upgrade the state of $c'$ for address $a$ to $x$, but
it is sent via the cache $c$. The directory waits for a downgrade response along
with a \FwdAck{} from $c$ before processing further requests for $a$. On
receiving the response from $c$, the directory would change $dir[c'][a]$ to $x$.

If cache $c$ receives the downgrade request from the directory along with the
forward request \FwdReq{c'}{w}{x}, and if $state[c][a] > y$ $c$ will send a
response to $c'$ authorizing the upgrade of $state[c'][a]$ to $x$, along with
the data associated with address $a$, if $w = I$. It will also send a downgrade
response to the directory notifying that $c$ has downgraded state $a$ to
$toCompatible(x)$ along with a \FwdAck{}; and it sends the data to the memory if
$x = S$.

When $x > S$, then the data in memory is going to be stale anyway (since
$c'$ will eventually have the most up-to-date version), so $c$ does not have to
transfer the data to memory in this case. However, when $x = S$, then $c$ also
downgrades state of address $a$ to $S$ and hence the data in the memory has to
be kept up-to-date, in order to guarantee Read-from-last-writer Invariant
\ref{readLast} for subsequent cache requests.

However, on receiving the downgrade request from the directory, if cache
$state[c][a] \le y$ for address $a$ (because of eviction due to replacement),
then the cache simply drops the request just like in the MSI protocol discussed
earlier. When $c$ downgraded its state earlier for address $a$, just like in the
MSI protocol, it must have sent the downgrade response and the data. The data is
always transferred in this case (subject to MESI optimization) because
the state $state[c][a]$ right before sending the voluntary downgrade
response must have been $> S$ which means $dataToUpper(state[c][a])$ is True,
and so in the MSI protocol, it would have transferred data. But this
voluntary response would not have a \FwdAck{}.

If the directory receives a downgrade response for address $a$ from cache $c$
along with a \FwdAck{}, the directory knows that $c$ has sent a response to
cache $c'$ authorizing $c'$ to upgrade its state for $a$ to $x$. If $x = S$, the
memory is updated with the received data (subject to MESI optimization). It will
finally remove the outstanding request from $c'$ for address $a$ as the request
has been handled.

On the other hand, if the directory receives a downgrade response for address
$a$ from cache $c$ without a \FwdAck{}, the directory knows that $c$ has not
sent a response to cache $c'$. The memory is updated with the received data
(subject to MESI optimization) for address $a$. This case is handled just like
in the MSI protocol, as if the forwarding never happened. It sends a
response to $c'$ authorizing it to upgrade its state for address $a$ to $x$. It
also sends the (just updated) data from memory, if $dir[c'][a]$ was $I$ at the
time of receiving the upgrade to $x$ request from $c'$.

$c'$ would eventually get the upgrade to $x$ response for address $a$ either
from the directory or the cache $c$ -- it can not distinguish between the two
cases.

Figure \ref{msi-inter} shows the order of messages transferred between the
caches in the scenario when forwarding happens as well the in the scenario when
no forwarding happens, for the $M$-only cache-intervention based protocol.

%:------------------- If the
%:Single-writer Invariant \ref{singleWriter} is maintained (and we prove
%:informally below that it is indeed the case), then $dir[c][a] = M$ and
%:$dir[c'][a] = I$ at the time of reception of the downgrade message from $c$.

It can easily be seen that the protocol given above satisfies all the
constraints \ref{} to \ref{}, except, potentially, Constraint \ref{orderreq}.
Before the upgrade response for address $a$ sent from $c$ to $c'$ has reached
$c'$, $c'$ could have received a downgrade request sent from the directory. This
happens as follows (Figure \ref{msi-fifo-needed}, with the numbers indicating
the order in which the messages are sent, for outgoing edges, and received for
incoming edges). Cache $c$ sends the upgrade response (and data) to cache $c'$,
and its own downgrade response along with the \FwdAck{} to the directory. The
directory receives the downgrade response from $c$. Soon afterwards, the
directory wants to downgrade the state of $a$ in $c'$ to $I$ (let's say, because
some other cache wants to upgrade its state for $a$ to $M$).  The directory
sends a downgrade to $I$ request to $c'$ for address $a$. This downgrade request
to $c'$ from the directory reaches $c'$ before the (forwarded) upgrade response
from $c$, violating Constraint \ref{orderreq}. However, as we saw in Section
\ref{sec:nofifoorder}, Constraint \ref{orderreq} can be satisfied by tagging the
downgrade request from the directory with the state the directory was in for
cache $c'$ and address $a$ right before the request to $c'$ was sent -- FIFO
ordering is not required from the network. $c'$ will not process the downgrade
request till it has received the upgrade response from $c$.

Since the protocol satisfies all the constraints \ref{} to \ref{}, it satisfies
all the invariants in Figure \ref{invariants}. Thus the directory is guaranteed
to have a conservative version of the states of each of the caches for each
address by Invariant \ref{conservative}. When the directory sends a downgrade
request for address $a$ to $c$ tagged with a \FwdReq{c'}{x}, it essentially
sends the upgrade response to cache $c'$ for address $a$ simultaneously. This
seemingly violates the compatibility of states for address $a$. For example, if
$c'$ has requested for an upgrade to state $M$ for address $a$, and $c$ is
already in state $M$ for address $a$, if the directory sends an upgrade response
to cache $c'$ without receiving a downgrade response from state $c$, then
according to Constraint \ref{psendeffect}, according to the directory's notion
of caches' states, both $c$ and $c'$ are in state $M$ for address $a$. But the
upgrade response to $c'$ goes via cache $c$. So the state of address $a$ in $c'$
would not have been upgraded to $M$ till $c$ has downgraded its state to $I$ and
the data transferred to $c'$ would be the most up-to-date data, as modified by
$c$. This ensures both the Single-writer Invariant \ref{singleWriter} and the
Read-from-last-writer Invariant \ref{readLast}.

Thus we have mapped the complex protocol with intervention by a cache in $M$
state to a MSI protocol. We have provided an informal sketch of the
proof using our framework of constraints \ref{} to \ref{} and Invariants (Figure
\ref{invariants}); the formal proof is beyond the scope of this paper.

\subsubsection{MOSI/MOESI protocol:}
In the MOSI protocol, if cache $c$ is in $O$ state for an address $a$, informally
it signifies that $c$ is the ``owner'' of the data and hence has to supply the
data for any request from other caches. The data in a cache in state $O$ for an
address can be more up-to-date than the memory, and other caches can have copies
of this data.  In our framework, we extend the $toCompatible$ and $dataToUpper$
mappings, and the $<$ relation to accommodate the $O$ state as shown in Figure
\ref{fig:o-extend}.

\begin{figure}
\begin{subfigure}{.25\linewidth}
\centering
\begin{tabular}{|ccc|}
\hline
$M$ & $\rightarrow$ & $I$\\
$O$ & $\rightarrow$ & $S$\\
$S$ & $\rightarrow$ & $O$\\
$I$ & $\rightarrow$ & $I$\\
\hline
\end{tabular}
\subcaption*{$toCompatible$}
\end{subfigure}
\begin{subfigure}{.25\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $\checkmark$\\
$O$ & $\checkmark$\\
$S$ & $\times$\\
$I$ & $\times$\\
\hline
\end{tabular}
\subcaption*{$dataToUpper$}
\end{subfigure}
\begin{subfigure}{.48\linewidth}
\centering
\begin{tabular}{|c|cccc|}
\hline
& $M$ & $O$ & $S$ & $I$\\
\hline
$M$ & $=$ & $>$ & $>$ & $>$\\
$O$ & $<$ & $=$ & $>$ & $>$\\
$S$ & $<$ & $<$ & $=$ & $>$\\
$I$ & $<$ & $<$ & $<$ & $=$\\
\hline
\end{tabular}
\subcaption*{$<$ (and other) relations}
\end{subfigure}
\caption{MOSI protocol mappings and $<$ relations}
\label{fig:o-extend}
\end{figure}

A cache always requests an upgrade to $M$ or $O$ state. It never requests an
upgrade to an $S$ state, even if the cache wants permissions just to read the
location of the memory. So, in our protocol, the ``owner'' state for an address
keeps getting transferred to the last cache that requests read permissions for
the address. This makes the protocol fit into our framework easily.

The directory's and the cache's behavior on receiving a request is exactly the
same as described in the $M$-only cache-intervention based protocol (Section
\ref{sec:m-only}), if $toCompatible$ and $dataToUpper$ functions and the $<$
relations are substituted appropriately. The argument for its correctness is
also exactly the same.

\subsubsection{Salient features of our cache-intervention based protocols:}
%There are two interesting points worth noting in our protocols:

\noindent \paragraph{Acknowledgement that forwarding has taken place is sent
earlier:} \FwdAck{} is sent from the forwarding cache $c$ rather than the final
cache $c'$ which is the case in commonly used protocols \cite{hammer, gems}.
This unblocks the directory to serve further requests for the same address as
soon as \FwdAck{} is received from $c$, instead of waiting for an
acknowledgement from $c'$.

\noindent \paragraph{Writebacks require no acks:} When a cache $c$ is in the $M$
or $O$ state for address $a$, then the address can be evicted from the cache
anytime (for replacement) without having to wait for the directory to
acknowledge this eviction (which is the case in commonly used protocols
\cite{hammer, gems}. Any downgrade response can be sent anytime without
requesting the directory for authorization. This frees up the cache line to
store a different address, without having to wait for authorization from the
directory.

It is interesting to compare the performance of our approach in which a cache
line in $M$ or $O$ state is written back, and the data evicted, against the
commonly used approaches, in which the cache line stays in the cache till the
directory acknowledges the writeback. The difference in performance arises only
in the following scenario: Cache $c$ has an address $a$ in $M$ state, and
another cache $c'$ sends an upgrade from $I$ to $M$ request for address $a$ to
the directory, and the directory has sent a downgrade request along with
\FwdReq{c'}{I}{M} to $c$. By the time $c$ received the downgrade request, it has
already evicted address $a$.

If cache $c$ has evicted address $a$ and sent the downgrade response to the
directory just before it received the downgrade+forward request from the
directory, then that downgrade response has to first reach the memory (takes
almost one ``leg'') and then has to be transferred from the memory to $c'$
(takes another ``leg''). So it takes almost four legs from the time $c'$ sent
the request to the time $c'$ received the response. This is shown in Figure
\ref{fig:4legs}.
%The times at which messages are sent are marked in the outgoing
%edges and the times at which messages are received are marked in the incoming
%edges.

On the other hand, if the directory receives the downgrade response from $c$
right after sending the downgrade+forward request to $c$, then it takes only
slightly more than 2 legs from the time $c'$ sent the request to the time $c'$
received the response. This scenario is depicted in Figure \ref{fig:2legs}

In the other protocols, in both the scenarios $c'$ would receive the response in
exactly 3 legs, as the response would be forwarded by $c$ in both the scenarios.

\subsubsection{Extending the protocol for multiple levels of hierarchy:}
Extending this protocol to multiple levels of hierarchy follows the same
procedure as extending the MSI protocol to multiple hierarchical levels.
Just like in the multi-level MSI protocol, each node in the hierarchy has a
\emph{directoy} $dir[c][a]$, which maintains a conservative version of the
state of each address $a$ in each of its immediate child $c$, and a
\emph{coherence state} $state[a]$ for each address $a$ which represents the
state with respect to the node's parent. If $state[a] = M \text{ or } O$ and
$\forall c, dir[c][a] \le S$, then the directory has the most up-to-date
version of the data, but if $\exists c, dir[c][a] > S$, then the subtree of $c$
has the most up-to-date data for address $a$. This is  the same interpretation
as for $state[a] = M$ in the multi-level MSI protocol.

Note that in the case of multiple levels of hierarchy, only siblings which share
the same immediate parent can communicate with each other during cache
intervention. Thus, the overall behavior for a node to handle incoming downgrade
requests from its parent, and sending downgrade responses to its parent remains
the same -- the only change is the way a node interacts with its own children.

We believe it's hard to allow any node to communicate with any
other node for cache intervention; it certainly can not be handled within our
framework.

### 2012-11-19 paper/isca40format/Introduction.tex
stuff

@@
 
-Large shared-memory multicores are likely to use cache-coherence protocols
-which are directory-based, hierarchical, and implemented using message passing.
-Even though such protocols have been around for two decades (see, for example,
-DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
-such protocols remains a challenge. It is difficult to convince oneself that
-all corner cases have been handled adequately without a formal proof. And yet
-formal proofs of real implementations are practically non-existent. One can't
-settle for conservative choices in cache-coherence protocol designs because
-performance (\eg the number of messages needed to maintain coherence, the
-buffer requirements, etc) is as important as correctness. Consequently proving
-a simple protocol for a small number of caches, under unbounded buffer
-assumption is of limited use in gaining confidence in the correctness of a real
-protocol.
+Large shared-memory multicores are likely to use cache-coherence protocols which are directory-based, hierarchical, and implemented using message passing.
+Even though such protocols have been around for two decades (see, for example, DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of such protocols remains a challenge.
+It is difficult to convince oneself that all corner cases have been handled adequately without a formal proof.
+And yet formal proofs of real implementations are practically non-existent.
+One can't settle for conservative choices in cache-coherence protocol designs because performance (\eg the number of messages needed to maintain coherence, the buffer requirements, etc) is as important as correctness.
+Consequently proving a simple protocol for a small number of caches, under unbounded buffer assumption is of limited use in gaining confidence in the correctness of a real protocol.
 
-In this paper we will consider systems where caches form a tree hierarchy and
-caches are \emph{inclusive}, that is, if a cache contains address $a$ then its
-parent is guaranteed to contain address $a$, though the parent may have stale
-data for $a$. Every cache or node in the system has an associated directory
-which contains information about its children caches. The coherence state
-changes are orchestrated by passing requests and responses between the caches
-via the directory.  Unlike bus-based protocols which lock the bus till all the
-required state transitions are done thereby ensuring atomicity of state
-transitions, in a message-passing protocol, requests and responses are split
-leading to transient states. This creates scenarios not present in bus-based
-protocols, which can lead to coherence violations or deadlocks, if the protocol
-is not designed or implemented carefully.
+In this paper we will consider systems where caches form a tree hierarchy and caches are \emph{inclusive}, that is, if a cache contains address $a$ then its parent is guaranteed to contain address $a$, though the parent may have stale data for $a$.
+Every cache or node in the system has an associated directory which contains information about its children caches.
+The coherence state changes are orchestrated by passing requests and responses between the caches via the directory.
+Unlike bus-based protocols which lock the bus till all the required state transitions are done thereby ensuring atomicity of state transitions, in a message-passing protocol, requests and responses are split leading to transient states.
+This creates scenarios not present in bus-based protocols, which can lead to coherence violations or deadlocks, if the protocol is not designed or implemented carefully.
 
-Several factors contribute to the difficulty of designing protocols for such
-distributed, hierarchical and message-passing systems:
+Several factors contribute to the difficulty of designing protocols for such distributed, hierarchical and message-passing systems:
 
-\begin{enumerate}
-\item Lots of concurrent events must be dealt with without a central authority
-to impose a sequence on events;
-\item The state on which transitions are based is physically distributed in
-space and cannot be accessed atomically;
-\item Physical network may block the transmission of a message because of lack
-of communication buffers or head-of-the-line blocking; 
-\item The cache or directory may reach a state which can not handle a
-particular message. An incomplete set of transition rules may cause the
-protocol to deadlock.
+\begin{enumerate} \item Lots of concurrent events must be dealt with without a central authority to impose a sequence on events; \item The state on which transitions are based is physically distributed in space and cannot be accessed atomically; \item Physical network may block the transmission of a message because of lack of communication buffers or head-of-the-line blocking; \item The cache or directory may reach a state which can not handle a particular message.
+An incomplete set of transition rules may cause the protocol to deadlock.
 \end{enumerate}
 
-As examples of problems in designing correct protocols, we describe several
-concrete and familiar scenarios from two-level systems (a private L1 cache for
-each processor and a memory with a directory):
+As examples of problems in designing correct protocols, we describe several concrete and familiar scenarios from two-level systems (a private L1 cache for each processor and a memory with a directory).
+States $M, S$ and $I$ have their usual meaning in MSI protocol ($M$ represents read and write permissions for an address, $S$, read-only permissions and $I$, no permission)
 
-\noindent \emph{Example 1:} A cache in state $S$ for a line gets a store request
-from the processor and sends a request for changing the line's state to $M$ to
-the directory controller. The controller may eventually send a response
-permitting the cache to go to $M$ state for that line. The response would
-signify that the states in the other caches for that line \emph{were} $I$. But
-the messages take time to propagate and the response is necessarily based on the
-old states of other caches -- the current state of another cache may not be $I$,
-violating the Single-writer-at-a-time invariant and hence breaking cache
-coherence.
+\noindent \emph{Example 1:} A cache in state $S$ for a line gets a store request from the processor and sends a request for changing the line's state to $M$ to the directory controller.
+The controller may eventually send a response permitting the cache to go to $M$ state for that line.
+The response would signify that the states in the other caches for that line \emph{were} $I$.
+But the messages take time to propagate and the response is necessarily based on the old states of other caches -- the current state of another cache may not be $I$, violating the Single-writer-at-a-time invariant and hence breaking cache coherence.
 
-\noindent \emph{Example 2:} A cache in state $S$ for a line gets a store request
-from the processor and sends a request for changing the line's state to $M$ to
-the directory controller. Instead of getting a response permitting it to go to
-$M$ state, the cache may receive a request from the directory controller asking
-this cache to go to $I$ state (invalidate request)-- this could have happened if
-the directory is serving some other cache's request to go to $M$ state for the
-same line. It is not easy to determine the correct behavior of the former cache
-on receiving the invalidate request -- whether it should ignore the invalidate
-request and wait for a response to its request or whether it should invalidate
-the line. Would ignoring the invalidate request lead to a deadlock?
+\noindent \emph{Example 2:} A cache in state $S$ for a line gets a store request from the processor and sends a request for changing the line's state to $M$ to the directory controller.
+Instead of getting a response permitting it to go to $M$ state, the cache may receive a request from the directory controller asking this cache to go to $I$ state (invalidate request)-- this could have happened if the directory is serving some other cache's request to go to $M$ state for the same line.
+It is not easy to determine the correct behavior of the former cache on receiving the invalidate request -- whether it should ignore the invalidate request and wait for a response to its request or whether it should invalidate the line.
+Would ignoring the invalidate request lead to a deadlock?
 
-\noindent \emph{Example 3:} A cache receives an invalidate request from the
-directory controller for a line that is not present in the cache. Is such a
-request even possible? The directory does not have access to the up-to-date
-state of the cache, and so can potentially make this request. How should a
-cache handle this request; should it drop such a request, or respond saying
-that it does not have the line?
+\noindent \emph{Example 3:} A cache receives an invalidate request from the directory controller for a line that is not present in the cache.
+Is such a request even possible?
+The directory does not have access to the up-to-date state of the cache, and so can potentially make this request.
+How should a cache handle this request; should it drop such a request, or respond saying that it does not have the line?
 
-These examples give a glimpse of the complexity of designing and implementing a
-correct cache coherence protocol in a directory-based non-bus setting.  Even for
-a 2-level cache hierarchy, designing and implementing the protocol is quite
-difficult. This problem is greatly exacerbated in a multi-level cache hierarchy,
-which is becoming the norm because of the increasing core counts.
+These examples give a glimpse of the complexity of designing and implementing a correct cache coherence protocol in a directory-based non-bus setting.
+Even for a 2-level cache hierarchy, designing and implementing the protocol is quite difficult.
+This problem is greatly exacerbated in a multi-level cache hierarchy, which is becoming the norm because of the increasing core counts.
 
-In this paper, we present a substantially easier method to design
-directory-based cache-coherence protocols for such systems. The method breaks
-down the design problem in two parts: a part that is shared by a number of
-concrete protocols which we call \glob{}s, and another part that is specific to
-the protocol to be designed and implemented, called \policy{}. \glob{}s provide
-guarantees such as every request will eventually get a response, the
-directory's notion of the cache states is ``conservative'' \ie a cache can not
-have more permissions for a line than what the directory thinks it has, \etc;
-\glob{} enables \policy{} to focus on the high level properties such as the
-set of actions a cache or a directory should perform to maintain the
-Single-writer-at-a-time and the Read-from-last-writer invariants.
+In this paper, we present a substantially easier method to design directory-based cache-coherence protocols for such systems.
+The method breaks down the design problem in two parts: a part that is shared by a number of concrete protocols which we call \glob{}s, and another part that is specific to the protocol to be designed and implemented, called \policy{}.
+\glob{}s provide guarantees such as every request will eventually get a response, the directory's notion of the cache states is ``conservative'' \ie a cache can not have more permissions for a line than what the directory thinks it has, \etc; \glob{} enables \policy{} to focus on the high level properties such as the set of actions a cache or a directory should perform to maintain the Single-writer-at-a-time and the Read-from-last-writer invariants.
 
-The \glob{}s can in turn be guaranteed using a set of (\policy{} independent)
-\local{}s which dictate local properties of a node such as when a request can be
-handled, what state changes happen on sending a response, properties about the network,
-\etc. We have formally proved that \local{}s ensure \glob{}s using the
-Coq Theorem Prover \cite{}, however, a presentation of the formal proof is
-beyond the scope of this paper. The \glob{}s and the \local{}s form the
-cornerstone of our framework. The separation of \glob{}s and \local{}s is akin
-to having an abstraction layer which hides low level details.
+The \glob{}s can in turn be guaranteed using a set of (\policy{} independent) \local{}s which dictate local properties of a node such as when a request can be handled, what state changes happen on sending a response, properties about the network, \etc.
+We have formally proved that \local{}s ensure \glob{}s using the Coq Theorem Prover \cite{}, however, a presentation of the formal proof is beyond the scope of this paper.
+The \glob{}s and the \local{}s form the cornerstone of our framework.
+The separation of \glob{}s and \local{}s is akin to having an abstraction layer which hides low level details.
 
-The \glob{}s of our framework can be used to design a wide range of cache
-coherence protocols, notably MSI, MESI, MOSI, MOESI, as well as their
-hierarchical version. We will show how these are designed as \policy{}s,
-focusing on high-level details.
+The \glob{}s of our framework can be used to design a wide range of cache coherence protocols, notably MSI, MESI, MOSI, MOESI, as well as their hierarchical version.
+We will show how these are designed as \policy{}s, focusing on high-level details.
 
-Our method substantially reduces the design and proof burden for new protocols
-because of the common abstract protocol. The abstract protocol is a set of
-transitions that the caches and directories follow to send or receive a request
-or a response. We have formally proved that these transitions ensure the
-following properties: a) the directory's notion of a cache's state is
-``conservative'' in that a cache is never going to have more permissions that
-what the directory thinks the cache has; and b) any request will eventually get
-a corresponding response. A formal proof of the abstract protocol was written
-using the Coq theorem prover \cite{}, however, a presentation of the formal
-proof is beyond the scope of this paper.
+Our method substantially reduces the design and proof burden for new protocols because of the common abstract protocol.
+The abstract protocol is a set of transitions that the caches and directories follow to send or receive a request or a response.
+We have formally proved that these transitions ensure the following properties: a) the directory's notion of a cache's state is ``conservative'' in that a cache is never going to have more permissions that what the directory thinks the cache has; and b) any request will eventually get a corresponding response.
+A formal proof of the abstract protocol was written using the Coq theorem prover \cite{}, however, a presentation of the formal proof is beyond the scope of this paper.
 
-It should be noted that the flexibility of our methodology to design multiple
-cache coherence protocols comes from the fact that we do not assign any meaning
-to the value of a coherence state beyond upgrades or downgrades. Thus, how or
-why a particular cache state implies that a particular address may not be
-present in any other cache stems from protocol specific ``compatibility of
-state'' check.
+It should be noted that the flexibility of our methodology to design multiple cache coherence protocols comes from the fact that we do not assign any meaning to the value of a coherence state beyond upgrades or downgrades.
+Thus, how or why a particular cache state implies that a particular address may not be present in any other cache stems from protocol specific ``compatibility of state'' check.
 
-The main contributions of this paper are: 1. A new abstract protocol that
-satisfies two important invariants for distributed and hierarchal
-message-passing protocols; 2. Instantiation of the abstract protocol to
-implement four well-known protocols (MSI, MESI, MOSI, MOESI) and their
-variations and showing why their correctness is guaranteed as long as the
-abstract protocol is correct, \ie it provably implements the two invariants;
-and 3. Our method does not introduce any overhead over direct implementation of
-these protocols.  We think that this new method of designing protocols
-substantially reduces the required effort, and that our abstract protocol is
-flexible enough to implement (while ensuring correctness of) most if not all
-inclusive, message-passing, directory-based protocols.
+The main contributions of this paper are: 1.
+A new abstract protocol that satisfies two important invariants for distributed and hierarchal message-passing protocols; 2.
+Instantiation of the abstract protocol to implement four well-known protocols (MSI, MESI, MOSI, MOESI) and their variations and showing why their correctness is guaranteed as long as the abstract protocol is correct, \ie it provably implements the two invariants; and 3.
+Our method does not introduce any overhead over direct implementation of these protocols.
+We think that this new method of designing protocols substantially reduces the required effort, and that our abstract protocol is flexible enough to implement (while ensuring correctness of) most if not all inclusive, message-passing, directory-based protocols.
 
-\paragraph{Paper organization:} Section \ref{sec:abstract} describes the abstract
-protocol; Section \ref{sec:properties} gives the two properties that are
-guaranteed by the abstract protocol and provides some intuition for proving
-these properties. We also discuss the communication buffer requirements for the
-abstract protocol and show that the requirement is only proportional to the
-number of levels and not to the number of processors. In Section
-\ref{sec:protocols}, we show how the abstract protocol can be used to implement
-MSI, MESI, MOSI and MOESI protocols and their variants. In Section
-\ref{sec:performance}, we show that these formally verified family of protocols
-require no extra state bits, nor extra message transfers compared to other
-published implementations. In Section \ref{sec:related}, we present some of the
-related work and finally in Section \ref{sec:conclusion}, we present some
-directions for the future work.
+\paragraph{Paper organization:} Section \ref{sec:abstract} describes the abstract protocol; Section \ref{sec:properties} gives the two properties that are guaranteed by the abstract protocol and provides some intuition for proving these properties.
+We also discuss the communication buffer requirements for the abstract protocol and show that the requirement is only proportional to the number of levels and not to the number of processors.
+In Section \ref{sec:protocols}, we show how the abstract protocol can be used to implement MSI, MESI, MOSI and MOESI protocols and their variants.
+In Section \ref{sec:performance}, we show that these formally verified family of protocols require no extra state bits, nor extra message transfers compared to other published implementations.
+In Section \ref{sec:related}, we present some of the related work and finally in Section \ref{sec:conclusion}, we present some directions for the future work.

### 2012-11-19 paper/isca40format/Invariants.tex
stuff

--- 2012-11-09 paper/isca40format/Invariants.tex
@@
 \label{sec:invariants}
-
-\newcommand{\threeAngle}[4]{\text{$#1\langle#2,#3,#4\rangle$}}
-
-\newcommand{\Req}[3]{\threeAngle{Req}{#1}{#2}{#3}}
-\newcommand{\Resp}[3]{\threeAngle{Resp}{#1}{#2}{#3}}
-\newcommand{\Inv}{\text{$Inv$\;}}
-
-The values of the coherence state (and hence the directory state) belong to a
-totally ordered set, \ie a set in which a \emph{less than} ($<$) relation is
-defined between any two set elements and this relation is transitive. If an
-address is not present, then the coherence state corresponding to that address
-is the lowest value in the set (which we denote by \Inv. For instance, in an
-MSI protocol, the values of the coherence state are from the set $\{M, S, I\}$
-where $I < S$, $S < M$ and $I < M$; if an address is not present, the state
-corresponding to that address is $I$. Our abstract protocol does not assign any
-semantics or meaning to the values of the coherence states. In a concrete cache
-coherence protocol, the values of the states are associated with permissions
-for reading and writing, and the state with fewer permissions is $<$ the state
-with more permissions.
-
-\begin{figure}\centering
-\begin{tabularx}{\linewidth}{|l|X|}
-\hline
-$state[a]$ & Coherence state for address $a$\\
-\hline
-$dir[a][c]$ & Directory's version of the coherence state for address $a$ in child
-$c$\\
-\hline
-\end{tabularx}
-\caption{Information stored for each cache line}
-\label{table:storage}
-\end{figure}
-
-\begin{figure}\centering
-\begin{tabularx}{\linewidth}{|l|X|}
-\hline
-$isPresent(a)$ & Is line for address $a$ present in this node?\\
-\hline
-$isWait(n, a)$ & Is there a pending response from node $n$ (either parent $P$
-or a child) for address $a$?\\
-\hline
-$isReplacing(a)$ & Is address $a$ currently being evicted?\\
-\hline
-$isEnterPending(a)$ & Is there any (child) request for address $a$ waiting to get
-a cache line entry in the cache?\\
-\hline
-$isHandling(n, a)$ & Is there any request for address $a$ from node $n$ (either
-parent $P$ or a child) that has started to be handled?\\
-\hline
-%$replacingAddr(a)$ & The address that is going to replace the line
-%corresponding to address $a$\\
-%\hline
-%$cWaitState(c, a)$ & The state to which the node has requested its child $c$ to
-%downgrade to and hence waiting for a response\\
-%\hline
-\end{tabularx}
-\caption{Information that can be computed for each address (typically from the
-MSHR)}
-\label{table:functions}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|l|l|X|}
-\hline
-\multicolumn{3}{|c|}{\Req{n}{a}{x}}\\
-\hline
-\multirow{4}{*}{$n = P$} & Incoming & Request from parent for changing my
-                                      $state[a]$ to $x$\\
-\cline{2-3}
-                         & Outgoing & Request to parent for permission to change
-                                      my $state[a]$ to $x$\\
-\hline
-\multirow{4}{*}{$n \neq P$} & Incoming & Request from child $n$ for permission
-                                         to change $n$'s $state[a]$ to $x$\\
-\cline{2-3}
-                            & Outgoing & Request to child $n$ to change $n$'s
-                                         $state[a]$ to $x$\\
-\hline
-\multicolumn{3}{|c|}{\Resp{n}{a}{x}}\\
-\hline
-\multirow{4}{*}{$n = P$} & Incoming & Response from parent granting permission
-                                      to change my $state[a]$ to $x$\\
-\cline{2-3}
-                         & Outgoing & Response to parent indicating that my
-                                      $state[a]$ has changed to $x$\\
-\hline
-\multirow{4}{*}{$n \neq P$} & Incoming & Response from child indicating that
-                                         $n$'s $state[a]$ has changed to $x$\\
-\cline{2-3}
-                            & Outgoing & Response to child $n$ granting permission
-                                         to change $n$'s $state[a]$ to $x$\\
-\hline
-\end{tabularx}
-\caption{Incoming and outgoing messages}
-\label{table:messages}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-A response for address $a$ can be sent only if the line for $a$ is present in
-the current node\\
-\hline
-\multicolumn{1}{|c|}{Response to child $c$ for address $a$}\\
-\hline
-\begin{enumerate}
-\item A response to child $c$ for address $a$ can be sent only if a
-request from child $c$ for $a$ is being handled
-\item Response \Resp{c}{a}{x} can be sent only if $x > dir[a][c]$, $\forall c',
-c' \neq c \rightarrow isCompatible(x, dir[a][c'])$ and  $x \le state[a]$
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{Response to parent $P$ for address $a$}\\
-\hline
-\begin{enumerate}
-\item If there is a pending response from the parent for address $a$, a
-response to the parent for $a$ can be sent only if a request from
-the parent for $a$ is being handled
-\item If there is no pending response from the parent for address $a$, a
-response to the parent for $a$ can be sent any time
-\item Response \Resp{P}{a}{x} can be sent only if $x < state[a]$ and $\forall
-c, x \ge dir[a][c]$
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Sending a response: Preconditions}
-\label{sendRespPre}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-A response for address $a$ can be sent only if the line for $a$ is present in
-the current node\\
-\hline
-\multicolumn{1}{|c|}{Response to child $c$ for address $a$}\\
-\hline
-\begin{enumerate}
-\item[] On sending \Resp{c}{a}{x}, $dir[a][c]$ changes to $x$
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{Response to parent $P$ for address $a$}\\
-\hline
-\begin{enumerate}
-\item[] On sending \Resp{P}{a}{x}, $state[a]$ changes to $x$
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Sending a response: Effects}
-\label{sendRespEff}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\multicolumn{1}{|c|}{On receiving a response from child $c$ for address $a$}\\
-\hline
-\begin{enumerate}
-\item $dir[a][c]$ changes to $x$
-\item Response \Resp{c}{a}{x} is removed
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{On receiving a response from parent $c$ for address $a$}\\
-\hline
-\begin{enumerate}
-\item $state[a]$ changes to $x$
-\item Response \Resp{P}{a}{x} is removed
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Receiving a response: Effects}
-\label{recvResp}
-\end{figure}
 
@@
 
+The caches are assumed to be organized in the form of a tree
+(Figure~\ref{fig:hierarchy}). The leaves of this tree represent the L1 caches,
+with each L1 cache connected to exactly one core. The root of the tree is a
+single Last-level cache (LLC), connected to the main memory. A node can
+communicate only with its parents or with its children via a request or
+response message whose format is in Figure~\ref{message}.
+
 \begin{figure}
+\begin{inv}
+An incoming request message from a child should not block an incoming request
+message from the parent, even for a different address\label{cReqNoBlockPReq}
+\end{inv}
+\begin{inv}
+An incoming request messages for an address $a$ should not block an incoming
+response message for address $a$, even from a different
+source.\label{reqNoBlockReq}
+\end{inv}
+\caption{Blocking restrictions}
+\label{blocking}
+\end{figure}
+
+\begin{figure}
+\begin{inv}
+An incoming response message for an address $a$ from a child $c$ should not be
+received before another incoming response message from the same $c$ and the
+same $a$ sent earlier has been received\label{cRespFifo}
+\end{inv}
+\begin{inv}
+An incoming request message for an address $a$ from a source $n$ should not be
+received before another incoming response message from the same $n$ and the
+same $a$ sent earlier has been received\label{reqNoOvertakeResp}
+\end{inv}
+\caption{Ordering requirements}
+\label{order}
+\end{figure}
+
+\begin{figure}
+\begin{inv}
+A response \Resp{p}{c}{a}{x}, where $p = c.parent$, can be sent from $p$ only
+if
 \begin{enumerate}
-\item $state[a]$ can change only on sending or receiving a response from the
-parent
-\item $dir[a][c]$ can change only on sending or receiving a response from child
-$c$
+\item a thread that handling a request from $c$ to $p$ for
+address $a$ is executing, and
+\item $x > p.dir[c][a]$
 \end{enumerate}
-\caption{State and directory changes: Preconditions}
+\label{cSendRespPre}
+\end{inv}
+\begin{inv}
+If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$
+\label{cSendRespPost}
+\end{inv}
+\begin{inv}
+A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
+if $x < c.state[a]$ \label{pSendRespPre1}
+\end{inv}
+\begin{inv}
+If there is a thread in node $c$ that is waiting to receive a response from its
+parent $p$ for an address $a$, then response \Resp{c}{p}{a}{x} can be sent only
+on executing a thread that is handling a request from $p$ to $c$ for address
+$a$. For instance, $a$ can not be evicted from $c$, if there is a thread
+handling a request from $p$ to $c$ for address $a$.
+\label{pSendRespPre2}
+\end{inv}
+\begin{inv}
+If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$
+\label{pSendRespPost}
+\end{inv}
+\caption{Invariants for sending responses}
+\label{sendResp}
+\end{figure}
+
+\begin{figure}
+\begin{inv}
+On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
+\gets x$
+\label{pRecvResp}
+\end{inv}
+\begin{inv}
+On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
+\gets x$
+\label{cRecvResp}
+\end{inv}
+\caption{Invariants for receiving responses}
+\label{recvResp}
+\end{figure}
+
+\begin{figure}
+\begin{inv}
+$c.state[a]$ can change only on $c$ sending a response to or receiving a
+response from $c.parent$
+\label{cState}
+\end{inv}
+\begin{inv}
+$p.dir[c][a]$ can change only on $p$ sending or receiving a response from
+its child $c$
+\label{pState}
+\end{inv}
+\caption{Invariants for state changes}
 \label{stateChange}
@@
 
-\floatstyle{plain}
-\restylefloat{figure}
-
 \begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\multicolumn{1}{|c|}{Request to child $c$ for address $a$}\\
-\hline
-\begin{enumerate}
-\item A request can be sent to child $c$ for address $a$ only if there is no
-pending response from child $c$ for $a$
-\item Request \Req{c}{a}{x} can be sent only if $x < dir[a][c]$
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{Request to parent $P$ for address $a$}\\
-\hline
-\begin{enumerate}
-\item A request can be sent to the parent for address $a$ only if
-  \begin{enumerate}
-  \item there is no pending response from the parent for address $a$, and
-  \item line in address $a$ is not being replaced
-  \end{enumerate}
-\item Request \Req{P}{a}{x} can be sent only if $x > state[a]$
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Sending a request: Preconditions}
+\begin{inv}
+\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$
+\label{pSendReqPre}
+\end{inv}
+\begin{inv}
+\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$
+\label{cSendReqPre}
+\end{inv}
+\caption{Invariants for sending a request}
 \label{sendReq}
@@
 \begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is being handled}\\
-\hline
-\begin{itemize}
-\item[] If $x \le state[a] \wedge \forall c', c' \neq c \rightarrow isCompatible(x, dir[a][c])$\newline
- then the node must send a response \Resp{c}{a}{x}
-to $c$
-\end{itemize}\\
-\hline
-\multicolumn{1}{|c|}{Address $a$ is marked as being replaced}\\
-\hline
+\begin{inv}
+$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
+if either
 \begin{enumerate}
-\item[] If $\forall c, dir[a][c] \le \Inv$, then the node must send a response
-\Resp{P}{a}{\Inv} to the parent
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is being
-handled}\\
-\hline
+\item address $a$ has been chosen for eviction by some thread and that thread
+has not yet replaced $a$, or
+\item a thread in $p$ is handling a request for address
+$a$ from any source
+\end{enumerate}
+\label{pHandleReq}
+\end{inv}
+\begin{inv}
+$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
+if either
 \begin{enumerate}
-\item[] If $\forall c, dir[a][c] \le x$, then the node must send a response
-\Resp{P}{a}{x} to the parent
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Forced responses: Effects}
-\label{respForReq}
+\item address $a$ has been chosen for eviction by some thread and that thread
+has not yet replaced $a$, or
+\item a thread in $p$ is handling a request for address
+$a$ from $p$, or
+\item a thread in $p$ is handling a request for address
+$a$ from one of $p$'s children $c'$, and that thread will not send any
+(downgrade) request to any child for address $a$ anymore
+\end{enumerate}
+\label{cHandleReq}
+\end{inv}
+\caption{Invariants for receiving a request (used by the scheduler)}
+\label{recvReq}
 \end{figure}
@@
 \begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is received}\\
-\hline
+\begin{inv}
+Address $a$ present in node $c$ can not be chosen for eviction by a thread
+executing in $c$ if either
 \begin{enumerate}
-\item If $a$ is present in the node, handling of request \Req{c}{a}{x} can be
-started only if
-  \begin{enumerate}
-  \item no request \Req{n}{a}{x} from any node $n$ for address $a$ is being
-  handled
-  \item address $a$ is not currently being replaced
+\item address $a$ has been chosen for eviction by some thread and that thread
+has not yet replaced $a$, or
+\item a thread in $c$ is handling a request for address
+$a$ from any source
 \end{enumerate}
-\item If $a$ is not present in the node, handling of request \Req{c}{a}{x} can be
-started only if there is a line with address $a'$ such that
-  \begin{enumerate}
-  \item no request \Req{n}{a'}{x} from any node $n$ for address $a'$ is being
-  handled
-  \item address $a'$ is not currently being evicted
-  \end{enumerate}
-\Req{n}{a'}{x} has started
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is received}\\
-\hline
-\begin{enumerate}
-\item[] Handling of request \Req{P}{a}{x} can be started only if there are no
-pending responses from any child for address $a$
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Start handling a request: Preconditions}
-\label{respForReqPre}
+\label{evict}
+\caption{Invariants for suspending a thread during choosing an address for eviction}
+\end{inv}
 \end{figure}
@@
 \begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is received}\\
-\hline
-\begin{enumerate}
-\item Start handling the request \Req{c}{a}{x}
-\item If $a$ is not present in the node, get a line with address $a'$ such that
-  \begin{enumerate}
-  \item no request \Req{n}{a'}{x} from any node $n$ for address $a'$ is being
-  handled
-  \item address $a'$ is not currently being evicted
-  \end{enumerate}
-Mark $a'$ as being evicted
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is received}\\
-\hline
-\begin{enumerate}
-\item[] Start handling the request \Req{P}{a}{x}
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Start handling a request: Effects}
-\label{respForReqEff}
+\begin{inv}
+Starvation-freedom: A thread that keeps becoming ready to execute will
+eventually be scheduled by the scheduler to execute.
+\end{inv}
 \end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\multicolumn{1}{|c|}{Request \Req{c}{a}{x} from child $c$ is being handled}\\
-\hline
-\begin{enumerate}
-\item[] If $x \le dir[a][c]$, remove the request
-\end{enumerate}\\
-\hline
-\multicolumn{1}{|c|}{Request \Req{P}{a}{x} from parent $P$ is being
-handled}\\
-\hline
-\begin{enumerate}
-\item[] If $x \ge state[a]$, remove the request
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Finish handling a request: Effects}
-\label{finishHandle}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-Request \Req{c}{a}{x} from child $c$ is being handled, and address $a'$ is
-replaced by $a$\\
-\hline
-\begin{enumerate}
-\item If $state[a'] <= \Inv$, replace address of the line for $a'$ as $a$,
-$state[a']$ as \Inv and $\forall c, dir(c, a)$ as \Inv
-\item Unmark $a'$ as being evicted
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Finish replacing line: Effects}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\begin{enumerate}
-\item[] If the state or dir of an address (either the requested address, or the
-evicted address in case of a replacement) for some node $n$ is not satisfactory,
-and there are no pending responses from $n$, send a request to make the state
-or dir satisfactory
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Handling of a request: Effects}
-\label{handling}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\begin{enumerate}
-\item[] The effects during sending a response or sending a request is atomic,
-\ie if there is no space available in the sending buffer, then the whole action
-is retried later
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Order of receiving messages}
-\label{atomicity}
-\end{figure}
-
-\begin{figure}
-\begin{tabularx}{\linewidth}{|X|}
-\hline
-\begin{enumerate}
-\item A request from a child should not block (\ie prevent start of the
-handling of) a request from the parent
-\item Requests for an address $a$ should not block (\ie prevent reception of)
-responses for address $a$, even from a different source
-\item A response for an address $a$ from a child $c$ should not be received
-before a response from the same $c$ for the same $a$ sent earlier has been
-received and removed
-\item The handling of a request for an address $a$ from a source $n$ should not
-be started before a response from the same $n$ for the same $a$ sent earlier
-has been received and removed
-\end{enumerate}\\
-\hline
-\end{tabularx}
-\caption{Order of receiving messages}
-\label{mesgOrder}
-\end{figure}

### 2012-11-19 paper/isca40format/Theorems.tex
stuff

\section{Theorems}
\label{sec:theorems}

If Invariants \ref{blocking} to \ref{evict} are obeyed, then we have formally
proved that the following theorems hold. The formal proof is beyond the scope of
the paper.

\begin{theorem}
It a thread has to send a message, it will eventually be able to send the
message.
\label{canSend}
\end{theorem}

\begin{theorem}
If a thread sends a request, it will eventually get a response.
\label{willRecv}
\end{theorem}

\begin{theorem}
If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
just before $p$ sends the response is the same as $c.state[a]$ just before $c$
receives the response.
\label{pcSame}
\end{theorem}

\begin{theorem}
If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
receives the response.
\label{cpSame}
\end{theorem}

\begin{theorem}
Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
p.dir[c][a] \ge c.state[a]$.
\label{conservative}
\end{theorem}

These theorems are needed to guarantee that Figure \ref{realistic} protocol is
indeed correct. If, for instance, a thread wants to send a message in the output
channel of the node and is not able to send one forever, then the thread can not
proceed further. Similarly, if a thread is waiting for a response (for a request
previously sent), and it never gets the response, the thread can never proceed
further. These scenarios create a deadlock.

We will illustrate the need for Theorems \ref{pcSame}, \ref{cpSame} and
\ref{conservative} using a 2-level of caches having two L1 caches $c_1$ and
$c_2$, and one shared L2 cache $p$.

Theorem \ref{pcSame} and \ref{cpSame} are important to make sure that the data
transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
$c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
sends \Req{c_1}{p}{a}{M} to $p$. $p$ receives the request, but assumes that
$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Theorem \ref{pcSame}
is violated. So, it does not transfer the data to $c_1$, resulting in $c_1$
waiting for data, and hence deadlock. A dual condition can be constructed
similarly, in which $c_1$ downgrades its state for address $a$ from $S$ to $I$,
and the parent assumes that $c_1$ is in state $M$, and keeps waiting for data.

Theorem \ref{conservative} is important because a parent node consults its
directory to decide the children it needs to send a downgrade request to.  Let
$p.dir[c_1][a] = S$, $c_1.state[a] = S$, $p.dir[c_2][a] = I$ but
$c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
\Req{c_1}{p}{a}{M} to $p$, $p$ will assume that $c_2$'s state for address $a$
is also $I$ and hence sends \Resp{p}{c_1}{a}{M} to $c_1$. But since the real
state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
\ref{singleWriter}.

If Theorems \ref{canSend}, to \ref{cpSame} holds, then the procedures given in
Figure \ref{realistic} can not deadlock because it will always be able to a)
send a request, b) send back a response, c) get back a response for a request it
sent, d) receive data when required. Procedure \uReq{} in Figure
\ref{msi-template} relies on Theorem \ref{conservative} to maintain the
Single-writer Invariant \ref{singleWriter}.

### 2012-11-19 paper/isca40format/Threads.tex
stuff

@@
 
-Earlier, each thread is started,, because of an incoming message, with a
-handler \textbf{procedure} particular to the type of the message. A procedure
-groups together actions which include state updates ($s \gets value$),
-\textbf{send} messages into output channels and \textbf{pop} messages from
-input channels. \textbf{send} and \textbf{pop} are blocking calls, as mentioned
-earlier. Procedures can \textbf{call} other procedures but they can not be
-recursive.
+\begin{figure}
+\begin{tabularx}{\linewidth}{|l|X|}
+\hline
+\send{} $m$ & Send a message $m$ to the node's output channel. This action
+blocks till the channel has space to send.\\
+\receive{} $m$ & Receive and remove a message $m$ from the node's input channel.
+This action blocks till a message is received.\\
+$x \gets value$ & Local state $x$ of the node gets updates to $value$.\\
+%\call{} $f(args)$ & Execute \textbf{procedure} $f$ with $args$ arguments.\\
+%\start{} $f(args)$ & Start a new thread that executes \textbf{procedure} $f$
+%with $args$ arguments. This action is performed only by the scheduler.\\
+\resume{} $t$ & Resume thread $t$ from where it stopped execution. This action
+is performed only by the scheduler\\
+\hline
+\end{tabularx}
+\caption{Actions used for describing a node's behavior}
+\label{actions}
+\end{figure}
+
+When a new message arrives in an input channel that requires creation of a
+thread, the scheduler first removes the message from the channel and then
+creates a thread that starts executing the \procedure{} specific to the type of
+the incoming message. Figure \ref{actions} show the notations we use to
+describe the actions performed by an executing thread. Procedures can call
+other procedures but they can not be recursive. We usually denote the execution
+of actions \receive{} and \send{} using their verb forms, for instance,
+,``receiving'' or ``sending''. We also denote the thread created by the
+scheduler to handle an incoming message as the thread ``handling'' that
+message.

### 2012-11-19 paper/isca40format/Invariants.tex
stuff

@@
 \restylefloat{figure}
-
-The caches are assumed to be organized in the form of a tree
-(Figure~\ref{fig:hierarchy}). The leaves of this tree represent the L1 caches,
-with each L1 cache connected to exactly one core. The root of the tree is a
-single Last-level cache (LLC), connected to the main memory. A node can
-communicate only with its parents or with its children via a request or
-response message whose format is in Figure~\ref{message}.
 

### 2012-11-19 paper/isca40format/Msi.tex
stuff

@@
 
-\newcommand{\printall}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
-\newcommand{\Req}[4]{\printall{Req}{#1}{#2}{#3}{#4}}
-\newcommand{\Resp}[4]{\printall{Resp}{#1}{#2}{#3}{#4}}
-\newcommand{\Data}[5]{\printall{Data}{#1}{#2}{#3}{#4}{#5}}
-
-\newcommand{\lett}{\textbf{let}}
-\newcommand{\send}{\textbf{send}}
-\newcommand{\receive}{\textbf{receive}}
-\newcommand{\pop}{\textbf{pop}}
-\newcommand{\remove}{\textbf{remove}}
-\newcommand{\assert}{\textbf{assert}}
-\newcommand{\call}{\textbf{call}}
-\newcommand{\dReq}{\textsc{handleDownReq}}
-\newcommand{\uReq}{\textsc{handleUpReq}}
-\newcommand{\dReqL}{\textsc{sendDownReq}}
-\newcommand{\uReqL}{\textsc{sendUpReq}}
-\newcommand{\dResp}{\textsc{sendDownResp}}
-\newcommand{\uResp}{\textsc{sendUpResp}}
-
 Let us walk through our framework using the example of a hierarchical MSI
 protocol.
+
+\floatstyle{plain}
+\restylefloat{figure}
 
@@
 $c.data[a]$ & data in cache $c$ for address $a$\\
-%$c.dir[n][a]$ & $c$'s notion of the coherence state of $n$ for address $a$\\
+%$p.dir[n][a]$ & $p$'s notion of the coherence state of $d$ for address $a$. This is meaningful only if $p$ is not an L1 cache\\
+$memory[a]$ & data in memory for address $a$\\
 \hline
 \end{tabularx}
-\caption{Notations}
-\label{table:lineinfo}
+\caption{Notations for information associated with an address in a cache}
+\label{notation}
 \end{figure}
 
-\begin{wrapfigure}{r}{.4\linewidth}
+\begin{figure}
+\begin{subfigure}{.43\linewidth}
 \centering
@@
 \end{tabular}
-\caption{$<$ and other relations for MSI states}
+\caption{$<$ and other relations for states}
 \label{msi<}
-\end{wrapfigure}
-
-Each cache stores the information given in Figure \ref{table:lineinfo}. The
-coherence state for an address denotes the (usual) permission that a cache has
-for the address: $M = \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$
-relation for coherence state values based on the permissions the states
-represent. This is shown in Figure \ref{msi<}. State $I$ is akin to address $a$
-not present in the cache.
-
-\begin{wrapfigure}{r}{.46\linewidth}
-\begin{subfigure}{.27\linewidth}
+\end{subfigure}
+\begin{subfigure}{.31\linewidth}
 \centering
@@
 \end{subfigure}
-~~~~~~~~~~~~~~
-\begin{subfigure}{.21\linewidth}
+\begin{subfigure}{.22\linewidth}
 \centering
@@
 \end{subfigure}
-\caption{$toCompatible$ and $isModified$ mappings}
-\label{fig:msimap}
-\end{wrapfigure}
+\caption{MSI state abstraction functions}
+\label{funcs}
+\end{figure}
 
-If a cache $c$ receives a request for an address $a$ from the processor, it can
-service the request straightaway if it has enough permissions for address $a$.
-Otherwise, it has to \emph{upgrade} to a state $x$, which has enough
-permissions. The cache can not upgrade its state arbitrarily, it has to ensure
-that such a state change does not violate the single-writer property (Invariant
-\label{singleWriter}), by \emph{downgrading}, \ie lowering the states of other
-caches appropriately. We define a $toCompatible(x)$ mapping (Figure
-~\ref{toCompat}), which specifies the highest state a cache can be, if another
-cache is in state $x$ for the same address. We will also define a
-$isModified(x)$ mapping to specify if an address in state $x$ in a cache can
-have data different from what's present in the memory.
+The caches are logically organized in the form of a tree
+(Figure~\ref{fig:hierarchy}). The leaves of this tree represent the L1 caches,
+with each L1 cache connected to exactly one core. The root of the tree is the
+Last-Level-Cache, and its connected to the memory. The cache hierarchy is
+inclusive, \ie if an address is present in a cache $c$, then it's present it its
+parent $c.parent$. Note that by address we mean cache-line address, not a byte
+address or a word address.
 
+Each cache stores the coherence state and the data for each address that it
+contains.
+%A cache which is not L1 (\ie an internal node in the tree of caches)
+%also stores a directory containing its notion of the coherence states of its
+%children.
+Figure \ref{notation} gives the notation used to represent all this
+information. For an L1 cache, the coherence state for an address denotes the
+(usual) permission that a core connected to the L1 cache has for the address: $M
+= \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$ relation for coherence
+state values based on the permissions the states represent. This is shown in
+Figure \ref{msi<}.
+%When an address $a$ not present in the L1 cache, then the
+%core associated with that L1 can neither read nor write that address, so it's
+%equivalent to the state of that address being $I$.
+
+For a non L1 cache, the meaning of coherence states is different. For an
+internal node $n$, if $n.state[a] = I$, then, the entire sub-tree rooted at $n$
+will have coherence state $I$ for address $a$. In particular, no L1 cache in
+this subtree will have any permission for $a$. If $n.state[a] = S$, then its
+children can either be in $S$ or $I$ state. Finally, if $n.state[a] = M$, then
+at most one of its children can be in $M$ state. This guarantees that at most
+one of the L1 caches can have write permissions, ensuring the following
+invariant:
+
+\begin{inv}
+\textbf{Single-writer Invariant}: When a cache has permission to modify data for
+an address, no other cache has read or write permission for that
+address.
+\label{singleWriter}
+\end{inv}
+
+We define a $toCompatible(x)$ mapping (Figure ~\ref{toCompat}), which specifies
+the highest state a cache can be to not violate the Single-writer Invariant, if
+another sibling cache is in state $x$ for the same address.
+
+The coherence states also denote the consistency of data in a cache with respect
+to its parent. If a cache (both L1 and non-L1) is in state $I$ for an address $a$,
+then it does not have the data corresponding to that address. If it is in state
+$S$, then its data is consistent with that of its parent. Finally, if it is in
+state $M$, then its data is not consistent with that of its parent. If none of
+its children are in state $M$, then the data in this cache is the most
+up-to-date value. But if one of its children $c$ is in the $M$ state, then the
+data in this cache is stale -- the most up-to-date value is present in the
+subtree rooted at $c$.
+
+We define a $isModified(x)$ mapping to specify if an address in state $x$ in a
+cache can have data different from what's present in its parent. The mappings
+$toCompatible$ and $isModified$ abstract away the semantics of the states so
+that our framework can be extended to other protocols like MOSI/MOESI.
+
+If a cache having no permission for an address requires read or write
+permissions for that address, then the most up-to-date value must be supplied,
+\ie:
+
+\begin{inv}
+\textbf{Read-from-last-writer}: When a cache reads data for an address, it reads
+the last modified version of the data for the address. If the data for an
+address has not been modified, the cache reads the data from memory.
+\label{lastRead}
+\end{inv}
+
+Invariants \ref{singleWriter} and \ref{lastRead} together ensure cache coherence
+\cite{Patterson}.
+
+We will first describe an atomic version of the hierarchical MSI protocol. Each
+of the caches in the hierarchy can be treated like C++/Java objects.
 \floatstyle{boxed}
@@
 \begin{figure}
+\small
+\begin{algorithmic}
+\Proc {\textsc{coreReq}}{$c, a, op, data$}
+  \If {$op = Ld$}
+    \State \textbf{if} ($c.state[a] \ge S$)
+    \State \;\;\;\; \call{} $c.$\uReq($this, a, S$);
+    \State \textbf{return} $c.data[a]$;
+  \ElsIf {$op = St$}
+    \State \textbf{if} ($c.state[a] = M$)
+    \State \;\;\;\; \call{} $c.$\uReq($this, a, M$);
+    \State $update(c.data[a], data)$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\caption{core handling a load or a store request by calling the request
+handler in its L1 cache}
+\label{coreHandle}
+\end{figure}
+
+In the atomic protocol, a core invokes the L1 cache's request handler
+\textsc{coreReq} (Figure \ref{coreHandle}). The L1 cache checks if the request
+can already be serviced, if not, it invokes its upgrade request handler \uReq{},
+which upgrades the state of the requested address. Finally, in the case of a
+load request, the data is sent back to the core, and in the case of a store, the
+cache line is updated. Note that the core supplies words or long-words while the
+cache line sizes are typically 64 bytes; only the appropriate word in the cache
+line is updated in case of a store.
+
+\begin{figure}
+\small
 
@@
 \begin{algorithmic}
-\Proc {\uReq}{$c , p, a, x$}
-  \If {$a$ is not present in $p$}
+\Proc {\uReq}{$c , a, x$}
+    \If {$a$ is not present in $this$}
+      \If {$this$ has no space for $a$}
         \State Choose address $a'$ for eviction;
-    \ForAll {$c' \in p.children$}
-      \State \call{} \dReq($p , c', a', I$);
-    \EndFor
-    \State \call{} \dResp($p , p.parent, a', I$);
-    \State Replace $a'$ with $a$ (initializing $p.state[a]$ to $I$);
+        \State \textbf{forall} ($c' \in this.children$)
+        \State \;\;\;\; \textbf{if} ($c'.state[a] > I$)
+        \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', I$);
+        \State \textbf{if} ($this.state[a'] > I$)
+        \State \;\;\;\; \call{} $this.$\dResp{}($this.parent, a', I$);
       \EndIf
-  \ForAll {$c' \in p.children, s.t., c' \neq c$}
-    \State \dReq($p , c', a, toCompatible(x)$);
-  \EndFor
-  \State \uReqL($p , p.parent, a, x$);
-  \State \uResp($p , c, a, x$);
+      \State Allocate a line for $a$;
+    \EndIf
+    \State // $a$ is already present in $this$
+    \State \textbf{forall} ($c' \in this.children$)
+    \State \;\;\;\; \textbf{if} ($c'.state[a] > toCompatible(x)$)
+    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', toCompatible(x)$);
+    \State \textbf{if} ($this$ is Last-Level-Cache \&\& $this.state[a] = I$)
+    \State \;\;\;\; \call{} $this$.\sendMem{}($a$);
+    \State \textbf{else}
+    \State \;\;\;\; \call{} $this.$\uReqL{}($this.parent, a, x$);
+  \State \textbf{if} ($c \neq core$)
+  \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
 \EndProc
@@
 \end{boxedminipage}
-\caption{Handling upgrade to $x$ request from $c$ to $p$ for address $a$}
+\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
+\label{atomicUReq1}
+\end{subfigure}
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\dReq}{$p , a, x$}
+  \If {$this.state[a] > x$}
+    \State \textbf{forall} ($c' \in this.children$)
+    \State \;\;\;\; \textbf{if} (c'.state[a] > x)
+    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a, x$);
+    \State \call{} $this.$\dResp($p, a, x$);
+  \EndIf
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{$this$ cache's downgrade request handler -- request from $this$'
+parent, $p$, to upgrade $p$'s state to $x$ for address $a$ }
+\label{atomicDReq1}
+\end{subfigure}
+
+\caption{MSI protocol}
+\label{msi-template}
+\end{figure}
+
+Figure \ref{msi-template} gives the methods of a cache to handle the upgrade
+requests from its children (or core) and downgrade requests from its parent.
+When the upgrade request handler \uReq{} for a cache $this$ is invoked by a
+cache $c$ to go to a particular state $x$, $this$ first makes sure that its
+other children are compatible with $c$ being in state $x$ for the requested
+address. This involves downgrading their states to $toCompatible(x)$. Cache
+$this$ has to upgrade its own state before upgrading the state of its child $c$,
+if $this$' state is less than the requested state. In this case, $this$ invokes
+its parent's handler for upgrade requests. Finally, when state of $this$ is not
+less than t$x$ and all its children are compatible with $x$, $this$ can upgrade
+$c$'s state, and send the data if $c$'s state was $I$ previously. This is shown
+in Figure \ref{atomicUReq1}.
+
+A cache, $this$', downgrade handler \dReq{} can also be invoked (for instance
+when its parent wants $this$ to downgrade its state to $x$ because the parent's
+upgrade request handler was invoked by some other cache. In this case, $this$
+makes sure that each of its child's state is not higher than $x$, by invoking
+their respective downgrade methods. When all the children have downgraded, the
+cache downgrades its own state to $x$, and sends data back to the parent, if it
+has been modified (\ie $isModified(x)$ is $True$).
+
+\begin{figure}
+\small
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uReqL}{$p, a, x$}
+  \State \call{} $p.$\uReq($this, a, x$);
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{Upgrading address $a$ of $this$ to $x$, by invoking the upgrade handler
+of $this$'s parent, $p$}
 \label{uReq}
@@
 \begin{algorithmic}
-\Proc {\dReq}{$p , c, a, x$}
-  \ForAll {$c' \in c.children$}
-    \State \dReqL($c , c', a, x$);
-  \EndFor
-  \State \dResp($c , p, a, x$);
+\Proc {\dReqL}{$c , a, x$}
+  \State \call{} $c.$\dReq($this, a, x$);
 \EndProc
@@
 \end{boxedminipage}
-\caption{Handling downgrade to $x$ request from $p$ to $c$ for address $a$}
-\label{dReq}
-\end{subfigure}
-
-\caption{MSI protocol}
-\label{msi}
-\end{figure}
-
-\begin{figure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\uReqL}{$c , p, a, x$}
-  \State \call{} \uReq($c, p, a, x$);
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending upgrade to $x$ request from $c$ to $p$ for address $a$}
-\label{uReq}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\dReq}{$p , c, a, x$}
-  \State \call{} \dReq($p, c, a, x$);
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending downgrade to $x$ request from $p$ to $c$ for address $a$}
+\caption{Downgrading address $a$ of $this$'s child $c$ to $x$, by invoking
+$c$'s downgrade handler}
 \label{dReq}
@@
 \begin{algorithmic}
-\Proc {\uResp}{$p , c, a, x$}
-  \If {$c.state[a] < x$}
+\Proc {\uResp}{$c, a, x$}
   \If {$c.state[a] = I$}
-      \State $c.data[a] \gets p.data[a]$;
+    \State $c.data[a] \gets this.data[a]$;
   \EndIf
   \State $c.state[a] \gets x$;
-  \EndIf
 \EndProc
@@
 \end{boxedminipage}
-\caption{Sending upgrade to $x$ response from $p$ to $c$ for address $a$}
+\caption{$this$ finally upgrading address $a$ of its child $c$ to $x$, and
+updating the data of $c$ if necessary}
 \label{uResp}
@@
 \begin{algorithmic}
-\Proc {\dResp}{$c , p, a, x$}
-  \If {$c.state[a] > x$}
+\Proc {\dResp}{$p, a, x$}
   \If {$isModified(c.state[a])$}
-      \State $p.data[a] \gets c.data[a]$;
+    \State $p.data[a] \gets this.data[a]$;
   \EndIf
-    \State $c.state[a] \gets x$;
-  \EndIf
+  \State $this.state[a] \gets x$;
 \EndProc
@@
 \end{boxedminipage}
-\caption{Sending downgrade to $x$ response from $c$ to $p$ for address $a$}
+\caption{$this$ finally downgrading its state for address $a$ to $x$ and
+updating data of its parent $p$ if necessary}
 \label{dResp}
@@
 \begin{algorithmic}
-\Proc {\uResp}{$p , core, a, x$}
-  \State $c.data[a] \gets p.data[a]$;
+\Proc {\sendMem{}}{$a$}
+  \State $this.data[a] \gets memory[a]$;
+  \State $this.state[a] \gets M$;
 \EndProc
@@
 \end{boxedminipage}
-\caption{Sending upgrade to $x$ response from $p$ to $core$ for address $a$}
+\caption{Last-Level-Cache $this$ reading data from memory for address $a$}
 \label{uResp}
@@
 
-\caption{Atomic MSI}
+\caption{Methods used in the MSI protocol dealing with other objects}
 \label{atomic}
@@
 
+Figure \ref{msi-template} gives the pseudo-code involved, including the
+actions needed to evict a cache line and replace it with another line, in case
+there are no free slots available for an address. For reasons that will become
+clear in the distributed setting, each cache can directly read or write its own
+state and data for each address. It can also directly read it's child's state,
+but not directly write it. It can not read it's parents state directly. It can
+not directly read or write other cache's data. In order to perform the required
+actions indirectly, the cache can invoke other cache's corresponding
+\textbf{method}s packaged together as shown in Figure \ref{atomic}. To be
+precise we use $this$ explicity even if the object invokes its own methods.
+
+In the case of an L1 cache, once the cache's parent has upgraded its state,
+there is no child to upgrade to. Similarly, in the case of a Last-Level-Cache,
+the cache simply reads data from the memory if the line is not present in the
+cache, since it has no parent whose upgrade request handler it can invoke. All
+lines in the Last-Level-Cache are upgraded automatically to be in $M$ state.
+
+\subsection{Refactoring the handler methods}
+Figure \ref{msi-template} shows how the request handler methods \uReq{} and
+\dReq{} can be refactored to make it amenable to distributed implementation. The
+upgrade and downgrade request handlers are not allowed to invoke other cache's
+upgrade or downgrade request handlers, or change other cache's states. Instead,
+the same calls are made via 5 more local methods listed in Figure \ref{atomic}.
+These methods in turn invoke the other caches upgrade and downgrade request
+handlers, or change the states of other caches.
+
 \begin{figure}
+\small
 
@@
 \begin{algorithmic}
-\Proc {\uReqL}{$c , p, a, x$}
-  \State \send{} \Req{c}{p}{a}{x};
-  \State \pop{} \Resp{p}{c}{a}{z};
-  \If {$c.state[a] = I$}
-    \State \pop{} \Data{p}{c}{a}{d};
-    \State $c.data[a] \gets d$;
+\Proc {\uReqL}{$p, a, x$}
+  \State \send{} \Req{this}{p}{a}{x};
+  \State \receive{} \Resp{p}{this}{a}{z};
+  \If {$this.state[a] = I$}
+    \State \receive{} \Data{p}{this}{a}{d};
+    \State $this.data[a] \gets d$;
   \EndIf
-  \State $c.state[a] \gets z$;
+  \State $this.state[a] \gets z$;
 \EndProc
@@
 \end{boxedminipage}
-\caption{Sending upgrade to $x$ request from $c$ to $p$ for address $a$}
+\caption{Upgrading address $a$ of $this$ to $x$, by sending a request to
+$this$'s parent $p$}
 \label{uReqLocal}
@@
 \begin{algorithmic}
-\Proc {\dReqL}{$p , c, a, x$}
-  \State \send{} \Req{p}{c}{a}{x};
-  \State \pop{} \Resp{c}{p}{a}{z};
-  \If {$isModified(p.dir[c][a])$}
-    \State \pop{} \Data{c}{p}{a}{d};
-    \State $p.data[a] \gets d$;
+\Proc {\dReqL}{$c, a, x$}
+  \State \send{} \Req{this}{c}{a}{x};
+  \State \receive{} \Resp{c}{this}{a}{z};
+  \If {$isModified(this.dir[c][a])$}
+    \State \receive{} \Data{c}{this}{a}{d};
+    \State $this.data[a] \gets d$;
   \EndIf
-  \State $p.dir[c][a] \gets z$;
+  \State $this.dir[c][a] \gets z$;
 \EndProc
@@
 \end{boxedminipage}
-\caption{Sending downgrade to $x$ request from $p$ to $c$ for address $a$}
+\caption{Downgrading address $a$ of $this$'s child $c$ to $x$, by sending a
+request to $c$}
+\label{dReqLocal}
+\end{subfigure}
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uResp}{$c, a, x$}
+  \State \textbf{if} ($this.dir[c][a] = I$)
+  \State \;\;\;\; \send{} \Data{this}{c}{a}{this.data[a]};
+  \State $this.dir[c][a] \gets x$;
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{$this$ finally upgrading address $a$ of its child $c$ to $x$, sending a
+response to $c$ notifying of the upgrade and sending data to $c$ if necessary}
+\label{uResp}
+\end{subfigure}
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\dResp}{$p, a, x$}
+  \State \textbf{if} ($isModified(this.state[a])$)
+  \State \;\;\;\; \send{} \Data{this}{p}{a}{this.data[a]};
+  \State $this.state[a] \gets x$;
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{$this$ finally downgrading its state for address $a$ to $x$, sending a
+response to $c$ notifying of the downgrade and sending data to $c$ if necessary}
 \label{dResp}
@@
 \begin{algorithmic}
-\Proc {\uResp}{$p , c, a, x$}
-  \If {$p.dir[c][a] < x$}
-    \If {$p.dir[c][a] = I$}
-      \State \send{} \Data{p}{c}{a}{p.data[a]};
-    \EndIf
-    \State $p.dir[c][a] \gets x$;
+\Proc {\dRespL}{$c , a, x$}
+  \If {$isModified(this.dir[c][a])$}
+    \State \receive{} \Data{c}{this}{a}{d};
+    \State $this.data[a] \gets d$;
   \EndIf
@@
 \end{boxedminipage}
-\caption{Sending upgrade to $x$ response from $p$ to $c$ for address $a$}
+\caption{$this$ handling unsolicited downgrade to $x$ response from $c$ for
+address $a$. This happens if $c$ evicted address $a$}
 \label{uResp}
@@
 \begin{algorithmic}
-\Proc {\dResp}{$c , p, a, x$}
-  \If {$c.state[a] > x$}
-    \If {$isModified(c.state[a])$}
-      \State \send{} \Data{c}{p}{a}{c.data[a]};
-    \EndIf
-    \State $c.state[a] \gets x$;
-  \EndIf
+\Proc {\sendMem{}}{$a$}
+  \State \send{} \Req{this}{Memory}{a}{M};
+  \State \receive{} \Data{Memory}{this}{a}{d};
+  \State $this.data[a] \gets d$;
+  \State $this.state[a] \gets M$;
 \EndProc
@@
 \end{boxedminipage}
-\caption{Sending downgrade to $x$ response from $c$ to $p$ for address $a$}
-\label{dResp}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\uResp}{$p , core, a, x$}
-  \State \send{} \Data{p}{c}{a}{p.data[a]};
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending upgrade to $x$ response from $p$ to $core$ for address $a$}
+\caption{Last-Level-Cache $this$ sending and receivging data from memory for
+address $a$}
 \label{uResp}
@@
 
-\caption{Realistic MSI protocol}
+\caption{Non atomic implementation of MSI protocol}
 \label{realistic}

### 2012-11-19 paper/isca40format/Introduction.tex
stuff

@@
 
-Large shared-memory multicores are likely to use cache-coherence protocols which are directory-based, hierarchical, and implemented using message passing.
-Even though such protocols have been around for two decades (see, for example, DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of such protocols remains a challenge.
-It is difficult to convince oneself that all corner cases have been handled adequately without a formal proof.
-And yet formal proofs of real implementations are practically non-existent.
-One can't settle for conservative choices in cache-coherence protocol designs because performance (\eg the number of messages needed to maintain coherence, the buffer requirements, etc) is as important as correctness.
-Consequently proving a simple protocol for a small number of caches, under unbounded buffer assumption is of limited use in gaining confidence in the correctness of a real protocol.
+Large shared-memory multicores are likely to use cache-coherence protocols
+which are directory-based, hierarchical, and implemented using message passing.
+Even though such protocols have been around for two decades (see, for example,
+DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
+such protocols remains a challenge. It is difficult to convince oneself that
+all corner cases have been handled adequately without a formal proof. And yet
+formal proofs of real implementations are practically non-existent. One can't
+settle for conservative choices in cache-coherence protocol designs because
+performance (\eg the number of messages needed to maintain coherence, the
+buffer requirements, etc) is as important as correctness. Consequently proving
+a simple protocol for a small number of caches, under unbounded buffer
+assumption is of limited use in gaining confidence in the correctness of a real
+protocol.
 
-In this paper we will consider systems where caches form a tree hierarchy and caches are \emph{inclusive}, that is, if a cache contains address $a$ then its parent is guaranteed to contain address $a$, though the parent may have stale data for $a$.
-Every cache or node in the system has an associated directory which contains information about its children caches.
-The coherence state changes are orchestrated by passing requests and responses between the caches via the directory.
-Unlike bus-based protocols which lock the bus till all the required state transitions are done thereby ensuring atomicity of state transitions, in a message-passing protocol, requests and responses are split leading to transient states.
-This creates scenarios not present in bus-based protocols, which can lead to coherence violations or deadlocks, if the protocol is not designed or implemented carefully.
+In this paper we will consider systems where caches form a tree hierarchy and
+caches are \emph{inclusive}, that is, if a cache contains address $a$ then its
+parent is guaranteed to contain address $a$, though the parent may have stale
+data for $a$. Every cache or node in the system has an associated directory
+which contains information about its children caches. The coherence state
+changes are orchestrated by passing requests and responses between the caches
+via the directory.  Unlike bus-based protocols which lock the bus till all the
+required state transitions are done thereby ensuring atomicity of state
+transitions, in a message-passing protocol, requests and responses are split
+leading to transient states. This creates scenarios not present in bus-based
+protocols, which can lead to coherence violations or deadlocks, if the protocol
+is not designed or implemented carefully.
 
-Several factors contribute to the difficulty of designing protocols for such distributed, hierarchical and message-passing systems:
+Several factors contribute to the difficulty of designing protocols for such
+distributed, hierarchical and message-passing systems:
 
-\begin{enumerate} \item Lots of concurrent events must be dealt with without a central authority to impose a sequence on events; \item The state on which transitions are based is physically distributed in space and cannot be accessed atomically; \item Physical network may block the transmission of a message because of lack of communication buffers or head-of-the-line blocking; \item The cache or directory may reach a state which can not handle a particular message.
-An incomplete set of transition rules may cause the protocol to deadlock.
-\end{enumerate}
+\begin{enumerate} \item Lots of concurrent events must be dealt with without a
+central authority to impose a sequence on events; \item The state on which
+transitions are based is physically distributed in space and cannot be accessed
+atomically; \item Physical network may block the transmission of a message
+because of lack of communication buffers or head-of-the-line blocking; \item
+The cache or directory may reach a state which can not handle a particular
+message. An incomplete set of transition rules may cause the protocol to
+deadlock.  \end{enumerate}
 
-As examples of problems in designing correct protocols, we describe several concrete and familiar scenarios from two-level systems (a private L1 cache for each processor and a memory with a directory).
-States $M, S$ and $I$ have their usual meaning in MSI protocol ($M$ represents read and write permissions for an address, $S$, read-only permissions and $I$, no permission)
+As examples of problems in designing correct protocols, we describe several
+concrete and familiar scenarios from two-level systems (a private L1 cache for
+each processor and a memory with a directory). States $M, S$ and $I$ have their
+usual meaning in MSI protocol ($M$ represents read and write permissions for an
+address, $S$, read-only permissions and $I$, no permission)
 
-\noindent \emph{Example 1:} A cache in state $S$ for a line gets a store request from the processor and sends a request for changing the line's state to $M$ to the directory controller.
-The controller may eventually send a response permitting the cache to go to $M$ state for that line.
-The response would signify that the states in the other caches for that line \emph{were} $I$.
-But the messages take time to propagate and the response is necessarily based on the old states of other caches -- the current state of another cache may not be $I$, violating the Single-writer-at-a-time invariant and hence breaking cache coherence.
+\noindent \emph{Example 1:} A cache in state $S$ for a line gets a store
+request from the processor and sends a request for changing the line's state to
+$M$ to the directory controller. The controller may eventually send a response
+permitting the cache to go to $M$ state for that line. The response would
+signify that the states in the other caches for that line \emph{were} $I$. But
+the messages take time to propagate and the response is necessarily based on
+the old states of other caches -- the current state of another cache may not be
+$I$, violating the Single-writer-at-a-time invariant and hence breaking cache
+coherence.
 
-\noindent \emph{Example 2:} A cache in state $S$ for a line gets a store request from the processor and sends a request for changing the line's state to $M$ to the directory controller.
-Instead of getting a response permitting it to go to $M$ state, the cache may receive a request from the directory controller asking this cache to go to $I$ state (invalidate request)-- this could have happened if the directory is serving some other cache's request to go to $M$ state for the same line.
-It is not easy to determine the correct behavior of the former cache on receiving the invalidate request -- whether it should ignore the invalidate request and wait for a response to its request or whether it should invalidate the line.
-Would ignoring the invalidate request lead to a deadlock?
+\noindent \emph{Example 2:} A cache in state $S$ for a line gets a store
+request from the processor and sends a request for changing the line's state to
+$M$ to the directory controller. Instead of getting a response permitting it to
+go to $M$ state, the cache may receive a request from the directory controller
+asking this cache to go to $I$ state (invalidate request)-- this could have
+happened if the directory is serving some other cache's request to go to $M$
+state for the same line. It is not easy to determine the correct behavior of
+the former cache on receiving the invalidate request -- whether it should
+ignore the invalidate request and wait for a response to its request or whether
+it should invalidate the line. Would ignoring the invalidate request lead to a
+deadlock?
 
-\noindent \emph{Example 3:} A cache receives an invalidate request from the directory controller for a line that is not present in the cache.
-Is such a request even possible?
-The directory does not have access to the up-to-date state of the cache, and so can potentially make this request.
-How should a cache handle this request; should it drop such a request, or respond saying that it does not have the line?
+\noindent \emph{Example 3:} A cache receives an invalidate request from the
+directory controller for a line that is not present in the cache. Is such a
+request even possible? The directory does not have access to the up-to-date
+state of the cache, and so can potentially make this request. How should a
+cache handle this request; should it drop such a request, or respond saying
+that it does not have the line?
 
-These examples give a glimpse of the complexity of designing and implementing a correct cache coherence protocol in a directory-based non-bus setting.
-Even for a 2-level cache hierarchy, designing and implementing the protocol is quite difficult.
-This problem is greatly exacerbated in a multi-level cache hierarchy, which is becoming the norm because of the increasing core counts.
+These examples give a glimpse of the complexity of designing and implementing a
+correct cache coherence protocol in a directory-based non-bus setting.  Even
+for a 2-level cache hierarchy, designing and implementing the protocol is quite
+difficult. This problem is greatly exacerbated in a multi-level cache
+hierarchy, which is becoming the norm because of the increasing core counts.
 
-In this paper, we present a substantially easier method to design directory-based cache-coherence protocols for such systems.
-The method breaks down the design problem in two parts: a part that is shared by a number of concrete protocols which we call \glob{}s, and another part that is specific to the protocol to be designed and implemented, called \policy{}.
-\glob{}s provide guarantees such as every request will eventually get a response, the directory's notion of the cache states is ``conservative'' \ie a cache can not have more permissions for a line than what the directory thinks it has, \etc; \glob{} enables \policy{} to focus on the high level properties such as the set of actions a cache or a directory should perform to maintain the Single-writer-at-a-time and the Read-from-last-writer invariants.
+In this paper, we present a substantially easier method to design cache
+coherence protocols for such systems. The user specifies a protocol for a
+two-level system without message passing. The protocol rules can simultaneously
+or \emph{atomically} examine and change the state of all the caches in the
+system. For example, the MSI protocol may be described as follows:
 
-The \glob{}s can in turn be guaranteed using a set of (\policy{} independent) \local{}s which dictate local properties of a node such as when a request can be handled, what state changes happen on sending a response, properties about the network, \etc.
-We have formally proved that \local{}s ensure \glob{}s using the Coq Theorem Prover \cite{}, however, a presentation of the formal proof is beyond the scope of this paper.
-The \glob{}s and the \local{}s form the cornerstone of our framework.
-The separation of \glob{}s and \local{}s is akin to having an abstraction layer which hides low level details.
+***** Murli insert the description of MSI **** We give a procedure that
+transforms such simple and global protocol descriptions into a distributed,
+message-passing protocol for hierarchal caches. Our procedure is based on
+preserving some invariants in rule transformations and has been shown to be
+sound and complete with respect the global description using the coq theorem
+prover \cite{}, however, a discussion of the formal proof is beyond the scope
+of this paper. Both the procedure and the proof are parameterized by the number
+of cache levels and the number of caches in each level of the hierarchy as well
+as by a predicate which involves a local ``compatibility of state'' test.
 
-The \glob{}s of our framework can be used to design a wide range of cache coherence protocols, notably MSI, MESI, MOSI, MOESI, as well as their hierarchical version.
-We will show how these are designed as \policy{}s, focusing on high-level details.
+It should be noted that the flexibility of our procedure comes from the fact
+that we do not assign any meaning to the value of a coherence state beyond
+upgrades or downgrades. Thus, how or why a particular cache state implies that
+a particular address may not be present in any other cache stems from protocol
+specific ``compatibility of state'' check.
 
-Our method substantially reduces the design and proof burden for new protocols because of the common abstract protocol.
-The abstract protocol is a set of transitions that the caches and directories follow to send or receive a request or a response.
-We have formally proved that these transitions ensure the following properties: a) the directory's notion of a cache's state is ``conservative'' in that a cache is never going to have more permissions that what the directory thinks the cache has; and b) any request will eventually get a corresponding response.
-A formal proof of the abstract protocol was written using the Coq theorem prover \cite{}, however, a presentation of the formal proof is beyond the scope of this paper.
+The main contributions of this paper are: 1. An automatic procedure to refine
+centralized protocols without message passing into protocols for distributed,
+hierarchal, cache system with message-passing; 2. Specification of the
+invariants to be preserved by the procedure; and 3. Showing that our procedure
+does not introduce any overhead over direct implementation of these protocols.
+We think that our method of designing protocols substantially reduces the
+required effort, and that our procedure is flexible enough to implement (while
+ensuring correctness of) most, if not all, inclusive message-passing
+directory-based protocols.
 
-It should be noted that the flexibility of our methodology to design multiple cache coherence protocols comes from the fact that we do not assign any meaning to the value of a coherence state beyond upgrades or downgrades.
-Thus, how or why a particular cache state implies that a particular address may not be present in any other cache stems from protocol specific ``compatibility of state'' check.
-
-The main contributions of this paper are: 1.
-A new abstract protocol that satisfies two important invariants for distributed and hierarchal message-passing protocols; 2.
-Instantiation of the abstract protocol to implement four well-known protocols (MSI, MESI, MOSI, MOESI) and their variations and showing why their correctness is guaranteed as long as the abstract protocol is correct, \ie it provably implements the two invariants; and 3.
-Our method does not introduce any overhead over direct implementation of these protocols.
-We think that this new method of designing protocols substantially reduces the required effort, and that our abstract protocol is flexible enough to implement (while ensuring correctness of) most if not all inclusive, message-passing, directory-based protocols.
-
-\paragraph{Paper organization:} Section \ref{sec:abstract} describes the abstract protocol; Section \ref{sec:properties} gives the two properties that are guaranteed by the abstract protocol and provides some intuition for proving these properties.
-We also discuss the communication buffer requirements for the abstract protocol and show that the requirement is only proportional to the number of levels and not to the number of processors.
-In Section \ref{sec:protocols}, we show how the abstract protocol can be used to implement MSI, MESI, MOSI and MOESI protocols and their variants.
-In Section \ref{sec:performance}, we show that these formally verified family of protocols require no extra state bits, nor extra message transfers compared to other published implementations.
-In Section \ref{sec:related}, we present some of the related work and finally in Section \ref{sec:conclusion}, we present some directions for the future work.
+\paragraph{Paper organization:} *** to be redone *** Section \ref{sec:abstract}
+describes the abstract protocol; Section \ref{sec:properties} gives the two
+properties that are guaranteed by the abstract protocol and provides some
+intuition for proving these properties. We also discuss the communication
+buffer requirements for the abstract protocol and show that the requirement is
+only proportional to the number of levels and not to the number of processors.
+In Section \ref{sec:protocols}, we show how the abstract protocol can be used
+to implement MSI, MESI, MOSI and MOESI protocols and their variants.  In
+Section \ref{sec:performance}, we show that these formally verified family of
+protocols require no extra state bits, nor extra message transfers compared to
+other published implementations.  In Section \ref{sec:related}, we present some
+of the related work and finally in Section \ref{sec:conclusion}, we present
+some directions for the future work.

### 2012-11-19 paper/isca40format/Invariants.tex
stuff

@@
 
-\begin{figure}
+\begin{figure}\small
 \begin{inv}
-An incoming request message from a child should not block an incoming request
-message from the parent, even for a different address\label{cReqNoBlockPReq}
+An incoming request message from a child or the core should not block an
+incoming request message from the parent, even for a different
+address\label{cReqNoBlockPReq}
 \end{inv}
@@
 response message for address $a$, even from a different
-source.\label{reqNoBlockReq}
+source.\label{reqNoBlockResp}
 \end{inv}
@@
 
-\begin{figure}
+\begin{figure}\small
 \begin{inv}
@@
 
-\begin{figure}
+\begin{figure}\small
+\begin{inv}
+Each node has 
+\begin{enumerate}
+\item at least one dedicated thread resource for handling only messages (requests or
+responses) from the parent,
+\item at least one dedicated thread resource for handling only requests from any
+child, and
+\item at least one dedicated thread resource for handling only responses (from children
+or from the parent).
+\end{enumerate}
+\label{dedicate}
+\end{inv}
+\caption{Dedicated resources in a node}
+\end{figure}
+
+\begin{figure}\small
 \begin{inv}
 Starvation-freedom: A thread that keeps becoming ready to execute will
-eventually be scheduled by the scheduler to execute.
+eventually be scheduled to execute.
 \end{inv}

### 2012-11-19 paper/isca40format/Theorems.tex
stuff

@@
 \label{willRecv}
+\end{theorem}
+
+\begin{theorem}
+If a thread is waiting for a line to be allocated, \ie for any address to become
+available for eviction, then some address will eventually become available for
+eviction
+\label{evictDead}
 \end{theorem}
@@
 These theorems are needed to guarantee that Figure \ref{realistic} protocol is
-indeed correct. If, for instance, a thread wants to send a message in the output
-channel of the node and is not able to send one forever, then the thread can not
-proceed further. Similarly, if a thread is waiting for a response (for a request
+indeed correct. If, for instance, a thread wants to send a message in the
+output channel of the node and is not able to send one forever, then the thread
+can not proceed further. If a thread is waiting for a response (for a request
 previously sent), and it never gets the response, the thread can never proceed
-further. These scenarios create a deadlock.
+further. A thread can also not proceed if it's waiting for being allocated in a
+free line but never gets one because all the lines are locked up. These
+scenarios create a deadlock.
 

### 2012-11-20 paper/isca40format/Introduction.tex
stuff

@@
 
-Large shared-memory multicores are likely to use cache-coherence protocols
-which are directory-based, hierarchical, and implemented using message passing.
-Even though such protocols have been around for two decades (see, for example,
+Large cache-coherent shared-memory multicores are likely to use protocols which
+are directory-based, hierarchical, and implemented using message passing.  Even
+though such protocols have been around for two decades (see, for example,
 DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
@@
 
-In this paper we will consider systems where caches form a tree hierarchy and
-caches are \emph{inclusive}, that is, if a cache contains address $a$ then its
-parent is guaranteed to contain address $a$, though the parent may have stale
-data for $a$. Every cache or node in the system has an associated directory
-which contains information about its children caches. The coherence state
-changes are orchestrated by passing requests and responses between the caches
-via the directory.  Unlike bus-based protocols which lock the bus till all the
-required state transitions are done thereby ensuring atomicity of state
-transitions, in a message-passing protocol, requests and responses are split
-leading to transient states. This creates scenarios not present in bus-based
-protocols, which can lead to coherence violations or deadlocks, if the protocol
-is not designed or implemented carefully.
+In this paper we will consider systems where caches logically form a tree
+hierarchy and caches are \emph{inclusive}, that is, if a cache contains address
+$a$ then its parent is guaranteed to contain address $a$, though the parent may
+have stale data for $a$ (Figure \ref{hierarchy}). Physically the system looks
+like a collection of caches connected to a message passing network (Figure
+\ref{physical}). Caches typically contain data and state for each cache line,
+and have two pairs of channels for communicating with parent and children
+caches. For the lowest level caches (L1's) the channels for communicating with
+children are connected to processors, while for the last-level caches the
+channels for communicating with the parents are connected to the memory.
+
+Every cache, unless it is an L1 cache, has a directory which contains
+information about the state of its children caches. The cache state changes to
+maintain coherence are orchestrated by passing requests and responses between
+the caches using the local directory information.  Unlike bus-based protocols
+which lock the bus till all the required state transitions are done thereby
+ensuring atomicity of state transitions, in a message-passing protocol,
+requests and responses are split leading to transient states. This creates
+scenarios not present in bus-based protocols, which can lead to coherence
+violations or deadlocks, if the protocol is not designed or implemented
+carefully.
 
@@
 central authority to impose a sequence on events; \item The state on which
-transitions are based is physically distributed in space and cannot be accessed
-atomically; \item Physical network may block the transmission of a message
-because of lack of communication buffers or head-of-the-line blocking; \item
-The cache or directory may reach a state which can not handle a particular
-message. An incomplete set of transition rules may cause the protocol to
-deadlock.  \end{enumerate}
+transitions are based conceptually is physically distributed and cannot be
+accessed atomically; \item Physical network may block the transmission of a
+message because of lack of communication buffers or head-of-the-line blocking;
+\item The cache or directory may reach a state in which a particular message
+cannot be handled. An incomplete set of transition rules may cause the protocol
+to deadlock.  \end{enumerate}
 
@@
 These examples give a glimpse of the complexity of designing and implementing a
-correct cache coherence protocol in a directory-based non-bus setting.  Even
-for a 2-level cache hierarchy, designing and implementing the protocol is quite
-difficult. This problem is greatly exacerbated in a multi-level cache
-hierarchy, which is becoming the norm because of the increasing core counts.
+correct cache coherence protocol in a distributed setting.  Even for a 2-level
+cache hierarchy, designing and implementing the protocol is non-trivial; the
+problem gets greatly exacerbated in large multi-level cache system.
 
 In this paper, we present a substantially easier method to design cache
-coherence protocols for such systems. The user specifies a protocol for a
-two-level system without message passing. The protocol rules can simultaneously
-or \emph{atomically} examine and change the state of all the caches in the
-system. For example, the MSI protocol may be described as follows:
+coherence protocols for such systems. The user specifies a protocol for a cache
+hierarchy as sequential methods to process requests. There is no message
+passing so the protocol procedure can simultaneously or \emph{atomically}
+examine and change the state of all the caches in the system. We then
+systematically transform such a protocol description into a distributed,
+message-passing protocol. The resulting protocol is presented as a collection
+of cooperating sequential methods which operate on ``cache objects''.  These
+sequential methods communicate with each other via blocking \emph{send} and
+\emph{receive} commands and thus can get suspended. Our transformation is based
+on preserving some invariants which guarantee that each request eventually gets
+a response and thus, our threads can never wait indefinitely.  It has been
+shown using the Coq theorem prover \cite{}, that the distributed protocol we
+derive is correct.  Both the transformation procedure and the proof are
+parameterized by the number of cache levels and the number of caches in each
+level of the hierarchy as well as by a predicate which involves a local
+``compatibility of state'' test. A discussion of the formal proof is beyond the
+scope of this paper.
 
-***** Murli insert the description of MSI **** We give a procedure that
-transforms such simple and global protocol descriptions into a distributed,
-message-passing protocol for hierarchal caches. Our procedure is based on
-preserving some invariants in rule transformations and has been shown to be
-sound and complete with respect the global description using the coq theorem
-prover \cite{}, however, a discussion of the formal proof is beyond the scope
-of this paper. Both the procedure and the proof are parameterized by the number
-of cache levels and the number of caches in each level of the hierarchy as well
-as by a predicate which involves a local ``compatibility of state'' test.
+It should be noted that the generality of our procedure comes from the fact
+that we do not assign any meaning to a coherence state beyond upgrades and
+downgrades. Thus, how or why a particular cache state implies that a particular
+address may not be present in any other cache stems from protocol specific
+``compatibility of state'' check.
 
-It should be noted that the flexibility of our procedure comes from the fact
-that we do not assign any meaning to the value of a coherence state beyond
-upgrades or downgrades. Thus, how or why a particular cache state implies that
-a particular address may not be present in any other cache stems from protocol
-specific ``compatibility of state'' check.
+The main contributions of this paper are: 1. An automatic procedure to refine a
+centralized MSI protocol without message passing into protocols for
+distributed, hierarchical, cache system with message-passing; 2. Specification
+of the invariants that ensure the correctness of the transformation procedure;
+and 3. Giving a provable lower bound on network virtual channels and buffers to
+avoid deadlocks; and 4. Showing that our procedure does not introduce any
+overhead over direct implementation of these protocols.  We think that our
+method of designing protocols substantially reduces the required effort in
+implementing provably correct protocols.
 
-The main contributions of this paper are: 1. An automatic procedure to refine
-centralized protocols without message passing into protocols for distributed,
-hierarchal, cache system with message-passing; 2. Specification of the
-invariants to be preserved by the procedure; and 3. Showing that our procedure
-does not introduce any overhead over direct implementation of these protocols.
-We think that our method of designing protocols substantially reduces the
-required effort, and that our procedure is flexible enough to implement (while
-ensuring correctness of) most, if not all, inclusive message-passing
-directory-based protocols.
-
-\paragraph{Paper organization:} *** to be redone *** Section \ref{sec:abstract}
-describes the abstract protocol; Section \ref{sec:properties} gives the two
-properties that are guaranteed by the abstract protocol and provides some
-intuition for proving these properties. We also discuss the communication
-buffer requirements for the abstract protocol and show that the requirement is
-only proportional to the number of levels and not to the number of processors.
-In Section \ref{sec:protocols}, we show how the abstract protocol can be used
-to implement MSI, MESI, MOSI and MOESI protocols and their variants.  In
-Section \ref{sec:performance}, we show that these formally verified family of
-protocols require no extra state bits, nor extra message transfers compared to
-other published implementations.  In Section \ref{sec:related}, we present some
-of the related work and finally in Section \ref{sec:conclusion}, we present
-some directions for the future work.
+\paragraph{Paper organization:} Section \ref{sec:GlobalMsi} describes the MSI
+protocol for a hierarchy of caches using nonsuspensive sequential methods. We
+use it a specification for proving the correctness of distributed MSI
+protocols. Section \ref{sec:DistributedMsi} systematically transforms the
+protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. Section
+\ref{sec:properties} gives a set of global properties or invariants that are
+needed to preserve the correctness of the transformation procedure. The
+desirability of these invariants is intuitively obvious. Section
+\ref{sec:localProperties} describes local properties that are needed to
+guarantee the desirable invariants discussed in Section \ref{sec:properties};
+it can be easily seen by inspection (because of the local nature of the
+properties) that the distributed protocol obeys these properties. Section
+\ref{sec:network} specifies the minimum virtual channel requirements to ensure
+the local properties specified in Section \ref{sec:localProperties}. We show
+that these formally verified family of protocols require no extra state bits,
+nor extra message transfers compared to other published implementations.  In
+Section \ref{sec:related}, we present some of the related work and finally in
+Section \ref{sec:conclusion}, we present some directions for the future work.

### 2012-11-20 paper/isca40format/Invariants.tex
stuff

@@
-\section{Invariants}
-\label{sec:invariants}
+\section{Local Properties}
+\label{sec:localProperties}
+
+In this section we list the local properties that are required to ensure the
+invariants in Section \ref{sec:properties}. We can easily see by inspection that the distributed MSI protocol in Figure \ref{realistic} does not violate the local properties in this section.
+
+ which in turn makes the 
+distributed cache coherence protocol given in Figure \ref{realistic} correct. We
+ give the intuition behind the these invariants whenever applicable, though the
+formal proof on why these local properties guarantee the invariants of Section
+\ref{sec:theorem} is beyond the scope of this paper.
+
+In all the scenarios that we present below, we will use a system containing two
+levels of caches, L1 and L2. There are 2 L1 caches $c_1$ and $c_2$, and both are
+connected to the shared L2. Both $c_1$ and $c_2$ can handle multiple outstanding
+requests from the cores.
 
@@
 \restylefloat{figure}
-
-\begin{figure}\small
-\begin{inv}
-An incoming request message from a child or the core should not block an
-incoming request message from the parent, even for a different
-address\label{cReqNoBlockPReq}
-\end{inv}
-\begin{inv}
-An incoming request messages for an address $a$ should not block an incoming
-response message for address $a$, even from a different
-source.\label{reqNoBlockResp}
-\end{inv}
-\caption{Blocking restrictions}
-\label{blocking}
-\end{figure}
 
@@
 
+Let's illustrate the need for Local-Property \ref{cRespFifo}. Let's say $c_1$ is
+in state $M$ and $c_2$ in state $I$ for address $a$, and the corresponding
+directory states in L2 match with their actual states. $c_2$ gets a store
+request from its processor and sends an upgrade request to L2. L2 sends a
+downgrade to $S$ request to $c_1$. $c_1$ receives the downgrade to $S$ request
+and downgrades to $S$, sending a downgrade response to L2. Let's say $c_1$
+decides to evict address $a$ to make room for another address. $c_1$ sends a
+downgrade to $I$ response to L2. Let's say the second response (downgrade to
+$I$) reaches L2 first, violating Local-Property \ref{cRespFifo}. L2 sends a
+upgrade to $S$ response to $c_2$. Let's say $c_2$ received the upgrade response
+and changed state to $S$.  L2 then receives the second downgrade to $S$ response
+from $c_1$.  Now L2 will change its directory state to $S$ for $c_1$ because it
+received a downgrade to $S$ response last. Now let's say $c_2$ sent an upgrade
+to $M$ request to L2 because of a store from its core. L2 will send an downgrade
+to $I$ request to $c_1$, but $c_1$ drops the request since it's already in state
+$I$.  Hence L2 will never receive a response from $c_1$ leading to a deadlock.
+This is illustrated in Figure \ref{complicated}.
+
+Let's illustrate the need for Local-Property \ref{reqNoOvertakeResp}. Let's say
+$c_1$ is in state $S$ and $c_2$ in state $I$ for address $a$. Suppose $c_1$
+receives a store request and sends an upgrade to $M$ request to L2. L2 sends
+back an upgrade to $M$ response to $c_1$. $c_2$ gets a load request for address
+$a$ and sends an upgrade to $S$ request to L2. L2 sends a request to $S$ to
+$c_1$. Suppose Local-Property \ref{reqNoOvertakeResp} is violated and the
+request to $S$ reaches $c_1$ before the response to $M$. $c_1$ will drop the
+request since its already in state $S$, and L2 will never get a response leading
+to a deadlock. This is illustrated in Figure \ref{lessComplicated}.
+
 \begin{figure}\small
@@
 address $a$ is executing, and
-\item $x > p.dir[c][a]$
+\item $x > y$, where $p.dir[c][a] = y$ just before sending the response
 \end{enumerate}
@@
 \begin{inv}
-If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$
-\label{cSendRespPost}
-\end{inv}
-\begin{inv}
 A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
-if $x < c.state[a]$ \label{pSendRespPre1}
+if $x < y$, where $c.state[a] = y$ just before sending the
+response\label{pSendRespPre1}
 \end{inv}
@@
 \end{inv}
-\begin{inv}
-If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$
-\label{pSendRespPost}
-\end{inv}
-\caption{Invariants for sending responses}
+\caption{Local-Properties for sending responses}
 \label{sendResp}
@@
 
+We will illustrate why Local-Properties \ref{cSendRespPre} and
+\ref{pSendRespPre2} are important. Let's say $c_1$ is in $S$ and $c_2$ is in $I$
+for address $a$ the corresponding directory states match with the actual states.
+
+In the first scenario, let $c_1$ receive a store request for address $a$. It
+sends an upgrade to $M$ request to L2. Then $c_1$ gets another load request from
+the core for a different address $a'$ which is not present in $c_1$. If $c_1$
+violates \ref{cSendRespPre} and evicts address $a$ to allocate a line for
+address $a$, then it sends a downgrade to $I$ response for address $a$ to L2.
+Meanwhile L2 has sent an upgrade to $M$ response to $c_1$. After sending the
+response, L2 receives the downgrade to $I$ response form $c_1$. So L2 now
+assumes that $c_1$ is in $I$ while in reality, $c_1$ is in $M$. This breaks
+Invariant \ref{conservative} which creates the problems described earlier. This
+is shown in Figure \ref{checkit}
+
+In the second scenario, let's say L2 violates \ref{pSendRespPre2} and
+voluntarily sends an upgrade to $M$ response for address $a$ to $c_1$.
+Meanwhile, $c_1$ receives a load request from the core for a different address
+$a'$ which is not present in $c_1$, and hence evicts $a$ to allocate a line for
+$a$. $c_1$ then gets the upgrade to $M$ response from L2 while L2 gets the
+downgrade to $I$ response, breaking Invariant \ref{conservative} again. This is
+shown in Figure \ref{checkit2}.
+
+In general, any crossing of response messages between a cache and its parent,
+like the ones shown in Figures \ref{checkit} and \ref{checkit2} results in
+violation of Invariant \ref{conservative}.
+
 \begin{figure}\small
 \begin{inv}
+If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$\label{cSendRespPost}
+\end{inv}
+\begin{inv}
+If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$\label{pSendRespPost}
+\end{inv}
+\begin{inv}
 On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
-\gets x$
-\label{pRecvResp}
+\gets x$\label{pRecvResp}
 \end{inv}
@@
 On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
-\gets x$
-\label{cRecvResp}
+\gets x$\label{cRecvResp}
 \end{inv}
-\caption{Invariants for receiving responses}
-\label{recvResp}
-\end{figure}
-
-\begin{figure}\small
 \begin{inv}
 $c.state[a]$ can change only on $c$ sending a response to or receiving a
-response from $c.parent$
-\label{cState}
+response from $c.parent$\label{cState}
 \end{inv}
@@
 $p.dir[c][a]$ can change only on $p$ sending or receiving a response from
-its child $c$
-\label{pState}
+its child $c$\label{pState}
 \end{inv}
-\caption{Invariants for state changes}
+\caption{Local-Properties for state changes}
 \label{stateChange}
@@
 \begin{inv}
-\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$
-\label{pSendReqPre}
+\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$\label{pSendReqPre}
 \end{inv}
 \begin{inv}
-\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$
-\label{cSendReqPre}
+If \Req{p}{c}{a}{x}, where $p = c.parent$, is received, if $x \ge c.state[a]$,
+then the request is dropped\label{pSendReqPost}
 \end{inv}
-\caption{Invariants for sending a request}
+\begin{inv}
+\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$\label{cSendReqPre}
+\end{inv}
+\begin{inv}
+If \Req{c}{p}{a}{x}, where $p = c.parent$, is received, it $x \le p.state[a]$,
+then the request is dropped\label{cSendReqPost}
+\end{inv}
+\caption{Local-Properties for sending and receiving a request}
 \label{sendReq}
+\end{figure}
+
+Local-Properties \ref{pSendReqPre} and \ref{pSendReqPost} are needed together
+because an incoming request is dropped if the state is already lower than or
+equal to the requested state without sending a response. So, the cache waiting
+for a response will never receive one, causing a deadlock.  Similarly,
+Local-Properties \ref{cSendReqPre} and \ref{cSendReqPost} are also needed
+together.
+
+\begin{figure}\small
+\begin{inv}
+An incoming request message from a child or the core should not block an
+incoming request message from the parent, even for a different
+address\label{cReqNoBlockPReq}
+\end{inv}
+\begin{inv}
+An incoming request messages for an address $a$ should not block an incoming
+response message for address $a$, even from a different
+source.\label{reqNoBlockResp}
+\end{inv}
+\caption{Blocking restrictions}
+\label{blocking}
 \end{figure}
@@
 \end{inv}
-\caption{Invariants for receiving a request (used by the scheduler)}
-\label{recvReq}
+\caption{Local-Properties for receiving a request (used by the scheduler)}\label{recvReq}
 \end{figure}
@@
 $a$ from any source
-\end{enumerate}
-\label{evict}
-\caption{Invariants for suspending a thread during choosing an address for eviction}
+\end{enumerate}\label{evict}
+\caption{Local-Properties for suspending a thread during choosing an address for eviction}
 \end{inv}
@@
 \begin{inv}
-Each node has 
-\begin{enumerate}
-\item at least one dedicated thread resource for handling only messages (requests or
-responses) from the parent,
-\item at least one dedicated thread resource for handling only requests from any
-child, and
-\item at least one dedicated thread resource for handling only responses (from children
-or from the parent).
-\end{enumerate}
-\label{dedicate}
+Each node has at least one dedicated thread resource for handling only messages
+(requests or responses) from the parent\label{dedicate1}
+\end{inv}
+\begin{inv}
+Each node has at least one dedicated thread resource for handling only responses
+(from children or from the parent)\label{dedicate2}
+\end{inv}
+\begin{inv}
+Each node has at least three thread resources\label{dedicate3}
 \end{inv}
@@
 
+Local-Properties \ref{evict}, \ref{pHandleReq} and \ref{cHandleReq} are needed
+because once a thread has finished up-to a point in its execution, it can not
+re-execute that portion. For instance, consider a thread handling an upgrade
+request from a child. Let's say thread $t$ has progressed to the point where it
+is simply waiting to send an upgrade response to the child. Now, if the line
+gets evicted because it was chosen for replacement, or it got downgraded by the
+parent, then thread $t$ can not re-execute its procedure again. To prevent this,
+only one thread is allowed to operate on a particular address, except in the
+case of downgrade requests from the parent. In this case, a thread is created to
+handle the downgrade request from the parent even if there is another thread
+handling a request for the same address from a child, if the latter thread is
+just waiting for a response from the parent. This is done to avoid the deadlock
+shown in the following scenario.
+
+Let's illustrate the need for Local-Properties \ref{cReqNoBlockPReq},
+\ref{dedicate1}, \ref{cReqNoBlockPReq} and \ref{dedicate2}. Let's say both $c_1$
+and $c_2$ are in state $S$ for address $a$, and the corresponding directory
+states in L2 match with their actual states. Both $c_1$ and $c_2$ get store
+requests from the processor for address $a$, and they send upgrade to $M$
+requests to L2. Let's say that L2 received the request from $c_2$ first.  L2
+sends a downgrade to $I$ request to $c1$.
+
+Let's first say Local-Property \ref{cReqNoBlockPReq} is not held, and instead some
+request from the core is blocking $c_1$ from receiving the request. The $c_1$
+will never respond back to L2. But L2 will never start handling the request from
+$c_1$ because of Local-Property \ref{cHandleReq}. This will create a deadlock.
+The same deadlock is also created when $c_1$ is out of thread resources to
+create a thread to handle the downgrade request from $L2$ (violating Local-Property
+\ref{dedicate1}).%Figure \ref{deadlockByPBlock} depicts this scenario.
+
+Let's extend this scenario further. Instead of violating Local-Properties
+\ref{cReqNoBlockPReq} or \ref{dedicate1} and $c_1$ not responding, let $c_1$
+send a downgrade to $I$ response to L2. Let's instead assume that Local-Property
+\ref{reqNoBlockResp} is violated in L2. So, the downgrade to $I$ response from
+$c_1$ is not received by L2, leading to the same deadlock. In the same
+manner, if Local-Property \ref{dedicate2} is violated, the response can not be
+handled by L2 creating the same deadlock yet again.
+
+The need for Local-Property \ref{dedicate3} can be shown easily. Let's say that a
+cache has exactly 2 thread resources, one for handling messages from parents,
+and one for handling responses from either the parent or the children. Now, if
+it receives a request from one of its children, it can not use either of the
+thread resources. Thus, at least one more thread resource is needed for
+handling requests from children. This thread resource can also be used for
+handling any type of incoming message.
+
 \begin{figure}\small
@@
 Starvation-freedom: A thread that keeps becoming ready to execute will
-eventually be scheduled to execute.
+eventually be scheduled to execute.\label{starvation}
 \end{inv}
 \end{figure}
+
+Local starvation-freedom is required in each cache node. If a node never
+executes a particular suspended thread, even if it keeps getting ready to
+execute, then the thread will not progress, leading to a deadlock.

### 2012-11-20 paper/isca40format/Theorems.tex
stuff

@@
-\section{Theorems}
-\label{sec:theorems}
+\section{Invariants}
+\label{sec:properties}
 
-If Invariants \ref{blocking} to \ref{evict} are obeyed, then we have formally
-proved that the following theorems hold. The formal proof is beyond the scope of
-the paper.
+We give a list of invariants or global properties that are needed to ensure that
+the distributed, hierarchical, MSI protocol in Figure \ref{realistic} is correct.
 
@@
 
-These theorems are needed to guarantee that Figure \ref{realistic} protocol is
-indeed correct. If, for instance, a thread wants to send a message in the
-output channel of the node and is not able to send one forever, then the thread
-can not proceed further. If a thread is waiting for a response (for a request
-previously sent), and it never gets the response, the thread can never proceed
-further. A thread can also not proceed if it's waiting for being allocated in a
-free line but never gets one because all the lines are locked up. These
-scenarios create a deadlock.
+The need for these invariants is intuitive. If, for instance, a thread wants to
+send a message in the output channel of the node and is not able to send one
+forever, then the thread can not proceed further. If a thread is waiting for a
+response (for a request previously sent), and it never gets the response, the
+thread can never proceed further. A thread can also not proceed if it's waiting
+for being allocated in a free line but never gets one because all the lines are
+locked up. These scenarios create a deadlock.
 
-We will illustrate the need for Theorems \ref{pcSame}, \ref{cpSame} and
+We will illustrate the need for Invariants \ref{pcSame}, \ref{cpSame} and
 \ref{conservative} using a 2-level of caches having two L1 caches $c_1$ and
@@
 
-Theorem \ref{pcSame} and \ref{cpSame} are important to make sure that the data
+Invariant \ref{pcSame} and \ref{cpSame} are important to make sure that the data
 transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
@@
 sends \Req{c_1}{p}{a}{M} to $p$. $p$ receives the request, but assumes that
-$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Theorem \ref{pcSame}
+$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant \ref{pcSame}
 is violated. So, it does not transfer the data to $c_1$, resulting in $c_1$
@@
 
-Theorem \ref{conservative} is important because a parent node consults its
+Invariant \ref{conservative} is important because a parent node consults its
 directory to decide the children it needs to send a downgrade request to.  Let
@@
 
-If Theorems \ref{canSend}, to \ref{cpSame} holds, then the procedures given in
+If Invariants \ref{canSend}, to \ref{cpSame} holds, then the procedures given in
 Figure \ref{realistic} can not deadlock because it will always be able to a)
@@
 sent, d) receive data when required. Procedure \uReq{} in Figure
-\ref{msi-template} relies on Theorem \ref{conservative} to maintain the
+\ref{msi-template} relies on Invariant \ref{conservative} to maintain the
 Single-writer Invariant \ref{singleWriter}.
+
+In order to guarantee these invariants, we need each node to obey certain local
+properties that we describe in the next section. Since these properties are
+localized to a node, it's easy to see that the distributed, hierarchical, MSI
+protocol we derive in Section \ref{sec:DistributedMsi} obeys these properties
+just by inspection.

### 2012-11-20 paper/isca40format/Invariants.tex
stuff

--- 2012-11-19 paper/isca40format/Invariants.tex
@@
-\section{Invariants}
-\label{sec:invariants}
+\subsection{Local Properties}
+\label{sec:localProperties}
 
@@
 
-\begin{figure}
-\begin{inv}
-An incoming request message from a child should not block an incoming request
-message from the parent, even for a different address\label{cReqNoBlockPReq}
-\end{inv}
-\begin{inv}
-An incoming request messages for an address $a$ should not block an incoming
-response message for address $a$, even from a different
-source.\label{reqNoBlockReq}
-\end{inv}
-\caption{Blocking restrictions}
-\label{blocking}
-\end{figure}
-
-\begin{figure}
-\begin{inv}
-An incoming response message for an address $a$ from a child $c$ should not be
-received before another incoming response message from the same $c$ and the
-same $a$ sent earlier has been received\label{cRespFifo}
-\end{inv}
-\begin{inv}
-An incoming request message for an address $a$ from a source $n$ should not be
-received before another incoming response message from the same $n$ and the
-same $a$ sent earlier has been received\label{reqNoOvertakeResp}
-\end{inv}
-\caption{Ordering requirements}
-\label{order}
-\end{figure}
-
-\begin{figure}
+\begin{figure}\small
 \begin{inv}
@@
 \begin{enumerate}
-\item a thread that handling a request from $c$ to $p$ for
-address $a$ is executing, and
-\item $x > p.dir[c][a]$
+\item a request \Req{c}{p}{a}{z} was received from $c$
+\item $x > y$, where $p.dir[c][a] = y$ just before sending the response
 \end{enumerate}
-\label{cSendRespPre}
-\end{inv}
-\begin{inv}
-If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$
-\label{cSendRespPost}
+\label{pSendRespPre}
 \end{inv}
@@
 A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
-if $x < c.state[a]$ \label{pSendRespPre1}
+if $x < y$, where $c.state[a] = y$ just before sending the
+response\label{pSendRespPre1}
 \end{inv}
 \begin{inv}
-If there is a thread in node $c$ that is waiting to receive a response from its
-parent $p$ for an address $a$, then response \Resp{c}{p}{a}{x} can be sent only
-on executing a thread that is handling a request from $p$ to $c$ for address
-$a$. For instance, $a$ can not be evicted from $c$, if there is a thread
-handling a request from $p$ to $c$ for address $a$.
-\label{pSendRespPre2}
+If $c$ has sent a request \Req{c}{p}{a}{x}, then $c$ can send a response
+\Resp{c}{p}{a}{y} only on receiving a request from $p$ for address $a$.
+\label{cSendRespPre2}
+\end{inv}
+%\caption{Local-Properties for sending responses}
+%\label{sendResp}
+%\end{figure}
+%
+%\begin{figure}\small
+\begin{inv}
+If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$\label{cSendRespPost}
 \end{inv}
 \begin{inv}
-If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$
-\label{pSendRespPost}
+If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$\label{pSendRespPost}
 \end{inv}
-\caption{Invariants for sending responses}
-\label{sendResp}
-\end{figure}
-
-\begin{figure}
 \begin{inv}
 On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
-\gets x$
-\label{pRecvResp}
+\gets x$\label{pRecvResp}
 \end{inv}
@@
 On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
-\gets x$
-\label{cRecvResp}
+\gets x$\label{cRecvResp}
 \end{inv}
-\caption{Invariants for receiving responses}
-\label{recvResp}
-\end{figure}
-
-\begin{figure}
 \begin{inv}
 $c.state[a]$ can change only on $c$ sending a response to or receiving a
-response from $c.parent$
-\label{cState}
+response from $c.parent$\label{cState}
 \end{inv}
@@
 $p.dir[c][a]$ can change only on $p$ sending or receiving a response from
-its child $c$
-\label{pState}
+its child $c$\label{pState}
 \end{inv}
-\caption{Invariants for state changes}
-\label{stateChange}
-\end{figure}
-
-\begin{figure}
+%\caption{Local-Properties for state changes}
+%\label{stateChange}
+%\end{figure}
+%
+%\begin{figure}\small
 \begin{inv}
-\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$
-\label{pSendReqPre}
+\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$\label{pSendReqPre}
 \end{inv}
 \begin{inv}
-\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$
-\label{cSendReqPre}
+If \Req{p}{c}{a}{x}, where $p = c.parent$, is received, if $x \ge c.state[a]$,
+then the request is dropped\label{pSendReqPost}
 \end{inv}
-\caption{Invariants for sending a request}
+\begin{inv}
+\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$\label{cSendReqPre}
+\end{inv}
+\begin{inv}
+If \Req{c}{p}{a}{x}, where $p = c.parent$, is received, it $x \le p.state[a]$,
+then the request is dropped\label{cSendReqPost}
+\end{inv}
+\begin{inv}
+A request can be sent by a cache node $n$ for address $n$ to another cache node $m$ only if
+no thread in $n$ is waiting for a response for address $a$ from $m$.
+\end{inv}
+\caption{Local-Properties for sending and receiving a request}
 \label{sendReq}
@@
 
-\begin{figure}
-\begin{inv}
-$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
-if either
-\begin{enumerate}
-\item address $a$ has been chosen for eviction by some thread and that thread
-has not yet replaced $a$, or
-\item a thread in $p$ is handling a request for address
-$a$ from any source
-\end{enumerate}
-\label{pHandleReq}
-\end{inv}
-\begin{inv}
-$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
-if either
-\begin{enumerate}
-\item address $a$ has been chosen for eviction by some thread and that thread
-has not yet replaced $a$, or
-\item a thread in $p$ is handling a request for address
-$a$ from $p$, or
-\item a thread in $p$ is handling a request for address
-$a$ from one of $p$'s children $c'$, and that thread will not send any
-(downgrade) request to any child for address $a$ anymore
-\end{enumerate}
-\label{cHandleReq}
-\end{inv}
-\caption{Invariants for receiving a request (used by the scheduler)}
-\label{recvReq}
-\end{figure}
+The Local-Properties capture the core properties of the distributed MSI
+protocol given in Section \ref{sec:distributedMsi}, and the requirements
+mentioned in Section \ref{sec:Network}.  We will use the violation of
+Local-Properties \ref{cSendRespPre} and \ref{pSendRespPre2} as examples to
+illustrate the importance of local properties.  In our usual system, let's say
+$c_1.state[a] = p.dir[c_1][a] = I$, $c_2.state[a] = p.dir[c_2][a] = I$.
 
-\begin{figure}
-\begin{inv}
-Address $a$ present in node $c$ can not be chosen for eviction by a thread
-executing in $c$ if either
-\begin{enumerate}
-\item address $a$ has been chosen for eviction by some thread and that thread
-has not yet replaced $a$, or
-\item a thread in $c$ is handling a request for address
-$a$ from any source
-\end{enumerate}
-\label{evict}
-\caption{Invariants for suspending a thread during choosing an address for eviction}
-\end{inv}
-\end{figure}
+Let $c_1$ receive a store request for address $a$. It sends an upgrade to $M$
+request to $p$. Then $c_1$ violates \ref{cSendRespPre} and evicts $a$ to
+replace it with another address. It sends a downgrade to $I$ response for
+address $a$ to $p$.  Meanwhile $p$ has sent an upgrade to $M$ response to
+$c_1$. $c_1$ receives the response and upgrades address $a$ to $M$. $p$ then
+receives the downgrade to $I$ response form $c_1$. So $p$ now assumes that
+$c_1$ is in $I$ while in reality, $c_1$ is in $M$. This breaks Invariant
+\ref{conservative} which creates the problems described earlier. This is shown
+in Figure \ref{checkit}
 
-\begin{figure}
-\begin{inv}
-Starvation-freedom: A thread that keeps becoming ready to execute will
-eventually be scheduled by the scheduler to execute.
-\end{inv}
-\end{figure}
+Let's say $p$ violates \ref{pSendRespPre2} and voluntarily sends an upgrade to
+$M$ response for address $a$ to $c_1$.  Meanwhile, $c_1$ evicts $a$ to replace
+it with another address. $c_1$ then gets the upgrade to $M$ response from $p$
+while $p$ gets the downgrade to $I$ response, breaking Invariant
+\ref{conservative} again. This is shown in Figure \ref{checkit2}.
+
+In general, any crossing of response messages between a cache and its parent,
+like the ones shown in Figures \ref{checkit} and \ref{checkit2} results in
+violation of Invariant \ref{conservative}.

### 2012-11-20 paper/isca40format/Theorems.tex
stuff

@@
 
-We give a list of invariants or global properties that are needed to ensure that
-the distributed, hierarchical, MSI protocol in Figure \ref{realistic} is correct.
+We first give a list of invariants or global properties that are needed to
+ensure that the distributed, hierarchical, MSI protocol in Figure
+\ref{realistic} is correct.
 
@@
 \begin{theorem}
-If a thread sends a request, it will eventually get a response.
+If a cache sends an upgrade to $x$ request to its parent, it will eventually
+get an upgrade to $y$ response from the parent where $y \ge x$.
 \label{willRecv}
+\end{theorem}
+
+\begin{theorem}
+If a cache sends an downgrade to $x$ request to one of its children, it will
+eventually get a downgrade to $z$ response from the child where $z \le x$.
+\label{willRecv2}
 \end{theorem}
@@
 We will illustrate the need for Invariants \ref{pcSame}, \ref{cpSame} and
-\ref{conservative} using a 2-level of caches having two L1 caches $c_1$ and
-$c_2$, and one shared L2 cache $p$.
+\ref{conservative} using the same example that we used in Section
+\ref{sec:network}.
 
@@
 $c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
-sends \Req{c_1}{p}{a}{M} to $p$. $p$ receives the request, but assumes that
+sends an upgrade to $M$ to $p$. $p$ receives the request, but assumes that
 $c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant \ref{pcSame}
@@
 $c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
-\Req{c_1}{p}{a}{M} to $p$, $p$ will assume that $c_2$'s state for address $a$
-is also $I$ and hence sends \Resp{p}{c_1}{a}{M} to $c_1$. But since the real
+an upgrade to $M$ request to $p$, $p$ will assume that $c_2$'s state for address $a$
+is also $I$ and hence send an upgrade to $M$ response to $c_1$. But since the real
 state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
@@
 
-If Invariants \ref{canSend}, to \ref{cpSame} holds, then the procedures given in
-Figure \ref{realistic} can not deadlock because it will always be able to a)
-send a request, b) send back a response, c) get back a response for a request it
-sent, d) receive data when required. Procedure \uReq{} in Figure
-\ref{msi-template} relies on Invariant \ref{conservative} to maintain the
-Single-writer Invariant \ref{singleWriter}.
+If Invariants \ref{canSend}, to \ref{cpSame} hold, then the procedures given
+in Figure \ref{realistic} can not deadlock because it will always be able to a)
+send a request, b) send back a response, c) get back a response for a request
+it sent, d) receive data when required.
 
-In order to guarantee these invariants, we need each node to obey certain local
-properties that we describe in the next section. Since these properties are
-localized to a node, it's easy to see that the distributed, hierarchical, MSI
-protocol we derive in Section \ref{sec:DistributedMsi} obeys these properties
-just by inspection.
+Procedure \uReq{} in Figure \ref{msi-template} relies on the Invariant
+\ref{conservative} being true in order to maintain the Single-writer Invariant
+\ref{singleWriter}.
+
+In procedures \uResp{}, \dResp{} and \dRespL{} (Figure \ref{realistic}), the
+data transfers rely on Invariants \ref{pcSame} and \ref{cpSame} being true in
+order to maintain the Read-from-last-writer Invariant \ref{readLast}.
+
+Thus, if we can somehow prove that Invariants \ref{canSend} to
+\ref{conservative} holds for the distributed protocol given in Figures
+\ref{msi-template}, \ref{realistic} and \ref{scheduler}, then the protocol will
+implement cache coherence correctly, without deadlocks. The formal proof of why
+the distributed protocol obeys the above invariants is beyond the scope of the
+paper. However, we will list some of the local properties that were extracted
+from the distributed protocol that was useful in proving Invariants
+\ref{canSend} to \ref{conservative}. We will also give intuition about how
+violating a particular local property can lead to deadlocks or violation of
+Invariant \ref{conservative}. It can easily be seen that the distributed
+protocol indeed satisfies these local properties. MUST REWRITE THIS WITH NETWORKS

### 2012-11-20 paper/isca40format/Introduction.tex
will be removing evict req

@@
 protocol.
+
+\begin{figure}
+\centering
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[scale=.4]{hierarchy}
+\caption{Logical topology}
+\label{hierarchy}
+\end{subfigure}
+\begin{subfigure}{.45\linewidth}
+\centering
+\includegraphics[scale=.4]{physical}
+\caption{Physical topology}
+\label{hierarchy}
+\end{subfigure}
+\caption{Logical and physical topology of cache hierarchies}
+\end{figure}
 
@@
 
-\begin{enumerate} \item Lots of concurrent events must be dealt with without a
-central authority to impose a sequence on events; \item The state on which
-transitions are based conceptually is physically distributed and cannot be
-accessed atomically; \item Physical network may block the transmission of a
-message because of lack of communication buffers or head-of-the-line blocking;
+\begin{enumerate}
+\item Lots of concurrent events must be dealt with without a central authority
+to impose a sequence on events;
+\item The state on which transitions are based conceptually is physically
+distributed and cannot be accessed atomically;
+\item Physical network may block the transmission of a message because of lack
+of communication buffers or head-of-the-line blocking;
 \item The cache or directory may reach a state in which a particular message
 cannot be handled. An incomplete set of transition rules may cause the protocol
-to deadlock.  \end{enumerate}
+to deadlock.
+\end{enumerate}
 
@@
 concrete and familiar scenarios from two-level systems (a private L1 cache for
-each processor and a memory with a directory). States $M, S$ and $I$ have their
+each processor and a shared L2 with a directory). States $M, S$ and $I$ have their
 usual meaning in MSI protocol ($M$ represents read and write permissions for an
@@
 \noindent \emph{Example 1:} A cache in state $S$ for a line gets a store
-request from the processor and sends a request for changing the line's state to
-$M$ to the directory controller. The controller may eventually send a response
-permitting the cache to go to $M$ state for that line. The response would
-signify that the states in the other caches for that line \emph{were} $I$. But
-the messages take time to propagate and the response is necessarily based on
-the old states of other caches -- the current state of another cache may not be
-$I$, violating the Single-writer-at-a-time invariant and hence breaking cache
-coherence.
+request from the processor and sends an upgrade to $M$ request to L2.  L2 may
+eventually send an upgrade to $M$ response. The response would signify that the
+states in the other caches for that line \emph{were} $I$. But the messages take
+time to propagate and the response is necessarily based on the old states of
+other caches -- the current state of another cache may be $M$, which means two
+cores can write to the same location simultaneously, violating cache coherence.
 
 \noindent \emph{Example 2:} A cache in state $S$ for a line gets a store
-request from the processor and sends a request for changing the line's state to
-$M$ to the directory controller. Instead of getting a response permitting it to
-go to $M$ state, the cache may receive a request from the directory controller
+request from the processor and sends an upgrade to $M$ request to L2. Instead
+of getting an upgrade to $M$ response, the cache may receive a request from L2
 asking this cache to go to $I$ state (invalidate request)-- this could have
-happened if the directory is serving some other cache's request to go to $M$
-state for the same line. It is not easy to determine the correct behavior of
-the former cache on receiving the invalidate request -- whether it should
+happened if the directory in L2 is serving some other cache's request to go to
+$M$ state for the same line. It is not easy to determine the correct behavior
+of the former cache on receiving the invalidate request -- whether it should
 ignore the invalidate request and wait for a response to its request or whether
@@
 \noindent \emph{Example 3:} A cache receives an invalidate request from the
-directory controller for a line that is not present in the cache. Is such a
-request even possible? The directory does not have access to the up-to-date
-state of the cache, and so can potentially make this request. How should a
-cache handle this request; should it drop such a request, or respond saying
-that it does not have the line?
+L2 for a line that is not present in the cache. Is such a request even
+possible? The directory does not have access to the up-to-date state of the
+cache, and so can potentially make this request. How should a cache handle this
+request; should it drop such a request, or respond saying that it does not have
+the line?
 
@@
 distributed, hierarchical, cache system with message-passing; 2. Specification
-of the invariants that ensure the correctness of the transformation procedure;
-and 3. Giving a provable lower bound on network virtual channels and buffers to
-avoid deadlocks; and 4. Showing that our procedure does not introduce any
+of the requirements that ensure the correctness of the generated protocol;
+3. A discussion of the invariants that simplifies the proof of correctness of
+the protocol considerably;
+4. Giving a provable lower bound on network virtual channels and buffers to
+avoid deadlocks; and 5. Showing that our procedure does not introduce any
 overhead over direct implementation of these protocols. We think that our
@@
 protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. Section
-\ref{sec:properties} gives a set of global properties or invariants that are
-needed to preserve the correctness of the transformation procedure. The
-desirability of these invariants is intuitively obvious. Section
-\ref{sec:localProperties} describes local properties that are needed to
-guarantee the desirable invariants discussed in Section \ref{sec:properties};
-it can be easily seen by inspection (because of the local nature of the
-properties) that the distributed protocol obeys these properties. Section
-\ref{sec:network} specifies the minimum virtual channel requirements to ensure
-the local properties specified in Section \ref{sec:localProperties}. We show
-that these formally verified family of protocols require no extra state bits,
-nor extra message transfers compared to other published implementations.  In
-Section \ref{sec:related}, we present some of the related work and finally in
-Section \ref{sec:conclusion}, we present some directions for the future work.
+\ref{sec:network} specifies the network, resource and other requirements that
+are needed for the distributed MSI protocol. Section \ref{sec:properties} gives
+a set of desirable global properties or invariants that are needed to easily
+establish the correctness of the distributed protocol.
+%It also lists core properties of the distributed protocol that are needed to establish the invariants.
+Section \ref{sec:implementation} specifies the minimum virtual
+channel requirements to ensure the local properties specified in Section
+\ref{sec:localProperties}. We show that these formally verified family of
+protocols require no extra state bits, nor extra message transfers compared to
+other published implementations.  In Section \ref{sec:related}, we present some
+of the related work and finally in Section \ref{sec:conclusion}, we present
+some directions for future work.

### 2012-11-20 paper/isca40format/Invariants.tex
will be removing evict req

@@
 if $x < y$, where $c.state[a] = y$ just before sending the
-response\label{pSendRespPre1}
+response\label{cSendRespPre1}
 \end{inv}
@@
 A request can be sent by a cache node $n$ for address $n$ to another cache node $m$ only if
-no thread in $n$ is waiting for a response for address $a$ from $m$.
+no $n$ is not waiting for a response for address $a$ from $m$.
 \end{inv}
-\caption{Local-Properties for sending and receiving a request}
+\caption{Local-Properties useful in proving the correctness of MSI protocol}
 \label{sendReq}
@@
 protocol given in Section \ref{sec:distributedMsi}, and the requirements
-mentioned in Section \ref{sec:Network}.  We will use the violation of
-Local-Properties \ref{cSendRespPre} and \ref{pSendRespPre2} as examples to
+mentioned in Section \ref{sec:network}. We will use the violation of
+Local-Properties \ref{pSendRespPre} and \ref{cSendRespPre2} as examples to
 illustrate the importance of local properties.  In our usual system, let's say
@@
 Let $c_1$ receive a store request for address $a$. It sends an upgrade to $M$
-request to $p$. Then $c_1$ violates \ref{cSendRespPre} and evicts $a$ to
+request to $p$. Then $c_1$ violates \ref{cSendRespPre2} and evicts $a$ to
 replace it with another address. It sends a downgrade to $I$ response for
@@
 
-Let's say $p$ violates \ref{pSendRespPre2} and voluntarily sends an upgrade to
+Let's say $p$ violates \ref{pSendRespPre} and voluntarily sends an upgrade to
 $M$ response for address $a$ to $c_1$.  Meanwhile, $c_1$ evicts $a$ to replace

### 2012-11-21 paper/isca40format/Introduction.tex
stuff

@@
 Large cache-coherent shared-memory multicores are likely to use protocols which
-are directory-based, hierarchical, and implemented using message passing.  Even
+are directory-based, hierarchical, and implemented using packet-switched networks.  Even
 though such protocols have been around for two decades (see, for example,
@@
 \caption{Physical topology}
-\label{hierarchy}
+\label{physical}
 \end{subfigure}
@@
 have stale data for $a$ (Figure \ref{hierarchy}). Physically the system looks
-like a collection of caches connected to a message passing network (Figure
+like a collection of caches connected to a packet switched network (Figure
 \ref{physical}). Caches typically contain data and state for each cache line,
@@
 Every cache, unless it is an L1 cache, has a directory which contains
-information about the state of its children caches. The cache state changes to
+information about the state of its children caches. The cache state-changes to
 maintain coherence are orchestrated by passing requests and responses between
@@
 which lock the bus till all the required state transitions are done thereby
-ensuring atomicity of state transitions, in a message-passing protocol,
+ensuring atomicity of state transitions, in a distributed protocol,
 requests and responses are split leading to transient states. This creates
@@
 Several factors contribute to the difficulty of designing protocols for such
-distributed, hierarchical and message-passing systems:
+distributed, hierarchical and packet-switched systems:
 
@@
 \item The cache or directory may reach a state in which a particular message
-cannot be handled. An incomplete set of transition rules may cause the protocol
+cannot be handled, \ie, an incomplete set of transition rules may cause the protocol
 to deadlock.
@@
 \noindent \emph{Example 1:} A cache in state $S$ for a line gets a store
-request from the processor and sends an upgrade to $M$ request to L2.  L2 may
-eventually send an upgrade to $M$ response. The response would signify that the
+request from the processor and sends an upgrade-to-$M$ request to L2.  L2 may
+eventually send an upgrade-to-$M$ response. The response would signify that the
 states in the other caches for that line \emph{were} $I$. But the messages take
@@
 \noindent \emph{Example 2:} A cache in state $S$ for a line gets a store
-request from the processor and sends an upgrade to $M$ request to L2. Instead
-of getting an upgrade to $M$ response, the cache may receive a request from L2
+request from the processor and sends an upgrade-to-$M$ request to L2. Instead
+of getting an upgrade-to-$M$ response, the cache may receive a request from L2
 asking this cache to go to $I$ state (invalidate request)-- this could have
@@
 coherence protocols for such systems. The user specifies a protocol for a cache
-hierarchy as sequential methods to process requests. There is no message
-passing so the protocol procedure can simultaneously or \emph{atomically}
+hierarchy as sequential methods to process requests. It is not distributed, so the protocol procedure can simultaneously or \emph{atomically}
 examine and change the state of all the caches in the system. We then
-systematically transform such a protocol description into a distributed,
-message-passing protocol. The resulting protocol is presented as a collection
+systematically transform such a protocol description into a distributed protocol. The resulting protocol is presented as a collection
 of cooperating sequential methods which operate on ``cache objects''.  These
@@
 The main contributions of this paper are: 1. An automatic procedure to refine a
-centralized MSI protocol without message passing into protocols for
-distributed, hierarchical, cache system with message-passing; 2. Specification
+centralized MSI protocol into protocols for
+distributed, hierarchical, cache system with packet-switching; 2. Specification
 of the requirements that ensure the correctness of the generated protocol;
@@
 the protocol considerably;
-4. Giving a provable lower bound on network virtual channels and buffers to
-avoid deadlocks; and 5. Showing that our procedure does not introduce any
-overhead over direct implementation of these protocols. We think that our
-method of designing protocols substantially reduces the required effort in
-implementing provably correct protocols.
+4. Giving a provable lower bound on the number of network virtual channels and
+buffers to avoid deadlocks; and 5. Showing that our procedure does not
+introduce any overhead over direct implementation of these protocols. We think
+that our method of designing protocols substantially reduces the required
+effort in implementing provably correct protocols.
 
@@
 \ref{sec:network} specifies the network, resource and other requirements that
-are needed for the distributed MSI protocol. Section \ref{sec:properties} gives
-a set of desirable global properties or invariants that are needed to easily
-establish the correctness of the distributed protocol.
+are needed for the distributed MSI protocol. We also give the minimum virtual
+channel requirements to ensure the requirements specified in Section
+\ref{sec:network}. We show that these formally verified family of protocols
+require no extra state bits, nor extra message transfers compared to other
+published implementations.  Section \ref{sec:properties} gives a set of
+desirable global properties or invariants that are needed to easily establish
+the correctness of the distributed protocol.
 %It also lists core properties of the distributed protocol that are needed to establish the invariants.
-Section \ref{sec:implementation} specifies the minimum virtual
-channel requirements to ensure the local properties specified in Section
-\ref{sec:localProperties}. We show that these formally verified family of
-protocols require no extra state bits, nor extra message transfers compared to
-other published implementations.  In Section \ref{sec:related}, we present some
-of the related work and finally in Section \ref{sec:conclusion}, we present
-some directions for future work.
+In Section \ref{sec:related}, we present some of the related work and finally
+in Section \ref{sec:conclusion}, we present some directions for future work.

### 2012-11-21 paper/isca40format/Invariants.tex
stuff

@@
 \begin{enumerate}
-\item a request \Req{c}{p}{a}{z} was received from $c$
+\item a request \Req{c}{p}{a}{z} was received from $c$,
+\item $p$ is not waiting for any response for address $a$ from $c$, and
 \item $x > y$, where $p.dir[c][a] = y$ just before sending the response
@@
 no $n$ is not waiting for a response for address $a$ from $m$.
+\label{nodoublereq}
 \end{inv}
@@
 The Local-Properties capture the core properties of the distributed MSI
-protocol given in Section \ref{sec:distributedMsi}, and the requirements
+protocol given in Section \ref{sec:DistributedMsi}, and the requirements
 mentioned in Section \ref{sec:network}. We will use the violation of
@@
 illustrate the importance of local properties.  In our usual system, let's say
-$c_1.state[a] = p.dir[c_1][a] = I$, $c_2.state[a] = p.dir[c_2][a] = I$.
+$c_1.state[a] = p.dir[c_1][a] = S$, $c_2.state[a] = p.dir[c_2][a] = I$.
 
-Let $c_1$ receive a store request for address $a$. It sends an upgrade to $M$
+\floatstyle{plain}
+\restylefloat{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=0.34]{checkit}
+\caption{Effect of violating Invariant \ref{cSendRespPre2} (child evicts a ``pending'' line)}
+\label{checkit}
+\end{figure}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+Let $c_1$ receive a store request for address $a$. It sends an upgrade-to-$M$
 request to $p$. Then $c_1$ violates \ref{cSendRespPre2} and evicts $a$ to
 replace it with another address. It sends a downgrade to $I$ response for
-address $a$ to $p$.  Meanwhile $p$ has sent an upgrade to $M$ response to
+address $a$ to $p$.  Meanwhile $p$ has sent an upgrade-to-$M$ response to
 $c_1$. $c_1$ receives the response and upgrades address $a$ to $M$. $p$ then
@@
 
-Let's say $p$ violates \ref{pSendRespPre} and voluntarily sends an upgrade to
-$M$ response for address $a$ to $c_1$.  Meanwhile, $c_1$ evicts $a$ to replace
-it with another address. $c_1$ then gets the upgrade to $M$ response from $p$
-while $p$ gets the downgrade to $I$ response, breaking Invariant
+\floatstyle{plain}
+\restylefloat{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=0.34]{checkit2}
+\caption{Effect of violating Invariant \ref{pSendRespPre} (parent sends a voluntary response)}
+\label{checkit2}
+\end{figure}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+Let's say $p$ violates \ref{pSendRespPre} and voluntarily sends an
+upgrade-to-$M$ response for address $a$ to $c_1$.  Meanwhile, $c_1$ evicts $a$
+to replace it with another address. $c_1$ then gets the upgrade-to-$M$ response
+from $p$ while $p$ gets the downgrade to $I$ response, breaking Invariant
 \ref{conservative} again. This is shown in Figure \ref{checkit2}.

### 2012-11-21 paper/isca40format/Msi.tex
stuff

--- 2012-11-18 paper/isca40format/Msi.tex
@@
 \section{MSI protocol}
-\label{sec:msi}
+\label{sec:GlobalMsi}
 
-\newcommand{\printall}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
-\newcommand{\Req}[4]{\printall{Req}{#1}{#2}{#3}{#4}}
-\newcommand{\Resp}[4]{\printall{Resp}{#1}{#2}{#3}{#4}}
-\newcommand{\Data}[5]{\printall{Data}{#1}{#2}{#3}{#4}{#5}}
+The caches are logically organized in the form of a tree
+(Figure~\ref{hierarchy}) where the leaves represent L1 caches and the root is the last level cache (LLC).
+Each L1 is connected to exactly one core and the LLC is connected to the memory.
+The cache hierarchy is
+inclusive, \ie if an address is present in a cache $c$, then it must also be present in its parent $c.parent$. Note that by address we mean cache-line address, not a byte address or a word address.
 
-\newcommand{\lett}{\textbf{let}}
-\newcommand{\send}{\textbf{send}}
-\newcommand{\receive}{\textbf{receive}}
-\newcommand{\pop}{\textbf{pop}}
-\newcommand{\remove}{\textbf{remove}}
-\newcommand{\assert}{\textbf{assert}}
-\newcommand{\call}{\textbf{call}}
-\newcommand{\dReq}{\textsc{handleDownReq}}
-\newcommand{\uReq}{\textsc{handleUpReq}}
-\newcommand{\dReqL}{\textsc{sendDownReq}}
-\newcommand{\uReqL}{\textsc{sendUpReq}}
-\newcommand{\dResp}{\textsc{sendDownResp}}
-\newcommand{\uResp}{\textsc{sendUpResp}}
+Each cache stores the coherence state and the data for each address that it
+contains. 
+% Figure \ref{notation} gives the notation used to represent all this information. 
+For an L1 cache, the coherence state for an address denotes the
+(usual) permission that a core connected to the L1 cache has for the address: $M
+= \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$ relation for coherence
+state values based on the permissions the states represent, that is, $I < S < M$. 
+For an internal or non-L1 cache, the meaning of the coherence states has to be generalized.
+For cache $c$, $c.state[a] = I$ means that each cache in the entire sub-tree rooted at $c$ will have coherence state $I$ for address $a$.
+$c.state[a] = S$ means that the children of $c$ can either be in $S$ or $I$ state. Finally, $c.state[a] = M$ means that 
+at most one of c's children can be in state $M$. If the state bits are manipulated properly then the following invariants will hold.
 
-Let us walk through our framework using the example of a hierarchical MSI
-protocol.
+\begin{theorem}
+\textbf{Single-writer Invariant}: When an L1 cache has permission to modify data for
+an address, no other L1 cache has read or write permission for that address. 
+\label{singleWriter}
+\end{theorem}
 
-\begin{figure}
-\begin{tabularx}{\linewidth}{|cX|}
-\hline
-$c.state[a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for address $a$\\
-$c.data[a]$ & data in cache $c$ for address $a$\\
-%$c.dir[n][a]$ & $c$'s notion of the coherence state of $n$ for address $a$\\
-\hline
-\end{tabularx}
-\caption{Notations}
-\label{table:lineinfo}
-\end{figure}
+\begin{theorem}
+\textbf{Read-from-last-writer Invariant}: When a core reads data for an address
+in L1, it gets the value that was either last stored in that address by some
+core, or if no core has stored any value in that address previously, the
+initial value from the memory.
+\label{lastRead}
+\end{theorem}
 
-\begin{wrapfigure}{r}{.4\linewidth}
-\centering
-\begin{tabular}{|c|ccc|}
-\hline
-& $M$ & $S$ & $I$\\
-\hline
-$M$ & $=$ & $>$ & $>$\\
-$S$ & $<$ & $=$ & $>$\\
-$I$ & $<$ & $<$ & $=$\\
-\hline
-\end{tabular}
-\caption{$<$ and other relations for MSI states}
-\label{msi<}
-\end{wrapfigure}
-
-Each cache stores the information given in Figure \ref{table:lineinfo}. The
-coherence state for an address denotes the (usual) permission that a cache has
-for the address: $M = \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$
-relation for coherence state values based on the permissions the states
-represent. This is shown in Figure \ref{msi<}. State $I$ is akin to address $a$
-not present in the cache.
-
-\begin{wrapfigure}{r}{.46\linewidth}
-\begin{subfigure}{.27\linewidth}
+\begin{wrapfigure}{r}{.6\linewidth}
+\begin{subfigure}{.55\linewidth}
 \centering
@@
 \end{tabular}
-\caption{$toCompatible$}
+\caption{$maxCompatible$}
 \label{toCompat}
 \end{subfigure}
-~~~~~~~~~~~~~~
-\begin{subfigure}{.21\linewidth}
+\begin{subfigure}{.24\linewidth}
 \centering
@@
 \hline
-$M$ & $\checkmark$\\
-$S$ & $\times$\\
-$I$ & $\times$\\
+$M$ & $True$\\
+$S$ & $False$\\
+$I$ & $False$\\
 \hline
@@
 \end{subfigure}
-\caption{$toCompatible$ and $isModified$ mappings}
-\label{fig:msimap}
+\caption{MSI auxiliary functions ($I < S < M$)}
+\label{funcs}
 \end{wrapfigure}
+In Figure \ref{funcs} we define some auxiliary functions $maxCompatible(x)$ and $isModified(x)$ which are useful in describing protocols to maintain the invariant.
 
-If a cache $c$ receives a request for an address $a$ from the processor, it can
-service the request straightaway if it has enough permissions for address $a$.
-Otherwise, it has to \emph{upgrade} to a state $x$, which has enough
-permissions. The cache can not upgrade its state arbitrarily, it has to ensure
-that such a state change does not violate the single-writer property (Invariant
-\label{singleWriter}), by \emph{downgrading}, \ie lowering the states of other
-caches appropriately. We define a $toCompatible(x)$ mapping (Figure
-~\ref{toCompat}), which specifies the highest state a cache can be, if another
-cache is in state $x$ for the same address. We will also define a
-$isModified(x)$ mapping to specify if an address in state $x$ in a cache can
-have data different from what's present in the memory.
+The MSI protocol maintains the invariant essentially by a cache requesting its parent a state upgrade, \ie $I$ to $S$, $I$ to $M$ or $S$ to $M$.
+The parent grants this request by downgrading its other children as necessary or requesting an upgrade of its own state from its parent.
+The data is also transferred from one cache to another as appropriate.
+In our protocol descriptions each cache in the hierarchy is treated as a C++/Java like object with methods to process upgrade and downgrade requests.
+A special method called \textsc{coreReq} is provided for L1's so that the core can initiate a load or a store request as shown in Figure \ref{coreHandle}.
+The L1 cache checks if the request
+can already be serviced, if not, it invokes its upgrade request handler \uReq{},
+which eventually upgrades the state of the requested address. Finally, in the case 
+of a load request, the data is sent back to the core, and in the case of a store, the
+cache line is updated. Note that the core supplies words or long-words while the
+cache line sizes are typically 64 bytes; only the appropriate word in the cache
+line is updated in case of a store.
 
@@
 \begin{figure}
+\small
+\begin{algorithmic}
+\Proc {\textsc{coreReq}}{$core, a, op, data$}
+  \If {$op = Ld$}
+    \State \textbf{if} ($this.state[a] < S$)
+    \State \;\;\;\; \call{} $this.$\uReq($core, a, S$);
+    \State \send{} $this.data[a]$ (to $core$);
+  \ElsIf {$op = St$}
+    \State \textbf{if} ($this.state[a] < M$)
+    \State \;\;\;\; \call{} $this.$\uReq($core, a, M$);
+    \State $update(this.data[a], data)$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\caption{L1 cache $this$ handling a load or a store request sent by $core$ by
+calling the request handler}
+\label{coreHandle}
+\end{figure}
+
+\floatstyle{boxed}
+\restylefloat{figure}
+
+\begin{figure}
+\small
 
@@
 \begin{algorithmic}
-\Proc {\uReq}{$c , p, a, x$}
-  \If {$a$ is not present in $p$}
+\Proc {\uReq}{$c , a, x$}
+    \If {$a$ is not present in $this$}
+      \If {$this$ has no space for $a$}
         \State Choose address $a'$ for eviction;
-    \ForAll {$c' \in p.children$}
-      \State \call{} \dReq($p , c', a', I$);
-    \EndFor
-    \State \call{} \dResp($p , p.parent, a', I$);
-    \State Replace $a'$ with $a$ (initializing $p.state[a]$ to $I$);
+        \State \textbf{forall} ($c' \in this.children$)
+        \State \;\;\;\; \textbf{if} ($c'.state[a] > I$)
+        \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', I$);
+        \State \textbf{if} ($this.state[a'] > I$)
+        \State \;\;\;\; \call{} $this.$\dResp{}($this.parent, a', I$);
       \EndIf
-  \ForAll {$c' \in p.children, s.t., c' \neq c$}
-    \State \dReq($p , c', a, toCompatible(x)$);
-  \EndFor
-  \State \uReqL($p , p.parent, a, x$);
-  \State \uResp($p , c, a, x$);
+      \State Allocate a line for $a$;
+    \EndIf
+    \State // $a$ is already present in $this$
+    \State \textbf{forall} ($c' \in this.children$)
+    \State \;\;\;\; \textbf{if} ($c'.state[a] > maxCompatible(x)$)
+    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', maxCompatible(x)$);
+    \State \call{} $this.$\uReqL{}($this.parent, a, x$);
+  \State \textbf{if} ($c \neq core$)
+  \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
 \EndProc
@@
 \end{boxedminipage}
-\caption{Handling upgrade to $x$ request from $c$ to $p$ for address $a$}
-\label{uReq}
+\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
+\label{atomicUReq1}
 \end{subfigure}
@@
 \begin{algorithmic}
-\Proc {\dReq}{$p , c, a, x$}
-  \ForAll {$c' \in c.children$}
-    \State \dReqL($c , c', a, x$);
-  \EndFor
-  \State \dResp($c , p, a, x$);
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Handling downgrade to $x$ request from $p$ to $c$ for address $a$}
-\label{dReq}
-\end{subfigure}
-
-\caption{MSI protocol}
-\label{msi}
-\end{figure}
-
-\begin{figure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\uReqL}{$c , p, a, x$}
-  \State \call{} \uReq($c, p, a, x$);
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending upgrade to $x$ request from $c$ to $p$ for address $a$}
-\label{uReq}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\dReq}{$p , c, a, x$}
-  \State \call{} \dReq($p, c, a, x$);
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending downgrade to $x$ request from $p$ to $c$ for address $a$}
-\label{dReq}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\uResp}{$p , c, a, x$}
-  \If {$c.state[a] < x$}
-    \If {$c.state[a] = I$}
-      \State $c.data[a] \gets p.data[a]$;
-    \EndIf
-    \State $c.state[a] \gets x$;
+\Proc {\dReq}{$p , a, x$}
+  \If {$this.state[a] > x$}
+    \State \textbf{forall} ($c' \in this.children$)
+    \State \;\;\;\; \textbf{if} (c'.state[a] > x)
+    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a, x$);
+    \State \call{} $this.$\dResp($p, a, x$);
   \EndIf
@@
 \end{boxedminipage}
-\caption{Sending upgrade to $x$ response from $p$ to $c$ for address $a$}
-\label{uResp}
+\caption{$this$ cache's downgrade request handler -- request from $this$'
+parent, $p$, to upgrade $p$'s state to $x$ for address $a$ }
+\label{atomicDReq1}
 \end{subfigure}
 
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
+\caption{MSI protocol}
+\label{msi-template}
+\end{figure}
+
+We will first describe the upgrade and downgrade methods as sequential procedures which can access and modify the state of any cache in the system (see
+Figure \ref{msi-template}).  
+When the upgrade request handler \uReq{} for a cache $this$ is invoked by a
+cache $c$ to go to a particular state $x$, $this$ first makes sure that it has space to process it.
+This means that if $a$ is not present in $this$, it allocates a new line for that address, sometimes by evicting another line.
+This requires downgrading the status of the evicted address in all its children.
+Next, $this$ ensures that its
+other children are ``compatible'' with $c$ being in state $x$ for the requested
+address.
+For example, for $c$ to go to state $M$, all the children of $this$, except for $c$ must be in state $I$.
+If some children are not in state $I$ then the parent (\ie $this$) must downgrade them to go into state $I$.
+In the protocol, this is accomplished by the parent asking the child to downgrade their state to $maxCompatible(x)$.
+If the child's state is such that $isModified(state)$ is true, then it will contain the latest version of the data for address $a$, and hence this data has to be written back into the parent's cache.
+If $this$' state is less than the requested state it invokes
+its parent's handler for an upgrade request to $x$. Finally, when state of $this$ is $x$ and all its children are compatible with $x$, $this$ can upgrade
+$c$'s state, and send the data if $c$'s state was $I$ previously. This is shown
+in Figure \ref{atomicUReq1}.
+
+Any cache except the LLC can get a downgrade request from its parent. Figure \ref{atomicDReq1} defines this downgrade method \dReq{} where the cache
+ensures that each of its children's state is not higher than $x$, by invoking
+their respective downgrade methods. When all the children have downgraded, the
+cache downgrades its own state to $x$, and sends data back to the parent, if it
+has been modified.
+
+\begin{figure}
+\small
+
 \begin{algorithmic}
-\Proc {\dResp}{$c , p, a, x$}
-  \If {$c.state[a] > x$}
-    \If {$isModified(c.state[a])$}
-      \State $p.data[a] \gets c.data[a]$;
+\State // Upgrading $this$ by invoking parent's upgrade handler
+\Proc {\uReqL}{$p, a, x$}
+  \State \textbf{if} ($this$ is LLC \&\& $this.state[a] = I$) \bopen
+  \State \;\;\;\; $this.data[a] \gets memory[a]$;
+  \State \;\;\;\; $this.state[a] \gets M$;
+  \State \bclose \textbf{else}
+  \State \;\;\;\; \call{} $p.$\uReq($this, a, x$);  
+\EndProc
+
+\State // Invoking the child's downgrade handler
+\Proc {\dReqL}{$c , a, x$}
+  \State \call{} $c.$\dReq($this, a, x$);
+\EndProc
+
+\State // $this$ finally upgrades the state (and if necessary, data) of its child
+\Proc {\uResp}{$c, a, x$}
+  \If {$c.state[a] = I$}
+    \State $c.data[a] \gets this.data[a]$;
   \EndIf
   \State $c.state[a] \gets x$;
+\EndProc
+
+\State // $this$ downgrades its own state and updates the parent's data
+\Proc {\dResp}{$p, a, x$}
+  \If {$isModified(c.state[a])$}
+    \State $p.data[a] \gets this.data[a]$;
+    \State // $p.data$ refers to $memory$ if $this$ is LLC
   \EndIf
+  \State $this.state[a] \gets x$;
+\EndProc
+
+\State // In case of LLC, $this$ reads data from memory
+\Proc {\sendMem{}}{$a$}
 \EndProc
 \end{algorithmic}
-\end{boxedminipage}
-\caption{Sending downgrade to $x$ response from $c$ to $p$ for address $a$}
-\label{dResp}
-\end{subfigure}
 
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\uResp}{$p , core, a, x$}
-  \State $c.data[a] \gets p.data[a]$;
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending upgrade to $x$ response from $p$ to $core$ for address $a$}
-\label{uResp}
-\end{subfigure}
-
-\caption{Atomic MSI}
+\caption{Auxiliary methods used in the MSI protocol of Figure \ref{msi-template}.
+In all the methods $p$ is the parent of $this$, $c$ is a child of $this$, and the methods work on address $a$ to change its state to $x$}
 \label{atomic}
@@
 
-\begin{figure}
+The methods to handle upgrade and downgrade requests in turn use 5 auxiliary methods defined in Figure \ref{atomic}.
+These methods are very simple and could have been inlined.
+However, they do recursively call $this$' parent's \uReq{} or $this$' children's \uReq{}.
+We have chosen not to inline them because these functions encapsulate all accesses to other caches when we systematically transform this protocol for distributed implementation.
+Note however that \uReq{} and \dReq{} read the states of its children.
+We will make those reads local by keeping a directory as discussed in the next section.
 
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\uReqL}{$c , p, a, x$}
-  \State \send{} \Req{c}{p}{a}{x};
-  \State \pop{} \Resp{p}{c}{a}{z};
-  \If {$c.state[a] = I$}
-    \State \pop{} \Data{p}{c}{a}{d};
-    \State $c.data[a] \gets d$;
-  \EndIf
-  \State $c.state[a] \gets z$;
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending upgrade to $x$ request from $c$ to $p$ for address $a$}
-\label{uReqLocal}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\dReqL}{$p , c, a, x$}
-  \State \send{} \Req{p}{c}{a}{x};
-  \State \pop{} \Resp{c}{p}{a}{z};
-  \If {$isModified(p.dir[c][a])$}
-    \State \pop{} \Data{c}{p}{a}{d};
-    \State $p.data[a] \gets d$;
-  \EndIf
-  \State $p.dir[c][a] \gets z$;
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending downgrade to $x$ request from $p$ to $c$ for address $a$}
-\label{dResp}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\uResp}{$p , c, a, x$}
-  \If {$p.dir[c][a] < x$}
-    \If {$p.dir[c][a] = I$}
-      \State \send{} \Data{p}{c}{a}{p.data[a]};
-    \EndIf
-    \State $p.dir[c][a] \gets x$;
-  \EndIf
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending upgrade to $x$ response from $p$ to $c$ for address $a$}
-\label{uResp}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\dResp}{$c , p, a, x$}
-  \If {$c.state[a] > x$}
-    \If {$isModified(c.state[a])$}
-      \State \send{} \Data{c}{p}{a}{c.data[a]};
-    \EndIf
-    \State $c.state[a] \gets x$;
-  \EndIf
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending downgrade to $x$ response from $c$ to $p$ for address $a$}
-\label{dResp}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
-\Proc {\uResp}{$p , core, a, x$}
-  \State \send{} \Data{p}{c}{a}{p.data[a]};
-\EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{Sending upgrade to $x$ response from $p$ to $core$ for address $a$}
-\label{uResp}
-\end{subfigure}
-
-\caption{Realistic MSI protocol}
-\label{realistic}
-\end{figure}
+All lines in the LLC are always in $M$ state.

### 2012-11-21 paper/isca40format/Theorems.tex
stuff

@@
 \begin{theorem}
-If a cache sends an upgrade to $x$ request to its parent, it will eventually
+If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
 get an upgrade to $y$ response from the parent where $y \ge x$.
@@
 \begin{theorem}
-If a cache sends an downgrade to $x$ request to one of its children, it will
-eventually get a downgrade to $z$ response from the child where $z \le x$.
+If a cache sends an downgrade-to-$x$ request to one of its children, it will
+eventually get a downgrade-to-$z$ response from the child where $z \le x$.
 \label{willRecv2}
@@
 $c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
-sends an upgrade to $M$ to $p$. $p$ receives the request, but assumes that
+sends an upgrade-to-$M$ to $p$. $p$ receives the request, but assumes that
 $c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant \ref{pcSame}
@@
 $c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
-an upgrade to $M$ request to $p$, $p$ will assume that $c_2$'s state for address $a$
-is also $I$ and hence send an upgrade to $M$ response to $c_1$. But since the real
+an upgrade-to-$M$ request to $p$, $p$ will assume that $c_2$'s state for address $a$
+is also $I$ and hence send an upgrade-to-$M$ response to $c_1$. But since the real
 state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
@@
 data transfers rely on Invariants \ref{pcSame} and \ref{cpSame} being true in
-order to maintain the Read-from-last-writer Invariant \ref{readLast}.
+order to maintain the Read-from-last-writer Invariant \ref{lastRead}.
 

### 2012-11-21 paper/isca40format/DistributedMsi.tex
before arvind is going to re-org

\section{Distributed MSI}
\label{sec:DistributedMsi}

In order to transfer the Global MSI protocol into a distribute MSI protocol, we
need to make sure that a method can only read and write the state of the cache
with which the method is associated, and it affects the state of other caches
only by sending messages. In order to read the state of a child's cache we
maintain a shadow state of in the form of a directory. Thus the coherence state
of a cache $c$, in addition to the $c.state[a]$ field, also contains a
directory field containing the shadow state of each child's state field, \ie,
$c.dir[c'][a]$, for child $c'$.

We extend our sequential language for writing methods with two communication
primitives: \send{} $(Req|Resp|Data)\langle c_1 \rightarrow c_2, a,
(x|data)\rangle$ and \receive{} $(Req|Resp|Data)(\langle c_1 \rightarrow c_2,
a, (x|data)\rangle$ where caches $c_1$ and $c_2$ must satisfy the parent-child
relation. Both of these primitives are \emph{blocking}, that is, \send{} may
block because the network is clogged, \ie, has no buffer space to hold a
message and a \receive{} may block because the message has not arrived or cache
has no resources to process it. These primitives make our sequential language
methods ``suspensive'. Request or response messages have very different
characteristics and a data message is conceptually always associated with some
response message. We think of each \textsc{coreReq} as launching a thread in
L1. This thread in turn may get processed, send request or response messages,
may get suspended and eventually die. Thus potentially every request message
can launch a new thread in the cache that receives it. A response message,
however, can only die or generate more response messages; it can never generate
a request message. However a response message can also block waiting for the
corresponding data message. Each thread has an address associated with it which
corresponds to the message that created it.

In order to write the distributed MSI protocol using suspensive threads, we will
maintain the following sets of invariants. In the next section, we will show in
the requirements in the network and the bounds on the number of virtual channels
to avoid head of line blocking. We will also show the other resource
requirements, as well as the properties of the thread scheduler that will ensure
these invariants there.

\subsection{Relation between directory and state of the children's caches}
\begin{theorem}
Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
p.dir[c][a] \ge c.state[a]$.  \label{conservative}
\end{theorem}

This ensures that if a cache makes decision about compatibility of states of
its children using the local directory information, then it does not violate the
compatibility of the children's real states.

\subsection{Network properties}
Regarding the \send{} and \receive{} operations, \ie the network, the following
properties can assume the following:
\begin{theorem}
It a thread has to send a message, it will eventually be able to send the
message.
\label{canSend}
\end{theorem}

\begin{theorem}
If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
get an upgrade-to-$y$ response from the parent where $y \ge x$.
\label{willRecv}
\end{theorem}

\begin{theorem}
If a cache sends an downgrade-to-$x$ request to one of its children, it will
eventually get a downgrade-to-$y$ response from the child where $y \le x$.
\label{willRecv2}
\end{theorem}

\subsection{Eviction}
If handling of a request requires eviction of another cache line, then too the
thread handling the request may get suspended. This would happen if all the
lines that can be allocated for this request may accessed by other suspended
threads. But we still assume the following invariant.

\begin{theorem}
If a thread is waiting for a line to be allocated, \ie for any address to become
available for eviction, then some address will eventually become available for
eviction
\label{evictDead}
\end{theorem}

It this is not true, a thread that is waiting for being allocated a free line
will never proceed leading to a deadlock.

\subsection{Data transfer}
Regarding the responses received, the designer can assume the following 2
invariants. These invariants are used to wait for receiving data messages
appropriately, as only some state transfers also cause data transfers.
We will show the exact need for these invariants later.

\begin{theorem}
If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
just before $p$ sends the response is the same as $c.state[a]$ just before $c$
receives the response.
\label{pcSame}
\end{theorem}

\begin{theorem}
If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
receives the response.
\label{cpSame}
\end{theorem}

\subsection{Need for invariants}
The need for these network invariants is intuitive. If, for instance, a thread wants to
send a message in the output channel of the node and is not able to send one
forever, then the thread can not proceed further. If a thread is waiting for a
response (for a request previously sent), and it never gets the response, the
thread can never proceed further. These scenarios create a deadlock.

We will illustrate the need for Invariants \ref{pcSame}, \ref{cpSame} and
\ref{conservative} using an example of a 2-level system containing two L1
caches $c_1$ and $c_2$ with a shared L2 cache $p$.

Invariant \ref{pcSame} and \ref{cpSame} are important to make sure that the
data transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
$c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
sends an upgrade-to-$M$ to $p$. $p$ receives the request, but assumes that
$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant
\ref{pcSame} is violated. So, it does not transfer the data to $c_1$, resulting
in $c_1$ waiting for data, and hence deadlock. A dual condition can be
constructed similarly, in which $c_1$ downgrades its state for address $a$ from
$S$ to $I$, and the parent assumes that $c_1$ is in state $M$, and keeps
waiting for data.

Invariant \ref{conservative} is important because a parent node consults its
directory to decide the children it needs to send a downgrade request to.  Let
$p.dir[c_1][a] = S$, $c_1.state[a] = S$, $p.dir[c_2][a] = I$ but
$c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
an upgrade-to-$M$ request to $p$, $p$ will assume that $c_2$'s state for address $a$
is also $I$ and hence send an upgrade-to-$M$ response to $c_1$. But since the real
state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
\ref{singleWriter}.

Thus, if all the invariants above (\ref{conservative}, to \ref{cpSame}) hold,
then the threads given in Figure \ref{realistic} can not deadlock because it
will always be able to a) send a request, b) send back a response, c) get back
a response for a request it sent, d) receive data when required. Though the
threads executing in a cache are actually suspensive, these invariants give an
illusion of the threads being non-suspensive barring the following exceptions.
Here $this$ represents the cache under under consideration.

\begin{enumerate}
\item $this.dir[c][a]$ can keep decreasing while a thread corresponding to
address $a$ is executing.
\item $this.dir[c][a']$ can keep decreasing while a thread which is evicting
line $a'$ is executing.
\item If a thread receives responses from the parent, then $this.state[a]$ can
decrease between the operation of sending a request to the parent and the
operation of receiving a response from the parent.
\end{enumerate}

As can be seen, these interactions are minimal, and can be ignored since they
affect neither the handling of upgrade nor downgrade requests. Hence, the
correctness of the protocol can be argued in terms of a single threaded,
non-suspensive, execution model. The formal proof for this, as well as the
invariants stated below is beyond the scope of this paper.

In order to design the distributed MSI protocol, the following considerations
must be kept in mind.
\begin{enumerate}
\item There are 3 kinds of incoming messages to be handled: request from that parent,
requests from children and responses from children.
\item A downgrade-to-$x$ message might be received from the parent even if the
state of the cache is $ \le x$. This has to be handled.
\item A parent sends a response to the child only when a request is received. But a child can send an voluntary/unsolicited response to the parent during evictions.
\item A request should be sent to the parent only if the current cache's state
is lower than what's desired to handle its incoming request. Similarly, a
request should sent to its child only if the directory's notion of that cache's
state is higher than what's desired.
\item A second request is not sent for the same address from the same cache to the same
destination, until a response is received for the first request.
\item A cache can not evict a line if it is currently waiting for a response from the parent for the address corresponding to the line.
\end{enumerate}

%We always reserve resources, \eg, cache line slot for an address before sending
%an upgrade request for that address.

The handling of messages follow the exact same algorithm as given for the
global MSI protocol in Figure \ref{msi-template} with the caveat that instead
of reading a child's state, the cache $this$ can only read its own directory.
So, it suffices to simply replace $c.state[a]$ by $this.dir[c][a]$ every time
the state of child $c$ is read by $this$.

As we mentioned above, a new handler has to be written for handling voluntary
responses. This is shown in Figure \ref{msi-unsolicited}.

\begin{figure}
\small
\begin{algorithmic}
\Proc {\dRespL}{$c , a, x$}
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
\EndProc
\end{algorithmic}
\caption{Handling an voluntary response message from a child $c$, sent if the
child evicts line $a$ and downgrades the line to $x$}
\label{msi-unsolicited}
\end{figure}

\begin{figure}
\small

\begin{algorithmic}
\State // Upgrading $this$ by sending request to parent and getting a response
\Proc {\uReqL}{$p, a, x$}
  \If {$this$ is LLC \&\& $this.state[a] = I$}
    \State \send{} \Req{this}{Memory}{a}{M};
    \State \receive{} \Data{Memory}{this}{a}{d};
    \State $this.data[a] \gets d$;
    \State $this.state[a] \gets M$;
  \ElsIf
  \State \send{} \Req{this}{p}{a}{x};
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
  \EndIf
\EndProc
\State // Downgrading $this$' child by sending request to child and getting a
response
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  \While {$this.dir[c][a] > x$}
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
  \EndWhile
\EndProc
\State // $this$ finally upgrading the directory state of its child and sending
a response to its child
\Proc {\uResp}{$c, a, x$}
  \State \textbf{if} ($this.dir[c][a] = I$)
  \State \;\;\;\; \send{} \Data{this}{c}{a}{this.data[a]};
  \State $this.dir[c][a] \gets x$;
\EndProc
\State // $this$ finally downgrading its state and sending a response to its
parent
\Proc {\dResp}{$p, a, x$}
  \State \textbf{if} ($isModified(this.state[a])$)
  \State \;\;\;\; \send{} \Data{this}{p}{a}{this.data[a]};
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}
\caption{Distributed MSI protocol. In all the methods $p$ is the parent of
$this$, $c$ is a child of $this$, and the methods work on address $a$ to change
its state (or directory) to $x$. This function only shows the auxiliary
functions and the handler for voluntary responses from children. The handler
for requests are exactly the same as in Figure \ref{msi-template}, with every
occurence of $c'.state[a]$ replaced by $this.dir[c'][a]$}
\label{realistic}
\end{figure}

### 2012-11-21 paper/isca40format/Implementation.tex
before arvind is going to re-org

\section{Implementation overhead}
\label{sec:implementation}
In this section we show how the requirements presented in Section
\ref{sec:network} can be implemented in a realistic system.

\subsubsection{Implementation using virtual channels:}

Let's first analyse ways to implement the ordering requirements (Figure
\ref{order}) and the blocking restrictions (Figure \ref{blocking}) in an actual
packet-switched network using virtual channels.

Consider $2N$ channels, where $N + 1$ is the number of levels in the hierarchy
(counting both the L1 caches and the LLC). Let's say each of these channels are
indexed by a pair $ReqChannel\langle x, direction \rangle$, $x$ denotes a level
in the hierarchy and $direction$ can either be $ToC$ or $ToP$.  Virtual
channels $1, ToC$ and $N + 1, ToP$ do not exist, giving a total of $2N$
channels. We will now map the requests sent from a cache to another node to
these channels as follows: a request sent from a cache in level $x$ to its
parent uses channel $ReqChannel\langle x, ToP \rangle$, and a request sent from
a cache in level $x$ to any of its children uses channel $ReqChannel\langle x,
ToC \rangle$.

This virtual channel mapping is sufficient for handling any request message in
the system, the reason being as follows: A request message can potentially
produce more request messages. However, a request message from a cache node in
level $x$ to a cache node in level $y$ will not produce any more request
messages sent from any cache node in level $x$ to any cache node in level $y$. A
downgrade request from the parent only produces more downgrade requests to its
children, recursively, so the above invariant is maintained. Similarly, an
upgrade request $R$ from a child $c$ to a parent $p$ produces more upgrade
requests from the parent to its own parent, recursively, and more downgrade
requests going from cache nodes at a higher level to a cache node at a lower
level. But no request message is sent from the level of child $c$ to the level
of parent $p$ to handle request $R$. This mapping also ensures that Requirement
\ref{cReqNoBlockPReq} is met.

Since responses from a child of a cache node should not be blocked at the cache
node by any incoming requests (Requirement\ref{reqNoBlockResp}), the responses
from a cache node at a lower level (children) to a cache node at a higher level
(parent) should not use the same share use any of the request channels. Since a
response is always removed by the system, just one channel, $RespChannel$, is
sufficient for transmitting responses from any child to its parent, in the
whole system.

Since responses from the parent $p$ of a cache node $c$ should not be blocked
at the cache node by a request from $c$'s child $c'$ (Requirement
\ref{reqBlockResp2}), these responses can not use the same channel as that used
by a request from $c'$ to $c$. However, they can use the same virtual channel
used by the requests from $p$ to $c$, $ReqChannel\langle x, ToC \rangle$, where
$x$ is the level of $p$. This is because no blocking restrictions of Figure
\ref{blocking} are violated.

\subsubsection{Ensuring the ordering requirements}
There are no ordering requirements between requests sent from the same source
to the same destination according to Figure \ref{order}.  However, requests
should not overtake responses to the same address, because of Requirement
\ref{reqNoOvertakeResp}. Requests and responses from a child to its parent come
on different virtual channels. Hence, maintaining Requirement
\ref{reqNoOvertakeResp} for messages from a child to its parent is difficult,
since it's difficult to guarantee point-to-point FIFO ordering the messages
come on different virtual channels.

Point-to-point FIFO ordering for messages from a child to its parent can be
ensured by adding more information to the messages rather than relying on the
underlying network. Each message for address $a$ from child $c$ to its parent
$p$ will contain an additional tag which specifies the state the child was in
at the time of sending the message. The new message becomes \Req{c}{p}{a}{x
\rightarrow y} or \Resp{c}{p}{a}{x \rightarrow y}, where $x$ specifies the
state of the cache at the time the request was sent. At the parent, a thread
for handling this message will be created, and this message will be removed,
only after $p.dir[c][a]$ becomes less than or equal to $x$, where $x$ is the
state of the child at the time the message was sent. The intuition behind why
this works is as follows (we do not give the formal proof here). Requests can
not be sent from a child when there is a pending response from the parent
. Similarly, a response can not be sent from
a child when there is a pending response from the parent. So, a response from the parent can never be in flight
when a request or a response from the child is in flight, since the
parent only sends responses on requests from the child . So the directory state of the parent at the time of
receiving a message can never be higher than the state of the child just before
sending the message, if the message did not overtake a previous response.

One issue with this for responses is that if only one virtual channel is used
for responses from any child to its parent throughout the system is as follows.
A response sent later for an address from a child to its parent can block a
response for the same address from the same child to its parent sent earlier.
This violates Requirement \ref{cRespFifo}. If two separate virtual channel are
used, one for downgrade-to-$S$ responses ($RespChannel_S$) and one for
downgrade-to$I$ responses ($RespChannel_I$), then using the response sent later
will not block an earlier response.

The same technique can not be used to guarantee point-to-point FIFO ordering
for transmitting messages (both requests and responses) from a parent to a
child; the intuition is that requests from the parent and a response from the
child can both be in flight at the same time since the child can voluntarily
downgrade due to evictions.  If the channel can not guarantee point-to-point
FIFO ordering, then an ACK must be sent by the child on receiving any response
from the parent to guarantee Requirements \ref{cRespFifo} and
\ref{reqNoOvertakeResp}. This ACK can use either of the channels meant for
responses from a child to its parent ($RespChannel_I$ or $RespChannel_S$).  The
protocol must change slightly; the child has to send an ACK after receiving
responses, and the parent has to wait for the ACK before freeing the thread.

This gives a total of $2N + 2$ virtual channels where $N + 1$ is the number of
levels in the cache hierarchy with the following ordering constraints:
(FIFO-ordered or ack-ordered) $ReqChannel\langle x, ToC \rangle$ for requests and
responses from a node in level $x$ to its child, unordered $ReqChannel\langle x,
ToP \rangle$ for requests from a node in level $x$ to its parent, unordered
$RespChannel_I$ for downgrade-to-$I$ responses from any node to its parent and
finally, unordered $RespChannel_S$ for downgrade-to-$S$ responses from any node
to its parent.

\subsubsection{Bits required for the directory:}
\begin{wrapfigure}{r}{.4\linewidth}
\begin{tabularx}{\linewidth}{|c|X|}
\hline
State & Directory \\
\hline
$M$ & one $M$, rest $I$\\
$M$ & some $S$, rest $I$\\
$S$ & some $S$, rest $I$\\
$I$ & all $I$\\
\hline
\end{tabularx}
\caption{Possible values of coherence state and directory for each cache line}
\label{encoding}
\end{wrapfigure}
The directory of a node store its version of the state for each of the node's
children. However, one can use the fact that for a node $n$, if $n.dir[c][a] =
M$, then $\forall c' \neq c, n.dir[c'][a] = I$. This comes from Invariants
\ref{singleWriter} and \ref{conservative}. Thus, storing just the index of the
cache is sufficient if $n.dir[c][a] = M$ for some cache $c$. Figure
\ref{encoding} shows the possible values of the directory and coherence state
for each line in the cache. It can clearly be seen that this requires
$max(\lceil log_2(n)\rceil, n) + 2$ bits per line, for the most efficient
encoding.

\subsubsection{MSHRs to implement threads}
The threads we describe essentially represent state stored in Miss Status Handling Registers (MSHRs). To
satisfy the requirements of Figure \ref{dedicated}, there must at least be 3
MSHR registers, in which one can be used only for messages from the parent, one
only for responses (both from children and the parent) and one general purpose
MSHR. We used the terms threads to describe our protocol because we give the
illusion of only one thread executing in the system though the threads are
actually suspensive. Hence the software terminology fits our model quite well.

### 2012-11-21 paper/isca40format/Introduction.tex
before arvind is going to re-org

@@
 though such protocols have been around for two decades (see, for example,
-DASH\cite{DASH}, Alewife\cite{Anant}, ...) correct design and implementation of
+DASH\cite{DASH}, Alewife\cite{Anant}) correct design and implementation of
 such protocols remains a challenge. It is difficult to convince oneself that
@@
 4. Giving a provable lower bound on the number of network virtual channels and
-buffers to avoid deadlocks; and 5. Showing that our procedure does not
+other resources to avoid deadlocks; and 5. Showing that our procedure does not
 introduce any overhead over direct implementation of these protocols. We think
@@
 channel requirements to ensure the requirements specified in Section
-\ref{sec:network}. We show that these formally verified family of protocols
+\ref{sec:network}. In Section \ref{sec:implementation}, we show that these formally verified family of protocols
 require no extra state bits, nor extra message transfers compared to other
-published implementations.  Section \ref{sec:properties} gives a set of
-desirable global properties or invariants that are needed to easily establish
-the correctness of the distributed protocol.
+published implementations. Section \ref{sec:properties} gives the local
+properties of the distributed protocol that we rely on to easily establish its
+correctness, and show how it can be generalized to design other protocols like MOSI.
 %It also lists core properties of the distributed protocol that are needed to establish the invariants.

### 2012-11-21 paper/isca40format/Invariants.tex
before arvind is going to re-org

@@
-\subsection{Local Properties}
-\label{sec:localProperties}
+\section{Local properties that ensure global invariants}
+\label{sec:properties}
+
+In this section, we will show some of the core properties that are extracted
+from the requirements spelt out in Section \ref{sec:network} as well as the
+implementation of the distributed protocol's auxiliary functions given in Figure
+\ref{realistic}. These properties can be independently used to prove the
+invariants presented in Section \ref{sec:DistributedMsi} (Invariants
+\ref{conservative} to \ref{cpSame}). We do not present the proof here, but give
+some intuition. Figure \ref{sendReq} lists these local properties.
 
@@
 
-The Local-Properties capture the core properties of the distributed MSI
-protocol given in Section \ref{sec:DistributedMsi}, and the requirements
-mentioned in Section \ref{sec:network}. We will use the violation of
-Local-Properties \ref{pSendRespPre} and \ref{cSendRespPre2} as examples to
-illustrate the importance of local properties.  In our usual system, let's say
-$c_1.state[a] = p.dir[c_1][a] = S$, $c_2.state[a] = p.dir[c_2][a] = I$.
+ We will use the violation of Local-Properties \ref{pSendRespPre} and
+\ref{cSendRespPre2} as examples to illustrate the importance of local
+properties.  In our usual system, let's say $c_1.state[a] = p.dir[c_1][a] = S$,
+$c_2.state[a] = p.dir[c_2][a] = I$.
 
@@
 \centering
-\includegraphics[scale=0.34]{checkit}
-\caption{Effect of violating Invariant \ref{cSendRespPre2} (child evicts a ``pending'' line)}
+\includegraphics[scale=0.4]{checkit}
+\caption{Effect of violating Local-Property \ref{cSendRespPre2} (child evicts a ``pending'' line)}
 \label{checkit}
@@
 \centering
-\includegraphics[scale=0.34]{checkit2}
-\caption{Effect of violating Invariant \ref{pSendRespPre} (parent sends a voluntary response)}
+\includegraphics[scale=0.4]{checkit2}
+\caption{Effect of violating Local-Property \ref{pSendRespPre} (parent sends a voluntary response)}
 \label{checkit2}
@@
 
-In general, any crossing of response messages between a cache and its parent,
-like the ones shown in Figures \ref{checkit} and \ref{checkit2} results in
-violation of Invariant \ref{conservative}.
+In general, any if response messages from both a cache to its parent and from
+the parent to the same cache are in flight together, like the scenarios in
+Figures \ref{checkit} and \ref{checkit2} Invariant \ref{conservative} will be
+violated.
+
+\begin{figure}
+\begin{subfigure}{.25\linewidth}
+\centering
+\begin{tabular}{|ccc|}
+\hline
+$M$ & $\rightarrow$ & $I$\\
+$O$ & $\rightarrow$ & $S$\\
+$S$ & $\rightarrow$ & $O$\\
+$I$ & $\rightarrow$ & $I$\\
+\hline
+\end{tabular}
+\subcaption*{$toCompatible$}
+\end{subfigure}
+\begin{subfigure}{.25\linewidth}
+\centering
+\begin{tabular}{|c|c|}
+\hline
+$M$ & $\checkmark$\\
+$O$ & $\checkmark$\\
+$S$ & $\times$\\
+$I$ & $\times$\\
+\hline
+\end{tabular}
+\subcaption*{$dataToUpper$}
+\end{subfigure}
+\begin{subfigure}{.48\linewidth}
+\centering
+\begin{tabular}{|c|cccc|}
+\hline
+& $M$ & $O$ & $S$ & $I$\\
+\hline
+$M$ & $=$ & $>$ & $>$ & $>$\\
+$O$ & $<$ & $=$ & $>$ & $>$\\
+$S$ & $<$ & $<$ & $=$ & $>$\\
+$I$ & $<$ & $<$ & $<$ & $=$\\
+\hline
+\end{tabular}
+\subcaption*{$<$ (and other) relations}
+\end{subfigure}
+\caption{MOSI protocol mappings and $<$ relations}
+\label{mosi}
+\end{figure}
+
+These properties do not require messages to physically come from the sources as
+indicated by the requests or responses -- some other cache node can forward
+these messages. These can be used to generalize our procedure to
+cache-intervention based protocols like MOSI. We generalize the states of MOSI
+using the $<$ relation and auxiliary functions $maxCompatible$ and $isModified$,
+as shown in Figure \ref{mosi}.
+
+In a hierarchical setting for the MOSI protocol, any cache can send data
+with the response granting permissions to the requesting L1 cache, instead of
+going through the hierarchy.  On a load request, L1 cache $c$ sends an
+upgrade-to-$O$ request, instead of an upgrade-to-$S$ request. The request keeps
+going up the hierarchy, sending an upgrade-to-$O$ request to each of the parent
+on the way, till it reaches a cache $p$ which has that line in $M$ or $O$
+state. If some other child is in $O$ or $M$ state, this node sends a
+upgrade-$c$-to-$O$ in addition to a downgrade-to-$S$ request to that cache.
+This upgrade-$c$-to-$O$ + downgrade-to-$S$ keeps propogating down the tree till
+it reaches a node $n$ whose children are all in a state $< O$. $n$ then
+downgrades its state, forwards the upgrade-to-$O$ response to $c$ and
+simultaneously sends a fwd-and-downgrade-to-$S$ response to its parent, which
+keeps forwarding that till it reaches $p$. If $c$ gets a store request, it will
+send an upgrade-to-$M$ request, and $p$ would send a downgrade-to-$I$ request
+instead, to its children.
+
+The caches encountered on the way from $n$ to $p$ should handle an extra
+scenario where it gets a downgrade response, without a forward response.In this
+case, the data and permission forwarding has to be done from this cache to $c$.
+It is the designer's responsibility to ensure that this case is handled.
+
+This protocol obeys all the local properties and the requirements mentioned in
+this section except that a request sent to a cache might overtake a response
+forwarded to that cache. To avoid this, once the forwarding message is
+delivered to the intended destination ($c$ in our case), then $c$ sends back an
+ACK. $c$'s parent is allowed to to start sending downgrade requests for that
+address only after receiving this ACK. This ACK is sent back to each of the
+parents till it reaches $p$, which allows each of these caches to send
+downgrade requests for that address towards $c$.  This satisfies all the
+requirements of Section \ref{sec:network} and the local properties given in
+Figure \ref{sendReq}, thereby automatically being correct. The reasoning for
+its correctness gets localized to checking if the ordering Requirement
+\ref{reqNoOvertakeResp} is maintained.

### 2012-11-21 paper/isca40format/Mine.tex
before arvind is going to re-org

\section{Distributed MSI}
\label{sec:DistributedMsi}

In order to transfer the Global MSI protocol into a distribute MSI protocol, we
need to make sure that a method can only read and write the state of the cache
with which the method is associated, and it affects the state of other caches
only by sending messages. In order to read the state of a child's cache we
maintain a shadow state of in the form of a directory. Thus the coherence state
of a cache $c$, in addition to the $c.state[a]$ field, also contains a
directory field containing the shadow state of each child's state field, \ie,
$c.dir[c'][a]$, for child $c'$.

We extend our sequential language for writing methods with two communication
primitives: \send{} $(Req|Resp|Data)\langle c_1 \rightarrow c_2, a,
(x|data)\rangle$ and \receive{} $(Req|Resp|Data)(\langle c_1 \rightarrow c_2,
a, (x|data)\rangle$ where caches $c_1$ and $c_2$ must satisfy the parent-child
relation. Both of these primitives are \emph{blocking}, that is, \send{} may
block because the network is clogged, \ie, has no buffer space to hold a
message and a \receive{} may block because the message has not arrived or cache
has no resources to process it. These primitives make our sequential language
methods ``suspensive'. Request or response messages have very different
characteristics and a data message is conceptually always associated with some
response message. We think of each \textsc{coreReq} as launching a thread in
L1. This thread in turn may get processed, send request or response messages,
may get suspended and eventually die. Thus potentially every request message
can launch a new thread in the cache that receives it. A response message,
however, can only die or generate more response messages; it can never generate
a request message. However a response message can also block waiting for the
corresponding data message. Each message and thread has an address associated
with it.

Though the threads are suspensive, the designer is given an illusion of the
threads being non-suspensive barring a few exceptions. In particular, the
\send{} and \receive{} operations can be considered as non-blocking operations
while designing the handlers for requests and responses. The exceptions that the
designer must be aware of which violates the illusion of non-suspensive threads
is as follows: ($this$ represents the cache under under consideration)

\begin{enumerate}
\item $this.dir[c][a]$ can keep decreasing while a thread corresponding to
address $a$ is executing.
\item $this.dir[c][a']$ can keep decreasing while a thread which is evicting
line $a'$ is executing.
\item If a thread receives responses from the parent, then $this.state[a]$ can
decrease between the operation of sending a request to the parent and the
operation of receiving a response from the parent.
\end{enumerate}

As can be seen, these interactions are minimal, and hence the correctness of
the protocol can be argued in terms of a single threaded, non-suspensive, execution model.
The formal proof for this, as well as the invariants stated below is beyond the
scope of this paper.

The designer must also be aware of the following:
\begin{enumerate}
\item There are 3 kinds of incoming messages to be handled: request from that parent,
requests from children and responses from children.
\item A downgrade-to-$x$ message might be received from the parent even if the
state of the cache is $ \le x$. This has to be handled.
\end{enumerate}

The designer can assume the following invariants regarding the directory and the state
of the children's caches:
\begin{theorem}
Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
p.dir[c][a] \ge c.state[a]$.  \label{conservative}
\end{theorem}

The reason why the designer can think a thread being non-suspensive because of
the following 4 invariants.

Regarding the \send{} and \receive{} operations, \ie the network, the designer
can assume the following:
\begin{theorem}
It a thread has to send a message, it will eventually be able to send the
message.
\label{canSend}
\end{theorem}

\begin{theorem}
If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
get an upgrade-to-$y$ response from the parent where $y \ge x$.
\label{willRecv}
\end{theorem}

\begin{theorem}
If a cache sends an downgrade-to-$x$ request to one of its children, it will
eventually get a downgrade-to-$y$ response from the child where $y \le x$.
\label{willRecv2}
\end{theorem}

The following invariant ensures that a line can always be obtained for
replacement .
\begin{theorem}
If a thread is waiting for a line to be allocated, \ie for any address to become
available for eviction, then some address will eventually become available for
eviction
\label{evictDead}
\end{theorem}

All the complexity of suspending a thread, scheduling a thread, waking up a
thread, \etc are hidden from the designer.

Regarding the responses received, the designer can assume the following 2
invariants:

\begin{theorem}
If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
just before $p$ sends the response is the same as $c.state[a]$ just before $c$
receives the response.
\label{pcSame}
\end{theorem}

\begin{theorem}
If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
receives the response.
\label{cpSame}
\end{theorem}

These invariants can be used by the designer to wait for receiving data messages
appropriately.

Our protocol also obeys the following properties:

We always reserve resources, \eg, cache line slot for an address before sending
an upgrade request for that address.
A second request is not sent for the same address from the same cache to the same
destination, until a response is received for the first request.

After the distributed protocol is designed, we
will show the bounds on the number of virtual channels to avoid head of the
line blocking and the minimum thread resource requirements to avoid deadlocks.

Figure \ref{msi-template} shows how the request handler methods \uReq{} and
\dReq{} can be refactored to make it amenable to distributed implementation. The
upgrade and downgrade request handlers are not allowed to invoke other cache's
upgrade or downgrade request handlers, or change other cache's states. Instead,
the same calls are made via 5 more local methods listed in Figure \ref{atomic}.
These methods in turn invoke the other caches upgrade and downgrade request
handlers, or change the states of other caches.

\begin{figure}
\small

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReqL}{$p, a, x$}
  \State \send{} \Req{this}{p}{a}{x};
  \While {$this.state[a] < x$}
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
  \EndWhile
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Upgrading address $a$ of $this$ to $x$, by sending a request to
$this$'s parent $p$}
\label{uReqLocal}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  \While {$this.dir[c][a] > x$}
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
  \EndWhile
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Downgrading address $a$ of $this$'s child $c$ to $x$, by sending a
request to $c$}
\label{dReqLocal}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uResp}{$c, a, x$}
  \State \textbf{if} ($this.dir[c][a] = I$)
  \State \;\;\;\; \send{} \Data{this}{c}{a}{this.data[a]};
  \State $this.dir[c][a] \gets x$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ finally upgrading address $a$ of its child $c$ to $x$, sending a
response to $c$ notifying of the upgrade and sending data to $c$ if necessary}
\label{uResp1}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dResp}{$p, a, x$}
  \State \textbf{if} ($isModified(this.state[a])$)
  \State \;\;\;\; \send{} \Data{this}{p}{a}{this.data[a]};
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ finally downgrading its state for address $a$ to $x$, sending a
response to $c$ notifying of the downgrade and sending data to $c$ if necessary}
\label{dResp1}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dRespL}{$c , a, x$}
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ handling unsolicited downgrade to $x$ response from $c$ for
address $a$. This happens if $c$ evicted address $a$}
\label{dRespL1}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\sendMem{}}{$a$}
  \State \send{} \Req{this}{Memory}{a}{M};
  \State \receive{} \Data{Memory}{this}{a}{d};
  \State $this.data[a] \gets d$;
  \State $this.state[a] \gets M$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Last-Level-Cache $this$ sending and receivging data from memory for
address $a$}
\label{sendMem2}
\end{subfigure}

\caption{Non atomic implementation of MSI protocol}
\label{realistic}
\end{figure}

\subsection{Caches as a distributed system of multi-threaded execution engines}

The memory hierarchy in a processor can be specified as a collection of nodes
(Figure \ref{distNodes}). Each node has local state associated with it. The
state may contain the cache and the directory among other book-keeping
information. A node also has multiple input and multiple output channels
through which it can send and receive messages. Most of the channels are
connected to a network, which delivers messages between nodes
%; the message delivery order is left unspecified
. Some of the channels may not be connected to the network; they represent
inputs and outputs to the memory system, \viz the requests from and responses to
cores.

The behavior of each node can be best described as a system of suspendable local
threads, with a scheduler local to that node. A thread can be in one of the two
states: \emph{executing} or \emph{suspended}. At a given time, either one
thread or the scheduler will be executing. When the scheduler is executing,
it does one of the following: a) picks one of suspended threads
which has become ready to execute due to changed conditions or b) creates a new
thread to handle an incoming message. An executing thread can read the local
state of the node, send messages to output channels, receive messages from
input channels, \etc{}% or spawn more threads . The input and output channels are
blocking, so if a thread tries to send to a full output channel or receive from
an empty input channel, the thread would not be able to proceed. The thread can
also fail to proceed due to some condition in the local state of the node. In
all these cases, the thread gets suspended, giving back control to the
scheduler. It remains suspended till the required conditions are met (for
instance, the output channel becomes free, the input channel receives a
pertinent message, or other local state changes), Once the required conditions
for a suspended thread are met the scheduler may pick it for execution. The
execution is resumed from the exact point where it was suspended. Once the
thread finishes its execution, all the book-keeping resources associated with
that thread is freed up and reclaimed.

Incoming messages remain in the channel until they are explicitly removed by
the thread dealing with that message. If the scheduler sees an incoming message
that requires a new thread to be created for handling, and if the required
book-keeping resources are available, then a thread is created.
%Different threads require different amounts of book-keeping resources.
If the
required book-keeping resources are not available, the scheduler might pick
another suspended thread which is ready to execute, or if that's not possible,
spin-waits.  The threads and incoming messages have a memory address associated
with them. For an incoming message that does not require a new thread to be
created, its address is used by the scheduler to decide if a suspended thread
is ready to execute. Similarly, the address associated with a thread that has
just finished executing is used to determine if the local state of the node
associated with that address has changed so that a suspended thread associated
with that address can resume execution.

Note that the thread analogy we described above has nothing to do with software
threads. We are describing the behavior of a cache controller using the familiar
notion of suspendable threads. The threads essentially correspond to the various
requests being handled by a cache controller's Miss Status Handler Register
(MSHR).

\begin{figure}
\begin{tabularx}{\linewidth}{|l|X|}
\hline
\send{} $m$ & Send a message $m$ to the node's output channel. This action
blocks till the channel has space to send.\\
\receive{} $m$ & Receive and remove a message $m$ from the node's input channel.
This action blocks till a message is received.\\
$x \gets value$ & Local state $x$ of the node gets updates to $value$.\\
%\call{} $f(args)$ & Execute \textbf{procedure} $f$ with $args$ arguments.\\
%\start{} $f(args)$ & Start a new thread that executes \textbf{procedure} $f$
%with $args$ arguments. This action is performed only by the scheduler.\\
\resume{} $t$ & Resume thread $t$ from where it stopped execution. This action
is performed only by the scheduler\\
\hline
\end{tabularx}
\caption{Actions used for describing a node's behavior}
\label{actions}
\end{figure}

When a new message arrives in an input channel that requires creation of a
thread, the scheduler first removes the message from the channel and then
creates a thread that starts executing the \procedure{} specific to the type of
the incoming message. Figure \ref{actions} show the notations we use to
describe the actions performed by an executing thread. Procedures can call
other procedures but they can not be recursive. We usually denote the execution
of actions \receive{} and \send{} using their verb forms, for instance,
,``receiving'' or ``sending''. We also denote the thread created by the
scheduler to handle an incoming message as the thread ``handling'' that
message.

### 2012-11-21 paper/isca40format/MyMsi.tex
before arvind is going to re-org

\section{MSI protocol}
\label{sec:GlobalMsi}

Let us walk through our framework using the example of a hierarchical MSI
protocol.

\floatstyle{plain}
\restylefloat{figure}

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$c.state[a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for address $a$\\
$c.data[a]$ & data in cache $c$ for address $a$\\
%$p.dir[n][a]$ & $p$'s notion of the coherence state of $d$ for address $a$. This is meaningful only if $p$ is not an L1 cache\\
$memory[a]$ & data in memory for address $a$\\
\hline
\end{tabularx}
\caption{Notations for information associated with an address in a cache}
\label{notation}
\end{figure}

\begin{figure}
\begin{subfigure}{.43\linewidth}
\centering
\begin{tabular}{|c|ccc|}
\hline
& $M$ & $S$ & $I$\\
\hline
$M$ & $=$ & $>$ & $>$\\
$S$ & $<$ & $=$ & $>$\\
$I$ & $<$ & $<$ & $=$\\
\hline
\end{tabular}
\caption{$<$ and other relations for states}
\label{msi<}
\end{subfigure}
\begin{subfigure}{.31\linewidth}
\centering
\begin{tabular}{|ccc|}
\hline
$M$ & $\rightarrow$ & $I$\\
$S$ & $\rightarrow$ & $S$\\
$I$ & $\rightarrow$ & $I$\\
\hline
\end{tabular}
\caption{$toCompatible$}
\label{toCompat}
\end{subfigure}
\begin{subfigure}{.22\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $\checkmark$\\
$S$ & $\times$\\
$I$ & $\times$\\
\hline
\end{tabular}
\caption{$isModified$}
\label{isModified}
\end{subfigure}
\caption{MSI state abstraction functions}
\label{funcs}
\end{figure}

The caches are logically organized in the form of a tree
(Figure~\ref{hierarchy}). The leaves of this tree represent the L1 caches,
with each L1 cache connected to exactly one core. The root of the tree is the
Last-Level-Cache, and its connected to the memory. The cache hierarchy is
inclusive, \ie if an address is present in a cache $c$, then it's also present it its
parent $c.parent$. Note that by address we mean cache-line address, not a byte
address or a word address.

Each cache stores the coherence state and the data for each address that it
contains.
%A cache which is not L1 (\ie an internal node in the tree of caches)
%also stores a directory containing its notion of the coherence states of its
%children.
Figure \ref{notation} gives the notation used to represent all this
information. For an L1 cache, the coherence state for an address denotes the
(usual) permission that a core connected to the L1 cache has for the address: $M
= \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$ relation for coherence
state values based on the permissions the states represent. This is shown in
Figure \ref{msi<}.
%When an address $a$ not present in the L1 cache, then the
%core associated with that L1 can neither read nor write that address, so it's
%equivalent to the state of that address being $I$.

For a non L1 cache, the meaning of coherence states is different. For an
internal cache node $n$, if $n.state[a] = I$, then, the entire sub-tree rooted at $n$
will have coherence state $I$ for address $a$. In particular, no L1 cache in
this subtree will have any permission for $a$. If $n.state[a] = S$, then its
children can either be in $S$ or $I$ state. Finally, if $n.state[a] = M$, then
at most one of its children can be in $M$ state. This guarantees that at most
one of the L1 caches can have write permissions, ensuring the following
invariant:

\begin{theorem}
\textbf{Single-writer Invariant}: When a cache has permission to modify data for
an address, no other cache has read or write permission for that
address.
\label{singleWriter}
\end{theorem}

We define a $toCompatible(x)$ mapping (Figure ~\ref{toCompat}), which specifies
the highest state a cache can be to not violate the Single-writer Invariant, if
another sibling cache is in state $x$ for the same address.

The coherence states also denote the consistency of data in a cache with respect
to its parent. If a cache (both L1 and non-L1) is in state $I$ for an address $a$,
then it does not have the data corresponding to that address. If it is in state
$S$, then its data is consistent with that of its parent. Finally, if it is in
state $M$, then its data is not consistent with that of its parent. If none of
its children are in state $M$, then the data in this cache is the most
up-to-date value. But if one of its children $c$ is in the $M$ state, then the
data in this cache is stale -- the most up-to-date value is present in the
subtree rooted at $c$.

We define a $isModified(x)$ mapping to specify if an address in state $x$ in a
cache can have data different from what's present in its parent. The mappings
$toCompatible$ and $isModified$ abstract away the semantics of the states so
that our framework can be extended to other protocols like MOSI/MOESI.

If a cache having no permission for an address requires read or write
permissions for that address, then the most up-to-date value must be supplied,
\ie:

\begin{theorem}
\textbf{Read-from-last-writer}: When a cache reads data for an address, it reads
the last modified version of the data for the address. If the data for an
address has not been modified, the cache reads the data from memory.
\label{lastRead}
\end{theorem}

Invariants \ref{singleWriter} and \ref{lastRead} together ensure cache coherence
\cite{Patterson}.

We will first describe an atomic version of the hierarchical MSI protocol. Each
of the caches in the hierarchy can be treated like C++/Java objects.
\floatstyle{boxed}
\restylefloat{figure}

\begin{figure}
\small
\begin{algorithmic}
\Proc {\textsc{coreReq}}{$core, a, op, data$}
  \If {$op = Ld$}
    \State \textbf{if} ($this.state[a] < S$)
    \State \;\;\;\; \call{} $this.$\uReq($core, a, S$);
    \State \send{} $this.data[a]$;
  \ElsIf {$op = St$}
    \State \textbf{if} ($this.state[a] < M$)
    \State \;\;\;\; \call{} $this.$\uReq($core, a, M$);
    \State $update(this.data[a], data)$;
  \EndIf
\EndProc
\end{algorithmic}
\caption{L1 cache handling a load or a store request sent by the core by
calling the request handler}
\label{coreHandle}
\end{figure}

In the atomic protocol, a core invokes the L1 cache's request handler
\textsc{coreReq} (Figure \ref{coreHandle}). The L1 cache checks if the request
can already be serviced, if not, it invokes its upgrade request handler \uReq{},
which upgrades the state of the requested address. Finally, in the case of a
load request, the data is sent back to the core, and in the case of a store, the
cache line is updated. Note that the core supplies words or long-words while the
cache line sizes are typically 64 bytes; only the appropriate word in the cache
line is updated in case of a store.

\begin{figure}
\small

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReq}{$c , a, x$}
  \If {$c.state[a] < x$}
    \If {$a$ is not present in $this$}
      \If {$this$ has no space for $a$}
        \State Choose address $a'$ for eviction;
        \State \textbf{forall} ($c' \in this.children$)
        \State \;\;\;\; \textbf{if} ($c'.state[a] > I$)
        \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', I$);
        \State \textbf{if} ($this.state[a'] > I$)
        \State \;\;\;\; \call{} $this.$\dResp{}($this.parent, a', I$);
      \EndIf
      \State Allocate a line for $a$;
    \EndIf
    \State // $a$ is already present in $this$
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} ($c'.state[a] > toCompatible(x)$)
    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a', toCompatible(x)$);
    \State \textbf{if} ($this$ is Last-Level-Cache \&\& $this.state[a] = I$)
    \State \;\;\;\; \call{} $this$.\sendMem{}($a$);
    \State \textbf{else}
    \State \;\;\;\; \call{} $this.$\uReqL{}($this.parent, a, x$);
  \State \textbf{if} ($c \neq core$)
  \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
\label{atomicUReq1}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReq}{$p , a, x$}
  \If {$this.state[a] > x$}
    \State \textbf{forall} ($c' \in this.children$)
    \State \;\;\;\; \textbf{if} (c'.state[a] > x)
    \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a, x$);
    \State \call{} $this.$\dResp($p, a, x$);
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ cache's downgrade request handler -- request from $this$'
parent, $p$, to upgrade $p$'s state to $x$ for address $a$ }
\label{atomicDReq1}
\end{subfigure}

\caption{MSI protocol}
\label{msi-template}
\end{figure}

Figure \ref{msi-template} gives the methods of a cache to handle the upgrade
requests from its children (or core) and downgrade requests from its parent.
When the upgrade request handler \uReq{} for a cache $this$ is invoked by a
cache $c$ to go to a particular state $x$, $this$ first makes sure that its
other children are compatible with $c$ being in state $x$ for the requested
address. This involves downgrading their states to $toCompatible(x)$. Cache
$this$ has to upgrade its own state before upgrading the state of its child $c$,
if $this$' state is less than the requested state. In this case, $this$ invokes
its parent's handler for upgrade requests. Finally, when state of $this$ is not
less than t$x$ and all its children are compatible with $x$, $this$ can upgrade
$c$'s state, and send the data if $c$'s state was $I$ previously. This is shown
in Figure \ref{atomicUReq1}.

A cache, $this$', downgrade handler \dReq{} can also be invoked (for instance
when its parent wants $this$ to downgrade its state to $x$ because the parent's
upgrade request handler was invoked by some other cache. In this case, $this$
makes sure that each of its child's state is not higher than $x$, by invoking
their respective downgrade methods. When all the children have downgraded, the
cache downgrades its own state to $x$, and sends data back to the parent, if it
has been modified (\ie $isModified(x)$ is $True$).

\begin{figure}
\small

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReqL}{$p, a, x$}
  \State \call{} $p.$\uReq($this, a, x$);
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Upgrading address $a$ of $this$ to $x$, by invoking the upgrade handler
of $this$'s parent, $p$}
\label{uReq}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReqL}{$c , a, x$}
  \State \call{} $c.$\dReq($this, a, x$);
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Downgrading address $a$ of $this$'s child $c$ to $x$, by invoking
$c$'s downgrade handler}
\label{dReq}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uResp}{$c, a, x$}
  \If {$c.state[a] = I$}
    \State $c.data[a] \gets this.data[a]$;
  \EndIf
  \State $c.state[a] \gets x$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ finally upgrading address $a$ of its child $c$ to $x$, and
updating the data of $c$ if necessary}
\label{uResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dResp}{$p, a, x$}
  \If {$isModified(c.state[a])$}
    \State $p.data[a] \gets this.data[a]$;
  \EndIf
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ finally downgrading its state for address $a$ to $x$ and
updating data of its parent $p$ if necessary}
\label{dResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\sendMem{}}{$a$}
  \State $this.data[a] \gets memory[a]$;
  \State $this.state[a] \gets M$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Last-Level-Cache $this$ reading data from memory for address $a$}
\label{uResp}
\end{subfigure}

\caption{Methods used in the MSI protocol dealing with other objects}
\label{atomic}
\end{figure}

Figure \ref{msi-template} gives the pseudo-code involved, including the
actions needed to evict a cache line and replace it with another line, in case
there are no free slots available for an address. For reasons that will become
clear in the distributed setting, each cache can directly read or write its own
state and data for each address. It can also directly read it's child's state,
but not directly write it. It can not read it's parents state directly. It can
not directly read or write other cache's data. In order to perform the required
actions indirectly, the cache can invoke other cache's corresponding
\textbf{method}s packaged together as shown in Figure \ref{atomic}. To be
precise we use $this$ explicity even if the object invokes its own methods.

In the case of an L1 cache, once the cache's parent has upgraded its state,
there is no child to upgrade to. Similarly, in the case of a Last-Level-Cache,
the cache simply reads data from the memory if the line is not present in the
cache, since it has no parent whose upgrade request handler it can invoke. All
lines in the Last-Level-Cache are upgraded automatically to be in $M$ state.

\subsection{Refactoring the handler methods}
Figure \ref{msi-template} shows how the request handler methods \uReq{} and
\dReq{} can be refactored to make it amenable to distributed implementation. The
upgrade and downgrade request handlers are not allowed to invoke other cache's
upgrade or downgrade request handlers, or change other cache's states. Instead,
the same calls are made via 5 more local methods listed in Figure \ref{atomic}.
These methods in turn invoke the other caches upgrade and downgrade request
handlers, or change the states of other caches.

\begin{figure}
\small

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReqL}{$p, a, x$}
  \State \send{} \Req{this}{p}{a}{x};
  \While {$this.state[a] < x$}
  \State \receive{} \Resp{p}{this}{a}{z};
  \If {$this.state[a] = I$}
    \State \receive{} \Data{p}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.state[a] \gets z$;
  \EndWhile
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Upgrading address $a$ of $this$ to $x$, by sending a request to
$this$'s parent $p$}
\label{uReqLocal}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReqL}{$c, a, x$}
  \State \send{} \Req{this}{c}{a}{x};
  \While {$this.dir[c][a] > x$}
  \State \receive{} \Resp{c}{this}{a}{z};
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
  \State $this.dir[c][a] \gets z$;
  \EndWhile
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Downgrading address $a$ of $this$'s child $c$ to $x$, by sending a
request to $c$}
\label{dReqLocal}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uResp}{$c, a, x$}
  \State \textbf{if} ($this.dir[c][a] = I$)
  \State \;\;\;\; \send{} \Data{this}{c}{a}{this.data[a]};
  \State $this.dir[c][a] \gets x$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ finally upgrading address $a$ of its child $c$ to $x$, sending a
response to $c$ notifying of the upgrade and sending data to $c$ if necessary}
\label{uResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dResp}{$p, a, x$}
  \State \textbf{if} ($isModified(this.state[a])$)
  \State \;\;\;\; \send{} \Data{this}{p}{a}{this.data[a]};
  \State $this.state[a] \gets x$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ finally downgrading its state for address $a$ to $x$, sending a
response to $c$ notifying of the downgrade and sending data to $c$ if necessary}
\label{dResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dRespL}{$c , a, x$}
  \If {$isModified(this.dir[c][a])$}
    \State \receive{} \Data{c}{this}{a}{d};
    \State $this.data[a] \gets d$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{$this$ handling unsolicited downgrade to $x$ response from $c$ for
address $a$. This happens if $c$ evicted address $a$}
\label{uResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\sendMem{}}{$a$}
  \State \send{} \Req{this}{Memory}{a}{M};
  \State \receive{} \Data{Memory}{this}{a}{d};
  \State $this.data[a] \gets d$;
  \State $this.state[a] \gets M$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Last-Level-Cache $this$ sending and receivging data from memory for
address $a$}
\label{uResp}
\end{subfigure}

\caption{Non atomic implementation of MSI protocol}
\label{realistic}
\end{figure}

### 2012-11-21 paper/isca40format/Network.tex
before arvind is going to re-org

\section{Network, resource and scheduler requirements}
\label{sec:network}

In this section we will spell out all the requirements that are needed in terms
of ordering of messages between a pair of caches, blocking of messages in the
network, the minimum amount of thread resources needed, and the requirements
for scheduling a thread. This hides all the complexity of the distributed MSI
protocol, so that the protocol can be reasoned about one thread at a time, as
if the threads in each node were non-suspensive.

\subsection{Message ordering requirements}
\floatstyle{boxed}
\restylefloat{figure}

\begin{figure}\small
\begin{requirement}
An incoming response for an address $a$ from a child $c$ should not be
received before another incoming response from the same $c$ and the
same $a$ sent earlier has been received\label{cRespFifo}
\end{requirement}
\begin{requirement}
An incoming request for an address $a$ from a source $n$ should not be
received before another incoming response from the same $n$ and the
same $a$ sent earlier has been received\label{reqNoOvertakeResp}
\end{requirement}
\caption{Ordering requirements}
\label{order}
\end{figure}

In the distributed MSI protocol presented in Section
\ref{sec:DistributedMsi}, we were agnostic about the properties of the input and
output channels of the caches. Figure \ref{order} gives the exact ordering
requirements between messages transmitted between the caches.

We will illustrate the need for Requirements \ref{cRespFifo} and
\ref{reqNoOvertakeResp} using a 2-level of caches having two L1 caches $c_1$ and
$c_2$, and one shared L2 cache $p$. Each core can have several outstanding
memory requests.

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.4]{complicated}
\caption{Effect of violating Requirement \ref{cRespFifo} (order between responses)}
\label{complicated}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

Let's say $c_1.state[a] = p.dir[c_1][a] = M$, $c_2.state[a] = p.dir[c_2][a] =
I$. $c_2$ gets a load request from its processor for address $a$ and sends an
upgrade request to $p$. $p$ sends a downgrade-to-$S$ request to $c_1$. $c_1$
receives the downgrade-to-$S$ request and downgrades to $S$, sending a downgrade
response to $p$. $c_1$ decides to evict address $a$ to make room for another
address $a'$, because of a request from the core for $a'$ not present in $c_1$.
$c_1$ sends a downgrade-to-$I$ response to $p$. Let's say the second downgrade
to $I$ response reaches $p$ first, violating Requirement \ref{cRespFifo}. $p$
sends an upgrade-to-$S$ response to $c_2$. $c_2$ receives the upgrade response
and changes state to $S$. $p$ then receives the second downgrade-to-$S$ response
from $c_1$, changing its $p.dir[c_1][a]$ to $S$. $c_2$ sends an upgrade-to-$M$
request to $p$ because of a store from its core for $a$. $p$ will send a
downgrade-to-$I$ request to $c_1$, but $c_1$ drops the request as seen in
procedure \dReq{} in Figure \ref{realistic}. So $p$ will never receive a
response from $c_1$ leading to a deadlock. This is illustrated in Figure
\ref{complicated}.

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.4]{lessComplicated}
\caption{Effect of violating Requirement \ref{reqNoOvertakeResp} (requests overtaking responses)}
\label{lessComplicated}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

Consider another scenario starting with $c_1.state[a] = p.dir[c_1][a] = S$ and
$c_2.state[a] = p.dir[c_2][a] = I$. $c_1$ receives a store request and sends an
upgrade-to-$M$ request to $p$. $p$ sends back an upgrade-to-$M$ response to
$c_1$. $c_2$ gets a load request for address $a$ and sends an upgrade-to-$S$
request to $p$. $p$ sends a request to $S$ to $c_1$. Suppose Requirement
\ref{reqNoOvertakeResp} is violated and the request to $S$ reaches $c_1$ before
the response to $M$. $c_1$ will drop the request since its already in state $S$,
and so $p$ will never get a response leading to a deadlock. This is illustrated
in Figure \ref{lessComplicated}.

\subsection{Resource requirements and blocking restrictions}

\begin{figure}\small
\begin{requirement}
At a cache node, an incoming request from the parent should not be blocked by any
incoming request from its children.\label{cReqNoBlockPReq}
\end{requirement}
\begin{requirement}
At a cache node, an incoming response from any of its children should not
be blocked by any incoming requests.\label{reqNoBlockResp}
\end{requirement}
\begin{requirement}
At a cache node, an incoming response for address $a$ from its parent should not be blocked by
an incoming request from its child for address $a$. \label{reqBlockResp2}
\end{requirement}
\caption{Blocking restrictions}
\label{blocking}
\end{figure}

Figure \ref{blocking} shows the restrictions that the input channels should
obey with respect to an incoming message blocking other incoming messages.

\begin{figure}\small
\begin{requirement}
Each cache node has at least one dedicated thread resource for handling only messages
(requests or responses) from the parent\label{dedicate1}
\end{requirement}
\begin{requirement}
Each cache node has at least one dedicated thread resource for handling only responses
(from children or from the parent)\label{dedicate2}
\end{requirement}
\begin{requirement}
Each cache node has at least three thread resources\label{dedicate3}
\end{requirement}
\caption{Dedicated resources in a cache node}
\label{dedicated}
\end{figure}

Figure \ref{dedicated} shows the minimum number of thread resources that should
be present in each cache node. Lack of required thread resouces leads to the same
kindof deadlock scenarios as incorrect blocking does. This is illustrated in the
following:

Let's say both $c_1.state[a] = p.dir[c_1][a] = S$ and $c_2.state[a] =
p.dir[c_2][a] = S$. Both $c_1$ and $c_2$ get store requests from the processor
for address $a$, and they send upgrade-to-$M$ requests to $p$.  $p$ receives the
request from $c_2$ first. $p$ sends a downgrade-to-$I$ request to $c1$.

\floatstyle{plain}
\restylefloat{figure}
\begin{figure}
\centering
\includegraphics[scale=0.4]{unknown}
\caption{Effect of violating Requirement \ref{cReqNoBlockPReq} (requests blocking responses from parent)}
\label{unknown}
\end{figure}
\floatstyle{boxed}
\restylefloat{figure}

Let's say Requirement \ref{cReqNoBlockPReq} is not held, and instead some request
from the core is blocking $c_1$ from receiving the request. $c_1$ will never
respond back to $p$. But $p$ will be waiting in procedure \dReqL{}, and will
never start handling the request from $c_1$. This will create a deadlock.  The
same deadlock is also created when $c_1$ is out of thread resources to create a
thread to handle the downgrade request from $p$ (violating Requirement
\ref{dedicate1}). Figure \ref{unknown} depicts this scenario.

Let's extend this scenario further. Instead of violating Requirements
\ref{cReqNoBlockPReq} or \ref{dedicate1}, let $c_1$
send a downgrade-to-$I$ response to $p$. Let's instead assume that Requirement
\ref{reqNoBlockResp} is violated in $p$. So, the downgrade-to-$I$ response from
$c_1$ is not received by $p$, leading to the same deadlock. This is the same if
Requirement \ref{dedicate2} is violated and $p$ is out of thread resources to
create a thread to handle the downgrade response.

The need for Requirement \ref{dedicate3} can be shown easily. Let's say that a
cache has exactly 2 thread resources, one for handling messages from parents,
and one for handling responses from either the parent or the children. Now, if
it receives a request from one of its children, it can not use either of the
thread resources. Thus, at least one more thread resource is needed for
handling requests from children. This thread resource can also be used for
handling other types of incoming message.

\subsection{Scheduler requirements}

We will now list the requirements that a scheduler must obey to avoid
deadlocks as well as to give the illusion of non-suspensive threads.

\begin{figure}\small
\begin{requirement}
$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{pHandleReq}
\end{requirement}
\begin{requirement}
$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from $p$, or
\item a thread in $c$ is handling a request for address
$a$ from one of $c$'s children $c'$, but that thread is not waiting for a
response from $p$
\end{enumerate}
\label{cHandleReq}
\end{requirement}
\caption{Requirements for receiving a request (used by the scheduler)}\label{recvReq}
\label{}
\end{figure}

\begin{figure}\small
\begin{requirement}
Address $a$ present in cache node $c$ can not be chosen for eviction by a thread
executing in $c$ if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from any source
\end{enumerate}\label{evict}
\caption{Requirements for suspending a thread during choosing an address for eviction}
\end{requirement}
\end{figure}

Requirements \ref{pHandleReq}, \ref{cHandleReq} and \ref{evict} are needed
because once a thread has finished up-to a point in its execution, it can not
re-execute that portion. For instance, consider a thread handling an upgrade
request from a child. Let's say thread $t$ has progressed to the point where it
is simply waiting to send an upgrade response to the child. Now, if the line
gets evicted because it was chosen for replacement, or it got downgraded by the
parent, then thread $t$ can not re-execute its procedure again. To prevent this,
only one thread is allowed to operate on a particular address, except in the
case of downgrade requests from the parent. In this case, a thread is created to
handle the downgrade request from the parent even if there is another thread
handling a request for the same address from a child, if the latter thread is
just waiting for a response from the parent. This is done to avoid deadlocks
like the ones shown in Figure \ref{unknown}.

\begin{figure}\small
\begin{requirement}
Starvation-freedom: A thread that keeps becoming ready to execute will
eventually be scheduled to execute.\label{starvation}
\end{requirement}
\end{figure}

Local starvation-freedom is required in each cache node. If a cache node never
executes a particular suspended thread, even if it keeps getting ready to
execute, then the thread will not progress, leading to a deadlock.

A scheduler can be implemented as described in Figure \ref{scheduler}, obeying
Requirements \ref{pHandleReq} and \ref{cHandleReq}. It should perform some form of
fair arbiration among its suspended threads to preserve Requirement
\ref{starvation}.

\newcommand{\lWhile}{\textbf{while}}
\newcommand{\lIf}{\textbf{if}}
\newcommand{\lElsIf}{\textbf{else if}}
\newcommand{\lElse}{\textbf{else}}

\floatstyle{plain}
\restylefloat{figure}

\begin{figure}
\small
\begin{boxedminipage}{\linewidth}
\begin{alltt}
\normalfont
\lWhile{} (\(True\)) \bopen
      \lIf (\Resp{c}{n}{a}{x} is in the input channel,
            where \(n = c.parent\)) \bopen
            \lIf (suspended thread \(t\) is waiting for 
                 a response from \(c\) for address \(a\)) \bopen
                   \resume{} \(t\);
            \bclose \lElse \bopen
                   \receive{} \Resp{c}{n}{a}{x};
                   \start{} \dRespL(\(c, n, a, x\));
            \bclose
      \bclose \lElsIf (\Resp{p}{n}{a}{x} is in the input channel,
            where \(p = n.parent\)) \bopen
                   // There must be a suspended thread \(t\)
                   waiting for a response from \(p\) for address \(a\)
                   \resume{} \(t\);
            \bclose
      \bclose \lElsIf (\Req{c}{n}{a}{x} is in the input channel,
                     where \(n = c.parent\)) \bopen
             \lIf (Local-Property \ref{pHandleReq} is satisfied) \bopen
                   \receive{} \Req{c}{n}{a}{x};
                   \start{} \uReq(\(c, n, a, x\));
             \bclose
      \bclose \lElsIf (\Req{p}{n}{a}{x} is in the input channel,
                  where \(p = n.parent\)) \bopen
             \lIf (Local-Property \ref{cHandleReq} is satisfied) \bopen
                   \receive{} \Req{p}{n}{a}{x};
                   \start{} \dReq(\(p, n, a, x\));
             \bclose
      \bclose \lElsIf (suspended thread \(t\) is ready to execute) 
             \resume{} \(t\);
\bclose
\end{alltt}
\end{boxedminipage}
\caption{Implementation of a scheduler for cache node $n$}
\label{scheduler}
\end{figure}

### 2012-11-21 paper/isca40format/RelatedWork.tex
before arvind is going to re-org

\section{Related work}
\label{sec:related}

Model-checking based techniques have been used before in trying to prove cache
coherence protocols \cite{Clarke}, \cite{Chou}, \cite{ip}, \cite{McMillan},
\cite{pong}, \cite{Stern}. They try to tackle model-checking's state explosion
problem using innovative techniques, but the fundamental issue remains -- large
systems can not be verified by model-checking.

Zhang \etal proposed Fractal Coherence \cite{Zhang} for verifying protocols
using model-checking. A minimum system is verified and the protocol has to be
designed in such a way that if a subtree is replaced by a single node, it has
the same interface behavior (the fractal property). The difficulty in this
approach is that fractal property is not common in hierarchical protocols, extra
non-intuitive messages have to be passed around to ensure the property.

There has also been some work on using theorem proving techniques for proving
the correctness of directory-based cache coherence protocols \cite{loew},
\cite{park}. They rely on using abstraction functions to simplify the proof of
coherence protocols, which is similar to our approach. We take this approach one
step further by doing automating the procedure.

\cite{Beu} and\cite{edya} present protocols which is similar to the final
distributed protocol that we present. But they do not present any reasoning
about the correctness of the protocol, or an abstract automated procedure with
a set of invariants to generate further protocols.

### 2012-11-21 paper/isca40format/Theorems.tex
before arvind is going to re-org

@@
 \ref{realistic} is correct.
-
-\begin{theorem}
-It a thread has to send a message, it will eventually be able to send the
-message.
-\label{canSend}
-\end{theorem}
-
-\begin{theorem}
-If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
-get an upgrade to $y$ response from the parent where $y \ge x$.
-\label{willRecv}
-\end{theorem}
-
-\begin{theorem}
-If a cache sends an downgrade-to-$x$ request to one of its children, it will
-eventually get a downgrade-to-$z$ response from the child where $z \le x$.
-\label{willRecv2}
-\end{theorem}
-
-\begin{theorem}
-If a thread is waiting for a line to be allocated, \ie for any address to become
-available for eviction, then some address will eventually become available for
-eviction
-\label{evictDead}
-\end{theorem}
-
-\begin{theorem}
-If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
-just before $p$ sends the response is the same as $c.state[a]$ just before $c$
-receives the response.
-\label{pcSame}
-\end{theorem}
-
-\begin{theorem}
-If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
-just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
-receives the response.
-\label{cpSame}
-\end{theorem}
-
-\begin{theorem}
-Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
-p.dir[c][a] \ge c.state[a]$.
-\label{conservative}
-\end{theorem}
-
 The need for these invariants is intuitive. If, for instance, a thread wants to

### 2012-11-21 paper/isca40format/stuff/MsiArvind2.tex
before arvind is going to re-org

--- 2012-11-21 paper/isca40format/MyMsi.tex
@@
 \section{MSI protocol}
-\label{sec:GlobalMsi}
+\label{sec:msi}
 
@@
 The caches are logically organized in the form of a tree
-(Figure~\ref{hierarchy}). The leaves of this tree represent the L1 caches,
+(Figure~\ref{fig:hierarchy}). The leaves of this tree represent the L1 caches,
 with each L1 cache connected to exactly one core. The root of the tree is the
 Last-Level-Cache, and its connected to the memory. The cache hierarchy is
-inclusive, \ie if an address is present in a cache $c$, then it's also present it its
+inclusive, \ie if an address is present in a cache $c$, then it's present it its
 parent $c.parent$. Note that by address we mean cache-line address, not a byte
@@
 For a non L1 cache, the meaning of coherence states is different. For an
-internal cache node $n$, if $n.state[a] = I$, then, the entire sub-tree rooted at $n$
+internal node $n$, if $n.state[a] = I$, then, the entire sub-tree rooted at $n$
 will have coherence state $I$ for address $a$. In particular, no L1 cache in
@@
 
-\begin{theorem}
+\begin{inv}
 \textbf{Single-writer Invariant}: When a cache has permission to modify data for
@@
 \label{singleWriter}
-\end{theorem}
+\end{inv}
 
@@
 
-\begin{theorem}
+\begin{inv}
 \textbf{Read-from-last-writer}: When a cache reads data for an address, it reads
@@
 \label{lastRead}
-\end{theorem}
+\end{inv}
 
@@
 
-We will first describe an atomic version of the hierarchical MSI protocol. Each
-of the caches in the hierarchy can be treated like C++/Java objects.
+We will first describe an atomic version of the hierarchical MSI protocol. In
+the atomic protocol, when a core connected to an L1 cache initiates an upgrade
+for an address, no other core can initiate an upgrade for any address till the
+former L1 cache has finished upgrading. This can be implemented in hardware
+using a bus-based system, which locks the bus till an initiated upgrade has
+completed.
+
 \floatstyle{boxed}
 \restylefloat{figure}
-
-\begin{figure}
-\small
-\begin{algorithmic}
-\Proc {\textsc{coreReq}}{$core, a, op, data$}
-  \If {$op = Ld$}
-    \State \textbf{if} ($this.state[a] < S$)
-    \State \;\;\;\; \call{} $this.$\uReq($core, a, S$);
-    \State \send{} $this.data[a]$;
-  \ElsIf {$op = St$}
-    \State \textbf{if} ($this.state[a] < M$)
-    \State \;\;\;\; \call{} $this.$\uReq($core, a, M$);
-    \State $update(this.data[a], data)$;
-  \EndIf
-\EndProc
-\end{algorithmic}
-\caption{L1 cache handling a load or a store request sent by the core by
-calling the request handler}
-\label{coreHandle}
-\end{figure}
-
-In the atomic protocol, a core invokes the L1 cache's request handler
-\textsc{coreReq} (Figure \ref{coreHandle}). The L1 cache checks if the request
-can already be serviced, if not, it invokes its upgrade request handler \uReq{},
-which upgrades the state of the requested address. Finally, in the case of a
-load request, the data is sent back to the core, and in the case of a store, the
-cache line is updated. Note that the core supplies words or long-words while the
-cache line sizes are typically 64 bytes; only the appropriate word in the cache
-line is updated in case of a store.
 
@@
 \begin{algorithmic}
+\Proc {\textsc{processReq}}{$c, a, op, data$}
+  \If {$op = Ld$}
+    \State \textbf{if} ($c.state[a] \ge S$)
+    \State \;\;\;\; \call{} $c.$\uReq($this, a, S$);
+    \State \textbf{return} $c.data[a]$;
+  \ElsIf {$op = St$}
+    \State \textbf{if} ($c.state[a] = M$)
+    \State \;\;\;\; \call{} $c.$\uReq($this, a, M$);
+    \State $update(c.data[a], data)$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{core handling a load or a store request by calling the request
+handler in its L1 cache}
+\label{coreHandle}
+\end{subfigure}
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
 \Proc {\uReq}{$c , a, x$}
-  \If {$c.state[a] < x$}
+    \If {$a$ is not present in $this$}
+      \If {$this$ has no space for $a$}
+        \State Choose address $a'$ for eviction;
+        \State \textbf{forall} ($c' \in this.children$)
+        \State \;\;\;\; \textbf{if} ($c'.state[a'] > I$)
+        \State \;\;\;\;\;\;\;\; \call{} $c'.$\dReq($this, a', I$);
+        \If {$this.state[a'] > I$}
+          \State \textbf{if} ($this.state[a'] = M$)
+          \State \;\;\;\; $(this.parent).data[a'] \gets this.data[a']$;
+          \State $this.state[a'] \gets I$;
+        \EndIf
+      \EndIf
+      \State Allocate a line for $a$;
+    \EndIf
+    \State // $a$ is already present in $this$
+    \State \textbf{forall} ($c' \in this.children$)
+    \State \;\;\;\; \textbf{if} ($c'.state[a] > toCompatible(x)$)
+    \State \;\;\;\;\;\;\;\; \call{} $c'.$\dReq($this, a', toCompatible(x)$);
+    \State \textbf{if} ($this$ is Last-Level-Cache \&\& $this.state[a] = I$) \bopen
+    \State \;\;\;\; $this.data[a] \gets memory[a]$;
+    \State \;\;\;\; $this.state[a] \gets M$;
+    \State \bclose \;\textbf{else}
+    \State \;\;\;\; \call{} $(this.parent.)$\uReq{}($this, a, x$);
+  \If {$c \neq core$}
+  \State \textbf{if} ($c.state[a] = I$)
+  \State \;\;\;\; $c.data[a] \gets this.data[a]$;
+  \State $c.state[a] \gets x$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
+\label{atomicUReq1}
+\end{subfigure}
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\dReq}{$p , a, x$}
+  \If {$this.state[a] > x$}
+    \State \textbf{forall} ($c' \in this.children$)
+    \State \;\;\;\; \textbf{if} (c'.state[a] > x)
+    \State \;\;\;\;\;\;\;\; \call{} $c'.$\dReq($this, a, x$);
+    \State \textbf{if} ($this.state[a] = M$)
+    \State \;\;\;\; $p.data[a] \gets this.data[a]$;
+    \State $this.state[a] \gets x$;
+  \EndIf
+\EndProc
+\end{algorithmic}
+\end{boxedminipage}
+\caption{$this$ cache's downgrade request handler -- request from $this$' parent, $p$, to upgrade $this$' state to $x$ for address $a$ }
+\label{atomicDReq1}
+\end{subfigure}
+
+\caption{Atomic MSI protocol}
+\label{msi-simple}
+\end{figure}
+
+Figure \ref{msi-simple} gives the pseudo-code for handling a request from a
+core. The caches and the cores can be thought of as objects with methods. The
+core has a \textsc{processReq} method which it invokes to process a load or
+store request. The core first checks if its L1 cache has the required state. If
+not, it invokes the L1 cache's upgrade request handler \uReq{}. \uReq{} in turn
+calls the upgrade request handler of which would upgrade the state of the
+L1-cache, and finally the request from the core will be serviced. Note that the
+core supplies words or long-words while the cache line sizes are typically 64
+bytes; only the appropriate word in the cache line is updated in case of a
+store.
+
+In the case of a non-L1 cache, $this$, when its upgrade request handler is
+invoked by its child $c$, if the requested addres is not already in $this$, it
+tries to get a line for the address. If the cache has no empty slots for the
+requested address, an existing address is evicted. In order to evict an address
+$a'$, to maintain the inclusiveness property, the address $a'$ has to be evicted
+in all of the $this$' children. This is done by calling the downgrade request
+handler \dReq{} of each of its child which is at a higher state than $I$ for
+that address; such a child is asked to downgrade to $I$ state. Once the children
+have evicted, if data in $this$ is modified with respect to its parent, then the
+data is written back. At this point a line is allocated for the originally
+requested address.
+
+Once $this$ has established that it has a line for the requested address, it now
+has to ensure that the states of its other children are compatible with the
+requested upgrade; $this$ again invokes its children's downgrade request handler
+\dReq{}, for all the children who are at a higher state than what is compatible,
+to downgrade them to the compatible state. $this$ invokes its parents upgrade
+request handler ($(this.parent).\uReq{}$), which in turn will will do the same
+sequence of operations and eventually upgrade $this$' state. Finally, $this$
+upgrades the state of the requesting child $c$, and sends data to $c$ if
+necessary.
+
+When a cache, $this$', downgrade handler is invoked by its parent, it in turn
+invokes the downgrade handler of all its children which are at a higher state
+than what the parent requested. Once all the children have downgraded their
+respective states, $this$ downgrades its own state and writes the data back to
+its parent if it has been modified with respect to the parent, \ie if $this$ is
+in $M$ state.
+
+In the case of an L1 cache, once the cache's parent has upgraded its state,
+there is no child to upgrade to. Similarly, in the case of a Last-Level-Cache,
+the cache simply reads data from the memory if the line is not present in the
+cache, since it has no parent whose upgrade request handler it can invoke. All
+lines in the Last-Level-Cache are made to be in $M$ state.
+
+\subsection{Refactoring the handler methods}
+\begin{figure}
+\small
+
+\begin{subfigure}{\linewidth}
+\begin{boxedminipage}{\linewidth}
+\begin{algorithmic}
+\Proc {\uReq}{$c , a, x$}
     \If {$a$ is not present in $this$}
@@
   \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
-  \EndIf
 \EndProc
@@
 
-\caption{MSI protocol}
+\caption{MSI protocol -- general template}
 \label{msi-template}
 \end{figure}
-
-Figure \ref{msi-template} gives the methods of a cache to handle the upgrade
-requests from its children (or core) and downgrade requests from its parent.
-When the upgrade request handler \uReq{} for a cache $this$ is invoked by a
-cache $c$ to go to a particular state $x$, $this$ first makes sure that its
-other children are compatible with $c$ being in state $x$ for the requested
-address. This involves downgrading their states to $toCompatible(x)$. Cache
-$this$ has to upgrade its own state before upgrading the state of its child $c$,
-if $this$' state is less than the requested state. In this case, $this$ invokes
-its parent's handler for upgrade requests. Finally, when state of $this$ is not
-less than t$x$ and all its children are compatible with $x$, $this$ can upgrade
-$c$'s state, and send the data if $c$'s state was $I$ previously. This is shown
-in Figure \ref{atomicUReq1}.
-
-A cache, $this$', downgrade handler \dReq{} can also be invoked (for instance
-when its parent wants $this$ to downgrade its state to $x$ because the parent's
-upgrade request handler was invoked by some other cache. In this case, $this$
-makes sure that each of its child's state is not higher than $x$, by invoking
-their respective downgrade methods. When all the children have downgraded, the
-cache downgrades its own state to $x$, and sends data back to the parent, if it
-has been modified (\ie $isModified(x)$ is $True$).
 
@@
 
-\caption{Methods used in the MSI protocol dealing with other objects}
+\caption{Atomic implementation of MSI protocol}
 \label{atomic}
@@
 
-Figure \ref{msi-template} gives the pseudo-code involved, including the
-actions needed to evict a cache line and replace it with another line, in case
-there are no free slots available for an address. For reasons that will become
-clear in the distributed setting, each cache can directly read or write its own
-state and data for each address. It can also directly read it's child's state,
-but not directly write it. It can not read it's parents state directly. It can
-not directly read or write other cache's data. In order to perform the required
-actions indirectly, the cache can invoke other cache's corresponding
-\textbf{method}s packaged together as shown in Figure \ref{atomic}. To be
-precise we use $this$ explicity even if the object invokes its own methods.
-
-In the case of an L1 cache, once the cache's parent has upgraded its state,
-there is no child to upgrade to. Similarly, in the case of a Last-Level-Cache,
-the cache simply reads data from the memory if the line is not present in the
-cache, since it has no parent whose upgrade request handler it can invoke. All
-lines in the Last-Level-Cache are upgraded automatically to be in $M$ state.
-
-\subsection{Refactoring the handler methods}
 Figure \ref{msi-template} shows how the request handler methods \uReq{} and
@@
   \State \send{} \Req{this}{p}{a}{x};
-  \While {$this.state[a] < x$}
   \State \receive{} \Resp{p}{this}{a}{z};
@@
   \State $this.state[a] \gets z$;
-  \EndWhile
 \EndProc
@@
   \State \send{} \Req{this}{c}{a}{x};
-  \While {$this.dir[c][a] > x$}
   \State \receive{} \Resp{c}{this}{a}{z};
@@
   \State $this.dir[c][a] \gets z$;
-  \EndWhile
 \EndProc

### 2012-11-22 paper/isca40format/Conclusion.tex
before reorging scheduler etc

\section{Conclusions}
\label{sec:conclusion}

We have presented an automatic procedure to generate a distributed cache
coherence protocol from a centralized protocol. We have shown how this can be
implemented realistically and give the lower bound on the number of virtual
channels to make this coherence protocol deadlock-free and the storage
requirements. We have also provided intuition for the correctness of the
generated protocol. We have shown that our method of designing protocols
substantially reduces the required effort in implementing provably correct
distributed protocols.

### 2012-11-22 paper/isca40format/DistributedMsi.tex
before reorging scheduler etc

@@
 
-In order to transfer the Global MSI protocol into a distribute MSI protocol, we
-need to make sure that a method can only read and write the state of the cache
-with which the method is associated, and it affects the state of other caches
-only by sending messages. In order to read the state of a child's cache we
-maintain a shadow state of in the form of a directory. Thus the coherence state
-of a cache $c$, in addition to the $c.state[a]$ field, also contains a
-directory field containing the shadow state of each child's state field, \ie,
-$c.dir[c'][a]$, for child $c'$.
+We will take the protocol designer's perspective, and first describe what the
+designer has to specify to create a distributed protocol.
 
-We extend our sequential language for writing methods with two communication
-primitives: \send{} $(Req|Resp|Data)\langle c_1 \rightarrow c_2, a,
-(x|data)\rangle$ and \receive{} $(Req|Resp|Data)(\langle c_1 \rightarrow c_2,
-a, (x|data)\rangle$ where caches $c_1$ and $c_2$ must satisfy the parent-child
-relation. Both of these primitives are \emph{blocking}, that is, \send{} may
-block because the network is clogged, \ie, has no buffer space to hold a
-message and a \receive{} may block because the message has not arrived or cache
-has no resources to process it. These primitives make our sequential language
-methods ``suspensive'. Request or response messages have very different
-characteristics and a data message is conceptually always associated with some
-response message. We think of each \textsc{coreReq} as launching a thread in
-L1. This thread in turn may get processed, send request or response messages,
-may get suspended and eventually die. Thus potentially every request message
-can launch a new thread in the cache that receives it. A response message,
-however, can only die or generate more response messages; it can never generate
-a request message. However a response message can also block waiting for the
-corresponding data message. Each thread has an address associated with it which
-corresponds to the message that created it.
+Each cache can only read and write its local state. A cache maintains a shadow
+ of all its children's states in the form of a directory. The designer can
+always assume that the state of an actual child is what is given in the
+directory. The directory entry of address $a$ for $this$' child $c$ is denoted
+using $this.dir[c][a]$.
 
-In order to write the distributed MSI protocol using suspensive threads, we will
-maintain the following sets of invariants. In the next section, we will show in
-the requirements in the network and the bounds on the number of virtual channels
-to avoid head of line blocking. We will also show the other resource
-requirements, as well as the properties of the thread scheduler that will ensure
-these invariants there.
+The designer has to specify the handlers for \uReq{} and \dReq{}. In the
+following, we assume that the cache whose handler the designer is writing is
+$this$ and the relevant address is $a$.
 
-\subsection{Relation between directory and state of the children's caches}
-\begin{theorem}
-Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
-p.dir[c][a] \ge c.state[a]$.  \label{conservative}
-\end{theorem}
+In order to read a child's state, the designer simply reads the state of the
+directory for that child. But to modify the state of child $c$ to $x$, the
+designer specifies a command \send{} \Req{this}{c}{a}{x}, followed by \receive{}
+\Resp{c}{this}{a}{y}. Command \send{} \Req{this}{c}{a}{x} can be called only when $x
+< this.dir[c][a]$. $y$ denotes a free variable which gets filled with the value of
+the state $c$ has changed to. $y$ is guaranteed to be $\le x$. $this.dir[c][a]$
+should be updated with this value.
 
-This ensures that if a cache makes decision about compatibility of states of
-its children using the local directory information, then it does not violate the
-compatibility of the children's real states.
+In order to read the data of another node $n$ (parent or child),
+the designer issues the command \receive{} \Data{this}{n}{a}{d}, where $d$ is a free
+variable that gets filled with the data. To write the data into address $a$ of
+another node $n$ (parent or child), the designer issues the command \send{}
+\Data{this}{n}{a}{this.data[a]}.
 
-\subsection{Network properties}
-Regarding the \send{} and \receive{} operations, \ie the network, the following
-properties can assume the following:
-\begin{theorem}
-It a thread has to send a message, it will eventually be able to send the
-message.
-\label{canSend}
-\end{theorem}
+In order to upgrade the state of $this$ to $x$, the designer has to
+specify the command \send{} \Req{this}{p}{a}{x} followed by \receive{}
+\Resp{p}{this}{a}{y}. Command \send{} \Req{this}{p}{a}{x} can be called only when $x
+> this.state[x]$. $y$ denotes a free variable which gets filled with the value of
+the state $this$ should be upgraded to. $y$ is guaranteed to be $\ge x$.
 
-\begin{theorem}
-If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
-get an upgrade-to-$y$ response from the parent where $y \ge x$.
-\label{willRecv}
-\end{theorem}
-
-\begin{theorem}
-If a cache sends an downgrade-to-$x$ request to one of its children, it will
-eventually get a downgrade-to-$y$ response from the child where $y \le x$.
-\label{willRecv2}
-\end{theorem}
-
-\subsection{Eviction}
-If handling of a request requires eviction of another cache line, then too the
-thread handling the request may get suspended. This would happen if all the
-lines that can be allocated for this request may accessed by other suspended
-threads. But we still assume the following invariant.
-
-\begin{theorem}
-If a thread is waiting for a line to be allocated, \ie for any address to become
-available for eviction, then some address will eventually become available for
-eviction
-\label{evictDead}
-\end{theorem}
-
-It this is not true, a thread that is waiting for being allocated a free line
-will never proceed leading to a deadlock.
-
-\subsection{Data transfer}
-Regarding the responses received, the designer can assume the following 2
-invariants. These invariants are used to wait for receiving data messages
-appropriately, as only some state transfers also cause data transfers.
-We will show the exact need for these invariants later.
-
-\begin{theorem}
-If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
-just before $p$ sends the response is the same as $c.state[a]$ just before $c$
-receives the response.
-\label{pcSame}
-\end{theorem}
-
-\begin{theorem}
-If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
-just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
-receives the response.
-\label{cpSame}
-\end{theorem}
-
-\subsection{Need for invariants}
-The need for these network invariants is intuitive. If, for instance, a thread wants to
-send a message in the output channel of the node and is not able to send one
-forever, then the thread can not proceed further. If a thread is waiting for a
-response (for a request previously sent), and it never gets the response, the
-thread can never proceed further. These scenarios create a deadlock.
-
-We will illustrate the need for Invariants \ref{pcSame}, \ref{cpSame} and
-\ref{conservative} using an example of a 2-level system containing two L1
-caches $c_1$ and $c_2$ with a shared L2 cache $p$.
-
-Invariant \ref{pcSame} and \ref{cpSame} are important to make sure that the
-data transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
-$c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
-sends an upgrade-to-$M$ to $p$. $p$ receives the request, but assumes that
-$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant
-\ref{pcSame} is violated. So, it does not transfer the data to $c_1$, resulting
-in $c_1$ waiting for data, and hence deadlock. A dual condition can be
-constructed similarly, in which $c_1$ downgrades its state for address $a$ from
-$S$ to $I$, and the parent assumes that $c_1$ is in state $M$, and keeps
-waiting for data.
-
-Invariant \ref{conservative} is important because a parent node consults its
-directory to decide the children it needs to send a downgrade request to.  Let
-$p.dir[c_1][a] = S$, $c_1.state[a] = S$, $p.dir[c_2][a] = I$ but
-$c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
-an upgrade-to-$M$ request to $p$, $p$ will assume that $c_2$'s state for address $a$
-is also $I$ and hence send an upgrade-to-$M$ response to $c_1$. But since the real
-state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
-\ref{singleWriter}.
-
-Thus, if all the invariants above (\ref{conservative}, to \ref{cpSame}) hold,
-then the threads given in Figure \ref{realistic} can not deadlock because it
-will always be able to a) send a request, b) send back a response, c) get back
-a response for a request it sent, d) receive data when required. Though the
-threads executing in a cache are actually suspensive, these invariants give an
-illusion of the threads being non-suspensive barring the following exceptions.
-Here $this$ represents the cache under under consideration.
-
+Using these commands, the designer can now easily specify a distributed MSI
+protocol, by specifying the methods \uReq{} and \dReq{}. Three other
+obligations to take care of:
 \begin{enumerate}
-\item $this.dir[c][a]$ can keep decreasing while a thread corresponding to
-address $a$ is executing.
-\item $this.dir[c][a']$ can keep decreasing while a thread which is evicting
-line $a'$ is executing.
-\item If a thread receives responses from the parent, then $this.state[a]$ can
-decrease between the operation of sending a request to the parent and the
-operation of receiving a response from the parent.
+\item Handler \uReq{}($c, a, x$) can be called even if $x \ge
+this.dir[c]$. In this case, it should be dropped.
+\item Handler \dReq{}($p, a, x$) can be called even if $x \le
+this.state[a]$. In this case, it should be dropped.
+\item A new handler, \dRespL{}($c, a, x$), where $c$ is a child of cache $this$,
+should be provided, and this informs $this$ that $c$ has changed its state to
+$x$.
 \end{enumerate}
 
-As can be seen, these interactions are minimal, and can be ignored since they
-affect neither the handling of upgrade nor downgrade requests. Hence, the
-correctness of the protocol can be argued in terms of a single threaded,
-non-suspensive, execution model. The formal proof for this, as well as the
-invariants stated below is beyond the scope of this paper.
-
-In order to design the distributed MSI protocol, the following considerations
-must be kept in mind.
-\begin{enumerate}
-\item There are 3 kinds of incoming messages to be handled: request from that parent,
-requests from children and responses from children.
-\item A downgrade-to-$x$ message might be received from the parent even if the
-state of the cache is $ \le x$. This has to be handled.
-\item A parent sends a response to the child only when a request is received. But a child can send an voluntary/unsolicited response to the parent during evictions.
-\item A request should be sent to the parent only if the current cache's state
-is lower than what's desired to handle its incoming request. Similarly, a
-request should sent to its child only if the directory's notion of that cache's
-state is higher than what's desired.
-\item A second request is not sent for the same address from the same cache to the same
-destination, until a response is received for the first request.
-\item A cache can not evict a line if it is currently waiting for a response from the parent for the address corresponding to the line.
-\end{enumerate}
-
-%We always reserve resources, \eg, cache line slot for an address before sending
-%an upgrade request for that address.
-
-The handling of messages follow the exact same algorithm as given for the
-global MSI protocol in Figure \ref{msi-template} with the caveat that instead
-of reading a child's state, the cache $this$ can only read its own directory.
-So, it suffices to simply replace $c.state[a]$ by $this.dir[c][a]$ every time
-the state of child $c$ is read by $this$.
-
-As we mentioned above, a new handler has to be written for handling voluntary
-responses. This is shown in Figure \ref{msi-unsolicited}.
+Using the above set of rules, the designer can specify the distributed MSI protocol as shown in
+Figure \ref{realistic}. Handlers \uReq{} and \dReq{} are exactly the same as
+shown in Figure \ref{msi-template}, except that whenever the state of a child
+$c'$ is read, it should be replaced syntactically by $this.dir[c'][a]$.
 
@@
     \State $this.state[a] \gets M$;
-  \ElsIf
+  \Else
   \State \send{} \Req{this}{p}{a}{x};
@@
 for requests are exactly the same as in Figure \ref{msi-template}, with every
-occurence of $c'.state[a]$ replaced by $this.dir[c'][a]$}
+occurrence of $c'.state[a]$ replaced by $this.dir[c'][a]$}
 \label{realistic}
 \end{figure}
+
+Even though the designer only specifies the distributed protocol in the manner we have
+written in Figure \ref{realistic}, we automatically
+create a correct implementation of the cache controller from this
+specification.  The commands \send{} and \receive{} get translated into
+appropriate request and response messages which get sent to or received from the respective
+nodes, and they invoke the corresponding message handlers. Note that in the final
+implementation, sending of a message blocks if there is no space in the output
+channel and receiving of a message blocks if there is no message available for
+reception, though the designer is oblivious to this fact. We will describe how
+the specification is translated to an actual implementation, and show why the
+implementation is correct.
+
+\subsection{Caches as a system executing suspensive threads}
+The behavior of each cache node can be best described as a system of suspensive
+threads. These correspond to Miss Status Handling Register (MSHRs) in an actual
+implementation. A scheduler creates and schedules threads within the cache.  A
+new thread is created whenever a new message arrives in the input channel, and
+no other thread is executing. When a thread is executing a \send{} command, then
+an actual message is created and sent to the appropriate cache. The thread then
+goes into a suspend state till it receives a response message indicating that
+the requested state change has taken place. In this case, the suspended thread is
+woken up by the scheduler to perform the rest of its actions. Once all of its
+actions are complete, a thread dies, and its resources are freed.
+
+Each request goes through several stages of processing in a cache (as shown in
+Figure \ref{msi-template}). This is the reason for using the vocabulary of
+threads instead of MSHRs -- a thread makes the processing stage a particular
+request is currently in, implicit.
+
+Now that we have established the vocabulary of threads, we will proceed to
+describe the requirements that a network should guarantee in order to make such
+an implementation possible. We will also show the scheduling algorithm that we
+generate.
+
+\subsection{Message ordering requirements}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+%The basic ordering requirement is that responses should not overtake each other
+%when they are sent from the same source to the same destination. Similarly,
+%requests should not overtake responses from the same source to the same
+%destination.
+
+\begin{figure}\small
+\begin{requirement}
+An incoming response for an address $a$ from a child $c$ should not be
+received before another incoming response from the same $c$ and the
+same $a$ sent earlier has been received\label{cRespFifo}
+\end{requirement}
+\begin{requirement}
+An incoming request for an address $a$ from a source $n$ should not be
+received before another incoming response from the same $n$ and the
+same $a$ sent earlier has been received\label{reqNoOvertakeResp}
+\end{requirement}
+\caption{Ordering requirements}
+\label{order}
+\end{figure}
+
+Figure \ref{order} gives the exact ordering
+requirements between messages transmitted between the caches.
+
+We will illustrate the need for Requirements \ref{cRespFifo} and
+\ref{reqNoOvertakeResp} using a 2-level cache hierarchy with two L1 caches $c_1$ and
+$c_2$, and one shared L2 cache $p$. Each core can have several outstanding
+memory requests.
+
+\floatstyle{plain}
+\restylefloat{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=0.4]{complicated}
+\caption{Effect of violating Requirement \ref{cRespFifo} (order between responses)}
+\label{complicated}
+\end{figure}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+Let's say $c_1.state[a] = p.dir[c_1][a] = M$, $c_2.state[a] = p.dir[c_2][a] =
+I$. $c_2$ gets a load request from its processor for address $a$ and sends an
+upgrade request to $p$. $p$ sends a downgrade-to-$S$ request to $c_1$. $c_1$
+receives the downgrade-to-$S$ request and downgrades to $S$, sending a downgrade
+response to $p$. $c_1$ decides to evict address $a$ to make room for another
+address $a'$, because of a request from the core for $a'$ not present in $c_1$.
+$c_1$ sends a downgrade-to-$I$ response to $p$. Let's say the second downgrade
+to $I$ response reaches $p$ first, violating Requirement \ref{cRespFifo}. $p$
+sends an upgrade-to-$S$ response to $c_2$. $c_2$ receives the upgrade response
+and changes state to $S$. $p$ then receives the second downgrade-to-$S$ response
+from $c_1$, changing its $p.dir[c_1][a]$ to $S$. $c_2$ sends an upgrade-to-$M$
+request to $p$ because of a store from its core for $a$. $p$ will send a
+downgrade-to-$I$ request to $c_1$, but $c_1$ drops the request as seen in
+procedure \dReq{} in Figure \ref{realistic}. So $p$ will never receive a
+response from $c_1$ leading to a deadlock. This is illustrated in Figure
+\ref{complicated}. In the figure, the timeline moves from left to right.
+Whenever the state of the directory or the cache changes, we show the new state.
+Messages are shown using arrows from the source to the destination, moving
+across time.
+
+\floatstyle{plain}
+\restylefloat{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=0.4]{lessComplicated}
+\caption{Effect of violating Requirement \ref{reqNoOvertakeResp} (requests overtaking responses)}
+\label{lessComplicated}
+\end{figure}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+Consider another scenario starting with $c_1.state[a] = p.dir[c_1][a] = S$ and
+$c_2.state[a] = p.dir[c_2][a] = I$. $c_1$ receives a store request and sends an
+upgrade-to-$M$ request to $p$. $p$ sends back an upgrade-to-$M$ response to
+$c_1$. $c_2$ gets a load request for address $a$ and sends an upgrade-to-$S$
+request to $p$. $p$ sends a request to $S$ to $c_1$. Suppose Requirement
+\ref{reqNoOvertakeResp} is violated and the request to $S$ reaches $c_1$ before
+the response to $M$. $c_1$ will drop the request since its already in state $S$,
+and so $p$ will never get a response leading to a deadlock. This is illustrated
+in Figure \ref{lessComplicated}.
+
+These scenarios show that these ordering requirements are indeed necessary for a
+correct cache coherence protocol.
+
+\subsection{Resource requirements and blocking restrictions}
+
+\begin{figure}\small
+\begin{requirement}
+At a cache node, an incoming request from the parent should not be blocked by any
+incoming request from its children.\label{cReqNoBlockPReq}
+\end{requirement}
+\begin{requirement}
+At a cache node, an incoming response from any of its children should not
+be blocked by any incoming requests.\label{reqNoBlockResp}
+\end{requirement}
+\begin{requirement}
+At a cache node, an incoming response for address $a$ from its parent should not be blocked by
+an incoming request from its child for address $a$. \label{reqBlockResp2}
+\end{requirement}
+\caption{Blocking restrictions}
+\label{blocking}
+\end{figure}
+
+Figure \ref{blocking} shows the restrictions that the input channels should
+obey with respect to an incoming message blocking other incoming messages.
+
+\begin{figure}\small
+\begin{requirement}
+Each cache node has at least one dedicated thread resource for handling only messages
+(requests or responses) from the parent\label{dedicate1}
+\end{requirement}
+\begin{requirement}
+Each cache node has at least one dedicated thread resource for handling only responses
+(from children or from the parent)\label{dedicate2}
+\end{requirement}
+\begin{requirement}
+Each cache node has at least three thread resources\label{dedicate3}
+\end{requirement}
+\caption{Dedicated resources in a cache node}
+\label{dedicated}
+\end{figure}
+
+Figure \ref{dedicated} shows the minimum number of thread resources that should
+be present in each cache node. Lack of required thread resouces leads to the same
+kind of deadlock scenarios as incorrect blocking does. This is illustrated in the
+following:
+
+Let's say both $c_1.state[a] = p.dir[c_1][a] = S$ and $c_2.state[a] =
+p.dir[c_2][a] = S$. Both $c_1$ and $c_2$ get store requests from the processor
+for address $a$, and they send upgrade-to-$M$ requests to $p$.  $p$ receives the
+request from $c_2$ first. $p$ sends a downgrade-to-$I$ request to $c1$.
+
+\floatstyle{plain}
+\restylefloat{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=0.4]{unknown}
+\caption{Effect of violating Requirement \ref{cReqNoBlockPReq} (requests blocking responses from parent)}
+\label{unknown}
+\end{figure}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+Let's say Requirement \ref{cReqNoBlockPReq} is not held, and instead some request
+from the core is blocking $c_1$ from receiving the request. $c_1$ will never
+respond back to $p$. But $p$ will be waiting in procedure \dReqL{}, and will
+never start handling the request from $c_1$. This will create a deadlock.  The
+same deadlock is also created when $c_1$ is out of thread resources to create a
+thread to handle the downgrade request from $p$ (violating Requirement
+\ref{dedicate1}). Figure \ref{unknown} depicts this scenario.
+
+Let's extend this scenario further. Instead of violating Requirements
+\ref{cReqNoBlockPReq} or \ref{dedicate1}, let $c_1$
+send a downgrade-to-$I$ response to $p$. Let's instead assume that Requirement
+\ref{reqNoBlockResp} is violated in $p$. So, the downgrade-to-$I$ response from
+$c_1$ is not received by $p$, leading to the same deadlock. This is the same if
+Requirement \ref{dedicate2} is violated and $p$ is out of thread resources to
+create a thread to handle the downgrade response.
+
+The need for Requirement \ref{dedicate3} can be shown easily. Let's say that a
+cache has exactly 2 thread resources, one for handling messages from parents,
+and one for handling responses from either the parent or the children. Now, if
+it receives a request from one of its children, it can not use either of the
+thread resources. Thus, at least one more thread resource is needed for
+handling requests from children. This thread resource can also be used for
+handling other types of incoming message.
+
+\subsection{Scheduler requirements}
+
+We will now list the requirements that a scheduler must obey to give the
+illusion of non-suspensive threads and avoid deadlocks.
+
+\begin{figure}\small
+\begin{requirement}
+$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
+if either
+\begin{enumerate}
+\item address $a$ has been chosen for eviction by some thread and that thread
+has not yet replaced $a$, or
+\item a thread in $p$ is handling a request for address
+$a$ from any source
+\end{enumerate}
+\label{pHandleReq}
+\end{requirement}
+\begin{requirement}
+$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
+if either
+\begin{enumerate}
+\item address $a$ has been chosen for eviction by some thread and that thread
+has not yet replaced $a$, or
+\item a thread in $c$ is handling a request for address
+$a$ from $p$, or
+\item a thread in $c$ is handling a request for address
+$a$ from one of $c$'s children $c'$, but that thread is not waiting for a
+response from $p$
+\end{enumerate}
+\label{cHandleReq}
+\end{requirement}
+\caption{Requirements for receiving a request (used by the scheduler)}\label{recvReq}
+\label{}
+\end{figure}
+
+\begin{figure}\small
+\begin{requirement}
+Address $a$ present in cache node $c$ can not be chosen for eviction by a thread
+executing in $c$ if either
+\begin{enumerate}
+\item address $a$ has been chosen for eviction by some thread and that thread
+has not yet replaced $a$, or
+\item a thread in $c$ is handling a request for address
+$a$ from any source
+\end{enumerate}\label{evict}
+\caption{Requirements for suspending a thread during choosing an address for eviction}
+\end{requirement}
+\end{figure}
+
+Requirements \ref{pHandleReq}, \ref{cHandleReq} and \ref{evict} are needed
+because once a thread has finished up-to a point in its execution, it can not
+re-execute that portion. For instance, consider a thread handling an upgrade
+request from a child. Let's say thread $t$ has progressed to the point where it
+is simply waiting to send an upgrade response to the child. Now, if the line
+gets evicted because it was chosen for replacement, or it got downgraded by the
+parent, then thread $t$ can not re-execute its procedure again. To prevent this,
+only one thread is allowed to operate on a particular address, except in the
+case of downgrade requests from the parent. In this case, a thread is created to
+handle the downgrade request from the parent even if there is another thread
+handling a request for the same address from a child, if the latter thread is
+just waiting for a response from the parent. This is done to avoid deadlocks
+like the ones shown in Figure \ref{unknown}.
+
+\begin{figure}\small
+\begin{requirement}
+Starvation-freedom: A thread that keeps becoming ready to execute will
+eventually be scheduled to execute.\label{starvation}
+\end{requirement}
+\end{figure}
+
+Local starvation-freedom is required in each cache node. If a cache node never
+executes a particular suspended thread, even if it keeps getting ready to
+execute, then the thread will not progress, leading to a deadlock.
+
+A scheduler can be implemented as described in Figure \ref{scheduler}, obeying
+Requirements \ref{pHandleReq} and \ref{cHandleReq}. It should perform some form of
+fair arbiration among its suspended threads to avoid starvation (Requirement
+\ref{starvation}).
+
+\newcommand{\lWhile}{\textbf{while}}
+\newcommand{\lIf}{\textbf{if}}
+\newcommand{\lElsIf}{\textbf{else if}}
+\newcommand{\lElse}{\textbf{else}}
+
+\floatstyle{plain}
+\restylefloat{figure}
+
+\begin{figure}
+\small
+\begin{boxedminipage}{\linewidth}
+\begin{alltt}
+\normalfont
+\lWhile{} (\(True\)) \bopen
+      \lIf (\Resp{c}{n}{a}{x} is in the input channel,
+            where \(n = c.parent\)) \bopen
+            \lIf (suspended thread \(t\) is waiting for 
+                 a response from \(c\) for address \(a\)) \bopen
+                   \resume{} \(t\);
+            \bclose \lElse \bopen
+                   \receive{} \Resp{c}{n}{a}{x};
+                   \start{} \dRespL(\(c, n, a, x\));
+            \bclose
+      \bclose \lElsIf (\Resp{p}{n}{a}{x} is in the input channel,
+            where \(p = n.parent\)) \bopen
+                   // There must be a suspended thread \(t\)
+                   waiting for a response from \(p\) for address \(a\)
+                   \resume{} \(t\);
+            \bclose
+      \bclose \lElsIf (\Req{c}{n}{a}{x} is in the input channel,
+                     where \(n = c.parent\)) \bopen
+             \lIf (Local-Property \ref{pHandleReq} is satisfied) \bopen
+                   \receive{} \Req{c}{n}{a}{x};
+                   \start{} \uReq(\(c, n, a, x\));
+             \bclose
+      \bclose \lElsIf (\Req{p}{n}{a}{x} is in the input channel,
+                  where \(p = n.parent\)) \bopen
+             \lIf (Local-Property \ref{cHandleReq} is satisfied) \bopen
+                   \receive{} \Req{p}{n}{a}{x};
+                   \start{} \dReq(\(p, n, a, x\));
+             \bclose
+      \bclose \lElsIf (suspended thread \(t\) is ready to execute) 
+             \resume{} \(t\);
+\bclose
+\end{alltt}
+\end{boxedminipage}
+\caption{Implementation of a scheduler for cache node $n$}
+\label{scheduler}
+\end{figure}

### 2012-11-22 paper/isca40format/Implementation.tex
before reorging scheduler etc

@@
 \label{sec:implementation}
-In this section we show how the requirements presented in Section
-\ref{sec:network} can be implemented in a realistic system.
+In this section we show how the requirements presented in Section \ref{sec:DistributedMsi} can be implemented in a realistic system.
 
@@
 Since responses from a child of a cache node should not be blocked at the cache
-node by any incoming requests (Requirement\ref{reqNoBlockResp}), the responses
+node by any incoming requests (Requirement \ref{reqNoBlockResp}), the responses
 from a cache node at a lower level (children) to a cache node at a higher level
-(parent) should not use the same share use any of the request channels. Since a
+(parent) should not use any of the request channels. Since a
 response is always removed by the system, just one channel, $RespChannel$, is
@@
 
-\subsubsection{Ensuring the ordering requirements}
+\subsubsection{Ensuring the ordering requirements:}
 There are no ordering requirements between requests sent from the same source
@@
 \ref{reqNoOvertakeResp} for messages from a child to its parent is difficult,
-since it's difficult to guarantee point-to-point FIFO ordering the messages
-come on different virtual channels.
+since it's difficult to guarantee point-to-point FIFO ordering of the messages
+coming on different virtual channels.
 
@@
 this works is as follows (we do not give the formal proof here). Requests can
-not be sent from a child when there is a pending response from the parent
-. Similarly, a response can not be sent from
+not be sent from a child when there is a pending response from the parent. Similarly, a response can not be sent from
 a child when there is a pending response from the parent. So, a response from the parent can never be in flight
@@
 used, one for downgrade-to-$S$ responses ($RespChannel_S$) and one for
-downgrade-to$I$ responses ($RespChannel_I$), then using the response sent later
+downgrade-to-$I$ responses ($RespChannel_I$), then using the response sent later
 will not block an earlier response.
@@
 
-\subsubsection{MSHRs to implement threads}
+\subsubsection{MSHRs to implement threads:}
 The threads we describe essentially represent state stored in Miss Status Handling Registers (MSHRs). To
@@
 only for responses (both from children and the parent) and one general purpose
-MSHR. We used the terms threads to describe our protocol because we give the
+MSHR. We used the term threads to describe our protocol because we give the
 illusion of only one thread executing in the system though the threads are

### 2012-11-22 paper/isca40format/Introduction.tex
before reorging scheduler etc

@@
 though such protocols have been around for two decades (see, for example,
-DASH\cite{DASH}, Alewife\cite{Anant}) correct design and implementation of
+\cite{DASH, Anant, c1, c2, c3, c4, c5}) correct design and implementation of
 such protocols remains a challenge. It is difficult to convince oneself that
@@
 Every cache, unless it is an L1 cache, has a directory which contains
-information about the state of its children caches. The cache state-changes to
+information about the state of its children caches. The changes in the cache states to
 maintain coherence are orchestrated by passing requests and responses between
 the caches using the local directory information.  Unlike bus-based protocols
-which lock the bus till all the required state transitions are done thereby
+which conceptually lock the bus till all the required state transitions are done thereby
 ensuring atomicity of state transitions, in a distributed protocol,
@@
 a response and thus, our threads can never wait indefinitely.  It has been
-shown using the Coq theorem prover \cite{}, that the distributed protocol we
+shown using the Coq theorem prover \cite{coq}, that the distributed protocol we
 derive is correct.  Both the transformation procedure and the proof are
@@
 downgrades. Thus, how or why a particular cache state implies that a particular
-address may not be present in any other cache stems from protocol specific
+address may not be present in any other cache stems from a protocol specific
 ``compatibility of state'' check.
@@
 The main contributions of this paper are: 1. An automatic procedure to refine a
-centralized MSI protocol into protocols for
-distributed, hierarchical, cache system with packet-switching; 2. Specification
-of the requirements that ensure the correctness of the generated protocol;
-3. A discussion of the invariants that simplifies the proof of correctness of
-the protocol considerably;
-4. Giving a provable lower bound on the number of network virtual channels and
-other resources to avoid deadlocks; and 5. Showing that our procedure does not
-introduce any overhead over direct implementation of these protocols. We think
+centralized MSI protocol into a protocol for
+distributed, hierarchical cache system with packet-switching; 2. Specification
+of requirements for network, resources and local scheduling to ensure the correctness of the generated protocol;
+3. A provable lower bound on the number of virtual channels and storage
+ to avoid deadlocks and showing that these requirements do not introduce any overhead over published implementations. We think
 that our method of designing protocols substantially reduces the required
-effort in implementing provably correct protocols.
+effort in implementing provably correct distributed protocols.
 
@@
 protocols. Section \ref{sec:DistributedMsi} systematically transforms the
-protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. Section
-\ref{sec:network} specifies the network, resource and other requirements that
-are needed for the distributed MSI protocol. We also give the minimum virtual
-channel requirements to ensure the requirements specified in Section
-\ref{sec:network}. In Section \ref{sec:implementation}, we show that these formally verified family of protocols
-require no extra state bits, nor extra message transfers compared to other
-published implementations. Section \ref{sec:properties} gives the local
+protocol in Section \ref{sec:GlobalMsi} into a distributed protocol. We specify
+the assumptions of the network, and other requirements that are needed for the
+distributed MSI protocol. We also give the minimum virtual channel requirements
+to ensure that the network assumptions are satisfied. In Section
+\ref{sec:implementation}, we show that these formally verified family of
+protocols require no extra state bits, nor extra message transfers compared to
+other published implementations. Section \ref{sec:properties} gives the local
 properties of the distributed protocol that we rely on to easily establish its
-correctness, and show how it can be generalized to design other protocols like MOSI.
-%It also lists core properties of the distributed protocol that are needed to establish the invariants.
+correctness, and show how it can be generalized to design other protocols like
+MOSI.
+%It also lists core properties of the distributed protocol that are needed to
+%establish the invariants.
 In Section \ref{sec:related}, we present some of the related work and finally
-in Section \ref{sec:conclusion}, we present some directions for future work.
+in Section \ref{sec:conclusion}, we present some conclusions.

### 2012-11-22 paper/isca40format/Invariants.tex
before reorging scheduler etc

@@
-\section{Local properties that ensure global invariants}
+\section{Intuition underlying the automatic implementation}
 \label{sec:properties}
 
-In this section, we will show some of the core properties that are extracted
-from the requirements spelt out in Section \ref{sec:network} as well as the
-implementation of the distributed protocol's auxiliary functions given in Figure
-\ref{realistic}. These properties can be independently used to prove the
-invariants presented in Section \ref{sec:DistributedMsi} (Invariants
-\ref{conservative} to \ref{cpSame}). We do not present the proof here, but give
-some intuition. Figure \ref{sendReq} lists these local properties.
+In this section, give the intuition on how we ensured that the implementation of
+the hierarchical MSI protocol is correct.
+
+The most important invariant that the generated protocol maintains between the
+caches and its parent is as follows:
+\begin{theorem}
+Conservative: For two nodes $p, c$, where $p = c.parent$, $\forall a,
+p.dir[c][a] \ge c.state[a]$.  \label{conservative}
+\end{theorem}
+
+This ensures that if a cache makes a decision about compatibility of states of
+its children using the local directory information, then it does not violate the
+compatibility of the children's real states. This essentially ensures that
+reading the directory is sufficient to know the value of a cache's state.
+
+Regarding the \send{} and \receive{} operations, the following
+invariants are very critical to ensure that a thread completes. If they are not
+true, then a thread might block waiting for a response, or being unable to send a
+message, leading to a deadlock. Our generated protocol maintains this.
+
+\begin{theorem}
+It a thread has to send a message, it will eventually be able to send the
+message.
+\label{canSend}
+\end{theorem}
+
+\begin{theorem}
+If a cache sends an upgrade-to-$x$ request to its parent, it will eventually
+get an upgrade-to-$y$ response from the parent where $y \ge x$.
+\label{willRecv}
+\end{theorem}
+
+\begin{theorem}
+If a cache sends an downgrade-to-$x$ request to one of its children, it will
+eventually get a downgrade-to-$y$ response from the child where $y \le x$.
+\label{willRecv2}
+\end{theorem}
+
+Another scenario where a thread can potentially get suspended is if handling of
+a request requires eviction of another cache line. The thread would get
+suspended if all the lines that can be allocated for this request are being 
+accessed by other suspended threads. But the generated protocol still guarantees
+the following invariant.
+
+\begin{theorem}
+If a thread is waiting for a line to be allocated, \ie for any address to become
+available for eviction, then some address will eventually become available for
+eviction
+\label{evictDead}
+\end{theorem}
+
+It this is not true, a thread that is waiting for being allocated a free line
+may never proceed, leading to a deadlock.
+
+If a cache receives a response message, it must know whether or not a) it will
+be receiving data, and b) whether it has to update the address with the received
+data, in case it does receive data. The first part can be indicated by the
+message itself, but the second part has to be decided only after knowing the
+state of the cache which sent the response. In this regard, the following two
+invariants are very useful.
+
+\begin{theorem}
+If $p$ sends \Resp{p}{c}{a}{x} to $c$, where $p = c.parent$, then $p.dir[c][a]$
+just before $p$ sends the response is the same as $c.state[a]$ just before $c$
+receives the response.
+\label{pcSame}
+\end{theorem}
+
+\begin{theorem}
+If $c$ sends \Resp{c}{p}{a}{x} to $p$, where $p = c.parent$, then $c.state[a]$
+just before $c$ sends the response is the same as $p.dir[c][a]$ just before $p$
+receives the response.
+\label{cpSame}
+\end{theorem}
+
+We will illustrate the need for Invariants \ref{pcSame}, \ref{cpSame} and
+\ref{conservative} using an example of a 2-level system containing two L1
+caches $c_1$ and $c_2$ with a shared L2 cache $p$.
+
+Invariant \ref{pcSame} and \ref{cpSame} are important to make sure that the
+data transfers happen correctly. Let $p.dir[c_1][a] = S$, $p.dir[c_2][a] = I$,
+$c_1.state[a] = S$, and $c_2.state[a] = I$. $c_2$ gets a store request, and
+sends an upgrade-to-$M$ request to $p$. $p$ receives the request, but assumes that
+$c_1$ already has data for $a$ since $p.dir[c_1][a] = S$. Invariant
+\ref{pcSame} is violated. So, it does not transfer the data to $c_1$, resulting
+in $c_1$ waiting for data, and hence deadlock. A dual condition can be
+constructed similarly, in which $c_1$ downgrades its state for address $a$ from
+$S$ to $I$, and the parent assumes that $c_1$ is in state $M$, and keeps
+waiting for data.
+
+Invariant \ref{conservative} is important because a parent node consults its
+directory to decide the children it needs to send a downgrade request to.  Let
+$p.dir[c_1][a] = S$, $c_1.state[a] = S$, $p.dir[c_2][a] = I$ but
+$c_2.state[a] = S$. If $c_1$ gets a store request from the processor and sends
+an upgrade-to-$M$ request to $p$, $p$ will assume that $c_2$'s state for address $a$
+is also $I$ and hence send an upgrade-to-$M$ response to $c_1$. But since the real
+state of  $c_2$ for address $a$ is $S$, it will break Single-writer Invariant
+\ref{singleWriter}.
+
+Thus, if all the invariants above (\ref{conservative}, to \ref{cpSame}) hold,
+then the threads given in Figure \ref{realistic} cannot deadlock because it
+will always be able to a) send a request, b) send back a response, c) get back
+a response for a request it sent, d) receive data when required. Though the
+threads executing in a cache are actually suspensive, these invariants give an
+
+We will show some of the core properties that are extracted
+from the requirements spelt out in Section \ref{sec:DistributedMsi} and the
+specification of the protocol (Figures \ref{msi-template} and
+\ref{realistic}). These properties alone can be used to prove the above
+invariants (Invariants \ref{conservative} to \ref{cpSame}). We do not present
+the proof here, but give some intuition. Figure \ref{sendReq} lists these local
+properties.
 
@@
 \end{inv}
-%\caption{Local-Properties for sending responses}
-%\label{sendResp}
-%\end{figure}
-%
-%\begin{figure}\small
+\caption{Local-Properties for sending responses}
+\label{sendResp}
+\end{figure}
+
+The properties in Figure \ref{sendResp} restrict when responses can be sent by
+ a cache (either to its children or to its parent).
+
+\begin{figure}\small
 \begin{inv}
@@
 \end{inv}
-%\caption{Local-Properties for state changes}
-%\label{stateChange}
-%\end{figure}
-%
-%\begin{figure}\small
+\caption{Local-Properties for state changes}
+\label{stateChange}
+\end{figure}
+
+The properties in Figure \ref{stateChange} specify exactly when the coherence state of
+a line (both directory and state) changes.
+
+\begin{figure}\small
 \begin{inv}
@@
 A request can be sent by a cache node $n$ for address $n$ to another cache node $m$ only if
-no $n$ is not waiting for a response for address $a$ from $m$.
+node $n$ is not waiting for a response for address $a$ from $m$.
 \label{nodoublereq}
@@
 upgrade-$c$-to-$O$ in addition to a downgrade-to-$S$ request to that cache.
-This upgrade-$c$-to-$O$ + downgrade-to-$S$ keeps propogating down the tree till
+This upgrade-$c$-to-$O$ + downgrade-to-$S$ keeps propagating down the tree till
 it reaches a node $n$ whose children are all in a state $< O$. $n$ then
@@
 delivered to the intended destination ($c$ in our case), then $c$ sends back an
-ACK. $c$'s parent is allowed to to start sending downgrade requests for that
+ACK. $c$'s parent is allowed to start sending downgrade requests for that
 address only after receiving this ACK. This ACK is sent back to each of the
@@
 downgrade requests for that address towards $c$.  This satisfies all the
-requirements of Section \ref{sec:network} and the local properties given in
-Figure \ref{sendReq}, thereby automatically being correct. The reasoning for
+requirements of Section \ref{sec:DistributedMsi} and the local properties given in
+Figures \ref{sendResp}, \ref{stateChange} and \ref{sendReq}, thereby automatically being correct. The reasoning for
 its correctness gets localized to checking if the ordering Requirement
 \ref{reqNoOvertakeResp} is maintained.
+
+The protocols can also be extended in a different dimension. The proof is
+oblivious to the states, so any downgrade of a state is treated the same. This
+means that a cache line can downgrade from $M$ to $S$ state instead of going all
+the way down to $I$ state. This flexibility can potentially be exploited, and we
+will study its effects in the future.

### 2012-11-22 paper/isca40format/Msi.tex
before arvind is going to re-org
before reorging scheduler etc

@@
 
-\begin{wrapfigure}{r}{.6\linewidth}
-\begin{subfigure}{.55\linewidth}
+%\begin{wrapfigure}{r}{.6\linewidth}
+\begin{figure}
+\begin{subfigure}{.4\linewidth}
 \centering
-\begin{tabular}{|ccc|}
+\begin{tabular}{|c|c|}
 \hline
-$M$ & $\rightarrow$ & $I$\\
-$S$ & $\rightarrow$ & $S$\\
-$I$ & $\rightarrow$ & $I$\\
+$M$ & $I$\\
+$S$ & $S$\\
+$I$ & $M$\\
 \hline
@@
 \end{subfigure}
-\begin{subfigure}{.24\linewidth}
+\begin{subfigure}{.4\linewidth}
 \centering
@@
 \label{funcs}
-\end{wrapfigure}
+%\end{wrapfigure}
+\end{figure}
 In Figure \ref{funcs} we define some auxiliary functions $maxCompatible(x)$ and $isModified(x)$ which are useful in describing protocols to maintain the invariant.
@@
 
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
 \begin{algorithmic}
+\State //$this$ cache's upgrade request handler. Handles request from $c$ to upgrade $c$'s state to $x$ for address $a$
 \Proc {\uReq}{$c , a, x$}
+  \If {$x > c.state[a]$}
     \If {$a$ is not present in $this$}
@@
   \State \;\;\;\; \call{} $this$.\uResp{}($c, a, x$);
+  \EndIf
 \EndProc
-\end{algorithmic}
-\end{boxedminipage}
-\caption{$this$ cache's upgrade request handler -- request from one of $this$' children, $c$, to upgrade $c$'s state to $x$ for address $a$ }
-\label{atomicUReq1}
-\end{subfigure}
-
-\begin{subfigure}{\linewidth}
-\begin{boxedminipage}{\linewidth}
-\begin{algorithmic}
+\State //$this$ cache's downgrade request handler. Handles request from 
+$p$ to downgrade $this$' state to $x$ for address $a$ 
 \Proc {\dReq}{$p , a, x$}
@@
     \State \textbf{forall} ($c' \in this.children$)
-    \State \;\;\;\; \textbf{if} (c'.state[a] > x)
+    \State \;\;\;\; \textbf{if} ($c'.state[a] > x$)
     \State \;\;\;\;\;\;\;\; \call{} $this.$\dReqL($c', a, x$);
@@
 \end{algorithmic}
-\end{boxedminipage}
-\caption{$this$ cache's downgrade request handler -- request from $this$'
-parent, $p$, to upgrade $p$'s state to $x$ for address $a$ }
-\label{atomicDReq1}
-\end{subfigure}
-
-\caption{MSI protocol}
+\caption{MSI protocol. In all the methods $p$ is the parent of $this$, $c$ is a child of $this$, and the methods work on address $a$ to change its state to $x$}
 \label{msi-template}
@@
 $c$'s state, and send the data if $c$'s state was $I$ previously. This is shown
-in Figure \ref{atomicUReq1}.
+in Figure \ref{msi-template}.
 
-Any cache except the LLC can get a downgrade request from its parent. Figure \ref{atomicDReq1} defines this downgrade method \dReq{} where the cache
+Any cache except the LLC can get a downgrade request from its parent. Figure \ref{msi-template} also defines this downgrade method \dReq{} where the cache
 ensures that each of its children's state is not higher than $x$, by invoking
@@
 \EndProc
-
-\State // In case of LLC, $this$ reads data from memory
-\Proc {\sendMem{}}{$a$}
-\EndProc
 \end{algorithmic}
@@
 
-The methods to handle upgrade and downgrade requests in turn use 5 auxiliary methods defined in Figure \ref{atomic}.
+The methods to handle upgrade and downgrade requests in turn use four auxiliary methods defined in Figure \ref{atomic}.
 These methods are very simple and could have been inlined.

### 2012-11-22 paper/isca40format/RelatedWork.tex
before reorging scheduler etc

@@
 \label{sec:related}
+
+The work on Manager-Client Pairing \cite{mcp} is closest to our work. It
+presents a unifying methodology for designing multi-tier coherence protocols by
+defining the interactions between levels within a coherence hierarchy to enable
+composition. They define interfaces which stand-alone coherence protocols can
+adhere to maintain coherence. We differ from their work because we generate
+the whole coherence protocol automatically.
+
+Token coherence \cite{token} proposes an alternative framework to build
+coherence protocols, by separating performance from correctness. This avoids
+races in the protocol by having a slow correctness substrate-basedd fall-back
+mechanism in case of a memory race (detected by a time out).
 
@@
 coherence protocols, which is similar to our approach. We take this approach one
-step further by doing automating the procedure.
+step further by automating the procedure.
 
-\cite{Beu} and\cite{edya} present protocols which is similar to the final
+Ladan-Mozes \etal \cite{edya} present a protocol which are similar to the final
 distributed protocol that we present. But they do not present any reasoning
@@
 a set of invariants to generate further protocols.
+
+Other techniques have been proposed which simplifies cache coherence using high
+speed optical connections \cite{lipasti}. They revert back to atomic protocols
+with a high-speed bus to lock addresses.

### 2012-11-29 paper/isca40format/DistributedMsi.tex
before reorging scheduler etc

@@
 should be provided, and this informs $this$ that $c$ has changed its state to
-$x$.
+$x$. It is the designer's responsibility to update $this.dir[c][a]$ and his
+copy of the data if need be.
 \end{enumerate}
@@
 Using the above set of rules, the designer can specify the distributed MSI protocol as shown in
-Figure \ref{realistic}. Handlers \uReq{} and \dReq{} are exactly the same as
+Figures \ref{msi-unsolicited} and \ref{realistic}. Handlers \uReq{} and \dReq{} are exactly the same as
 shown in Figure \ref{msi-template}, except that whenever the state of a child
@@
   \State \send{} \Req{this}{c}{a}{x};
-  \While {$this.dir[c][a] > x$}
+  %\While {$this.dir[c][a] > x$}
   \State \receive{} \Resp{c}{this}{a}{z};
@@
   \State $this.dir[c][a] \gets z$;
-  \EndWhile
+  %\EndWhile
 \EndProc
@@
 
-Now that we have established the vocabulary of threads, we will proceed to
-describe the requirements that a network should guarantee in order to make such
-an implementation possible. We will also show the scheduling algorithm that we
-generate.
+Now that we have established the vocabulary of threads, we will show the
+scheduling algorithm that we generate.
 
-\subsection{Message ordering requirements}
+\subsection{Scheduling algorithm}
+
+\floatstyle{plain}
+\restylefloat{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=.4]{Scheduler}
+\caption{Scheduler or the cache controller}
+\label{Scheduler}
+\end{figure}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+The cache controller, or, extending the thread analogy further, the scheduler
+is shown in Figure \ref{Scheduler} We list the scheduling algorithm that
+enables the designer to have an illusion of non-suspensive threads while
+guaranteeing deadlock freedom in Figure \ref{scheduler}.
+
+\newcommand{\lWhile}{\textbf{while}}
+\newcommand{\lIf}{\textbf{if}}
+\newcommand{\lElsIf}{\textbf{else if}}
+\newcommand{\lElse}{\textbf{else}}
+
+\begin{figure}
+\small
+%\begin{boxedminipage}{\linewidth}
+\begin{alltt}
+\normalfont
+\lWhile{} (\(True\)) \bopen
+      \lIf (\Resp{c}{n}{a}{x} is in the input response channel
+             from children, where \(n = c.parent\)) \bopen
+            \lIf (suspended thread \(t\) is waiting for 
+                  a response from \(c\) for address \(a\)) \bopen
+                   \resume{} \(t\);
+            \bclose \lElsIf ($n$ has a free resource for responses) \bopen
+                   \receive{} \Resp{c}{n}{a}{x};
+                   \start{} \dRespL(\(c, n, a, x\));
+            \bclose
+      \bclose \lElsIf (\Resp{p}{n}{a}{x} is in the input channel
+                       from the parent, where \(p = n.parent\)) \bopen
+                   // There must be a suspended thread \(t\)
+                   waiting for a response from \(p\) for address \(a\)
+                   \resume{} \(t\);
+            \bclose
+      \bclose \lElsIf (suspended thread \(t\) is ready to execute)
+             \resume{} \(t\);
+      \bclose \lElsIf (\Req{c}{n}{a}{x} is in the input request
+                      channel from children, where \(n = c.parent\)) \bopen
+             \lIf ((\textbf{!}(address \(a\) has been chosen for eviction by 
+                       some other thread in \(n\) and has not been evicted) \textbf{or}
+                     (a thread in \(n\) is handling a request for address \(a\))) \textbf{and}
+                    \(n\) has a free resource for requests from children) \bopen
+                   \receive{} \Req{c}{n}{a}{x};
+                   \start{} \uReq(\(c, n, a, x\));
+             \bclose
+      \bclose \lElsIf (\Req{p}{n}{a}{x} is in the input channel
+                        from the parent, where \(p = n.parent\)) \bopen
+             \lIf (\textbf{!}((address \(a\) has been chosen for eviction by 
+                       some other thread in \(n\) and has not been evicted) \textbf{or}
+                     (a thread in \(n\) is handling a request from \(p\)
+                       for address \(a\)) \textbf{or}
+                     (a thread \(t\) in \(n\) is handling a request from child \(c\) for
+                       address \(a\) and \(t\) is not waiting for a response from \(p\))) \textbf{and}
+                    \(n\) has a free resource for requests from parent) \bopen
+                   \receive{} \Req{p}{n}{a}{x};
+                   \start{} \dReq(\(p, n, a, x\));
+             \bclose
+      \bclose
+\bclose
+\end{alltt}
+%\end{boxedminipage}
+\caption{Implementation of a scheduler for cache node $n$}
+\label{scheduler}
+\end{figure}
+
+It should be noted that if a request message arrives, even if there are
+resources available for starting the processing of the message, the request is
+sometimes not handled, \ie the scheduler does not start a new thread for
+handling the message. Intuitively, this is to guarantee that only one thread is
+modifying a particular address in the cache -- till the thread finishes its
+work, no other thread should be allowed to modify that address. If another
+thread is allowed to modify the address, then the assumptions that the first
+thread has about the state of the address will be voided. Either the thread has
+to restart from the beginning, or we can avoid this scenario by not scheduling
+the second thread. The only exception to this rule is when a thread has been
+suspended by sending a request to the parent and another request from the parent
+for the same address is received. In this case, a new thread has to be created
+for the request from the parent to avoid deadlocks like the ones shown in Figure
+\ref{unknown}.
+
+The scheduler must also ensure that it does some fair arbitration of threads
+that are suspended. If a particular address remains suspended forever, it will
+lead to starvation. The static priority that we present in the algorithm -- a
+suspended thread that is ready to run always gets priority over creating new
+threads for request messages, will ensure that no thread gets starved.
+
+\begin{figure}\small
+\begin{requirement}
+An incoming request from the parent should not be blocked by any
+incoming request from its children.\label{cReqNoBlockPReq}
+\end{requirement}
+\begin{requirement}
+A response should not be blocked by a request.
+\label{reqNoBlockResp}
+\end{requirement}
+\caption{Blocking restrictions}
+\label{blocking}
+\end{figure}
+
+Note that we present three incoming channels to the scheduler and three incoming
+channels to the scheduler. This is to ensure that a message from one of these
+channels does not block a message in another queue. The exact requirements for
+which messages are allowed to block are stated in Figure \ref{blocking}.
+
+To avoid blocking, a scheduler has to have at least one dedicate resource
+(MSHR) for for handling requests and responses from the parent, and one for
+handling responses from children. There should at least be one more since
+requests from children can not use the other two.
+
+The channels that we depict in Figure \ref{Scheduler} are all FIFO queues. But
+there is an ordering requirement even between queues between requests and
+responses sent from the children. We specify the minimum ordering requirements
+in Figure \ref{order}. Though the figure specifies FIFO requirements for every
+address, an implementation which ensures the corresponding FIFO requirements
+overall, even between addresses, will guarantee the stated requirements.
+
+\subsection{Illustrating the need for FIFO queues and blocking restrictions}
+We will illustrate the need to avoid blocking using the example of a 2-level
+cache hierarchy with two L1 caches $c_1$ and $c_2$, and one shared L2 cache
+$p$. Each core can have several outstanding memory requests.
+
+Let's say both $c_1.state[a] = p.dir[c_1][a] = S$ and $c_2.state[a] =
+p.dir[c_2][a] = S$. Both $c_1$ and $c_2$ get store requests from the processor
+for address $a$, and they send upgrade-to-$M$ requests to $p$.  $p$ receives the
+request from $c_2$ first. $p$ sends a downgrade-to-$I$ request to $c_1$.
+
+\floatstyle{plain}
+\restylefloat{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=0.4]{unknown}
+\caption{Effect of violating Requirement \ref{cReqNoBlockPReq} (requests blocking responses from parent)}
+\label{unknown}
+\end{figure}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+Let's say some request from the core is blocking $c_1$ from receiving the
+request. $c_1$ will never respond back to $p$. But $p$ will be waiting in
+procedure \dReqL{}, and will never start handling the request from $c_1$. This
+will create a deadlock.  The same deadlock is also created when $c_1$ is out of
+thread resources to create a thread to handle the downgrade request from $p$.
+Figure \ref{unknown} depicts this scenario.
+
+Let's extend this scenario further. Let $c_1$ send a downgrade-to-$I$ response
+to $p$. Let's assume that the downgrade-to-$I$ response from $c_1$ is being
+blocked. This means $p$ doesn't receive the downgrade-to-$I$ response, leading
+to the same deadlock. This is also the same if there isn't a dedicated resource
+in $p$ to handle the downgrade response.
+
 \floatstyle{boxed}
@@
 
-\subsection{Resource requirements and blocking restrictions}
-
-\begin{figure}\small
-\begin{requirement}
-At a cache node, an incoming request from the parent should not be blocked by any
-incoming request from its children.\label{cReqNoBlockPReq}
-\end{requirement}
-\begin{requirement}
-At a cache node, an incoming response from any of its children should not
-be blocked by any incoming requests.\label{reqNoBlockResp}
-\end{requirement}
-\begin{requirement}
-At a cache node, an incoming response for address $a$ from its parent should not be blocked by
-an incoming request from its child for address $a$. \label{reqBlockResp2}
-\end{requirement}
-\caption{Blocking restrictions}
-\label{blocking}
-\end{figure}
-
-Figure \ref{blocking} shows the restrictions that the input channels should
-obey with respect to an incoming message blocking other incoming messages.
-
-\begin{figure}\small
-\begin{requirement}
-Each cache node has at least one dedicated thread resource for handling only messages
-(requests or responses) from the parent\label{dedicate1}
-\end{requirement}
-\begin{requirement}
-Each cache node has at least one dedicated thread resource for handling only responses
-(from children or from the parent)\label{dedicate2}
-\end{requirement}
-\begin{requirement}
-Each cache node has at least three thread resources\label{dedicate3}
-\end{requirement}
-\caption{Dedicated resources in a cache node}
-\label{dedicated}
-\end{figure}
-
-Figure \ref{dedicated} shows the minimum number of thread resources that should
-be present in each cache node. Lack of required thread resouces leads to the same
-kind of deadlock scenarios as incorrect blocking does. This is illustrated in the
-following:
-
-Let's say both $c_1.state[a] = p.dir[c_1][a] = S$ and $c_2.state[a] =
-p.dir[c_2][a] = S$. Both $c_1$ and $c_2$ get store requests from the processor
-for address $a$, and they send upgrade-to-$M$ requests to $p$.  $p$ receives the
-request from $c_2$ first. $p$ sends a downgrade-to-$I$ request to $c1$.
-
-\floatstyle{plain}
-\restylefloat{figure}
-\begin{figure}
-\centering
-\includegraphics[scale=0.4]{unknown}
-\caption{Effect of violating Requirement \ref{cReqNoBlockPReq} (requests blocking responses from parent)}
-\label{unknown}
-\end{figure}
-\floatstyle{boxed}
-\restylefloat{figure}
-
-Let's say Requirement \ref{cReqNoBlockPReq} is not held, and instead some request
-from the core is blocking $c_1$ from receiving the request. $c_1$ will never
-respond back to $p$. But $p$ will be waiting in procedure \dReqL{}, and will
-never start handling the request from $c_1$. This will create a deadlock.  The
-same deadlock is also created when $c_1$ is out of thread resources to create a
-thread to handle the downgrade request from $p$ (violating Requirement
-\ref{dedicate1}). Figure \ref{unknown} depicts this scenario.
-
-Let's extend this scenario further. Instead of violating Requirements
-\ref{cReqNoBlockPReq} or \ref{dedicate1}, let $c_1$
-send a downgrade-to-$I$ response to $p$. Let's instead assume that Requirement
-\ref{reqNoBlockResp} is violated in $p$. So, the downgrade-to-$I$ response from
-$c_1$ is not received by $p$, leading to the same deadlock. This is the same if
-Requirement \ref{dedicate2} is violated and $p$ is out of thread resources to
-create a thread to handle the downgrade response.
-
-The need for Requirement \ref{dedicate3} can be shown easily. Let's say that a
-cache has exactly 2 thread resources, one for handling messages from parents,
-and one for handling responses from either the parent or the children. Now, if
-it receives a request from one of its children, it can not use either of the
-thread resources. Thus, at least one more thread resource is needed for
-handling requests from children. This thread resource can also be used for
-handling other types of incoming message.
-
-\subsection{Scheduler requirements}
-
-We will now list the requirements that a scheduler must obey to give the
-illusion of non-suspensive threads and avoid deadlocks.
-
-\begin{figure}\small
-\begin{requirement}
-$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
-if either
-\begin{enumerate}
-\item address $a$ has been chosen for eviction by some thread and that thread
-has not yet replaced $a$, or
-\item a thread in $p$ is handling a request for address
-$a$ from any source
-\end{enumerate}
-\label{pHandleReq}
-\end{requirement}
-\begin{requirement}
-$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
-if either
-\begin{enumerate}
-\item address $a$ has been chosen for eviction by some thread and that thread
-has not yet replaced $a$, or
-\item a thread in $c$ is handling a request for address
-$a$ from $p$, or
-\item a thread in $c$ is handling a request for address
-$a$ from one of $c$'s children $c'$, but that thread is not waiting for a
-response from $p$
-\end{enumerate}
-\label{cHandleReq}
-\end{requirement}
-\caption{Requirements for receiving a request (used by the scheduler)}\label{recvReq}
-\label{}
-\end{figure}
-
-\begin{figure}\small
-\begin{requirement}
-Address $a$ present in cache node $c$ can not be chosen for eviction by a thread
-executing in $c$ if either
-\begin{enumerate}
-\item address $a$ has been chosen for eviction by some thread and that thread
-has not yet replaced $a$, or
-\item a thread in $c$ is handling a request for address
-$a$ from any source
-\end{enumerate}\label{evict}
-\caption{Requirements for suspending a thread during choosing an address for eviction}
-\end{requirement}
-\end{figure}
-
-Requirements \ref{pHandleReq}, \ref{cHandleReq} and \ref{evict} are needed
-because once a thread has finished up-to a point in its execution, it can not
-re-execute that portion. For instance, consider a thread handling an upgrade
-request from a child. Let's say thread $t$ has progressed to the point where it
-is simply waiting to send an upgrade response to the child. Now, if the line
-gets evicted because it was chosen for replacement, or it got downgraded by the
-parent, then thread $t$ can not re-execute its procedure again. To prevent this,
-only one thread is allowed to operate on a particular address, except in the
-case of downgrade requests from the parent. In this case, a thread is created to
-handle the downgrade request from the parent even if there is another thread
-handling a request for the same address from a child, if the latter thread is
-just waiting for a response from the parent. This is done to avoid deadlocks
-like the ones shown in Figure \ref{unknown}.
-
-\begin{figure}\small
-\begin{requirement}
-Starvation-freedom: A thread that keeps becoming ready to execute will
-eventually be scheduled to execute.\label{starvation}
-\end{requirement}
-\end{figure}
-
-Local starvation-freedom is required in each cache node. If a cache node never
-executes a particular suspended thread, even if it keeps getting ready to
-execute, then the thread will not progress, leading to a deadlock.
-
-A scheduler can be implemented as described in Figure \ref{scheduler}, obeying
-Requirements \ref{pHandleReq} and \ref{cHandleReq}. It should perform some form of
-fair arbiration among its suspended threads to avoid starvation (Requirement
-\ref{starvation}).
-
-\newcommand{\lWhile}{\textbf{while}}
-\newcommand{\lIf}{\textbf{if}}
-\newcommand{\lElsIf}{\textbf{else if}}
-\newcommand{\lElse}{\textbf{else}}
-
-\floatstyle{plain}
-\restylefloat{figure}
-
-\begin{figure}
-\small
-\begin{boxedminipage}{\linewidth}
-\begin{alltt}
-\normalfont
-\lWhile{} (\(True\)) \bopen
-      \lIf (\Resp{c}{n}{a}{x} is in the input channel,
-            where \(n = c.parent\)) \bopen
-            \lIf (suspended thread \(t\) is waiting for 
-                 a response from \(c\) for address \(a\)) \bopen
-                   \resume{} \(t\);
-            \bclose \lElse \bopen
-                   \receive{} \Resp{c}{n}{a}{x};
-                   \start{} \dRespL(\(c, n, a, x\));
-            \bclose
-      \bclose \lElsIf (\Resp{p}{n}{a}{x} is in the input channel,
-            where \(p = n.parent\)) \bopen
-                   // There must be a suspended thread \(t\)
-                   waiting for a response from \(p\) for address \(a\)
-                   \resume{} \(t\);
-            \bclose
-      \bclose \lElsIf (\Req{c}{n}{a}{x} is in the input channel,
-                     where \(n = c.parent\)) \bopen
-             \lIf (Local-Property \ref{pHandleReq} is satisfied) \bopen
-                   \receive{} \Req{c}{n}{a}{x};
-                   \start{} \uReq(\(c, n, a, x\));
-             \bclose
-      \bclose \lElsIf (\Req{p}{n}{a}{x} is in the input channel,
-                  where \(p = n.parent\)) \bopen
-             \lIf (Local-Property \ref{cHandleReq} is satisfied) \bopen
-                   \receive{} \Req{p}{n}{a}{x};
-                   \start{} \dReq(\(p, n, a, x\));
-             \bclose
-      \bclose \lElsIf (suspended thread \(t\) is ready to execute) 
-             \resume{} \(t\);
-\bclose
-\end{alltt}
-\end{boxedminipage}
-\caption{Implementation of a scheduler for cache node $n$}
-\label{scheduler}
-\end{figure}
+\subsection{Suspension due to eviction}
+A thread can also get suspended if it needs to evict another line because of
+lack of space in the cache for the address corresponding to the thread. This is
+because, an address that is currently being processed by another thread should
+not be evicted. Another point to note is that an address which is already
+selected for eviction, but is waiting for the children to downgrade to $I$,
+should not be selected for eviction by another address.

### 2012-11-29 paper/isca40format/Implementation.tex
before reorging scheduler etc

@@
 Since responses from the parent $p$ of a cache node $c$ should not be blocked
-at the cache node by a request from $c$'s child $c'$ (Requirement
-\ref{reqBlockResp2}), these responses can not use the same channel as that used
-by a request from $c'$ to $c$. However, they can use the same virtual channel
-used by the requests from $p$ to $c$, $ReqChannel\langle x, ToC \rangle$, where
-$x$ is the level of $p$. This is because no blocking restrictions of Figure
-\ref{blocking} are violated.
+at the cache node by a request from $c$'s child $c'$, these responses can not
+use the same channel as that used by a request from $c'$ to $c$. However, they
+can use the same virtual channel used by the requests from $p$ to $c$,
+$ReqChannel\langle x, ToC \rangle$, where $x$ is the level of $p$. This is
+because no blocking restrictions of Figure \ref{blocking} are violated.
 
@@
 \subsubsection{Bits required for the directory:}
+\floatstyle{plain}
+\restylefloat{figure}
 \begin{wrapfigure}{r}{.4\linewidth}
@@
 encoding.
-
-\subsubsection{MSHRs to implement threads:}
-The threads we describe essentially represent state stored in Miss Status Handling Registers (MSHRs). To
-satisfy the requirements of Figure \ref{dedicated}, there must at least be 3
-MSHR registers, in which one can be used only for messages from the parent, one
-only for responses (both from children and the parent) and one general purpose
-MSHR. We used the term threads to describe our protocol because we give the
-illusion of only one thread executing in the system though the threads are
-actually suspensive. Hence the software terminology fits our model quite well.

### 2013-12-04 paper/Introduction.tex
added conflict for modules with parameters

--- 2012-10-30 paper/isca40format/Body.tex
@@
 \section{Introduction}
-
-This document provides the formatting instructions for submissions to the 40th
-Annual International Symposium on Computer Architecture,
-2013. In an effort to respect the efforts of reviewers and in
-the interest of fairness to all prospective authors, we request that all
-submissions to ISCA-40 follow the formatting and submission rules detailed
-below.  Submissions that violate these instructions may not be
-reviewed, at the discretion of the program chair, in order to maintain a review
-process that is fair to all potential authors.  This is a generous format,
-with plenty of space -- there should be no need to tweak it in any
-significant way.
-
-An example submission (formatted using the ISCA-40 submission format) that
-contains the submission and formatting guidelines can be downloaded from
-the conference website at {\em http://isca2013.eew.technion.ac.il}.
-
-All questions regarding paper formatting and submission should be directed to
-the program chair.
-
-\section{Preparation Instructions}
-
-\subsection{Paper Formatting}
-
-All submissions should contain a maximum of 12 pages of single-spaced
-two-column text. If you are using \LaTeX~\cite{lamport94} 
-to typeset your paper, then we 
-strongly suggest
-that you use the template available at
-http://isca2013.eew.technion.ac.il -- this
-document was prepared with that template.  
-If you are using a different
-software package to typeset your paper, then please adhere to the guidelines
-given in Table~\ref{table:formatting}.
-
-\begin{table}[h!]
-  \centering
-  \begin{tabular}{|l|l|}
-    \hline
-    \textbf{Field} & \textbf{Value}\\
-    \hline
-    \hline
-    Page limit & 12 pages\\
-    \hline
-    Paper size & US Letter 8.5in $\times$ 11in\\
-    \hline
-    Top margin & 1in\\
-    \hline
-    Bottom margin & 1in\\
-    \hline
-    Left margin & 0.75in\\
-    \hline
-    Right margin & 0.75in\\
-    \hline
-    Separation between columns & 0.25in\\
-    \hline
-    Body font & 10pt\\
-    \hline
-    Abstract font & 10pt, italicized\\
-    \hline
-    Section heading font & 12pt, bold\\
-    \hline
-    Subsection heading font & 10pt, bold\\
-    \hline
-    Caption font & 9pt, bold\\
-    \hline
-    References & 8pt\\
-    \hline
-  \end{tabular}
-  \caption{Formatting guidelines for submission. }
-  \label{table:formatting}
-\end{table}
-
-\textbf{Please ensure that you include page numbers with your
-submission}. This makes it easier for the reviewers to refer to
-different parts of your paper when they provide comments.
-
-Also, please ensure that your submission has a banner at the top of 
-the title page, similar to this one, which contains the submission
-number and the notice of confidentiality.  If using the template,
-just replace XXX in the template with the submission number
-you receive from the submission website.
-
-If you use bibtex, please note that the references.bib file provided in the
-template example includes some format-specific incantations at the top of the
-file.  If you substitute your own bib file, you will probably want to include
-these incantations at the top of it.
-
-\subsection{Content}
-
-\noindent\textbf{\sout{Author List.}} All submissions are double
-blind. Therefore, please do not include any author names in the
-submission. You must also ensure that the metadata included in the
-PDF does not give away the authors. If you are improving upon your
-prior work, refer to your prior work in the third person and include
-a full citation for the work in the bibliography. For example, if
-you are building on {\em your own} prior work in the papers \cite{nicepaper,nicepaper2}, 
-you would say something like "While the authors of \cite{nicepaper,nicepaper2} did X and Y,
-this paper does X, Y and Z, and is therefore much better."  Do NOT omit or
-anonymize references for blind review.
-
-\noindent\textbf{Figures and Tables.} Ensure that the figures and
-tables are legible.  Please also ensure that you refer to your
-figures in the main text. The proceedings will be printed in
-gray-scale, and many reviewers print the papers in
-gray-scale. Therefore, if you must use colors for your figures, ensure
-that the different colors are highly distinguishable in gray-scale.
-If a figure is not easily understandable in gray-scale, then assume
-it will not be understood by the reviewers.  In many cases, it
-is better to just prepare your documents without color.
-
-\noindent\textbf{Main Body.} Avoid bad page or column breaks in
-your main text, i.e., last line of a paragraph at the top of a
-column or first line of a paragraph at the end of a column. If you
-begin a new section or sub-section near the end of a column,
-ensure that you have at least 2 lines of body text on the same
-column.
-
-\section{Submission Instructions}
-
-\subsection{Paper Authors}
-
-Declare all the authors of the paper upfront. Addition/removal of authors once
-the paper is accepted will have to be approved by the program chair.  The
-paper selection process is carefully run in a way that maximizes fairness
-by seeking to eliminate all conflicts of interest.  Late changes to author 
-lists can invalidate that process.
-
-\subsection{Declaring Conflicts of Interest}
-
-The authors must register all their conflicts into the paper submission site.
-Conflicts are needed to resolve assignment of reviewers. 
-Please get the conflicts
-right.  You have a week between the registration of the paper and final
-submission -- there is no need to do the conflicts in a rush at the last
-second. If a paper is found to
-have an undeclared conflict that causes a problem OR if a paper is found to declare false conflicts in order to abuse or ``game'' the review system, the paper may be rejected.
-
-Please declare a conflict of interest (COI) if any of the following exist for any author of a paper:
-
-\begin{enumerate}
-\item Your Ph.D. advisor and Ph.D. students forever.
-\item Family relations by blood or marriage forever (if they might be potential reviewers).
-\item Other past or current advisors
-\item People with whom you collaborated in the last five years. Collaborators include:
-\begin{itemize}
-\item Co-authors on an accepted/rejected/pending research paper, 
-\item Co-PIs on an accepted/rejected/pending grant, 
-\item Those who are funders (decision-makers) regarding your research grants, and researchers whom you fund. 
-\end{itemize}
-You many exclude ``service'' collaborations like writing a CSTB report or serving on a program committee together.
-\item People who shared your primary institution in the last five years. Note that if either you or they have moved, there could be several institutions to consider.
-\end{enumerate}
-
-There may be others with whom you know a conflict of interest exists.  However,
-you will need to justify this conflict of interest.  Please be reasonable.  For
-example, just because a reviewer works on similar topics as the paper you are
-submitting is on, you cannot declare a conflict of interest with them. The PC Chair may contact co-authors to explain COIs whose origin is not clear.
-
-You will have to declare all conflicts with PC members as well as non-PC members 
-with whom you
-have a conflict of interest.  When in doubt, contact the program chair.
-
-\subsection{Concurrent Submissions and Resubmissions of Already Published Papers}
-
-By submitting a manuscript to ISCA-40, the authors guarantee that the
-manuscript has not been previously published or accepted for publication in a
-substantially similar form in any conference or journal. The authors also
-guarantee that no paper which contains significant overlap with the
-contributions of the submitted paper is under review to any other conference or
-journal or workshop, or will be submitted to one of them during the ISCA-40
-review period. Violation of any of these conditions will lead to rejection.
-
-Extended versions of papers accepted to IEEE Computer Architecture Letters can
-be submitted to ISCA-19.  If you are in doubt, contact the program chair.
-
-Finally, we also note that the ACM Plagiarism Policy ({\em http://www.acm.org/publications/policies/plagiarism\_policy}) covers a range of ethical issues concerning the misrepresentation of other works or one's own work.
+\label{sec:Introduction}

### 2013-12-04 paper/Title.tex
added conflict for modules with parameters

%\begin{titlepage}

\title{Stuff
}

%       {\Large (Regular Submission)}}

%\author{Names withheld for submission}%\vspace{30pt}}

\author{
    Muralidaran Vijayaraghavan\\Massachusetts Institute of Technology\\vmurali@csail.mit.edu\and
    Nirav Dave\\SRI International\\ndave@csl.sri.com\and
    Arvind\\Massachusetts Institute of Technology\\arvind@csail.mit.edu
}

\date{}

\maketitle % \thispagestyle{empty}

%=========================================================================
%  Abstract
%=========================================================================

\begin{abstract}
\input{Abstract}
\end{abstract}

%\bigskip

%\centerline{\footnotesize {\bf Keywords}: Guarded Atomic Actions, HW Synthesis, Modular Composition}

%\end{titlepage}

### 2014-01-04 paper/Background.tex
added more explanations

\section{Background}

In order to obtain the execution semantics of a multi-threaded program on a
multi-processor system, one has to understand the memory consistency model of the
underlying system. This is necessary because, as opposed to a direct access to
a single-ported memory, the processors a) have store buffers which reorders the
sequence of memory requests from the processor, for faster access (say by bypassing
a store-value to a subsequent load request), and b) communicate with the memory
through several levels of caches, which cache the data for faster access. The
memory consistency model describes the semantic behavior of such requests and
the implementation of the system ensures that the cache hierarchy and the store
buffers do not violate the specified semantics.

Memory consistency models can be partitioned into two sub-parts
\cite{Arvind-memory-model}: a) the store-atomicity property for each
location/address and b) the reordering restrictions between requests to
different locations within each processor.  The store-atomicity property per address
can be informally explained as follows: each location behaves like an atomic
register in which no two writes (stores) to that location can happen
simultaneously, and any read of that location (load) returns the previously
written value in that location. Most memory consistency models require that
the memory subsystem obeys the store-atomicity property, which is a global
property for the whole multi-processor system. Unlike the store-atomicity property,
the reordering restrictions dictated by the memory consistency model pertain to
reordering requests within the same processor, but across multiple addresses. For
instance, it dictates whether a store request to an address $x$ can be issued
to the memory sub-system before a load request to a different address $y$,
which was issued by the same processor earlier (\ie load $y$ happened before store
$x$ is program order, but the order of issue to memory sub-system is reversed).

The mechanism employed by the memory sub-system to ensure store-atomicity per
location, in the presence of caches, is known as \emph{cache coherence}. In the
absence of caches, since there is exactly one copy of the data for each
address, it is easy to ensure cache coherence. The problem comes in the
presence of shared caches -- multiple copies of the same address can exist in
different caches, and if these locations are independently written, then the
same address can end up having different values in two different caches. Each
location would no longer behave like an atomic register unless some sort of
co-ordination happens between the different caches -- this co-ordination is
summed up as the \emph{cache coherence protocol}.

\subsection{Formal definition of store-atomicity}
In this section we give the formal definition of the store-atomicity property.

A request issued by a processor to the memory sub-system will contain the
address corresponding to the request, and a type indicating whether the request
is a load or a store. In case of a store request, it will normally contain the
data to be stored. We augment the request message with two special fields: a
\mylabel which uniquely identifies each request, and an \myindex which
is a natural number denoting the position of a request amongst those emanating
from a particular processor. We omit the data field because it can be subsumed
by the unique label (the data and the unique label together forms another
unique label).

A response issued by the memory sub-system to the processor will contain a
label to identify the request corresponding to the response and the processor
to which the response is sent to. In the case of a load request, the memory
sub-system returns \stl,  the label of the last store to that address. In
case there has been no store to that address, a special value called Initial is
returned. The memory sub-system also returns a natural number, \mytime
which represents the position at which this request has been handled globally
among all the requests issued from all the processors. Both the store label (or
Initial) and the time are needed just for the definition of the store-atomicity
property -- a real system will simply return the data corresponding to the load
request's address.

\begin{figure}
\centering
\begin{tabular}{|p{.02\textwidth}lp{.9\textwidth}|}
\hline
\multicolumn{3}{|c|}{\Request}\\
\hline
&\proc:& Processor issuing the request\\
&\mylabel:& A unique label to identify the request\\
&\loc:& Address associated with the request\\
&\desc:& Request type (\Ld{} or \St)\\
&\myindex:& A unique natural number indicating the local order of requests\\
\hline
\hline
\multicolumn{3}{|c|}{\Response}\\
\hline
&\proc:& Processor which receives the response\\
&\mylabel:& Label identifying the request corresponding to this response\\
&\stl:& The label of the last store to that address (or the special value,
       Initial) in case of a load response\\
&\mytime:& A natural number indicating the global order of (processing the) requests\\
\hline
\end{tabular}
\caption{Data types for request and response}
\mylabel{req-resp}
\end{figure}

The uniqueness of request labels (globally) and request indices (per processor) can be defined as follows:
\begin{defn}
\begin{multline*}
\textit{uniqReqLabels} := \forall q_1, q_2\in\Request, q_1.\mylabel = q_2.\mylabel @-> q_1 = q_2\\
\end{multline*}
\end{defn}
\begin{defn}
\begin{multline*}
\textit{uniqIndices} := \forall q_1, q_2\in\Request,
     q_1.\proc = q_2.\proc @-> q_1.\myindex = q_2.\myindex \\ @-> q_1 = q_2
\end{multline*}
\end{defn}

The responses should obey the following sanity conditions:
\begin{defn} (Each response corresponds to a request.)
\begin{equation*}
\textit{respHasReq} := \forall r\in\Response, \exists q\in\Request, r.\mylabel = q.\mylabel
\end{equation*}
\end{defn}

\begin{defn} (A response comes back to the same processor that sent the corresponding request.)
\begin{multline*}
\textit{respProc} := \forall (r\in\Response) (q\in\Request), r.\mylabel = q.\mylabel \\ @-> r.\proc = q.\proc
\end{multline*}
\newline
\end{defn}

\begin{defn} (The labels in the responses are unique. Since \stl{} field is irrelecant for
store requests, they are checked only for load requests.)
\begin{multline*}
\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response), r_1.\mylabel = r_2.\mylabel @-> \\
\hspace{1cm} r_1.\proc = r_2.\proc \wedge r_1.\mytime = r_2.\mytime \wedge \hfill \\
\forall (q \in \Response), q.\mylabel = r_1.\mylabel @->
q.\desc = \Ld @-> r_1.\stl = r_2.\stl
\end{multline*}
\end{defn}

\begin{defn} (No response can share the same time as another store response.)
\begin{multline*}
\textit{uniqStRespTimes} := 
\forall (r_1, r_2 \in Response), r_1.\mytime = r_2.\mytime @-> \\
\forall (q_1, q_2 \in Request), r_1.\mylabel = q_1.\mylabel @->
r_2.\mylabel = q_2.\mylabel -> \\
q_1.\loc = q_2.\loc @-> q_1.\desc = \St @-> \\ r_1.\mylabel = r_2.\mylabel
\end{multline*}
\end{defn}

In addition, the global ordering of responses to the same processor should not
violate the local ordering of the corresponding requests.
\begin{defn}
\begin{multline*}
\textit{localOrdering} :=
\forall (r_1, r_2 \in \Response) (q_1, q_2 \in \Request), \\ r_1.\mylabel
= q_1.\mylabel @-> r_2.\mylabel = q_2.\mylabel @-> \\ q_1.\proc = q_2.\proc @->
q_1.\loc = q_2.\loc @-> \\ q_1.\myindex < q_2.\myindex @-> r_1.\mytime > r_2.\mytime
  @-> \False
\end{multline*}
\end{defn}

This brings us to the main condition for store atomicity which states that the
store label returned in a load response corresponds to the \emph{last earlier}
store request that has been processed in the system for that address (lastEarlierValue). The
\emph{last} and \emph{earlier} are defined according to the field \mytime{}
returned in the responses. In case no store has happened earlier for a
particular address when a load is processed, then the load response returns a
special value \Initial, which is used to initialize all the addresses (initialValue).

\begin{defn}
\begin{multline*}
\textit{initialValue}:= r.\stl = \Initial \wedge \\
\forall (r' \in \Response) (q' \in \Request),
r'.\mylabel = q'.\mylabel @-> \\ 0 \le r'.\mytime < r.\mytime @->
\neg (q.\loc = q'.\loc \wedge q'.\desc = \St)
\end{multline*}
\end{defn}

\begin{defn}
\begin{multline*}
\textit{lastEarlierValue} :=
\exists m, r.\stl = \Store m \wedge \\
\exists (r_m \in \Response) (q_m \in \Request),
r_m.\mylabel = m \wedge q_m.\mylabel = m \wedge \\
r_m.\mytime < r.\mytime \wedge q_m.\loc = q.\loc \wedge q_m.\desc = \St \wedge  \\
\forall (r' \in \Response) (q' \in \Request),
r'.\mylabel = q'.\mylabel @-> \\ r_m.\mytime < r'.\mytime < r.\mytime ->
\neg (q.\loc = q'.\loc \wedge q'.\desc = \St)
\end{multline*}
\end{defn}

Using the above definitions, we can define the store-atomicity theorem as follows:
\begin{thm}
\begin{multline*}
storeAtomicity := \textit{uniqReqLabels} @-> \textit{uniqIndices} @-> \\
\textit{respHasReq} \wedge \textit{respProc} \wedge \textit{uniqRespLabels} \wedge
\textit{uniqStRespTimes} \wedge \textit{localOrdering}\\ \wedge 
(\textit{initialValue} \vee \textit{lastEarlierValue})
\end{multline*}
\end{thm}

### 2014-01-04 paper/System.tex
added more explanations

\section{System}
\label{sec:System}

The system we are considering consists of a tree hierarchy of caches. These
caches are inclusive, \ie a cache at a higher level (near the root) contains
all the addresses present in caches in the lower levels of its substree. The
root of the communicates with the main memory, and the leaf caches communicate
with the processor (Figure \ref{hier}).

Each cache has a \emph{coherence state} associated with each address. The
coherence state can be one of the three values: \Mo (Modified), \Sh (Shared) or
\In (Invalid). \Mo state denotes that the subtree rooted at the cache in \Mo
state for that address can modify the data, \ie both read and write the data
corresponding to the address. \Sh state denotes that the subtree rooted at the
cache in \Sh state for that address can only read the data corresponding to
that address and \In state denotes that the subtree rooted at the cache in \In
state can neither read nor write the data corresponding to the address. The
meanings of the state is clear, but one must verify the actual cache coherence
protocol for violations of these invariants, which is what the proof in this
paper intends to do. There's a logical ordering between the states, in terms of
the permissions that they have: $\Mo > \Sh > \In$ (the notation should be read
assuming that the order is transitively closed).
In addition, each non-leaf cache maintains a \emph{directory}, which contains a
local version of the state of each of its children for a particular address.
This local version may be stale, but it should not be lower than the real state
of the child (again this property has to be proven for a cache coherence
protocol)

In order for a cache to change to a higher state for a given address, \ie
\emph{upgrade}, it has to \emph{request} its parents for increasing the state.
For a cache to change to a lower state, \ie \emph{downgrade}, it has to notify
its parents indicating the downgrade -- it uses \emph{response} messages for
this purpose. On the other end, for a parent to force a child to downgrade the
child's state for an address, or equivalently, to downgrade the parent's
directory state for the child, it has to \emph{request} that child to
downgrade, and a parent can notify a child about upgrading the child's state,
or equivalently, about upgrading the parent's directory state for the child, by
sending a \emph{response} message.

Each cache is connected to its parent cache using 3 logical connections: a)
\cpReq{} channel to transmit requests from the child to the parent, b) \cpResp{}
channel to transmit responses from the child to the parent, and c) \pc{}
\textbf{FIFO} channel to transmit both request and response messages from
parent to child. In addition, the leaf caches (called L1 caches usually)
communicate with the processor through two channels: a) \cproc{} \textbf{FIFO}
channel to transmit requests from the processor to L1 cache, and b) \procc{}
\textbf{FIFO} channel to transmit responses from the L1 cache to the processor.
There is exactly one of \cproc{} and \procc{} channels between the L1 and the processor.

In this paper, we will assume that each channel has a dedicated link between
caches. In a real system, these channels will be logically implemented over an
interconnect network, which poses additional problems that have to be proven --
namely that of deadlock and starvation-freedom. Though we have proved these
results requiring the network to obey very relaxed and practically feasible
conditions, we completely omit the formal discussion of deadlock and starvation
freedom in this paper because of lack of space. We \xxx{might later} informally
discuss how deadlock and starvation freedom properties are proved for our
system.

Since we have dedicated channel between each pair of child-parent, we will
index the three channels (\cpReq{}, \cpResp{} and \pc{}) by the name of the
child that uses that channel. We will also index the two channels (\cproc{} and
\procc{}) between the processor and L1 cache by the name of the L1 cache.

Figure \ref{format} shows the message format transmitted in the 3 channels
between caches. The messages transmitted between the processor and the L1 cache
is the same as in Figure \ref{req-resp}.
\begin{figure}
\centering
\begin{tabular}{|l|lp{.9\textwidth}|}
\hline
\multirow{2}{*}{\cpReq} & \from: & Current state of child\\
& \myto: & Desired (higher) state of child\\
\hline
\multirow{3}{*}{\cpResp} & \from: & Current state of child\\
& \myto: & Downgraded state of child\\
& \data: & Data associated with the address, if necessary\\
\multirow{4}{*}{\pc} & \type: & \Req{} (for requests) and \Resp{} (for responses)\\
\hline
& \from: & Current state of parent's directory for the child\\
& \myto: & Desired (lower) directory state of a child for \Req{} and upgraded
directory state of a child for \Resp{}\\
& \data: & Data associated with the address if necessary (for \Resp{} only)\\
\hline
\end{tabular}
\caption{Data types messages between caches}
\label{format}
\end{figure}

### 2014-01-04 paper/System.tex
added more explanations

@@
 Each cache has a \emph{coherence state} associated with each address. The
-coherence state can be one of the three values: \Mo (Modified), \Sh (Shared) or
-\In (Invalid). \Mo state denotes that the subtree rooted at the cache in \Mo
-state for that address can modify the data, \ie both read and write the data
-corresponding to the address. \Sh state denotes that the subtree rooted at the
-cache in \Sh state for that address can only read the data corresponding to
-that address and \In state denotes that the subtree rooted at the cache in \In
+coherence state can be one of the three values: \Mo{} (Modified), \Sh{} (Shared) or
+\In{} (Invalid). \Mo{} state denotes that the subtree rooted at the cache in \Mo{}
+state for that address can modify the data, \ie{} both read and write the data
+corresponding to the address. \Sh{} state denotes that the subtree rooted at the
+cache in \Sh{} state for that address can only read the data corresponding to
+that address and \In{} state denotes that the subtree rooted at the cache in \In{}
 state can neither read nor write the data corresponding to the address. The
@@
 of the child (again this property has to be proven for a cache coherence
-protocol)
+protocol).
+
+Though this paper restricts the formal discussion to the simple MSI protocol,
+extension of this proof technique as well as major portions of the proof to
+other protocols like MOSI and MOESI is straightforward.
 
@@
 between caches. The messages transmitted between the processor and the L1 cache
-is the same as in Figure \ref{req-resp}.
+is the same as in Figure \ref{req-resp} (except for the omission of the \proc{}
+field, which is indicated by the channel's index).
 \begin{figure}
@@
 \hline
-\multirow{2}{*}{\cpReq} & \from: & Current state of child\\
+\multirow{3}{*}{\cpReq} & \from: & Current state of child\\
 & \myto: & Desired (higher) state of child\\
+& \addr: & Address associated with the request\\
 \hline
-\multirow{3}{*}{\cpResp} & \from: & Current state of child\\
+\multirow{4}{*}{\cpResp} & \from: & Current state of child\\
 & \myto: & Downgraded state of child\\
+& \addr: & Address associated with the request\\
 & \data: & Data associated with the address, if necessary\\
@@
 \hline
-& \from: & Current state of parent's directory for the child\\
+%& \from: & Current state of parent's directory for the child\\
 & \myto: & Desired (lower) directory state of a child for \Req{} and upgraded
 directory state of a child for \Resp{}\\
+& \addr: & Address associated with the request\\
 & \data: & Data associated with the address if necessary (for \Resp{} only)\\
@@
 \end{figure}
+
+The description of the behavior of each of the caches can be given in terms of
+``rules'' for state transitions. Each rule specifies a series of actions that
+are performed atomically, and during the execution of a rule, no other
+(conflicting) rule interferes with it. Thus, the behavior of any rule-based
+system can be understood as if the rules executed one at a time.
+
+Such a rule-based description can be automatically converted into synchronous
+hardware, for instance using Bluespec \cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
+Bluespec executes each atomic rule into a single hardware cycle, which
+simultaneously executing non-conflicting rules \ie{} rules that do not access the
+same state (the exact conditions for simultaneous scheduling are specified in
+the citations). Though the rules are logically executed atomically, their execution
+can be spread across multiple cycles (as long as other conflicting rules are
+not executed during the whole period of current rule's execution). Synchronous
+hardware generation for multi-cycle atomic rules has also been studied
+\cite{Karczmarek}.

### 2014-01-23 paper/Introduction.tex
all working - reorg
more cleanup before next venture

@@
-\section{Introduction}
+h\section{Introduction}
 \label{sec:Introduction}
+
+Shared memory is a common abstraction for distributed systems as it
+provides an easily understood model of interaction. To achieve this
+abstraction efficiently, many hardware systems implement a cache
+coherence implementation.
+
+Since the correctness of this abstraction is paramount in the correct
+functioning of the system, significant effort is placed in making sure
+that the cache coherence engine is correct. In practice, this is done
+in two steps. First, a cache coherence protocol is constructed and
+formally verified. This protocol generally directly represents many of
+the key concurrent aspects of an implementation, \eg\ reads and
+writes to the system are broken into local atomic microoperations, but
+ignores implementation details \eg\ buffer sizing and memory access
+times. Second, once the protocol has been verified it is handed to
+hardware experts who fill in the missing implementation details and
+realize it at the Register Transfer Level (RTL) of gates, wires, and
+registers. This effort is not a one time translation, as performance
+properties are not easily estimated; a significant amount of
+performance testing is requires to select a good implementation.
+
+From a methodological perspective, this two step process represent a
+significant methodological failure. We spent a significant amount of
+time verifying the protocol and then invalidate any correctness
+guarantees to the implementation by performing not only a non-trivial
+translation from the protocol representation to hardware, but also
+the refinement to add missing implementation details.
+
+Previously, it was shown that, because verified protocols are often
+described in as via Term Rewrite Systems (TRS), one could directly
+transliterate a verified protocol into Bluespec, a hardware
+description language on Guarded Atomic Actions, a restriction of
+TRS. This reduces the correctness of the implementation to the
+correctness of the protocol, effectively removing any implementation
+bugs. This requires the protocol include many implementation-level
+details which had previously been elided for simplicity from the
+protocol so that it is realizable; practically, beyond some additional
+descriptional noise in the proof this is not a show stopper.
+
+The issue with this approach is that design of a hardware system
+requires many design tweaks to reach optimal performance as one cannot
+practically know the correct low-level hardware results without
+evaluating the low-level representation. In this protocol
+transliteration scheme, each of these changes can potentially breaks
+correctness and a full reverification of correctness is needed in
+principle. As hardware designers are unlikely to have the expertise or
+skill this effectively nullifies the verification guarantee in design.
+
+To fully realize this approach of design we need to provide not just a
+single proof of correctness, but a proof framework covering any
+variation of a design that might be of interest for performance
+reasons. This would enable implementation level changes without
+compromising on the ability to verify correctness of the design.
+
+In practice, protocol variants rely on a small set of general
+``building block'' properties, \eg\ messages always get from source to
+destination or that the design's reaction to a protocol message
+results in the appropriate messages being sent, about parts of the
+design to be true for the full system correctness. The details of
+these properties hold vary depend on the precise design variant. For
+instance, to show a message is handled in a finite amount of time it
+is sufficient to check that there is a maximum time bound between
+start and completion; this may rely on the number of messages in
+flight which is in turn affected by buffering size.
+
+In this paper, we describe an approach to provide this proof framework
+for any particular cache coherence protocol via the example of a
+MOESI-style protocol. We show of the modular structure of the proof
+and design can be shared and how modification of microarchtectural
+aspects of design directly translation to reverification of our key
+building block properties which are effectively trivial for the
+microarchitectural variations required.

### 2014-01-23 paper/Introduction.tex
making it cleaner

--- 2013-12-04 paper/Introduction.tex
@@
-\section{Introduction}
+:qh\section{Introduction}
 \label{sec:Introduction}
+
+Shared memory is a common abstraction for distributed systems as it
+provides an easily understood model of interaction. To achieve this
+abstraction efficiently, many hardware systems implement a cache
+coherence implementation.
+
+Since the correctness of this abstraction is paramount in the correct
+functioning of the system, significant effort is placed in making sure
+that the cache coherence engine is correct. In practice, this is done
+in two steps. First, a cache coherence protocol is constructed and
+verified. This protocol generally directly represents many of
+the key concurrent aspects of an implementation, \eg reads and
+writes to the system are broken into local atomic micro-operations, but
+ignores implementation details \eg buffer sizing and memory access
+times. Second, once the protocol has been verified it is handed to
+hardware experts who fill in the missing implementation details and
+realize it at the Register Transfer Level (RTL) of gates, wires, and
+registers. This involves further breaking down of local micro-operations to
+make it realizable in hardware. This effort is not a one time translation, as
+performance properties are not easily estimated; a significant amount of
+performance testing is requires to select a good implementation.
+
+From a methodological perspective, this two step process represent a
+significant methodological failure. We spent a significant amount of
+time verifying the protocol and then invalidate any correctness
+guarantees to the implementation by performing not only a non-trivial
+translation from the protocol representation to hardware, but also
+the refinement to add missing implementation details.
+
+Because verified protocols are often described via Term Rewrite Systems (TRS).
+it was shown by Dave \etal{} that a verified protocol could be transliterated
+into Bluespec, a hardware description language based on Guarded Atomic Actions,
+a restriction of TRS \cite{Hoe:TCAD, DNA:CoherenceImplementation}. This reduces
+the correctness of the implementation to the correctness of the protocol,
+effectively removing any implementation bugs. Though in the current versions of
+the Bluespec language, each atomic action corresponds to one hardware clock
+cycle, there has been some work to reduce this restriction \cite{Karczmarek}.
+%This requires that the protocol description includes many important
+%implementation-level details which had previously been elided for simplicity
+%of the description. Practically, beyond some additional descriptional noise in
+%the proof this is not a show stopper.
+
+The issue with this protocol transliteration approach is that design of a
+hardware system requires many design tweaks to reach optimal performance as one
+cannot practically know the correct low-level hardware results without
+evaluating the low-level representation.
+Each of these changes can potentially breaks
+correctness and a full reverification of correctness is needed in
+principle. As hardware designers are unlikely to have the expertise or
+skill this effectively nullifies the verification guarantee in design.
+
+To fully realize this approach of design we need to provide not just a
+single proof of correctness, but a proof framework covering any
+variation of a design that might be of interest for performance
+reasons. This would enable implementation level changes without
+compromising on the ability to verify correctness of the design.
+
+In practice, protocol variants rely on a small set of general ``local building
+block'' properties, \eg{} messages always get from source to destination or
+that the local cache's reaction to receiving a protocol message results in the
+appropriate messages being sent, the appropriate local state being changed \etc{}.
+% about parts of the
+%design to be true for the full system correctness.
+These properties are easily verifiable since they correspond to a small
+partition of the system, for instance a single cache.
+The details of how these properties hold vary depending on the precise design
+variant. For instance, the reason why some property holds, like the
+point-to-point FIFO ordering for messages transferred between two nodes,
+depends on the type of network; the proof will be very different for a mesh
+network compared to a point-to-point network.
+
+In this paper, we describe an approach to provide this proof framework for any
+particular cache coherence protocol via the example of a MSI-style protocol and
+show how it can be extended to other protocols like MESI and MOESI. We show how
+the modular structure of the proof and design can be shared and how
+modification of micro-archtectural aspects of design directly translates to
+just reverification of our local building block properties. We use Coq for our
+proof framework. We also show how an atomic action based MSI protocol
+description can be directly transliterated into an Inductive definition in Coq,
+and verify its local building block properties formally, thus completing the
+picture.

### 2014-01-23 paper/Background.tex
added more explanations
typo errors
reverted to complete proof

@@
 a store-value to a subsequent load request), and b) communicate with the memory
-through several levels of caches, which cache the data for faster access. The
+through several levels of data caches for faster access. The
 memory consistency model describes the semantic behavior of such requests and
-the implementation of the system ensures that the cache hierarchy and the store
+the system ensures that the cache hierarchy and the store
 buffers do not violate the specified semantics.
@@
 Memory consistency models can be partitioned into two sub-parts
-\cite{Arvind-memory-model}: a) the store-atomicity property for each
-location/address and b) the reordering restrictions between requests to
-different locations within each processor.  The store-atomicity property per address
-can be informally explained as follows: each location behaves like an atomic
-register in which no two writes (stores) to that location can happen
+\cite{Arvind-memory-model}: a) the ``global'' store-atomicity property for each
+location/address and b) the ``local'' reordering restrictions between requests
+to different locations within each processor. The store-atomicity property per
+address can be informally explained as follows: each location behaves like an
+atomic register in which no two writes (stores) to that location can happen
 simultaneously, and any read of that location (load) returns the previously
-written value in that location. Most memory consistency models require that
-the memory subsystem obeys the store-atomicity property, which is a global
-property for the whole multi-processor system. Unlike the store-atomicity property,
-the reordering restrictions dictated by the memory consistency model pertain to
-reordering requests within the same processor, but across multiple addresses. For
-instance, it dictates whether a store request to an address $x$ can be issued
-to the memory sub-system before a load request to a different address $y$,
-which was issued by the same processor earlier (\ie load $y$ happened before store
-$x$ is program order, but the order of issue to memory sub-system is reversed).
+written value in that location. Most memory consistency models require that the
+memory subsystem obeys the store-atomicity property. The reordering
+restrictions dictated by the memory consistency model pertain to reordering
+requests within the same processor (unlike the store-atomicty property which is
+global across the entire system), but across multiple addresses. For example,
+it dictates whether a store request to an address $x$ can be issued to the
+memory sub-system before a load request to a different address $y$, which was
+generated by the same processor earlier (\ie load $y$ happened before store $x$
+is program order, but the order of issue to memory sub-system is reversed).
 
@@
 
-A request issued by a processor to the memory sub-system will contain the
+Figure \ref{req-resp} shows the data-types corresponding to a request issued
+from the processor to the memory subsystem and a response issued from the
+memory sub-system to the processor. A request contains the
 address corresponding to the request, and a type indicating whether the request
-is a load or a store. In case of a store request, it will normally contain the
+is a load or a store. In case of a store request, a real system will contain the
 data to be stored. We augment the request message with two special fields: a
@@
 
-A response issued by the memory sub-system to the processor will contain a
-label to identify the request corresponding to the response and the processor
-to which the response is sent to. In the case of a load request, the memory
-sub-system returns \stl,  the label of the last store to that address. In
-case there has been no store to that address, a special value called Initial is
-returned. The memory sub-system also returns a natural number, \mytime
-which represents the position at which this request has been handled globally
-among all the requests issued from all the processors. Both the store label (or
-Initial) and the time are needed just for the definition of the store-atomicity
-property -- a real system will simply return the data corresponding to the load
-request's address.
+A response issued contains a label to identify the request corresponding to the
+response and the processor to which the response is sent to. In the case of a
+load request, the memory sub-system returns \stl, the label of the last store
+request to that address. In case there has been no store to that address, a
+special value called Initial is returned. The memory sub-system also returns a
+natural number, \mytime which represents the position at which this request has
+been handled globally among all the requests issued from all the processors.
+Both the store label (or Initial) and the time are needed just for the
+definition of the store-atomicity property -- a real system will simply return
+the data corresponding to the load request's address.
 
@@
 \caption{Data types for request and response}
-\mylabel{req-resp}
+\label{req-resp}
 \end{figure}
 
-The uniqueness of request labels (globally) and request indices (per processor) can be defined as follows:
+The \Request set in the following contains only those requests processed by the
+memory subsystem and the \Response set contains only the responses generated by
+the memory subsystem.
+
+The uniqueness of request labels (globally), which is needed for defining store
+atomicity, can be defined as follows:
 \begin{defn}
@@
 \end{defn}
+
+Requests from the same processor to the memory subsystem are sent in the order
+of their \myindex field. Thus, is request $r$ is sent to a processor $c$, then
+the next request that can be sent to the processor obeys the following
+property.
 \begin{defn}
 \begin{multline*}
-\textit{uniqIndices} := \forall q_1, q_2\in\Request,
-     q_1.\proc = q_2.\proc @-> q_1.\myindex = q_2.\myindex \\ @-> q_1 = q_2
+\textit{nextReq} r c := 
 \end{multline*}
@@
 = q_1.\mylabel @-> r_2.\mylabel = q_2.\mylabel @-> \\ q_1.\proc = q_2.\proc @->
-q_1.\loc = q_2.\loc @-> \\ q_1.\myindex < q_2.\myindex @-> r_1.\mytime > r_2.\mytime
-  @-> \False
+q_1.\loc = q_2.\loc @-> \\ q_1.\myindex < q_2.\myindex @-> r_1.\mytime \le r_2.\mytime
 \end{multline*}
@@
 \begin{multline*}
-\textit{initialValue}:= r.\stl = \Initial \wedge \\
+\textit{initialValue}:= \forall (r \in \Request) (q \in \Response), r.\mylabel = q.\mylabel @->
+q.\desc = \Ld @-> \\(r.\stl = \Initial \wedge
 \forall (r' \in \Response) (q' \in \Request),
 r'.\mylabel = q'.\mylabel @-> \\ 0 \le r'.\mytime < r.\mytime @->
-\neg (q.\loc = q'.\loc \wedge q'.\desc = \St)
+\neg (q.\loc = q'.\loc \wedge q'.\desc = \St))
 \end{multline*}
@@
 \begin{multline*}
-\textit{lastEarlierValue} :=
-\exists m, r.\stl = \Store m \wedge \\
+\textit{lastEarlierValue} := \forall (r \in \Request) (q \in \Response), r.\mylabel = q.\mylabel @-> \\
+q.\desc = \Ld @-> 
+(\exists m, r.\stl = \Store m \wedge \\
 \exists (r_m \in \Response) (q_m \in \Request), 
@@
 r'.\mylabel = q'.\mylabel @-> \\ r_m.\mytime < r'.\mytime < r.\mytime ->
-\neg (q.\loc = q'.\loc \wedge q'.\desc = \St)
+\neg (q.\loc = q'.\loc \wedge q'.\desc = \St))
 \end{multline*}

### 2014-01-27 paper/Background.tex
paper

@@
 In order to obtain the execution semantics of a multi-threaded program on a
-multi-processor system, one has to understand the memory consistency model of the
-underlying system. This is necessary because, as opposed to a direct access to
-a single-ported memory, the processors a) have store buffers which reorders the
-sequence of memory requests from the processor, for faster access (say by bypassing
-a store-value to a subsequent load request), and b) communicate with the memory
-through several levels of data caches for faster access. The
-memory consistency model describes the semantic behavior of such requests and
-the system ensures that the cache hierarchy and the store
-buffers do not violate the specified semantics.
+multi-processor system, one has to understand the memory consistency model of
+the underlying system. This is necessary because, as opposed to a direct access
+to a single-ported memory, the processors a) have store buffers which reorders
+the sequence of memory requests from the processor, for faster access (say by
+bypassing a store-value to a subsequent load request), and b) communicate with
+the memory through several levels of data caches for faster access. The memory
+consistency model describes the semantic behavior of such requests and the
+system ensures that the cache hierarchy and the store buffers do not violate
+the specified semantics.
 
@@
 
-Figure \ref{req-resp} shows the data-types corresponding to a request issued
-from the processor to the memory subsystem and a response issued from the
-memory sub-system to the processor. A request contains the
-address corresponding to the request, and a type indicating whether the request
-is a load or a store. In case of a store request, a real system will contain the
-data to be stored. We augment the request message with two special fields: a
-\mylabel which uniquely identifies each request, and an \myindex which
-is a natural number denoting the position of a request amongst those emanating
-from a particular processor. We omit the data field because it can be subsumed
-by the unique label (the data and the unique label together forms another
-unique label).
+The requests issued by the processors to the memory subsystem can be modeled
+using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
+processor $p$, \ie $\reqFn: (\textit{Processor} * \mathbb{N}) @-> \Request$.
+The processor name and the position of the request within the processor
+uniquely identifies the request.  Figure \ref{req-resp} shows the data-type
+corresponding to a request issued from the processor to the memory subsystem.
+A request contains the address (\addrQ) corresponding to the request, and a
+type indicating whether the request is a load or a store (\type). In case of a
+store request, it also contains data (\dataQ) supplied by the processor for
+that address.
 
-A response issued contains a label to identify the request corresponding to the
-response and the processor to which the response is sent to. In the case of a
-load request, the memory sub-system returns \stl, the label of the last store
-request to that address. In case there has been no store to that address, a
-special value called Initial is returned. The memory sub-system also returns a
-natural number, \mytime which represents the position at which this request has
-been handled globally among all the requests issued from all the processors.
-Both the store label (or Initial) and the time are needed just for the
-definition of the store-atomicity property -- a real system will simply return
-the data corresponding to the load request's address.
+Figure \ref{req-resp} also shows the data-type corresponding to a response
+issued from the memory subsystem to the processor.  A response contains a label
+(\labelR) to identify the request corresponding to the response. The label
+contains a processor name and a position, which uniquely identifies a request.
+In the case of a load request, the memory sub-system returns \dataR, the data
+returned by the memory subsystem for that address. The memory sub-system also
+returns a natural number, \timeR, which is used solely in the definition of
+store atomicity, and is not returned in a real system. \timeR can be informally
+thought to represent the position at which this request has been handled
+globally among all the requests issued from all the processors.
 
@@
 \hline
-&\proc:& Processor issuing the request\\
-&\mylabel:& A unique label to identify the request\\
-&\loc:& Address associated with the request\\
+&\addrQ:& Address associated with the request\\
 &\desc:& Request type (\Ld{} or \St)\\
-&\myindex:& A unique natural number indicating the local order of requests\\
+&\dataQ:& Data associated with a \St{} request\\
 \hline
@@
 \hline
-&\proc:& Processor which receives the response\\
-&\mylabel:& Label identifying the request corresponding to this response\\
-&\stl:& The label of the last store to that address (or the special value,
-       Initial) in case of a load response\\
-&\mytime:& A natural number indicating the global order of (processing the) requests\\
+&\labelR:& Label identifying the request corresponding to this response\\
+&\dataR:& Data returned by the memory subsystem for a \Ld{} request\\
+&\timeR:& Natural number representing global order of (processing) requests\\
 \hline
@@
 
-The \Request set in the following contains only those requests processed by the
-memory subsystem and the \Response set contains only the responses generated by
-the memory subsystem.
-
-The uniqueness of request labels (globally), which is needed for defining store
-atomicity, can be defined as follows:
-\begin{defn}
-\begin{multline*}
-\textit{uniqReqLabels} := \forall q_1, q_2\in\Request, q_1.\mylabel = q_2.\mylabel @-> q_1 = q_2\\
-\end{multline*}
-\end{defn}
-
-Requests from the same processor to the memory subsystem are sent in the order
-of their \myindex field. Thus, is request $r$ is sent to a processor $c$, then
-the next request that can be sent to the processor obeys the following
-property.
-\begin{defn}
-\begin{multline*}
-\textit{nextReq} r c := 
-\end{multline*}
-\end{defn}
+In the following, a \Response set only contains those responses sent by the
+memory subsystem.
 
 The responses should obey the following sanity conditions:
-\begin{defn} (Each response corresponds to a request.)
-\begin{equation*}
-\textit{respHasReq} := \forall r\in\Response, \exists q\in\Request, r.\mylabel = q.\mylabel
-\end{equation*}
-\end{defn}
 
-\begin{defn} (A response comes back to the same processor that sent the corresponding request.)
+\begin{defn} (The labels in the responses are unique. Since \dataR{} field is
+irrelevant for store requests, so uniqueness of responses does not depend on
+\dataR{} field for stores.)
 \begin{multline*}
-\textit{respProc} := \forall (r\in\Response) (q\in\Request), r.\mylabel = q.\mylabel \\ @-> r.\proc = q.\proc
-\end{multline*}
-\newline
-\end{defn}
-
-\begin{defn} (The labels in the responses are unique. Since \stl{} field is irrelecant for
-store requests, they are checked only for load requests.)
-\begin{multline*}
-\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response), r_1.\mylabel = r_2.\mylabel @-> \\
-\hspace{1cm} r_1.\proc = r_2.\proc \wedge r_1.\mytime = r_2.\mytime \wedge \hfill \\
-\forall (q \in \Response), q.\mylabel = r_1.\mylabel @->
-q.\desc = \Ld @-> r_1.\stl = r_2.\stl
+\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),\\
+r_1.\labelR = r_2.\labelR @-> r_1.\timeR = r_2.\timeR \wedge\\
+(\reqFn(r_1.\labelR) = \Ld @-> r_1.\dataR = r_2.\dataR)
 \end{multline*}
@@
 \textit{uniqStRespTimes} := 
-\forall (r_1, r_2 \in Response), r_1.\mytime = r_2.\mytime @-> \\
-\forall (q_1, q_2 \in Request), r_1.\mylabel = q_1.\mylabel @->
-r_2.\mylabel = q_2.\mylabel -> \\
-q_1.\loc = q_2.\loc @-> q_1.\desc = \St @-> \\ r_1.\mylabel = r_2.\mylabel
+\forall (r_1, r_2 \in \Response),
+r_1.\timeR = r_2.\timeR @-> \\
+(\reqFn (r_1.\labelR)).\addrQ = (\reqFn (r_2.\labelR)).\addrQ @->\\
+(\reqFn (r_1.\labelR)).\desc = \St @->
+r_1.\labelR = r_2.\labelR
 \end{multline*}
@@
 \textit{localOrdering} :=
-\forall (r_1, r_2 \in \Response) (q_1, q_2 \in \Request), \\ r_1.\mylabel
-= q_1.\mylabel @-> r_2.\mylabel = q_2.\mylabel @-> \\ q_1.\proc = q_2.\proc @->
-q_1.\loc = q_2.\loc @-> \\ q_1.\myindex < q_2.\myindex @-> r_1.\mytime \le r_2.\mytime
+\forall (r_1, r_2 \in \Response), (r_1.\labelR).\fst = (r_2.\labelR).\fst @->\\
+(r_1.\labelR).\snd < (r2.\labelR).\snd @-> r_1.\timeR \le r_2.\timeR
 \end{multline*}
@@
 This brings us to the main condition for store atomicity which states that the
-store label returned in a load response corresponds to the \emph{last earlier}
-store request that has been processed in the system for that address (lastEarlierValue). The
-\emph{last} and \emph{earlier} are defined according to the field \mytime{}
-returned in the responses. In case no store has happened earlier for a
-particular address when a load is processed, then the load response returns a
-special value \Initial, which is used to initialize all the addresses (initialValue).
-
+value returned for a load response corresponds to the \emph{last earlier} store
+request that has been processed in the system for that address
+(\textit{lastEarlierValue}). The \emph{last} and \emph{earlier} are defined
+according to the field \timeR{} returned in the responses. In case no store has
+happened earlier for a particular address when a load is processed, then the
+load response returns the initial value for the address (\textit{initialValue}).
 \begin{defn}
 \begin{multline*}
-\textit{initialValue}:= \forall (r \in \Request) (q \in \Response), r.\mylabel = q.\mylabel @->
-q.\desc = \Ld @-> \\(r.\stl = \Initial \wedge
-\forall (r' \in \Response) (q' \in \Request),
-r'.\mylabel = q'.\mylabel @-> \\ 0 \le r'.\mytime < r.\mytime @->
-\neg (q.\loc = q'.\loc \wedge q'.\desc = \St))
+\textit{initialValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin
+q.\desc = \Ld @-> \\ r.\dataR = \initData (r.\addrQ) \wedge \\
+\forall (r' \in \Response), \mylet q' := \reqFn (r'.\labelR) \myin 0 \le r'.\timeR < r.\timeR @->\\
+\neg (q'.\addrQ = q.\addrQ \wedge q'.\desc = \St)
 \end{multline*}
@@
 \begin{multline*}
-\textit{lastEarlierValue} := \forall (r \in \Request) (q \in \Response), r.\mylabel = q.\mylabel @-> \\
-q.\desc = \Ld @-> 
-(\exists m, r.\stl = \Store m \wedge \\
-\exists (r_m \in \Response) (q_m \in \Request), 
-r_m.\mylabel = m \wedge q_m.\mylabel = m \wedge \\
-r_m.\mytime < r.\mytime \wedge q_m.\loc = q.\loc \wedge q_m.\desc = \St \wedge  \\
-\forall (r' \in \Response) (q' \in \Request),
-r'.\mylabel = q'.\mylabel @-> \\ r_m.\mytime < r'.\mytime < r.\mytime ->
-\neg (q.\loc = q'.\loc \wedge q'.\desc = \St))
+\textit{lastEarlierValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin
+q.\desc = \Ld @-> \\ \exists r_m, \mylet q_m := \reqFn (r_m.\labelR) \myin
+r.\dataR = q_m.\dataQ \wedge \\
+\forall (r' \in \Response), \mylet q' := \reqFn (r'.\labelR) \myin \\ r_m.\timeR \le r'.\timeR < r.\timeR @->
+\neg (q'.\addrQ = q.\addrQ \wedge q'.\desc = \St)
 \end{multline*}
@@
 \begin{multline*}
-storeAtomicity := \textit{uniqReqLabels} @-> \textit{uniqIndices} @-> \\
-\textit{respHasReq} \wedge \textit{respProc} \wedge \textit{uniqRespLabels} \wedge
+storeAtomicity := 
+\textit{uniqRespLabels} \wedge
 \textit{uniqStRespTimes} \wedge \textit{localOrdering}\\ \wedge 
-(\textit{initialValue} \vee \textit{lastEarlierValue})
+(\forall (r \in \Response), \textit{initialValue}(r) \vee \textit{lastEarlierValue}(r))
 \end{multline*}

### 2014-01-27 paper/Background.tex
stuff

@@
 A request contains the address (\addrQ) corresponding to the request, and a
-type indicating whether the request is a load or a store (\type). In case of a
+type indicating whether the request is a load or a store (\desc). In case of a
 store request, it also contains data (\dataQ) supplied by the processor for
@@
 returns a natural number, \timeR, which is used solely in the definition of
-store atomicity, and is not returned in a real system. \timeR can be informally
+store atomicity, and is not returned in a real system. \timeR{} can be informally
 thought to represent the position at which this request has been handled
@@
 \centering
-\begin{tabular}{|p{.02\textwidth}lp{.9\textwidth}|}
+\begin{subfigure}{5.5cm}
+\begin{tabular}{|lp{4.5cm}|}
 \hline
-\multicolumn{3}{|c|}{\Request}\\
+\multicolumn{2}{|c|}{\Request}\\
 \hline
-&\addrQ:& Address associated with the request\\
-&\desc:& Request type (\Ld{} or \St)\\
-&\dataQ:& Data associated with a \St{} request\\
-\hline
-\hline
-\multicolumn{3}{|c|}{\Response}\\
-\hline
-&\labelR:& Label identifying the request corresponding to this response\\
-&\dataR:& Data returned by the memory subsystem for a \Ld{} request\\
-&\timeR:& Natural number representing global order of (processing) requests\\
+\addrQ:& Address associated with the request\\
+\desc:& Request type (\Ld{} or \St)\\
+\dataQ:& Data associated with a \St{} request\\
 \hline
 \end{tabular}
+\end{subfigure}~~~~
+\begin{subfigure}{5.5cm}
+\begin{tabular}{|lp{4.5cm}|}
+\hline
+\multicolumn{2}{|c|}{\Response}\\
+\hline
+\labelR:& Label identifying the request corresponding to this response\\
+\dataR:& Data returned by the memory subsystem for a \Ld{} request\\
+\timeR:& Natural number representing global order of (processing) requests\\
+\hline
+\end{tabular}
+\end{subfigure}
 \caption{Data types for request and response}
@@
 
-In the following, a \Response set only contains those responses sent by the
+In the following, a \Response{} set only contains those responses sent by the
 memory subsystem.
@@
 \begin{multline*}
-\textit{initialValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin
-q.\desc = \Ld @-> \\ r.\dataR = \initData (r.\addrQ) \wedge \\
+\textit{initialValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin\\
+q.\desc = \Ld @-> r.\dataR = \initData (r.\addrQ) \wedge \\
 \forall (r' \in \Response), \mylet q' := \reqFn (r'.\labelR) \myin 0 \le r'.\timeR < r.\timeR @->\\
@@
 \begin{multline*}
-\textit{lastEarlierValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin
-q.\desc = \Ld @-> \\ \exists r_m, \mylet q_m := \reqFn (r_m.\labelR) \myin
-r.\dataR = q_m.\dataQ \wedge \\
+\textit{lastEarlierValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin\\
+q.\desc = \Ld @-> \exists r_m, \mylet q_m := \reqFn (r_m.\labelR) \myin\\
+r.\dataR = q_m.\dataQ \wedge
 \forall (r' \in \Response), \mylet q' := \reqFn (r'.\labelR) \myin \\ r_m.\timeR \le r'.\timeR < r.\timeR @->

### 2014-01-27 paper/System.tex
stuff

@@
 
-The system we are considering consists of a tree hierarchy of caches. These
+For proving store-atomicity of the memory subsystem, it is sufficient to model
+just the memory subsystem, instead of modeling the processors.
+The memory subsystem consists of a tree hierarchy of caches. These
 caches are inclusive, \ie a cache at a higher level (near the root) contains
@@
 In addition, each non-leaf cache maintains a \emph{directory}, which contains a
-local version of the state of each of its children for a particular address.
+local version of the state of each of its children caches for a particular address.
 This local version may be stale, but it should not be lower than the real state
@@
 In order for a cache to change to a higher state for a given address, \ie
-\emph{upgrade}, it has to \emph{request} its parents for increasing the state.
-For a cache to change to a lower state, \ie \emph{downgrade}, it has to notify
-its parents indicating the downgrade -- it uses \emph{response} messages for
-this purpose. On the other end, for a parent to force a child to downgrade the
-child's state for an address, or equivalently, to downgrade the parent's
-directory state for the child, it has to \emph{request} that child to
-downgrade, and a parent can notify a child about upgrading the child's state,
-or equivalently, about upgrading the parent's directory state for the child, by
-sending a \emph{response} message.
+\emph{upgrade}, it has to \emph{request} its parent cache for increasing the
+state.  For a cache to change to a lower state, \ie \emph{downgrade}, it has to
+notify its parent cache indicating the downgrade -- it uses \emph{response}
+messages for this purpose. On the other hand, for a parent cache to force a
+child cache to downgrade the child's state for an address (in order to give
+permissions to a different child cache), or equivalently, to downgrade the
+parent's directory state for the child, it has to \emph{request} that child
+cache to downgrade, and a parent can notify a child about upgrading the child's
+state, or equivalently, about upgrading the parent's directory state for the
+child, by sending a \emph{response} message.
 
-Each cache is connected to its parent cache using 3 logical connections: a)
+Each cache is connected to its parent cache using 3 logical FIFO connections: a)
 \cpReq{} channel to transmit requests from the child to the parent, b) \cpResp{}
 channel to transmit responses from the child to the parent, and c) \pc{}
-\textbf{FIFO} channel to transmit both request and response messages from
-parent to child. In addition, the leaf caches (called L1 caches usually)
-communicate with the processor through two channels: a) \cproc{} \textbf{FIFO}
-channel to transmit requests from the processor to L1 cache, and b) \procc{}
-\textbf{FIFO} channel to transmit responses from the L1 cache to the processor.
-There is exactly one of \cproc{} and \procc{} channels between the L1 and the processor.
+channel to transmit both request and response messages from
+parent to child.
+
+%In addition, the leaf caches (called L1 caches usually)
+%communicate with the processor through two channels: a) \cproc{} \textbf{FIFO}
+%channel to transmit requests from the processor to L1 cache, and b) \procc{}
+%\textbf{FIFO} channel to transmit responses from the L1 cache to the processor.
+%There is exactly one of \cproc{} and \procc{} channels between the L1 and the processor.
 
@@
 namely that of deadlock and starvation-freedom. Though we have proved these
-results requiring the network to obey very relaxed and practically feasible
-conditions, we completely omit the formal discussion of deadlock and starvation
-freedom in this paper because of lack of space. We \xxx{might later} informally
-discuss how deadlock and starvation freedom properties are proved for our
-system.
+results for a general interconnect network as opposed to a point-to-point
+connection between the communicating caches, we completely omit the formal
+discussion of deadlock and starvation freedom in this paper because of lack of
+space. We \xxx{might later} informally discuss how deadlock and starvation
+freedom properties are proved for our system.
 
@@
 index the three channels (\cpReq{}, \cpResp{} and \pc{}) by the name of the
-child that uses that channel. We will also index the two channels (\cproc{} and
-\procc{}) between the processor and L1 cache by the name of the L1 cache.
-
+child that uses that channel.
+%We will also index the two channels (\cproc{} and
+%\procc{}) between the processor and L1 cache by the name of the L1 cache.
 Figure \ref{format} shows the message format transmitted in the 3 channels
-between caches. The messages transmitted between the processor and the L1 cache
-is the same as in Figure \ref{req-resp} (except for the omission of the \proc{}
-field, which is indicated by the channel's index).
+between caches.
+%The messages transmitted between the processor and the L1 cache
+%is the same as in Figure \ref{req-resp} (except for the omission of the \proc{}
+%field, which is indicated by the channel's index).
 \begin{figure}
-\centering
-\begin{tabular}{|l|lp{.9\textwidth}|}
+\begin{subfigure}{6.8cm}
+\begin{tabular}{|lp{5.8cm}|}
 \hline
-\multirow{3}{*}{\cpReq} & \from: & Current state of child\\
-& \myto: & Desired (higher) state of child\\
-& \addr: & Address associated with the request\\
+\multicolumn{2}{|c|}{\Reqcp}\\
 \hline
-\multirow{4}{*}{\cpResp} & \from: & Current state of child\\
-& \myto: & Downgraded state of child\\
-& \addr: & Address associated with the request\\
-& \data: & Data associated with the address, if necessary\\
-\multirow{4}{*}{\pc} & \type: & \Req{} (for requests) and \Resp{} (for responses)\\
+\from: & Current state of child\\
+\myto: & Desired (higher) state of child\\
+\addr: & Address associated with the request\\
 \hline
-%& \from: & Current state of parent's directory for the child\\
-& \myto: & Desired (lower) directory state of a child for \Req{} and upgraded
-directory state of a child for \Resp{}\\
-& \addr: & Address associated with the request\\
-& \data: & Data associated with the address if necessary (for \Resp{} only)\\
+\hline
+\multicolumn{2}{|c|}{\Respcp}\\
+\hline
+\from: & Current state of child\\
+\myto: & Downgraded state of child\\
+\addr: & Address associated with the request\\
+\data: & Data associated with the address, if necessary\\
 \hline
 \end{tabular}
-\caption{Data types messages between caches}
+\end{subfigure}
+\begin{subfigure}{5.4cm}
+\begin{tabular}{|lp{4.4cm}|}
+\hline
+\multicolumn{2}{|c|}{\Mpc}\\
+\hline
+\typ: & \Req{} (for requests) and \Resp{} (for responses)\\
+\myto: & Desired (lower) directory state of a child for \Req{} and upgraded
+directory state of a child for \Resp{}\\
+\addr: & Address associated with the request\\
+\data: & Data associated with the address if necessary (for \Resp{} only)\\
+\hline
+\end{tabular}
+\end{subfigure}
+\caption{Data types for messages between caches}
 \label{format}
@@
 
-The description of the behavior of each of the caches can be given in terms of
-``rules'' for state transitions. Each rule specifies a series of actions that
-are performed atomically, and during the execution of a rule, no other
-(conflicting) rule interferes with it. Thus, the behavior of any rule-based
-system can be understood as if the rules executed one at a time.
+The behavior of the system can be given in terms of atomic state transitions.
+These transitions have two parts: a) the guard, which dictates when the
+transition can fire, and b) the action, which dictates how the state of the
+system changes when the transition happens. These transitions or \emph{guarded
+atomic actions} logically happen one by one, making the overall behavior of the
+system well defined.
 
-Such a rule-based description can be automatically converted into synchronous
-hardware, for instance using Bluespec \cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
-Bluespec executes each atomic rule into a single hardware cycle, which
-simultaneously executing non-conflicting rules \ie{} rules that do not access the
-same state (the exact conditions for simultaneous scheduling are specified in
-the citations). Though the rules are logically executed atomically, their execution
-can be spread across multiple cycles (as long as other conflicting rules are
-not executed during the whole period of current rule's execution). Synchronous
-hardware generation for multi-cycle atomic rules has also been studied
-\cite{Karczmarek}.
+The atomic transitions access local states, \ie a transition can read or write
+states only corresponding to a single cache and/or the channel the cache is
+connected to. This restriction allows such a system to be implemented directly
+into hardware. Bluespec System Verilog (BSV), for instance, directly converts
+these transitions into efficient synchronous hardware. Though these transitions
+logically happen one by one, BSV executes several transitions simultaneously,
+as long as they do not access the same state, \ie do not \emph{conflict}
+\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}. Sometimes, these transitions (both the
+guards and the actions) may not be amenable to single cycle implementations in
+hardware. For example, writing a register array (like a cache) can actually
+take several hardware clock cycles, though logically it is a single transition.
+Karczmarek \etal \cite{Karczmarek} has provided a scheme to convert atomic
+transitions into synchronous hardware in which each transition can span several
+clock cycles.

### 2014-01-27 paper/Background.tex
stuff

@@
 \dataR{} field for stores.)
+\small
 \begin{multline*}
@@
 \begin{defn} (No response can share the same time as another store response.)
+\small
 \begin{multline*}
-\textit{uniqStRespTimes} := 
+\textit{uniqStTimes} := 
 \forall (r_1, r_2 \in \Response),
@@
 \begin{defn}
+\small
 \begin{multline*}
@@
 value returned for a load response corresponds to the \emph{last earlier} store
-request that has been processed in the system for that address
-(\textit{lastEarlierValue}). The \emph{last} and \emph{earlier} are defined
-according to the field \timeR{} returned in the responses. In case no store has
-happened earlier for a particular address when a load is processed, then the
-load response returns the initial value for the address (\textit{initialValue}).
+request that has been processed in the system for that address This is
+indicated by the proposition \textit{lastEarlierValue}$(r)$ for response $r$.
+The \emph{last} and \emph{earlier} are defined according to the field \timeR{}
+returned in the responses. In case no store has happened earlier for a
+particular address when a load is processed, then the load response returns the
+initial value for the address. This is indicated by the proposition
+\textit{initialValue}$(r)$ for response $r$.
 \begin{defn}
+\small
 \begin{multline*}
-\textit{initialValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin\\
-q.\desc = \Ld @-> r.\dataR = \initData (r.\addrQ) \wedge \\
-\forall (r' \in \Response), \mylet q' := \reqFn (r'.\labelR) \myin 0 \le r'.\timeR < r.\timeR @->\\
-\neg (q'.\addrQ = q.\addrQ \wedge q'.\desc = \St)
+\textit{initialValue} := \lambda (r \in \Response) \Rightarrow
+(\reqFn (r.\labelR)).\desc = \Ld @->\\ r.\dataR = \initData (r.\addrQ) \wedge 
+\forall (r' \in \Response), 0 \le r'.\timeR < r.\timeR @->\\
+\neg ((\reqFn (r'.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ \wedge (\reqFn (r'.\labelR)).\desc = \St)
 \end{multline*}
@@
 \begin{defn}
+\small
 \begin{multline*}
-\textit{lastEarlierValue} (r: \Response) := \mylet q := \reqFn (r.\labelR) \myin\\
-q.\desc = \Ld @-> \exists r_m, \mylet q_m := \reqFn (r_m.\labelR) \myin\\
-r.\dataR = q_m.\dataQ \wedge
-\forall (r' \in \Response), \mylet q' := \reqFn (r'.\labelR) \myin \\ r_m.\timeR \le r'.\timeR < r.\timeR @->
-\neg (q'.\addrQ = q.\addrQ \wedge q'.\desc = \St)
+\textit{initialValue} := \lambda (r \in \Response) \Rightarrow
+(\reqFn (r.\labelR)).\desc = \Ld @->
+\exists (r_m \in \Response),\\ r.\dataR = (\reqFn (r_m.\labelR)).\dataQ \wedge 
+(\reqFn (r_m.\labelR)).\desc = \St \\ \wedge (\reqFn (r_m.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ\\ \wedge
+\forall (r' \in \Response), r_m.\timeR \le r'.\timeR < r.\timeR @->\\
+\neg ((\reqFn (r'.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ \wedge (\reqFn (r'.\labelR)).\desc = \St)
 \end{multline*}
@@
 \begin{thm}
+\small
 \begin{multline*}
@@
 \textit{uniqRespLabels} \wedge
-\textit{uniqStRespTimes} \wedge \textit{localOrdering}\\ \wedge 
+\textit{uniqStTimes} \wedge \textit{localOrdering}\\ \wedge 
 (\forall (r \in \Response), \textit{initialValue}(r) \vee \textit{lastEarlierValue}(r))

### 2014-01-27 paper/System.tex
stuff

@@
 root of the communicates with the main memory, and the leaf caches communicate
-with the processor (Figure \ref{hier}).
+with the processor (Figure \ref{hier}). Whether a cache is a leaf or not is
+denoted by the relation \leaf$(c)$.
 
-Each cache has a \emph{coherence state} associated with each address. The
-coherence state can be one of the three values: \Mo{} (Modified), \Sh{} (Shared) or
-\In{} (Invalid). \Mo{} state denotes that the subtree rooted at the cache in \Mo{}
-state for that address can modify the data, \ie{} both read and write the data
-corresponding to the address. \Sh{} state denotes that the subtree rooted at the
-cache in \Sh{} state for that address can only read the data corresponding to
-that address and \In{} state denotes that the subtree rooted at the cache in \In{}
-state can neither read nor write the data corresponding to the address. The
-meanings of the state is clear, but one must verify the actual cache coherence
-protocol for violations of these invariants, which is what the proof in this
-paper intends to do. There's a logical ordering between the states, in terms of
-the permissions that they have: $\Mo > \Sh > \In$ (the notation should be read
-assuming that the order is transitively closed).
-In addition, each non-leaf cache maintains a \emph{directory}, which contains a
-local version of the state of each of its children caches for a particular address.
-This local version may be stale, but it should not be lower than the real state
-of the child (again this property has to be proven for a cache coherence
-protocol).
+We will restrict the discussion of the formal proofs to the MSI protocol
+\cite{MSI}. Extension of this proof framework to other protocols like MESI and
+MOESI is straightforward and will be discussed informally.
 
-Though this paper restricts the formal discussion to the simple MSI protocol,
-extension of this proof technique as well as major portions of the proof to
-other protocols like MOSI and MOESI is straightforward.
+Each leaf cache $c$ has a \emph{position} denoted by $c.\pos$, which represents
+the position of the next request from the processor corresponding to $c$ that
+it has to process. Since it has to process the first request first, it is set
+to $0$ initially.
+
+Each cache $c$, in addition to the data associated with each address $a$ (denoted
+by $c.\data[a]$), also has a \emph{coherence state} associated with the
+address (denoted by $c.\state[a]$). The coherence state can be
+one of the three values: \Mo{} (Modified), \Sh{} (Shared) or \In{} (Invalid).
+\Mo{} state denotes that the subtree rooted at the cache in \Mo{} state for
+that address can modify the data, \ie{} both read and write the data
+corresponding to the address. \Sh{} state denotes that the subtree rooted at
+the cache in \Sh{} state for that address can only read the data corresponding
+to that address and \In{} state denotes that the subtree rooted at the cache in
+\In{} state can neither read nor write the data corresponding to the address.
+The meanings of the state is clear, but one must verify the actual cache
+coherence protocol for violations of these invariants, which is what the proof
+in this paper intends to do.  There's a logical ordering between the states, in
+terms of the permissions that they have: $\Mo > \Sh > \In$ (the notation should
+be read assuming that the order is transitively closed). The states for each of
+the addresses in each of the caches, other than the root cache, is set to \In,
+while states for each address in the root cache is set to \Mo. This is because
+only the main memory (connected to the root cache) has the initial copy of
+data.  In addition, each non-leaf cache maintains a \emph{directory}, which
+contains a local version of the state of each of its children caches for a
+particular address.  For a non-leaf cache $p$, the state in its directory for
+child cache $c$ is denoted by $p.\dstate[c][a]$, for address $a$.  The relation
+$\parent(c, p)$ denotes that $p$ is the parent of cache $c$.  This local
+version may be stale, but it should not be lower than the real state of the
+child (again this property has to be proven for a cache coherence protocol).
+All the directory states are set to \In initially because initially, all the
+non-root caches have their states for all the addresses set to \In.
 
@@
 In this paper, we will assume that each channel has a dedicated link between
-caches. In a real system, these channels will be logically implemented over an
-interconnect network, which poses additional problems that have to be proven --
-namely that of deadlock and starvation-freedom. Though we have proved these
-results for a general interconnect network as opposed to a point-to-point
-connection between the communicating caches, we completely omit the formal
+the communicating caches. In a real system, these channels will be logically
+implemented over an interconnect network, which poses additional problems that
+have to be proven -- namely that of deadlock and starvation-freedom. Though we
+have proved these results for a general interconnect network as opposed to a
+point-to-point connection between caches, we completely omit the formal
 discussion of deadlock and starvation freedom in this paper because of lack of
-space. We \xxx{might later} informally discuss how deadlock and starvation
-freedom properties are proved for our system.
+space. We will informally discuss how deadlock and starvation freedom
+properties are proved for our system.
 
@@
 index the three channels (\cpReq{}, \cpResp{} and \pc{}) by the name of the
-child that uses that channel.
-%We will also index the two channels (\cproc{} and
-%\procc{}) between the processor and L1 cache by the name of the L1 cache.
-Figure \ref{format} shows the message format transmitted in the 3 channels
-between caches.
+child that uses that channel (using the notation $\cpReq[c]$, \etc for the
+\cpReq{} channel associated with cache $c$).  Figure \ref{format} shows the
+message format transmitted in the 3 channels between caches. In each of the
+channels $ch$, a message $m$ can be enqueued ($\enq(ch, m)$) or dequeued
+($\deq(ch)$), and the channels can be examined for the first element
+($\first(ch)$) or for availability of any message in the channel ($\avail(ch)$).
+Each of these channels start out empty \ie $\forall ch, \neg \avail(ch)$, initially.
 %The messages transmitted between the processor and the L1 cache
@@
 
+In addition to the actual coherence states, each cache also maintains a
+temporary \emph{wait state} for each of its addresses. The wait state contains
+two parts: a) a bool representing whether the cache has sent a request to its
+parent cache earlier and is waiting for a response (denoted by $c.\wait[a]$ for
+address $a$) and b) one of the three values \Mo, \Sh{} or \In representing the
+state upgrade that the cache is waiting for (denoted by $c.\waitS[a]$ for
+address $a$). Similarly, each non-leaf cache has the corresponding wait states
+for its directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting for a
+response from its child $c$ for address $a$, and $p.\dwaitS[c][a]$ denotes the
+state that cache $p$ is waiting for $c$ to downgrade to. Both \wait{} and
+\dwait{} are \False{} initially for each of the caches, and each of the
+addresses, because no cache is waiting for any response from either its child
+or its parent initially.
+
 The behavior of the system can be given in terms of atomic state transitions.
@@
 atomic actions} logically happen one by one, making the overall behavior of the
-system well defined.
+system well defined. Figure \ref{trans} gives the atomic transitions for a
+cache coherent memory subsystem. The actions in the transitions repesent only
+the specific system states that change. We use $x \Leftarrow y$ to indicate
+setting of a specific state $x$ of the overall system to value $y$ because of a
+transition.
+\begin{figure}
+\small
+\centering
+\begin{subfigure}{\textwidth}
+\centering
+\begin{tabular}{|ll|}
+\hline
+\multicolumn{2}{|c|}{ChildSendReq$(c, x, a)$: Child $c$ sending request to upgrade to $x$ for address $a$}\\
+\hline
+Guard: & $c.\state[a] \le x \wedge c.\wait[a] = \False$\\
+\hline
+Action: & $\enq(\cpReq[c], \Req\langle c.\state[a], x, a \rangle)$;\\
+\hline
+\hline
+\multicolumn{2}{|c|}{ParentRecvReq$(p, c)$: Parent $p$ receiving a request from child $c$}\\
+\hline
+Guard: & 
+$\parent(c,p) \wedge \avail(\cpReq[c]) \wedge \mylet q := \first(\cpReq[c]) \myin$\\
+& $\dwait[p][c][q.\addr] = \False \wedge \dstate[p][c][q.\addr] \le q.\from \wedge$\\
+& $\compat(p, c, (q.\myto) \wedge q.\myto \le p.\state[q.\addr]$\\
+\hline
+Action: & $\enq(\pc[c], \langle \Resp, q.\myto, q.\addr, $\\
+& $\;\;\;\myif \dstate[p][c][q.\addr] = \In \mythen \data[p][a] \rangle \myelse \_\rangle)$;\\
+& $\dstate[p][c][q.\addr] \Leftarrow q.\myto$; $\deq(\cpReq[c])$;\\
+\hline
+\hline
+\multicolumn{2}{|c|}{ChildRecvResp$(c)$: Child $c$ receiving a response}\\
+\hline
+Guard: & 
+$\avail(\pc[c]) \wedge \mylet r := \first(\pc[c]) \myin$\\
+\hline
+Action: & $\myif \state[c][q.\addr] = \In \mythen \data[c][a] \Leftarrow q.\data$;\\
+&$\myif \waitS[c][q.\addr] \le q.\myto \mythen c.\wait[a] \Leftarrow \False $;\\
+& $c.\state[q.\addr] \Leftarrow q.\to$; $\deq(\pc[c])$;\\
+\hline
+\end{tabular}
+\caption{Child sending upgrade request and getting back the response}
+\end{subfigure}
+\begin{subfigure}{\textwidth}
+\centering
+\begin{tabular}{|ll|}
+\hline
+\multicolumn{2}{|c|}{LoadReq$(c)$:  Handling a processor's load request}\\
+\hline
+Guard: & $\leaf(c) \wedge \reqFn(c, c.\pos) = \Ld \wedge c.\state[\addrQ(\reqFn(c, c.\pos))] \ge \Sh$\\
+\hline
+Action:& $c.\pos \Leftarrow c.\pos + 1$;\\
+\hline
+\hline
+\multicolumn{2}{|c|}{StoreReq$(c)$:  Handling a processor's store request}\\
+\hline
+Guard: & $\leaf(c) \wedge \reqFn(c, c.\pos) = \St \wedge c.\state[\addrQ(\reqFn(c, c.\pos))] = \Mo$\\
+\hline
+Action:& $c.\pos \Leftarrow c.\pos + 1$;\\
+\hline
+\end{tabular}
+\caption{Handling requests from the processor}
+\end{subfigure}
+\caption{Atomic transitions for a cache coherent memory subsystem}
+\label{trans}
+\end{figure}
 
-The atomic transitions access local states, \ie a transition can read or write
-states only corresponding to a single cache and/or the channel the cache is
-connected to. This restriction allows such a system to be implemented directly
-into hardware. Bluespec System Verilog (BSV), for instance, directly converts
-these transitions into efficient synchronous hardware. Though these transitions
-logically happen one by one, BSV executes several transitions simultaneously,
-as long as they do not access the same state, \ie do not \emph{conflict}
-\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}. Sometimes, these transitions (both the
-guards and the actions) may not be amenable to single cycle implementations in
-hardware. For example, writing a register array (like a cache) can actually
-take several hardware clock cycles, though logically it is a single transition.
-Karczmarek \etal \cite{Karczmarek} has provided a scheme to convert atomic
-transitions into synchronous hardware in which each transition can span several
-clock cycles.
+Note that he atomic transitions in Figure \ref{trans} access only local states,
+\ie a transition can read or write states only corresponding to a single cache
+and/or the channel the cache is connected to. This restriction allows such a
+system to be implemented directly into hardware. Bluespec System Verilog (BSV),
+for instance, directly converts these transitions into efficient synchronous
+hardware. Though these transitions logically happen one by one, BSV executes
+several transitions simultaneously, as long as they do not access the same
+state, \ie do not \emph{conflict} \cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
+Sometimes, these transitions (both the guards and the actions) may not be
+amenable to single cycle implementations in hardware. For example, writing a
+register array (like a cache) can actually take several hardware clock cycles,
+though logically it is a single transition.  Karczmarek \etal \cite{Karczmarek}
+has provided a scheme to convert atomic transitions into synchronous hardware
+in which each transition can span several clock cycles.

### 2014-01-28 paper/System.tex
stuff

@@
 \hline
-\multicolumn{2}{|c|}{ChildSendReq$(c, x, a)$: Child $c$ sending request to upgrade to $x$ for address $a$}\\
+\multicolumn{2}{|l|}{\textbf{ChildSendReq}$(c, x, a)$: Child $c$ sending request to upgrade to $x$ for address $a$}\\
 \hline
-Guard: & $c.\state[a] \le x \wedge c.\wait[a] = \False$\\
+Guard: & $c.\state[a] < x \wedge c.\wait[a] = \False$\\
 \hline
-Action: & $\enq(\cpReq[c], \Req\langle c.\state[a], x, a \rangle)$;\\
+Action: & $\enq(\cpReq[c], \langle c.\state[a], x, a \rangle)$;\\
 \hline
 \hline
-\multicolumn{2}{|c|}{ParentRecvReq$(p, c)$: Parent $p$ receiving a request from child $c$}\\
+\multicolumn{2}{|l|}{\textbf{ParentRecvReq}$(p, c)$: Parent $p$ receiving a request from child $c$}\\
 \hline
@@
 $\parent(c,p) \wedge \avail(\cpReq[c]) \wedge \mylet q := \first(\cpReq[c]) \myin$\\
-& $\dwait[p][c][q.\addr] = \False \wedge \dstate[p][c][q.\addr] \le q.\from \wedge$\\
-& $\compat(p, c, (q.\myto) \wedge q.\myto \le p.\state[q.\addr]$\\
+& $p.\dwait[c][q.\addr] = \False \wedge p.\dstate[c][q.\addr] \le q.\from \wedge$\\
+& $\compat(p, c, q.\myto) \wedge q.\myto \le p.\state[q.\addr]$\\
 \hline
-Action: & $\enq(\pc[c], \langle \Resp, q.\myto, q.\addr, $\\
-& $\;\;\;\myif \dstate[p][c][q.\addr] = \In \mythen \data[p][a] \rangle \myelse \_\rangle)$;\\
-& $\dstate[p][c][q.\addr] \Leftarrow q.\myto$; $\deq(\cpReq[c])$;\\
+Action: & $\enq(\pc[c], \langle \Resp, p.\state[c][q.\addr], q.\myto, q.\addr, $\\
+& $\;\;\;%
+\myif p.\dstate[c][q.\addr] = \In \mythen p.\data[q.\addr] \myelse \_\rangle)$;\\
+& $p.\dstate[c][q.\addr] \Leftarrow q.\myto$; $\deq(\cpReq[c])$;\\
 \hline
 \hline
-\multicolumn{2}{|c|}{ChildRecvResp$(c)$: Child $c$ receiving a response}\\
+\multicolumn{2}{|l|}{\textbf{ChildRecvResp}$(c)$: Child $c$ receiving a response}\\
 \hline
 Guard: & 
-$\avail(\pc[c]) \wedge \mylet r := \first(\pc[c]) \myin$\\
+$\avail(\pc[c]) \wedge \mylet r := \first(\pc[c]) \myin r.\typ = \Resp$\\
 \hline
-Action: & $\myif \state[c][q.\addr] = \In \mythen \data[c][a] \Leftarrow q.\data$;\\
-&$\myif \waitS[c][q.\addr] \le q.\myto \mythen c.\wait[a] \Leftarrow \False $;\\
-& $c.\state[q.\addr] \Leftarrow q.\to$; $\deq(\pc[c])$;\\
+Action: & $\myif c.\state[r.\addr] = \In \mythen c.\data[a] \Leftarrow r.\data$;\\
+&$\myif c.\waitS[r.\addr] \le r.\myto \mythen c.\wait[a] \Leftarrow \False $;\\
+& $c.\state[r.\addr] \Leftarrow r.\myto$; $\deq(\pc[c])$;\\
 \hline
@@
 \end{subfigure}
+
 \begin{subfigure}{\textwidth}
@@
 \hline
-\multicolumn{2}{|c|}{LoadReq$(c)$:  Handling a processor's load request}\\
+\multicolumn{2}{|l|}{\textbf{ParentSendReq}$(p, c, x, a)$: Parent $p$ sending request to child $c$ to downgrade to $x$ for address $a$}\\
+\hline
+Guard: & $\parent(c,p) \wedge p.\dstate[c][a] > x \wedge p.\dwait[c][a] = \False$\\
+\hline
+Action: & $\enq(\pc[c], \langle \Req, p.\dstate[c][a], x, a, \_ \rangle)$;\\
+\hline
+\hline
+\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildRecvReq}$(c)$: Child $c$ receiving a request from its parent and has to downgrade}\\
+\hline
+Guard: & 
+$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge$ \\
+& $(\forall i, \parent(i, c) \rightarrow c.\dstate[i][q.\addr] \le q.\myto) \wedge q.\myto < c.\state[q.\addr]$\\
+\hline
+Action: & $\enq(\cpResp[c], \langle c.\state[q.\addr], q.\myto, q.\addr,$\\
+& $\;\;\;%
+\myif c.\state[q.\addr] = \Mo \mythen c.\data[q.\addr] \myelse \_\rangle)$;\\
+& $c.\state[q.\addr] \Leftarrow q.\myto$; $\deq(\pc[c])$;\\
+\hline
+\hline
+\multicolumn{2}{|l|}{\textbf{ParentRecvResp}$(p, c)$: Parent $p$ receiving a response from child $c$}\\
+\hline
+Guard: & 
+$\parent(c,p) \wedge \avail(\cpResp[c]) \wedge \mylet r := \first(\cpResp[c]) \myin$\\
+& $r.\from = c.\state[r.\addr]$\\
+\hline
+Action: & $\myif p.\dstate[c][r.\addr] = \Mo \mythen p.\data[r.\addr] \Leftarrow r.\data$;\\
+&$\myif p.\dwaitS[c][r.\addr] \ge r.\myto \mythen p.\dwait[c][r.\addr] \Leftarrow \False $;\\
+& $p.\dstate[c][r.\addr] \Leftarrow r.\myto$; $\deq(\cpResp[c])$;\\
+\hline
+\end{tabular}
+\caption{Parent sending downgrade request and getting back the response}
+\end{subfigure}
+
+\begin{subfigure}{\textwidth}
+\centering
+\begin{tabular}{|ll|}
+\hline
+\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildVolResp}$(c, x, )$: Child $c$ sending a response to downgrade to $x$ for address $a$ without any request from its parent}\\
+\hline
+Guard: & $(\forall i, \parent(i, c) \rightarrow c.\dstate[i][a] \le x) \wedge x < c.\state[q.\addr]$\\
+\hline
+Action: & $\enq(\cpResp[c], \langle c.\state[q.\addr], x, a,$\\
+& $\;\;\;%
+\myif c.\state[a] = \Mo \mythen c.\data[a] \myelse \_\rangle)$;\\
+& $c.\state[a] \Leftarrow x$;\\
+\hline
+\hline
+\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildRecvBadReq}$(c)$: Child $c$ receiving a request from its parent and has already downgraded}\\
+\hline
+Guard: & 
+$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge$\\
+& $q.\myto \ge c.\state[q.\addr]$\\
+\hline
+Action: & $\deq(\pc[c])$;\\
+\hline
+\end{tabular}
+\caption{Other transitions at a non-root cache}
+\end{subfigure}
+
+\begin{subfigure}{\textwidth}
+\centering
+\begin{tabular}{|ll|}
+\hline
+\multicolumn{2}{|l|}{\textbf{LoadReq}$(c)$: Handling a processor's load request}\\
 \hline
@@
 \hline
-\multicolumn{2}{|c|}{StoreReq$(c)$:  Handling a processor's store request}\\
+\multicolumn{2}{|l|}{\textbf{StoreReq}$(c)$: Handling a processor's store request}\\
 \hline

### 2014-01-28 paper/System.tex
stuff

@@
 \hline
-Action: & $\enq(\cpReq[c], \langle c.\state[a], x, a \rangle)$;\\
+Action: & $c.\waitS[a] \Leftarrow x$; $c.wait[a] \Leftarrow \True$; $\enq(\cpReq[c], \langle c.\state[a], x, a \rangle)$;\\
 \hline
@@
 \hline
-Action: & $\enq(\pc[c], \langle \Resp, p.\state[c][q.\addr], q.\myto, q.\addr, $\\
+Action: & $\enq(\pc[c], \langle \Resp, q.\myto, q.\addr, $\\
 & $\;\;\;%
@@
 \hline
-Action: & $\enq(\pc[c], \langle \Req, p.\dstate[c][a], x, a, \_ \rangle)$;\\
+Action: & $p.\dwaitS[c][a] \Leftarrow x$; $p.\dwait[c][a] \Leftarrow \True$; $\enq(\pc[c], \langle \Req, x, a, \_ \rangle)$;\\
 \hline
@@
 \hline
-\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildVolResp}$(c, x, )$: Child $c$ sending a response to downgrade to $x$ for address $a$ without any request from its parent}\\
+\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildVolResp}$(c, x, a)$: Child $c$ sending a response to downgrade to $x$ for address $a$ without any request from its parent}\\
 \hline
-Guard: & $(\forall i, \parent(i, c) \rightarrow c.\dstate[i][a] \le x) \wedge x < c.\state[q.\addr]$\\
+Guard: & $(\forall i, \parent(i, c) \rightarrow c.\dstate[i][a] \le x) \wedge x < c.\state[a]$\\
 \hline
-Action: & $\enq(\cpResp[c], \langle c.\state[q.\addr], x, a,$\\
-& $\;\;\;%
+Action: & $\enq(\cpResp[c], \langle c.\state[a], x, a,%$\\
+%& $\;\;\;%
 \myif c.\state[a] = \Mo \mythen c.\data[a] \myelse \_\rangle)$;\\
@@
 Guard: & 
-$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge$\\
-& $q.\myto \ge c.\state[q.\addr]$\\
+$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge%$\\
+%& $%
+q.\myto \ge c.\state[q.\addr]$\\
 \hline
@@
 \hline
-\multicolumn{2}{|l|}{\textbf{LoadReq}$(c)$: Handling a processor's load request}\\
+\multicolumn{2}{|l|}{\textbf{LoadReq}$(c)$: Handling a processor's load request at a leaf cache}\\
 \hline
-Guard: & $\leaf(c) \wedge \reqFn(c, c.\pos) = \Ld \wedge c.\state[\addrQ(\reqFn(c, c.\pos))] \ge \Sh$\\
+Guard: & $\leaf(c) \wedge \reqFn(c, c.\pos) = \Ld \wedge c.\state[(\reqFn(c, c.\pos)).\addrQ] \ge \Sh$\\
 \hline
@@
 \hline
-\multicolumn{2}{|l|}{\textbf{StoreReq}$(c)$: Handling a processor's store request}\\
+\multicolumn{2}{|l|}{\textbf{StoreReq}$(c)$: Handling a processor's store request at a leaf cache}\\
 \hline
-Guard: & $\leaf(c) \wedge \reqFn(c, c.\pos) = \St \wedge c.\state[\addrQ(\reqFn(c, c.\pos))] = \Mo$\\
+Guard: & $\leaf(c) \wedge \reqFn(c, c.\pos) = \St \wedge c.\state[(\reqFn(c, c.\pos)).\addrQ] = \Mo$\\
 \hline

### 2014-01-29 paper/Background.tex
Background cleanup

--- 2014-01-27 paper/Background.tex
@@
-\section{Background}
+\section{Correctness of our Cache Coherence System}
 
-In order to obtain the execution semantics of a multi-threaded program on a
-multi-processor system, one has to understand the memory consistency model of
-the underlying system. This is necessary because, as opposed to a direct access
-to a single-ported memory, the processors a) have store buffers which reorders
-the sequence of memory requests from the processor, for faster access (say by
-bypassing a store-value to a subsequent load request), and b) communicate with
-the memory through several levels of data caches for faster access. The memory
-consistency model describes the semantic behavior of such requests and the
-system ensures that the cache hierarchy and the store buffers do not violate
-the specified semantics.
+Before we can discuss our protocol framework and it's verification we
+must first establish exactly what it means for a system to be
+correct.
 
-Memory consistency models can be partitioned into two sub-parts
-\cite{Arvind-memory-model}: a) the ``global'' store-atomicity property for each
-location/address and b) the ``local'' reordering restrictions between requests
-to different locations within each processor. The store-atomicity property per
-address can be informally explained as follows: each location behaves like an
-atomic register in which no two writes (stores) to that location can happen
-simultaneously, and any read of that location (load) returns the previously
-written value in that location. Most memory consistency models require that the
-memory subsystem obeys the store-atomicity property. The reordering
-restrictions dictated by the memory consistency model pertain to reordering
-requests within the same processor (unlike the store-atomicty property which is
-global across the entire system), but across multiple addresses. For example,
-it dictates whether a store request to an address $x$ can be issued to the
-memory sub-system before a load request to a different address $y$, which was
-generated by the same processor earlier (\ie load $y$ happened before store $x$
-is program order, but the order of issue to memory sub-system is reversed).
+Intuitively, the easiest view of shared memory is as a single ported
+memory multiplexed amongst the many individual client/processors. In
+this view, all operations in the system appear to have a single unique
+total ordering.
 
-The mechanism employed by the memory sub-system to ensure store-atomicity per
-location, in the presence of caches, is known as \emph{cache coherence}. In the
-absence of caches, since there is exactly one copy of the data for each
-address, it is easy to ensure cache coherence. The problem comes in the
-presence of shared caches -- multiple copies of the same address can exist in
-different caches, and if these locations are independently written, then the
-same address can end up having different values in two different caches. Each
-location would no longer behave like an atomic register unless some sort of
-co-ordination happens between the different caches -- this co-ordination is
-summed up as the \emph{cache coherence protocol}.
+While simple, it is often useful for performance to permit some
+reorderings. For instance, in the context of a processor with local
+store buffer, we may wish to allow store values to bypass subsequent
+load requests. In general, any potential reordering may be considered
+a valid memory consistency model, though they may result in a stronger
+verification requirement.
 
-\subsection{Formal definition of store-atomicity}
-In this section we give the formal definition of the store-atomicity property.
+As part of our design exploration for our concrete protocol example,
+we may want to change the memory consistency model. To this end, we
+are using the notion of store-atomicity to define our memory
+model. Most memory consistency models require that the memory
+subsystem obeys the store-atomicity property and as such our design
+should be modifiable to any of these designs.
 
-The requests issued by the processors to the memory subsystem can be modeled
-using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
-processor $p$, \ie $\reqFn: (\textit{Processor} * \mathbb{N}) @-> \Request$.
-The processor name and the position of the request within the processor
-uniquely identifies the request.  Figure \ref{req-resp} shows the data-type
-corresponding to a request issued from the processor to the memory subsystem.
-A request contains the address (\addrQ) corresponding to the request, and a
-type indicating whether the request is a load or a store (\desc). In case of a
-store request, it also contains data (\dataQ) supplied by the processor for
-that address.
+A store-atomic memory model can be partitioned into two
+parts~\cite{Arvind-memory-model}. First, a ``global'' store-atomicity
+property stating that for every read and write access to a particular
+location in memory occur in some total order with each operation
+seeing the effects of previous operations and not later ones. Seconds
+a ``local'' reordering restrictions between requests to different
+locations within each processor. The advantage of this approach is
+that reordering restrictions pertain to reordering requests locally of
+the same processor, while the global store-atomicity property is
+constant throughout. Thus changing the memory model should not incur a
+reverification of the store-atomicity property. For our concrete
+protocol example, we will assume that our memory model is weakly
+ordered, in that there are inter-address dependencies between
+processors.
 
-Figure \ref{req-resp} also shows the data-type corresponding to a response
-issued from the memory subsystem to the processor.  A response contains a label
+Now that we have informally discussed the meaning store atomicity, we
+can now provide a formal definition.
+
+%% For example, it dictates whether a store request to an address $x$ can
+%% be issued to the memory sub-system before a load request to a
+%% different address $y$, which was generated by the same processor
+%% earlier (\ie load $y$ happened before store $x$ is program order, but
+%% the order of issue to memory sub-system is reversed).
+
+%% The mechanism employed by the memory sub-system to ensure store-atomicity per
+%% location, in the presence of caches, is known as \emph{cache coherence}. In the
+%% absence of caches, since there is exactly one copy of the data for each
+%% address, it is easy to ensure cache coherence. The problem comes in the
+%% presence of shared caches -- multiple copies of the same address can exist in
+%% different caches, and if these locations are independently written, then the
+%% same address can end up having different values in two different caches. Each
+%% location would no longer behave like an atomic register unless some sort of
+%% co-ordination happens between the different caches -- this co-ordination is
+%% summed up as the \emph{cache coherence protocol}.
+
+\subsection{Formal definition of store atomicity}
+
+To properly define the interface with between the processor and the
+caching system we must fine the request and responses for each
+processor.  Figure~\ref{req-resp} shows the
+data-type corresponding to a request issued from the processor to the
+memory subsystem.  A request contains the address (\addrQ)
+corresponding to the request, and a type indicating whether the
+request is a load or a store (\desc). In case of a store request, it
+also contains data (\dataQ) supplied by the processor for that
+address.A response contains a label
 (\labelR) to identify the request corresponding to the response. The label
@@
 globally among all the requests issued from all the processors.
+
+We model request of a processor formally with the function
+\reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
+processor $p$, \ie{} $\reqFn: (\textit{Processor} * \mathbb{N}) @->
+\Request$.  The processor name and the position of the request within
+the processor uniquely identifies a request.
 
@@
 violate the local ordering of the corresponding requests.
+
 \begin{defn}

### 2014-01-29 paper/Introduction.tex
Background cleanup

@@
-:qh\section{Introduction}
+\section{Introduction}
 \label{sec:Introduction}
@@
 verified. This protocol generally directly represents many of
-the key concurrent aspects of an implementation, \eg reads and
+the key concurrent aspects of an implementation, \eg{} reads and
 writes to the system are broken into local atomic micro-operations, but
-ignores implementation details \eg buffer sizing and memory access
+ignores implementation details \eg{} buffer sizing and memory access
 times. Second, once the protocol has been verified it is handed to
@@
 
-Because verified protocols are often described via Term Rewrite Systems (TRS).
-it was shown by Dave \etal{} that a verified protocol could be transliterated
-into Bluespec, a hardware description language based on Guarded Atomic Actions,
-a restriction of TRS \cite{Hoe:TCAD, DNA:CoherenceImplementation}. This reduces
-the correctness of the implementation to the correctness of the protocol,
-effectively removing any implementation bugs. Though in the current versions of
-the Bluespec language, each atomic action corresponds to one hardware clock
-cycle, there has been some work to reduce this restriction \cite{Karczmarek}.
-%This requires that the protocol description includes many important
-%implementation-level details which had previously been elided for simplicity
-%of the description. Practically, beyond some additional descriptional noise in
-%the proof this is not a show stopper.
+Because verified protocols are often described via Term Rewrite
+Systems (TRS).  it was shown by Dave \etal{} that a verified protocol
+could be transliterated into Bluespec, a hardware description language
+based on Guarded Atomic Actions, a restriction of
+TRS~\cite{Hoe:TCAD,DNA:CoherenceImplementation}. This reduces the
+correctness of the implementation to the correctness of the protocol,
+effectively removing any implementation bugs. Though in the current
+versions of the Bluespec language, each atomic action corresponds to
+one hardware clock cycle, there has been some work to reduce this
+restriction~\cite{Karczmarek}.
 
-The issue with this protocol transliteration approach is that design of a
-hardware system requires many design tweaks to reach optimal performance as one
-cannot practically know the correct low-level hardware results without
-evaluating the low-level representation.
-Each of these changes can potentially breaks
-correctness and a full reverification of correctness is needed in
-principle. As hardware designers are unlikely to have the expertise or
-skill this effectively nullifies the verification guarantee in design.
+The issue with this protocol transliteration approach is that design
+of a hardware system requires many design tweaks to reach optimal
+performance as one cannot practically know the correct low-level
+hardware results without evaluating the low-level representation.
+Each of these changes can potentially breaks correctness and a full
+reverification of correctness is needed in principle. As hardware
+designers are unlikely to have the expertise or skill this effectively
+nullifies the verification guarantee in design.
 
@@
 
-In practice, protocol variants rely on a small set of general ``local building
-block'' properties, \eg{} messages always get from source to destination or
-that the local cache's reaction to receiving a protocol message results in the
-appropriate messages being sent, the appropriate local state being changed \etc{}.
-% about parts of the
-%design to be true for the full system correctness.
-These properties are easily verifiable since they correspond to a small
-partition of the system, for instance a single cache.
-The details of how these properties hold vary depending on the precise design
-variant. For instance, the reason why some property holds, like the
-point-to-point FIFO ordering for messages transferred between two nodes,
-depends on the type of network; the proof will be very different for a mesh
+In practice, protocol variants rely on a small set of general ``local
+building block'' properties, \eg{} messages always get from source to
+destination or that the local cache's reaction to receiving a protocol
+message results in the appropriate messages being sent, the
+appropriate local state being changed \etc{}.  These properties are
+easily verifiable since they correspond to a small partition of the
+system, for instance a single cache.  The details of how these
+properties hold vary depending on the precise design variant. For
+instance, the reason why some property holds, like the point-to-point
+FIFO ordering for messages transferred between two nodes, depends on
+the type of network; the proof will be very different for a mesh
 network compared to a point-to-point network.
 
-In this paper, we describe an approach to provide this proof framework for any
-particular cache coherence protocol via the example of a MSI-style protocol and
-show how it can be extended to other protocols like MESI and MOESI. We show how
-the modular structure of the proof and design can be shared and how
-modification of micro-archtectural aspects of design directly translates to
-just reverification of our local building block properties. We use Coq for our
-proof framework. We also show how an atomic action based MSI protocol
-description can be directly transliterated into an Inductive definition in Coq,
-and verify its local building block properties formally, thus completing the
-picture.
+In this paper, we describe an approach to provide this proof framework
+for any particular cache coherence protocol via the example of a
+MSI-style protocol and show how it can be extended to other protocols
+like MESI and MOESI. We show how the modular structure of the proof
+and design can be shared and how modification of microarchitectural
+aspects of design directly translates to just reverification of our
+local building block properties. We use Coq for our proof
+framework. We also show how an atomic action based MSI protocol
+description can be directly transliterated into an Inductive
+definition in Coq, and verify its local building block properties
+formally, thus completing the picture.
+
+\noindent\textbf{Paper Outline}. In Section~\ref{Sec:Background} we
+discuss our notion of correctness for our cache coherence protocol
+framework. We discuss our MSI protocol example and its implementation
+in Section~\ref{Sec:Protocol}. In Section~\ref{Sec:Proof} we discuss
+the structuring of our proof framework and verification of the base
+protocol. We show implementation refinments in
+Section~\ref{Sec:Refinements} and conclude with a general discussion
+of the applicability of our technique.

### 2014-01-29 paper/System.tex
stuff

@@
 We will restrict the discussion of the formal proofs to the MSI protocol
-\cite{MSI}. Extension of this proof framework to other protocols like MESI and
-MOESI is straightforward and will be discussed informally.
+\cite{MSI}. In the MSI protocol, each cache maintains a permission for each
+address, denoting whether that address can be read and modified (\Mo), only
+read (\Sh) or neither read nor modified (\In) in that cache. Consequently, in a
+leaf cache, a load and store request can be processed by a cache only if the
+cache is at least in \Sh{} and \Mo{} states, respectively, for the corresponding
+address. For each address, caches which are in the subtree of a cache can not
+have higher permissions than the latter. Extension of this proof framework to
+other protocols like MESI and MOESI is straightforward and will be discussed
+informally.
 
@@
 transition.
+
+Figure \ref{childside} shows the transitions that result in an upgrade of the
+state of a cache. It first involves a cache sending an upgrade request to the
+its parent when the cache is not already waiting for a response for that
+address from the parent (ChildSendReq). This request is received by the parent
+(ParentRecvReq), and if the rest of the parent's children caches are
+\emph{compatible}, and the parents state also has the permission being
+requested, it sends an upgrade response to the cache, while changing the
+appropriate directory state. Compatibility of an upgrade request with respect
+to a cache's siblings can be defined as follows:
+\begin{multline*}
+\compat = \lambda (p \in Cache) (c \in Cache) (a \in \textit{Addr}) (s \in \{\Mo, \Sh, \In\}) \Rightarrow \forall i, i \ne c @-> \\
+\mylet s_i := p.\dstate[i][a] \myin\\
+ (s = \Mo @-> s_i = \In) \wedge (s = \Sh @-> s_i = \Sh) \wedge (s = \In @-> s_i = \Mo)
+\end{multline*}
+Finally, the child cache receives the response from the parent (ChildRecvResp)
+and sets its state while reseting its wait state if the desired upgrade is
+indicated by the response.
+
+Figure \ref{parentside} shows the transitions that result in a downgrade of the
+state of a cache because of a request by the cache's parent when the parent is
+not already waiting for a response from that cache (ParentSendReq).  This
+request is received by the child (ChildRecvReq), and if the state of each of
+the child's own children is not greater than the requested downgrade, the cache
+downgrades its state and sends a notification response to the parent. Finally,
+the parent receives the response from the cache (ParentRecvResp) and sets the
+appropriate directory state while reseting the appropriate directory wait state
+if the desired downgrade is indicated by the response.
+
+Figure \ref{childextra} shows transitions that happen in a non-root cache
+because of the presence of \emph{voluntary} responses. Whenever a cache runs
+out of space to store the data corresponding to an address, it evicts the data
+corresponding to one or more addresses to make room for the new address. This
+eviction causes the state of the cache for that address to go to $\In$, which
+has to be informed to its parent. We use the same \cpResp{} FIFO (and the same
+response format) to indicate this. Such a voluntary response for an address can
+take place only if the cache is not waiting for a response from its parent for
+that address (ChildVolResp).
+
+Because of the presence of voluntary response, a child cache could have already
+downgraded to a state required by a downgrade request from its parent. In such
+a case, the cache simply drops the downgrade request (ChildDropReq).
+
+Figure \ref{procside} shows the transitions that actually serve a processors
+request in the leaf caches. LoadReq and StoreReq show the processing of a load
+and store request, respectively, from a processor. In the case of a store
+request, the data in the cache for the requested address is updated with that
+supplied in the request. We do not explicitly show the response generated by
+the cache in case of a load request in our memory subsystem model; in a real
+system, the data in the cache for the requested address is sent back to the
+processor. Our definition of store atomicity requires responses for both load
+and store requests, and the response field contains the
+$(\textit{Processor}*\mathbb{N})$ pair identifying the request corresponding to
+the response, the data from the cache in case of a load response, and a value
+for the field \tme. A \Response{} is defined whenever LoadReq or StoreReq
+transition takes place. Let $c$ be a leaf cache, and $t$ be the position of a
+LoadReq$(c)$ or StoreReq$(c)$ transition in the ordered list of transitions of
+the full system starting from the initial state. The fields of the response are
+as follows:
+\begin{multline*}
+\small\langle (c, c.\pos), t, \mylet q := \reqFn(c, c.\pos) \myin\\
+\small\myif q.\desc = \Ld \mythen c.\data[q.\addr] \myelse \_ \rangle
+\end{multline*}
+
 \begin{figure}
@@
 & $p.\dwait[c][q.\addr] = \False \wedge p.\dstate[c][q.\addr] \le q.\from \wedge$\\
-& $\compat(p, c, q.\myto) \wedge q.\myto \le p.\state[q.\addr]$\\
+& $\compat(p, c, q.\addr, q.\myto) \wedge q.\myto \le p.\state[q.\addr]$\\
 \hline
@@
 \caption{Child sending upgrade request and getting back the response}
+\label{childside}
 \end{subfigure}
@@
 $\parent(c,p) \wedge \avail(\cpResp[c]) \wedge \mylet r := \first(\cpResp[c]) \myin$\\
-& $r.\from = c.\state[r.\addr]$\\
+& $r.\from = p.\state[c][r.\addr]$\\
 \hline
@@
 \caption{Parent sending downgrade request and getting back the response}
+\label{parentside}
 \end{subfigure}
@@
 \hline
-Guard: & $(\forall i, \parent(i, c) \rightarrow c.\dstate[i][a] \le x) \wedge x < c.\state[a]$\\
+Guard: & $(\forall i, \parent(i, c) \rightarrow c.\dstate[i][a] \le x) \wedge x < c.\state[a] \wedge c.\wait[a] = \False$\\
 \hline
@@
 \caption{Other transitions at a non-root cache}
+\label{childextra}
 \end{subfigure}
@@
 \hline
-Guard: & $\leaf(c) \wedge \reqFn(c, c.\pos) = \Ld \wedge c.\state[(\reqFn(c, c.\pos)).\addrQ] \ge \Sh$\\
+Guard: & $\leaf(c) \wedge \mylet q := \reqFn(c, c.\pos) \myin q.\desc = \Ld \wedge c.\state[q.\addrQ] \ge \Sh$\\
 \hline
@@
 \hline
-Guard: & $\leaf(c) \wedge \reqFn(c, c.\pos) = \St \wedge c.\state[(\reqFn(c, c.\pos)).\addrQ] = \Mo$\\
+Guard: & $\leaf(c) \wedge \mylet q := \reqFn(c, c.\pos) \myin q.\desc = \St \wedge c.\state[q.\addrQ] = \Mo$\\
 \hline
-Action:& $c.\pos \Leftarrow c.\pos + 1$;\\
+\hline
+Action:& $c.\pos \Leftarrow c.\pos + 1$; $c.\data[q.\addrQ] \Leftarrow q.\dataQ$;\\
 \hline
@@
 \caption{Handling requests from the processor}
+\label{procside}
 \end{subfigure}
@@
 
-Note that he atomic transitions in Figure \ref{trans} access only local states,
+Note that the atomic transitions in Figure \ref{trans} access only local states,
 \ie a transition can read or write states only corresponding to a single cache

### 2014-01-29 paper/Background.tex
stuff
some modifications of background

@@
 Intuitively, the easiest view of shared memory is as a single ported
-memory multiplexed amongst the many individual client/processors. In
+memory without caches multiplexed amongst the many individual clients/processors. In
 this view, all operations in the system appear to have a single unique
@@
 
-While simple, it is often useful for performance to permit some
-reorderings. For instance, in the context of a processor with local
-store buffer, we may wish to allow store values to bypass subsequent
-load requests. In general, any potential reordering may be considered
-a valid memory consistency model, though they may result in a stronger
-verification requirement.
+All modern systems have caches between the processor and memory for faster memory access.
+It is often also useful for performance to permit some reorderings fo requests from a processor to the memory.
+For instance, in the context of a processor with local store buffer, we may
+wish to allow store values to bypass subsequent load requests. The specification of
+ which reorderings are permitted and which are restricted defines the memory consistency
+model.
+%, though they may result in a stronger verification requirement.
 
-As part of our design exploration for our concrete protocol example,
-we may want to change the memory consistency model. To this end, we
-are using the notion of store-atomicity to define our memory
-model. Most memory consistency models require that the memory
-subsystem obeys the store-atomicity property and as such our design
-should be modifiable to any of these designs.
+%As part of our design exploration for our concrete protocol example,
+%we may want to change the memory consistency model. To this end, we
+%are using the notion of store-atomicity to define our memory
+%model. Most memory consistency models require that the memory
+%subsystem obeys the store-atomicity property and as such our design
+%should be modifiable to any of these designs.
 
-A store-atomic memory model can be partitioned into two
-parts~\cite{Arvind-memory-model}. First, a ``global'' store-atomicity
-property stating that for every read and write access to a particular
-location in memory occur in some total order with each operation
-seeing the effects of previous operations and not later ones. Seconds
-a ``local'' reordering restrictions between requests to different
+Most memory consistency models can be partitioned into two
+parts~\cite{Arvind-memory-model}: a), a ``global'' store-atomicity
+property stating that for each location/address in memory, every read and write access to that 
+location occurs in some total order with each operation
+seeing the effects of previous operations and not later ones (each location behaves like an atomic register) and b),
+a ``local'' reordering restriction between requests to different
 locations within each processor. The advantage of this approach is
-that reordering restrictions pertain to reordering requests locally of
-the same processor, while the global store-atomicity property is
-constant throughout. Thus changing the memory model should not incur a
-reverification of the store-atomicity property. For our concrete
-protocol example, we will assume that our memory model is weakly
-ordered, in that there are inter-address dependencies between
-processors.
+that reordering restrictions pertain to reordering requests locally within
+the same processor, while the store-atomicity property is across the whole
+system. Thus changing the memory model
+by modifying the reordering restrictions will not incur a reverification of
+the store-atomicity property.
 
-Now that we have informally discussed the meaning store atomicity, we
-can now provide a formal definition.
+%For our concrete
+%protocol example, we will assume that our memory model is weakly
+%ordered, in that there are inter-address dependencies between
+%processors.
 
@@
 
-%% The mechanism employed by the memory sub-system to ensure store-atomicity per
-%% location, in the presence of caches, is known as \emph{cache coherence}. In the
-%% absence of caches, since there is exactly one copy of the data for each
-%% address, it is easy to ensure cache coherence. The problem comes in the
-%% presence of shared caches -- multiple copies of the same address can exist in
-%% different caches, and if these locations are independently written, then the
-%% same address can end up having different values in two different caches. Each
-%% location would no longer behave like an atomic register unless some sort of
-%% co-ordination happens between the different caches -- this co-ordination is
-%% summed up as the \emph{cache coherence protocol}.
+The mechanism employed by the memory sub-system to ensure store-atomicity per
+location, in the presence of caches, is known as \emph{cache coherence}. In the
+absence of caches, since there is exactly one copy of the data for each
+address, it is easy to ensure cache coherence. The problem comes in the
+presence of shared caches -- multiple copies of the same address can exist in
+different caches, and if these locations are independently written, then the
+same address can end up having different values in two different caches, and
+each location may not behave like an atomic register unless some sort of
+co-ordination happens between the different caches -- this co-ordination is
+summed up as the \emph{cache coherence protocol}.
+
+Now that we have informally discussed the meaning of cache coherence and store
+atomicity, we can now provide a formal definition.
 
@@
 To properly define the interface with between the processor and the
-caching system we must fine the request and responses for each
-processor.  Figure~\ref{req-resp} shows the
+caching system we must define the request and responses for each
+processor. 
+The requests issued by the processors to the memory sub-system can be modeled
+using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
+processor $p$.
+The processor name and the position of the request within the processor
+uniquely identifies the request.
+
+Figure~\ref{req-resp} shows the
 data-type corresponding to a request issued from the processor to the
-memory subsystem.  A request contains the address (\addrQ)
+memory sub-system.  A request contains the address (\addrQ)
 corresponding to the request, and a type indicating whether the
@@
 also contains data (\dataQ) supplied by the processor for that
-address.A response contains a label
+address. A response contains a label
 (\labelR) to identify the request corresponding to the response. The label
@@
 In the case of a load request, the memory sub-system returns \dataR, the data
-returned by the memory subsystem for that address. The memory sub-system also
+returned by the memory sub-system for that address. The memory sub-system also
 returns a natural number, \timeR, which is used solely in the definition of
@@
 
-We model request of a processor formally with the function
-\reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
-processor $p$, \ie{} $\reqFn: (\textit{Processor} * \mathbb{N}) @->
-\Request$.  The processor name and the position of the request within
-the processor uniquely identifies a request.
+%We model request of a processor formally with the function
+%\reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
+%processor $p$, \ie{} $\reqFn: (\textit{Processor} * \mathbb{N}) @->
+%\Request$.  The processor name and the position of the request within
+%the processor uniquely identifies a request.
 
@@
 \labelR:& Label identifying the request corresponding to this response\\
-\dataR:& Data returned by the memory subsystem for a \Ld{} request\\
 \timeR:& Natural number representing global order of (processing) requests\\
+\dataR:& Data returned by the memory sub-system for a \Ld{} request\\
 \hline
@@
 In the following, a \Response{} set only contains those responses sent by the
-memory subsystem.
+memory sub-system.
 

### 2014-01-30 paper/Introduction.tex
System is a little more motivated. Pull out bits of Proof structure to end and drop into new section

@@
 framework. We discuss our MSI protocol example and its implementation
-in Section~\ref{Sec:Protocol}. In Section~\ref{Sec:Proof} we discuss
+in Section~\ref{Sec:Protocol}. In Section~\ref{Sec:ProofStructure} we discuss
 the structuring of our proof framework and verification of the base
-protocol. We show implementation refinments in
+protocol. We show implementation refinements in
 Section~\ref{Sec:Refinements} and conclude with a general discussion

### 2014-01-30 paper/ProofStructure.tex
System is a little more motivated. Pull out bits of Proof structure to end and drop into new section

\section{Understanding the Correctness of the Protocol}
\label{Sec:ProofStructure}

For our protocol to be correct it must conform to our notion of
correct behavior \eg{} store atomicity Intuitively, we can understand
the correctness at a high-level from the following two facts.  that
particular address. Second, that any child node upgrades its local
state from \In to either \Sh or \Mo, it will get the same version of
data as it's parent which is the current global value.

%%%%%%%%%%%%%%%%%

Now that we have established a high-level view of the base protocol,
we will concretely discuss the structure of the proof. Much of the
structure discussed is necessary for the proof of correctness only and
not the actual implementation, \eg{} time. These additional components
can easily be optimized away as they're values are not used in the
actual handling of cache operations.

### 2014-01-30 paper/System.tex
System is a little more motivated. Pull out bits of Proof structure to end and drop into new section

@@
 
-For proving store-atomicity of the memory subsystem, it is sufficient to model
-just the memory subsystem, instead of modeling the processors.
-The memory subsystem consists of a tree hierarchy of caches. These
-caches are inclusive, \ie a cache at a higher level (near the root) contains
-all the addresses present in caches in the lower levels of its substree. The
-root of the communicates with the main memory, and the leaf caches communicate
-with the processor (Figure \ref{hier}). Whether a cache is a leaf or not is
-denoted by the relation \leaf$(c)$.
+We have chosen as our concrete example is a multi-step hierarchical
+MSI protocol~\cite{MSI}. Conceptually, our cache coherence system is
+organized as in a tree-like structure with the root node as the main
+memory, the leaves as the last local cache for individual processors
+interfaces and the remaining nodes as intermediate cache structures
+(see Figure~\ref{hier}). Whether a cache is a leaf or not is denoted
+by the relation \leaf$(c)$.
 
-We will restrict the discussion of the formal proofs to the MSI protocol
-\cite{MSI}. In the MSI protocol, each cache maintains a permission for each
-address, denoting whether that address can be read and modified (\Mo), only
-read (\Sh) or neither read nor modified (\In) in that cache. Consequently, in a
-leaf cache, a load and store request can be processed by a cache only if the
-cache is at least in \Sh{} and \Mo{} states, respectively, for the corresponding
-address. For each address, caches which are in the subtree of a cache can not
-have higher permissions than the latter. Extension of this proof framework to
-other protocols like MESI and MOESI is straightforward and will be discussed
-informally.
+Input and output happen at the leaves in the form of memory requests
+issued by the associated processor $c$. These will be represented by
+an unbounded input and output arrays and position value denoted
+$c.\pos$, which represents the position of the next request to
+process; it is set to $0$ initially.
 
-Each leaf cache $c$ has a \emph{position} denoted by $c.\pos$, which represents
-the position of the next request from the processor corresponding to $c$ that
-it has to process. Since it has to process the first request first, it is set
-to $0$ initially.
+In the MSI protocol, each node maintains the apermission state for
+each address, denoting what operations may be performed on the
+data. An address may be in the modified state (\Mo) representing that
+it may be both read and modified, or in the shared state (\Sh) where
+it may only be read, or in the Invalid state (\In), meaning it can be
+neither modified or read. This forms a natural total order of
+permissions (\In $<$ \Sh $<$ \Mo). As we will discuss later, this can
+naturally extended to MESI and MOESI based protocols.
 
-Each cache $c$, in addition to the data associated with each address $a$ (denoted
-by $c.\data[a]$), also has a \emph{coherence state} associated with the
-address (denoted by $c.\state[a]$). The coherence state can be
-one of the three values: \Mo{} (Modified), \Sh{} (Shared) or \In{} (Invalid).
-\Mo{} state denotes that the subtree rooted at the cache in \Mo{} state for
-that address can modify the data, \ie{} both read and write the data
-corresponding to the address. \Sh{} state denotes that the subtree rooted at
-the cache in \Sh{} state for that address can only read the data corresponding
-to that address and \In{} state denotes that the subtree rooted at the cache in
-\In{} state can neither read nor write the data corresponding to the address.
-The meanings of the state is clear, but one must verify the actual cache
-coherence protocol for violations of these invariants, which is what the proof
-in this paper intends to do.  There's a logical ordering between the states, in
-terms of the permissions that they have: $\Mo > \Sh > \In$ (the notation should
-be read assuming that the order is transitively closed). The states for each of
-the addresses in each of the caches, other than the root cache, is set to \In,
-while states for each address in the root cache is set to \Mo. This is because
-only the main memory (connected to the root cache) has the initial copy of
-data.  In addition, each non-leaf cache maintains a \emph{directory}, which
-contains a local version of the state of each of its children caches for a
-particular address.  For a non-leaf cache $p$, the state in its directory for
-child cache $c$ is denoted by $p.\dstate[c][a]$, for address $a$.  The relation
-$\parent(c, p)$ denotes that $p$ is the parent of cache $c$.  This local
-version may be stale, but it should not be lower than the real state of the
-child (again this property has to be proven for a cache coherence protocol).
-All the directory states are set to \In initially because initially, all the
-non-root caches have their states for all the addresses set to \In.
+Each cache in addition to the data associated with each address $a$
+(denoted by $c.\data[a]$). It also has a \emph{coherence state}
+associated with the address (denoted by $c.\state[a]$) holding the
+permissions on that address (\Mo, \Sh, or \In). At this level, each
+node has a complete cache of the entire memory system; we will
+restrict these caches to be partial as a refinement later. Initially,
+all address in each cache is in the \In state, save for those in the
+root cache which are in the \Mo state.
 
-In order for a cache to change to a higher state for a given address, \ie
-\emph{upgrade}, it has to \emph{request} its parent cache for increasing the
-state.  For a cache to change to a lower state, \ie \emph{downgrade}, it has to
-notify its parent cache indicating the downgrade -- it uses \emph{response}
-messages for this purpose. On the other hand, for a parent cache to force a
-child cache to downgrade the child's state for an address (in order to give
-permissions to a different child cache), or equivalently, to downgrade the
-parent's directory state for the child, it has to \emph{request} that child
-cache to downgrade, and a parent can notify a child about upgrading the child's
-state, or equivalently, about upgrading the parent's directory state for the
-child, by sending a \emph{response} message.
+Non-leaf nodes also keep a copy of the coherence state of each of its
+children in a \emph{directory}, which contains a the state of each of
+its children caches for a particular address.  For a non-leaf cache
+$p$, the state in its directory for child cache $c$ is denoted by
+$p.\dstate[c][a]$, for address $a$.  The relation $\parent(c, p)$
+denotes that $p$ is the parent of cache $c$.  This local version may
+be stale, in that the child may have changed it state but the parent
+has not yet updated itself. However, the parent state will always be a
+conservative overapproximation of the child state.
 
-Each cache is connected to its parent cache using 3 logical FIFO connections: a)
-\cpReq{} channel to transmit requests from the child to the parent, b) \cpResp{}
-channel to transmit responses from the child to the parent, and c) \pc{}
-channel to transmit both request and response messages from
-parent to child.
+In order for a cache to change to a higher state for a given address,
+\ie{} \emph{upgrade}, it has to \emph{request} its parent.  For a
+cache to change to a lower state, \ie \emph{downgrade}, it has to
+notify its parent cache indicating the downgrade -- it uses
+\emph{response} messages for this purpose. Conversely, for a parent
+cache to force a child cache to downgrade the child's state for an
+address (in order to give permissions to a different child cache), or
+equivalently, to downgrade the parent's directory state for the child,
+it has to \emph{request} that child cache to downgrade, and a parent
+can notify a child about upgrading the child's state, or equivalently,
+about upgrading the parent's directory state for the child, by sending
+a \emph{response} message.
 
-%In addition, the leaf caches (called L1 caches usually)
-%communicate with the processor through two channels: a) \cproc{} \textbf{FIFO}
-%channel to transmit requests from the processor to L1 cache, and b) \procc{}
-%\textbf{FIFO} channel to transmit responses from the L1 cache to the processor.
-%There is exactly one of \cproc{} and \procc{} channels between the L1 and the processor.
+Each cache is connected to its parent cache using 3 logical FIFO
+connections: a) \cpReq{} channel to transmit requests from the child
+to the parent, b) \cpResp{} channel to transmit responses from the
+child to the parent, and c) \pc{} channel to transmit both request and
+response messages from parent to child.
 
-In this paper, we will assume that each channel has a dedicated link between
-the communicating caches. In a real system, these channels will be logically
-implemented over an interconnect network, which poses additional problems that
-have to be proven -- namely that of deadlock and starvation-freedom. Though we
-have proved these results for a general interconnect network as opposed to a
-point-to-point connection between caches, we completely omit the formal
-discussion of deadlock and starvation freedom in this paper because of lack of
-space. We will informally discuss how deadlock and starvation freedom
-properties are proved for our system.
+Since we have dedicated channel between each parent-child pair, we
+will index the three channels (\cpReq{}, \cpResp{} and \pc{}) by the
+name of the child that uses that channel (using the notation
+$\cpReq[c]$, \etc for the \cpReq{} channel associated with cache $c$).
+Figure \ref{format} shows the message format transmitted in the 3
+channels between caches. In each of the channels $ch$, a message $m$
+can be enqueued ($\enq(ch, m)$) or dequeued ($\deq(ch)$), and the
+channels can be examined for the first element ($\first(ch)$) or for
+availability of any message in the channel ($\avail(ch)$).  Each of
+these channels start out empty \ie $\forall ch, \neg \avail(ch)$,
+initially.
 
-Since we have dedicated channel between each pair of child-parent, we will
-index the three channels (\cpReq{}, \cpResp{} and \pc{}) by the name of the
-child that uses that channel (using the notation $\cpReq[c]$, \etc for the
-\cpReq{} channel associated with cache $c$).  Figure \ref{format} shows the
-message format transmitted in the 3 channels between caches. In each of the
-channels $ch$, a message $m$ can be enqueued ($\enq(ch, m)$) or dequeued
-($\deq(ch)$), and the channels can be examined for the first element
-($\first(ch)$) or for availability of any message in the channel ($\avail(ch)$).
-Each of these channels start out empty \ie $\forall ch, \neg \avail(ch)$, initially.
 %The messages transmitted between the processor and the L1 cache
@@
 
-In addition to the actual coherence states, each cache also maintains a
-temporary \emph{wait state} for each of its addresses. The wait state contains
-two parts: a) a bool representing whether the cache has sent a request to its
-parent cache earlier and is waiting for a response (denoted by $c.\wait[a]$ for
-address $a$) and b) one of the three values \Mo, \Sh{} or \In representing the
-state upgrade that the cache is waiting for (denoted by $c.\waitS[a]$ for
-address $a$). Similarly, each non-leaf cache has the corresponding wait states
-for its directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting for a
-response from its child $c$ for address $a$, and $p.\dwaitS[c][a]$ denotes the
-state that cache $p$ is waiting for $c$ to downgrade to. Both \wait{} and
-\dwait{} are \False{} initially for each of the caches, and each of the
-addresses, because no cache is waiting for any response from either its child
-or its parent initially.
+In addition to the actual coherence states, each cache also maintains
+a temporary \emph{wait state} for each address to deal with
+transitioning between permission levels. The wait state consists of
+two parts: a) a boolean representing whether the cache has sent a
+request to its parent cache earlier and is waiting for a response
+(denoted by $c.\wait[a]$ for address $a$) and b) one of the three
+values \Mo, \Sh{} or \In representing the state upgrade that the cache
+is waiting to make (denoted by $c.\waitS[a]$ for address
+$a$). Similarly, each non-leaf cache has the corresponding wait states
+for its directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting
+for a response from its child $c$ for address $a$, and
+$p.\dwaitS[c][a]$ denotes the state to which that cache $p$ is waiting
+for $c$ to downgrade.
 
-The behavior of the system can be given in terms of atomic state transitions.
-These transitions have two parts: a) the guard, which dictates when the
-transition can fire, and b) the action, which dictates how the state of the
+%% Both \wait{} and \dwait{} are \False{} initially
+%% for each of the caches, and each of the addresses, because no cache is
+%% waiting for any response from either its child or its parent
+%% initially.
+
+The behavior of the system are described via atomic state transitions.
+These transitions have two parts: the guard dictating when the
+transition can fire, and the action defining how the state of the
 system changes when the transition happens. These transitions or \emph{guarded
 atomic actions} logically happen one by one, making the overall behavior of the
-system well defined. Figure \ref{trans} gives the atomic transitions for a
+system well defined. Figure~\ref{trans} gives the atomic transitions for a
 cache coherent memory subsystem. The actions in the transitions repesent only
@@
 
-Figure \ref{childside} shows the transitions that result in an upgrade of the
+Figure~\ref{childside} shows the transitions that result in an upgrade of the
 state of a cache. It first involves a cache sending an upgrade request to the
@@
 
-Figure \ref{parentside} shows the transitions that result in a downgrade of the
+Figure~\ref{parentside} shows the transitions that result in a downgrade of the
 state of a cache because of a request by the cache's parent when the parent is
@@
 
-Figure \ref{childextra} shows transitions that happen in a non-root cache
+Figure~\ref{childextra} shows transitions that happen in a non-root cache
 because of the presence of \emph{voluntary} responses. Whenever a cache runs
@@
 
-Figure \ref{procside} shows the transitions that actually serve a processors
+Figure~\ref{procside} shows the transitions that actually serve a processors
 request in the leaf caches. LoadReq and StoreReq show the processing of a load
@@
 Guard: & 
-$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge%$\\
+%$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge%$\\
 %& $%
-q.\myto \ge c.\state[q.\addr]$\\
+%q.\myto \ge c.\state[q.\addr]$\\
+$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge q.\myto \ge c.\state[q.\addr]$\\
 \hline
@@
 
-Note that the atomic transitions in Figure \ref{trans} access only local states,
-\ie a transition can read or write states only corresponding to a single cache
+Note that the atomic transitions in Figure~\ref{trans} access only local states,
+\ie{} a transition can read or write states only corresponding to a single cache
 and/or the channel the cache is connected to. This restriction allows such a
@@
 several transitions simultaneously, as long as they do not access the same
-state, \ie do not \emph{conflict} \cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
+state, \ie{} do not \emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
 Sometimes, these transitions (both the guards and the actions) may not be
@@
 register array (like a cache) can actually take several hardware clock cycles,
-though logically it is a single transition.  Karczmarek \etal \cite{Karczmarek}
+though logically it is a single transition.  Karczmarek \etal~\cite{Karczmarek}
 has provided a scheme to convert atomic transitions into synchronous hardware
 in which each transition can span several clock cycles.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+
+%In addition, the leaf caches (called L1 caches usually)
+%communicate with the processor through two channels: a) \cproc{} \textbf{FIFO}
+%channel to transmit requests from the processor to L1 cache, and b) \procc{}
+%\textbf{FIFO} channel to transmit responses from the L1 cache to the processor.
+%There is exactly one of \cproc{} and \procc{} channels between the L1 and the processor.

### 2014-01-30 paper/Abstract.tex
Abstract added
Clean up the intro in light of the abstract. Sounds a little more natural now.

Correct implementation of cache coherent systems of high importance in
industrial use. While significant efforts have been made in formally
verifying that the underlying protocol, this provides no formal
guarantees to the actual physically implemented hardware cache
coherence engine, which are generally derived manually and entail
nontrivial design-oriented refinements, \eg{} buffer sizing, caching
policies, multiplexing of distinct communication channels. Previously,
it was shown that that a formal model protocol description could be
almost automatically translated into a hardware description allowing
protocol proof to be reused for the implementation; however, any
hardware-oriented change to the design resulted in a new protocol and
required a complete reverification. We propose a metholoogy where the
proof is structured to be modular and shown mechnically allowing
designers to make desired changes to the implementation without
inducing further verification overhead. We show this via the concrete
example of an hierarchical MSI protocol and many interesting design variants.

### 2014-01-30 paper/Introduction.tex
Abstract added
Clean up the intro in light of the abstract. Sounds a little more natural now.

@@
 registers. This involves further breaking down of local micro-operations to
-make it realizable in hardware. This effort is not a one time translation, as
+make it realizable in hardware. This effort is not a one-time translation, as
 performance properties are not easily estimated; a significant amount of
@@
 
-From a methodological perspective, this two step process represent a
-significant methodological failure. We spent a significant amount of
-time verifying the protocol and then invalidate any correctness
-guarantees to the implementation by performing not only a non-trivial
-translation from the protocol representation to hardware, but also
-the refinement to add missing implementation details.
+From a methodological perspective, this two-step process represent a
+significant methodological failure. The significant effort of
+verification is effectively lost by the concurrent non-trivial tasks
+of translation from the protocol representation to a hardware
+description and the necessary refinement to add missing implementation
+details to the protocol.
 
 Because verified protocols are often described via Term Rewrite
-Systems (TRS).  it was shown by Dave \etal{} that a verified protocol
+Systems (TRS). It was shown by Dave
+\etal{}~\cite{DNA:CoherenceImplementation} that a verified protocol
 could be transliterated into Bluespec, a hardware description language
 based on Guarded Atomic Actions, a restriction of
-TRS~\cite{Hoe:TCAD,DNA:CoherenceImplementation}. This reduces the
-correctness of the implementation to the correctness of the protocol,
-effectively removing any implementation bugs. Though in the current
-versions of the Bluespec language, each atomic action corresponds to
-one hardware clock cycle, there has been some work to reduce this
-restriction~\cite{Karczmarek}.
+TRS~\cite{Hoe:TCAD}. This reduces the correctness of the
+implementation to the correctness of the protocol, effectively
+removing any implementation bugs. While the current Bluespec compiler
+has certain restrictions which limit the expressiveness of certain
+operations \eg{} fairness is not guaranteed, atomic actions are always
+implemented in a single cycle, work has been done to remove these
+restrictions~\cite{DNA:CoherenceImplementation, Karczmarek}.
 
-The issue with this protocol transliteration approach is that design
-of a hardware system requires many design tweaks to reach optimal
-performance as one cannot practically know the correct low-level
-hardware results without evaluating the low-level representation.
-Each of these changes can potentially breaks correctness and a full
-reverification of correctness is needed in principle. As hardware
-designers are unlikely to have the expertise or skill this effectively
-nullifies the verification guarantee in design.
+While better, this approach of protocol transliteration has
+flaws. Design of a hardware system requires many tweaks to reach
+optimal performance as one cannot practically know the correct
+low-level hardware results without evaluating the low-level
+representation.  Each of these changes can potentially breaks
+correctness and a full reverification of correctness is needed in
+principle. As hardware designers are unlikely to have the expertise or
+skill to do make this verification, this greatly weakens the ability
+to carry over our verification guarantee to the design.
 
-To fully realize this approach of design we need to provide not just a
-single proof of correctness, but a proof framework covering any
+For this approach to be truly viable, we need not a single proof of
+correctness of a single design, but a proof framework covering any
 variation of a design that might be of interest for performance
@@
 local building block properties. We use Coq for our proof
-framework. We also show how an atomic action based MSI protocol
+framework. We also show how an atomic-action-based MSI protocol
 description can be directly transliterated into an Inductive
 definition in Coq, and verify its local building block properties
-formally, thus completing the picture.
+formally.
 

### 2014-01-30 paper/Title.tex
Abstract added
Clean up the intro in light of the abstract. Sounds a little more natural now.

@@
 
-\title{Stuff
-}
+\title{Proof Modularization: Enabling Verified Design-Oriented Modifications in Cache Coherence Implementation}
 
@@
 \author{
-    Muralidaran Vijayaraghavan\\Massachusetts Institute of Technology\\vmurali@csail.mit.edu\and
-    Nirav Dave\\SRI International\\ndave@csl.sri.com\and
-    Arvind\\Massachusetts Institute of Technology\\arvind@csail.mit.edu
+%    Muralidaran Vijayaraghavan\\Massachusetts Institute of Technology\\vmurali@csail.mit.edu\and
+%    Nirav Dave\\SRI International\\ndave@csl.sri.com\and
+%    Adam Chlipala\\Massachusetts Institute of Technology\\adamc@csail.mit.edu\and
+%    Arvind\\Massachusetts Institute of Technology\\arvind@csail.mit.edu
 }

### 2014-01-31 paper/Abstract.tex
stuff

@@
 Correct implementation of cache coherent systems of high importance in
-industrial use. While significant efforts have been made in formally
-verifying that the underlying protocol, this provides no formal
-guarantees to the actual physically implemented hardware cache
-coherence engine, which are generally derived manually and entail
-nontrivial design-oriented refinements, \eg{} buffer sizing, caching
+industrial use.
+
+Model checking techniques which are typically used to verify
+the protocols require both scaling down and simplifying the system rather than
+modeling the full implementation.
+
+ can be used to tackle these
+issues by providing parameterized proofs. 
+We use the Coq theorem prover to
+
+We use the Coq theorem
+prover to provide a modular proof of safety properties for two distinct cache coherence
+protocols (MSI and MOESI).
+We identify a set of invariants which all these protocols obey  all of which obey common invariants.
+We use the example of an MSI protocol specified as a set of local atomic transitio.
+We also show how large portions of this
 policies, multiplexing of distinct communication channels. Previously,

### 2014-01-31 paper/System.tex
stuff

@@
 
-We have chosen as our concrete example is a multi-step hierarchical
-MSI protocol~\cite{MSI}. Conceptually, our cache coherence system is
+We have chosen as our concrete example is a hierarchical
+MSI protocol~\cite{MSI}. As we will
+discuss later, this can naturally extended to MESI and MOESI based protocols.
+Conceptually, our cache coherence system is
 organized as in a tree-like structure with the root node as the main
@@
 (see Figure~\ref{hier}). Whether a cache is a leaf or not is denoted
-by the relation \leaf$(c)$.
+by the relation \leaf$(c)$.  The relation $\parent(c, p)$
+denotes that $p$ is the parent of cache $c$.
 
 Input and output happen at the leaves in the form of memory requests
-issued by the associated processor $c$. These will be represented by
-an unbounded input and output arrays and position value denoted
-$c.\pos$, which represents the position of the next request to
-process; it is set to $0$ initially.
+issued by the associated processor. These will be represented by
+unbounded input and output arrays. For each leaf $c$, $c.\pos$ represents the
+position of the next request to process; it is set to $0$ initially.
 
-In the MSI protocol, each node maintains the apermission state for
+Each node maintains the a permission state for
 each address, denoting what operations may be performed on the
@@
 neither modified or read. This forms a natural total order of
-permissions (\In $<$ \Sh $<$ \Mo). As we will discuss later, this can
-naturally extended to MESI and MOESI based protocols.
+permissions ($\In < \Sh < \Mo$). For leaf caches, the meaning of these
+permission states is straightfoward, a load request can be processed only if the
+leaf is at least in the \Sh{} state, and a store request can be processed only
+if it is in the \Mo{} state. For the non-leaf cache, these states denote the
+highest state present in the subtree rooted at this cache, denoting that some
+cache in the subtree has the specified permissions (we are modeling inclusive
+caches). Of course, all these invariants have to be enforced by the protocol.
 
 Each cache in addition to the data associated with each address $a$
-(denoted by $c.\data[a]$). It also has a \emph{coherence state}
+(denoted by $c.\data[a]$) also has a \emph{coherence state}
 associated with the address (denoted by $c.\state[a]$) holding the
-permissions on that address (\Mo, \Sh, or \In). At this level, each
-node has a complete cache of the entire memory system; we will
-restrict these caches to be partial as a refinement later. Initially,
-all address in each cache is in the \In state, save for those in the
-root cache which are in the \Mo state.
+permissions on that address (\Mo, \Sh, or \In). If a cache does not have a
+particular location, the it is equivalent to its permission being \In.
+Initially, all addresses in each cache is in the \In state, save for those in
+the root cache which are in the \Mo state.
 
@@
 $p$, the state in its directory for child cache $c$ is denoted by
-$p.\dstate[c][a]$, for address $a$.  The relation $\parent(c, p)$
-denotes that $p$ is the parent of cache $c$.  This local version may
+$p.\dstate[c][a]$, for address $a$. This local version may
 be stale, in that the child may have changed it state but the parent
@@
 
-%The messages transmitted between the processor and the L1 cache
-%is the same as in Figure \ref{req-resp} (except for the omission of the \proc{}
-%field, which is indicated by the channel's index).
 \begin{figure}
@@
 for a response from its child $c$ for address $a$, and
-$p.\dwaitS[c][a]$ denotes the state to which that cache $p$ is waiting
-for $c$ to downgrade.
-
-%% Both \wait{} and \dwait{} are \False{} initially
-%% for each of the caches, and each of the addresses, because no cache is
-%% waiting for any response from either its child or its parent
-%% initially.
+$p.\dwaitS[c][a]$ denotes the state to which $p$ is waiting
+for $c$ to downgrade. Both \wait{} and \dwait{} are \False{} initially
+for each of the caches, and each of the addresses, because no cache is
+waiting for any response from either its child or its parent
+initially.
 
@@
 \hline
-\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildRecvBadReq}$(c)$: Child $c$ receiving a request from its parent and has already downgraded}\\
+\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildDropReq}$(c)$: Child $c$ receiving a request from its parent and has already downgraded}\\
 \hline

### 2014-02-03 paper/Abstract.tex
stuff

--- 2014-01-30 paper/Abstract.tex
@@
-Correct implementation of cache coherent systems of high importance in
-industrial use. While significant efforts have been made in formally
-verifying that the underlying protocol, this provides no formal
-guarantees to the actual physically implemented hardware cache
-coherence engine, which are generally derived manually and entail
-nontrivial design-oriented refinements, \eg{} buffer sizing, caching
-policies, multiplexing of distinct communication channels. Previously,
-it was shown that that a formal model protocol description could be
-almost automatically translated into a hardware description allowing
-protocol proof to be reused for the implementation; however, any
-hardware-oriented change to the design resulted in a new protocol and
-required a complete reverification. We propose a metholoogy where the
-proof is structured to be modular and shown mechnically allowing
-designers to make desired changes to the implementation without
-inducing further verification overhead. We show this via the concrete
-example of an hierarchical MSI protocol and many interesting design variants.
+Correct implementation of cache coherent systems is of high importance in
+industrial use. Model-checking is inadequate for verifying a large system with
+hierarchical caches because of its size and complexity. We provide a
+theorem-prover verified proof for the safety properties of a size-parameterized
+hierarchical MSI-based coherent system. The system is specified in terms of
+atomic transitions accessing only local state.  Previously, it has been shown
+that these atomic transitions can be translated automatically into efficient
+hardware. Verifying the specification based on atomic transitions therefore
+guarantees the correctness of the actual implementation.
+The proof is built in a modular fashion, identifying key invariants common with
+other protocols like MESI, MOSI, \etc. We informally show how these other
+protocols can be proved reusing modules from the proof of the MSI protocol.

### 2014-02-03 paper/System.tex
stuff

@@
 Input and output happen at the leaves in the form of memory requests
-issued by the associated processor. These will be represented by
-unbounded input and output arrays. For each leaf $c$, $c.\pos$ represents the
+issued by and responses issued to the associated processor. These will be
+represented by two mappings one from $\mathbb{N}$ to requests and the other
+from $\mathbb{N}$ to responses. For each leaf $c$, $c.\pos$ represents the
 position of the next request to process; it is set to $0$ initially.
@@
 \begin{multline*}
-\compat = \lambda (p \in Cache) (c \in Cache) (a \in \textit{Addr}) (s \in \{\Mo, \Sh, \In\}) \Rightarrow \forall i, i \ne c @-> \\
+\compat = \lambda (p \in Cache) (c \in Cache) (a \in \textit{Addr}) (s \in
+\{\Mo, \Sh, \In\}) \Rightarrow\\ \forall i, i \ne c @->
 \mylet s_i := p.\dstate[i][a] \myin\\

### 2014-02-03 paper/Title.tex
stuff

@@
 
-\title{Proof Modularization: Enabling Verified Design-Oriented Modifications in Cache Coherence Implementation}
+\title{Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof}
+
+% hierarchical cache coherence
+% modular proof
 
@@
 
-%\centerline{\footnotesize {\bf Keywords}: Guarded Atomic Actions, HW Synthesis, Modular Composition}
+\centerline{\footnotesize {\bf Keywords}: Cache coherence, theorem prover, cache hierarchy, guarded atomic actions}
 

### 2014-02-03 paper/Intuition.tex
stuff

\section{Intuition}

In this section we give the proof of the MSI cache coherence protocol. We give
the proof in a top-down fashion starting with the store atomicity theorem. We
specify important invariants and show how they are proved in the MSI protocol.
In Section \ref{Refinements} show how these invariants hold in other protocols
like MESI and MOSI. These invariants are true at each step in the (infinite)
list of transitions starting from the initial state. We use the term time $t$
to specify the global state reached after $t$ transitions.  All the invariants
below pertain to a particular address; we will implicitly assume a universal
quantification over all address as a prefix to each of the invariants (if not
explicitly mentioned).

\subsection{Top invariants}
\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address iff the state of the
cache is at least \Sh{} and the state of all its children is at most \Sh.
\end{defn}
\begin{inv}
At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
up-to-date value for that address, \ie
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request has been processed at
any time $< t$, or
\item There is a store request $q$ that was processed at time $t'$ such that
$t' < t$ and $a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t_i$ such that $t' < t_i < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}

The following invariants hold for leaf caches.
\begin{inv}
\textit{uniqM}: At all times, at most one leaf cache can be in \Mo{} state.
\label{uniqM}
\end{inv}

\begin{inv}
\textit{processReq}: At all times, a load or a store request at a leaf cache can be processed
only if it is at least in \Sh{} or \Mo{} state at that time, respectively.
\label{processReq}
\end{inv}

\begin{inv}
\textit{noReqAgain}: A leaf cache $c$ processes the $c.\pos^{th}$ request iff
$c.\pos$ changes. $c.\pos$ only gets incremented when it changes.
\label{noReqAgain}
\end{inv}

Using Invariants \ref{latestValue} to \ref{noReqAgain}, it is easy to see that
Theorem \textit{storeAtomicity} can be proved. Since a store can happen only
when the state of a leaf cache is in \Mo{} state (Invariant
\textit{processReq}), because of Invariant \textit{uniqM} only one store
request can be processed at a time throughout the system, thereby satisfying
\textit{uniqStTimes}. Because of Invariant \textit{noReqAgain}, once a response
has been produced for a request $i$, the cache can only process requests $j, j
> i$. Thus there is at most one response to each request, satisfying
\textit{uniqRespLabels}. A load request can be processed only if a cache is at
least in \Sh{} state according to Invariant \textit{processReq}.  When a leaf
cache is at least in \Sh{} state, since it has no children, it is \clean{}.
Thus, by Invariant \textit{lastestValue}, the final condition for Theorem
\textit{storeAtomicity} is satisfied.

Note that Invariants \textit{processReq} and \textit{noReqAgain} can be proved
directly from the atomic transitions describing the system (Figure \ref{trans}
for MSI; other protocols must also have similar transitions). Proving Invariants
\textit{uniqM} and \textit{latestValue} requires the following invariants.

\begin{inv}
\textit{conservative}: At all times, the state of a cache is not greater than
its parent's directory corresponding to the cache.
\label{conservative}
\end{inv}

\begin{inv}
\textit{stateCompat}: At all times, the state of a cache is not less than its
directory corresponding to any of its children.
\label{stateCompat}
\end{inv}

We use the definition of compatibility between two caches as follows:
\begin{defn}
\textit{compatibile}: At some time, cache $c_1$ is compatible with $c_2$ iff
at that time,
\begin{enumerate}
\item $c_1.\state[a] = \In$ whenever $c_2.\state[a] = \Mo$ and
\item $c_1.\state[a] \le \Sh$ whenever $c_2.\state[a] = \Sh$
\end{enumerate}
\label{compatible}
\end{defn}

\begin{inv}
\textit{siblingCompat}: At all times, two sibling caches (\ie caches with the
same parent) are compatible.
\label{siblingCompat}
\end{inv}

\begin{inv}
\textit{noTransitWrite}: When a cache $c$ transitions from non-clean at time
$t$ to clean at time $t+1$ for an address $a$, then it receives data $a$ from
some cache $s$ such that the $s$ is clean at the time $t'$ for address $a$, and
no store request was processed for $a$ at time $t_i$ such that $t' \le t_i \le
t$.
\label{noTransitWrite}
\end{inv}

For any two caches $c$ and $p$, such that $c$ is a descendent of $p$ in the cache
hierarchy (the descendent relation between two caches is the transitive
reflexive closure of the parent relation), from Invariants
\textit{conservative} and \textit{stateCompat}, it is clear that for any
address $a$, $c.\state[a] \le p.\state[a]$. Any two caches $c_1, c_2$ in which
neither is the decendent of the other will have a lowest common ancestor $x$
such that both the caches are descendents of $x$ and $c_1, c_2$ are descendents
of distinct caches $x_1, x_2$, respectively, such that $x_1, x_2$ are children
of $x$. $x_1$ and $x_2$ are compatible because of Invariant
\textit{siblingCompat}. $c_1.\state[a] \le x_1.\state[a]$ and
$c_2.\state[a] \le x_2.\state[a]$ because of Invariants \textit{conservative}
and \textit{stateCompat}. From this, it can easily be seen from the definition
of compatible that $c_1$ and $c_2$ are compatible whenever $x_1$ and $x_2$ are
compatible. From the definition of compatible it can be seen that both caches can not
be in the \Mo{} state for them to be compatible. This directly proves Invariant
\textit{uniqM}.

Invariant \textit{latestValue} can be proved using induction on time. Let us
assume that all caches which are clean at time $t'$ such that $t'<t+1$ have the
latest value. If a cache $c$ is clean at time $t$, then none of its descendents
are in \Mo, and hence none of its descendents can process a store request at
$t$. Any non descendent is compatible with it (as shown in the previous
paragraph) and so no non-descendent cache can be in \Mo, since a clean cache is
at least in \Sh. So only $c$ can process a store request at $t$ (assuming $c$
is a leaf cache and is in state \Mo) when it is clean, and no other cache can
process a store request. By induction hypothesis, it contained the latest value
at time $t$ which remains true (because of no processing of store request) at
time $t+1$. Now, if a cache was not clean at time $t$, then it would have
received the data from some cache which was clean at the time of sending the
data (by Invariant \textit{noTransitWrite}), which means the data is the latest
value by induction hypothesis, so the new data it receives at time $t$ is the
latest value, making the clean cache at time $t+1$ have the latest value.

Any earlier clean cache has the latestValue, and data is
obtained from a clean cache

%Top
%a) conservative: forall p c a t, parent c p -> dir p c a t >= state c a t
%b) forall p c a t, parent c p -> state p a t >= dir p c a t
%c) forall c1 c2 a t, sibling c1 c2 -> compatible c1 c2 a t
%d) nextClean: ~ clean c a t -> clean c a (S t) -> exists s d ts, recv data s a t d /\ send data s a ts d /\ clean s a ts /\ ts <= t /\ forall ti, ts <= ti < t ->  ~ write a ti
%
%(from=state(t1)=dir(t2) -> betweenPSendCRecv:dir(t)>=to&state(t)<=from, betweenCSendPRecv:state(t)<=to&dir(t)>=from), noMesgImp:state(t)=dir(t) -> conservative (ignore from=, use only a, b)
%a) Child sends message iff it downgrades
%b) Parent sends message iff it upgrades
%c) State is changed only on sending or receiving a message
%d) FIFO messages
%e) cross: Message from child and parent do not cross
%
%cross:
%a) parent send message only on receiving request
%b) req from child does not overtake response from child (prove using the from states)
%c) resp from parent does not overtake request from parent
%d) voluntary response sent from child only when not waiting
%
%In this section we specify a set of important invariants for a parent-child pair of caches.

### 2014-02-03 paper/Intuition.tex
stuff

@@
 
-Any earlier clean cache has the latestValue, and data is
-obtained from a clean cache
+Invariant \ref{conservative} can be proved by
+further breaking them into the following invariants between a parent-child pair
+of caches.
+
+\begin{inv}
+\textit{stateChange}:
+\begin{enumerate}
+\item A child sends a response to its parent iff it downgrades its state.
+\item A parent sends a response to a child iff it upgrades the directory of
+that child.
+\item The state and the directory can change only on sending and receiving a
+response.
+\end{enumerate}
+\label{stateChange}
+\end{inv}
+
+Invariant \textit{stateChange} can be proved directly from the atomic
+transitions for MSI. Other protocols have similar atomic transitions obeying
+these invariants.
+
+\begin{inv}
+\textit{inTransitChange}:
+\begin{enumerate}
+\item When a response is sent from a child $c$ at time $t$ for an address $a$,
+and has not been received by the parent $p$ before time $t'$, $\forall t_i, t <
+t_i \le t' @-> c.\state[a]$ at $t_i \le c.\state[a]$ at $t+1$ and $\forall t_i,
+t \le t_i \le t' @->$ $p.\dstate[c][a]$ at $t_i \ge c.\state[a]$ at $t$.
+\item When a response is sent from parent $p$ at time $t$ for an address $a$,
+and has not been received by its child $c$ before time $t'$, $\forall t_i, t <
+t_i \le t' @-> p.\dstate[c][a]$ at $t_i \ge p.\dstate[c][a]$ at $t+1$ and
+$\forall t_i, t \le t_i \le t' @->$ $c.\state[a]$ at $t_i \le p.\dstate[c][a]$ at
+$t$.
+\end{enumerate}
+\label{inTransitChange}
+\end{inv}
+
+\begin{inv}
+\textit{noTransitEqual}: When no response is in transit for an address, then
+the state of a cache for that address is the same as its parent's directory for
+that cache and address.
+\label{noTransitEqual}
+\end{inv}
+
+Using Invariants \ref{stateChange} to \ref{noTransit}, it is easy to prove
+Invariant \ref{conservative}. If no response is in flight for an address, then
+the parent's directory state of the cache for the address matches that of the
+cache's state for that address. If there is a response from parent to child in
+flight, then the corresponding directory at the parent will be no less than the
+directory immediately after the response was sent and the state of the child
+will be no greater than the directory immediately before the response was sent
+(by Invariant \textit{inTransitChange}). By Invariant \textit{stateChange}, the
+parent upgrades the directory when sending the response, thus conforming to
+Invariant \textit{conservative}. Similarly, it can be shown that if there is a
+response from child to parent, Invariant \textit{coservative} will hold.
+
+\begin{inv}
+\textit{prevState}: When a response is received from a child to the parent
+(parent to child) for an address, the state of the child for that address at
+the time of sending (receiving) the response is the same as the parent's
+directory for that child and address at the time of receiving (sending) the
+response.
+\label{prevState}
+\end{inv}
 

### 2014-02-03 paper/Intuition.tex
stuff

@@
 \begin{inv}
-\textit{inTransitChange}:
+\textit{inTransitCondition}:
 \begin{enumerate}
@@
 \end{enumerate}
-\label{inTransitChange}
+\label{inTransitCondition}
 \end{inv}
@@
 
-Using Invariants \ref{stateChange} to \ref{noTransit}, it is easy to prove
-Invariant \ref{conservative}. If no response is in flight for an address, then
+Using Invariants \ref{stateChange} to \ref{noTransitEqual}, it is easy to prove
+Invariant \textit{conservative}. If no response is in flight for an address, then
 the parent's directory state of the cache for the address matches that of the
@@
 will be no greater than the directory immediately before the response was sent
-(by Invariant \textit{inTransitChange}). By Invariant \textit{stateChange}, the
+(by Invariant \textit{inTransitCondition}). By Invariant \textit{stateChange}, the
 parent upgrades the directory when sending the response, thus conforming to
 Invariant \textit{conservative}. Similarly, it can be shown that if there is a
-response from child to parent, Invariant \textit{coservative} will hold.
+response from child to parent, Invariant \textit{conservative} will hold.
+
+Proving Invariants \textit{inTransitCondition} and \textit{noTransitEqual},
+needs two further invariants, along with Invariant \textit{stateChange}.
+
+\begin{inv}
+\textit{respFIFO}: Responses are received in the order in which they are sent,
+for a source-destination pair.
+\label{respFIFO}
+\end{inv}
+
+\begin{inv}
+\textit{noCross}: A response from a cache to its child for an address can not
+be in transit while another response from the same child to the parent for the
+same address is in transit.
+\label{noCross}
+\end{inv}
+
+Invariant \textit{inTransitCondition} can be proved as follows. Whenever a
+response from a child to the parent is in transit, the parent can not send any
+response to the child, or it will violate Invariant \textit{noCross}. Using the fact that
 

### 2014-02-03 paper/System.tex
System Update for grammar/spelling

@@
 permissions ($\In < \Sh < \Mo$). For leaf caches, the meaning of these
-permission states is straightfoward, a load request can be processed only if the
+permission states is straightforward, a load request can be processed only if the
 leaf is at least in the \Sh{} state, and a store request can be processed only
@@
 $p$, the state in its directory for child cache $c$ is denoted by
-$p.\dstate[c][a]$, for address $a$. This local version may
-be stale, in that the child may have changed it state but the parent
-has not yet updated itself. However, the parent state will always be a
-conservative overapproximation of the child state.
+$p.\dstate[c][a]$, for address $a$. This local version may be stale,
+in that the child may have changed it state but the parent has not yet
+updated itself. However, the parent state will always be a
+conservative over-approximation of the child state.
 
@@
 system well defined. Figure~\ref{trans} gives the atomic transitions for a
-cache coherent memory subsystem. The actions in the transitions repesent only
+cache coherent memory subsystem. The actions in the transitions represent only
 the specific system states that change. We use $x \Leftarrow y$ to indicate
@@
 Finally, the child cache receives the response from the parent (ChildRecvResp)
-and sets its state while reseting its wait state if the desired upgrade is
+and sets its state while resetting its wait state if the desired upgrade is
 indicated by the response.
@@
 the parent receives the response from the cache (ParentRecvResp) and sets the
-appropriate directory state while reseting the appropriate directory wait state
+appropriate directory state while resetting the appropriate directory wait state
 if the desired downgrade is indicated by the response.
@@
 in which each transition can span several clock cycles.
-
-%%%%%%%%%%%%%%%%%%%%%%%
-
-%In addition, the leaf caches (called L1 caches usually)
-%communicate with the processor through two channels: a) \cproc{} \textbf{FIFO}
-%channel to transmit requests from the processor to L1 cache, and b) \procc{}
-%\textbf{FIFO} channel to transmit responses from the L1 cache to the processor.
-%There is exactly one of \cproc{} and \procc{} channels between the L1 and the processor.

### 2014-02-03 paper/Intuition.tex
some modifications of background

@@
 \item The state and the directory can change only on sending and receiving a
-response.
+response; the receiving entity changes its value to that of the sending entity
+immediately after sending the response message.
+\item 
 \end{enumerate}
@@
 
+\begin{inv}
+\textit{respFIFO}: Responses are received in the order in which they are sent,
+for a source-destination pair, for the same address.
+\label{respFIFO}
+\end{inv}
+
+\begin{inv}
+\textit{noCross}: A response from a cache to its child for an address can not
+be in transit while another response from the same child to the parent for the
+same address is in transit.
+\label{noCross}
+\end{inv}
+
+Invariant \textit{conservative} at any time $t$ for child cache $c$ and parent
+$p$ for an address $a$ can be proved as follows. Consider the case when
+$c.\state[a]$ never changed before $t$, then $c.\state[a] = \In$ because of the
+initialization conditions described in Section \ref{System}. No response would
+have been sent by $c$ to its parent for address $a$ before $t$, otherwise
+$c.\state[a]$ would have changed. Thus the parent would not have received any
+response from $c$, $p.\dstate[c][a]$ could have changed only by sending a
+response to $c$ which can only upgrade it from the initial value of $\In$. Thus
+at $t$, $p.\dstate[c][a] \ge c.state[a]$.
+
+Now consider the case when $c.\state[a]$ last changed at $t'$ such that $t' <
+t$ because of receiving a response from $p$, sent at $t''$. $c.\state[a]$ at
+$t'+1 = p.\dstate[c][a]$ at $t''+1$ because of Invariant \textit{stateChange}.
+$c$ neiter sends nor receives a response at any $t_i$ such that $t' < t_i < t$.
+If $p$ received a response at some $t_m$ such that $t'' < t_m < t$,
+then it would either violate Invariant \textit{noCross}, or the fact that no
+message was sent between $t' < t_i < t$ from $c$. Thus, $p.\dstate[c][a]$ at $t
+\ge p.\dstate[c][a]$ at $t''+1$ because any message sent can only upgrade
+$p.\dstate[a]$. This ensures that Invariant \textit{conservative} is obeyed. We
+can similarly prove this invariant holds when $c.\state[a]$ last changes before
+$t$ by sending a response.
+
 Invariant \textit{stateChange} can be proved directly from the atomic
@@
 these invariants.
+
+Invariant \textit{respFIFO} can be enforced by the interconnect network which
+ensures a FIFO ordering for messages sent between a pair of nodes.
+
+%Invariant \textit{respFIFO} is known as point-to-point FIFO ordering in the
+%Network-on-Chip NoC community, and is enforced by common routing algorithms
+%like Dimension-ordered routing \cite{dimension}. Only the parent-to-child
+%channels need to be FIFO ordered, the FIFO ordering in child-to-parent channels
+%can be enforced using the protocol itself.
+
+---------------------------------------------
+
+\begin{inv}
+\textit{noTransitEqual}: When no response is in transit for an address, then
+the state of a cache for that address is the same as its parent's directory for
+that cache and address.
+\label{noTransitEqual}
+\end{inv}
+
+Proving Invariants \textit{inTransitCondition} and \textit{noTransitEqual},
+needs two further invariants, along with Invariant \textit{stateChange}.
+
+Invariant \textit{inTransitCondition} can be proved as follows. Whenever a
+response from a child to the parent is in transit, the parent can not send any
+response to the child, or it will violate Invariant \textit{noCross}. Using the fact that
 
@@
 \end{inv}
-
-\begin{inv}
-\textit{noTransitEqual}: When no response is in transit for an address, then
-the state of a cache for that address is the same as its parent's directory for
-that cache and address.
-\label{noTransitEqual}
-\end{inv}
-
-Using Invariants \ref{stateChange} to \ref{noTransitEqual}, it is easy to prove
-Invariant \textit{conservative}. If no response is in flight for an address, then
-the parent's directory state of the cache for the address matches that of the
-cache's state for that address. If there is a response from parent to child in
-flight, then the corresponding directory at the parent will be no less than the
-directory immediately after the response was sent and the state of the child
-will be no greater than the directory immediately before the response was sent
-(by Invariant \textit{inTransitCondition}). By Invariant \textit{stateChange}, the
-parent upgrades the directory when sending the response, thus conforming to
-Invariant \textit{conservative}. Similarly, it can be shown that if there is a
-response from child to parent, Invariant \textit{conservative} will hold.
-
-Proving Invariants \textit{inTransitCondition} and \textit{noTransitEqual},
-needs two further invariants, along with Invariant \textit{stateChange}.
-
-\begin{inv}
-\textit{respFIFO}: Responses are received in the order in which they are sent,
-for a source-destination pair.
-\label{respFIFO}
-\end{inv}
-
-\begin{inv}
-\textit{noCross}: A response from a cache to its child for an address can not
-be in transit while another response from the same child to the parent for the
-same address is in transit.
-\label{noCross}
-\end{inv}
-
-Invariant \textit{inTransitCondition} can be proved as follows. Whenever a
-response from a child to the parent is in transit, the parent can not send any
-response to the child, or it will violate Invariant \textit{noCross}. Using the fact that
 

### 2014-02-03 paper/Introduction.tex
Intuition

@@
 Systems (TRS). It was shown by Dave
-\etal{}~\cite{DNA:CoherenceImplementation} that a verified protocol
-could be transliterated into Bluespec, a hardware description language
-based on Guarded Atomic Actions, a restriction of
+\etal{}~\cite{DNA:CoherenceImplementation} that such a verified
+protocol could be transliterated into Bluespec, a hardware description
+language based on Guarded Atomic Actions, a restriction of
 TRS~\cite{Hoe:TCAD}. This reduces the correctness of the
@@
 
-For this approach to be truly viable, we need not a single proof of
-correctness of a single design, but a proof framework covering any
-variation of a design that might be of interest for performance
-reasons. This would enable implementation level changes without
-compromising on the ability to verify correctness of the design.
+For this approach to be truly viable, designers cannot be handed a
+single proof of correctnessof a single design; rather, they need a
+proof framework covering any variation of a design that might be of
+interest for performance reasons. This would enable implementation
+level changes without compromising on the ability to verify
+correctness of the design.
 

### 2014-02-03 paper/Intuition.tex
some modifications of background
Intuition

@@
 \section{Intuition}
+\label{Sec:ProofStructure}
 
-In this section we give the proof of the MSI cache coherence protocol. We give
-the proof in a top-down fashion starting with the store atomicity theorem. We
-specify important invariants and show how they are proved in the MSI protocol.
-In Section \ref{Refinements} show how these invariants hold in other protocols
-like MESI and MOSI. These invariants are true at each step in the (infinite)
-list of transitions starting from the initial state. We use the term time $t$
-to specify the global state reached after $t$ transitions.  All the invariants
-below pertain to a particular address; we will implicitly assume a universal
-quantification over all address as a prefix to each of the invariants (if not
-explicitly mentioned).
+In this section, we structure our proof framework for our cache
+coherence protocol. Specifically, we formally reduce the proof of the
+whole system store atomicity theorem to the proof of a set of easily
+proven invariants based on the structure of the protocol and design
+structure which capture the ``intuitive'' proof of correctness of our
+protocol design. As we show, these invariants are easily shown in the
+base protocol design; further, in Section~\ref{Sec:Refinements}, we
+show how the verification of a design refinement requires only that we
+prove these invariants for the new system which is relatively
+straightforward.
 
-\subsection{Top invariants}
+Our invariants hold before and after any execution of the component
+transactions of our design reachable from the initial state. We use
+the term time $t$ to specify the global state reached after $t$
+transitions.  All the invariants below pertain to a particular
+address; we will implicitly assume a universal quantification over all
+address as a prefix to each of the invariants (if not explicitly
+mentioned).
+
+While we have proofs in Coq of each step in this framework our
+discussion will be mainly informal for brevity. A full description can
+be found at~\textbf{MURALI PUT REFERENCE HERE}.
+
+\subsection{Reducing correctness to Node-level Invariants}
+
+We've defined correctness is a whole system property. The first step
+in our framework is to reduce the store atomicity property down to the
+following 4 global invariants about nodes:
+
 \begin{defn}
-\textit{clean}: A cache is said to be \clean{} for an address iff the state of the
-cache is at least \Sh{} and the state of all its children is at most \Sh.
+\textit{clean}: A cache is said to be \clean{} for an address $a$ iff
+the state of the cache for that address is at least \Sh{} and the
+state of all its children is at most \Sh.
 \end{defn}
@@
 At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
-up-to-date value for that address, \ie
+up-to-date value for that address, \ie{}
 \begin{enumerate}
-\item $c.\data[a] = \initData(a)$ and no store request has been processed at
-any time $< t$, or
+
+\item $c.\data[a] = \initData(a)$ and no store request has been
+  processed at any time $< t$, or
+
 \item There is a store request $q$ that was processed at time $t'$ such that
@@
 
-Using Invariants \ref{latestValue} to \ref{noReqAgain}, it is easy to see that
-Theorem \textit{storeAtomicity} can be proved. Since a store can happen only
-when the state of a leaf cache is in \Mo{} state (Invariant
-\textit{processReq}), because of Invariant \textit{uniqM} only one store
-request can be processed at a time throughout the system, thereby satisfying
-\textit{uniqStTimes}. Because of Invariant \textit{noReqAgain}, once a response
-has been produced for a request $i$, the cache can only process requests $j, j
-> i$. Thus there is at most one response to each request, satisfying
-\textit{uniqRespLabels}. A load request can be processed only if a cache is at
-least in \Sh{} state according to Invariant \textit{processReq}.  When a leaf
-cache is at least in \Sh{} state, since it has no children, it is \clean{}.
-Thus, by Invariant \textit{lastestValue}, the final condition for Theorem
-\textit{storeAtomicity} is satisfied.
+We can see that the correctness of our system
+(Theorem~\textit{storeAtomicity}) follows naturally from Invariants
+\ref{latestValue} to \ref{noReqAgain}. Since a store may happen only
+when leaf cache is in \Mo{} state (Invariant~\textit{processReq}), and
+only one cache may cause a store request per address at a particular
+point in time (Invariant \textit{uniqM}), Theorem~\textit{uniqStTimes}
+holds. Given Invariant~\textit{noReqAgain}, once a response has been
+produced for a request $i$, the cache may only process later requests
+$j, j > i$. Thus there is at most one response to each request,
+satisfying Theorem~\textit{uniqRespLabels}. A load request can be
+processed only if a cache is at least in \Sh{} state according to
+Invariant \textit{processReq}.  When a leaf cache is at least in \Sh{}
+state, since it has no children, it is \clean{}.  Thus, by
+Invariant~\textit{lastestValue}, the final condition for
+Theorem~\textit{storeAtomicity} is satisfied.
 
-Note that Invariants \textit{processReq} and \textit{noReqAgain} can be proved
-directly from the atomic transitions describing the system (Figure \ref{trans}
-for MSI; other protocols must also have similar transitions). Proving Invariants
-\textit{uniqM} and \textit{latestValue} requires the following invariants.
+\subsection{Reducing to Pair-wise Invariants}
+
+Invariants~\textit{processReq} and \textit{noReqAgain} can be proved
+directly from the atomic transitions describing the system
+(Figure~\ref{trans} for MSI; other protocols must also have similar
+transitions). Proving invariants~\textit{uniqM} and
+\textit{latestValue} needs some knowledge of the hierarchy. We can
+prove both using the following invariants on parent-child
+relationships.
 
 \begin{inv}
-\textit{conservative}: At all times, the state of a cache is not greater than
-its parent's directory corresponding to the cache.
+\textit{conservative}: At all times, the state of an address in a
+cache is not greater than its that in the corresponding entry in the
+parent's directory.
 \label{conservative}
@@
 For any two caches $c$ and $p$, such that $c$ is a descendent of $p$ in the cache
-hierarchy (the descendent relation between two caches is the transitive
+hierarchy (the descendant relation between two caches is the transitive
 reflexive closure of the parent relation), from Invariants
@@
 address $a$, $c.\state[a] \le p.\state[a]$. Any two caches $c_1, c_2$ in which
-neither is the decendent of the other will have a lowest common ancestor $x$
-such that both the caches are descendents of $x$ and $c_1, c_2$ are descendents
+neither is the decendant of the other will have a lowest common ancestor $x$
+such that both the caches are descendants of $x$ and $c_1, c_2$ are descendants
 of distinct caches $x_1, x_2$, respectively, such that $x_1, x_2$ are children
@@
 
-Invariant \textit{latestValue} can be proved using induction on time. Let us
+Invariant~\textit{latestValue} can be proved using induction on time. Let us
 assume that all caches which are clean at time $t'$ such that $t'<t+1$ have the
@@
 at time $t$ which remains true (because of no processing of store request) at
-time $t+1$. Now, if a cache was not clean at time $t$, then it would have
+time $t+1$. If a cache was not clean at time $t$, then it would have
 received the data from some cache which was clean at the time of sending the
-data (by Invariant \textit{noTransitWrite}), which means the data is the latest
+data (by Invariant~\textit{noTransitWrite}), which means the data is the latest
 value by induction hypothesis, so the new data it receives at time $t$ is the
@@
 
-Invariant \ref{conservative} can be proved by
+Invariant~\textit{conservative} can be proved by
 further breaking them into the following invariants between a parent-child pair
@@
 immediately after sending the response message.
-\item 
 \end{enumerate}
@@
 
-Invariant \textit{conservative} at any time $t$ for child cache $c$ and parent
+Invariant~\textit{conservative} at any time $t$ for child cache $c$ and parent
 $p$ for an address $a$ can be proved as follows. Consider the case when
@@
 
+\subsection{Final reduction}
+
+---------------------------------------------
+
 %Invariant \textit{respFIFO} is known as point-to-point FIFO ordering in the
@@
 %can be enforced using the protocol itself.
-
----------------------------------------------
 

### 2014-02-03 paper/System.tex
Intuition

@@
 
-The behavior of the system are described via atomic state transitions.
-These transitions have two parts: the guard dictating when the
-transition can fire, and the action defining how the state of the
-system changes when the transition happens. These transitions or \emph{guarded
-atomic actions} logically happen one by one, making the overall behavior of the
-system well defined. Figure~\ref{trans} gives the atomic transitions for a
-cache coherent memory subsystem. The actions in the transitions represent only
-the specific system states that change. We use $x \Leftarrow y$ to indicate
-setting of a specific state $x$ of the overall system to value $y$ because of a
-transition.
+The behavior of the system are described via the atomic state
+transitions.  These transitions have two parts: the guard dictating
+when the transition can fire, and the action defining how the state of
+the system changes when the transition happens. In our generated
+implementation, these transitions or \emph{guarded atomic actions}
+happen concurrently in a single hardware clock, but they will
+logically happen happen one at a time. Figure~\ref{trans} gives the
+atomic transitions for a cache coherent memory subsystem. The actions
+in the transitions represent only the specific system states that
+change. We use $x \Leftarrow y$ to indicate setting of a specific
+state $x$ of the overall system to value $y$ because of a transition.
 

### 2014-02-04 paper/Intuition.tex
intuition

--- 2014-02-03 paper/Intuition.tex
@@
 \section{Intuition}
+\label{Sec:ProofStructure}
 
-In this section we give the proof of the MSI cache coherence protocol. We give
-the proof in a top-down fashion starting with the store atomicity theorem. We
-specify important invariants and show how they are proved in the MSI protocol.
-In Section \ref{Refinements} show how these invariants hold in other protocols
-like MESI and MOSI. These invariants are true at each step in the (infinite)
-list of transitions starting from the initial state. We use the term time $t$
-to specify the global state reached after $t$ transitions.  All the invariants
-below pertain to a particular address; we will implicitly assume a universal
-quantification over all address as a prefix to each of the invariants (if not
-explicitly mentioned).
+In this section, we present the proof structure for our MSI-based cache
+coherence protocol of Figure \ref{trans}. Specifically, we formally reduce the
+proof of the store atomicity theorem for the whole system to the proof of a set
+of invariants. In order to prove these invariants we introduce the main
+``intuition'' behind the correctness of the protocol. This intuition is not
+restricted to the MSI-based protocol; Section \ref{Sec:Refinements} shows that
+it applies to other protocols like MOSI.
 
-\subsection{Top invariants}
+Our invariants hold after any series of executions of the component
+transactions of our design reachable from the initial state. We use
+the term time $t$ to specify the global state reached after firing $t$
+transitions.
+
+While we have proofs in Coq of each step in this framework our
+discussion will be mainly informal for brevity. A full description can
+be found at~\xxx{MURALI PUT REFERENCE HERE}.
+
+\subsection{Reducing correctness to Node-level Invariants}
+
+We've defined store atomicity to be a whole system property. The first step in
+our framework is to reduce the store atomicity property down to the following 4
+global invariants about nodes:
+
 \begin{defn}
-\textit{clean}: A cache is said to be \clean{} for an address iff the state of the
-cache is at least \Sh{} and the state of all its children is at most \Sh.
+\textit{clean}: A cache is said to be \clean{} for an address $a$ iff
+the state of the cache for that address is at least \Sh{} and the
+state of all its children is at most \Sh.
 \end{defn}
@@
 At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
-up-to-date value for that address, \ie
+up-to-date value for that address, \ie{}
 \begin{enumerate}
-\item $c.\data[a] = \initData(a)$ and no store request has been processed at
-any time $< t$, or
+\item $c.\data[a] = \initData(a)$ and no store request has been
+  processed at any time $t_i$ such that $t_i < t$, or
 \item There is a store request $q$ that was processed at time $t'$ such that
@@
 
-The following invariants hold for leaf caches.
+We use the definition of compatibility between two caches as follows:
+\begin{defn}
+\textit{compatibile}: Values of states $x_1$ and $x_2$ are compatible iff
+either $x_1$ or $x_2$ is \Mo{}, then the other is \In.
+\label{compatible}
+\end{defn}
+
 \begin{inv}
-\textit{uniqM}: At all times, at most one leaf cache can be in \Mo{} state.
-\label{uniqM}
+\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
+that neither is an ancestor of the other in the cache tree hierarchy, for each
+address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are compatible.
+\label{nonAncestorCompatible}
 \end{inv}
@@
 \begin{inv}
-\textit{processReq}: At all times, a load or a store request at a leaf cache can be processed
-only if it is at least in \Sh{} or \Mo{} state at that time, respectively.
+\textit{processReq}: A load (store) request for an address at a leaf cache can
+be processed only if it is at least in the \Sh{} (\Mo) state for that address.
 \label{processReq}
@@
 \begin{inv}
-\textit{noReqAgain}: A leaf cache $c$ processes the $c.\pos^{th}$ request iff
-$c.\pos$ changes. $c.\pos$ only gets incremented when it changes.
+\textit{noReqAgain}:
+\item For a leaf cache $c$, $c.\pos$ changes only on processing a request in
+which case $c.\pos$ is incremented.
 \label{noReqAgain}
@@
 
-Using Invariants \ref{latestValue} to \ref{noReqAgain}, it is easy to see that
-Theorem \textit{storeAtomicity} can be proved. Since a store can happen only
-when the state of a leaf cache is in \Mo{} state (Invariant
-\textit{processReq}), because of Invariant \textit{uniqM} only one store
-request can be processed at a time throughout the system, thereby satisfying
-\textit{uniqStTimes}. Because of Invariant \textit{noReqAgain}, once a response
-has been produced for a request $i$, the cache can only process requests $j, j
-> i$. Thus there is at most one response to each request, satisfying
-\textit{uniqRespLabels}. A load request can be processed only if a cache is at
-least in \Sh{} state according to Invariant \textit{processReq}.  When a leaf
-cache is at least in \Sh{} state, since it has no children, it is \clean{}.
-Thus, by Invariant \textit{lastestValue}, the final condition for Theorem
-\textit{storeAtomicity} is satisfied.
+We can see that the correctness of our system (Theorem~\textit{storeAtomicity})
+follows naturally from Invariants \ref{latestValue} to \ref{noReqAgain}.  A
+store for an address may happen only when leaf cache is in \Mo{} state for that
+address (Invariant~\textit{processReq}). Any two distinct leaf caches are not
+ancestors of each other, so by Invariant \textit{nonAncestorCompatible}, if a
+leaf cache is in \Mo{} state, all other leaf caches will be in \In{} state
+which prevents processing of both load and store requests for the same address
+in other leaf caches (again by Invariant \textit{processReq}). Thus, Definition
+\textit{uniqStTimes} holds.  Given Invariant~\textit{noReqAgain}, once a
+response has been produced for a request $i$, the cache may subsequently only
+process $j^{th}$ request such that $j > i$. Thus there is at most one response
+to each request, satisfying Definition~\textit{uniqRespLabels}. A load request
+can be processed only if a cache is at least in \Sh{} state (again by Invariant
+\textit{processReq}). When a leaf cache is at least in \Sh{} state, since it
+has no children, it is \clean. A load response supplies the data for that
+address present in the cache, which, by Invariant~\textit{lastestValue},
+contains the ``latest value''. Thus, the final condition for
+Theorem~\textit{storeAtomicity} is also satisfied.
 
-Note that Invariants \textit{processReq} and \textit{noReqAgain} can be proved
-directly from the atomic transitions describing the system (Figure \ref{trans}
-for MSI; other protocols must also have similar transitions). Proving Invariants
-\textit{uniqM} and \textit{latestValue} requires the following invariants.
+\subsection{Reducing to Pair-wise Invariants}
+\label{pairwise}
+
+Invariants~\textit{processReq} and \textit{noReqAgain} can be proved directly
+from the atomic transitions describing the system (Figure~\ref{trans} for MSI;
+other protocols must also have similar transitions). Proving
+invariants~\textit{uniqM} and \textit{latestValue} needs further work. We can
+prove both using the following invariants on parent-child relationships.
 
 \begin{inv}
-\textit{conservative}: At all times, the state of a cache is not greater than
-its parent's directory corresponding to the cache.
+\textit{conservative}: The state of an address in a cache is never greater than
+the corresponding entry in the parent's directory.
 \label{conservative}
@@
 \begin{inv}
-\textit{stateCompat}: At all times, the state of a cache is not less than its
-directory corresponding to any of its children.
-\label{stateCompat}
-\end{inv}
-
-We use the definition of compatibility between two caches as follows:
-\begin{defn}
-\textit{compatibile}: At some time, cache $c_1$ is compatible with $c_2$ iff
-at that time,
+\textit{localCompatibility}:
 \begin{enumerate}
-\item $c_1.\state[a] = \In$ whenever $c_2.\state[a] = \Mo$ and
-\item $c_1.\state[a] \le \Sh$ whenever $c_2.\state[a] = \Sh$
+\item The state of a cache is never less than its directory entry for any of
+its children.
+\item Let $c_1$ and $c_2$ be distinct caches with the same parent $p$. Then,
+$p.\dstate[c_1][a]$ and $p.\dstate[c_2][a]$ are compatible.
 \end{enumerate}
-\label{compatible}
-\end{defn}
-
-\begin{inv}
-\textit{siblingCompat}: At all times, two sibling caches (\ie caches with the
-same parent) are compatible.
-\label{siblingCompat}
+\label{localCompatibility}
 \end{inv}
@@
 \begin{inv}
-\textit{noTransitWrite}: When a cache $c$ transitions from non-clean at time
-$t$ to clean at time $t+1$ for an address $a$, then it receives data $a$ from
-some cache $s$ such that the $s$ is clean at the time $t'$ for address $a$, and
-no store request was processed for $a$ at time $t_i$ such that $t' \le t_i \le
-t$.
+\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
+$\forall t, t_s \le t \le t_r$ where $t_s$ is the time of sending the data and
+$t_r$ the time of receiving the data), no cache can process a store request for
+$a$.
 \label{noTransitWrite}
@@
 
-For any two caches $c$ and $p$, such that $c$ is a descendent of $p$ in the cache
-hierarchy (the descendent relation between two caches is the transitive
-reflexive closure of the parent relation), from Invariants
-\textit{conservative} and \textit{stateCompat}, it is clear that for any
-address $a$, $c.\state[a] \le p.\state[a]$. Any two caches $c_1, c_2$ in which
-neither is the decendent of the other will have a lowest common ancestor $x$
-such that both the caches are descendents of $x$ and $c_1, c_2$ are descendents
-of distinct caches $x_1, x_2$, respectively, such that $x_1, x_2$ are children
-of $x$. $x_1$ and $x_2$ are compatible because of Invariant
-\textit{siblingCompat}. $c_1.\state[a] \le x_1.\state[a]$ and
-$c_2.\state[a] \le x_2.\state[a]$ because of Invariants \textit{conservative}
-and \textit{stateCompat}. From this, it can easily be seen from the definition
-of compatible that $c_1$ and $c_2$ are compatible whenever $x_1$ and $x_2$ are
-compatible. From the definition of compatible it can be seen that both caches can not
-be in the \Mo{} state for them to be compatible. This directly proves Invariant
-\textit{uniqM}.
+For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in the
+cache tree hierarchy, from Invariants \textit{conservative} and the first part
+of \textit{localCompatibility}, we can show that for any address address $a$,
+$c.\state[a] \le p.\state[c][a]$. This is because the ancestor relation is the
+reflective-transitive closure of the parent relation, and Invariants
+\textit{conservative} and the first part \textit{localCompatibility} of defines
+a $\ge$ relation between the states of a parent and its child.
 
-Invariant \textit{latestValue} can be proved using induction on time. Let us
-assume that all caches which are clean at time $t'$ such that $t'<t+1$ have the
-latest value. If a cache $c$ is clean at time $t$, then none of its descendents
-are in \Mo, and hence none of its descendents can process a store request at
-$t$. Any non descendent is compatible with it (as shown in the previous
-paragraph) and so no non-descendent cache can be in \Mo, since a clean cache is
-at least in \Sh. So only $c$ can process a store request at $t$ (assuming $c$
-is a leaf cache and is in state \Mo) when it is clean, and no other cache can
-process a store request. By induction hypothesis, it contained the latest value
-at time $t$ which remains true (because of no processing of store request) at
-time $t+1$. Now, if a cache was not clean at time $t$, then it would have
-received the data from some cache which was clean at the time of sending the
-data (by Invariant \textit{noTransitWrite}), which means the data is the latest
-value by induction hypothesis, so the new data it receives at time $t$ is the
-latest value, making the clean cache at time $t+1$ have the latest value.
+Any two caches $c_1, c_2$ in which neither is an ancestor of the other will
+have a lowest common ancestor $x$ such that $x$ is an ancestor of both the
+caches and $x$ has two distinct children $x_1$ and $x_2$ such that they are the
+ancestors of $c_1$ and $c_2$ respectively. For any address $a$,
+$x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are compatible by the second part
+of Invariant \ref{localCompatibility}. From this, it straightaway follows that
+$x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from Invariant
+\ref{conservative}). From the previous paragraph, it follows that $c_1$ and
+$c_2$ are also compatible. This proves Invariant \textit{nonAncestorCompatibility}.
 
-Invariant \ref{conservative} can be proved by
-further breaking them into the following invariants between a parent-child pair
-of caches.
+We can prove that Invariant \textit{latestValue} is obeyed by strong induction
+on time. Consider two cases: cache $c$ is already \clean{} for an address $a$
+at time $t$ or the contrary.  If \clean{}, then it has the latest value at time
+$t$. Since none of its children's directory entries can be in \Mo{}, all its
+descendents should be no greater than \Sh{} for that address (as discussed
+above). Any non-ancestor of $c$, and non-descendent of $c$ must be compatible
+(as discussed above). Since $c$ is \clean, it is at least in the \Sh{} state,
+so compatibility means that non-descendents can not be in \Mo{} state. All of
+$c$'s ancestors are non leaves, so they can not process any store request. So,
+no cache could have processed a store request other than $c$ itself, which
+means that $c$ contains the latest value for address $a$ at time $t+1$ if it
+was \clean{} at time $t$.
 
-\begin{inv}
-\textit{stateChange}:
-\begin{enumerate}
-\item A child sends a response to its parent iff it downgrades its state.
-\item A parent sends a response to a child iff it upgrades the directory of
-that child.
-\item The state and the directory can change only on sending and receiving a
-response.
-\end{enumerate}
-\label{stateChange}
-\end{inv}
+If $c$ was not clean at time $t$, it must have received data from either its
+parent or its child.  As can be seen in Figure \ref{trans}, whenever data is
+sent either from a parent to a child, or from a child to the parent, the sender
+is \clean. By the induction hypothesis, it means that the sender has the latest
+value for the address. Because of Invariant \ref{noTransitWrite}, no store
+request can be processed when the data is in transit, which means $c$ will have
+the latest value for address $a$ at time $t+1$.
 
-Invariant \textit{stateChange} can be proved directly from the atomic
-transitions for MSI. Other protocols have similar atomic transitions obeying
-these invariants.
+\subsection{Main intuition to prove the pair-wise properties}
+We present two basic invariants that the MSI protocol of Figure \ref{trans}
+obeys. In fact the protocol (\ie the request/response message semantics) was
+designed to ensure the following two invariants. We believe that most
+cache-coherence protocols would end up obeying these two invariants; we show
+how MOSI, for instance, obeys these invariants in Section
+\ref{sec:Refinements}. These two invariants together can prove all the pair-wise
+properties of Section \ref{pairwise}.
 
-\begin{inv}
-\textit{inTransitCondition}:
-\begin{enumerate}
-\item When a response is sent from a child $c$ at time $t$ for an address $a$,
-and has not been received by the parent $p$ before time $t'$, $\forall t_i, t <
-t_i \le t' @-> c.\state[a]$ at $t_i \le c.\state[a]$ at $t+1$ and $\forall t_i,
-t \le t_i \le t' @->$ $p.\dstate[c][a]$ at $t_i \ge c.\state[a]$ at $t$.
-\item When a response is sent from parent $p$ at time $t$ for an address $a$,
-and has not been received by its child $c$ before time $t'$, $\forall t_i, t <
-t_i \le t' @-> p.\dstate[c][a]$ at $t_i \ge p.\dstate[c][a]$ at $t+1$ and
-$\forall t_i, t \le t_i \le t' @->$ $c.\state[a]$ at $t_i \le p.\dstate[c][a]$ at
-$t$.
-\end{enumerate}
-\label{inTransitCondition}
-\end{inv}
-
-\begin{inv}
-\textit{noTransitEqual}: When no response is in transit for an address, then
-the state of a cache for that address is the same as its parent's directory for
-that cache and address.
-\label{noTransitEqual}
-\end{inv}
-
-Using Invariants \ref{stateChange} to \ref{noTransitEqual}, it is easy to prove
-Invariant \textit{conservative}. If no response is in flight for an address, then
-the parent's directory state of the cache for the address matches that of the
-cache's state for that address. If there is a response from parent to child in
-flight, then the corresponding directory at the parent will be no less than the
-directory immediately after the response was sent and the state of the child
-will be no greater than the directory immediately before the response was sent
-(by Invariant \textit{inTransitCondition}). By Invariant \textit{stateChange}, the
-parent upgrades the directory when sending the response, thus conforming to
-Invariant \textit{conservative}. Similarly, it can be shown that if there is a
-response from child to parent, Invariant \textit{conservative} will hold.
-
-Proving Invariants \textit{inTransitCondition} and \textit{noTransitEqual},
-needs two further invariants, along with Invariant \textit{stateChange}.
-
-\begin{inv}
-\textit{respFIFO}: Responses are received in the order in which they are sent,
-for a source-destination pair.
-\label{respFIFO}
-\end{inv}
-
+The transitions of the protocol are setup so that the following invariant holds:
 \begin{inv}
@@
 
-Invariant \textit{inTransitCondition} can be proved as follows. Whenever a
-response from a child to the parent is in transit, the parent can not send any
-response to the child, or it will violate Invariant \textit{noCross}. Using the fact that
-
+In addition, the following invariant must hold regarding the order of responses
+between the same pair of nodes:
 \begin{inv}
-\textit{prevState}: When a response is received from a child to the parent
-(parent to child) for an address, the state of the child for that address at
-the time of sending (receiving) the response is the same as the parent's
-directory for that child and address at the time of receiving (sending) the
-response.
-\label{prevState}
+\textit{respFIFO}: Responses are received in the order in which they are sent,
+for a source-destination pair, for the same address.
 \end{inv}
 
-%Top
-%a) conservative: forall p c a t, parent c p -> dir p c a t >= state c a t
-%b) forall p c a t, parent c p -> state p a t >= dir p c a t
-%c) forall c1 c2 a t, sibling c1 c2 -> compatible c1 c2 a t
-%d) nextClean: ~ clean c a t -> clean c a (S t) -> exists s d ts, recv data s a t d /\ send data s a ts d /\ clean s a ts /\ ts <= t /\ forall ti, ts <= ti < t ->  ~ write a ti
+The MSI-based protocol in Figure \ref{trans} is such that any change in either
+the state of a cache or its directory entry takes place only on sending or
+receiving a response. The parent upgrades the corresponding directory entry on
+sending a response to a child and the child downgrades its directory entry on
+sending a response to the parent.
+
+To prove Invariant \textit{conservative} at some time $t$, consider the case
+when a child cache $c$ has sent to or received from the parent $p$ some message
+before $t$. Consider the last such message $m$ sent by or received at $c$ at
+$t'$. Let's say $m$ is sent by $c$. Then $p$ can not have received any message
+from $c$ after sending $m$ in order to not violate Invariant \textit{noCross}.
+Instead, if $c$ receives $m$ at $t'$, $p$ will not receive any message after
+sending $m$ in order to not violate Invariants \textit{noCross} and
+\textit{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
+$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ (because
+the response messages sync up the next-state according to the transitions in
+Figure \ref{trans}). Since $p$ can not receive any message between $t''+1$ and
+$t$, it can only upgrade $p.\dstate[c][a]$. Since $c.\state[a]$ does not change
+between $t'+1$ and $t$, Invariant \textit{conservative} holds at $t$. If no
+message has been sent or received by $c$ before $t$, then $p$
+wouldn't have received any message sent by $c$ before $t$ as it means that the
+message must be sent by $c$ before $t$. Since the initial states of the
+directory and the child's state are both set to \In{}, the same argument holds
+(in which the parent can only upgrade), thus maintaining Invariant
+\textit{conservative}.
+
+To prove Invariant \textit{localCompatibility}, it is enough to show that a
+child $c$ upgrades $c.\state[a]$ on receiving a response for $a$ from its
+parent $p$ and $p.\dstate[c][a]$ is downgraded on receiving a response for $a$
+from $c$. If this is true, then Invariant \textit{localCompatibility} will not
+be violated on receiving a response. The transitions of Figure \ref{trans}
+ensures that Invariant \textit{localCompatibility} is not violated when sending
+a response.
+
+If the value of the state of a child at the time of sending a response is the
+same as the value of the directory for that child at the parent at the time of
+receiving the response, it follows straightaway that the parent downgrades its
+directory on receiving a response from the child, since a child downgrades on
+sending a response and the next-states are synced on receiving the response.
+The proof for this follows exactly the same arguments presented above to prove
+Invariant \textit{conservative} -- we prove by considering the last message
+sent or received at the child before $t$ (and the case when no message has been
+sent or received).
+
+Invariant \textit{noTransitWrite} can be proved as follows. Consider a response
+sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
+other response can be received by the parent at time $t_p$ such that $t+1 \le
+t_p \le t'$ to not violate Invariant \textit{noCross}, so $p.\dstate[c][a]$ at
+$t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
+directory for $c$. Similarly, no response can be sent by the child at time
+$t_c$ such that $t \le t_c \le t'$ for the same reason and hence $c.\state[a]$
+at $t \le c.\state[a]$ at $t'$ since the child can only downgrade its state.
+This means that during $t_p$ such that $t+1 \le t_p \le t'$, $p.\dstate[c][a]
+> \In$ since the parent upgraded its directory for $c$ at $t$. Similarly,
+during $t_c$ such that $t \le t_c \le t'$, $c.\state[a] < \Mo$ since the
+child upgrades its state at $t'$. Using arguments similar to the proof of
+Invariant \label{latestValue}, we can establish that no cache can be in \Mo{}
+state for address $a$ when a response is in transit from $p$ to $c$.
+Similarly, we can prove that no cache can be in \Mo{} state for address $a$
+when a response is in transit for address $a$ from $c$ to $p$. Since data is in
+transit at the same time as when the response is in transit, Invariant
+\textit{noTransitWrite} follows.
+
+%from the parent and child respectively, it is sufficient to show that the previous state
+%or the 
 %
-%(from=state(t1)=dir(t2) -> betweenPSendCRecv:dir(t)>=to&state(t)<=from, betweenCSendPRecv:state(t)<=to&dir(t)>=from), noMesgImp:state(t)=dir(t) -> conservative (ignore from=, use only a, b)
-%a) Child sends message iff it downgrades
-%b) Parent sends message iff it upgrades
-%c) State is changed only on sending or receiving a message
-%d) FIFO messages
-%e) cross: Message from child and parent do not cross
+%Invariant~\textit{conservative} can be proved by
+%further breaking them into the following invariants between a parent-child pair
+%of caches.
 %
-%cross:
-%a) parent send message only on receiving request
-%b) req from child does not overtake response from child (prove using the from states)
-%c) resp from parent does not overtake request from parent
-%d) voluntary response sent from child only when not waiting
+%\begin{inv}
+%\textit{stateChange}:
+%\begin{enumerate}
+%\item A child sends a response to its parent iff it downgrades its state.
+%\item A parent sends a response to a child iff it upgrades the directory of
+%that child.
+%\item The state and the directory can change only on sending and receiving a
+%response; the receiving entity changes its value to that of the sending entity
+%immediately after sending the response message.
+%\end{enumerate}
+%\label{stateChange}
+%\end{inv}
 %
-%In this section we specify a set of important invariants for a parent-child pair of caches.
+%\begin{inv}
+%\textit{respFIFO}: Responses are received in the order in which they are sent,
+%for a source-destination pair, for the same address.
+%\label{respFIFO}
+%\end{inv}
+%
+%\begin{inv}
+%\textit{noCross}: A response from a cache to its child for an address can not
+%be in transit while another response from the same child to the parent for the
+%same address is in transit.
+%\label{noCross}
+%\end{inv}
+%
+%Invariant~\textit{conservative} at any time $t$ for child cache $c$ and parent
+%$p$ for an address $a$ can be proved as follows. Consider the case when
+%$c.\state[a]$ never changed before $t$, then $c.\state[a] = \In$ because of the
+%initialization conditions described in Section \ref{System}. No response would
+%have been sent by $c$ to its parent for address $a$ before $t$, otherwise
+%$c.\state[a]$ would have changed. Thus the parent would not have received any
+%response from $c$, $p.\dstate[c][a]$ could have changed only by sending a
+%response to $c$ which can only upgrade it from the initial value of $\In$. Thus
+%at $t$, $p.\dstate[c][a] \ge c.state[a]$.
+%
+%Now consider the case when $c.\state[a]$ last changed at $t'$ such that $t' <
+%t$ because of receiving a response from $p$, sent at $t''$. $c.\state[a]$ at
+%$t'+1 = p.\dstate[c][a]$ at $t''+1$ because of Invariant \textit{stateChange}.
+%$c$ neiter sends nor receives a response at any $t_i$ such that $t' < t_i < t$.
+%If $p$ received a response at some $t_m$ such that $t'' < t_m < t$,
+%then it would either violate Invariant \textit{noCross}, or the fact that no
+%message was sent between $t' < t_i < t$ from $c$. Thus, $p.\dstate[c][a]$ at $t
+%\ge p.\dstate[c][a]$ at $t''+1$ because any message sent can only upgrade
+%$p.\dstate[a]$. This ensures that Invariant \textit{conservative} is obeyed. We
+%can similarly prove this invariant holds when $c.\state[a]$ last changes before
+%$t$ by sending a response.
+%
+%Invariant \textit{stateChange} can be proved directly from the atomic
+%transitions for MSI. Other protocols have similar atomic transitions obeying
+%these invariants.
+%
+%Invariant \textit{respFIFO} can be enforced by the interconnect network which
+%ensures a FIFO ordering for messages sent between a pair of nodes.
+%
+%---------------------------------------------
+%
+%%Invariant \textit{respFIFO} is known as point-to-point FIFO ordering in the
+%%Network-on-Chip NoC community, and is enforced by common routing algorithms
+%%like Dimension-ordered routing \cite{dimension}. Only the parent-to-child
+%%channels need to be FIFO ordered, the FIFO ordering in child-to-parent channels
+%%can be enforced using the protocol itself.
+%
+%\begin{inv}
+%\textit{noTransitEqual}: When no response is in transit for an address, then
+%the state of a cache for that address is the same as its parent's directory for
+%that cache and address.
+%\label{noTransitEqual}
+%\end{inv}
+%
+%Proving Invariants \textit{inTransitCondition} and \textit{noTransitEqual},
+%needs two further invariants, along with Invariant \textit{stateChange}.
+%
+%Invariant \textit{inTransitCondition} can be proved as follows. Whenever a
+%response from a child to the parent is in transit, the parent can not send any
+%response to the child, or it will violate Invariant \textit{noCross}. Using the fact that 
+%
+%
+%
+%
+%
+%
+%
+%\begin{inv}
+%\textit{inTransitCondition}:
+%\begin{enumerate}
+%\item When a response is sent from a child $c$ at time $t$ for an address $a$,
+%and has not been received by the parent $p$ before time $t'$, $\forall t_i, t <
+%t_i \le t' @-> c.\state[a]$ at $t_i \le c.\state[a]$ at $t+1$ and $\forall t_i,
+%t \le t_i \le t' @->$ $p.\dstate[c][a]$ at $t_i \ge c.\state[a]$ at $t$.
+%\item When a response is sent from parent $p$ at time $t$ for an address $a$,
+%and has not been received by its child $c$ before time $t'$, $\forall t_i, t <
+%t_i \le t' @-> p.\dstate[c][a]$ at $t_i \ge p.\dstate[c][a]$ at $t+1$ and
+%$\forall t_i, t \le t_i \le t' @->$ $c.\state[a]$ at $t_i \le p.\dstate[c][a]$ at
+%$t$.
+%\end{enumerate}
+%\label{inTransitCondition}
+%\end{inv}
+%
+%\begin{inv}
+%\textit{prevState}: When a response is received from a child to the parent
+%(parent to child) for an address, the state of the child for that address at
+%the time of sending (receiving) the response is the same as the parent's
+%directory for that child and address at the time of receiving (sending) the
+%response.
+%\label{prevState}
+%\end{inv}
+%
+%
+%
+%%Top
+%%a) conservative: forall p c a t, parent c p -> dir p c a t >= state c a t
+%%b) forall p c a t, parent c p -> state p a t >= dir p c a t
+%%c) forall c1 c2 a t, sibling c1 c2 -> compatible c1 c2 a t
+%%d) nextClean: ~ clean c a t -> clean c a (S t) -> exists s d ts, recv data s a t d /\ send data s a ts d /\ clean s a ts /\ ts <= t /\ forall ti, ts <= ti < t ->  ~ write a ti
+%%
+%%(from=state(t1)=dir(t2) -> betweenPSendCRecv:dir(t)>=to&state(t)<=from, betweenCSendPRecv:state(t)<=to&dir(t)>=from), noMesgImp:state(t)=dir(t) -> conservative (ignore from=, use only a, b)
+%%a) Child sends message iff it downgrades
+%%b) Parent sends message iff it upgrades
+%%c) State is changed only on sending or receiving a message
+%%d) FIFO messages
+%%e) cross: Message from child and parent do not cross
+%%
+%%cross:
+%%a) parent send message only on receiving request
+%%b) req from child does not overtake response from child (prove using the from states)
+%%c) resp from parent does not overtake request from parent
+%%d) voluntary response sent from child only when not waiting
+%%
+%%In this section we specify a set of important invariants for a parent-child pair of caches.

### 2014-02-04 paper/System.tex
intuition

@@
 
-The behavior of the system are described via the atomic state
+The behavior of the system is described via the atomic state
 transitions. These transitions have two parts: the guard dictating
@@
 logically happen happen one at a time. Figure~\ref{trans} gives the
-atomic transitions for a cache coherent memory subsystem. The actions
+atomic transitions for an MSI-based cache coherent memory subsystem. The actions
 in the transitions represent only the specific system states that

### 2014-02-04 paper/Intuition.tex
intuition

@@
-\section{Intuition}
-\label{Sec:ProofStructure}
+\section{Proof of MSI protocol being sound wrt Store atomicity}
+\label{safety}
 
-In this section, we present the proof structure for our MSI-based cache
-coherence protocol of Figure \ref{trans}. Specifically, we formally reduce the
-proof of the store atomicity theorem for the whole system to the proof of a set
-of invariants. In order to prove these invariants we introduce the main
-``intuition'' behind the correctness of the protocol. This intuition is not
-restricted to the MSI-based protocol; Section \ref{Sec:Refinements} shows that
-it applies to other protocols like MOSI.
+In this section, we present the proof structure for the soundness property of our
+MSI-based cache coherence protocol of Figure \ref{trans}. Specifically, we
+formally reduce the proof of the store atomicity theorem for the whole system
+to the proof of a set of invariants. In order to prove these invariants we
+introduce the main ``intuition'' behind the correctness of the protocol. This
+intuition is not restricted to the MSI-based protocol; Section
+\ref{Sec:Refinements} shows that it applies to other protocols like MOSI.
 

### 2014-02-04 paper/Deadlock.tex
intuition

\section{Liveness properties}
\label{liveness}

In Section \ref{safety} we showed how the MSI protocol of Figure \ref{trans}
obeys the Store atomicity theorem. In addition to this criteria, we must
establish that the system makes forward progress.

Forward progress can be either of the following:
\begin{enumerate}
\item Weak forward progress: The system never reaches a state where none of the transitions of Figure
\ref{trans} apply or
\item Strong forward progress: At any time, eventually a new request from some processor will be
processed
\end{enumerate}

In order to prove strong forward progress, we need to
establish some more conditions on the strategy to execute the transitions of
Figure \ref{trans} in each cache -- the caches must ensure local
starvation-freedom in the processing of messages when sufficient resources are
available. We do not discuss this issue in this paper and instead focus on 
weak forward progress.

In order to establish that some transition always executes, we prove the
following important invariant.

\begin{inv}
\textit{waitCond}: When a cache $x$ has sent a request to another cache $y$ and
is waiting for a response from $y$, then either $y$ has not yet received the
request, or it has received the request and sent the appropriate response to
$x$.
\end{inv}

This condition ensures that every request gets a response eventually. Using
this invariant, it is easy to see that there will always be some cache in the
system which can either send a request to another cache (ChildSendReq or
ParentSendReq) or process a request (ParentRecvReq, ChildRecvReq or
ChildDropReq) or accept a response from another cache (ChildRecvResp or
ParentRecvResp). We do not give the formal proof here due to lack of space.

### 2014-02-04 paper/Intuition.tex
intuition

@@
 intuition is not restricted to the MSI-based protocol; Section
-\ref{Sec:Refinements} shows that it applies to other protocols like MOSI.
+\ref{sec:Refinements} shows that it applies to other protocols like MOSI.
 
@@
 child upgrades its state at $t'$. Using arguments similar to the proof of
-Invariant \label{latestValue}, we can establish that no cache can be in \Mo{}
+Invariant \textit{latestValue}, we can establish that no cache can be in \Mo{}
 state for address $a$ when a response is in transit from $p$ to $c$.

### 2014-02-04 paper/Intuition.tex
Intuition cleanup

@@
 
-In this section, we present the proof structure for the soundness property of our
-MSI-based cache coherence protocol of Figure \ref{trans}. Specifically, we
-formally reduce the proof of the store atomicity theorem for the whole system
-to the proof of a set of invariants. In order to prove these invariants we
-introduce the main ``intuition'' behind the correctness of the protocol. This
-intuition is not restricted to the MSI-based protocol; Section
-\ref{sec:Refinements} shows that it applies to other protocols like MOSI.
+In this section, we present the proof structure for the soundness
+property of our MSI-based cache coherence protocol of
+Figure~\ref{trans}. Specifically, we formally reduce the proof of the
+store atomicity theorem for the whole system to the proof of a set of
+simple invariants by leveraging the ``intuitive'' justification of
+correctness of our protocol.
+
+%% This intuition
+%% is not restricted to the MSI-based protocol;
+%% Section~\ref{sec:Refinements} shows that it applies can apply to other
+%% protocols.
 
@@
 
-We've defined store atomicity to be a whole system property. The first step in
-our framework is to reduce the store atomicity property down to the following 4
-global invariants about nodes:
+The store atomicity is a whole system property. Intuitively our
+protocol is correct because it guarantees that at most one node can
+write at a time, and that any local cache that attempts to read has
+the ``latest'' value. We can state this formally with the following 4
+invariants about nodes:
 
 \begin{defn}
-\textit{clean}: A cache is said to be \clean{} for an address $a$ iff
+\textit{clean}: A cache is said to be \clean{} for an address $a$ if and only if
 the state of the cache for that address is at least \Sh{} and the
@@
 \begin{defn}
-\textit{compatibile}: Values of states $x_1$ and $x_2$ are compatible iff
+\textit{compatibile}: Values of states $x_1$ and $x_2$ are compatible if and only if
 either $x_1$ or $x_2$ is \Mo{}, then the other is \In.
@@
 
-We can see that the correctness of our system (Theorem~\textit{storeAtomicity})
-follows naturally from Invariants \ref{latestValue} to \ref{noReqAgain}.  A
-store for an address may happen only when leaf cache is in \Mo{} state for that
-address (Invariant~\textit{processReq}). Any two distinct leaf caches are not
-ancestors of each other, so by Invariant \textit{nonAncestorCompatible}, if a
-leaf cache is in \Mo{} state, all other leaf caches will be in \In{} state
-which prevents processing of both load and store requests for the same address
-in other leaf caches (again by Invariant \textit{processReq}). Thus, Definition
-\textit{uniqStTimes} holds.  Given Invariant~\textit{noReqAgain}, once a
-response has been produced for a request $i$, the cache may subsequently only
-process $j^{th}$ request such that $j > i$. Thus there is at most one response
-to each request, satisfying Definition~\textit{uniqRespLabels}. A load request
-can be processed only if a cache is at least in \Sh{} state (again by Invariant
-\textit{processReq}). When a leaf cache is at least in \Sh{} state, since it
-has no children, it is \clean. A load response supplies the data for that
-address present in the cache, which, by Invariant~\textit{lastestValue},
-contains the ``latest value''. Thus, the final condition for
-Theorem~\textit{storeAtomicity} is also satisfied.
+Given these invariants, we codify our intuition by showing that
+Theorem~\textit{storeAtomicity} follows naturally from these
+invariants.  A store may happen only when leaf cache is in \Mo{} state
+for that address (Invariant~\textit{processReq}). Leaf caches cannot
+be ancestors of any other node, so by Invariant
+\textit{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
+other leaf caches will be in \In{} state preventing the possibility of
+load and store requests for the same address (by Invariant
+\textit{processReq}), showing Definition~\textit{uniqStTimes} holds.
+
+Given Invariant~\textit{noReqAgain}, once a response has been produced
+for a request $i$, the cache may subsequently only process $j^{th}$
+request such that $j > i$. Thus there is at most one response to each
+request, satisfying Definition~\textit{uniqRespLabels}. A load request
+may be processed only if a cache is at least in \Sh{} state (Invariant
+\textit{processReq}). As leaf caches have no children, a leaf cache in
+at least in \Sh{} state for address $a$ is \clean for that address. A
+load response supplies the data for that address present in the cache,
+which, by Invariant~\textit{lastestValue}, contains the ``latest
+value''. Thus, the final condition for Theorem~\textit{storeAtomicity}
+is also satisfied. All that remains is to show that these properties hold.
 
@@
 
-Invariants~\textit{processReq} and \textit{noReqAgain} can be proved directly
-from the atomic transitions describing the system (Figure~\ref{trans} for MSI;
-other protocols must also have similar transitions). Proving
-invariants~\textit{uniqM} and \textit{latestValue} needs further work. We can
-prove both using the following invariants on parent-child relationships.
+We can directly show that Invariants~\textit{processReq} and
+\textit{noReqAgain} hold at the initial configuration and are
+preserved through each atomic transition of our protocol
+(Figure~\ref{trans}). Invariants~\textit{uniqM} and
+\textit{latestValue} correctness rely on properties of the node
+hierarchy. We can prove both using the following invariants on
+parent-child relationships.
 
@@
 
-For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in the
-cache tree hierarchy, from Invariants \textit{conservative} and the first part
-of \textit{localCompatibility}, we can show that for any address address $a$,
-$c.\state[a] \le p.\state[c][a]$. This is because the ancestor relation is the
-reflective-transitive closure of the parent relation, and Invariants
-\textit{conservative} and the first part \textit{localCompatibility} of defines
-a $\ge$ relation between the states of a parent and its child.
+For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
+the cache tree hierarchy, from Invariants \textit{conservative} and
+\textit{localCompatibility}, we can show that for any address $a$,
+$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
+relation is the reflective-transitive closure of the parent relation,
+and Invariants \textit{conservative} and the first part
+\textit{localCompatibility} of defines a $\ge$ relation between the
+states of a parent and its child.
 
-Any two caches $c_1, c_2$ in which neither is an ancestor of the other will
-have a lowest common ancestor $x$ such that $x$ is an ancestor of both the
-caches and $x$ has two distinct children $x_1$ and $x_2$ such that they are the
-ancestors of $c_1$ and $c_2$ respectively. For any address $a$,
-$x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are compatible by the second part
-of Invariant \ref{localCompatibility}. From this, it straightaway follows that
-$x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from Invariant
-\ref{conservative}). From the previous paragraph, it follows that $c_1$ and
-$c_2$ are also compatible. This proves Invariant \textit{nonAncestorCompatibility}.
+Any two caches $c_1, c_2$ neither of which are an ancestor of the
+other will have a lowest common ancestor $x$ which is an ancestor of
+both $c_1$ and $c_2$ and has two distinct children $x_1$ and $x_2$
+such that they are the ancestors of $c_1$ and $c_2$ respectively. For
+any address $a$, $x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are
+compatible by the second part of
+Invariant~\ref{localCompatibility}. From this, it straightaway follows
+that $x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from
+Invariant~\ref{conservative}). From the previous paragraph, it follows
+that $c_1$ and $c_2$ are also compatible. This proves
+Invariant \textit{nonAncestorCompatibility}.
 
-We can prove that Invariant \textit{latestValue} is obeyed by strong induction
-on time. Consider two cases: cache $c$ is already \clean{} for an address $a$
-at time $t$ or the contrary.  If \clean{}, then it has the latest value at time
-$t$. Since none of its children's directory entries can be in \Mo{}, all its
-descendents should be no greater than \Sh{} for that address (as discussed
-above). Any non-ancestor of $c$, and non-descendent of $c$ must be compatible
-(as discussed above). Since $c$ is \clean, it is at least in the \Sh{} state,
-so compatibility means that non-descendents can not be in \Mo{} state. All of
-$c$'s ancestors are non leaves, so they can not process any store request. So,
-no cache could have processed a store request other than $c$ itself, which
-means that $c$ contains the latest value for address $a$ at time $t+1$ if it
-was \clean{} at time $t$.
+We prove Invariant \textit{latestValue} by strong induction on
+time. Consider two cases: cache $c$ is already \clean{} for an address
+$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
+value at time $t$. Since none of its children's directory entries can
+be in \Mo{}, all its descendants should be no greater than \Sh{} for
+that address (as discussed above). Any non-ancestor of $c$, and
+non-descendant of $c$ must be compatible (as discussed above). Since
+$c$ is \clean, it is at least in the \Sh{} state, so compatibility
+means that non-descendants can not be in \Mo{} state. All of $c$'s
+ancestors are non-leaves, so they can not process any store
+request. Therefore, no cache save $c$ may have process a store
+request, which means that $c$ contains the latest value for address
+$a$ at time $t+1$ if it was \clean{} at time $t$.
 
-If $c$ was not clean at time $t$, it must have received data from either its
-parent or its child.  As can be seen in Figure \ref{trans}, whenever data is
-sent either from a parent to a child, or from a child to the parent, the sender
-is \clean. By the induction hypothesis, it means that the sender has the latest
-value for the address. Because of Invariant \ref{noTransitWrite}, no store
-request can be processed when the data is in transit, which means $c$ will have
-the latest value for address $a$ at time $t+1$.
+If $c$ was not clean at time $t$, it must have received data from
+either its parent or its child.  As can be seen in Figure~\ref{trans},
+whenever data is sent either from a parent to a child, or from a child
+to the parent, the sender is \clean. By the induction hypothesis, it
+means that the sender has the latest value for the address. Because of
+Invariant~\ref{noTransitWrite}, no store request can be processed when
+the data is in transit, which means $c$ will have the latest value for
+address $a$ at time $t+1$.
 
-\subsection{Main intuition to prove the pair-wise properties}
-We present two basic invariants that the MSI protocol of Figure \ref{trans}
-obeys. In fact the protocol (\ie the request/response message semantics) was
-designed to ensure the following two invariants. We believe that most
-cache-coherence protocols would end up obeying these two invariants; we show
-how MOSI, for instance, obeys these invariants in Section
-\ref{sec:Refinements}. These two invariants together can prove all the pair-wise
-properties of Section \ref{pairwise}.
+\subsection{Proving parent-child hierarchy properties}
+
+%ndave: This needs some repointing
+
+We present two basic invariants that the MSI protocol of
+Figure~\ref{trans} obeys. In fact the protocol (\ie{} the
+request/response message semantics) was designed to ensure the
+following two invariants. We believe that most cache-coherence
+protocols would end up obeying these two invariants; we show how MOSI,
+for instance, obeys these invariants in
+Section~\ref{sec:Refinements}. These two invariants together can prove
+all the pair-wise properties of Section~\ref{pairwise}.
 
@@
 
-The MSI-based protocol in Figure \ref{trans} is such that any change in either
+The MSI-based protocol in Figure~\ref{trans} is such that any change in either
 the state of a cache or its directory entry takes place only on sending or
@@
 %%In this section we specify a set of important invariants for a parent-child pair of caches.
+
+%%  LocalWords:  Invariants invariants

### 2014-02-04 paper/Background.tex
System is a little more motivated. Pull out bits of Proof structure to end and drop into new section
stuff

@@
-\section{Correctness of our Cache Coherence System}
+\section{Correctness of a Cache Coherence Protocol}
+\label{Background}
 
-Before we can discuss our protocol framework and it's verification we
-must first establish exactly what it means for a system to be
-correct.
+Before we can discuss our protocol and proof framework we must first establish
+exactly what it means for a cache coherence protocol to be correct.
 
-Intuitively, the easiest view of shared memory is as a single ported
-memory without caches multiplexed amongst the many individual clients/processors. In
-this view, all operations in the system appear to have a single unique
-total ordering.
+Intuitively, the simplest view of shared memory is as a single ported memory
+without caches multiplexed amongst the many individual clients/processors. In
+this view, all operations in the system appear to have a single unique total
+ordering.
 
-All modern systems have caches between the processor and memory for faster memory access.
-It is often also useful for performance to permit some reorderings fo requests from a processor to the memory.
-For instance, in the context of a processor with local store buffer, we may
-wish to allow store values to bypass subsequent load requests. The specification of
- which reorderings are permitted and which are restricted defines the memory consistency
+All modern systems have caches between the processor and memory for faster
+memory access.  It is often also useful to permit some reorderings of requests
+from a processor to the memory to improve performance.  For instance, in the
+context of a processor with local store buffer, we may wish to allow a store to
+bypass a subsequent load request to a different location. The specification of
+which reorderings are permitted or restricted defines the memory consistency
 model.
-%, though they may result in a stronger verification requirement.
-
-%As part of our design exploration for our concrete protocol example,
-%we may want to change the memory consistency model. To this end, we
-%are using the notion of store-atomicity to define our memory
-%model. Most memory consistency models require that the memory
-%subsystem obeys the store-atomicity property and as such our design
-%should be modifiable to any of these designs.
 
 Most memory consistency models can be partitioned into two
-parts~\cite{Arvind-memory-model}: a), a ``global'' store-atomicity
-property stating that for each location/address in memory, every read and write access to that 
-location occurs in some total order with each operation
-seeing the effects of previous operations and not later ones (each location behaves like an atomic register) and b),
-a ``local'' reordering restriction between requests to different
-locations within each processor. The advantage of this approach is
-that reordering restrictions pertain to reordering requests locally within
-the same processor, while the store-atomicity property is across the whole
-system. Thus changing the memory model
-by modifying the reordering restrictions will not incur a reverification of
-the store-atomicity property.
-
-%For our concrete
-%protocol example, we will assume that our memory model is weakly
-%ordered, in that there are inter-address dependencies between
-%processors.
-
-%% For example, it dictates whether a store request to an address $x$ can
-%% be issued to the memory sub-system before a load request to a
-%% different address $y$, which was generated by the same processor
-%% earlier (\ie load $y$ happened before store $x$ is program order, but
-%% the order of issue to memory sub-system is reversed).
+parts~\cite{Arvind-memory-model}: a) a ``global'' store-atomicity property
+stating that for each location/address in memory, every read and write access
+to that location occurs in some total order with each operation seeing the
+effects of previous operations and not later ones (in other words, each
+location behaves like an atomic register) and b), a ``local'' reordering
+restriction between requests to different locations within each processor.
+Changing the memory model by modifying the reordering restrictions will not
+incur a reverification of the store-atomicity property.
 
@@
 location, in the presence of caches, is known as \emph{cache coherence}. In the
-absence of caches, since there is exactly one copy of the data for each
-address, it is easy to ensure cache coherence. The problem comes in the
-presence of shared caches -- multiple copies of the same address can exist in
-different caches, and if these locations are independently written, then the
-same address can end up having different values in two different caches, and
-each location may not behave like an atomic register unless some sort of
-co-ordination happens between the different caches -- this co-ordination is
-summed up as the \emph{cache coherence protocol}.
+presence of caches, multiple copies of the same address can exist in different
+caches. If these copies are independently written, the same address can end up
+having different values in different caches, violating the atomic register-like
+behavior unless the caches co-ordinate via a \emph{cache coherence protocol}.
 
-Now that we have informally discussed the meaning of cache coherence and store
-atomicity, we can now provide a formal definition.
+We will now provide the formal definition of store atomicity.
 
 \subsection{Formal definition of store atomicity}
+We will model an abstract memory sub-system, which interfaces with a set of
+processors. The requests issued by the processors to the memory sub-system can
+be modeled using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request
+issued by processor $p$.
 
-To properly define the interface with between the processor and the
-caching system we must define the request and responses for each
-processor. 
-The requests issued by the processors to the memory sub-system can be modeled
-using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
-processor $p$.
-The processor name and the position of the request within the processor
-uniquely identifies the request.
+%Figure~\ref{req-resp} shows the
+%data-type corresponding to a request issued from the processor to the
+%memory sub-system.
 
-Figure~\ref{req-resp} shows the
-data-type corresponding to a request issued from the processor to the
-memory sub-system.  A request contains the address (\addrQ)
-corresponding to the request, and a type indicating whether the
-request is a load or a store (\desc). In case of a store request, it
-also contains data (\dataQ) supplied by the processor for that
-address. A response contains a label
-(\labelR) to identify the request corresponding to the response. The label
-contains a processor name and a position, which uniquely identifies a request.
-In the case of a load request, the memory sub-system returns \dataR, the data
-returned by the memory sub-system for that address. The memory sub-system also
-returns a natural number, \timeR, which is used solely in the definition of
-store atomicity, and is not returned in a real system. \timeR{} can be informally
-thought to represent the position at which this request has been handled
-globally among all the requests issued from all the processors.
+A request contains the address (\addrQ) corresponding to the request, and a
+\desc{} indicating whether the request is a load or a store. A store request
+also contains the data supplied by the processor for that address (\dataQ). A
+response contains a label (\labelR) containing a processor name and a position
+to uniquely identify the request.  In the case of a load request, the memory
+sub-system returns \dataR, the data returned by the memory sub-system for that
+address. The memory sub-system also returns an $\mathbb{N}$, \timeR, which
+intuitively represents the position at which the request is handled among the
+global set of handled requests from all the processors. \timeR{} is used solely
+for the definition of store atomicity, a real system does not return such a
+field.
 
@@
 
-\begin{figure}
-\centering
-\begin{subfigure}{5.5cm}
-\begin{tabular}{|lp{4.5cm}|}
-\hline
-\multicolumn{2}{|c|}{\Request}\\
-\hline
-\addrQ:& Address associated with the request\\
-\desc:& Request type (\Ld{} or \St)\\
-\dataQ:& Data associated with a \St{} request\\
-\hline
-\end{tabular}
-\end{subfigure}~~~~
-\begin{subfigure}{5.5cm}
-\begin{tabular}{|lp{4.5cm}|}
-\hline
-\multicolumn{2}{|c|}{\Response}\\
-\hline
-\labelR:& Label identifying the request corresponding to this response\\
-\timeR:& Natural number representing global order of (processing) requests\\
-\dataR:& Data returned by the memory sub-system for a \Ld{} request\\
-\hline
-\end{tabular}
-\end{subfigure}
-\caption{Data types for request and response}
-\label{req-resp}
-\end{figure}
+%\begin{figure}
+%\centering
+%\begin{subfigure}{5.5cm}
+%\begin{tabular}{|lp{4.5cm}|}
+%\hline
+%\multicolumn{2}{|c|}{\Request}\\
+%\hline
+%\addrQ:& Address associated with the request\\
+%\desc:& Request type (\Ld{} or \St)\\
+%\dataQ:& Data associated with a \St{} request\\
+%\hline
+%\end{tabular}
+%\end{subfigure}~~~~
+%\begin{subfigure}{5.5cm}
+%\begin{tabular}{|lp{4.5cm}|}
+%\hline
+%\multicolumn{2}{|c|}{\Response}\\
+%\hline
+%\labelR:& Label identifying the request corresponding to this response\\
+%\timeR:& Natural number representing global order of (processing) requests\\
+%\dataR:& Data returned by the memory sub-system for a \Ld{} request\\
+%\hline
+%\end{tabular}
+%\end{subfigure}
+%\caption{Data types for request and response}
+%\label{req-resp}
+%\end{figure}
 
-In the following, a \Response{} set only contains those responses sent by the
-memory sub-system.
-
-The responses should obey the following sanity conditions:
-
-\begin{defn} (The labels in the responses are unique. Since \dataR{} field is
-irrelevant for store requests, so uniqueness of responses does not depend on
-\dataR{} field for stores.)
+The \Response{} set contains the set of responses from the memory sub-system.
+Responses should obey the following sanity conditions. 
+\begin{defn} (The labels in the responses are unique)
+%Since \dataR{} field is
+%irrelevant for store requests, so uniqueness of responses does not depend on
+%\dataR{} field for stores.)
 \small
@@
 
-\begin{defn} (No response can share the same time as another store response.)
+\begin{defn} (No response can share a \timeR{} value with another store response.)
 \small
@@
 This brings us to the main condition for store atomicity which states that the
-value returned for a load response corresponds to the \emph{last earlier} store
-request that has been processed in the system for that address This is
-indicated by the proposition \textit{lastEarlierValue}$(r)$ for response $r$.
-The \emph{last} and \emph{earlier} are defined according to the field \timeR{}
-returned in the responses. In case no store has happened earlier for a
+data returned for a load response corresponds to the \emph{last earlier} store
+request from any processor that has been handled for that address,
+\emph{last} and \emph{earlier} being defined according to the field \timeR{}
+returned in the responses.  This proposition is indicated by
+\textit{lastEarlierValue}$(a, t, d)$ for a load request for address $a$ handled
+at time $t$, returning data $d$.  In case no store has happened earlier for a
 particular address when a load is processed, then the load response returns the
-initial value for the address. This is indicated by the proposition
-\textit{initialValue}$(r)$ for response $r$.
+initial value for the address, the whole proposition indicated by
+\textit{initialValue}$(a, t, d)$.
 \begin{defn}
@@
 \begin{multline*}
-\textit{initialValue} := \lambda (r \in \Response) \Rightarrow
-(\reqFn (r.\labelR)).\desc = \Ld @->\\ r.\dataR = \initData (r.\addrQ) \wedge 
-\forall (r' \in \Response), 0 \le r'.\timeR < r.\timeR @->\\
-\neg ((\reqFn (r'.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ \wedge (\reqFn (r'.\labelR)).\desc = \St)
+\textit{initialValue} := \lambda (a \in Addr)\; (t \in \mathbb{N}) \; (d \in Data)\Rightarrow
+d = \initData (a) \wedge \\
+\forall (r' \in \Response), 0 \le r'.\timeR < t @->\\
+\neg ((\reqFn (r'.\labelR)).\addrQ = t \wedge (\reqFn (r'.\labelR)).\desc = \St)
 \end{multline*}
@@
 \begin{multline*}
-\textit{initialValue} := \lambda (r \in \Response) \Rightarrow
-(\reqFn (r.\labelR)).\desc = \Ld @->
-\exists (r_m \in \Response),\\ r.\dataR = (\reqFn (r_m.\labelR)).\dataQ \wedge 
-(\reqFn (r_m.\labelR)).\desc = \St \\ \wedge (\reqFn (r_m.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ\\ \wedge
-\forall (r' \in \Response), r_m.\timeR \le r'.\timeR < r.\timeR @->\\
-\neg ((\reqFn (r'.\labelR)).\addrQ = (\reqFn (r.\labelR)).\addrQ \wedge (\reqFn (r'.\labelR)).\desc = \St)
+\textit{lastEarlierValue} := \lambda (a \in Addr)\; (t \in \mathbb{N}) \; (d \in Data)\Rightarrow
+\exists (r_m \in \Response),\\ d = (\reqFn (r_m.\labelR)).\dataQ \wedge 
+(\reqFn (r_m.\labelR)).\desc = \St \wedge\\ (\reqFn (r_m.\labelR)).\addrQ = a \wedge
+\forall (r' \in \Response), r_m.\timeR \le r'.\timeR < t @->\\
+\neg ((\reqFn (r'.\labelR)).\addrQ = a \wedge (\reqFn (r'.\labelR)).\desc = \St)
 \end{multline*}
@@
 \textit{uniqRespLabels} \wedge
-\textit{uniqStTimes} \wedge \textit{localOrdering}\\ \wedge 
-(\forall (r \in \Response), \textit{initialValue}(r) \vee \textit{lastEarlierValue}(r))
+\textit{uniqStTimes} \wedge \textit{localOrdering} \wedge \\
+(\forall (r \in \Response), \mylet q := \reqFn(r.\labelR) \myin q.\desc = \Ld \wedge \\
+(\textit{initialValue}(q.\addrQ, r.\timeR, r.\dataR) \vee \textit{lastEarlierValue}(q.\addrQ, r.\timeR, r.\dataR)))
 \end{multline*}

### 2014-02-04 paper/System.tex
Refinements
some modifications of background

@@
-\section{System}
+\section{MSI-based cache coherence protocol}
 \label{sec:System}
 
-We have chosen as our concrete example is a hierarchical
-MSI protocol~\cite{MSI}. As we will
-discuss later, this can naturally extended to MESI and MOESI based protocols.
-Conceptually, our cache coherence system is
-organized as in a tree-like structure with the root node as the main
-memory, the leaves as the last local cache for individual processors
-interfaces and the remaining nodes as intermediate cache structures
-(see Figure~\ref{hier}). Whether a cache is a leaf or not is denoted
-by the relation \leaf$(c)$.  The relation $\parent(c, p)$
-denotes that $p$ is the parent of cache $c$.
+We have chosen as our concrete example is a hierarchical MSI
+protocol~\cite{MSI}. Logically, the system is organized as in a tree-like
+structure with the root connected to the main memory and each leaf interfaced
+with a processor (Figure~\ref{hier}). In the following, $\leaf(c)$ indicates
+that a cache is a leaf, and $\parent(c,p)$ indicates that $p$ is the parent of
+$c$.
 
-Input and output happen at the leaves in the form of memory requests
-issued by and responses issued to the associated processor. These will be
+Input and output happens at the leaves in the form of memory requests
+issued from and responses issued to the associated processor. These will be
 represented by two mappings one from $\mathbb{N}$ to requests and the other
@@
 
-Each node maintains the a permission state for
-each address, denoting what operations may be performed on the
-data. An address may be in the modified state (\Mo) representing that
-it may be both read and modified, or in the shared state (\Sh) where
-it may only be read, or in the Invalid state (\In), meaning it can be
-neither modified or read. This forms a natural total order of
-permissions ($\In < \Sh < \Mo$). For leaf caches, the meaning of these
-permission states is straightforward, a load request can be processed only if the
-leaf is at least in the \Sh{} state, and a store request can be processed only
-if it is in the \Mo{} state. For the non-leaf cache, these states denote the
-highest state present in the subtree rooted at this cache, denoting that some
-cache in the subtree has the specified permissions (we are modeling inclusive
-caches). Of course, all these invariants have to be enforced by the protocol.
+Each node maintains a permission state for each address $a$, $c.\state[a]$, in
+addition to the data associated with the address, $c.\data[a]$. The permission
+state denotes which operations may be performed on the data. An address may be
+in the modified state (\Mo) representing that it may be both read and modified,
+in the shared state (\Sh) in which it may only be read, or in the Invalid state
+(\In), indicating that it can be neither modified or read. This forms a natural
+total order of permissions ($\In < \Sh < \Mo$). For a leaf cache, the meaning
+of these permission states is straightforward, a load request can be processed
+only if the leaf is at least in the \Sh{} state, and a store request can be
+processed only if it is in the \Mo{} state. For a non-leaf cache, it denotes
+that no cache in the subtree rooted at that cache has more permissions (we are
+modeling inclusive caches).  If a cache does not have a particular location,
+the it is equivalent to its permission being \In.  Initially, all addresses in
+each cache is in the \In{} state, save for those in the root cache which are in
+the \Mo{} state.
 
-Each cache in addition to the data associated with each address $a$
-(denoted by $c.\data[a]$) also has a \emph{coherence state}
-associated with the address (denoted by $c.\state[a]$) holding the
-permissions on that address (\Mo, \Sh, or \In). If a cache does not have a
-particular location, the it is equivalent to its permission being \In.
-Initially, all addresses in each cache is in the \In state, save for those in
-the root cache which are in the \Mo state.
+%Of course, all these invariants have to be enforced by the protocol.
 
-Non-leaf nodes also keep a copy of the coherence state of each of its
-children in a \emph{directory}, which contains a the state of each of
-its children caches for a particular address. For a non-leaf cache
-$p$, the state in its directory for child cache $c$ is denoted by
-$p.\dstate[c][a]$, for address $a$. This local version may be stale,
-in that the child may have changed it state but the parent has not yet
-updated itself. However, the parent state will always be a
-conservative over-approximation of the child state.
+Non-leaf caches also keep a (possibly stale) copy of the permission state of
+each of its children in a \emph{directory} for every address. The directory
+entry for a cache $c$ whose parent is $p$ for address $a$ is denoted by
+$p.\dstate[c][a]$. The directory entry for a cache, for an address is never
+less than the actual cache's permission state for that address.
 
-In order for a cache to change to a higher state for a given address,
-\ie{} \emph{upgrade}, it has to \emph{request} its parent.  For a
-cache to change to a lower state, \ie \emph{downgrade}, it has to
-notify its parent cache indicating the downgrade -- it uses
-\emph{response} messages for this purpose. Conversely, for a parent
-cache to force a child cache to downgrade the child's state for an
-address (in order to give permissions to a different child cache), or
-equivalently, to downgrade the parent's directory state for the child,
-it has to \emph{request} that child cache to downgrade, and a parent
-can notify a child about upgrading the child's state, or equivalently,
-about upgrading the parent's directory state for the child, by sending
-a \emph{response} message.
+In order for a cache to change to a higher state for a given address, \ie{}
+\emph{upgrade}, it has to \emph{request} its parent.  For changing to a lower
+state, \ie{} \emph{downgrade}, it has to notify its parent of the downgrade
+using a \emph{response} message.  Similarly, for a parent to force a child to
+downgrade for an address (in order to give permissions to a different child),
+or equivalently, to downgrade the parent's directory for the child and address,
+it has to \emph{request} that child to downgrade, and a parent can notify a
+child about upgrading the child's state, or equivalently, about upgrading the
+parent's directory state for the child, by sending a \emph{response} message.
 
-Each cache is connected to its parent cache using 3 logical FIFO
-connections: a) \cpReq{} channel to transmit requests from the child
-to the parent, b) \cpResp{} channel to transmit responses from the
-child to the parent, and c) \pc{} channel to transmit both request and
-response messages from parent to child.
+Each cache is connected to its parent cache using 3 logical FIFO connections:
+a) \cpReq{} channel to transmit requests from the child to the parent, b)
+\cpResp{} channel to transmit responses from the child to the parent, and c)
+\pc{} channel to transmit both request and response messages from parent to
+child.
 
-Since we have dedicated channel between each parent-child pair, we
-will index the three channels (\cpReq{}, \cpResp{} and \pc{}) by the
-name of the child that uses that channel (using the notation
-$\cpReq[c]$, \etc for the \cpReq{} channel associated with cache $c$).
+We have dedicated channels between each parent-child pair. The three channels
+are indexed by the name of the child that uses that channel (denoted by
+$\textit{ch}[c]$ for the \textit{ch} channel associated with cache $c$).
 Figure \ref{format} shows the message format transmitted in the 3
@@
 
-In addition to the actual coherence states, each cache also maintains
-a temporary \emph{wait state} for each address to deal with
-transitioning between permission levels. The wait state consists of
-two parts: a) a boolean representing whether the cache has sent a
-request to its parent cache earlier and is waiting for a response
-(denoted by $c.\wait[a]$ for address $a$) and b) one of the three
-values \Mo, \Sh{} or \In representing the state upgrade that the cache
-is waiting to make (denoted by $c.\waitS[a]$ for address
-$a$). Similarly, each non-leaf cache has the corresponding wait states
-for its directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting
-for a response from its child $c$ for address $a$, and
-$p.\dwaitS[c][a]$ denotes the state to which $p$ is waiting
-for $c$ to downgrade. Both \wait{} and \dwait{} are \False{} initially
-for each of the caches, and each of the addresses, because no cache is
-waiting for any response from either its child or its parent
-initially.
+In addition to the actual permission states, each cache also maintains a
+temporary \emph{wait state} for each address to deal with transitioning between
+permission levels. The wait state consists of two parts: a) a boolean
+representing whether the cache has sent a request to its parent cache earlier
+and is waiting for a response (denoted by $c.\wait[a]$ for address $a$) and b)
+one of the three values \Mo, \Sh{} or \In{} representing the state upgrade that
+the cache is waiting to make (denoted by $c.\waitS[a]$ for address $a$).
+Similarly, each non-leaf cache has the corresponding wait states for its
+directory.  $p.\dwait[c][a]$ denotes that cache $p$ is waiting for a response
+from its child $c$ for address $a$, and $p.\dwaitS[c][a]$ denotes the state to
+which $p$ is waiting for $c$ to downgrade. Both \wait{} and \dwait{} are
+\False{} initially for each of the caches, and for each of the addresses,
+because no cache is waiting for any response from either its child or its
+parent initially.
 
-The behavior of the system is described via the atomic state
-transitions. These transitions have two parts: the guard dictating
-when the transition can fire, and the action defining how the state of
-the system changes when the transition happens. In our generated
-implementation, these transitions or \emph{guarded atomic actions}
-happen concurrently in a single hardware clock, but they will
-logically happen happen one at a time. Figure~\ref{trans} gives the
-atomic transitions for an MSI-based cache coherent memory subsystem. The actions
-in the transitions represent only the specific system states that
-change. We use $x \Leftarrow y$ to indicate setting of a specific
-state $x$ of the overall system to value $y$ because of a transition.
+The behavior of the system is described via the atomic state transitions. These
+transitions have two parts: a) the guard dictating when the transition can
+fire, and b) the action defining how the state of the system changes when the
+transition happens. In the actual implementation, several of these transitions
+or \emph{guarded atomic actions} happen concurrently in a single hardware
+clock, but they will logically happen happen one at a time. Figure~\ref{trans}
+gives the atomic transitions for the MIS protocol.
+The actions in the transitions in Figure \ref{trans} show only the specific
+states that change, the rest of the states remain as is. The notation ($x
+\Leftarrow y$;) in a transition indicates setting of a specific state $x$ of
+the overall system to value $y$.
 
 Figure~\ref{childside} shows the transitions that result in an upgrade of the
-state of a cache. It first involves a cache sending an upgrade request to the
-its parent when the cache is not already waiting for a response for that
-address from the parent (ChildSendReq). This request is received by the parent
-(ParentRecvReq), and if the rest of the parent's children caches are
-\emph{compatible}, and the parents state also has the permission being
-requested, it sends an upgrade response to the cache, while changing the
-appropriate directory state. Compatibility of an upgrade request with respect
-to a cache's siblings can be defined as follows:
+state of a cache. It involves a cache sending an upgrade request to the its
+parent if the cache is not already waiting for some response for that address
+from the parent (ChildSendReq). This request is received by the parent
+(ParentRecvReq). If the directory entries for all of the other children are
+\emph{compatible} with the upgrade request, and parent's permission state has
+at least the requested value, it sends an upgrade response to the cache, while
+changing the appropriate directory state. The notation \compat{} used in
+ParentRecvReq transition of Figure \ref{trans} is defined as follows:
 \begin{multline*}
@@
 Finally, the child cache receives the response from the parent (ChildRecvResp)
-and sets its state while resetting its wait state if the desired upgrade is
-indicated by the response.
+and sets its state while resetting its wait state if the response contains the
+desired upgrade.
 
-Figure~\ref{parentside} shows the transitions that result in a downgrade of the
-state of a cache because of a request by the cache's parent when the parent is
-not already waiting for a response from that cache (ParentSendReq).  This
-request is received by the child (ChildRecvReq), and if the state of each of
-the child's own children is not greater than the requested downgrade, the cache
-downgrades its state and sends a notification response to the parent. Finally,
-the parent receives the response from the cache (ParentRecvResp) and sets the
-appropriate directory state while resetting the appropriate directory wait state
-if the desired downgrade is indicated by the response.
+Figure~\ref{parentside} shows the similar sequence of transitions that results
+in downgrading a cache's directory entry for one of its chilren $c$. The
+difference is in the guard for ChildRecvReq transition: the directory entry for
+each of the $c$'s children should not be greater than the requested
+downgrade, in order for $c$ to accept a request and send a downgrade
+response.
 
 Figure~\ref{childextra} shows transitions that happen in a non-root cache
-because of the presence of \emph{voluntary} responses. Whenever a cache runs
-out of space to store the data corresponding to an address, it evicts the data
-corresponding to one or more addresses to make room for the new address. This
-eviction causes the state of the cache for that address to go to $\In$, which
-has to be informed to its parent. We use the same \cpResp{} FIFO (and the same
-response format) to indicate this. Such a voluntary response for an address can
-take place only if the cache is not waiting for a response from its parent for
+because of the presence of \emph{voluntary} responses. This models eviction of
+a location from a cache to make room for a new location.  This eviction usually
+causes the permission state of the cache for that address to go to $\In$, while
+we allow any downgrade to take place voluntarily.  We use the same \cpResp{}
+FIFO for such a message. Such a voluntary response for an address can take
+place only if the cache is not waiting for a response from its parent for
 that address (ChildVolResp).
 
-Because of the presence of voluntary response, a child cache could have already
+Because of the presence of voluntary response, a child could have already
 downgraded to a state required by a downgrade request from its parent. In such
-a case, the cache simply drops the downgrade request (ChildDropReq).
+a case, the child simply drops the downgrade request (ChildDropReq).
 
-Figure~\ref{procside} shows the transitions that actually serve a processors
-request in the leaf caches. LoadReq and StoreReq show the processing of a load
-and store request, respectively, from a processor. In the case of a store
-request, the data in the cache for the requested address is updated with that
-supplied in the request. We do not explicitly show the response generated by
-the cache in case of a load request in our memory subsystem model; in a real
-system, the data in the cache for the requested address is sent back to the
-processor. Our definition of store atomicity requires responses for both load
-and store requests, and the response field contains the
-$(\textit{Processor}*\mathbb{N})$ pair identifying the request corresponding to
-the response, the data from the cache in case of a load response, and a value
-for the field \tme. A \Response{} is defined whenever LoadReq or StoreReq
-transition takes place. Let $c$ be a leaf cache, and $t$ be the position of a
-LoadReq$(c)$ or StoreReq$(c)$ transition in the ordered list of transitions of
-the full system starting from the initial state. The fields of the response are
-as follows:
+Figure~\ref{procside} shows the transitions that serve a processors request in
+the leaf caches. LoadReq and StoreReq show the processing of a load and store
+request, respectively. In the case of a store request, the data in the cache
+for the requested address is updated with that supplied in the request. We do
+not explicitly show the response generated by the cache in case of a load
+request in our memory subsystem model; in a real system, the data in the cache
+for the requested address is sent back to the processor. Our definition of
+store atomicity requires responses for both load and store requests, and the
+response field contains the $(\textit{Processor}*\mathbb{N})$ pair identifying
+the request corresponding to the response, the data from the cache in case of a
+load response, and a value for the field \tme. A \Response{} is defined
+whenever LoadReq or StoreReq transition takes place. Let $c$ be a leaf cache,
+and $t$ be the position of a LoadReq$(c)$ or StoreReq$(c)$ transition in the
+ordered list of transitions of the full system starting from the initial state.
+The fields of the response are as follows:
 \begin{multline*}
@@
 \hline
-\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildVolResp}$(c, x, a)$: Child $c$ sending a response to downgrade to $x$ for address $a$ without any request from its parent}\\
+\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildVolResp}$(c, x, a)$: Child $c$
+sending a response to downgrade to $x$ for address $a$ without any request from
+its parent}\\
 \hline
@@
 \hline
-\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildDropReq}$(c)$: Child $c$ receiving a request from its parent and has already downgraded}\\
+\multicolumn{2}{|p{\textwidth}|}{\textbf{ChildDropReq}$(c)$: Child $c$
+receiving a request from its parent and has already downgraded}\\
 \hline

### 2014-02-04 paper/Conclusion.tex

\section{Discussion}

We illustrated a direct realization of our proposed proof framework
for a very flexible cache coherence protocol implementation codifying
the intuition behind the system's correctness as simple
invariants. While this was took significant expertise, once done, we
were able to show that many key hardware modifications,\eg\ caching
policies, could be made with zero or very small natural changes to the
proof. These bits were small enough that an informed hardware designer
implementing our protocols can reasonably be expected to provide such
information during refinement giving them a viable path for producing
a provably correct performant cache coherence engine implementation.

The protocol framework we showed was very flexible, and could easily
be extended for variants that go beyond microarchitectural variants,
\eg\ silent drops, extension to MOSI with a relatively small amount of
additional reasoning.  This gives us good confidence that we can
effectively produce frameworks for other classes protocol,
\eg\ update-based protocols, non-inclusive cache hierarchies. Further,
nothing about our methodology requires we limit ourselves to cache
coherence systems; this approach should naturally extend to any
hardware-based system.

While fundamentally this work meets our goal of carrying verification
into through the design refinement step, there are still some
important mechanization to make this ready for industrial
use. Foremost, the simple transliteration between the Coq transition
system and the Bluespec hardware description implementation should be
fully mechanized. Further, the proposed extension to provide fairness
of transaction scheduling in the Bluespec compiler should be
implemented. Finally, to fully guarantee correctness of the resulting
hardware FSM, we should also fully verify the hardware generation as
in the FeSi~\cite{Tomas} efforts.

### 2014-02-04 paper/Deadlock.tex

@@
 
-In Section \ref{safety} we showed how the MSI protocol of Figure \ref{trans}
-obeys the Store atomicity theorem. In addition to this criteria, we must
-establish that the system makes forward progress.
+In Section~\ref{safety} we showed how the MSI protocol of
+Figure~\ref{trans} obeys the store atomicity theorem, showing that
+only correct behavior occurs. To complete our correctness picture, we
+must also show that we avoid deadlock, \ie\ we make forward progress.
 
@@
 \begin{enumerate}
-\item Weak forward progress: The system never reaches a state where none of the transitions of Figure
-\ref{trans} apply or
-\item Strong forward progress: At any time, eventually a new request from some processor will be
-processed
+\item Weak forward progress: The system never reaches a state where
+  none of the transitions of Figure~\ref{trans} apply or
+\item Strong forward progress: At any time, eventually a new request
+  from some processor will be processed
 \end{enumerate}
 
-In order to prove strong forward progress, we need to
-establish some more conditions on the strategy to execute the transitions of
-Figure \ref{trans} in each cache -- the caches must ensure local
-starvation-freedom in the processing of messages when sufficient resources are
-available. We do not discuss this issue in this paper and instead focus on 
-weak forward progress.
+To prove strong forward progress given weak fairness, we need some
+guarantees on the choice of executions of the transitions
+Figure~\ref{trans} in each cache. Specifically, caches must ensure
+local starvation-freedom in the processing of messages when sufficient
+resources are available. This can be enforced via during hardware
+compilation; for this paper we will focus only on weak forward
+progress.
 

### 2014-02-04 paper/Introduction.tex
intuition

@@
 \etal{}~\cite{DNA:CoherenceImplementation} that such a verified
-protocol could be transliterated into Bluespec, a hardware description
-language based on Guarded Atomic Actions, a restriction of
-TRS~\cite{Hoe:TCAD}. This reduces the correctness of the
-implementation to the correctness of the protocol, effectively
+protocol could be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
+hardware description language based on Guarded Atomic Actions, a
+restriction of TRS~\cite{Hoe:TCAD}. This reduces the correctness of
+the implementation to the correctness of the protocol, effectively
 removing any implementation bugs. While the current Bluespec compiler
@@
 framework. We discuss our MSI protocol example and its implementation
-in Section~\ref{Sec:Protocol}. In Section~\ref{Sec:ProofStructure} we discuss
+in Section~\ref{Sec:System}. In Section~\ref{Sec:ProofStructure} we discuss
 the structuring of our proof framework and verification of the base

### 2014-02-04 paper/System.tex
some modifications of background

@@
 fire, and b) the action defining how the state of the system changes when the
-transition happens. In the actual implementation, several of these transitions
-or \emph{guarded atomic actions} happen concurrently in a single hardware
-clock, but they will logically happen happen one at a time. Figure~\ref{trans}
-gives the atomic transitions for the MIS protocol.
-The actions in the transitions in Figure \ref{trans} show only the specific
-states that change, the rest of the states remain as is. The notation ($x
-\Leftarrow y$;) in a transition indicates setting of a specific state $x$ of
-the overall system to value $y$.
+transition happens. Logically, these transitions happen one at a time.
+Figure~\ref{trans} gives the atomic transitions for the MIS protocol.  The
+actions in the transitions in Figure \ref{trans} show only the specific states
+that change, the rest of the states remain as is. The notation ($x \Leftarrow
+y$;) in a transition indicates setting of a specific state $x$ of the overall
+system to value $y$.
 
@@
 not explicitly show the response generated by the cache in case of a load
-request in our memory subsystem model; in a real system, the data in the cache
-for the requested address is sent back to the processor. Our definition of
-store atomicity requires responses for both load and store requests, and the
-response field contains the $(\textit{Processor}*\mathbb{N})$ pair identifying
-the request corresponding to the response, the data from the cache in case of a
-load response, and a value for the field \tme. A \Response{} is defined
-whenever LoadReq or StoreReq transition takes place. Let $c$ be a leaf cache,
-and $t$ be the position of a LoadReq$(c)$ or StoreReq$(c)$ transition in the
-ordered list of transitions of the full system starting from the initial state.
-The fields of the response are as follows:
-\begin{multline*}
-\small\langle (c, c.\pos), t, \mylet q := \reqFn(c, c.\pos) \myin\\
-\small\myif q.\desc = \Ld \mythen c.\data[q.\addr] \myelse \_ \rangle
-\end{multline*}
+request; in a real system, the data in the cache for the requested address is
+sent back to the processor. When a LoadReq or StoreReq happens at a cache $c$
+pertaining to address $a$, the \Response{}, needed in the definition of store
+atomicity, is generated as follows: the \labelR{} field contains $(c, c.\pos)$
+pair, the \dataR{} field contains the value $c.\data[a]$, and \timeR{} field
+contains the number of transitions that has taken place in the system before
+the current transition, intuitively denoting the time at which this transition
+takes place.
 
@@
 amenable to single cycle implementations in hardware. For example, writing a
-register array (like a cache) can actually take several hardware clock cycles,
-though logically it is a single transition.  Karczmarek \etal~\cite{Karczmarek}
-has provided a scheme to convert atomic transitions into synchronous hardware
-in which each transition can span several clock cycles.
+register array can actually take several hardware clock cycles, though
+logically it is a single transition.  Karczmarek \etal~\cite{Karczmarek} has
+provided a scheme to convert atomic transitions into synchronous hardware in
+which each transition can span several clock cycles.

### 2014-02-04 paper/Background.tex
some modifications of backgrounda

@@
 All modern systems have caches between the processor and memory for faster
-memory access.  It is often also useful to permit some reorderings of requests
+memory access.  It is often also useful to permit some re-orderings of requests
 from a processor to the memory to improve performance.  For instance, in the
@@
 bypass a subsequent load request to a different location. The specification of
-which reorderings are permitted or restricted defines the memory consistency
+which re-orderings are permitted or restricted defines the memory consistency
 model.

### 2014-02-04 paper/Murali.tex
some modifications of backgrounda

\section{Intuition}

In this section we give the proof of the MSI cache coherence protocol. We give
the proof in a top-down fashion starting with the store atomicity theorem. We
specify important invariants and show how they are proved in the MSI protocol.
In Section \ref{Refinements} show how these invariants hold in other protocols
like MESI and MOSI. These invariants are true at each step in the (infinite)
list of transitions starting from the initial state. We use the term time $t$
to specify the global state reached after $t$ transitions.  All the invariants
below pertain to a particular address; we will implicitly assume a universal
quantification over all address as a prefix to each of the invariants (if not
explicitly mentioned).

\subsection{Top invariants}
\begin{defn}
\textit{clean}: A cache is said to be \clean{} for an address iff the state of the
cache is at least \Sh{} and the state of all its children is at most \Sh.
\end{defn}
\begin{inv}
At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
up-to-date value for that address, \ie
\begin{enumerate}
\item $c.\data[a] = \initData(a)$ and no store request has been processed at
any time $< t$, or
\item There is a store request $q$ that was processed at time $t'$ such that
$t' < t$ and $a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
request $q'$ was processed at any time $t_i$ such that $t' < t_i < t \wedge a =
q'.\addrQ$.
\end{enumerate}
\label{latestValue}
\end{inv}

The following invariants hold for leaf caches.
\begin{inv}
\textit{uniqM}: At all times, at most one leaf cache can be in \Mo{} state.
\label{uniqM}
\end{inv}

\begin{inv}
\textit{processReq}: At all times, a load or a store request at a leaf cache can be processed
only if it is at least in \Sh{} or \Mo{} state at that time, respectively.
\label{processReq}
\end{inv}

\begin{inv}
\textit{noReqAgain}: A leaf cache $c$ processes the $c.\pos^{th}$ request iff
$c.\pos$ changes. $c.\pos$ only gets incremented when it changes.
\label{noReqAgain}
\end{inv}

Using Invariants \ref{latestValue} to \ref{noReqAgain}, it is easy to see that
Theorem \textit{storeAtomicity} can be proved. Since a store can happen only
when the state of a leaf cache is in \Mo{} state (Invariant
\textit{processReq}), because of Invariant \textit{uniqM} only one store
request can be processed at a time throughout the system, thereby satisfying
\textit{uniqStTimes}. Because of Invariant \textit{noReqAgain}, once a response
has been produced for a request $i$, the cache can only process requests $j, j
> i$. Thus there is at most one response to each request, satisfying
\textit{uniqRespLabels}. A load request can be processed only if a cache is at
least in \Sh{} state according to Invariant \textit{processReq}.  When a leaf
cache is at least in \Sh{} state, since it has no children, it is \clean{}.
Thus, by Invariant \textit{lastestValue}, the final condition for Theorem
\textit{storeAtomicity} is satisfied.

Note that Invariants \textit{processReq} and \textit{noReqAgain} can be proved
directly from the atomic transitions describing the system (Figure \ref{trans}
for MSI; other protocols must also have similar transitions). Proving Invariants
\textit{uniqM} and \textit{latestValue} requires the following invariants.

\begin{inv}
\textit{conservative}: At all times, the state of a cache is not greater than
its parent's directory corresponding to the cache.
\label{conservative}
\end{inv}

\begin{inv}
\textit{stateCompat}: At all times, the state of a cache is not less than its
directory corresponding to any of its children.
\label{stateCompat}
\end{inv}

We use the definition of compatibility between two caches as follows:
\begin{defn}
\textit{compatibile}: At some time, cache $c_1$ is compatible with $c_2$ iff
at that time,
\begin{enumerate}
\item $c_1.\state[a] = \In$ whenever $c_2.\state[a] = \Mo$ and
\item $c_1.\state[a] \le \Sh$ whenever $c_2.\state[a] = \Sh$
\end{enumerate}
\label{compatible}
\end{defn}

\begin{inv}
\textit{siblingCompat}: At all times, two sibling caches (\ie caches with the
same parent) are compatible.
\label{siblingCompat}
\end{inv}

\begin{inv}
\textit{noTransitWrite}: When a cache $c$ transitions from non-clean at time
$t$ to clean at time $t+1$ for an address $a$, then it receives data $a$ from
some cache $s$ such that the $s$ is clean at the time $t'$ for address $a$, and
no store request was processed for $a$ at time $t_i$ such that $t' \le t_i \le
t$.
\label{noTransitWrite}
\end{inv}

For any two caches $c$ and $p$, such that $c$ is a descendent of $p$ in the cache
hierarchy (the descendent relation between two caches is the transitive
reflexive closure of the parent relation), from Invariants
\textit{conservative} and \textit{stateCompat}, it is clear that for any
address $a$, $c.\state[a] \le p.\state[a]$. Any two caches $c_1, c_2$ in which
neither is the decendent of the other will have a lowest common ancestor $x$
such that both the caches are descendents of $x$ and $c_1, c_2$ are descendents
of distinct caches $x_1, x_2$, respectively, such that $x_1, x_2$ are children
of $x$. $x_1$ and $x_2$ are compatible because of Invariant
\textit{siblingCompat}. $c_1.\state[a] \le x_1.\state[a]$ and
$c_2.\state[a] \le x_2.\state[a]$ because of Invariants \textit{conservative}
and \textit{stateCompat}. From this, it can easily be seen from the definition
of compatible that $c_1$ and $c_2$ are compatible whenever $x_1$ and $x_2$ are
compatible. From the definition of compatible it can be seen that both caches can not
be in the \Mo{} state for them to be compatible. This directly proves Invariant
\textit{uniqM}.

Invariant \textit{latestValue} can be proved using induction on time. Let us
assume that all caches which are clean at time $t'$ such that $t'<t+1$ have the
latest value. If a cache $c$ is clean at time $t$, then none of its descendents
are in \Mo, and hence none of its descendents can process a store request at
$t$. Any non descendent is compatible with it (as shown in the previous
paragraph) and so no non-descendent cache can be in \Mo, since a clean cache is
at least in \Sh. So only $c$ can process a store request at $t$ (assuming $c$
is a leaf cache and is in state \Mo) when it is clean, and no other cache can
process a store request. By induction hypothesis, it contained the latest value
at time $t$ which remains true (because of no processing of store request) at
time $t+1$. Now, if a cache was not clean at time $t$, then it would have
received the data from some cache which was clean at the time of sending the
data (by Invariant \textit{noTransitWrite}), which means the data is the latest
value by induction hypothesis, so the new data it receives at time $t$ is the
latest value, making the clean cache at time $t+1$ have the latest value.

Invariant \textit{conservative} can be proved by
further breaking them into the following invariants between a parent-child pair
of caches.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A child sends a response to its parent iff it downgrades its state.
\item A parent sends a response to a child iff it upgrades the directory of
that child.
\item The state and the directory can change only on sending and receiving a
response; the receiving entity changes its value to that of the sending entity
immediately after sending the response message.
\end{enumerate}
\label{stateChange}
\end{inv}

\begin{inv}
\textit{respFIFO}: Responses are received in the order in which they are sent,
for a source-destination pair, for the same address.
\label{respFIFO}
\end{inv}

\begin{inv}
\textit{noCross}: A response from a cache to its child for an address can not
be in transit while another response from the same child to the parent for the
same address is in transit.
\label{noCross}
\end{inv}

Invariant \textit{conservative} at any time $t$ for child cache $c$ and parent
$p$ for an address $a$ can be proved as follows. Consider the case when
$c.\state[a]$ never changed before $t$, then $c.\state[a] = \In$ because of the
initialization conditions described in Section \ref{System}. No response would
have been sent by $c$ to its parent for address $a$ before $t$, otherwise
$c.\state[a]$ would have changed. Thus the parent would not have received any
response from $c$, $p.\dstate[c][a]$ could have changed only by sending a
response to $c$ which can only upgrade it from the initial value of $\In$. Thus
at $t$, $p.\dstate[c][a] \ge c.state[a]$.

Now consider the case when $c.\state[a]$ last changed at $t'$ such that $t' <
t$ because of receiving a response from $p$, sent at $t''$. $c.\state[a]$ at
$t'+1 = p.\dstate[c][a]$ at $t''+1$ because of Invariant \textit{stateChange}.
$c$ neiter sends nor receives a response at any $t_i$ such that $t' < t_i < t$.
If $p$ received a response at some $t_m$ such that $t'' < t_m < t$,
then it would either violate Invariant \textit{noCross}, or the fact that no
message was sent between $t' < t_i < t$ from $c$. Thus, $p.\dstate[c][a]$ at $t
\ge p.\dstate[c][a]$ at $t''+1$ because any message sent can only upgrade
$p.\dstate[a]$. This ensures that Invariant \textit{conservative} is obeyed. We
can similarly prove this invariant holds when $c.\state[a]$ last changes before
$t$ by sending a response.

Invariant \textit{stateChange} can be proved directly from the atomic
transitions for MSI. Other protocols have similar atomic transitions obeying
these invariants.

Invariant \textit{respFIFO} can be enforced by the interconnect network which
ensures a FIFO ordering for messages sent between a pair of nodes.

---------------------------------------------

Proving Invariant \textit{noCross} takes us further down the rabbit
hole; it requires the following invariants.

\begin{inv}
\textit{parentRespOnReq}: Parent sends a response to a child $c$ for address
$a$ only on receiving a request from $c$ for address $a$.
\label{parentRespOnReq}
\end{inv}

\begin{inv}
\textit{parentAnyFIFO}: Any message (request or response) sent from a parent
$p$ to its child $c$ for an address does not overtake another message (request
or response) sent earlier from $p$ to $c$ for the same address.
\label{parentRespReqFIFO}
\end{inv}

\begin{inv}
\textit{childWaitNoVoluntaryResp}: A child can send a response to its parent
without getting a request from the parent only if it is not in the wait state
(\ie waiting for a response from the parent for an upgrade request sent
earlier).
\label{childWaitNoVoluntaryResp}
\end{inv}

\begin{inv}
\textit{parentWaitNoResp}: A parent can send a response to its child
only if it is not in the wait state (\ie waiting for a response from the child
for a downgrade request sent earlier).
\label{parentWaitNoResp}
\end{inv}

\begin{inv}
\textit{waitNoReq}:
\begin{enumerate}
\item A child $c$ can not send a request to its parent for an address $a$ while
$c.\wait[a]$ is true.
\item A parent $p$ can not send a request to its child $c$ for an address $a$
while $p.\dwait[c][a]$ is true.
\end{enumerate}
\end{inv}

\begin{inv}
\textit{waitState}:
\begin{enumerate}
\item If a request to upgrade is sent from child $c$ to its parent for address
$a$, $c.\wait[a]$ is set to true till it receives the correct upgrade for that
address from $p$.
\item If a request to downgrade is sent from parent $p$ to its child $c$ for address
$a$, $p.\dwait[c][a]$ is set to true till it receives the correct downgrade for that
address from $c$.
\end{enumerate}
\label{waitState}
\end{inv}

\begin{inv}
\textit{childReqRespPair}: If a child $c$ sends a request $r$ to its parent $p$
at time $t$, $c$ will not send any request to $p$ on or before receiving a response
from the $p$ sent immediately on receiving request $r$.
\label{childReqRespPair}
\end{inv}

\begin{inv}
\textit{childRespReqFIFO}: A request sent from a child $c$ to the parent for an
address does not overtake a response sent earlier from $c$ to the parent for the
same address.
\label{childRespReqFIFO}
same address.
\end{inv}

The request-response mechanism in the MSI protocol is basically set up to
ensure Invariant \textit{noCross}. Other protocols also have a similar
request-response mechanism. Invariant \textit{noCross} can be proved by
contradiction as follows.

Consider a time $t$ when there is both a response from the parent $p$ to a
child $c$ and from $c$ to $p$ the same address $a$ in transit. Let the response
from $p$ to $c$ be sent at time $t_1$ and the response from $c$ to $p$ be sent
at $t_2$. The response from $p$ to $c$ must have been triggered because of a
request from the $c$ (by Invariant \textit{parentRespOnReq}). Let this request
be sent at time $t_3$ from $c$. Consider the case when $c.\wait[a] =
\False$ at $t_2$. Then it must be that $c$ received a response from $p$ between
$t_3$ and $t_2$. Let the earlier such responses be received at $t_4$ such that
$t_3 \le t_4 < t2$. Let this response be triggered by a request to $p$ from $c$
sent at $t_5$. Because of Invariant \textit{waitNoReq}, this request must have
been sent at time $t_6$ such that $t_6 < t_3$, but this would contradict
Invariant \textit{childReqRespPair}. Thus, $c.\wait[a] = \True$ at $t_2$.
This means that

If By Invariant \textit{childReqRespPair}, $c$
will not have

-----------------------------------------------------------------------------------------------------------------------------

%Invariant \textit{respFIFO} is known as point-to-point FIFO ordering in the
%Network-on-Chip NoC community, and is enforced by common routing algorithms
%like Dimension-ordered routing \cite{dimension}. Only the parent-to-child
%channels need to be FIFO ordered, the FIFO ordering in child-to-parent channels
%can be enforced using the protocol itself.

\begin{inv}
\textit{noTransitEqual}: When no response is in transit for an address, then
the state of a cache for that address is the same as its parent's directory for
that cache and address.
\label{noTransitEqual}
\end{inv}

Proving Invariants \textit{inTransitCondition} and \textit{noTransitEqual},
needs two further invariants, along with Invariant \textit{stateChange}.

Invariant \textit{inTransitCondition} can be proved as follows. Whenever a
response from a child to the parent is in transit, the parent can not send any
response to the child, or it will violate Invariant \textit{noCross}. Using the fact that

\begin{inv}
\textit{inTransitCondition}:
\begin{enumerate}
\item When a response is sent from a child $c$ at time $t$ for an address $a$,
and has not been received by the parent $p$ before time $t'$, $\forall t_i, t <
t_i \le t' @-> c.\state[a]$ at $t_i \le c.\state[a]$ at $t+1$ and $\forall t_i,
t \le t_i \le t' @->$ $p.\dstate[c][a]$ at $t_i \ge c.\state[a]$ at $t$.
\item When a response is sent from parent $p$ at time $t$ for an address $a$,
and has not been received by its child $c$ before time $t'$, $\forall t_i, t <
t_i \le t' @-> p.\dstate[c][a]$ at $t_i \ge p.\dstate[c][a]$ at $t+1$ and
$\forall t_i, t \le t_i \le t' @->$ $c.\state[a]$ at $t_i \le p.\dstate[c][a]$ at
$t$.
\end{enumerate}
\label{inTransitCondition}
\end{inv}

\begin{inv}
\textit{prevState}: When a response is received from a child to the parent
(parent to child) for an address, the state of the child for that address at
the time of sending (receiving) the response is the same as the parent's
directory for that child and address at the time of receiving (sending) the
response.
\label{prevState}
\end{inv}

%Top
%a) conservative: forall p c a t, parent c p -> dir p c a t >= state c a t
%b) forall p c a t, parent c p -> state p a t >= dir p c a t
%c) forall c1 c2 a t, sibling c1 c2 -> compatible c1 c2 a t
%d) nextClean: ~ clean c a t -> clean c a (S t) -> exists s d ts, recv data s a t d /\ send data s a ts d /\ clean s a ts /\ ts <= t /\ forall ti, ts <= ti < t ->  ~ write a ti
%
%(from=state(t1)=dir(t2) -> betweenPSendCRecv:dir(t)>=to&state(t)<=from, betweenCSendPRecv:state(t)<=to&dir(t)>=from), noMesgImp:state(t)=dir(t) -> conservative (ignore from=, use only a, b)
%a) Child sends message iff it downgrades
%b) Parent sends message iff it upgrades
%c) State is changed only on sending or receiving a message
%d) FIFO messages
%e) cross: Message from child and parent do not cross
%
%cross:
%a) parent send message only on receiving request
%b) req from child does not overtake response from child (prove using the from states)
%c) resp from parent does not overtake request from parent
%d) voluntary response sent from child only when not waiting
%
%In this section we specify a set of important invariants for a parent-child pair of caches.

### 2014-02-04 paper/Conclusion.tex
some modifications of backgrounda

@@
-\section{Discussion}
+\section{Conclusions and future work}
 
-We illustrated a direct realization of our proposed proof framework
-for a very flexible cache coherence protocol implementation codifying
-the intuition behind the system's correctness as simple
-invariants. While this was took significant expertise, once done, we
-were able to show that many key hardware modifications,\eg\ caching
-policies, could be made with zero or very small natural changes to the
-proof. These bits were small enough that an informed hardware designer
-implementing our protocols can reasonably be expected to provide such
-information during refinement giving them a viable path for producing
-a provably correct performant cache coherence engine implementation.
+We illustrated a direct realization of our proposed proof framework for a
+concrete MSI cache coherence protocol implementation codifying the intuition
+behind the system's correctness as simple invariants. While this took
+significant effort, once done, we showed that many key hardware
+modifications,\eg{} replacement policies, could be made with a very small
+addition to the proof. These bits were small enough that an informed hardware
+designer implementing our protocols can reasonably be expected to provide such
+additional proofs during refinement, producing a provably correct realistic
+cache coherence engine implementation.
 
-The protocol framework we showed was very flexible, and could easily
-be extended for variants that go beyond microarchitectural variants,
-\eg\ silent drops, extension to MOSI with a relatively small amount of
-additional reasoning.  This gives us good confidence that we can
-effectively produce frameworks for other classes protocol,
-\eg\ update-based protocols, non-inclusive cache hierarchies. Further,
-nothing about our methodology requires we limit ourselves to cache
-coherence systems; this approach should naturally extend to any
+The protocol framework we showed was very flexible, and could easily be
+extended for variants that go beyond micro-architectural variants, \eg{}
+extension to MOSI with a relatively small amount of additional reasoning. This
+gives us good confidence that we can effectively produce frameworks for other
+classes of protocols, \eg{} update-based protocols, non-inclusive cache
+hierarchies. Further, nothing about our methodology requires we limit ourselves
+to cache coherence systems; this approach should naturally extend to any
 hardware-based system.
 
-While fundamentally this work meets our goal of carrying verification
-into through the design refinement step, there are still some
-important mechanization to make this ready for industrial
-use. Foremost, the simple transliteration between the Coq transition
-system and the Bluespec hardware description implementation should be
-fully mechanized. Further, the proposed extension to provide fairness
-of transaction scheduling in the Bluespec compiler should be
-implemented. Finally, to fully guarantee correctness of the resulting
-hardware FSM, we should also fully verify the hardware generation as
-in the FeSi~\cite{Tomas} efforts.
+While fundamentally this work meets our goal of verification of synthesizable
+hardware specifications, there is still some important mechanization to make
+this ready for industrial use. Foremost, the simple transliteration between the
+inductively defined transition system in Coq and the Bluespec hardware
+description should be fully mechanized.
+%Further, the proposed extension to provide fairness of transaction scheduling
+%in the Bluespec compiler should be implemented.
+Further, to fully guarantee correctness of the resulting hardware, we should
+also verify the hardware synthesis compiler, possibly using FeSi~\cite{Tomas}.

### 2014-02-04 paper/Intuition.tex
some modifications of backgrounda

@@
 For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
-the cache tree hierarchy, from Invariants \textit{conservative} and
+the cache tree hierarchy, from Invariants \fullref{conservative} and
 \textit{localCompatibility}, we can show that for any address $a$,
@@
 relation is the reflective-transitive closure of the parent relation,
-and Invariants \textit{conservative} and the first part
+and Invariants \fullref{conservative} and the first part
 \textit{localCompatibility} of defines a $\ge$ relation between the
@@
 that $c_1$ and $c_2$ are also compatible. This proves
-Invariant \textit{nonAncestorCompatibility}.
+Invariant \fullref{nonAncestorCompatibility}.
 
-We prove Invariant \textit{latestValue} by strong induction on
+We prove Invariant \fullref{latestValue} by strong induction on
 time. Consider two cases: cache $c$ is already \clean{} for an address
@@
 
-To prove Invariant \textit{conservative} at some time $t$, consider the case
+To prove Invariant \fullref{conservative} at some time $t$, consider the case
 when a child cache $c$ has sent to or received from the parent $p$ some message
@@
 $t'$. Let's say $m$ is sent by $c$. Then $p$ can not have received any message
-from $c$ after sending $m$ in order to not violate Invariant \textit{noCross}.
+from $c$ after sending $m$ in order to not violate Invariant \fullref{noCross}.
 Instead, if $c$ receives $m$ at $t'$, $p$ will not receive any message after
-sending $m$ in order to not violate Invariants \textit{noCross} and
+sending $m$ in order to not violate Invariants \fullref{noCross} and
 \textit{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
@@
 $t$, it can only upgrade $p.\dstate[c][a]$. Since $c.\state[a]$ does not change
-between $t'+1$ and $t$, Invariant \textit{conservative} holds at $t$. If no
+between $t'+1$ and $t$, Invariant \fullref{conservative} holds at $t$. If no
 message has been sent or received by $c$ before $t$, then $p$
@@
 
-To prove Invariant \textit{localCompatibility}, it is enough to show that a
+To prove Invariant \fullref{localCompatibility}, it is enough to show that a
 child $c$ upgrades $c.\state[a]$ on receiving a response for $a$ from its
 parent $p$ and $p.\dstate[c][a]$ is downgraded on receiving a response for $a$
-from $c$. If this is true, then Invariant \textit{localCompatibility} will not
+from $c$. If this is true, then Invariant \fullref{localCompatibility} will not
 be violated on receiving a response. The transitions of Figure \ref{trans}
-ensures that Invariant \textit{localCompatibility} is not violated when sending
+ensures that Invariant \fullref{localCompatibility} is not violated when sending
 a response.
@@
 The proof for this follows exactly the same arguments presented above to prove
-Invariant \textit{conservative} -- we prove by considering the last message
+Invariant \fullref{conservative} -- we prove by considering the last message
 sent or received at the child before $t$ (and the case when no message has been
@@
 
-Invariant \textit{noTransitWrite} can be proved as follows. Consider a response
+Invariant \fullref{noTransitWrite} can be proved as follows. Consider a response
 sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
 other response can be received by the parent at time $t_p$ such that $t+1 \le
-t_p \le t'$ to not violate Invariant \textit{noCross}, so $p.\dstate[c][a]$ at
+t_p \le t'$ to not violate Invariant \fullref{noCross}, so $p.\dstate[c][a]$ at
 $t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
@@
 child upgrades its state at $t'$. Using arguments similar to the proof of
-Invariant \textit{latestValue}, we can establish that no cache can be in \Mo{}
+Invariant \fullref{latestValue}, we can establish that no cache can be in \Mo{}
 state for address $a$ when a response is in transit from $p$ to $c$.
@@
 %t$ because of receiving a response from $p$, sent at $t''$. $c.\state[a]$ at
-%$t'+1 = p.\dstate[c][a]$ at $t''+1$ because of Invariant \textit{stateChange}.
+%$t'+1 = p.\dstate[c][a]$ at $t''+1$ because of Invariant \fullref{stateChange}.
 %$c$ neiter sends nor receives a response at any $t_i$ such that $t' < t_i < t$.
 %If $p$ received a response at some $t_m$ such that $t'' < t_m < t$,
-%then it would either violate Invariant \textit{noCross}, or the fact that no
+%then it would either violate Invariant \fullref{noCross}, or the fact that no
 %message was sent between $t' < t_i < t$ from $c$. Thus, $p.\dstate[c][a]$ at $t
 %\ge p.\dstate[c][a]$ at $t''+1$ because any message sent can only upgrade
-%$p.\dstate[a]$. This ensures that Invariant \textit{conservative} is obeyed. We
+%$p.\dstate[a]$. This ensures that Invariant \fullref{conservative} is obeyed. We
 %can similarly prove this invariant holds when $c.\state[a]$ last changes before
@@
 %
-%Invariant \textit{stateChange} can be proved directly from the atomic
+%Invariant \fullref{stateChange} can be proved directly from the atomic
 %transitions for MSI. Other protocols have similar atomic transitions obeying
@@
 %
-%Invariant \textit{respFIFO} can be enforced by the interconnect network which
+%Invariant \fullref{respFIFO} can be enforced by the interconnect network which
 %ensures a FIFO ordering for messages sent between a pair of nodes.
@@
 %
-%%Invariant \textit{respFIFO} is known as point-to-point FIFO ordering in the
+%%Invariant \fullref{respFIFO} is known as point-to-point FIFO ordering in the
 %%Network-on-Chip NoC community, and is enforced by common routing algorithms
@@
 %
-%Proving Invariants \textit{inTransitCondition} and \textit{noTransitEqual},
-%needs two further invariants, along with Invariant \textit{stateChange}.
+%Proving Invariants \fullref{inTransitCondition} and \textit{noTransitEqual},
+%needs two further invariants, along with Invariant \fullref{stateChange}.
 %
-%Invariant \textit{inTransitCondition} can be proved as follows. Whenever a
+%Invariant \fullref{inTransitCondition} can be proved as follows. Whenever a
 %response from a child to the parent is in transit, the parent can not send any
-%response to the child, or it will violate Invariant \textit{noCross}. Using the fact that 
+%response to the child, or it will violate Invariant \fullref{noCross}. Using the fact that 
 %

### 2014-02-04 paper/Intuition.tex
some modifications of backgrounda

@@
 Given these invariants, we codify our intuition by showing that
-Theorem~\textit{storeAtomicity} follows naturally from these
+Theorem~\fullref{storeAtomicity} follows naturally from these
 invariants.  A store may happen only when leaf cache is in \Mo{} state
-for that address (Invariant~\textit{processReq}). Leaf caches cannot
+for that address (Invariant~\fullref{processReq}). Leaf caches cannot
 be ancestors of any other node, so by Invariant
-\textit{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
+\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
 other leaf caches will be in \In{} state preventing the possibility of
 load and store requests for the same address (by Invariant
-\textit{processReq}), showing Definition~\textit{uniqStTimes} holds.
+\fullref{processReq}), showing Definition~\fullref{uniqStTimes} holds.
 
-Given Invariant~\textit{noReqAgain}, once a response has been produced
+Given Invariant~\fullref{noReqAgain}, once a response has been produced
 for a request $i$, the cache may subsequently only process $j^{th}$
 request such that $j > i$. Thus there is at most one response to each
-request, satisfying Definition~\textit{uniqRespLabels}. A load request
+request, satisfying Definition~\fullref{uniqRespLabels}. A load request
 may be processed only if a cache is at least in \Sh{} state (Invariant
-\textit{processReq}). As leaf caches have no children, a leaf cache in
+\fullref{processReq}). As leaf caches have no children, a leaf cache in
 at least in \Sh{} state for address $a$ is \clean for that address. A
 load response supplies the data for that address present in the cache,
-which, by Invariant~\textit{lastestValue}, contains the ``latest
-value''. Thus, the final condition for Theorem~\textit{storeAtomicity}
+which, by Invariant~\fullref{lastestValue}, contains the ``latest
+value''. Thus, the final condition for Theorem~\fullref{storeAtomicity}
 is also satisfied. All that remains is to show that these properties hold.
@@
 
-We can directly show that Invariants~\textit{processReq} and
-\textit{noReqAgain} hold at the initial configuration and are
+We can directly show that Invariants~\fullref{processReq} and
+\fullref{noReqAgain} hold at the initial configuration and are
 preserved through each atomic transition of our protocol
-(Figure~\ref{trans}). Invariants~\textit{uniqM} and
-\textit{latestValue} correctness rely on properties of the node
+(Figure~\ref{trans}). Invariants~\fullref{uniqM} and
+\fullref{latestValue} correctness rely on properties of the node
 hierarchy. We can prove both using the following invariants on
@@
 the cache tree hierarchy, from Invariants \fullref{conservative} and
-\textit{localCompatibility}, we can show that for any address $a$,
+\fullref{localCompatibility}, we can show that for any address $a$,
 $c.\state[a] \le p.\state[c][a]$. This is because the ancestor
@@
 and Invariants \fullref{conservative} and the first part
-\textit{localCompatibility} of defines a $\ge$ relation between the
+\fullref{localCompatibility} of defines a $\ge$ relation between the
 states of a parent and its child.
@@
 sending $m$ in order to not violate Invariants \fullref{noCross} and
-\textit{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
+\fullref{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
 $p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ (because
@@
 (in which the parent can only upgrade), thus maintaining Invariant
-\textit{conservative}.
+\fullref{conservative}.
 
@@
 transit at the same time as when the response is in transit, Invariant
-\textit{noTransitWrite} follows.
+\fullref{noTransitWrite} follows.
 
@@
 %
-%Invariant~\textit{conservative} can be proved by
+%Invariant~\fullref{conservative} can be proved by
 %further breaking them into the following invariants between a parent-child pair
@@
 %
-%Invariant~\textit{conservative} at any time $t$ for child cache $c$ and parent
+%Invariant~\fullref{conservative} at any time $t$ for child cache $c$ and parent
 %$p$ for an address $a$ can be proved as follows. Consider the case when

### 2014-02-05 paper/Intuition.tex
intuition

@@
-\section{Proof of MSI protocol being sound wrt Store atomicity}
+\section{Proof of MSI protocol being preserving Store atomicity}
 \label{safety}
 
-In this section, we present the proof structure for the soundness
-property of our MSI-based cache coherence protocol of
-Figure~\ref{trans}. Specifically, we formally reduce the proof of the
-store atomicity theorem for the whole system to the proof of a set of
-simple invariants by leveraging the ``intuitive'' justification of
-correctness of our protocol.
+In this section, we present the structure of the proof of Theorem
+\fullref{storeAtomicity} for a system implementing the MSI protocol of
+Figure~\ref{trans}. We reduce the whole system proof to that of a set of
+invariants between any parent-child pair. These invariants in turn depend on
+exactly two invariants. Identifying these two basic invariants is one of the
+main contributions of the paper; they also hold for other protocols like MOSI
+(discussed in Section \ref{sec:Refinements}).
 
-%% This intuition
-%% is not restricted to the MSI-based protocol;
-%% Section~\ref{sec:Refinements} shows that it applies can apply to other
-%% protocols.
+The invariants hold after any number of executions of the transactions starting
+from the initial state. We use the term time $t$ to specify the global state
+reached after firing $t$ transitions.
 
-Our invariants hold after any series of executions of the component
-transactions of our design reachable from the initial state. We use
-the term time $t$ to specify the global state reached after firing $t$
-transitions.
-
-While we have proofs in Coq of each step in this framework our
-discussion will be mainly informal for brevity. A full description can
-be found at~\xxx{MURALI PUT REFERENCE HERE}.
+While we have proofs in Coq for each invariant our discussion will be mainly
+informal for brevity. A full description can be found at~\xxx{MURALI PUT
+REFERENCE HERE}.
 
@@
 The store atomicity is a whole system property. Intuitively our
-protocol is correct because it guarantees that at most one node can
-write at a time, and that any local cache that attempts to read has
+protocol is correct because it guarantees that at most one cache can
+write at a time, and that any cache that attempts to read has
 the ``latest'' value. We can state this formally with the following 4
-invariants about nodes:
+invariants.
 
-\begin{defn}
-\textit{clean}: A cache is said to be \clean{} for an address $a$ if and only if
-the state of the cache for that address is at least \Sh{} and the
-state of all its children is at most \Sh.
-\end{defn}
 \begin{inv}
@@
 \begin{enumerate}
-\item $c.\data[a] = \initData(a)$ and no store request has been
-  processed at any time $t_i$ such that $t_i < t$, or
-\item There is a store request $q$ that was processed at time $t'$ such that
-$t' < t$ and $a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
-request $q'$ was processed at any time $t_i$ such that $t' < t_i < t \wedge a =
+\item $c.\data[a] = \initData(a)$ and no store request $q'$ has been processed at
+any time $t'$ such that $t' < t \wedge a = q'.\addrQ$, or
+\item There is a store request $q$ that was processed at time $t_q$ such that
+$t_q < t \wedge a = q.\addrQ \wedge c.\data[a] = q.\dataQ$ and no other store
+request $q'$ was processed at any time $t'$ such that $t_q < t' < t \wedge a =
 q'.\addrQ$.
@@
 \end{inv}
-
-We use the definition of compatibility between two caches as follows:
+A clean cache is defined as follows.
 \begin{defn}
-\textit{compatibile}: Values of states $x_1$ and $x_2$ are compatible if and only if
-either $x_1$ or $x_2$ is \Mo{}, then the other is \In.
-\label{compatible}
+\textit{clean}: A cache is said to be \clean{} for an address $a$ if and only if
+the state of the cache for that address is at least \Sh{} and the directory 
+state for all its children is at most \Sh.
 \end{defn}
@@
 \textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
-that neither is an ancestor of the other in the cache tree hierarchy, for each
-address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are compatible.
+that neither is an ancestor of the other in the cache tree
+hierarchy (where ancestor relation is the reflexive-transitive closure of the parent relation), for each
+address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are \textit{compatible}.
 \label{nonAncestorCompatible}
 \end{inv}
+Compatibility of states is defined as follows.
+\begin{defn}
+\textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible} iff
+either $x_1$ or $x_2$ is \Mo{}, then the other is \In. Note that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also compatible if $x_1, x_2$ are compatible.
+\label{compatible}
+\end{defn}
 
-\begin{inv}
-\textit{processReq}: A load (store) request for an address at a leaf cache can
-be processed only if it is at least in the \Sh{} (\Mo) state for that address.
-\label{processReq}
-\end{inv}
+%\begin{inv}
+%\textit{processReq}: Load and store requests can be processed at a leaf cache
+%only if it is at least in the \Sh{} and \Mo{} states, respectively, for that
+%address.
+%\label{processReq}
+%\end{inv}
+%
+%\begin{inv}
+%\textit{noReqAgain}:
+%\item For a leaf cache $c$, $c.\pos$ changes only on processing a request. In
+%that case, $c.\pos$ gets incremented.
+%\label{noReqAgain}
+%\end{inv}
 
-\begin{inv}
-\textit{noReqAgain}:
-\item For a leaf cache $c$, $c.\pos$ changes only on processing a request in
-which case $c.\pos$ is incremented.
-\label{noReqAgain}
-\end{inv}
+Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
+follows. From the transitions of Figure \ref{trans}, a store may happen only
+when a leaf cache is in \Mo{} state for that
+address%(Invariant~\textit{processReq}).
+Leaf caches cannot be ancestors of any
+other node, so by Invariant \fullref{nonAncestorCompatible}, if a leaf cache is
+in \Mo{} state, all other leaf caches will be in \In{} state preventing the
+possibility of processing load and store requests for the same address at the
+same time. Thus, Definition~\fullref{uniqStTimes} holds.
 
-Given these invariants, we codify our intuition by showing that
-Theorem~\fullref{storeAtomicity} follows naturally from these
-invariants.  A store may happen only when leaf cache is in \Mo{} state
-for that address (Invariant~\fullref{processReq}). Leaf caches cannot
-be ancestors of any other node, so by Invariant
-\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
-other leaf caches will be in \In{} state preventing the possibility of
-load and store requests for the same address (by Invariant
-\fullref{processReq}), showing Definition~\fullref{uniqStTimes} holds.
-
-Given Invariant~\fullref{noReqAgain}, once a response has been produced
-for a request $i$, the cache may subsequently only process $j^{th}$
-request such that $j > i$. Thus there is at most one response to each
-request, satisfying Definition~\fullref{uniqRespLabels}. A load request
-may be processed only if a cache is at least in \Sh{} state (Invariant
-\fullref{processReq}). As leaf caches have no children, a leaf cache in
-at least in \Sh{} state for address $a$ is \clean for that address. A
-load response supplies the data for that address present in the cache,
-which, by Invariant~\fullref{lastestValue}, contains the ``latest
-value''. Thus, the final condition for Theorem~\fullref{storeAtomicity}
-is also satisfied. All that remains is to show that these properties hold.
+Once a response has been produced for a request $i$, the cache increments the
+counter for the request it processes next. Thus there is at most one response
+to each request, satisfying Definition~\fullref{uniqRespLabels}. As leaf caches
+have no children, a leaf cache in at least \Sh{} state for address $a$ is
+\clean{} for that address. Since processing a load request requires the cache
+to be at least in the \Sh{} state, its response, which supplies the data for
+that address present in the cache, supplies the latest value by
+Invariant~\fullref{lastestValue}.  Thus, the final condition for
+Theorem~\fullref{storeAtomicity} is also satisfied. All that remains is to show
+how these invariants hold.
 
@@
 
-We can directly show that Invariants~\fullref{processReq} and
-\fullref{noReqAgain} hold at the initial configuration and are
-preserved through each atomic transition of our protocol
-(Figure~\ref{trans}). Invariants~\fullref{uniqM} and
-\fullref{latestValue} correctness rely on properties of the node
-hierarchy. We can prove both using the following invariants on
-parent-child relationships.
+Correctness of Invariants \fullref{uniqM} and \fullref{latestValue} can be
+proved using the following invariants on a parent-child pair.
 
@@
 \textit{conservative}: The state of an address in a cache is never greater than
-the corresponding entry in the parent's directory.
+the corresponding entry in its parent's directory.
 \label{conservative}
@@
 \item Let $c_1$ and $c_2$ be distinct caches with the same parent $p$. Then,
-$p.\dstate[c_1][a]$ and $p.\dstate[c_2][a]$ are compatible.
+$p.\dstate[c_1][a]$ and $p.\dstate[c_2][a]$ are \textit{compatible}.
 \end{enumerate}
@@
 For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
-the cache tree hierarchy, from Invariants \fullref{conservative} and
-\fullref{localCompatibility}, we can show that for any address $a$,
-$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
-relation is the reflective-transitive closure of the parent relation,
-and Invariants \fullref{conservative} and the first part
-\fullref{localCompatibility} of defines a $\ge$ relation between the
-states of a parent and its child.
+the cache tree hierarchy,
 
-Any two caches $c_1, c_2$ neither of which are an ancestor of the
-other will have a lowest common ancestor $x$ which is an ancestor of
-both $c_1$ and $c_2$ and has two distinct children $x_1$ and $x_2$
-such that they are the ancestors of $c_1$ and $c_2$ respectively. For
-any address $a$, $x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are
-compatible by the second part of
-Invariant~\ref{localCompatibility}. From this, it straightaway follows
-that $x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from
-Invariant~\ref{conservative}). From the previous paragraph, it follows
-that $c_1$ and $c_2$ are also compatible. This proves
-Invariant \fullref{nonAncestorCompatibility}.
+Invariant \fullref{conservative} and the first part of Invariant
+\fullref{localCompatibility} defines a $\ge$ relation between the states of a
+parent and its child for any address.  Since the ancestor relation is the
+reflexive-transitive closure of the parent relation, the $\ge$ relation is
+preserved between the states of two caches $p$ and $c$ where $p$ is the
+ancestor of $c$.
 
-We prove Invariant \fullref{latestValue} by strong induction on
-time. Consider two cases: cache $c$ is already \clean{} for an address
-$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
-value at time $t$. Since none of its children's directory entries can
-be in \Mo{}, all its descendants should be no greater than \Sh{} for
-that address (as discussed above). Any non-ancestor of $c$, and
-non-descendant of $c$ must be compatible (as discussed above). Since
-$c$ is \clean, it is at least in the \Sh{} state, so compatibility
-means that non-descendants can not be in \Mo{} state. All of $c$'s
-ancestors are non-leaves, so they can not process any store
-request. Therefore, no cache save $c$ may have process a store
-request, which means that $c$ contains the latest value for address
-$a$ at time $t+1$ if it was \clean{} at time $t$.
+Any two caches $c_1, c_2$ neither of which are an ancestor of the other has a
+lowest common ancestor $x$ which is an ancestor of both $c_1$ and $c_2$. $x$
+has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
+$c_1$ and $c_2$ respectively. For any address $a$, $x.\dstate[x_1][a]$ and
+$x.\dstate[x_1][a]$ are compatible by the second part of
+Invariant~\fullref{localCompatibility}. It follows that $x_1.\state[a]$ and
+$x_2.\state[a]$ are compatible since these values are smaller than the
+respective directory values.  From the previous paragraph about the relation
+between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
+are also compatible. This proves Invariant \fullref{nonAncestorCompatible}.
 
-If $c$ was not clean at time $t$, it must have received data from
-either its parent or its child.  As can be seen in Figure~\ref{trans},
-whenever data is sent either from a parent to a child, or from a child
-to the parent, the sender is \clean. By the induction hypothesis, it
-means that the sender has the latest value for the address. Because of
-Invariant~\ref{noTransitWrite}, no store request can be processed when
-the data is in transit, which means $c$ will have the latest value for
-address $a$ at time $t+1$.
+We prove Invariant \fullref{latestValue} by strong induction on time. Consider
+two cases: a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
+b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
+none of its children's directory entries can be in \Mo{}, all its descendants
+should be no greater than \Sh{} for that address (because of Invariant
+\fullref{conservative} and the fact that $\ge$ relation is preserved for the
+state of a cache and that of its ancestor). Any non-ancestor of $c$, and
+non-descendant of $c$ must be compatible (as discussed in the previous
+paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
+compatibility means that non-descendants can not be in \Mo{} state. All of
+$c$'s ancestors are non-leaves, so they can not process any store request.
+Therefore, no cache save $c$ may have process a store request, which means that
+$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
+at time $t$.
+
+If $c$ was not clean at time $t$, it must have received data from either its
+parent or its child.  As can be seen in Figure~\ref{trans}, whenever data is
+sent either from a parent to a child, or from a child to the parent, the sender
+is \clean. By the induction hypothesis, it means that the sender has the latest
+value for the address. Because of Invariant~\ref{noTransitWrite}, no store
+request can be processed when the data is in transit, which means $c$ will have
+the latest value for address $a$ at time $t+1$. This establishes Invariant
+\fullref{latestValue}.
 
@@
 
-We present two basic invariants that the MSI protocol of
-Figure~\ref{trans} obeys. In fact the protocol (\ie{} the
-request/response message semantics) was designed to ensure the
-following two invariants. We believe that most cache-coherence
-protocols would end up obeying these two invariants; we show how MOSI,
-for instance, obeys these invariants in
-Section~\ref{sec:Refinements}. These two invariants together can prove
-all the pair-wise properties of Section~\ref{pairwise}.
+We present two basic invariants that the MSI protocol of Figure~\ref{trans}
+obeys. In fact the request/response message semantics was designed to satisfy
+the following two invariants. We believe that most cache-coherence protocols
+would end up obeying these two invariants. With two invariants we can
+prove all the pair-wise properties of Section~\ref{pairwise}.
 
-The transitions of the protocol are setup so that the following invariant holds:
 \begin{inv}
@@
 
-In addition, the following invariant must hold regarding the order of responses
-between the same pair of nodes:
 \begin{inv}
@@
 receiving a response. The parent upgrades the corresponding directory entry on
-sending a response to a child and the child downgrades its directory entry on
+sending a response to a child and the child downgrades its state on
 sending a response to the parent.
@@
 before $t$. Consider the last such message $m$ sent by or received at $c$ at
-$t'$. Let's say $m$ is sent by $c$. Then $p$ can not have received any message
-from $c$ after sending $m$ in order to not violate Invariant \fullref{noCross}.
-Instead, if $c$ receives $m$ at $t'$, $p$ will not receive any message after
-sending $m$ in order to not violate Invariants \fullref{noCross} and
-\fullref{respFIFO}. Let $m$ be sent or received by $p$ at $t''$. In both cases,
-$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ (because
+$t'$. Let's say it is sent by $c$ at $t'$, not received. Then $p$ can not have
+received any message from $c$ after sending $m$ and before $t$ in order to not
+violate Invariant \fullref{respFIFO}.  Instead, if $c$ receives $m$ at $t'$,
+$p$ will not receive any message after sending $m$ and before $t$ in order to
+not violate Invariant \fullref{noCross}. Thus, $p$ will not receive any message
+from $c$ after sending or receiving $m$ and before $t$.  Let $t''$ be the time
+at which $p$ has sent or received $m$. Whether $p$ sends or receives $m$,
+$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ because
 the response messages sync up the next-state according to the transitions in
-Figure \ref{trans}). Since $p$ can not receive any message between $t''+1$ and
-$t$, it can only upgrade $p.\dstate[c][a]$. Since $c.\state[a]$ does not change
-between $t'+1$ and $t$, Invariant \fullref{conservative} holds at $t$. If no
-message has been sent or received by $c$ before $t$, then $p$
-wouldn't have received any message sent by $c$ before $t$ as it means that the
-message must be sent by $c$ before $t$. Since the initial states of the
-directory and the child's state are both set to \In{}, the same argument holds
-(in which the parent can only upgrade), thus maintaining Invariant
-\fullref{conservative}.
+Figure \ref{trans}. Since $p$ can not receive any message between $t''+1$ and
+$t$, it can only upgrade $p.\dstate[c][a]$. $c.\state[a]$ does not change
+between $t'+1$ and $t$ as it never changes state. Therefore, Invariant
+\fullref{conservative} holds at $t$.
+
+If no message has been sent or received by $c$ before $t$, then $p$ wouldn't
+have received any message sent by $c$ before $t$. So $p.\dstate[c][a]$ at time
+$t \ge p.\dstate[c][a]$ at time $0$ . Since the initial values of the directory
+and the child's state are both set to \In{} Invariant \textit{conservative}
+holds.
 
 To prove Invariant \fullref{localCompatibility}, it is enough to show that a
-child $c$ upgrades $c.\state[a]$ on receiving a response for $a$ from its
-parent $p$ and $p.\dstate[c][a]$ is downgraded on receiving a response for $a$
-from $c$. If this is true, then Invariant \fullref{localCompatibility} will not
-be violated on receiving a response. The transitions of Figure \ref{trans}
-ensures that Invariant \fullref{localCompatibility} is not violated when sending
-a response.
+$c.\state[a]$ is upgraded on receiving a response for $a$ from $c$'s parent $p$
+and $p.\dstate[c][a]$ is downgraded on receiving a response for $a$ from $c$.
+If this is true, then Invariant \fullref{localCompatibility} will not be
+violated on receiving a response. The transitions of Figure \ref{trans} ensures
+that Invariant \fullref{localCompatibility} is not violated when sending a
+response. Since whenever the state or directory changes, the appropriate
+response is sent or received, Invariant \fullref{localCompatibility} holds.
 
@@
 receiving the response, it follows straightaway that the parent downgrades its
-directory on receiving a response from the child, since a child downgrades on
-sending a response and the next-states are synced on receiving the response.
-The proof for this follows exactly the same arguments presented above to prove
-Invariant \fullref{conservative} -- we prove by considering the last message
-sent or received at the child before $t$ (and the case when no message has been
-sent or received).
+directory on receiving a response from the child. This is because a child
+downgrades on sending a response and the next-states are synced on receiving
+the response.  This proof uses exactly the same arguments presented above to
+prove Invariant \fullref{conservative} -- we prove by considering the last
+message sent or received at the child before $t$ (and the case when no message
+has been sent or received). Similarly, we can show that a child upgrades its
+state on receiving a response, satisfying the conditions in the previous paragraph.
 
-Invariant \fullref{noTransitWrite} can be proved as follows. Consider a response
-sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
-other response can be received by the parent at time $t_p$ such that $t+1 \le
-t_p \le t'$ to not violate Invariant \fullref{noCross}, so $p.\dstate[c][a]$ at
-$t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
-directory for $c$. Similarly, no response can be sent by the child at time
-$t_c$ such that $t \le t_c \le t'$ for the same reason and hence $c.\state[a]$
-at $t \le c.\state[a]$ at $t'$ since the child can only downgrade its state.
-This means that during $t_p$ such that $t+1 \le t_p \le t'$, $p.\dstate[c][a]
-> \In$ since the parent upgraded its directory for $c$ at $t$. Similarly,
-during $t_c$ such that $t \le t_c \le t'$, $c.\state[a] < \Mo$ since the
-child upgrades its state at $t'$. Using arguments similar to the proof of
-Invariant \fullref{latestValue}, we can establish that no cache can be in \Mo{}
-state for address $a$ when a response is in transit from $p$ to $c$.
-Similarly, we can prove that no cache can be in \Mo{} state for address $a$
-when a response is in transit for address $a$ from $c$ to $p$. Since data is in
-transit at the same time as when the response is in transit, Invariant
-\fullref{noTransitWrite} follows.
+Invariant \fullref{noTransitWrite} can be proved as follows. Consider a
+response sent from a parent $p$ at time $t$ received by its child $c$ at time
+$t'$. No other response can be received by the parent at time $t_p$ such that
+$t+1 \le t_p \le t'$ in order to not violate Invariant \fullref{noCross}, so
+$p.\dstate[c][a]$ at $t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can
+only upgrade its directory for $c$. Similarly, no response can be sent by the
+child at time $t_c$ such that $t \le t_c \le t'$ for the same reason and hence
+$c.\state[a]$ at $t \le c.\state[a]$ at $t'$ since the child can only downgrade
+its state.  This means that during $t_p$ such that $t+1 \le t_p \le t'$,
+$p.\dstate[c][a] > \In$ and (hence $p.\state[a] > \In$ by the first part of
+Invariant \fullref{localCompatibility}) since the parent upgraded its directory
+for $c$ at $t$.  Similarly, during $t_c$ such that $t \le t_c \le t'$,
+$c.\state[a] < \Mo$ since the child upgrades its state at $t'$. We have already
+shown that (in the proof of Invariant \fullref{latestValue}) if any cache is
+not in \In{} state, no non-descendent cache can process a store request. In
+this case, during the times between $t$ and $t'$, $p$ is not in \In{} state.
+Among $p$'s directory entries, all of them save the entry for $c$ is not in
+\Mo{} state, so none of the descendents of $p$'s children, save $c$'s
+descendents can process a store request. During the times between $t$ and $t'$,
+$c$'s state is $< \Mo{}$, so neither $c$ nor its descendents can process a store
+request, proving Invariant \fullref{noTransitWrite}.
 
@@
 %
-%Invariant~\fullref{conservative} can be proved by
+%Invariant~\textit{conservative} can be proved by
 %further breaking them into the following invariants between a parent-child pair
@@
 %
-%Invariant~\fullref{conservative} at any time $t$ for child cache $c$ and parent
+%Invariant~\textit{conservative} at any time $t$ for child cache $c$ and parent
 %$p$ for an address $a$ can be proved as follows. Consider the case when

### 2014-02-05 paper/Deadlock.tex
intuition

--- 2014-02-04 paper/Deadlock.tex
@@
 
-In Section \ref{safety} we showed how the MSI protocol of Figure \ref{trans}
-obeys the Store atomicity theorem. In addition to this criteria, we must
-establish that the system makes forward progress.
+In Section~\ref{safety} we showed how the MSI protocol of Figure~\ref{trans}
+obeys the store atomicity theorem. To complete the picture, we must also show
+that it avoids deadlock, \ie{} the protocol makes forward progress.
 
@@
 \begin{enumerate}
-\item Weak forward progress: The system never reaches a state where none of the transitions of Figure
-\ref{trans} apply or
-\item Strong forward progress: At any time, eventually a new request from some processor will be
-processed
+\item Weak forward progress: The system never reaches a state where
+  none of the transitions of Figure~\ref{trans} apply, or
+\item Strong forward progress: At any time, eventually a new request
+  from some processor will be processed.
 \end{enumerate}
 
-In order to prove strong forward progress, we need to
-establish some more conditions on the strategy to execute the transitions of
-Figure \ref{trans} in each cache -- the caches must ensure local
-starvation-freedom in the processing of messages when sufficient resources are
-available. We do not discuss this issue in this paper and instead focus on 
-weak forward progress.
+To prove strong forward progress, we need some guarantees on the choice of
+executions of the transitions Figure~\ref{trans} in each cache. Each cache must
+ensure that it does not avoid receiving a particular available request from one
+of its children indefinitely, when the guard for the corresponding transition
+(ParentRecvReq) is true. Besides, all the transitions in Figure \ref{trans} are
+voluntary in nature. Thus, the system can execute a sequence of transitions
+which simply moves, say, the \Mo{} permission among the children of a cache,
+without doing any further work with that permission
+%This can potentially be enforced via during hardware
+%compilation;
+For this paper we focus only on weak forward progress, as we have not modeled
+the algorithm for correctly choosing the transitions to guarantee strong
+forward progress in Coq.
 
@@
 this invariant, it is easy to see that there will always be some cache in the
-system which can either send a request to another cache (ChildSendReq or
-ParentSendReq) or process a request (ParentRecvReq, ChildRecvReq or
-ChildDropReq) or accept a response from another cache (ChildRecvResp or
-ParentRecvResp). We do not give the formal proof here due to lack of space.
+system which can either send a request to another cache if the cache is not
+waiting for any response from the latter cache (ChildSendReq or ParentSendReq),
+or accept a response from another cache if such a response is available
+(ChildRecvResp or ParentRecvResp). But if all the caches are waiting for a
+response, and no response is available, then the leaf caches can execute
+ChildRecvReq transition (as the leaf does not have any children, and this
+transition is executed even if a cache is waiting for a response from its
+parent). Therefore, some transition can always be executed, proving weak
+forward progress.

### 2014-02-05 paper/Intuition.tex
intuition

@@
 \begin{inv}
+\textit{latestValue}:
 At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
@@
 that address present in the cache, supplies the latest value by
-Invariant~\fullref{lastestValue}.  Thus, the final condition for
+Invariant~\fullref{latestValue}.  Thus, the final condition for
 Theorem~\fullref{storeAtomicity} is also satisfied. All that remains is to show
@@
 
-Correctness of Invariants \fullref{uniqM} and \fullref{latestValue} can be
+Correctness of Invariants \fullref{localCompatibility} and \fullref{latestValue} can be
 proved using the following invariants on a parent-child pair.
@@
 for a source-destination pair, for the same address.
+\label{respFIFO}
 \end{inv}

### 2014-02-05 paper/Intuition.tex
Introduction put back in. fix some links

@@
 \section{Proof of MSI protocol being preserving Store atomicity}
-\label{safety}
+\label{Sec:Safety}
 
@@
 main contributions of the paper; they also hold for other protocols like MOSI
-(discussed in Section \ref{sec:Refinements}).
+(discussed in Section~\ref{Sec:Refinements}).
 

### 2014-02-05 paper/Intuition.tex
Cleanup Refinement text a bit.

@@
 
-The store atomicity is a whole system property. Intuitively our
-protocol is correct because it guarantees that at most one cache can
-write at a time, and that any cache that attempts to read has
-the ``latest'' value. We can state this formally with the following 4
+Store atomicity is a whole system property. Intuitively our
+protocol is correct because it guarantees at most one cache may
+write at a time, and that any cache permitted to read has
+the ``latest'' value. We state this formally with the following 4
 invariants.

### 2014-02-05 paper/Introduction.tex

@@
 \section{Introduction}
-\label{sec:Introduction}
+\label{Sec:Introduction}
 
@@
 For this approach to be truly viable, designers cannot be handed a
-single proof of correctnessof a single design; rather, they need a
+single proof of correctness of a single design; rather, they need a
 proof framework covering any variation of a design that might be of
@@
 framework. We discuss our MSI protocol example and its implementation
-in Section~\ref{Sec:System}. In Section~\ref{Sec:ProofStructure} we discuss
-the structuring of our proof framework and verification of the base
-protocol. We show implementation refinements in
-Section~\ref{Sec:Refinements} and conclude with a general discussion
-of the applicability of our technique.
+in Section~\ref{Sec:System}. In Sections~\ref{Sec:Safety} and
+\ref{Sec:Liveness} we discuss the structuring of our proof framework and
+verification of the base protocol. We show implementation refinements
+in Section~\ref{Sec:Refinements} and conclude with a general
+discussion of the applicability of our technique.

### 2014-02-05 paper/System.tex
some modifications of backgrounda
Introduction put back in. fix some links

@@
 \section{MSI-based cache coherence protocol}
-\label{sec:System}
+\label{Sec:System}
 
@@
 Figure~\ref{parentside} shows the similar sequence of transitions that results
-in downgrading a cache's directory entry for one of its chilren $c$. The
+in downgrading a cache's directory entry for one of its children $c$. The
 difference is in the guard for ChildRecvReq transition: the directory entry for
@@
 
-Figure~\ref{childextra} shows transitions that happen in a non-root cache
-because of the presence of \emph{voluntary} responses. This models eviction of
-a location from a cache to make room for a new location.  This eviction usually
-causes the permission state of the cache for that address to go to $\In$, while
-we allow any downgrade to take place voluntarily.  We use the same \cpResp{}
-FIFO for such a message. Such a voluntary response for an address can take
-place only if the cache is not waiting for a response from its parent for
-that address (ChildVolResp).
+Figure~\ref{childextra} shows transitions that happen in a non-root
+cache due to the presence of \emph{voluntary} responses. This models
+eviction from a cache. This eviction usually causes the permission
+state of the cache for that address to go to $\In$, while we allow any
+downgrade to take place voluntarily.  We use the same \cpResp{} FIFO
+for such a message. Such a voluntary response for an address can take
+place only if the cache is not waiting for a response from its parent
+for that address (ChildVolResp).
 
-Because of the presence of voluntary response, a child could have already
+Because of the presence of voluntary responses, a child may have already
 downgraded to a state required by a downgrade request from its parent. In such
-a case, the child simply drops the downgrade request (ChildDropReq).
+a case, the child drops the downgrade request (ChildDropReq).
 
@@
 Guard: & 
-%$\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge%$\\
-%& $%
-%q.\myto \ge c.\state[q.\addr]$\\
 $\avail(\pc[c]) \wedge \mylet q := \first(\pc[c]) \myin q.\typ = \Req \wedge q.\myto \ge c.\state[q.\addr]$\\

### 2014-02-05 paper/Intuition.tex
Text cleanup

@@
 \fullref{storeAtomicity} for a system implementing the MSI protocol of
-Figure~\ref{trans}. We reduce the whole system proof to that of a set of
-invariants between any parent-child pair. These invariants in turn depend on
-exactly two invariants. Identifying these two basic invariants is one of the
-main contributions of the paper; they also hold for other protocols like MOSI
-(discussed in Section~\ref{Sec:Refinements}).
+Figure~\ref{trans}. We reduce the whole system proof to that of a set
+of invariants between any parent-child pair. These invariants in turn
+depend on two smaller invariants which encapsulate the core aspects of
+correctness of the MSI protocol and allow the large scope of verified
+hardware refinements we present later (Section~\ref{Sec:Refinements}).
 
-The invariants hold after any number of executions of the transactions starting
-from the initial state. We use the term time $t$ to specify the global state
-reached after firing $t$ transitions.
+The invariants hold after any number of executions of the transactions
+starting from the initial state. We use the term time $t$ to specify
+the global state reached after firing $t$ transitions.
 

### 2014-02-05 paper/System.tex
Text cleanup

@@
 
-We have dedicated channels between each parent-child pair. The three channels
-are indexed by the name of the child that uses that channel (denoted by
-$\textit{ch}[c]$ for the \textit{ch} channel associated with cache $c$).
-Figure \ref{format} shows the message format transmitted in the 3
-channels between caches. In each of the channels $ch$, a message $m$
-can be enqueued ($\enq(ch, m)$) or dequeued ($\deq(ch)$), and the
-channels can be examined for the first element ($\first(ch)$) or for
-availability of any message in the channel ($\avail(ch)$).  Each of
-these channels start out empty \ie $\forall ch, \neg \avail(ch)$,
-initially.
+We have dedicated channels between each parent-child pair. The three
+channels are indexed by the name of the child that uses that channel
+(denoted by $\textit{ch}[c]$ for the \textit{ch} channel associated
+with cache $c$).  Figure~\ref{format} shows the message format
+transmitted in the 3 channels between caches. In each of the channels
+$ch$, a message $m$ can be enqueued ($\enq(ch, m)$) or dequeued
+($\deq(ch)$), and the channels can be examined for the first element
+($\first(ch)$) or for availability of any message in the channel
+($\avail(ch)$).  Each of these channels start out empty \ie{} $\forall
+ch, \neg \avail(ch)$, initially.
 
@@
 \hline
-Action: & $\enq(\cpResp[c], \langle c.\state[a], x, a,%$\\
-%& $\;\;\;%
+Action: & $\enq(\cpResp[c], \langle c.\state[a], x, a,
 \myif c.\state[a] = \Mo \mythen c.\data[a] \myelse \_\rangle)$;\\
@@
 
-Note that the atomic transitions in Figure~\ref{trans} access only local states,
-\ie{} a transition can read or write states only corresponding to a single cache
-and/or the channel the cache is connected to. This restriction allows such a
-system to be implemented directly into hardware. Bluespec System Verilog (BSV),
-for instance, directly converts these transitions into efficient synchronous
-hardware. Though these transitions logically happen one by one, BSV executes
-several transitions simultaneously, as long as they do not access the same
-state, \ie{} do not \emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
-Sometimes, these transitions (both the guards and the actions) may not be
-amenable to single cycle implementations in hardware. For example, writing a
-register array can actually take several hardware clock cycles, though
-logically it is a single transition.  Karczmarek \etal~\cite{Karczmarek} has
-provided a scheme to convert atomic transitions into synchronous hardware in
-which each transition can span several clock cycles.
+Note that the atomic transitions in Figure~\ref{trans} access only
+local states, \ie{} a transition can read or write states only
+corresponding to a single cache and/or the channel the cache is
+connected to. This restriction allows such a system to be implemented
+directly into hardware. Bluespec SystemVerilog (BSV), for instance,
+directly converts these transitions into efficient synchronous
+hardware. Though these transitions logically happen one by one, BSV
+executes several transitions simultaneously in a single clock cycle,
+as long as they do not access the same state, \ie{} do not
+\emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
+
+%ndave: I know this is mildly relevant, but it's making the current
+%hardware implementation seem worse and it's not adding any actual
+%knowledge to the reader who doesn't care about this level of detail
+
+%%  Sometimes,
+%% these transitions may not be amenable to single cycle implementations
+%% in hardware. For example, writing a memory can actually take several
+%% hardware clock cycles, though logically it is a single transition.
+%% Karczmarek \etal~\cite{Karczmarek} has provided a scheme to convert
+%% atomic transitions into synchronous hardware in which each transition
+%% can span several clock cycles.

### 2014-02-05 paper/Intuition.tex
finsih merging Sections. Few real changes, just bad auto merging. I may have lost some edits.

@@
 
-Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
-follows. From the transitions of Figure \ref{trans}, a store may happen only
-when a leaf cache is in \Mo{} state for that
-address%(Invariant~\textit{processReq}).
-Leaf caches cannot be ancestors of any
-other node, so by Invariant \fullref{nonAncestorCompatible}, if a leaf cache is
-in \Mo{} state, all other leaf caches will be in \In{} state preventing the
-possibility of processing load and store requests for the same address at the
-same time. Thus, Definition~\fullref{uniqStTimes} holds.
+Given these invariants, we codify our intuition by showing that
+Theorem~\fullref{storeAtomicity} follows naturally from these
+invariants.  A store may happen only when leaf cache is in \Mo{} state
+for that address (Invariant~\fullref{processReq}). Leaf caches cannot
+be ancestors of any other node, so by Invariant
+\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all
+other leaf caches will be in \In{} state preventing the possibility of
+load and store requests for the same address (by Invariant
+\fullref{processReq}), showing Definition~\fullref{uniqStTimes} holds.
 
-Once a response has been produced for a request $i$, the cache increments the
-counter for the request it processes next. Thus there is at most one response
-to each request, satisfying Definition~\fullref{uniqRespLabels}. As leaf caches
-have no children, a leaf cache in at least \Sh{} state for address $a$ is
-\clean{} for that address. Since processing a load request requires the cache
-to be at least in the \Sh{} state, its response, which supplies the data for
-that address present in the cache, supplies the latest value by
-Invariant~\fullref{latestValue}.  Thus, the final condition for
-Theorem~\fullref{storeAtomicity} is also satisfied. All that remains is to show
-how these invariants hold.
+Given Invariant~\fullref{noReqAgain}, once a response has been produced
+for a request $i$, the cache may subsequently only process $j^{th}$
+request such that $j > i$. Thus there is at most one response to each
+request, satisfying Definition~\fullref{uniqRespLabels}. A load request
+may be processed only if a cache is at least in \Sh{} state (Invariant
+\fullref{processReq}). As leaf caches have no children, a leaf cache in
+at least in \Sh{} state for address $a$ is \clean for that address. A
+load response supplies the data for that address present in the cache,
+which, by Invariant~\fullref{lastestValue}, contains the ``latest
+value''. Thus, the final condition for Theorem~\fullref{storeAtomicity}
+is also satisfied. All that remains is to show that these properties hold.
 
@@
 For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
-the cache tree hierarchy,
+the cache tree hierarchy, from Invariants \fullref{conservative} and
+\fullref{localCompatibility}, we can show that for any address $a$,
+$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
+relation is the reflective-transitive closure of the parent relation,
+and Invariants \fullref{conservative} and the first part
+\fullref{localCompatibility} of defines a $\ge$ relation between the
+states of a parent and its child.
 
-Invariant \fullref{conservative} and the first part of Invariant
-\fullref{localCompatibility} defines a $\ge$ relation between the states of a
-parent and its child for any address.  Since the ancestor relation is the
-reflexive-transitive closure of the parent relation, the $\ge$ relation is
-preserved between the states of two caches $p$ and $c$ where $p$ is the
-ancestor of $c$.
+Any two caches $c_1, c_2$ neither of which are an ancestor of the
+other will have a lowest common ancestor $x$ which is an ancestor of
+both $c_1$ and $c_2$ and has two distinct children $x_1$ and $x_2$
+such that they are the ancestors of $c_1$ and $c_2$ respectively. For
+any address $a$, $x.\dstate[x_1][a]$ and $x.\dstate[x_1][a]$ are
+compatible by the second part of
+Invariant~\ref{localCompatibility}. From this, it straightaway follows
+that $x_1.\state[a]$ and $x_2.\state[a]$ are compatible (from
+Invariant~\ref{conservative}). From the previous paragraph, it follows
+that $c_1$ and $c_2$ are also compatible. This proves
+Invariant \fullref{nonAncestorCompatibility}.
 
-Any two caches $c_1, c_2$ neither of which are an ancestor of the other has a
-lowest common ancestor $x$ which is an ancestor of both $c_1$ and $c_2$. $x$
-has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
-$c_1$ and $c_2$ respectively. For any address $a$, $x.\dstate[x_1][a]$ and
-$x.\dstate[x_1][a]$ are compatible by the second part of
-Invariant~\fullref{localCompatibility}. It follows that $x_1.\state[a]$ and
-$x_2.\state[a]$ are compatible since these values are smaller than the
-respective directory values.  From the previous paragraph about the relation
-between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
-are also compatible. This proves Invariant \fullref{nonAncestorCompatible}.
+We prove Invariant \fullref{latestValue} by strong induction on
+time. Consider two cases: cache $c$ is already \clean{} for an address
+$a$ at time $t$ or the contrary.  If \clean{}, then it has the latest
+value at time $t$. Since none of its children's directory entries can
+be in \Mo{}, all its descendants should be no greater than \Sh{} for
+that address (as discussed above). Any non-ancestor of $c$, and
+non-descendant of $c$ must be compatible (as discussed above). Since
+$c$ is \clean, it is at least in the \Sh{} state, so compatibility
+means that non-descendants can not be in \Mo{} state. All of $c$'s
+ancestors are non-leaves, so they can not process any store
+request. Therefore, no cache save $c$ may have process a store
+request, which means that $c$ contains the latest value for address
+$a$ at time $t+1$ if it was \clean{} at time $t$.
 
-We prove Invariant \fullref{latestValue} by strong induction on time. Consider
-two cases: a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
-b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
-none of its children's directory entries can be in \Mo{}, all its descendants
-should be no greater than \Sh{} for that address (because of Invariant
-\fullref{conservative} and the fact that $\ge$ relation is preserved for the
-state of a cache and that of its ancestor). Any non-ancestor of $c$, and
-non-descendant of $c$ must be compatible (as discussed in the previous
-paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
-compatibility means that non-descendants can not be in \Mo{} state. All of
-$c$'s ancestors are non-leaves, so they can not process any store request.
-Therefore, no cache save $c$ may have process a store request, which means that
-$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
-at time $t$.
-
-If $c$ was not clean at time $t$, it must have received data from either its
-parent or its child.  As can be seen in Figure~\ref{trans}, whenever data is
-sent either from a parent to a child, or from a child to the parent, the sender
-is \clean. By the induction hypothesis, it means that the sender has the latest
-value for the address. Because of Invariant~\ref{noTransitWrite}, no store
-request can be processed when the data is in transit, which means $c$ will have
-the latest value for address $a$ at time $t+1$. This establishes Invariant
-\fullref{latestValue}.
+If $c$ was not clean at time $t$, it must have received data from
+either its parent or its child.  As can be seen in Figure~\ref{trans},
+whenever data is sent either from a parent to a child, or from a child
+to the parent, the sender is \clean. By the induction hypothesis, it
+means that the sender has the latest value for the address. Because of
+Invariant~\ref{noTransitWrite}, no store request can be processed when
+the data is in transit, which means $c$ will have the latest value for
+address $a$ at time $t+1$.
 
@@
 receiving the response, it follows straightaway that the parent downgrades its
-directory on receiving a response from the child. This is because a child
-downgrades on sending a response and the next-states are synced on receiving
-the response.  This proof uses exactly the same arguments presented above to
-prove Invariant \fullref{conservative} -- we prove by considering the last
-message sent or received at the child before $t$ (and the case when no message
-has been sent or received). Similarly, we can show that a child upgrades its
-state on receiving a response, satisfying the conditions in the previous paragraph.
+directory on receiving a response from the child, since a child downgrades on
+sending a response and the next-states are synced on receiving the response.
+The proof for this follows exactly the same arguments presented above to prove
+Invariant \fullref{conservative} -- we prove by considering the last message
+sent or received at the child before $t$ (and the case when no message has been
+sent or received).
 
-Invariant \fullref{noTransitWrite} can be proved as follows. Consider a
-response sent from a parent $p$ at time $t$ received by its child $c$ at time
-$t'$. No other response can be received by the parent at time $t_p$ such that
-$t+1 \le t_p \le t'$ in order to not violate Invariant \fullref{noCross}, so
-$p.\dstate[c][a]$ at $t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can
-only upgrade its directory for $c$. Similarly, no response can be sent by the
-child at time $t_c$ such that $t \le t_c \le t'$ for the same reason and hence
-$c.\state[a]$ at $t \le c.\state[a]$ at $t'$ since the child can only downgrade
-its state.  This means that during $t_p$ such that $t+1 \le t_p \le t'$,
-$p.\dstate[c][a] > \In$ and (hence $p.\state[a] > \In$ by the first part of
-Invariant \fullref{localCompatibility}) since the parent upgraded its directory
-for $c$ at $t$.  Similarly, during $t_c$ such that $t \le t_c \le t'$,
-$c.\state[a] < \Mo$ since the child upgrades its state at $t'$. We have already
-shown that (in the proof of Invariant \fullref{latestValue}) if any cache is
-not in \In{} state, no non-descendent cache can process a store request. In
-this case, during the times between $t$ and $t'$, $p$ is not in \In{} state.
-Among $p$'s directory entries, all of them save the entry for $c$ is not in
-\Mo{} state, so none of the descendents of $p$'s children, save $c$'s
-descendents can process a store request. During the times between $t$ and $t'$,
-$c$'s state is $< \Mo{}$, so neither $c$ nor its descendents can process a store
-request, proving Invariant \fullref{noTransitWrite}.
+Invariant \fullref{noTransitWrite} can be proved as follows. Consider a response
+sent from a parent $p$ at time $t$ received by its child $c$ at time $t'$. No
+other response can be received by the parent at time $t_p$ such that $t+1 \le
+t_p \le t'$ to not violate Invariant \fullref{noCross}, so $p.\dstate[c][a]$ at
+$t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can only upgrade its
+directory for $c$. Similarly, no response can be sent by the child at time
+$t_c$ such that $t \le t_c \le t'$ for the same reason and hence $c.\state[a]$
+at $t \le c.\state[a]$ at $t'$ since the child can only downgrade its state.
+This means that during $t_p$ such that $t+1 \le t_p \le t'$, $p.\dstate[c][a]
+> \In$ since the parent upgraded its directory for $c$ at $t$. Similarly,
+during $t_c$ such that $t \le t_c \le t'$, $c.\state[a] < \Mo$ since the
+child upgrades its state at $t'$. Using arguments similar to the proof of
+Invariant \fullref{latestValue}, we can establish that no cache can be in \Mo{}
+state for address $a$ when a response is in transit from $p$ to $c$.
+Similarly, we can prove that no cache can be in \Mo{} state for address $a$
+when a response is in transit for address $a$ from $c$ to $p$. Since data is in
+transit at the same time as when the response is in transit, Invariant
+\fullref{noTransitWrite} follows.
 

### 2014-02-05 paper/Introduction.tex
A pass over the intro

@@
 
-Shared memory is a common abstraction for distributed systems as it
+Shared memory is a common abstraction for distributed systems, as it
 provides an easily understood model of interaction. To achieve this
-abstraction efficiently, many hardware systems implement a cache
-coherence implementation.
+abstraction efficiently, many hardware systems implement \emph{cache
+coherence protocols}.
 
 Since the correctness of this abstraction is paramount in the correct
-functioning of the system, significant effort is placed in making sure
-that the cache coherence engine is correct. In practice, this is done
+functioning of the system, significant effort is expended in making sure
+that the cache coherence engine is correct. In practice, verification proceeds
 in two steps. First, a cache coherence protocol is constructed and
 verified. This protocol generally directly represents many of
-the key concurrent aspects of an implementation, \eg{} reads and
-writes to the system are broken into local atomic micro-operations, but
-ignores implementation details \eg{} buffer sizing and memory access
-times. Second, once the protocol has been verified it is handed to
+the key concurrent aspects of an implementation (\eg{} reads and
+writes to the system are broken into local atomic micro-operations) but
+ignores implementation details (\eg{} buffer sizing and memory access
+times). Second, once the protocol has been verified, it is handed to
 hardware experts who fill in the missing implementation details and
 realize it at the Register Transfer Level (RTL) of gates, wires, and
-registers. This involves further breaking down of local micro-operations to
-make it realizable in hardware. This effort is not a one-time translation, as
+registers, via further breaking down of local micro-operations to
+make them realizable in hardware. This effort is not a one-time translation, as
 performance properties are not easily estimated; a significant amount of
@@
 
-From a methodological perspective, this two-step process represent a
-significant methodological failure. The significant effort of
+\adamc{I don't understand what this paragraph is saying.}
+From a methodological perspective, this two-step process leaves much
+to be desired. The significant effort of
 verification is effectively lost by the concurrent non-trivial tasks
@@
 
-Because verified protocols are often described via Term Rewrite
-Systems (TRS). It was shown by Dave
+Verified protocols are often described via Term Rewrite
+Systems (TRS) \adamc{citation for TRSes?}, which apply equational logic to describe systems at an
+interesting point on the spectrum between declarative and operational methods.
+It was shown by Dave
 \etal{}~\cite{DNA:CoherenceImplementation} that such a verified
@@
 hardware description language based on Guarded Atomic Actions, a
-restriction of TRS~\cite{Hoe:TCAD}. This reduces the correctness of
-the implementation to the correctness of the protocol, effectively
-removing any implementation bugs. While the current Bluespec compiler
-has certain restrictions which limit the expressiveness of certain
-operations \eg{} fairness is not guaranteed, atomic actions are always
-implemented in a single cycle, work has been done to remove these
-restrictions~\cite{DNA:CoherenceImplementation, Karczmarek}.
+restriction of TRS~\cite{Hoe:TCAD}. In this way one reduces the correctness of
+the implementation to the correctness of the protocol, with no need
+for separate reasoning about an implementation. While the current Bluespec compiler
+does not support some implementation elements that are common in practice
+(\eg{} fairness guarantees, atomic actions running over multiple clock cycles),
+recent work has provided key insights toward lifting the
+restrictions~\cite{DNA:CoherenceImplementation, Karczmarek}.  We
+believe that Bluespec and its future evolutions represent a
+very promising platform for productive implementation and
+verification of hardware components, including cache-coherent
+shared memories.
 
-While better, this approach of protocol transliteration has
-flaws. Design of a hardware system requires many tweaks to reach
-optimal performance as one cannot practically know the correct
-low-level hardware results without evaluating the low-level
-representation.  Each of these changes can potentially breaks
-correctness and a full reverification of correctness is needed in
-principle. As hardware designers are unlikely to have the expertise or
-skill to do make this verification, this greatly weakens the ability
-to carry over our verification guarantee to the design.
-
-For this approach to be truly viable, designers cannot be handed a
-single proof of correctness of a single design; rather, they need a
-proof framework covering any variation of a design that might be of
-interest for performance reasons. This would enable implementation
-level changes without compromising on the ability to verify
-correctness of the design.
+Even though the protocol transliteration approach cuts out one
+layer of verification effort, it may not be practical in practice.
+Design of a hardware system requires many tweaks to reach
+optimal performance, since low-level performance details often
+can only be found empirically.  Each performance tweak potentially
+breaks correctness, or at least breaks the structure of the old
+\emph{proof} of correctness, so that a full \emph{reverification}
+of correctness is needed in principle. Hardware designers are
+unlikely to have the expertise or skill to do each such
+verification.  Our motivating question in this paper is, \emph{can
+we modularize formal proofs of cache coherence protocols so that
+verification experts do the hard work once and for all, and then
+hardware designers can instantiate the results to different designs
+with much less effort?}
 
 In practice, protocol variants rely on a small set of general ``local
-building block'' properties, \eg{} messages always get from source to
-destination or that the local cache's reaction to receiving a protocol
-message results in the appropriate messages being sent, the
-appropriate local state being changed \etc{}.  These properties are
-easily verifiable since they correspond to a small partition of the
+building block'' properties, like guaranteed delivery of messages sent
+between processors, or guaranteed state changes in processors upon
+receiving certain messages.  These properties are
+easily verifiable, since they correspond to a small partition of the
 system, for instance a single cache.  The details of how these
@@
 In this paper, we describe an approach to provide this proof framework
-for any particular cache coherence protocol via the example of a
-MSI-style protocol and show how it can be extended to other protocols
-like MESI and MOESI. We show how the modular structure of the proof
-and design can be shared and how modification of microarchitectural
-aspects of design directly translates to just reverification of our
-local building block properties. We use Coq for our proof
-framework. We also show how an atomic-action-based MSI protocol
-description can be directly transliterated into an Inductive
-definition in Coq, and verify its local building block properties
-formally.
+for any particular cache coherence protocol via the example of an
+MSI-style protocol.  We also show how it can be extended to other protocols
+like MESI and MOESI.  We present a modular proof architecture,
+where modification of microarchitectural
+aspects of design only necessitates reverification of relatively shallow
+local building block properties. Our framework and several instantiations
+for it are mechanized in the Coq proof assistant system, providing the
+highest levels of formal assurance upon a small trusted code base.
+We implemented and verified one particular MSI protocol via an
+operational semantics in Coq.
 
@@
 discuss our notion of correctness for our cache coherence protocol
-framework. We discuss our MSI protocol example and its implementation
+framework. We describe our MSI protocol example and its implementation
 in Section~\ref{Sec:System}. In Sections~\ref{Sec:Safety} and
-\ref{Sec:Liveness} we discuss the structuring of our proof framework and
+\ref{Sec:Liveness}, we discuss the structuring of our proof framework and
 verification of the base protocol. We show implementation refinements

### 2014-02-05 paper/Background.tex
intuition
Introduction put back in. fix some links
Text cleanup
A pass over Background

--- 2014-02-04 paper/Background.tex
@@
 \section{Correctness of a Cache Coherence Protocol}
-\label{Background}
+\label{Sec:Background}
 
-Before we can discuss our protocol and proof framework we must first establish
+Before we can discuss our protocol and proof framework, we must first establish
 exactly what it means for a cache coherence protocol to be correct.
 
-Intuitively, the simplest view of shared memory is as a single ported memory
-without caches multiplexed amongst the many individual clients/processors. In
+Intuitively, the simplest view of shared memory is as a memory (array of addressible objects)
+multiplexed amongst the many individual clients/processors, which send read/write
+messages to the memory as needed. In
 this view, all operations in the system appear to have a single unique total
@@
 Most memory consistency models can be partitioned into two
-parts~\cite{Arvind-memory-model}: a) a ``global'' store-atomicity property
+parts~\cite{Arvind-memory-model}: (a) a ``global'' store-atomicity property
 stating that for each location/address in memory, every read and write access
-to that location occurs in some total order with each operation seeing the
+to that location occurs in some total order, with each operation seeing the
 effects of previous operations and not later ones (in other words, each
-location behaves like an atomic register) and b), a ``local'' reordering
+location behaves like an atomic register) and (b), a ``local'' reordering
 restriction between requests to different locations within each processor.
@@
 
-The mechanism employed by the memory sub-system to ensure store-atomicity per
+The mechanism employed by the memory subsystem to ensure store-atomicity per
 location, in the presence of caches, is known as \emph{cache coherence}. In the
 presence of caches, multiple copies of the same address can exist in different
-caches. If these copies are independently written, the same address can end up
+caches. If these copies are written independently, the same address can end up
 having different values in different caches, violating the atomic register-like
-behavior unless the caches co-ordinate via a \emph{cache coherence protocol}.
+behavior unless the caches coordinate via a \emph{cache coherence protocol}.
 
-We will now provide the formal definition of store atomicity.
-
-\subsection{Formal definition of store atomicity}
-We will model an abstract memory sub-system, which interfaces with a set of
-processors. The requests issued by the processors to the memory sub-system can
-be modeled using a function \reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request
+\subsection{Formal Definition of Store Atomicity}
+We will model an abstract memory subsystem, which interfaces with a set of
+processors. The requests issued by the processors to the memory subsystem can
+be modeled using a function \reqFn, where $\reqFn(p, i)$ is the $i^{th}$ request
 issued by processor $p$.
@@
 %data-type corresponding to a request issued from the processor to the
-%memory sub-system.
+%memory subsystem.
 
-A request contains the address (\addrQ) corresponding to the request, and a
+We define a record type modeling a request in terms of a set of named fields.
+A request contains the address (\addrQ) corresponding to the request, as well as a
 \desc{} indicating whether the request is a load or a store. A store request
 also contains the data supplied by the processor for that address (\dataQ). A
-response contains a label (\labelR) containing a processor name and a position
-to uniquely identify the request.  In the case of a load request, the memory
-sub-system returns \dataR, the data returned by the memory sub-system for that
-address. The memory sub-system also returns an $\mathbb{N}$, \timeR, which
-intuitively represents the position at which the request is handled among the
-global set of handled requests from all the processors. \timeR{} is used solely
-for the definition of store atomicity, a real system does not return such a
-field.
+response (modeled with a different record type) contains a label (\labelR)
+to identify the request uniquely, via a pair of a processor name and a processor-local
+sequence number.  In the case of a load request, the memory
+subsystem returns \dataR, the data returned by the memory subsystem for that
+address. The memory subsystem also returns $\timeR \in \mathbb{N}$, which
+intuitively represents the time at which the request is handled among the
+global set of handled requests from all the processors. The \timeR{} field is
+present solely as a specificational device to capture store atomicity, and it
+would not be present in a running system implementation.
 
@@
 %\timeR:& Natural number representing global order of (processing) requests\\
-%\dataR:& Data returned by the memory sub-system for a \Ld{} request\\
+%\dataR:& Data returned by the memory subsystem for a \Ld{} request\\
 %\hline
@@
 
-The \Response{} set contains the set of responses from the memory sub-system.
+We model a single protocol execution as a set of all the requests and responses
+that sent in the course of that execution.
+The fundamental parameter is a (potentially infinite) set \Response{} of all responses
+that the memory sends to processors.  From here, the associated requests follow by
+applying the function $\reqFn$ to map each response to the request that provoked it.
 Responses should obey the following sanity conditions. 
-\begin{defn} (The labels in the responses are unique)
+\begin{defn} (No label is reused across multiple distinct responses.)
 %Since \dataR{} field is
@@
 \end{multline*}
+\label{uniqRespLabels}
 \end{defn}
@@
 r_1.\timeR = r_2.\timeR @-> \\
-(\reqFn (r_1.\labelR)).\addrQ = (\reqFn (r_2.\labelR)).\addrQ @->\\
-(\reqFn (r_1.\labelR)).\desc = \St @->
+\reqFn (r_1.\labelR).\addrQ = \reqFn (r_2.\labelR).\addrQ @->\\
+\reqFn (r_1.\labelR).\desc = \St @->
 r_1.\labelR = r_2.\labelR
 \end{multline*}
+\label{uniqStTimes}
 \end{defn}
@@
 \textit{localOrdering} :=
-\forall (r_1, r_2 \in \Response), (r_1.\labelR).\fst = (r_2.\labelR).\fst @->\\
-(r_1.\labelR).\snd < (r2.\labelR).\snd @-> r_1.\timeR \le r_2.\timeR
+\forall (r_1, r_2 \in \Response), r_1.\labelR.\fst = r_2.\labelR.\fst @->\\
+r_1.\labelR.\snd < r2.\labelR.\snd @-> r_1.\timeR \le r_2.\timeR
 \end{multline*}
+\label{localOrdering}
 \end{defn}
 
-This brings us to the main condition for store atomicity which states that the
+This brings us to the main condition for store atomicity, which states that the
 data returned for a load response corresponds to the \emph{last earlier} store
@@
 \begin{multline*}
-\textit{initialValue} := \lambda (a \in Addr)\; (t \in \mathbb{N}) \; (d \in Data)\Rightarrow
-d = \initData (a) \wedge \\
+\textit{initialValue}(a, t, d) := d = \initData (a) \; \wedge \\
 \forall (r' \in \Response), 0 \le r'.\timeR < t @->\\
-\neg ((\reqFn (r'.\labelR)).\addrQ = t \wedge (\reqFn (r'.\labelR)).\desc = \St)
+\neg (\reqFn (r'.\labelR).\addrQ = a \wedge \reqFn (r'.\labelR).\desc = \St)
 \end{multline*}
+\label{initialValue}
 \end{defn}
@@
 \begin{multline*}
-\textit{lastEarlierValue} := \lambda (a \in Addr)\; (t \in \mathbb{N}) \; (d \in Data)\Rightarrow
-\exists (r_m \in \Response),\\ d = (\reqFn (r_m.\labelR)).\dataQ \wedge 
-(\reqFn (r_m.\labelR)).\desc = \St \wedge\\ (\reqFn (r_m.\labelR)).\addrQ = a \wedge
-\forall (r' \in \Response), r_m.\timeR \le r'.\timeR < t @->\\
-\neg ((\reqFn (r'.\labelR)).\addrQ = a \wedge (\reqFn (r'.\labelR)).\desc = \St)
+\textit{lastEarlierValue}(a, t, d) := \exists (r_m \in \Response),\\ \reqFn (r_m.\labelR).\dataQ = d \wedge 
+\reqFn (r_m.\labelR).\desc = \St \; \wedge\\ \reqFn (r_m.\labelR).\addrQ = a \wedge
+\forall (r' \in \Response), r_m.\timeR < r'.\timeR < t @->\\
+\neg (\reqFn (r'.\labelR).\addrQ = a \wedge \reqFn (r'.\labelR).\desc = \St)
 \end{multline*}
+\label{lastEarlierValue}
 \end{defn}
 
-Using the above definitions, we can define the store-atomicity theorem as follows:
+Using the above definitions, we can define the store atomicity theorem as follows:
 \begin{thm}
@@
 \begin{multline*}
-storeAtomicity := 
+\textit{storeAtomicity} := 
 \textit{uniqRespLabels} \wedge
-\textit{uniqStTimes} \wedge \textit{localOrdering} \wedge \\
-(\forall (r \in \Response), \mylet q := \reqFn(r.\labelR) \myin q.\desc = \Ld \wedge \\
+\textit{uniqStTimes} \wedge \textit{localOrdering} \; \wedge \\
+(\forall (r \in \Response), \mylet q := \reqFn(r.\labelR) \myin q.\desc = \Ld \; @-> \\
 (\textit{initialValue}(q.\addrQ, r.\timeR, r.\dataR) \vee \textit{lastEarlierValue}(q.\addrQ, r.\timeR, r.\dataR)))
 \end{multline*}
+\label{storeAtomicity}
 \end{thm}
+
+The technical meat of this paper is a framework for proving this store atomicity
+property for a whole family of cache coherence protocols, based on parameterization
+points corresponding to key architectural choices.

### 2014-02-05 paper/System.tex
A pass over System

@@
 
-We have chosen as our concrete example is a hierarchical MSI
-protocol~\cite{MSI}. Logically, the system is organized as in a tree-like
-structure with the root connected to the main memory and each leaf interfaced
+We have chosen as our concrete example a hierarchical MSI
+protocol~\cite{MSI}. Logically, the system is organized in a tree
+structure, with the root connected to the main memory and each leaf interfaced
 with a processor (Figure~\ref{hier}). In the following, $\leaf(c)$ indicates
-that a cache is a leaf, and $\parent(c,p)$ indicates that $p$ is the parent of
+that a cache $c$ is a leaf, and $\parent(c,p)$ indicates that $p$ is the parent of
 $c$.
@@
 issued from and responses issued to the associated processor. These will be
-represented by two mappings one from $\mathbb{N}$ to requests and the other
+represented by two mappings, one from $\mathbb{N}$ to requests and the other
 from $\mathbb{N}$ to responses. For each leaf $c$, $c.\pos$ represents the
 position of the next request to process; it is set to $0$ initially.
+\adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
+\adamc{I hope the diagram explains what these mappings mean, and in general what
+the tree structure really means.}
 
@@
 state denotes which operations may be performed on the data. An address may be
-in the modified state (\Mo) representing that it may be both read and modified,
-in the shared state (\Sh) in which it may only be read, or in the Invalid state
-(\In), indicating that it can be neither modified or read. This forms a natural
+in the \emph{read-write} state \Mo{} (for ``modifiable''),
+the \emph{read-only} state \Sh{} (for ``shared''), or in the \emph{invalid} state
+\In{} (indicating that it can be neither read nor written). The options form a natural
 total order of permissions ($\In < \Sh < \Mo$). For a leaf cache, the meaning
-of these permission states is straightforward, a load request can be processed
+of these permission states is straightforward: a load request can be processed
 only if the leaf is at least in the \Sh{} state, and a store request can be
-processed only if it is in the \Mo{} state. For a non-leaf cache, it denotes
-that no cache in the subtree rooted at that cache has more permissions (we are
-modeling inclusive caches).  If a cache does not have a particular location,
-the it is equivalent to its permission being \In.  Initially, all addresses in
-each cache is in the \In{} state, save for those in the root cache which are in
-the \Mo{} state.
+processed only if it is in the \Mo{} state. The state of a non-leaf cache indicates
+an \emph{upper bound} on the permission level of any cache in the associated subtree.
+If a cache does not include a particular location,
+we consider that address to have permission \In.  Initially, all addresses in
+each cache are in the \In{} state, save for those in the root cache (i.e., main memory),
+which are in the \Mo{} state.
 
@@
 
-Non-leaf caches also keep a (possibly stale) copy of the permission state of
+Each non-leaf cache also keeps a (possibly stale) copy of the permission state of
 each of its children in a \emph{directory} for every address. The directory
 entry for a cache $c$ whose parent is $p$ for address $a$ is denoted by
-$p.\dstate[c][a]$. The directory entry for a cache, for an address is never
-less than the actual cache's permission state for that address.
+$p.\dstate[c][a]$. A directory entry for a cache is never
+less than the actual cache's permission state for the corresponding address.
 
@@
 Each cache is connected to its parent cache using 3 logical FIFO connections:
-a) \cpReq{} channel to transmit requests from the child to the parent, b)
-\cpResp{} channel to transmit responses from the child to the parent, and c)
+(a) \cpReq{} channel to transmit requests from the child to the parent, (b)
+\cpResp{} channel to transmit responses from the child to the parent, and (c)
 \pc{} channel to transmit both request and response messages from parent to
@@
 (denoted by $\textit{ch}[c]$ for the \textit{ch} channel associated
-with cache $c$).  Figure~\ref{format} shows the message format
+with cache $c$).  Figure~\ref{format} shows the message formats
 transmitted in the 3 channels between caches. In each of the channels
@@
 ($\first(ch)$) or for availability of any message in the channel
-($\avail(ch)$).  Each of these channels start out empty \ie{} $\forall
+($\avail(ch)$).  Each of these channels start out empty, \ie{} $\forall
 ch, \neg \avail(ch)$, initially.
@@
 temporary \emph{wait state} for each address to deal with transitioning between
-permission levels. The wait state consists of two parts: a) a boolean
+permission levels. The wait state consists of two parts: (a) a Boolean
 representing whether the cache has sent a request to its parent cache earlier
-and is waiting for a response (denoted by $c.\wait[a]$ for address $a$) and b)
+and is waiting for a response (denoted by $c.\wait[a]$ for address $a$) and (b)
 one of the three values \Mo, \Sh{} or \In{} representing the state upgrade that
@@
 which $p$ is waiting for $c$ to downgrade. Both \wait{} and \dwait{} are
-\False{} initially for each of the caches, and for each of the addresses,
+\False{} initially for all caches and addresses,
 because no cache is waiting for any response from either its child or its
@@
 
-The behavior of the system is described via the atomic state transitions. These
-transitions have two parts: a) the guard dictating when the transition can
-fire, and b) the action defining how the state of the system changes when the
+The behavior of the system is described via the atomic state transitions, which
+are effectively a nondeterministic operational semantics. These
+transitions have two parts: (a) the guard dictating when the transition can
+fire, and (b) the action defining how the state of the system changes when the
 transition happens. Logically, these transitions happen one at a time.
@@
 actions in the transitions in Figure \ref{trans} show only the specific states
-that change, the rest of the states remain as is. The notation ($x \Leftarrow
+that change, and other state components are assumed to remain the same.
+The notation ($x \Leftarrow
 y$;) in a transition indicates setting of a specific state $x$ of the overall
@@
 (ParentRecvReq). If the directory entries for all of the other children are
-\emph{compatible} with the upgrade request, and parent's permission state has
+\emph{compatible} with the upgrade request, and the parent's permission state has
 at least the requested value, it sends an upgrade response to the cache, while
 changing the appropriate directory state. The notation \compat{} used in
-ParentRecvReq transition of Figure \ref{trans} is defined as follows:
+the ParentRecvReq transition of Figure \ref{trans} is defined as follows:
 \begin{multline*}
-\compat = \lambda (p \in Cache) (c \in Cache) (a \in \textit{Addr}) (s \in
-\{\Mo, \Sh, \In\}) \Rightarrow\\ \forall i, i \ne c @->
+\compat(p, c, a, s) = \forall i \ne c,
 \mylet s_i := p.\dstate[i][a] \myin\\
@@
 in downgrading a cache's directory entry for one of its children $c$. The
-difference is in the guard for ChildRecvReq transition: the directory entry for
-each of the $c$'s children should not be greater than the requested
+difference is in the guard for the ChildRecvReq transition: the directory entry for
+each of $c$'s children should not be greater than the requested
 downgrade, in order for $c$ to accept a request and send a downgrade
@@
 Figure~\ref{childextra} shows transitions that happen in a non-root
-cache due to the presence of \emph{voluntary} responses. This models
+cache due to the presence of \emph{voluntary} responses, which model
 eviction from a cache. This eviction usually causes the permission
@@
 
-Figure~\ref{procside} shows the transitions that serve a processors request in
+Figure~\ref{procside} shows the transitions that serve a processor's request in
 the leaf caches. LoadReq and StoreReq show the processing of a load and store
@@
 pertaining to address $a$, the \Response{}, needed in the definition of store
-atomicity, is generated as follows: the \labelR{} field contains $(c, c.\pos)$
-pair, the \dataR{} field contains the value $c.\data[a]$, and \timeR{} field
-contains the number of transitions that has taken place in the system before
+atomicity, is generated as follows: the \labelR{} field contains a pair $(c, c.\pos)$,
+the \dataR{} field contains the value $c.\data[a]$, and the \timeR{} field
+contains the number of transitions that have happened in the system before
 the current transition, intuitively denoting the time at which this transition
@@
 $\parent(c,p) \wedge \avail(\cpResp[c]) \wedge \mylet r := \first(\cpResp[c]) \myin$\\
-& $r.\from = p.\state[c][r.\addr]$\\
+& $r.\from = p.\dstate[c][r.\addr]$\\
 \hline
@@
 \end{subfigure}
-\caption{Atomic transitions for a cache coherent memory subsystem}
+\caption{Atomic transitions for a cache-coherent memory subsystem}
 \label{trans}

### 2014-02-05 paper/Intuition.tex
A pass over Intuition

@@
-\section{Proof of MSI protocol being preserving Store atomicity}
+\section{Proof of Store Atomicity for the MSI Protocol}
 \label{Sec:Safety}
@@
 of invariants between any parent-child pair. These invariants in turn
-depend on two smaller invariants which encapsulate the core aspects of
+depend on two smaller invariants that encapsulate the core aspects of
 correctness of the MSI protocol and allow the large scope of verified
@@
 The invariants hold after any number of executions of the transactions
-starting from the initial state. We use the term time $t$ to specify
+starting from the initial state. We write ``time $t$'' for
 the global state reached after firing $t$ transitions.
 
-While we have proofs in Coq for each invariant our discussion will be mainly
+While we have proofs in Coq for each invariant, our discussion will be mainly
 informal for brevity. A full description can be found at~\xxx{MURALI PUT
 REFERENCE HERE}.
+\adamc{I presume we'll want to submit a tarball of the Coq source code with the paper.
+Is that what you have in mind here?}
 
-\subsection{Reducing correctness to Node-level Invariants}
+\subsection{Reducing Correctness to Node-Level Invariants}
 
-Store atomicity is a whole system property. Intuitively our
+Store atomicity is a whole-system property. Intuitively our
 protocol is correct because it guarantees at most one cache may
 write at a time, and that any cache permitted to read has
-the ``latest'' value. We state this formally with the following 4
+the ``latest'' value. We formalize this intuition with the following 4
 invariants.
@@
 \textit{latestValue}:
-At any time $t$, a cache $c$ which is clean for an address $a$ will have the most
+At any time $t$, a cache $c$ that is clean for an address $a$ will have the most
 up-to-date value for that address, \ie{}
@@
 that neither is an ancestor of the other in the cache tree
-hierarchy (where ancestor relation is the reflexive-transitive closure of the parent relation), for each
+hierarchy (where the ancestor relation is the reflexive-transitive closure of the parent relation), for each
 address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are \textit{compatible}.
@@
 \textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible} iff
-either $x_1$ or $x_2$ is \Mo{}, then the other is \In. Note that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also compatible if $x_1, x_2$ are compatible.
+whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In. Note that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also compatible if $x_1, x_2$ are compatible.
 \label{compatible}
@@
 
+\adamc{Something needs to be fixed up here, for consistency in how many invariants there are.}
+
 Given these invariants, we codify our intuition by showing that
-Theorem~\fullref{storeAtomicity} follows naturally from these
-invariants.  A store may happen only when leaf cache is in \Mo{} state
+Theorem~\fullref{storeAtomicity} follows naturally from the
+invariants.  A store may happen only when a leaf cache is in the \Mo{} state
 for that address (Invariant~\fullref{processReq}). Leaf caches cannot
@@
 Given Invariant~\fullref{noReqAgain}, once a response has been produced
-for a request $i$, the cache may subsequently only process $j^{th}$
+for a request $i$, the cache may subsequently only process a $j^{th}$
 request such that $j > i$. Thus there is at most one response to each
@@
 \fullref{processReq}). As leaf caches have no children, a leaf cache in
-at least in \Sh{} state for address $a$ is \clean for that address. A
+at least the \Sh{} state for address $a$ is \clean{} for that address. A
 load response supplies the data for that address present in the cache,
 which, by Invariant~\fullref{lastestValue}, contains the ``latest
-value''. Thus, the final condition for Theorem~\fullref{storeAtomicity}
-is also satisfied. All that remains is to show that these properties hold.
+value.'' Thus, the final condition for Theorem~\fullref{storeAtomicity}
+is also satisfied. All that remains is to show that the system maintains the
+invariants.
 
-\subsection{Reducing to Pair-wise Invariants}
+\subsection{Reducing to Pairwise Invariants}
 \label{pairwise}
@@
 \fullref{localCompatibility}, we can show that for any address $a$,
-$c.\state[a] \le p.\state[c][a]$. This is because the ancestor
+$c.\state[a] \le p.\state[c][a]$. The reason is that the ancestor
 relation is the reflective-transitive closure of the parent relation,
 and Invariants \fullref{conservative} and the first part
-\fullref{localCompatibility} of defines a $\ge$ relation between the
+of \fullref{localCompatibility} imply a $\ge$ relation between the
 states of a parent and its child.
@@
 be in \Mo{}, all its descendants should be no greater than \Sh{} for
-that address (as discussed above). Any non-ancestor of $c$, and
+that address (as discussed above). Any non-ancestor of $c$ and
 non-descendant of $c$ must be compatible (as discussed above). Since
 $c$ is \clean, it is at least in the \Sh{} state, so compatibility
-means that non-descendants can not be in \Mo{} state. All of $c$'s
-ancestors are non-leaves, so they can not process any store
-request. Therefore, no cache save $c$ may have process a store
+means that non-descendants cannot be in \Mo{} state. All of $c$'s
+ancestors are non-leaves, so they cannot process any store
+request. Therefore, no cache save $c$ may have processed a store
 request, which means that $c$ contains the latest value for address
@@
 either its parent or its child.  As can be seen in Figure~\ref{trans},
-whenever data is sent either from a parent to a child, or from a child
+whenever data is sent either from a parent to a child or from a child
 to the parent, the sender is \clean. By the induction hypothesis, it
@@
 
-\subsection{Proving parent-child hierarchy properties}
+\subsection{Proving Parent-Child Hierarchy Properties}
 
 %ndave: This needs some repointing
+
+\adamc{This section seems like a prime place to reduce detail to save space, while also saving readers from trying to understand proofs that are mechanized, anyway.}
 
@@
 \begin{inv}
-\textit{noCross}: A response from a cache to its child for an address can not
+\textit{noCross}: A response from a cache to its child for an address cannot
 be in transit while another response from the same child to the parent for the
@@
 before $t$. Consider the last such message $m$ sent by or received at $c$ at
-$t'$. Let's say it is sent by $c$ at $t'$, not received. Then $p$ can not have
+$t'$. If it is sent rather than received by $c$ at $t'$, then $p$ cannot have
 received any message from $c$ after sending $m$ and before $t$ in order to not
@@
 the response messages sync up the next-state according to the transitions in
-Figure \ref{trans}. Since $p$ can not receive any message between $t''+1$ and
+Figure \ref{trans}. Since $p$ cannot receive any message between $t''+1$ and
 $t$, it can only upgrade $p.\dstate[c][a]$. $c.\state[a]$ does not change
@@
 
-If no message has been sent or received by $c$ before $t$, then $p$ wouldn't
+If no message has been sent or received by $c$ before $t$, then $p$ would not
 have received any message sent by $c$ before $t$. So $p.\dstate[c][a]$ at time
-$t \ge p.\dstate[c][a]$ at time $0$ . Since the initial values of the directory
-and the child's state are both set to \In{} Invariant \textit{conservative}
+$t$ is greater than $p.\dstate[c][a]$ at time $0$. Since the initial values of the directory
+and the child's state are both set to \In{}, Invariant \textit{conservative}
 holds.
@@
 The proof for this follows exactly the same arguments presented above to prove
-Invariant \fullref{conservative} -- we prove by considering the last message
+Invariant \fullref{conservative} -- we prove it by considering the last message
 sent or received at the child before $t$ (and the case when no message has been

### 2014-02-05 paper/Introduction.tex
Cut out text

@@
 
-\adamc{I don't understand what this paragraph is saying.}
-From a methodological perspective, this two-step process leaves much
-to be desired. The significant effort of
-verification is effectively lost by the concurrent non-trivial tasks
-of translation from the protocol representation to a hardware
-description and the necessary refinement to add missing implementation
-details to the protocol.
+Methodologically, this process is flawed. The guarantees gained from
+verifying the protocol have no bearing on the implementation if
+hardware designers must do a manual translation to RTL.
 
-Verified protocols are often described via Term Rewrite
-Systems (TRS) \adamc{citation for TRSes?}, which apply equational logic to describe systems at an
-interesting point on the spectrum between declarative and operational methods.
-It was shown by Dave
-\etal{}~\cite{DNA:CoherenceImplementation} that such a verified
-protocol could be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
-hardware description language based on Guarded Atomic Actions, a
-restriction of TRS~\cite{Hoe:TCAD}. In this way one reduces the correctness of
-the implementation to the correctness of the protocol, with no need
-for separate reasoning about an implementation. While the current Bluespec compiler
-does not support some implementation elements that are common in practice
-(\eg{} fairness guarantees, atomic actions running over multiple clock cycles),
-recent work has provided key insights toward lifting the
+Verified protocols are often described via Term Rewrite Systems (TRS),
+which apply equational logic to describe systems at an interesting
+point on the spectrum between declarative and operational methods.  It
+was shown by Dave \etal{}~\cite{DNA:CoherenceImplementation} that such
+a verified protocol could be transliterated into
+Bluespec~\cite{Bluespec:TFRG}, a hardware description language based
+on Guarded Atomic Actions, a restriction of TRS~\cite{Hoe:TCAD}. In
+this way one reduces the correctness of the implementation to the
+correctness of the protocol, with no need for separate reasoning about
+an implementation. While the current Bluespec compiler does not
+support some implementation elements that are common in practice
+(\eg{} fairness guarantees, atomic actions running over multiple clock
+cycles), recent work has provided key insights toward lifting the
 restrictions~\cite{DNA:CoherenceImplementation, Karczmarek}.  We
-believe that Bluespec and its future evolutions represent a
-very promising platform for productive implementation and
-verification of hardware components, including cache-coherent
-shared memories.
+believe that Bluespec and its future evolutions represent a very
+promising platform for productive implementation and verification of
+hardware components, including cache-coherent shared memories.
 

### 2014-02-05 paper/System.tex
Cut out text

@@
 protocol~\cite{MSI}. Logically, the system is organized in a tree
-structure, with the root connected to the main memory and each leaf interfaced
-with a processor (Figure~\ref{hier}). In the following, $\leaf(c)$ indicates
-that a cache $c$ is a leaf, and $\parent(c,p)$ indicates that $p$ is the parent of
-$c$.
+structure, with the root connected to the main memory and each leaf
+interfaced with a processor (Figure~\ref{hier}). In the following,
+$\leaf(c)$ indicates that a cache $c$ is a leaf, and $\parent(c,p)$
+indicates that $p$ is the parent of $c$.
 
-Input and output happens at the leaves in the form of memory requests
-issued from and responses issued to the associated processor. These will be
-represented by two mappings, one from $\mathbb{N}$ to requests and the other
-from $\mathbb{N}$ to responses. For each leaf $c$, $c.\pos$ represents the
-position of the next request to process; it is set to $0$ initially.
-\adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
-\adamc{I hope the diagram explains what these mappings mean, and in general what
-the tree structure really means.}
+Input and output happens at the leaves in the form of a stream of
+memory requests issued from and responses issued to the associated
+processor. These will be represented in Coq as two mappings each with
+integer ($c.\pos$) indicating how many elements in the stream we have
+handled so far.
+
+%% \adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
+%% \adamc{I hope the diagram explains what these mappings mean, and in general what
+%% the tree structure really means.}
 
@@
 which are in the \Mo{} state.
-
-%Of course, all these invariants have to be enforced by the protocol.
 

### 2014-02-06 paper/Abstract.tex
this the the correct version

@@
 Correct implementation of cache coherent systems is of high importance in
-industrial use. Model-checking is inadequate for verifying a large system with
-hierarchical caches because of its size and complexity. We provide a
-theorem-prover verified proof for the safety properties of a size-parameterized
-hierarchical MSI-based coherent system. The system is specified in terms of
-atomic transitions accessing only local state.  Previously, it has been shown
-that these atomic transitions can be translated automatically into efficient
-hardware. Verifying the specification based on atomic transitions therefore
-guarantees the correctness of the actual implementation.
-The proof is built in a modular fashion, identifying key invariants common with
-other protocols like MESI, MOSI, \etc. We informally show how these other
-protocols can be proved reusing modules from the proof of the MSI protocol.
+industrial use. We first provide a concise mathematical definition of a cache
+coherent system as one which obeys store atomicity. We give a Coq-based proof
+of an MSI cache coherence protocol for an infinite set of cache hierarchies. We
+specify a set of invariants which imply store atomicity, enabling other
+protocols to be verified for store atomicity automatically as long as they
+satisfy the invariants. Our key insight for proving these invariants is a set
+of fundamental properties.  We believe that these fundamental properties are
+true for many other invalidation-based protocols with no silent drops -- we
+show that informally for MESI, MOSI and MOESI protocols.
+
+%Previously, it has been shown
+%that these atomic transitions can be translated automatically into efficient
+%hardware. Verifying the specification based on atomic transitions therefore
+%guarantees the correctness of the actual implementation.
+%The proof is built in a modular fashion, identifying key invariants common with
+%other protocols like MESI, MOSI, \etc. We informally show how these other
+%protocols can be proved reusing modules from the proof of the MSI protocol.

### 2014-02-06 paper/Introduction.tex
this the the correct version

--- 2014-02-05 paper/Introduction.tex
@@
 \section{Introduction}
-\label{Sec:Introduction}
+\label{sec:Introduction}
 
@@
 abstraction efficiently, many hardware systems implement \emph{cache
-coherence protocols}.
+coherence implementation}.
 
 Since the correctness of this abstraction is paramount in the correct
-functioning of the system, significant effort is expended in making sure
-that the cache coherence engine is correct. In practice, verification proceeds
+functioning of the system, significant effort is expended in making sure that
+the cache coherence engine is correct. In practice, verification proceeds
 in two steps. First, a cache coherence protocol is constructed and
@@
 framework. We describe our MSI protocol example and its implementation
-in Section~\ref{Sec:System}. In Sections~\ref{Sec:Safety} and
-\ref{Sec:Liveness}, we discuss the structuring of our proof framework and
-verification of the base protocol. We show implementation refinements
-in Section~\ref{Sec:Refinements} and conclude with a general
-discussion of the applicability of our technique.
+in Section~\ref{Sec:System}. In Section~\ref{Sec:ProofStructure} we discuss
+the structuring of our proof framework and verification of the base
+protocol. We show implementation refinements in
+Section~\ref{Sec:Refinements} and conclude with a general discussion
+of the applicability of our technique.

### 2014-02-06 paper/System.tex
this the the correct version

--- 2014-02-05 paper/System.tex
@@
 \section{MSI-based cache coherence protocol}
-\label{Sec:System}
+\label{sec:System}
 
@@
 
-We have dedicated channels between each parent-child pair. The three
-channels are indexed by the name of the child that uses that channel
-(denoted by $\textit{ch}[c]$ for the \textit{ch} channel associated
+We have dedicated channels between each parent-child pair. The three channels
+are indexed by the name of the child that uses that channel (denoted by
+$\textit{ch}[c]$ for the \textit{ch} channel associated
 with cache $c$).  Figure~\ref{format} shows the message formats
-transmitted in the 3 channels between caches. In each of the channels
-$ch$, a message $m$ can be enqueued ($\enq(ch, m)$) or dequeued
-($\deq(ch)$), and the channels can be examined for the first element
-($\first(ch)$) or for availability of any message in the channel
-($\avail(ch)$).  Each of these channels start out empty, \ie{} $\forall
-ch, \neg \avail(ch)$, initially.
+transmitted in the 3
+channels between caches. In each of the channels $ch$, a message $m$
+can be enqueued ($\enq(ch, m)$) or dequeued ($\deq(ch)$), and the
+channels can be examined for the first element ($\first(ch)$) or for
+availability of any message in the channel
+($\avail(ch)$). Each of these channels start out empty, \ie{} $\forall ch,
+\neg \avail(ch)$, initially.
 
@@
 which $p$ is waiting for $c$ to downgrade. Both \wait{} and \dwait{} are
-\False{} initially for all caches and addresses,
++\False{} initially for all caches and addresses,
 because no cache is waiting for any response from either its child or its
@@
 Figure~\ref{childextra} shows transitions that happen in a non-root
-cache due to the presence of \emph{voluntary} responses, which model
-eviction from a cache. This eviction usually causes the permission
-state of the cache for that address to go to $\In$, while we allow any
-downgrade to take place voluntarily.  We use the same \cpResp{} FIFO
-for such a message. Such a voluntary response for an address can take
-place only if the cache is not waiting for a response from its parent
-for that address (ChildVolResp).
+cache due to the presence of \emph{voluntary} responses, which model eviction
+from a cache to make room for a new location. This eviction usually
+causes the permission state of the cache for that address to go to $\In$, while
+we allow any downgrade to take place voluntarily.  We use the same \cpResp{}
+FIFO for such a message. Such a voluntary response for an address can take
+place only if the cache is not waiting for a response from its parent for
+that address (ChildVolResp).
 
-Because of the presence of voluntary responses, a child may have already
+Because of the presence of voluntary response, a child could have already
 downgraded to a state required by a downgrade request from its parent. In such
-a case, the child drops the downgrade request (ChildDropReq).
+a case, the child simply drops the downgrade request (ChildDropReq).
 
@@
 & $p.\dwait[c][q.\addr] = \False \wedge p.\dstate[c][q.\addr] \le q.\from \wedge$\\
-& $\compat(p, c, q.\addr, q.\myto) \wedge q.\myto \le p.\state[q.\addr]$\\
+& $\compat(p, c, q.\addr, q.\myto) \wedge q.\myto \le p.\dstate[q.\addr]$\\
 \hline
@@
 \hline
-Action: & $\enq(\cpResp[c], \langle c.\state[a], x, a,
+Action: & $\enq(\cpResp[c], \langle c.\state[a], x, a,%$\\
+%& $\;\;\;%
 \myif c.\state[a] = \Mo \mythen c.\data[a] \myelse \_\rangle)$;\\
@@
 
-Note that the atomic transitions in Figure~\ref{trans} access only
-local states, \ie{} a transition can read or write states only
-corresponding to a single cache and/or the channel the cache is
-connected to. This restriction allows such a system to be implemented
-directly into hardware. Bluespec SystemVerilog (BSV), for instance,
-directly converts these transitions into efficient synchronous
-hardware. Though these transitions logically happen one by one, BSV
-executes several transitions simultaneously in a single clock cycle,
-as long as they do not access the same state, \ie{} do not
-\emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
-
-%ndave: I know this is mildly relevant, but it's making the current
-%hardware implementation seem worse and it's not adding any actual
-%knowledge to the reader who doesn't care about this level of detail
-
-%%  Sometimes,
-%% these transitions may not be amenable to single cycle implementations
-%% in hardware. For example, writing a memory can actually take several
-%% hardware clock cycles, though logically it is a single transition.
-%% Karczmarek \etal~\cite{Karczmarek} has provided a scheme to convert
-%% atomic transitions into synchronous hardware in which each transition
-%% can span several clock cycles.
+Note that the atomic transitions in Figure~\ref{trans} access only local states,
+\ie{} a transition can read or write states only corresponding to a single cache
+and/or the channel the cache is connected to. This restriction allows such a
+system to be implemented directly into hardware. Bluespec System Verilog (BSV),
+for instance, directly converts these transitions into efficient synchronous
+hardware. Though these transitions logically happen one by one, BSV executes
+several transitions simultaneously, as long as they do not access the same
+state, \ie{} do not \emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
+Sometimes, these transitions (both the guards and the actions) may not be
+amenable to single cycle implementations in hardware. For example, writing a
+register array can actually take several hardware clock cycles, though
+logically it is a single transition.  Karczmarek \etal~\cite{Karczmarek} has
+provided a scheme to convert atomic transitions into synchronous hardware in
+which each transition can span several clock cycles.

### 2014-02-06 paper/Abstract.tex
this the the correct version

@@
 satisfy the invariants. Our key insight for proving these invariants is a set
-of fundamental properties.  We believe that these fundamental properties are
-true for many other invalidation-based protocols with no silent drops -- we
-show that informally for MESI, MOSI and MOESI protocols.
+of fundamental properties.  We believe that these fundamental properties hold
+for many other invalidation-based protocols with no silent drops -- we show
+that informally for MESI, MOSI and MOESI protocols.
 

### 2014-02-06 paper/Abstract.tex
this the the correct version

@@
-Correct implementation of cache coherent systems is of high importance in
-industrial use. We first provide a concise mathematical definition of a cache
-coherent system as one which obeys store atomicity. We give a Coq-based proof
-of an MSI cache coherence protocol for an infinite set of cache hierarchies. We
-specify a set of invariants which imply store atomicity, enabling other
-protocols to be verified for store atomicity automatically as long as they
-satisfy the invariants. Our key insight for proving these invariants is a set
-of fundamental properties.  We believe that these fundamental properties hold
-for many other invalidation-based protocols with no silent drops -- we show
-that informally for MESI, MOSI and MOESI protocols.
+%Correct implementation of cache coherent systems is of high importance in
+%industrial use.
+We provide a concise mathematical definition of a cache coherent system as one
+which obeys store atomicity. We give a Coq-based proof of an MSI cache
+coherence protocol for an infinite family of cache hierarchies. We specify a
+set of invariants which imply store atomicity, enabling other protocols to be
+verified for store atomicity automatically as long as they satisfy the
+invariants. Our key insight for proving these invariants is a set of
+fundamental properties.  We believe that these fundamental properties hold for
+many other invalidation-based protocols with no silent drops -- we show that
+informally for MESI, MOSI and MOESI protocols.
 

### 2014-02-06 paper/Background.tex
this the the correct version

@@
 \section{Correctness of a Cache Coherence Protocol}
-\label{Sec:Background}
+\label{Background}
 
@@
 
-Intuitively, the simplest view of shared memory is as a memory (array of addressible objects)
-multiplexed amongst the many individual clients/processors, which send read/write
-messages to the memory as needed. In
-this view, all operations in the system appear to have a single unique total
-ordering.
+Intuitively, the simplest view of shared memory is as an array of addressible
+objects in which a single entry can be read or written by one of the many
+processors at a time.  In this view, all operations in the system appear to
+have a single unique total ordering.
 
 All modern systems have caches between the processor and memory for faster
-memory access.  It is often also useful to permit some reorderings of requests
-from a processor to the memory to improve performance.  For instance, in the
-context of a processor with local store buffer, we may wish to allow a store to
-bypass a subsequent load request to a different location. The specification of
-which reorderings are permitted or restricted defines the memory consistency
-model.
+memory access. It is often also useful to reorder requests from a processor to
+the memory to improve performance. For instance, a load may bypass a previous
+store to a different location. The \emph{memory consistency model} describes
+the semantics of a system in the presence of caches and reorderings.
 
@@
 stating that for each location/address in memory, every read and write access
-to that location occurs in some total order, with each operation seeing the
-effects of previous operations and not later ones (in other words, each
-location behaves like an atomic register) and (b), a ``local'' reordering
-restriction between requests to different locations within each processor.
-Changing the memory model by modifying the reordering restrictions will not
-incur a reverification of the store-atomicity property.
+to that location from any processor, occurs in some total order, with
+each operation seeing the effects of previous operations and not later ones (in
+other words, each location behaves like an atomic register) and (b), a
+``local'' reordering restriction between requests to different locations within
+each processor. Changing the memory model by modifying the reordering
+restrictions will not incur a reverification of the store-atomicity property.
 
@@
 
+We will now provide the formal definition of store atomicity.
+
 \subsection{Formal Definition of Store Atomicity}
-We will model an abstract memory subsystem, which interfaces with a set of
-processors. The requests issued by the processors to the memory subsystem can
-be modeled using a function \reqFn, where $\reqFn(p, i)$ is the $i^{th}$ request
-issued by processor $p$.
+A memory subsystem can be modeled as the following function. Its input models a
+set of infinite number of requests ever sent by any processor, and its output
+models a set of (possibly infinite number of) responses ever sent by the memory
+subsystem to any processor.
 
-%Figure~\ref{req-resp} shows the
-%data-type corresponding to a request issued from the processor to the
-%memory subsystem.
+Each input to the memory subsystem is isomorphic to a function
+$\reqFn:(Processor*\mathbb{N}) @-> \Request$, where $\reqFn(p, i)$ models the
+$i^{th}$ request issued by processor $p$. An element of the range-set of
+\reqFn{}, \viz \Request{} contains fields (a) address (\addrQ) corresponding to
+the request, (b) \desc{} indicating whether the request is a load or a store
+and (c) data (\dataQ) containing the data supplied by the processor for stores.
 
-We define a record type modeling a request in terms of a set of named fields.
-A request contains the address (\addrQ) corresponding to the request, as well as a
-\desc{} indicating whether the request is a load or a store. A store request
-also contains the data supplied by the processor for that address (\dataQ). A
-response (modeled with a different record type) contains a label (\labelR)
-to identify the request uniquely, via a pair of a processor name and a processor-local
-sequence number.  In the case of a load request, the memory
-subsystem returns \dataR, the data returned by the memory subsystem for that
-address. The memory subsystem also returns $\timeR \in \mathbb{N}$, which
-intuitively represents the time at which the request is handled among the
-global set of handled requests from all the processors. The \timeR{} field is
-present solely as a specificational device to capture store atomicity, and it
-would not be present in a running system implementation.
+Each output of the memory subsystem contains a set (called \Response) of
+elements containing the fields (a) \labelR, containing a pair from the set
+$Processor*\mathbb{N}$, which is meant to identify a request uniquely, (b)
+\dataR, meant to model the data returned from the memory subsystem and (c)
+\timeR, a value from set $\mathbb{N}$, meant to represent the ``time'' at which
+the corresponding request (identified by \labelR) is handled by the memory
+subsystem. \timeR{} does not have a counterpart in a real implementation, it is
+present solely as a specificational device to capture store atomicity. Since a
+request corresponding to a response is identified using the field \labelR, for
+notational convenience, we will access the fields of the corresponding request
+directly from the response. For instance, $r.\addrQ$ for $r \in \Response$
+denotes the $(r.\labelR).\addrQ$, \etc.
 
-%We model request of a processor formally with the function
-%\reqFn. $\reqFn(p, i)$ returns the $i^{th}$ request issued by
-%processor $p$, \ie{} $\reqFn: (\textit{Processor} * \mathbb{N}) @->
-%\Request$.  The processor name and the position of the request within
-%the processor uniquely identifies a request.
-
-%\begin{figure}
-%\centering
-%\begin{subfigure}{5.5cm}
-%\begin{tabular}{|lp{4.5cm}|}
-%\hline
-%\multicolumn{2}{|c|}{\Request}\\
-%\hline
-%\addrQ:& Address associated with the request\\
-%\desc:& Request type (\Ld{} or \St)\\
-%\dataQ:& Data associated with a \St{} request\\
-%\hline
-%\end{tabular}
-%\end{subfigure}~~~~
-%\begin{subfigure}{5.5cm}
-%\begin{tabular}{|lp{4.5cm}|}
-%\hline
-%\multicolumn{2}{|c|}{\Response}\\
-%\hline
-%\labelR:& Label identifying the request corresponding to this response\\
-%\timeR:& Natural number representing global order of (processing) requests\\
-%\dataR:& Data returned by the memory subsystem for a \Ld{} request\\
-%\hline
-%\end{tabular}
-%\end{subfigure}
-%\caption{Data types for request and response}
-%\label{req-resp}
-%\end{figure}
-
-We model a single protocol execution as a set of all the requests and responses
-that sent in the course of that execution.
-The fundamental parameter is a (potentially infinite) set \Response{} of all responses
-that the memory sends to processors.  From here, the associated requests follow by
-applying the function $\reqFn$ to map each response to the request that provoked it.
-Responses should obey the following sanity conditions. 
-\begin{defn} (No label is reused across multiple distinct responses.)
-%Since \dataR{} field is
-%irrelevant for store requests, so uniqueness of responses does not depend on
-%\dataR{} field for stores.)
+If \reqFn{} is an input to the memory subsystem, and the set \Response{} is the
+produced output, then the memory subsystem is said to be store atomic if it
+obeys Theorem \fullref{storeAtomicity} defined as follows. We use the word
+response to denote an element in the set \Response{}.
+%
+\begin{defn} (Responses have unique labels)
 \small
 \begin{multline*}
-\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),\\
-r_1.\labelR = r_2.\labelR @-> r_1.\timeR = r_2.\timeR \wedge\\
-(\reqFn(r_1.\labelR) = \Ld @-> r_1.\dataR = r_2.\dataR)
+\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),
+r_1.\labelR = r_2.\labelR @-> \\ r_1.\timeR = r_2.\timeR \wedge
+r_1.\dataR = r_2.\dataR
 \end{multline*}
@@
 \end{defn}
-
-\begin{defn} (No response can share a \timeR{} value with another store response.)
+%
+\begin{defn} (No response can share a \timeR{} value with another store
+response for the same address.)
 \small
@@
 r_1.\timeR = r_2.\timeR @-> \\
-\reqFn (r_1.\labelR).\addrQ = \reqFn (r_2.\labelR).\addrQ @->\\
-\reqFn (r_1.\labelR).\desc = \St @->
+r_1.\addrQ = r_2.\addrQ @->
+r_1.\desc = \St @->
 r_1.\labelR = r_2.\labelR
@@
 \end{defn}
-
-In addition, the global ordering of responses to the same processor should not
-violate the local ordering of the corresponding requests.
-
-\begin{defn}
+%
+\begin{defn} (The global ordering of responses to the same processor should not
+violate the ordering of the corresponding requests within the processor)
 \small
@@
 \forall (r_1, r_2 \in \Response), r_1.\labelR.\fst = r_2.\labelR.\fst @->\\
-r_1.\labelR.\snd < r2.\labelR.\snd @-> r_1.\timeR \le r_2.\timeR
+r_1.\labelR.\snd < r_2.\labelR.\snd @-> r_1.\timeR \le r_2.\timeR
 \end{multline*}
@@
 \end{defn}
-
+%
+\begin{defn} (If a response to a particular request from a processor exists in the \Response{} set,
+then responses for all previous requests from that processor must exist)
+\small
+\begin{multline*}
+\textit{allPrevious} :=
+\forall (r \in \Response), \forall i < r.\labelR.\snd, \\
+\exists (r' \in \Response) \text{ such that } r'.\labelR.\fst = r.\labelR.\fst \wedge r'.\labelR.\snd = i
+\end{multline*}
+\label{allPrevious}
+\end{defn}
+%
 This brings us to the main condition for store atomicity, which states that the
 data returned for a load response corresponds to the \emph{last earlier} store
-request from any processor that has been handled for that address,
-\emph{last} and \emph{earlier} being defined according to the field \timeR{}
-returned in the responses.  This proposition is indicated by
-\textit{lastEarlierValue}$(a, t, d)$ for a load request for address $a$ handled
-at time $t$, returning data $d$.  In case no store has happened earlier for a
-particular address when a load is processed, then the load response returns the
-initial value for the address, the whole proposition indicated by
-\textit{initialValue}$(a, t, d)$.
+request from any processor that has been handled for that address, \emph{last}
+and \emph{earlier} being defined according to the field \timeR{} returned in
+the responses.  This proposition is indicated by \textit{lastEarlierValue}$(r)$
+for a load request $r$.  In case no store has happened earlier for a particular
+address when a load is processed, then the load response returns the initial
+value for the address ($\initData(a)$), the whole proposition indicated by
+\textit{useInitialValue}$(r)$.
 \begin{defn}
@@
 \begin{multline*}
-\textit{initialValue}(a, t, d) := d = \initData (a) \; \wedge \\
-\forall (r' \in \Response), 0 \le r'.\timeR < t @->\\
-\neg (\reqFn (r'.\labelR).\addrQ = a \wedge \reqFn (r'.\labelR).\desc = \St)
+\textit{useInitialValue}(r) := r.\dataR = \initData(a) \; \wedge \neg \exists (r' \in \Response), \\
+r'.\timeR < t \wedge r'.\addrQ = r.\addrQ \wedge r'.\desc = \St
 \end{multline*}
-\label{initialValue}
+\label{useInitialValue}
 \end{defn}
-
+%
 \begin{defn}
@@
 \begin{multline*}
-\textit{lastEarlierValue}(a, t, d) := \exists (r_m \in \Response),\\ \reqFn (r_m.\labelR).\dataQ = d \wedge 
-\reqFn (r_m.\labelR).\desc = \St \; \wedge\\ \reqFn (r_m.\labelR).\addrQ = a \wedge
-\forall (r' \in \Response), r_m.\timeR < r'.\timeR < t @->\\
-\neg (\reqFn (r'.\labelR).\addrQ = a \wedge \reqFn (r'.\labelR).\desc = \St)
+\textit{lastEarlierValue}(r) := \exists (r_m \in \Response), r_m.\dataQ = r.\dataQ \; \wedge 
+\\ r_m.\desc = \St \wedge r_m.\addrQ = r.\addrQ \wedge
+\neg \exists (r' \in \Response), \\
+r_m.\timeR < r'.\timeR < t \wedge r'.\addrQ = r.\addrQ \wedge r'.\desc = \St
 \end{multline*}
@@
 \end{defn}
-
-Using the above definitions, we can define the store atomicity theorem as follows:
-\begin{thm}
+%
+Using the above definitions, we can define the store-atomicity theorem as follows:
+\begin{thm}\textit{storeAtomicity}:
 \small
 \begin{multline*}
-\textit{storeAtomicity} := 
 \textit{uniqRespLabels} \wedge
-\textit{uniqStTimes} \wedge \textit{localOrdering} \; \wedge \\
-(\forall (r \in \Response), \mylet q := \reqFn(r.\labelR) \myin q.\desc = \Ld \; @-> \\
-(\textit{initialValue}(q.\addrQ, r.\timeR, r.\dataR) \vee \textit{lastEarlierValue}(q.\addrQ, r.\timeR, r.\dataR)))
+\textit{uniqStTimes} \wedge \textit{localOrdering} \wedge \textit{allPrevious} \; \wedge \\
+(\forall (r \in \Response), r.\desc = \Ld \; @-> \\
+(\textit{useInitialValue}(r) \vee \textit{lastEarlierValue}(r)))
 \end{multline*}

### 2014-02-06 paper/System.tex
this the the correct version

@@
 
+Each node contains the fields mentioned in Figure \ref{hier}. We will
+explain what the fields mean in the following discussion.
+
+\begin{figure}
+\includegraphics[scale=.5]{hier}
+\caption{A cache hierarchy, and the fields in each node}
+\label{hier}
+\end{figure}
+
 Input and output happens at the leaves in the form of memory requests
@@
 from $\mathbb{N}$ to responses. For each leaf $c$, $c.\pos$ represents the
-position of the next request to process; it is set to $0$ initially.
+sequence number of the next request to process; it is set to $0$ initially.
 \adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}

### 2014-02-06 paper/System.tex
this the the correct version

@@
 structure, with the root connected to the main memory and each leaf interfaced
-with a processor (Figure~\ref{hier}). In the following, $\leaf(c)$ indicates
+with a processor (Figure~\ref{hier}). The system is parameterized by the shape
+of the tree, and each node can have any number of children independent of the
+number in other nodes. In the following, $\leaf(c)$ indicates
 that a cache $c$ is a leaf, and $\parent(c,p)$ indicates that $p$ is the parent of
@@
 The behavior of the system is described via the atomic state transitions, which
-are effectively a nondeterministic operational semantics. These
-transitions have two parts: (a) the guard dictating when the transition can
-fire, and (b) the action defining how the state of the system changes when the
-transition happens. Logically, these transitions happen one at a time.
-Figure~\ref{trans} gives the atomic transitions for the MIS protocol.  The
-actions in the transitions in Figure \ref{trans} show only the specific states
-that change, and other state components are assumed to remain the same.
-The notation ($x \Leftarrow
-y$;) in a transition indicates setting of a specific state $x$ of the overall
-system to value $y$.
+are effectively a nondeterministic operational semantics. These transitions
+have two parts: (a) the guard dictating when the transition can fire, and (b)
+the action defining how the state of the system changes when the transition
+happens. Logically, these transitions happen one at a time.  Figure~\ref{trans}
+gives the atomic transitions for the MIS protocol.  The actions in the
+transitions in Figure \ref{trans} show only the specific states that change,
+and other state components are assumed to remain the same.  The notation ($x
+\Leftarrow y$;) in a transition indicates setting of a specific state $x$ of
+the overall system to value $y$.
+
+We first give an intuition on how the transitions work in the common case.
+A cache can decide, unsolicitedly \xxx{is this a word?}, to upgrade its state,
+in which case it sends an upgrade request to its parent. The parent then makes
+a local decision on whether to send a response to the requesting child or not
+(based on its directory approximation, and its own state). If its own state is
+lower the requested upgrade, then it can not handle the request. The parent
+must instead, decide to upgrade its state -- the transition for doing so has
+nothing to do with the incoming request.  Once the parent's state is not lower
+than the requested upgrade, it makes sure that the rest of its children are
+``compatible'' with the requested upgrade. If not, the parent must, send
+requests to the uncompatible children to downgrade -- again the transition for
+doing so has nothing to do with the incoming request. If the children are
+compatible and its own state is higher, then the request can be handled and a
+response is sent. Each of the transitions is explained in greater detail below.
 
@@
 
-Figure~\ref{childextra} shows transitions that happen in a non-root
-cache due to the presence of \emph{voluntary} responses, which model eviction
-from a cache to make room for a new location. This eviction usually
+Figure~\ref{childextra} shows transitions that happen in a non-root cache due
+to the presence of \emph{voluntary} responses, which model eviction of a
+location from a cache to make room for a new location. This eviction usually
 causes the permission state of the cache for that address to go to $\In$, while
@@
 FIFO for such a message. Such a voluntary response for an address can take
-place only if the cache is not waiting for a response from its parent for
-that address (ChildVolResp).
+place only if the cache is not waiting for a response from its parent for that
+address (ChildVolResp).
 

### 2014-02-06 paper/Intuition.tex
this the the correct version

--- 2014-02-05 paper/Intuition.tex
@@
 \section{Proof of MSI protocol being preserving Store atomicity}
-\label{Sec:Safety}
+\label{safety}
 
@@
 \fullref{storeAtomicity} for a system implementing the MSI protocol of
-Figure~\ref{trans}. We reduce the whole system proof to that of a set
-of invariants between any parent-child pair. These invariants in turn
-depend on two smaller invariants which encapsulate the core aspects of
-correctness of the MSI protocol and allow the large scope of verified
-hardware refinements we present later (Section~\ref{Sec:Refinements}).
+Figure~\ref{trans}.
 
-The invariants hold after any number of executions of the transactions
-starting from the initial state. We use the term time $t$ to specify
-the global state reached after firing $t$ transitions.
+We first reduce the proof of store atomicity to 4 invariants, presented in
+Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
+invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
+ invariants is because they are not specific to MSI protocol
+alone; any other protocol which obeys these invariants obeys Theorem
+\fullref{storeAtomicity}. We finally present the key insight which we
+discovered for proving the second-level invariants in the form of 3 invariants.
+We believe that these properties are common in most, if not all
+invalidation-based protocols which do not perform silent drops (\ie{} protocols
+in which a cache that downgrades its state informs some other cache about the
+downgrade). We informally show why this is the case using the example of MESI,
+MOSI and MOESI in Section \ref{mosi}.
+
+%We use the terms write to mean handling a store request, \ie{} executing the
+%StoreReq transition and read to mean handling a load request, \ie{} executing
+%the LoadReq transition.
+
+These invariants hold after any number of executions of the transactions starting
+from the initial state. We use the term time $t$ to specify the global state
+reached after firing $t$ transitions.
 
@@
 
-\subsection{Reducing correctness to Node-level Invariants}
-
-Store atomicity is a whole system property. Intuitively our
-protocol is correct because it guarantees at most one cache may
-write at a time, and that any cache permitted to read has
-the ``latest'' value. We state this formally with the following 4
-invariants.
+\subsection{First level invariants}
+Intuitively our protocol is correct because it guarantees that at most one
+cache can handle a store request at a time, and that any cache that attempts to
+handle a load request has the ``latest'' value. We can state this formally with
+the following 4 invariants.
 
@@
 state for all its children is at most \Sh.
+\label{clean}
 \end{defn}
@@
 \textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
-that neither is an ancestor of the other in the cache tree
-hierarchy (where ancestor relation is the reflexive-transitive closure of the parent relation), for each
-address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are \textit{compatible}.
+that neither is an ancestor of the other in the cache tree hierarchy (where
+ancestor relation is the reflexive-transitive closure of the parent relation),
+for each address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are
+\textit{compatible}.
 \label{nonAncestorCompatible}
@@
 \begin{defn}
-\textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible} iff
-either $x_1$ or $x_2$ is \Mo{}, then the other is \In. Note that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also compatible if $x_1, x_2$ are compatible.
+\textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible}
+iff either $x_1$ or $x_2$ is \Mo{}, then the other is \In.
 \label{compatible}
@@
 
-%\begin{inv}
-%\textit{processReq}: Load and store requests can be processed at a leaf cache
-%only if it is at least in the \Sh{} and \Mo{} states, respectively, for that
-%address.
-%\label{processReq}
-%\end{inv}
-%
-%\begin{inv}
-%\textit{noReqAgain}:
-%\item For a leaf cache $c$, $c.\pos$ changes only on processing a request. In
-%that case, $c.\pos$ gets incremented.
-%\label{noReqAgain}
-%\end{inv}
+It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
+compatible if $x_1, x_2$ are compatible.
+
+\begin{inv}
+\textit{processReq}: Load and store requests can be processed only at a leaf
+node which is at least in the \Sh{} and \Mo{} states, respectively, for that
+address.
+\label{processReq}
+\end{inv}
+
+\begin{inv}
+\textit{noReqAgain}:
+\item For a leaf cache $c$, $c.\pos$ changes only on processing a request. In
+that case, $c.\pos$ gets incremented.
+\label{noReqAgain}
+\end{inv}
 
 Given these invariants, we will show how Theorem~\fullref{storeAtomicity}
-follows. From the transitions of Figure \ref{trans}, a store may happen only
-when a leaf cache is in \Mo{} state for that
-address%(Invariant~\textit{processReq}).
-Leaf caches cannot be ancestors of any
-other node, so by Invariant \fullref{nonAncestorCompatible}, if a leaf cache is
-in \Mo{} state, all other leaf caches will be in \In{} state preventing the
-possibility of processing load and store requests for the same address at the
-same time. Thus, Definition~\fullref{uniqStTimes} holds.
+follows.
+%From the transitions of Figure \ref{trans}, a store may happen only
+%when a leaf cache is in \Mo{} state for that
+%address (Invariant~\textit{processReq}).
+Leaf caches cannot be ancestors of any other node, so by Invariant
+\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
+leaf caches will be in \In{} state preventing the possibility of processing
+load and store requests (because of Invariant \ref{processReq}) for the same
+address at the same time at different leaf caches. This satisfies
+Definition~\fullref{uniqStTimes}.
 
-Once a response has been produced for a request $i$, the cache increments the
-counter for the request it processes next. Thus there is at most one response
-to each request, satisfying Definition~\fullref{uniqRespLabels}. As leaf caches
-have no children, a leaf cache in at least \Sh{} state for address $a$ is
-\clean{} for that address. Since processing a load request requires the cache
-to be at least in the \Sh{} state, its response, which supplies the data for
-that address present in the cache, supplies the latest value by
-Invariant~\fullref{latestValue}.  Thus, the final condition for
-Theorem~\fullref{storeAtomicity} is also satisfied. All that remains is to show
-how these invariants hold.
+It is easy to see that Definition \fullref{uniqRespLabels}, Definition
+\fullref{localOrdering} and Definition \fullref{allPrevious} directly follows
+from Invariant \fullref{noReqAgain}.  As leaf caches have no children, one in
+at least \Sh{} state for address $a$ is \clean{} for that address. Since
+processing a load request requires the cache to be at least in the \Sh{} state,
+its response, which supplies the data for that address present in the cache,
+supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
+conditions for Theorem~\fullref{storeAtomicity} is satisfied given Invariants
+\ref{latestValue} to \ref{noReqAgain}.
 
-\subsection{Reducing to Pair-wise Invariants}
-\label{pairwise}
+\subsection{Second-level Invariants}
+\label{second-level}
 
-Correctness of Invariants \fullref{localCompatibility} and \fullref{latestValue} can be
-proved using the following invariants on a parent-child pair.
+While Invariant \fullref{processReq} and \fullref{noReqAgain} can be seen
+directly from the MSI transitions, correctness of Invariants
+\fullref{localCompatibility} and \fullref{latestValue} require further
+decomposition. We present 4 more invariants for proving those two invariants.
 
@@
 \begin{inv}
-\textit{localCompatibility}:
-\begin{enumerate}
-\item The state of a cache is never less than its directory entry for any of
+\textit{localCompatibility1}:
+The state of a cache is never less than its directory entry for any of
 its children.
-\item Let $c_1$ and $c_2$ be distinct caches with the same parent $p$. Then,
-$p.\dstate[c_1][a]$ and $p.\dstate[c_2][a]$ are \textit{compatible}.
-\end{enumerate}
-\label{localCompatibility}
+\label{localCompatibility1}
+\end{inv}
+\begin{inv}
+\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
+same parent $p$. Then, $c_1.\state[a]$ and $c_2.\state[a]$ are
+\textit{compatible}.
+\label{localCompatibility2}
 \end{inv}
@@
 $t_r$ the time of receiving the data), no cache can process a store request for
-$a$.
+$a$, and the data must be sent from a cache which is \textit{clean}.
 \label{noTransitWrite}
@@
 
-For any two caches $c$ and $p$, such that $p$ is an ancestor of $c$ in
-the cache tree hierarchy,
-
-Invariant \fullref{conservative} and the first part of Invariant
-\fullref{localCompatibility} defines a $\ge$ relation between the states of a
-parent and its child for any address.  Since the ancestor relation is the
-reflexive-transitive closure of the parent relation, the $\ge$ relation is
-preserved between the states of two caches $p$ and $c$ where $p$ is the
-ancestor of $c$.
+Invariants \fullref{conservative} and \fullref{localCompatibility1} defines a
+$\ge$ relation between the states of a parent and its child for any address.
+Since the ancestor relation is the reflexive-transitive closure of the parent
+relation, the $\ge$ relation is preserved between the states of two caches $p$
+and $c$ where $p$ is the ancestor of $c$.
 
@@
 has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
-$c_1$ and $c_2$ respectively. For any address $a$, $x.\dstate[x_1][a]$ and
-$x.\dstate[x_1][a]$ are compatible by the second part of
+$c_1$ and $c_2$ respectively. For any address $a$, $x_1.\state[a]$ and
+$x_2.\state[a]$ are compatible by the second part of
 Invariant~\fullref{localCompatibility}. It follows that $x_1.\state[a]$ and
@@
 If $c$ was not clean at time $t$, it must have received data from either its
-parent or its child.  As can be seen in Figure~\ref{trans}, whenever data is
-sent either from a parent to a child, or from a child to the parent, the sender
-is \clean. By the induction hypothesis, it means that the sender has the latest
-value for the address. Because of Invariant~\ref{noTransitWrite}, no store
-request can be processed when the data is in transit, which means $c$ will have
-the latest value for address $a$ at time $t+1$. This establishes Invariant
-\fullref{latestValue}.
+parent or its child. And the sender is \clean{} by Invariant
+\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
+the latest value for the address. Again, because of
+Invariant~\ref{noTransitWrite}, no store request can be processed when the data
+is in transit, which means $c$ will have the latest value for address $a$ at
+time $t+1$. This establishes Invariant \fullref{latestValue}.
 
-\subsection{Proving parent-child hierarchy properties}
+\subsection{Key insight for proving second-level invariants}
 
 %ndave: This needs some repointing
+We present what we believe to be the key insight for proving any cache coherence protocol.
+In fact the request/response message semantics for the MSI protocol was designed to satisfy
+the following three invariants.
 
-We present two basic invariants that the MSI protocol of Figure~\ref{trans}
-obeys. In fact the request/response message semantics was designed to satisfy
-the following two invariants. We believe that most cache-coherence protocols
-would end up obeying these two invariants. With two invariants we can
-prove all the pair-wise properties of Section~\ref{pairwise}.
+\begin{inv}
+\textit{stateChange}:
+\begin{enumerate}
+\item A cache downgrades its state for an address iff it sends a response for that address
+\item A parent upgrades the directory entry for an address for one of its
+children iff it sends a response for that child and address
+\item The state or directory entry changes for an address only on sending and
+receiving an response for that address, and they change to the \myto-field of the response.
+\end{enumerate}
+\label{stateChange}
+\end{inv}
+
+Invariant \fullref{stateChange} ensures that a response message acts as a sync
+between a parent's directory entry for a child and the state of the child, as
+both the sender and the receiver change their respective directory entry or
+state to the \myto-field of the response.
 
@@
 
-The MSI-based protocol in Figure~\ref{trans} is such that any change in either
-the state of a cache or its directory entry takes place only on sending or
-receiving a response. The parent upgrades the corresponding directory entry on
-sending a response to a child and the child downgrades its state on
-sending a response to the parent.
+\begin{figure}
+\centering
+\includegraphics[scale=.5]{resps}
+\caption{An example showing a sequence of responses ping ponging between a child and its parent}
+\label{resps}
+\end{figure}
 
-To prove Invariant \fullref{conservative} at some time $t$, consider the case
-when a child cache $c$ has sent to or received from the parent $p$ some message
-before $t$. Consider the last such message $m$ sent by or received at $c$ at
-$t'$. Let's say it is sent by $c$ at $t'$, not received. Then $p$ can not have
-received any message from $c$ after sending $m$ and before $t$ in order to not
-violate Invariant \fullref{respFIFO}.  Instead, if $c$ receives $m$ at $t'$,
-$p$ will not receive any message after sending $m$ and before $t$ in order to
-not violate Invariant \fullref{noCross}. Thus, $p$ will not receive any message
-from $c$ after sending or receiving $m$ and before $t$.  Let $t''$ be the time
-at which $p$ has sent or received $m$. Whether $p$ sends or receives $m$,
-$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ because
-the response messages sync up the next-state according to the transitions in
-Figure \ref{trans}. Since $p$ can not receive any message between $t''+1$ and
-$t$, it can only upgrade $p.\dstate[c][a]$. $c.\state[a]$ does not change
-between $t'+1$ and $t$ as it never changes state. Therefore, Invariant
-\fullref{conservative} holds at $t$.
+Figure \ref{resps} shows an example of the effect of Invariants
+\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
+entry of the parent or the state of the child changes on sending and receiving
+a response and the new state is showed at the points of sending or receiving
+responses.
+Both the cache and its parent follow the same sequence state and directory
+entry changes; just that the receiver always lags slightly behind the sender.
+The receiver can not send any response before receiving because of
+\fullref{noCross}, thereby eliminating any scenarios where the lagging makes
+reasoning complicated.
 
-If no message has been sent or received by $c$ before $t$, then $p$ wouldn't
-have received any message sent by $c$ before $t$. So $p.\dstate[c][a]$ at time
-$t \ge p.\dstate[c][a]$ at time $0$ . Since the initial values of the directory
-and the child's state are both set to \In{} Invariant \textit{conservative}
-holds.
+Using these three invariants (Invariant \ref{stateChange} to \ref{respFIFO}, it
+is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
+back to Figure \ref{resps}, we can see that at all times, either there is no
+response in flight, or only responses from parent in flight, or responses from
+child in flight. A response from child downgrades the state and a response from
+parent upgrades the directory. Since the other side simply lags behind,
+Invariant \fullref{conservative} holds.
 
-To prove Invariant \fullref{localCompatibility}, it is enough to show that a
-$c.\state[a]$ is upgraded on receiving a response for $a$ from $c$'s parent $p$
-and $p.\dstate[c][a]$ is downgraded on receiving a response for $a$ from $c$.
-If this is true, then Invariant \fullref{localCompatibility} will not be
-violated on receiving a response. The transitions of Figure \ref{trans} ensures
-that Invariant \fullref{localCompatibility} is not violated when sending a
-response. Since whenever the state or directory changes, the appropriate
-response is sent or received, Invariant \fullref{localCompatibility} holds.
+Proving the other 3 invariants in Section \ref{second-level} requires using
+more properties of the MSI transistions of Figure \ref{trans}, though it is
+straightforward.
 
-If the value of the state of a child at the time of sending a response is the
-same as the value of the directory for that child at the parent at the time of
-receiving the response, it follows straightaway that the parent downgrades its
-directory on receiving a response from the child. This is because a child
-downgrades on sending a response and the next-states are synced on receiving
-the response.  This proof uses exactly the same arguments presented above to
-prove Invariant \fullref{conservative} -- we prove by considering the last
-message sent or received at the child before $t$ (and the case when no message
-has been sent or received). Similarly, we can show that a child upgrades its
-state on receiving a response, satisfying the conditions in the previous paragraph.
-
-Invariant \fullref{noTransitWrite} can be proved as follows. Consider a
-response sent from a parent $p$ at time $t$ received by its child $c$ at time
-$t'$. No other response can be received by the parent at time $t_p$ such that
-$t+1 \le t_p \le t'$ in order to not violate Invariant \fullref{noCross}, so
-$p.\dstate[c][a]$ at $t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can
-only upgrade its directory for $c$. Similarly, no response can be sent by the
-child at time $t_c$ such that $t \le t_c \le t'$ for the same reason and hence
-$c.\state[a]$ at $t \le c.\state[a]$ at $t'$ since the child can only downgrade
-its state.  This means that during $t_p$ such that $t+1 \le t_p \le t'$,
-$p.\dstate[c][a] > \In$ and (hence $p.\state[a] > \In$ by the first part of
-Invariant \fullref{localCompatibility}) since the parent upgraded its directory
-for $c$ at $t$.  Similarly, during $t_c$ such that $t \le t_c \le t'$,
-$c.\state[a] < \Mo$ since the child upgrades its state at $t'$. We have already
-shown that (in the proof of Invariant \fullref{latestValue}) if any cache is
-not in \In{} state, no non-descendent cache can process a store request. In
-this case, during the times between $t$ and $t'$, $p$ is not in \In{} state.
-Among $p$'s directory entries, all of them save the entry for $c$ is not in
-\Mo{} state, so none of the descendents of $p$'s children, save $c$'s
-descendents can process a store request. During the times between $t$ and $t'$,
-$c$'s state is $< \Mo{}$, so neither $c$ nor its descendents can process a store
-request, proving Invariant \fullref{noTransitWrite}.
+%To prove Invariant \fullref{conservative} at some time $t$, consider the case
+%when a child cache $c$ has sent to or received from the parent $p$ some message
+%before $t$. Consider the last such message $m$ sent by or received at $c$ at
+%$t'$. Let's say it is sent by $c$ at $t'$, not received. Then $p$ can not have
+%received any message from $c$ after sending $m$ and before $t$ in order to not
+%violate Invariant \fullref{respFIFO}.  Instead, if $c$ receives $m$ at $t'$,
+%$p$ will not receive any message after sending $m$ and before $t$ in order to
+%not violate Invariant \fullref{noCross}. Thus, $p$ will not receive any message
+%from $c$ after sending or receiving $m$ and before $t$.  Let $t''$ be the time
+%at which $p$ has sent or received $m$. Whether $p$ sends or receives $m$,
+%$p.\dstate[c][a]$ at $t''+1$ will be equal to $c.\state[a]$ at $t'+1$ because
+%the response messages sync up the next-state according to the transitions in
+%Figure \ref{trans}. Since $p$ can not receive any message between $t''+1$ and
+%$t$, it can only upgrade $p.\dstate[c][a]$. $c.\state[a]$ does not change
+%between $t'+1$ and $t$ as it never changes state. Therefore, Invariant
+%\fullref{conservative} holds at $t$.
+%
+%If no message has been sent or received by $c$ before $t$, then $p$ wouldn't
+%have received any message sent by $c$ before $t$. So $p.\dstate[c][a]$ at time
+%$t \ge p.\dstate[c][a]$ at time $0$ . Since the initial values of the directory
+%and the child's state are both set to \In{} Invariant \textit{conservative}
+%holds.
+%
+%To prove Invariant \fullref{localCompatibility}, it is enough to show that a
+%$c.\state[a]$ is upgraded on receiving a response for $a$ from $c$'s parent $p$
+%and $p.\dstate[c][a]$ is downgraded on receiving a response for $a$ from $c$.
+%If this is true, then Invariant \fullref{localCompatibility} will not be
+%violated on receiving a response. The transitions of Figure \ref{trans} ensures
+%that Invariant \fullref{localCompatibility} is not violated when sending a
+%response. Since whenever the state or directory changes, the appropriate
+%response is sent or received, Invariant \fullref{localCompatibility} holds.
+%
+%If the value of the state of a child at the time of sending a response is the
+%same as the value of the directory for that child at the parent at the time of
+%receiving the response, it follows straightaway that the parent downgrades its
+%directory on receiving a response from the child. This is because a child
+%downgrades on sending a response and the next-states are synced on receiving
+%the response.  This proof uses exactly the same arguments presented above to
+%prove Invariant \fullref{conservative} -- we prove by considering the last
+%message sent or received at the child before $t$ (and the case when no message
+%has been sent or received). Similarly, we can show that a child upgrades its
+%state on receiving a response, satisfying the conditions in the previous paragraph.
+%
+%Invariant \fullref{noTransitWrite} can be proved as follows. Consider a
+%response sent from a parent $p$ at time $t$ received by its child $c$ at time
+%$t'$. No other response can be received by the parent at time $t_p$ such that
+%$t+1 \le t_p \le t'$ in order to not violate Invariant \fullref{noCross}, so
+%$p.\dstate[c][a]$ at $t+1 \le p.\dstate[c][a]$ at $t'$, since the parent can
+%only upgrade its directory for $c$. Similarly, no response can be sent by the
+%child at time $t_c$ such that $t \le t_c \le t'$ for the same reason and hence
+%$c.\state[a]$ at $t \le c.\state[a]$ at $t'$ since the child can only downgrade
+%its state.  This means that during $t_p$ such that $t+1 \le t_p \le t'$,
+%$p.\dstate[c][a] > \In$ and (hence $p.\state[a] > \In$ by the first part of
+%Invariant \fullref{localCompatibility}) since the parent upgraded its directory
+%for $c$ at $t$.  Similarly, during $t_c$ such that $t \le t_c \le t'$,
+%$c.\state[a] < \Mo$ since the child upgrades its state at $t'$. We have already
+%shown that (in the proof of Invariant \fullref{latestValue}) if any cache is
+%not in \In{} state, no non-descendent cache can process a store request. In
+%this case, during the times between $t$ and $t'$, $p$ is not in \In{} state.
+%Among $p$'s directory entries, all of them save the entry for $c$ is not in
+%\Mo{} state, so none of the descendents of $p$'s children, save $c$'s
+%descendents can process a store request. During the times between $t$ and $t'$,
+%$c$'s state is $< \Mo{}$, so neither $c$ nor its descendents can process a store
+%request, proving Invariant \fullref{noTransitWrite}.
 

### 2014-02-06 paper/Intuition.tex
this the the correct version

@@
 \subsection{First level invariants}
+\label{firstLevel}
 Intuitively our protocol is correct because it guarantees that at most one
@@
 \subsection{Second-level Invariants}
-\label{second-level}
+\label{secondLevel}
 
@@
 directly from the MSI transitions, correctness of Invariants
-\fullref{localCompatibility} and \fullref{latestValue} require further
+\fullref{localCompatibility1}, \fullref{localCompatibility2} and \fullref{latestValue} require further
 decomposition. We present 4 more invariants for proving those two invariants.

### 2014-02-06 paper/Intuition.tex
this the the correct version

@@
 $c_1$ and $c_2$ respectively. For any address $a$, $x_1.\state[a]$ and
-$x_2.\state[a]$ are compatible by the second part of
-Invariant~\fullref{localCompatibility}. It follows that $x_1.\state[a]$ and
+$x_2.\state[a]$ are compatible by 
+Invariant~\fullref{localCompatibility2}. It follows that $x_1.\state[a]$ and
 $x_2.\state[a]$ are compatible since these values are smaller than the
@@
 
-Proving the other 3 invariants in Section \ref{second-level} requires using
+Proving the other 3 invariants in Section \ref{secondLevel} requires using
 more properties of the MSI transistions of Figure \ref{trans}, though it is

### 2014-02-06 paper/Abstract.tex
this the the correct version

@@
 %industrial use.
-We provide a concise mathematical definition of a cache coherent system as one
-which obeys store atomicity. We give a Coq-based proof of an MSI cache
+We provide a precise formal definition of a cache coherent system as one
+which obeys store atomicity. We give a Coq-based proof of a directory-based MSI cache
 coherence protocol for an infinite family of cache hierarchies. We specify a
@@
 invariants. Our key insight for proving these invariants is a set of
-fundamental properties.  We believe that these fundamental properties hold for
-many other invalidation-based protocols with no silent drops -- we show that
-informally for MESI, MOSI and MOESI protocols.
+fundamental properties which are shared with other protocols like MESI, MOSI
+and MOESI, enabling reuse of portions of our MSI proof for their verification.
 

### 2014-02-06 paper/Introduction.tex
this the the correct version

--- 2014-02-05 paper/Introduction.tex
@@
 \section{Introduction}
-\label{Sec:Introduction}
+\label{sec:Introduction}
 
-Shared memory is a common abstraction for distributed systems, as it
-provides an easily understood model of interaction. To achieve this
-abstraction efficiently, many hardware systems implement \emph{cache
-coherence protocols}.
+Shared memory is a common abstraction for distributed systems, as it provides
+an easily understood model of interaction. To achieve this abstraction
+efficiently, many hardware systems implement \emph{cache coherence protocols}.
+Since the correctness of this abstraction is paramount in the correct
+functioning of the system, it is imperative to ensure that the cache coherence
+protocol implementation is correct.
 
-Since the correctness of this abstraction is paramount in the correct
-functioning of the system, significant effort is expended in making sure
-that the cache coherence engine is correct. In practice, verification proceeds
-in two steps. First, a cache coherence protocol is constructed and
-verified. This protocol generally directly represents many of
-the key concurrent aspects of an implementation (\eg{} reads and
-writes to the system are broken into local atomic micro-operations) but
-ignores implementation details (\eg{} buffer sizing and memory access
-times). Second, once the protocol has been verified, it is handed to
-hardware experts who fill in the missing implementation details and
-realize it at the Register Transfer Level (RTL) of gates, wires, and
-registers, via further breaking down of local micro-operations to
-make them realizable in hardware. This effort is not a one-time translation, as
-performance properties are not easily estimated; a significant amount of
-performance testing is requires to select a good implementation.
+Formal verification of cache coherence protocols has traditionally been done
+using model-checking based approaches. This signficantly restricts the size as
+well as complexity of the system that has to be verified; most model-checking
+based approaches can not handle more than one level of hierarchy. To the best
+of our knowledge, the largest system that has been verified using
+model-checking based approaches has only 12 processors. \xxx{citation:
+http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
+data from their related work must cite)}.
 
-Methodologically, this process is flawed. The guarantees gained from
-verifying the protocol have no bearing on the implementation if
-hardware designers must do a manual translation to RTL.
+Apart from scalability concerns, another problem with model-checking based
+approaches is that it provides no intuition as to why a protocol is correct, if
+the model checker successfully proves a protocol. Developing a new protocol
+therefore therefore can not use any intuition that could have been gained from
+the first protocol. Moreover the new protocol has to be verified from scratch.
 
-Verified protocols are often described via Term Rewrite Systems (TRS),
-which apply equational logic to describe systems at an interesting
-point on the spectrum between declarative and operational methods.  It
-was shown by Dave \etal{}~\cite{DNA:CoherenceImplementation} that such
-a verified protocol could be transliterated into
-Bluespec~\cite{Bluespec:TFRG}, a hardware description language based
-on Guarded Atomic Actions, a restriction of TRS~\cite{Hoe:TCAD}. In
-this way one reduces the correctness of the implementation to the
-correctness of the protocol, with no need for separate reasoning about
-an implementation. While the current Bluespec compiler does not
-support some implementation elements that are common in practice
-(\eg{} fairness guarantees, atomic actions running over multiple clock
-cycles), recent work has provided key insights toward lifting the
-restrictions~\cite{DNA:CoherenceImplementation, Karczmarek}.  We
-believe that Bluespec and its future evolutions represent a very
-promising platform for productive implementation and verification of
-hardware components, including cache-coherent shared memories.
+We propose a theorem-prover based approach for proving cache coherence
+protocols. Using this approach eliminates the scalability and complexity
+problems faced by model-checkers. But, the downside is that considerable
+laborious effort is needed by verification people to provide a proof for the
+protocol, and the complexity of the protocol directly translates into effort.
+Our motivating question in this paper is,
+\emph{can we modularize formal proofs of cache coherence protocols so that
+verification experts do the hard work once and for all, and then hardware
+designers can instantiate the results to different designs with much less
+effort?}
 
-Even though the protocol transliteration approach cuts out one
-layer of verification effort, it may not be practical in practice.
-Design of a hardware system requires many tweaks to reach
-optimal performance, since low-level performance details often
-can only be found empirically.  Each performance tweak potentially
-breaks correctness, or at least breaks the structure of the old
-\emph{proof} of correctness, so that a full \emph{reverification}
-of correctness is needed in principle. Hardware designers are
-unlikely to have the expertise or skill to do each such
-verification.  Our motivating question in this paper is, \emph{can
-we modularize formal proofs of cache coherence protocols so that
-verification experts do the hard work once and for all, and then
-hardware designers can instantiate the results to different designs
-with much less effort?}
+While we have not reached a push-button state in our work, we do develop a
+modular proof strategy, where portions of our proof can be used \emph{as is} to
+prove other related but distinct protocols. Moreover, the proof gives solid
+insights on how the protocol works correctly, which can be used to design other
+protocols.
 
-In practice, protocol variants rely on a small set of general ``local
-building block'' properties, like guaranteed delivery of messages sent
-between processors, or guaranteed state changes in processors upon
-receiving certain messages.  These properties are
-easily verifiable, since they correspond to a small partition of the
-system, for instance a single cache.  The details of how these
-properties hold vary depending on the precise design variant. For
-instance, the reason why some property holds, like the point-to-point
-FIFO ordering for messages transferred between two nodes, depends on
-the type of network; the proof will be very different for a mesh
-network compared to a point-to-point network.
+The main contributions of this paper are as follows:
+\begin{enumerate}
+\item \emph{We give a precise formal definition of a cache coherent system as one
+which obeys store atomicity. The definition is abstract in that it does not
+rely on any hardware implementation.}
+\item \emph{We provide a Coq-based proof of the directory-based MSI cache coherence
+protocol with inclusive caches, and no silent drops.}
+\item \emph{We present a set of invariants which are not restricted to the MSI
+protocol, but would imply store atomicity. Thus, other protocols will also
+satisfy store atomicity as long as they obey the invariants.}
+\item \emph{We present what we believe to be the key insight behind the
+correctness of the MSI protocol. We believe that this property is true also of
+other protocols like MESI, MOSI and MOESI, enabling reuse of portions of our
+MSI proof for their verification.}
+\end{enumerate}
+%In practice, verification proceeds in
+%two steps. First, a cache coherence protocol is constructed and verified. This
+%protocol generally directly represents many of the key concurrent aspects of an
+%implementation (\eg{} reads and writes to the system are broken into local
+%atomic micro-operations) but ignores implementation details (\eg{} buffer
+%sizing and memory access times). Second, once the protocol has been verified,
+%it is handed to hardware experts who fill in the missing implementation details
+%and realize it at the Register Transfer Level (RTL) of gates, wires, and
+%registers, via further breaking down of local micro-operations to make them
+%realizable in hardware. This effort is not a one-time translation, as
+%performance properties are not easily estimated; a significant amount of
+%performance testing is requires to select a good implementation. 
+%
+%\adamc{I don't understand what this paragraph is saying.}
+%From a methodological perspective, this two-step process leaves much to be
+%desired. The significant effort of verification is effectively lost by the
+%concurrent non-trivial tasks of translation from the protocol representation to
+%a hardware description and the necessary refinement to add missing
+%implementation details to the protocol.
 
-In this paper, we describe an approach to provide this proof framework
-for any particular cache coherence protocol via the example of an
-MSI-style protocol.  We also show how it can be extended to other protocols
-like MESI and MOESI.  We present a modular proof architecture,
-where modification of microarchitectural
-aspects of design only necessitates reverification of relatively shallow
-local building block properties. Our framework and several instantiations
-for it are mechanized in the Coq proof assistant system, providing the
-highest levels of formal assurance upon a small trusted code base.
-We implemented and verified one particular MSI protocol via an
-operational semantics in Coq.
+Protocols, post verification, are often described via Term Rewrite Systems
+(TRS) \adamc{citation for TRSes?}, which apply equational logic to describe
+systems at an interesting point on the spectrum between declarative and
+operational methods.  It was shown by Dave
+\etal{}~\cite{DNA:CoherenceImplementation} that such a protocol, after its
+verification, can be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
+hardware description language based on Guarded Atomic Actions, a restriction of
+TRS~\cite{Hoe:TCAD}. Bluespec specification can be directly implemented into
+efficient hardware. In this way one reduces the correctness of the
+implementation to the correctness of the protocol, with no need for separate
+reasoning about an implementation. While the current Bluespec compiler does not
+support some implementation elements that are common in practice (\eg{}
+fairness guarantees, atomic actions running over multiple clock cycles), recent
+work has provided key insights toward lifting the
+restrictions~\cite{Karczmarek}. We believe that Bluespec and its future
+evolutions represent a very promising platform for productive implementation of
+hardware components, including cache coherent shared memories, starting from a
+description meant for making verification easier. Our model of the MSI protocol
+in Coq is described as an inductively defined set of transitions which is
+directly translatable into Guarded Atomic Actions. In fact, our Coq-based model
+is a direct transliteration of the MSI protocol specified in Bluespec.  This
+ensures that the final hardware synthesized from the specification is formally
+verified.
 
-\noindent\textbf{Paper Outline}. In Section~\ref{Sec:Background} we
-discuss our notion of correctness for our cache coherence protocol
-framework. We describe our MSI protocol example and its implementation
-in Section~\ref{Sec:System}. In Sections~\ref{Sec:Safety} and
-\ref{Sec:Liveness}, we discuss the structuring of our proof framework and
-verification of the base protocol. We show implementation refinements
-in Section~\ref{Sec:Refinements} and conclude with a general
-discussion of the applicability of our technique.
+%Even though the protocol transliteration approach cuts out one
+%layer of verification effort, it may not be practical in practice.
+%Design of a hardware system requires many tweaks to reach
+%optimal performance, since low-level performance details often
+%can only be found empirically.  Each performance tweak potentially
+%breaks correctness, or at least breaks the structure of the old
+%\emph{proof} of correctness, so that a full \emph{reverification}
+%of correctness is needed in principle. Hardware designers are
+%unlikely to have the expertise or skill to do each such
+%verification. 
+%
+%In practice, protocol variants rely on a small set of general ``local
+%building block'' properties, like guaranteed delivery of messages sent
+%between processors, or guaranteed state changes in processors upon
+%receiving certain messages.  These properties are
+%easily verifiable, since they correspond to a small partition of the
+%system, for instance a single cache.  The details of how these
+%properties hold vary depending on the precise design variant. For
+%instance, the reason why some property holds, like the point-to-point
+%FIFO ordering for messages transferred between two nodes, depends on
+%the type of network; the proof will be very different for a mesh
+%network compared to a point-to-point network.
+
+%In this paper, we describe an approach to provide this proof framework
+%for any particular cache coherence protocol via the example of an
+%MSI-style protocol.  We also show how it can be extended to other protocols
+%like MESI and MOESI.  We present a modular proof architecture,
+%where modification of microarchitectural
+%aspects of design only necessitates reverification of relatively shallow
+%local building block properties. Our framework and several instantiations
+%for it are mechanized in the Coq proof assistant system, providing the
+%highest levels of formal assurance upon a small trusted code base.
+%We implemented and verified one particular MSI protocol via an
+%operational semantics in Coq.
+
+\noindent\textbf{Paper Outline}. In Section~\ref{Background} we give a
+formal definition of cache coherent system as one which obeys store atomicity.
+We describe memory subsystems as mathematical functions and define store
+atomicity as a property of the input/output pairs of the memory subsystem
+function. We describe our MSI protocol and its representation as a
+transition relation in Coq in Section~\ref{sec:System}. In
+Section~\ref{safety} we discuss the structuring of our proof. We
+present the invariants that are sufficient to prove store atomicity and give
+the key insight that enables these invariants to be proved.
+%We discuss some refinements in
+%Section~\ref{Sec:Refinements} and conclude with a general discussion
+%of the applicability of our technique.

### 2014-02-06 paper/Intuition.tex
this the the correct version

@@
 discovered for proving the second-level invariants in the form of 3 invariants.
-We believe that these properties are common in most, if not all
-invalidation-based protocols which do not perform silent drops (\ie{} protocols
-in which a cache that downgrades its state informs some other cache about the
-downgrade). We informally show why this is the case using the example of MESI,
+We believe that other invalidation-based protocols with inclusive caches (\ie{}
+those cache hierarchies where a parent cache contains all the locations that
+any of its children contains) like MESI, MOSI and MOESI also
+satisfy these 3 fundamental properties, making their verification easier by reusing
+portions of our MSI proof.
+We informally show why this is the case using the example of MESI,
 MOSI and MOESI in Section \ref{mosi}.
@@
 directly from the MSI transitions, correctness of Invariants
-\fullref{localCompatibility1}, \fullref{localCompatibility2} and \fullref{latestValue} require further
+\fullref{nonAncestorCompatible} and \fullref{latestValue} require further
 decomposition. We present 4 more invariants for proving those two invariants.
@@
 parent upgrades the directory. Since the other side simply lags behind,
-Invariant \fullref{conservative} holds.
+Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
+\ref{respFIFO} deals with just one cache's state for a single address and the
+corresponding directory entry, these properties can potentially be verified
+using model-checking techniques for an invalidation-based protocol.
 
@@
 more properties of the MSI transistions of Figure \ref{trans}, though it is
-straightforward.
+straightforward. Analysing a parent-child pair will suffice in proving Invariants
+\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
+\fullref{noTransitWrite} is similar to the proof of Invariant
+\fullref{nonAncestorCompatible}.
 

### 2014-02-06 paper/Introduction.tex
Fix missing citation

@@
 of our knowledge, the largest system that has been verified using
-model-checking based approaches has only 12 processors. \xxx{citation:
-http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
-data from their related work must cite)}.
+model-checking based approaches has only 12 processors~\cite{Chen:2008:VHC:1559300}.
 

### 2014-02-06 paper/Abstract.tex
A pass through end of Introduction

@@
-%Correct implementation of cache coherent systems is of high importance in
-%industrial use.
-We provide a precise formal definition of a cache coherent system as one
-which obeys store atomicity. We give a Coq-based proof of a directory-based MSI cache
-coherence protocol for an infinite family of cache hierarchies. We specify a
-set of invariants which imply store atomicity, enabling other protocols to be
-verified for store atomicity automatically as long as they satisfy the
-invariants. Our key insight for proving these invariants is a set of
-fundamental properties which are shared with other protocols like MESI, MOSI
+Cache-coherent shared memory is the foundation for today's high-performance
+multicore hardware platforms.  Previous work has verified correctness of
+shared memory implementations for finite-state systems, for fixed topologies of
+processors and caches.  We present the first formal verification of correctness
+for an \emph{infinite} family of memory hierarchies, via machine-checked proofs
+in the Coq proof assistant.  First, we identify a set of general invariants
+and prove formally that they imply cache coherence.  Then we give a proof
+for the concrete example of the directory-based MSI protocol, establishing
+correct behavior for arbitrary tree-shaped memory hierarchies.
+Our key insight for proving the invariants is a set of
+fundamental properties shared with other protocols like MESI, MOSI,
 and MOESI, enabling reuse of portions of our MSI proof for their verification.
-
-%Previously, it has been shown
-%that these atomic transitions can be translated automatically into efficient
-%hardware. Verifying the specification based on atomic transitions therefore
-%guarantees the correctness of the actual implementation.
-%The proof is built in a modular fashion, identifying key invariants common with
-%other protocols like MESI, MOSI, \etc. We informally show how these other
-%protocols can be proved reusing modules from the proof of the MSI protocol.

### 2014-02-06 paper/Introduction.tex
A pass through end of Introduction

--- 2014-02-06 paper/Introduction.tex
@@
 Formal verification of cache coherence protocols has traditionally been done
-using model-checking based approaches. This signficantly restricts the size as
+using model checking. That family of analyses significantly restricts the size as
 well as complexity of the system that has to be verified; most model-checking
@@
 of our knowledge, the largest system that has been verified using
-model-checking based approaches has only 12 processors. \xxx{citation:
+model checking has only 12 processors, in an era when many observers are
+predicting mainstream systems with hundreds of processor cores.
+\xxx{citation:
 http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
@@
 
-Apart from scalability concerns, another problem with model-checking based
-approaches is that it provides no intuition as to why a protocol is correct, if
+Apart from scalability concerns, another downside of model checking
+is that it provides no intuition as to why a protocol is correct, if
 the model checker successfully proves a protocol. Developing a new protocol
-therefore therefore can not use any intuition that could have been gained from
-the first protocol. Moreover the new protocol has to be verified from scratch.
+therefore cannot use any intuition that could have been gained from
+the first protocol. Moreover, the new protocol has to be verified from scratch.
 
@@
 protocols. Using this approach eliminates the scalability and complexity
-problems faced by model-checkers. But, the downside is that considerable
-laborious effort is needed by verification people to provide a proof for the
+problems faced by model checkers, but the downside is that considerable
+laborious effort is needed by verification experts to provide a proof for the
 protocol, and the complexity of the protocol directly translates into effort.
@@
 
-While we have not reached a push-button state in our work, we do develop a
+While our present approach is not push-button, we have developed a
 modular proof strategy, where portions of our proof can be used \emph{as is} to
-prove other related but distinct protocols. Moreover, the proof gives solid
-insights on how the protocol works correctly, which can be used to design other
+prove other related but distinct protocols. Moreover, the proof gives
+insights on why the protocol works correctly, which can be used to design other
 protocols.
 
-The main contributions of this paper are as follows:
+The main contribution of this paper is \textbf{the first verification of a
+cache coherence protocol that is parametric over an infinite family of
+memory hierarchies}.  Our proof is implemented in the Coq proof assistant and
+checked via a very small trusted code base.  We also want to draw attention to
+a few key elements of our methodology that should be of interest for future
+verifications of this kind, both through literal code reuse and through insights
+about key invariants that should generalize across protocols.
 \begin{enumerate}
-\item \emph{We give a precise formal definition of a cache coherent system as one
-which obeys store atomicity. The definition is abstract in that it does not
-rely on any hardware implementation.}
-\item \emph{We provide a Coq-based proof of the directory-based MSI cache coherence
-protocol with inclusive caches, and no silent drops.}
-\item \emph{We present a set of invariants which are not restricted to the MSI
-protocol, but would imply store atomicity. Thus, other protocols will also
-satisfy store atomicity as long as they obey the invariants.}
-\item \emph{We present what we believe to be the key insight behind the
-correctness of the MSI protocol. We believe that this property is true also of
-other protocols like MESI, MOSI and MOESI, enabling reuse of portions of our
-MSI proof for their verification.}
+\item We give a \emph{precise formal definition of a cache-coherent system} as one
+that obeys \emph{store atomicity}. The definition is abstract in that it does not
+rely on any hardware implementation.
+\item We provide a \emph{Coq-based proof of the directory-based MSI cache coherence
+protocol} with inclusive caches and no silent drops.
+\item We present \emph{a set of invariants} that are not restricted to the MSI
+protocol but which imply store atomicity. Thus, other protocols will also
+satisfy store atomicity as long as they obey the invariants.
+\item We present what we believe to be \emph{the key insight behind the
+correctness of the MSI protocol}. We believe that this property is true also of
+other protocols like MESI, MOSI, and MOESI, enabling reuse of portions of our
+MSI proof for their verification.
 \end{enumerate}
@@
 hardware description language based on Guarded Atomic Actions, a restriction of
-TRS~\cite{Hoe:TCAD}. Bluespec specification can be directly implemented into
+TRS~\cite{Hoe:TCAD}. Bluespec specifications can be directly implemented into
 efficient hardware. In this way one reduces the correctness of the
@@
 evolutions represent a very promising platform for productive implementation of
-hardware components, including cache coherent shared memories, starting from a
-description meant for making verification easier. Our model of the MSI protocol
-in Coq is described as an inductively defined set of transitions which is
+hardware components, including cache-coherent shared memories, starting from a
+description designed to support verification. Our model of the MSI protocol
+in Coq is described as an inductively defined set of transitions that is
 directly translatable into Guarded Atomic Actions. In fact, our Coq-based model
-is a direct transliteration of the MSI protocol specified in Bluespec.  This
-ensures that the final hardware synthesized from the specification is formally
+is a direct transliteration of the MSI protocol specified in Bluespec,
+ensuring that the final hardware synthesized from the specification is formally
 verified.
@@
 \noindent\textbf{Paper Outline}. In Section~\ref{Background} we give a
-formal definition of cache coherent system as one which obeys store atomicity.
+formal definition of a cache-coherent system as one that obeys store atomicity.
 We describe memory subsystems as mathematical functions and define store

### 2014-02-06 paper/Title.tex
intuition
this the the correct version
A pass through end of Introduction

@@
 
-\title{Verifying Hierarchical Cache Coherence Protocols: a theorem-prover based modular, reusable proof}
+\title{Verifying Cache Coherence Protocols for an Infinite Family of Cache Hierarchies}
 
@@
 
-\centerline{\footnotesize {\bf Keywords}: Cache coherence, theorem prover, cache hierarchy, guarded atomic actions}
+\centerline{\footnotesize {\bf Keywords}: Cache coherence, proof assistant, cache hierarchy, guarded atomic actions}
 

### 2014-02-06 paper/Background.tex
this the the correct version
A pass through Background

@@
 stating that for each location/address in memory, every read and write access
-to that location from any processor, occurs in some total order, with
+to that location, from any processor, occurs in some total order, with
 each operation seeing the effects of previous operations and not later ones (in
@@
 \subsection{Formal Definition of Store Atomicity}
-A memory subsystem can be modeled as the following function. Its input models a
-set of infinite number of requests ever sent by any processor, and its output
-models a set of (possibly infinite number of) responses ever sent by the memory
-subsystem to any processor.
+A memory subsystem can be modeled as the following function. Its input is a
+potentially infinite set of requests sent by processors in the course of some
+full-system execution, and its output is a potentially infinite set of
+the responses sent by the memory subsystem processors.
 
 Each input to the memory subsystem is isomorphic to a function
-$\reqFn:(Processor*\mathbb{N}) @-> \Request$, where $\reqFn(p, i)$ models the
-$i^{th}$ request issued by processor $p$. An element of the range-set of
-\reqFn{}, \viz \Request{} contains fields (a) address (\addrQ) corresponding to
-the request, (b) \desc{} indicating whether the request is a load or a store
+$\reqFn : Processor \times \mathbb{N} @-> \Request$, where $\reqFn(p, i)$ models the
+$i^{th}$ request issued by processor $p$. An element of the set \Request{} (the
+range of \reqFn{}) contains fields (a) address (\addrQ) corresponding to
+the request, (b) \desc{} indicating whether the request is a load or a store,
 and (c) data (\dataQ) containing the data supplied by the processor for stores.
 
-Each output of the memory subsystem contains a set (called \Response) of
-elements containing the fields (a) \labelR, containing a pair from the set
-$Processor*\mathbb{N}$, which is meant to identify a request uniquely, (b)
-\dataR, meant to model the data returned from the memory subsystem and (c)
-\timeR, a value from set $\mathbb{N}$, meant to represent the ``time'' at which
+An output of the memory subsystem is a set \Response{} of exactly the responses generated
+in a particular protocol execution.  Each response includes the fields
+(a) \labelR, containing a pair from the set
+$Processor \times \mathbb{N}$, which is meant to identify a request uniquely, (b)
+\dataR, meant to model the data returned from the memory subsystem, and (c)
+\timeR{} in $\mathbb{N}$, meant to represent the ``time'' at which
 the corresponding request (identified by \labelR) is handled by the memory
-subsystem. \timeR{} does not have a counterpart in a real implementation, it is
+subsystem. This last field does not have a counterpart in a real implementation; it is
 present solely as a specificational device to capture store atomicity. Since a
@@
 directly from the response. For instance, $r.\addrQ$ for $r \in \Response$
-denotes the $(r.\labelR).\addrQ$, \etc.
+denotes $(r.\labelR).\addrQ$, \etc
 
 If \reqFn{} is an input to the memory subsystem, and the set \Response{} is the
-produced output, then the memory subsystem is said to be store atomic if it
+corresponding output, then the memory subsystem is said to be \textbf{store atomic} if it
 obeys Theorem \fullref{storeAtomicity} defined as follows. We use the word
-response to denote an element in the set \Response{}.
+``response'' to denote an element in the set \Response{}.
 %
-\begin{defn} (Responses have unique labels)
+\begin{defn} (Responses have unique labels.)
 \small
 \begin{multline*}
-\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response),
+\textit{uniqRespLabels} := \forall (r_1, r_2 \in \Response). \;
 r_1.\labelR = r_2.\labelR @-> \\ r_1.\timeR = r_2.\timeR \wedge
@@
 \end{defn}
-%
+
+\vspace{-.3in}
+
 \begin{defn} (No response can share a \timeR{} value with another store
-response for the same address.)
+response for the same address, imposing a total order on stores within an address.)
 \small
@@
 \textit{uniqStTimes} := 
-\forall (r_1, r_2 \in \Response),
+\forall (r_1, r_2 \in \Response). \;
 r_1.\timeR = r_2.\timeR @-> \\
@@
 \end{defn}
-%
+
+\vspace{-.3in}
+
 \begin{defn} (The global ordering of responses to the same processor should not
-violate the ordering of the corresponding requests within the processor)
+violate the ordering of the corresponding requests within the processor.)
 \small
@@
 \textit{localOrdering} :=
-\forall (r_1, r_2 \in \Response), r_1.\labelR.\fst = r_2.\labelR.\fst @->\\
+\forall (r_1, r_2 \in \Response). \; r_1.\labelR.\fst = r_2.\labelR.\fst @->\\
 r_1.\labelR.\snd < r_2.\labelR.\snd @-> r_1.\timeR \le r_2.\timeR
@@
 \end{defn}
-%
+
+\vspace{-.3in}
+
 \begin{defn} (If a response to a particular request from a processor exists in the \Response{} set,
-then responses for all previous requests from that processor must exist)
+then responses for all previous requests from that processor must exist.)
 \small
@@
 \textit{allPrevious} :=
-\forall (r \in \Response), \forall i < r.\labelR.\snd, \\
-\exists (r' \in \Response) \text{ such that } r'.\labelR.\fst = r.\labelR.\fst \wedge r'.\labelR.\snd = i
+\forall (r \in \Response). \; \forall i < r.\labelR.\snd. \\
+\exists (r' \in \Response). \; r'.\labelR.\fst = r.\labelR.\fst \wedge r'.\labelR.\snd = i
 \end{multline*}
@@
 \end{defn}
-%
-This brings us to the main condition for store atomicity, which states that the
+
+\vspace{-.3in}
+
+Now we are ready to give the main condition for store atomicity, which states that the
 data returned for a load response corresponds to the \emph{last earlier} store
@@
 \begin{multline*}
-\textit{useInitialValue}(r) := r.\dataR = \initData(a) \; \wedge \neg \exists (r' \in \Response), \\
-r'.\timeR < t \wedge r'.\addrQ = r.\addrQ \wedge r'.\desc = \St
+\textit{useInitialValue}(r) := r.\dataR = \initData(a) \; \wedge \neg \exists (r' \in \Response). \\
+r'.\timeR < r.\timeQ \wedge r'.\addrQ = r.\addrQ \wedge r'.\desc = \St
 \end{multline*}
@@
 \end{defn}
-%
+
+\vspace{-.3in}
+
 \begin{defn}
@@
 \begin{multline*}
-\textit{lastEarlierValue}(r) := \exists (r_m \in \Response), r_m.\dataQ = r.\dataQ \; \wedge 
+\textit{lastEarlierValue}(r) := \exists (r_m \in \Response). \; r_m.\dataQ = r.\dataQ \; \wedge 
 \\ r_m.\desc = \St \wedge r_m.\addrQ = r.\addrQ \wedge
-\neg \exists (r' \in \Response), \\
-r_m.\timeR < r'.\timeR < t \wedge r'.\addrQ = r.\addrQ \wedge r'.\desc = \St
+\neg \exists (r' \in \Response). \\
+r_m.\timeR < r'.\timeR < r.\timeQ \wedge r'.\addrQ = r.\addrQ \wedge r'.\desc = \St
 \end{multline*}
@@
 \end{defn}
-%
+
+\vspace{-.3in}
+
 Using the above definitions, we can define the store-atomicity theorem as follows:
@@
 \textit{uniqStTimes} \wedge \textit{localOrdering} \wedge \textit{allPrevious} \; \wedge \\
-(\forall (r \in \Response), r.\desc = \Ld \; @-> \\
+(\forall (r \in \Response). \; r.\desc = \Ld \; @-> \\
 (\textit{useInitialValue}(r) \vee \textit{lastEarlierValue}(r)))
@@
 
+\vspace{-.3in}
+
 The technical meat of this paper is a framework for proving this store atomicity
-property for a whole family of cache coherence protocols, based on parameterization
-points corresponding to key architectural choices.
+property independently of details of memory hierarchies.  In the next section,
+we formalize memory hierarchies and a specific nondeterministic algorithm for
+generating their possible event traces.

### 2014-02-06 paper/Intuition.tex
A pass through Intuition

@@
-\section{Proof of MSI protocol being preserving Store atomicity}
+\section{Proof that the MSI Protocol Preserves Store Atomicity}
 \label{safety}
@@
 invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
- invariants is because they are not specific to MSI protocol
-alone; any other protocol which obeys these invariants obeys Theorem
-\fullref{storeAtomicity}. We finally present the key insight which we
+ invariants is because they are not specific to the MSI protocol;
+any other protocol that obeys these invariants obeys Theorem
+\fullref{storeAtomicity}. We finally present the key insight that we
 discovered for proving the second-level invariants in the form of 3 invariants.
@@
 portions of our MSI proof.
-We informally show why this is the case using the example of MESI,
+We informally show why this is the case using the examples of MESI,
 MOSI and MOESI in Section \ref{mosi}.
+\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
 
@@
 
-These invariants hold after any number of executions of the transactions starting
-from the initial state. We use the term time $t$ to specify the global state
+These invariants hold after any number of executions of the transitions starting
+from the initial state. We write ``time $t$'' for the global state
 reached after firing $t$ transitions.
 
-While we have proofs in Coq for each invariant our discussion will be mainly
+While we have proofs in Coq for each invariant, our discussion will be mainly
 informal for brevity. A full description can be found at~\xxx{MURALI PUT
-REFERENCE HERE}.
+REFERENCE HERE}. \adamc{or just remind of the source code URL given earlier}
 
-\subsection{First level invariants}
+\subsection{First-Level Invariants}
 \label{firstLevel}
@@
 cache can handle a store request at a time, and that any cache that attempts to
-handle a load request has the ``latest'' value. We can state this formally with
-the following 4 invariants.
+handle a load request has the ``latest'' value. Four invariants formalize
+the intuition.
 
@@
 \end{inv}
+
+\vspace{-.15in}
+
 A clean cache is defined as follows.
@@
 that neither is an ancestor of the other in the cache tree hierarchy (where
-ancestor relation is the reflexive-transitive closure of the parent relation),
+the ancestor relation is the reflexive-transitive closure of the parent relation),
 for each address $a$, $c_1.\state[a]$ and $c_2.\state[a]$ are
@@
 \textit{compatibile}: Values of states $x_1$ and $x_2$ are \textit{compatible}
-iff either $x_1$ or $x_2$ is \Mo{}, then the other is \In.
+iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
 \label{compatible}
@@
 It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
-compatible if $x_1, x_2$ are compatible.
+compatible if $x_1$ and $x_2$ are compatible.
 
@@
 \textit{processReq}: Load and store requests can be processed only at a leaf
-node which is at least in the \Sh{} and \Mo{} states, respectively, for that
+node that is at least in the \Sh{} and \Mo{} states, respectively, for that
 address.
@@
 \begin{inv}
-\textit{noReqAgain}:
-\item For a leaf cache $c$, $c.\pos$ changes only on processing a request. In
-that case, $c.\pos$ gets incremented.
+\textit{noReqAgain}: For a leaf cache $c$, $c.\pos$ changes only on processing a request. In
+that case, $c.\pos$ is incremented.
 \label{noReqAgain}
@@
 \fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
-leaf caches will be in \In{} state preventing the possibility of processing
+leaf caches will be in \In{} state, blocking simultaneous
 load and store requests (because of Invariant \ref{processReq}) for the same
-address at the same time at different leaf caches. This satisfies
+address at the same time at different leaf caches. Thus, we have proved
 Definition~\fullref{uniqStTimes}.
@@
 It is easy to see that Definition \fullref{uniqRespLabels}, Definition
-\fullref{localOrdering} and Definition \fullref{allPrevious} directly follows
+\fullref{localOrdering}, and Definition \fullref{allPrevious} directly follow
 from Invariant \fullref{noReqAgain}.  As leaf caches have no children, one in
@@
 supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
-conditions for Theorem~\fullref{storeAtomicity} is satisfied given Invariants
+conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants
 \ref{latestValue} to \ref{noReqAgain}.
 
-\subsection{Second-level Invariants}
+These arguments are fleshed out in a machine-checked Coq proof that the invariants
+imply store atomicity.  Our proof may be reused for any protocol (i.e., alternative
+choice of rules like in Figure \ref{trans}) that satisfies the invariants,
+so that the only new verification effort is in proving the invariants.
+
+\subsection{Second-Level Invariants}
 \label{secondLevel}
 
-While Invariant \fullref{processReq} and \fullref{noReqAgain} can be seen
+While Invariants \fullref{processReq} and \fullref{noReqAgain} can be seen
 directly from the MSI transitions, correctness of Invariants
-\fullref{nonAncestorCompatible} and \fullref{latestValue} require further
-decomposition. We present 4 more invariants for proving those two invariants.
+\fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
+decomposition. We present 4 more invariants that are helpful in the proof.
 
@@
 \textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
-$\forall t, t_s \le t \le t_r$ where $t_s$ is the time of sending the data and
+$\forall t. \; t_s \le t \le t_r$ where $t_s$ is the time of sending the data and
 $t_r$ the time of receiving the data), no cache can process a store request for
-$a$, and the data must be sent from a cache which is \textit{clean}.
+$a$, and the data must be sent from a \textit{clean} cache.
 \label{noTransitWrite}
@@
 
-Any two caches $c_1, c_2$ neither of which are an ancestor of the other has a
-lowest common ancestor $x$ which is an ancestor of both $c_1$ and $c_2$. $x$
+Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
+lowest common ancestor $x$ that is an ancestor of both $c_1$ and $c_2$. $x$
 has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
@@
 We prove Invariant \fullref{latestValue} by strong induction on time. Consider
-two cases: a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
-b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
+two cases: (a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
+(b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
 none of its children's directory entries can be in \Mo{}, all its descendants
 should be no greater than \Sh{} for that address (because of Invariant
-\fullref{conservative} and the fact that $\ge$ relation is preserved for the
-state of a cache and that of its ancestor). Any non-ancestor of $c$, and
+\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
+state of a cache and that of its ancestor). Any non-ancestor of $c$ and
 non-descendant of $c$ must be compatible (as discussed in the previous
 paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
-compatibility means that non-descendants can not be in \Mo{} state. All of
-$c$'s ancestors are non-leaves, so they can not process any store request.
-Therefore, no cache save $c$ may have process a store request, which means that
+compatibility means that non-descendants cannot be in \Mo{} state. All of
+$c$'s ancestors are non-leaves, so they cannot process any store requests.
+Therefore, no cache save $c$ may have processed a store request, which means that
 $c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
@@
 If $c$ was not clean at time $t$, it must have received data from either its
-parent or its child. And the sender is \clean{} by Invariant
+parent or its child, and the sender is \clean{} by Invariant
 \ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
@@
 is in transit, which means $c$ will have the latest value for address $a$ at
-time $t+1$. This establishes Invariant \fullref{latestValue}.
+time $t+1$, establishing Invariant \fullref{latestValue}.
 
-\subsection{Key insight for proving second-level invariants}
+The full version of this proof is also implemented in Coq, providing an even
+higher-level starting point for verification of a new protocol.
+
+\subsection{Key Insight for Proving Second-Level Invariants}
 
 %ndave: This needs some repointing
-We present what we believe to be the key insight for proving any cache coherence protocol.
-In fact the request/response message semantics for the MSI protocol was designed to satisfy
+We present what we believe to be the key insight for proving any cache-coherence protocol.
+In fact, the request/response message semantics for the MSI protocol was designed to satisfy
 the following three invariants.
@@
 \begin{enumerate}
-\item A cache downgrades its state for an address iff it sends a response for that address
+\item A cache downgrades its state for an address iff it sends a response for that address.
 \item A parent upgrades the directory entry for an address for one of its
-children iff it sends a response for that child and address
+children iff it sends a response for that child and address.
 \item The state or directory entry changes for an address only on sending and
-receiving an response for that address, and they change to the \myto-field of the response.
+receiving a response for that address, and they change to the \myto-field of the response.
 \end{enumerate}
@@
 \end{inv}
+
+\vspace{-.1in}
 
@@
 \includegraphics[scale=.5]{resps}
-\caption{An example showing a sequence of responses ping ponging between a child and its parent}
+\caption{An example showing a sequence of responses ping-ponging between a child and its parent
+\adamc{For a camera-ready version, we should replace these bitmap images with vector images, for better printing.}}
 \label{resps}
@@
 entry of the parent or the state of the child changes on sending and receiving
-a response and the new state is showed at the points of sending or receiving
+a response and the new state is shown at the points of sending or receiving
 responses.
-Both the cache and its parent follow the same sequence state and directory
-entry changes; just that the receiver always lags slightly behind the sender.
-The receiver can not send any response before receiving because of
-\fullref{noCross}, thereby eliminating any scenarios where the lagging makes
-reasoning complicated.
+Both the cache and its parent follow the same sequence of state and directory
+entry changes, but the receiver always lags slightly behind the sender.
+The receiver cannot send any response before receiving because of
+\fullref{noCross}, thereby eliminating any scenarios where the lag complicates
+reasoning.
 
-Using these three invariants (Invariant \ref{stateChange} to \ref{respFIFO}, it
+Using these three invariants (Invariants \ref{stateChange} to \ref{respFIFO}), it
 is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
@@
 response in flight, or only responses from parent in flight, or responses from
-child in flight. A response from child downgrades the state and a response from
+child in flight. A response from child downgrades the state, and a response from
 parent upgrades the directory. Since the other side simply lags behind,
@@
 Proving the other 3 invariants in Section \ref{secondLevel} requires using
-more properties of the MSI transistions of Figure \ref{trans}, though it is
-straightforward. Analysing a parent-child pair will suffice in proving Invariants
+more properties of the MSI transistions of Figure \ref{trans}, though the argument is
+straightforward. Analyzing a parent-child pair will suffice in proving Invariants
 \fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant

### 2014-02-06 paper/Background.tex
Spelling and grammar in Intro + System

--- 2014-02-06 paper/Background.tex
@@
 
-Before we can discuss our protocol and proof framework, we must first establish
-exactly what it means for a cache coherence protocol to be correct.
-
-Intuitively, the simplest view of shared memory is as an array of addressible
-objects in which a single entry can be read or written by one of the many
-processors at a time.  In this view, all operations in the system appear to
-have a single unique total ordering.
+Before we can discuss our protocol and proof framework, we must first
+establish exactly what it means for a cache coherence protocol to be
+correct. Intuitively, the simplest view of shared memory is as an
+array of addressable objects in which a single entry can be read or
+written by one of the many processors at a time.  In this view, all
+operations in the system appear to have a single unique total
+ordering.
 
@@
 \subsection{Formal Definition of Store Atomicity}
+
 A memory subsystem can be modeled as the following function. Its input models a
@@
 directly from the response. For instance, $r.\addrQ$ for $r \in \Response$
-denotes the $(r.\labelR).\addrQ$, \etc.
+denotes $(r.\labelR).\addrQ$, \etc.
 

### 2014-02-06 paper/Introduction.tex
A pass through System
Spelling and grammar in Intro + System

--- 2014-02-06 paper/Introduction.tex
@@
 Formal verification of cache coherence protocols has traditionally been done
-using model-checking based approaches. This signficantly restricts the size as
+using model-checking based approaches. This significantly restricts the size as
 well as complexity of the system that has to be verified; most model-checking
-based approaches can not handle more than one level of hierarchy. To the best
-of our knowledge, the largest system that has been verified using
-model-checking based approaches has only 12 processors~\cite{Chen:2008:VHC:1559300}.
+based approaches cannot handle more than one level of hierarchy. To the best of
+our knowledge, the largest system that has been verified using model-checking
+based approaches has only 12 processors~\cite{Chen:2008:VHC:1559300}.
 
@@
 the model checker successfully proves a protocol. Developing a new protocol
-therefore therefore can not use any intuition that could have been gained from
-the first protocol. Moreover the new protocol has to be verified from scratch.
+therefore cannot use any intuition that could have been gained from the first
+protocol. Moreover the new protocol has to be verified from scratch.
 
@@
 
-While we have not reached a push-button state in our work, we do develop a
-modular proof strategy, where portions of our proof can be used \emph{as is} to
-prove other related but distinct protocols. Moreover, the proof gives solid
-insights on how the protocol works correctly, which can be used to design other
-protocols.
+While we have not reached a push-button state in our work, we do
+develop a modular proof strategy, where portions of our proof can be
+used \emph{as is} to prove other related but distinct
+protocols. Further, the proof gives solid insights on how the protocol
+works correctly, which can be used to design other protocols.
 
@@
 \begin{enumerate}
-\item \emph{We give a precise formal definition of a cache coherent system as one
-which obeys store atomicity. The definition is abstract in that it does not
-rely on any hardware implementation.}
-\item \emph{We provide a Coq-based proof of the directory-based MSI cache coherence
-protocol with inclusive caches, and no silent drops.}
-\item \emph{We present a set of invariants which are not restricted to the MSI
-protocol, but would imply store atomicity. Thus, other protocols will also
-satisfy store atomicity as long as they obey the invariants.}
-\item \emph{We present what we believe to be the key insight behind the
-correctness of the MSI protocol. We believe that this property is true also of
-other protocols like MESI, MOSI and MOESI, enabling reuse of portions of our
-MSI proof for their verification.}
+
+\item \emph{We give a precise formal definition of a cache coherent
+  system as one which obeys store atomicity. The definition is
+  abstract in that it does not rely on any hardware implementation.}
+
+\item \emph{We provide a Coq-based proof of the directory-based MSI
+  cache coherence protocol with inclusive caches, and no silent
+  drops.}
+
+\item \emph{We present a set of invariants which are not restricted to
+  the MSI protocol, but would imply store atomicity. Thus, other
+  protocols will also satisfy store atomicity as long as they obey the
+  invariants.}
+
+\item \emph{We present what we believe to be the key insight behind
+  the correctness of the MSI protocol. We believe that this property
+  is true also of other protocols like MESI, MOSI and MOESI, enabling
+  reuse of portions of our MSI proof for their verification.}
+
 \end{enumerate}
+
 %In practice, verification proceeds in
@@
 
-Protocols, post verification, are often described via Term Rewrite Systems
-(TRS) \adamc{citation for TRSes?}, which apply equational logic to describe
-systems at an interesting point on the spectrum between declarative and
-operational methods.  It was shown by Dave
-\etal{}~\cite{DNA:CoherenceImplementation} that such a protocol, after its
-verification, can be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
-hardware description language based on Guarded Atomic Actions, a restriction of
-TRS~\cite{Hoe:TCAD}. Bluespec specification can be directly implemented into
-efficient hardware. In this way one reduces the correctness of the
-implementation to the correctness of the protocol, with no need for separate
-reasoning about an implementation. While the current Bluespec compiler does not
-support some implementation elements that are common in practice (\eg{}
-fairness guarantees, atomic actions running over multiple clock cycles), recent
-work has provided key insights toward lifting the
+Protocols, post verification, are often described via Term Rewrite Systems (TRS)
+\adamc{citation for TRSes?}, which apply equational logic to describe systems at
+an interesting point on the spectrum between declarative and operational
+methods.  It was shown by Dave \etal{}~\cite{DNA:CoherenceImplementation} that
+such a protocol, after its verification, can be transliterated into
+Bluespec~\cite{Bluespec:TFRG}, a hardware description language based on Guarded
+Atomic Actions, a restriction of TRS~\cite{Hoe:TCAD}. Bluespec specification can
+be directly implemented into efficient hardware. In this way one reduces the
+correctness of the implementation to the correctness of the protocol, with no
+need for separate reasoning about an implementation. While the current Bluespec
+compiler does not support some implementation elements that are common in
+practice (\eg{} fairness guarantees, atomic actions running over multiple clock
+cycles), recent work has provided key insights toward lifting these
 restrictions~\cite{Karczmarek}. We believe that Bluespec and its future
-evolutions represent a very promising platform for productive implementation of
-hardware components, including cache coherent shared memories, starting from a
-description meant for making verification easier. Our model of the MSI protocol
-in Coq is described as an inductively defined set of transitions which is
-directly translatable into Guarded Atomic Actions. In fact, our Coq-based model
-is a direct transliteration of the MSI protocol specified in Bluespec.  This
-ensures that the final hardware synthesized from the specification is formally
-verified.
+evolutions represent a promising platform for productive implementation of
+hardware, including cache-coherent shared memories, starting from a
+verification-empowering description.
+
+Our model of the MSI protocol in Coq is described as an inductively defined set
+of transitions which is directly translatable into Guarded Atomic Actions. In
+fact, our Coq-based model is a direct transliteration of the MSI protocol
+specified in Bluespec.  This ensures that the final hardware synthesized from
+the specification is formally verified.
 
@@
 the key insight that enables these invariants to be proved.
+
 %We discuss some refinements in

### 2014-02-06 paper/Intuition.tex
Spelling and grammar in Intro + System

--- 2014-02-06 paper/Intuition.tex
@@
 We first reduce the proof of store atomicity to 4 invariants, presented in
-Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
-invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
+Section~\ref{firstLevel}. These invariants in turn can be reduced to 4 more
+invariants presented in Section~\ref{secondLevel}. The reason for choosing the 8
  invariants is because they are not specific to MSI protocol
@@
 We informally show why this is the case using the example of MESI,
-MOSI and MOESI in Section \ref{mosi}.
+MOSI and MOESI in Section~\ref{mosi}.
 

### 2014-02-06 paper/System.tex
this the the correct version
Spelling and grammar in Intro + System

@@
 
-Each node contains the fields mentioned in Figure \ref{hier}. We will
+Each node contains the fields mentioned in Figure~\ref{hier}. We will
 explain what the fields mean in the following discussion.
@@
 \begin{figure}
+\centering
 \includegraphics[scale=.5]{hier}
@@
 sequence number of the next request to process; it is set to $0$ initially.
-\adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
+\xxx{Fixed???? \adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
 \adamc{I hope the diagram explains what these mappings mean, and in general what
-the tree structure really means.}
+the tree structure really means.}}
 
@@
 we consider that address to have permission \In.  Initially, all addresses in
-each cache are in the \In{} state, save for those in the root cache (i.e., main memory),
+each cache are in the \In{} state, save for those in the root cache (\ie{} main memory),
 which are in the \Mo{} state.
@@
 
-In addition to the actual permission states, each cache also maintains a
+In addition to the permission states, each cache also maintains a
 temporary \emph{wait state} for each address to deal with transitioning between
@@
 because no cache is waiting for any response from either its child or its
-parent initially.
+parent.
 
@@
 happens. Logically, these transitions happen one at a time.  Figure~\ref{trans}
-gives the atomic transitions for the MIS protocol.  The actions in the
-transitions in Figure \ref{trans} show only the specific states that change,
+gives the atomic transitions for the MSI protocol.  The actions in the
+transitions in Figure~\ref{trans} show only the specific states that change,
 and other state components are assumed to remain the same.  The notation ($x
@@
 
-We first give an intuition on how the transitions work in the common case.
-A cache can decide, unsolicitedly \xxx{is this a word?}, to upgrade its state,
-in which case it sends an upgrade request to its parent. The parent then makes
-a local decision on whether to send a response to the requesting child or not
-(based on its directory approximation, and its own state). If its own state is
-lower the requested upgrade, then it can not handle the request. The parent
-must instead, decide to upgrade its state -- the transition for doing so has
-nothing to do with the incoming request.  Once the parent's state is not lower
-than the requested upgrade, it makes sure that the rest of its children are
-``compatible'' with the requested upgrade. If not, the parent must, send
-requests to the uncompatible children to downgrade -- again the transition for
-doing so has nothing to do with the incoming request. If the children are
-compatible and its own state is higher, then the request can be handled and a
-response is sent. Each of the transitions is explained in greater detail below.
+We first give an intuition on how the transitions work in the common case.  A
+cache may spontaneously choose to upgrade its state, in which case it sends an
+upgrade request to its parent. The parent then makes a local decision on whether
+to send a response to the requesting child or not (based on its directory
+approximation, and its own state). If its own state is lower the requested
+upgrade, then it can not handle the request. The parent must instead, decide to
+upgrade its state -- the transition for doing so has nothing to do with the
+incoming request.  Once the parent's state is not lower than the requested
+upgrade, it makes sure that the rest of its children are ``compatible'' with the
+requested upgrade. If not, the parent must, send requests to the incompatible
+children to downgrade -- again the transition for doing so has nothing to do
+with the incoming request. If the children are compatible and its own state is
+higher, then the request can be handled and a response is sent. Each of the
+transitions is explained in greater detail below.
 
@@
 changing the appropriate directory state. The notation \compat{} used in
-the ParentRecvReq transition of Figure \ref{trans} is defined as follows:
+the ParentRecvReq transition of Figure~\ref{trans} is defined as follows:
 \begin{multline*}
+\small
 \compat(p, c, a, s) = \forall i \ne c,
@@
 \end{multline*}
-Finally, the child cache receives the response from the parent (ChildRecvResp)
-and sets its state while resetting its wait state if the response contains the
+Finally, the child receives the response from the parent (ChildRecvResp)
+and sets its state, resetting its wait state if the response contains the
 desired upgrade.
@@
 downgraded to a state required by a downgrade request from its parent. In such
-a case, the child simply drops the downgrade request (ChildDropReq).
+cases, the child simply drops the downgrade request (ChildDropReq).
 
@@
 and/or the channel the cache is connected to. This restriction allows such a
-system to be implemented directly into hardware. Bluespec System Verilog (BSV),
+system to be implemented directly into hardware. Bluespec SystemVerilog (BSV),
 for instance, directly converts these transitions into efficient synchronous

### 2014-02-06 paper/Intuition.tex
Intuition

@@
 
-We prove Invariant \fullref{latestValue} by strong induction on time. Consider
+We prove Invariant \fullref{latestValue} via strong induction on time. Consider
 two cases: a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
@@
 
-%ndave: This needs some repointing
 We present what we believe to be the key insight for proving any cache coherence protocol.
@@
 \begin{enumerate}
-\item A cache downgrades its state for an address iff it sends a response for that address
-\item A parent upgrades the directory entry for an address for one of its
-children iff it sends a response for that child and address
+\item A cache downgrades its state for an address if and only if it sends a response for that address
+\item A parent upgrades the directory entry for an address for one of
+  its children if and only if it sends a response for that child and
+  address
 \item The state or directory entry changes for an address only on sending and
@@
 
-Figure \ref{resps} shows an example of the effect of Invariants
+Figure~\ref{resps} shows an example of the effect of Invariants
 \fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
@@
 
-Using these three invariants (Invariant \ref{stateChange} to \ref{respFIFO}, it
+Using these three invariants (Invariant~\ref{stateChange} to \ref{respFIFO}, it
 is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
-back to Figure \ref{resps}, we can see that at all times, either there is no
+back to Figure~\ref{resps}, we can see that at all times, either there is no
 response in flight, or only responses from parent in flight, or responses from
@@
 parent upgrades the directory. Since the other side simply lags behind,
-Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
+Invariant~\fullref{conservative} holds. Since Invariants~\ref{stateChange} to
 \ref{respFIFO} deals with just one cache's state for a single address and the
@@
 
-Proving the other 3 invariants in Section \ref{secondLevel} requires using
-more properties of the MSI transistions of Figure \ref{trans}, though it is
-straightforward. Analysing a parent-child pair will suffice in proving Invariants
-\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
-\fullref{noTransitWrite} is similar to the proof of Invariant
-\fullref{nonAncestorCompatible}.
+Proving the other 3 invariants in Section~\ref{secondLevel} requires
+using more properties of the MSI transistions of Figure~\ref{trans},
+though it is straightforward. Analysing a parent-child pair will
+suffice in proving Invariants~\fullref{localCompatibility1} and
+\fullref{localCompatibility2}. Proof of
+Invariant~\fullref{noTransitWrite} is similar to the proof of
+Invariant~\fullref{nonAncestorCompatible}.
 

### 2014-02-06 paper/System.tex
spell in later half of paper

@@
 which $p$ is waiting for $c$ to downgrade. Both \wait{} and \dwait{} are
-+\False{} initially for all caches and addresses,
+\False{} initially for all caches and addresses,
 because no cache is waiting for any response from either its child or its
@@
 register array can actually take several hardware clock cycles, though
-logically it is a single transition. Karczmarek \etal~\cite{Karczmarek} has
+logically it is a single transition. Karczmarek \etal \cite{Karczmarek} has
 provided a scheme to convert atomic transitions into synchronous hardware in

### 2014-02-06 paper/Background.tex

--- 2014-02-06 paper/Background.tex
@@
 \end{thm}
-
 \vspace{-.3in}

### 2014-02-06 paper/Introduction.tex

--- 2014-02-06 paper/Introduction.tex
@@
 Formal verification of cache coherence protocols has traditionally been done
-using model checking. That family of analyses significantly restricts the size as
+using model-checking based approaches. This significantly restricts the size as
 well as complexity of the system that has to be verified; most model-checking
-based approaches can not handle more than one level of hierarchy. To the best
-of our knowledge, the largest system that has been verified using
-model checking has only 12 processors, in an era when many observers are
-predicting mainstream systems with hundreds of processor cores.
-\xxx{citation:
-http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
-data from their related work must cite)}.
+based approaches cannot handle more than one level of hierarchy. To the best of
+our knowledge, the largest system that has been verified using model-checking
+based approaches has only 12 processors~\cite{Chen:2008:VHC:1559300}, in an era
+when many observers are predicting mainstream systems with hundreds of processor
+cores.
 
@@
 \begin{enumerate}
+
 \item We give a \emph{precise formal definition of a cache-coherent system} as one
@@
 MSI proof for their verification.
+
 \end{enumerate}
+
 %In practice, verification proceeds in
@@
 
-Protocols, post verification, are often described via Term Rewrite Systems
-(TRS) \adamc{citation for TRSes?}, which apply equational logic to describe
-systems at an interesting point on the spectrum between declarative and
-operational methods.  It was shown by Dave
-\etal{}~\cite{DNA:CoherenceImplementation} that such a protocol, after its
-verification, can be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
-hardware description language based on Guarded Atomic Actions, a restriction of
-TRS~\cite{Hoe:TCAD}. Bluespec specifications can be directly implemented into
-efficient hardware. In this way one reduces the correctness of the
-implementation to the correctness of the protocol, with no need for separate
-reasoning about an implementation. While the current Bluespec compiler does not
-support some implementation elements that are common in practice (\eg{}
-fairness guarantees, atomic actions running over multiple clock cycles), recent
-work has provided key insights toward lifting the
+Protocols, post verification, are often described via Term Rewrite Systems (TRS)
+\adamc{citation for TRSes?}, which apply equational logic to describe systems at
+an interesting point on the spectrum between declarative and operational
+methods.  It was shown by Dave \etal{}~\cite{DNA:CoherenceImplementation} that
+such a protocol, after its verification, can be transliterated into
+Bluespec~\cite{Bluespec:TFRG}, a hardware description language based on Guarded
+Atomic Actions, a restriction of TRS~\cite{Hoe:TCAD}. Bluespec specification can
+be directly implemented into efficient hardware. In this way one reduces the
+correctness of the implementation to the correctness of the protocol, with no
+need for separate reasoning about an implementation. While the current Bluespec
+compiler does not support some implementation elements that are common in
+practice (\eg{} fairness guarantees, atomic actions running over multiple clock
+cycles), recent work has provided key insights toward lifting these
 restrictions~\cite{Karczmarek}. We believe that Bluespec and its future
-evolutions represent a very promising platform for productive implementation of
-hardware components, including cache-coherent shared memories, starting from a
-description designed to support verification. Our model of the MSI protocol
-in Coq is described as an inductively defined set of transitions that is
-directly translatable into Guarded Atomic Actions. In fact, our Coq-based model
-is a direct transliteration of the MSI protocol specified in Bluespec,
-ensuring that the final hardware synthesized from the specification is formally
-verified.
+evolutions represent a promising platform for productive implementation of
+hardware, including cache-coherent shared memories, starting from a
+verification-empowering description.
+
+Our model of the MSI protocol in Coq is described as an inductively defined set
+of transitions which is directly translatable into Guarded Atomic Actions. In
+fact, our Coq-based model is a direct transliteration of the MSI protocol
+specified in Bluespec.  This ensures that the final hardware synthesized from
+the specification is formally verified.
 
@@
 %of the applicability of our technique.
+
+\adamc{This might be a good place to include a link to the Coq source code.}

### 2014-02-06 paper/Intuition.tex
spell in later half of paper

--- 2014-02-06 paper/Intuition.tex
@@
 \section{Proof that the MSI Protocol Preserves Store Atomicity}
+\label{Sec:Safety}
 \label{safety}
@@
 We first reduce the proof of store atomicity to 4 invariants, presented in
-Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
-invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
- invariants is because they are not specific to the MSI protocol;
-any other protocol that obeys these invariants obeys Theorem
-\fullref{storeAtomicity}. We finally present the key insight that we
+Section~\ref{firstLevel}. These invariants in turn can be reduced to 4 more
+invariants presented in Section~\ref{secondLevel}. The reason for choosing the 8
+ invariants is because they are not specific to MSI protocol
+alone; any other protocol which obeys these invariants obeys Theorem
+\fullref{storeAtomicity}. We finally present the key insight which we
 discovered for proving the second-level invariants in the form of 3 invariants.
@@
 portions of our MSI proof.
-We informally show why this is the case using the examples of MESI,
-MOSI and MOESI in Section \ref{mosi}.
+We informally show why this is the case using the example of MESI,
+MOSI and MOESI in Section~\ref{sec:Refinements}.
 \adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
@@
 
-%ndave: This needs some repointing
 We present what we believe to be the key insight for proving any cache-coherence protocol.
@@
 \begin{enumerate}
-\item A cache downgrades its state for an address iff it sends a response for that address.
-\item A parent upgrades the directory entry for an address for one of its
-children iff it sends a response for that child and address.
+\item A cache downgrades its state for an address if and only if it sends a response for that address
+\item A parent upgrades the directory entry for an address for one of
+  its children if and only if it sends a response for that child and
+  address
 \item The state or directory entry changes for an address only on sending and
@@
 \end{inv}
-
 \vspace{-.1in}
@@
 
-Using these three invariants (Invariants \ref{stateChange} to \ref{respFIFO}), it
+Using these three invariants (Invariant~\ref{stateChange} to \ref{respFIFO}, it
 is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
-back to Figure \ref{resps}, we can see that at all times, either there is no
+back to Figure~\ref{resps}, we can see that at all times, either there is no
 response in flight, or only responses from parent in flight, or responses from
@@
 
-Proving the other 3 invariants in Section \ref{secondLevel} requires using
-more properties of the MSI transistions of Figure \ref{trans}, though the argument is
-straightforward. Analyzing a parent-child pair will suffice in proving Invariants
-\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
-\fullref{noTransitWrite} is similar to the proof of Invariant
-\fullref{nonAncestorCompatible}.
+Proving the other 3 invariants in Section~\ref{secondLevel} requires using more
+properties of the MSI transistions of Figure~\ref{trans}, though it is
+straightforward. Analysing a parent-child pair will suffice in proving
+Invariants~\fullref{localCompatibility1} and
+\fullref{localCompatibility2}. Proof of Invariant~\fullref{noTransitWrite} is
+similar to the proof of Invariant~\fullref{nonAncestorCompatible}.
 

### 2014-02-06 paper/System.tex

@@
-\section{MSI-based cache coherence protocol}
+\section{MSI-Based Cache Coherence Protocol}
 \label{sec:System}
@@
 with a processor (Figure~\ref{hier}). The system is parameterized by the shape
-of the tree, and each node can have any number of children independent of the
-number in other nodes. In the following, $\leaf(c)$ indicates
+of the tree, and each node can have any number of children.
+In the following, $\leaf(c)$ indicates
 that a cache $c$ is a leaf, and $\parent(c,p)$ indicates that $p$ is the parent of
@@
 
-Input and output happens at the leaves in the form of memory requests
-issued from and responses issued to the associated processor. These will be
+Input and output happen at the leaves in the form of memory requests
+issued from and responses issued to the associated processors. These will be
 represented by two mappings, one from $\mathbb{N}$ to requests and the other
@@
 sequence number of the next request to process; it is set to $0$ initially.
-\xxx{Fixed???? \adamc{What is a ``position'' here?  Is ``sequence number'' a better choice?}
-\adamc{I hope the diagram explains what these mappings mean, and in general what
-the tree structure really means.}}
+\adamc{Add processors and main memory, labeled explicitly, to the diagram?}
 
@@
 the \emph{read-only} state \Sh{} (for ``shared''), or in the \emph{invalid} state
-\In{} (indicating that it can be neither read nor written). The options form a natural
+\In{} (indicating that it may be neither read nor written). The options form a natural
 total order of permissions ($\In < \Sh < \Mo$). For a leaf cache, the meaning
-of these permission states is straightforward: a load request can be processed
-only if the leaf is at least in the \Sh{} state, and a store request can be
+of these permission states is straightforward: a load request may be processed
+only if the leaf is at least in the \Sh{} state, and a store request may be
 processed only if it is in the \Mo{} state. The state of a non-leaf cache indicates
@@
 availability of any message in the channel
-($\avail(ch)$). Each of these channels start out empty, \ie{} $\forall ch,
+($\avail(ch)$). Each of these channels start out empty, \ie{} $\forall ch. \;
 \neg \avail(ch)$, initially.
@@
 \end{subfigure}
-\caption{Data types for messages between caches}
+\caption{Data types for messages between caches
+\adamc{Names here don't match those in the preceding text.}}
 \label{format}
@@
 
-We first give an intuition on how the transitions work in the common case.  A
-cache may spontaneously choose to upgrade its state, in which case it sends an
-upgrade request to its parent. The parent then makes a local decision on whether
-to send a response to the requesting child or not (based on its directory
-approximation, and its own state). If its own state is lower the requested
-upgrade, then it can not handle the request. The parent must instead, decide to
-upgrade its state -- the transition for doing so has nothing to do with the
-incoming request.  Once the parent's state is not lower than the requested
-upgrade, it makes sure that the rest of its children are ``compatible'' with the
-requested upgrade. If not, the parent must, send requests to the incompatible
-children to downgrade -- again the transition for doing so has nothing to do
-with the incoming request. If the children are compatible and its own state is
-higher, then the request can be handled and a response is sent. Each of the
+We first give an intuition on how the transitions work in the common
+case.  A cache can decide, unprovoked, to upgrade its state, in which
+case it sends an upgrade request to its parent. The parent then makes
+a local decision on whether to send a response to the requesting child
+or not (based on its directory approximation, and its own state). If
+its own state is lower the requested upgrade, then it cannot handle
+the request. The parent must instead decide to upgrade its state --
+the transition for doing so has nothing to do with the incoming
+request.  Once the parent's state is not lower than the requested
+upgrade, it makes sure that the rest of its children are
+``compatible'' with the requested upgrade. If not, the parent must
+send requests to the incompatible children to downgrade -- again the
+transition for doing so has nothing to do with the incoming
+request. If the children are compatible and its own state is higher,
+then the request can be handled and a response is sent. Each of the
 transitions is explained in greater detail below.
@@
 \small
-\compat(p, c, a, s) = \forall i \ne c,
+\compat(p, c, a, s) = \forall i \ne c. \;
 \mylet s_i := p.\dstate[i][a] \myin\\
@@
 
-Because of the presence of voluntary response, a child could have already
+Because of the presence of voluntary responses, a child could have already
 downgraded to a state required by a downgrade request from its parent. In such
@@
 Sometimes, these transitions (both the guards and the actions) may not be
-amenable to single cycle implementations in hardware. For example, writing a
+amenable to single-cycle implementations in hardware. For example, writing a
 register array can actually take several hardware clock cycles, though

### 2014-02-06 paper/Background.tex
don't like relying on fixity except for -> operator

--- 2014-02-06 paper/Background.tex
@@
 Each input to the memory subsystem is isomorphic to a function
-$\reqFn : Processor \times \mathbb{N} @-> \Request$, where $\reqFn(p, i)$ models the
+$\reqFn : (Processor \times \mathbb{N}) @-> \Request$, where $\reqFn(p, i)$ models the
 $i^{th}$ request issued by processor $p$. An element of the set \Request{} (the

### 2014-02-06 paper/System.tex
don't like relying on fixity except for -> operator

@@
 
-Each node contains the fields mentioned in Figure~\ref{hier}. We will
+Each node contains the fields mentioned in Figure \ref{hier}. We will
 explain what the fields mean in the following discussion.
@@
 we consider that address to have permission \In.  Initially, all addresses in
-each cache are in the \In{} state, save for those in the root cache (\ie{} main memory),
+each cache are in the \In{} state, save for those in the root cache (i.e., main memory),
 which are in the \Mo{} state.
@@
 
-In addition to the permission states, each cache also maintains a
+In addition to the actual permission states, each cache also maintains a
 temporary \emph{wait state} for each address to deal with transitioning between
@@
 because no cache is waiting for any response from either its child or its
-parent.
+parent initially.
 
@@
 gives the atomic transitions for the MSI protocol.  The actions in the
-transitions in Figure~\ref{trans} show only the specific states that change,
+transitions in Figure \ref{trans} show only the specific states that change,
 and other state components are assumed to remain the same.  The notation ($x
@@
 
-We first give an intuition on how the transitions work in the common
-case.  A cache can decide, unprovoked, to upgrade its state, in which
-case it sends an upgrade request to its parent. The parent then makes
-a local decision on whether to send a response to the requesting child
-or not (based on its directory approximation, and its own state). If
-its own state is lower the requested upgrade, then it cannot handle
-the request. The parent must instead decide to upgrade its state --
-the transition for doing so has nothing to do with the incoming
-request.  Once the parent's state is not lower than the requested
-upgrade, it makes sure that the rest of its children are
-``compatible'' with the requested upgrade. If not, the parent must
-send requests to the incompatible children to downgrade -- again the
-transition for doing so has nothing to do with the incoming
-request. If the children are compatible and its own state is higher,
-then the request can be handled and a response is sent. Each of the
-transitions is explained in greater detail below.
+We first give an intuition on how the transitions work in the common case.
+A cache can decide, unprovoked, to upgrade its state,
+in which case it sends an upgrade request to its parent. The parent then makes
+a local decision on whether to send a response to the requesting child or not
+(based on its directory approximation, and its own state). If its own state is
+lower the requested upgrade, then it cannot handle the request. The parent
+must instead decide to upgrade its state -- the transition for doing so has
+nothing to do with the incoming request.  Once the parent's state is not lower
+than the requested upgrade, it makes sure that the rest of its children are
+``compatible'' with the requested upgrade. If not, the parent must send
+requests to the incompatible children to downgrade -- again the transition for
+doing so has nothing to do with the incoming request. If the children are
+compatible and its own state is higher, then the request can be handled and a
+response is sent. Each of the transitions is explained in greater detail below.
 
@@
 changing the appropriate directory state. The notation \compat{} used in
-the ParentRecvReq transition of Figure~\ref{trans} is defined as follows:
+the ParentRecvReq transition of Figure \ref{trans} is defined as follows:
 \begin{multline*}
@@
 \end{multline*}
-Finally, the child receives the response from the parent (ChildRecvResp)
-and sets its state, resetting its wait state if the response contains the
+Finally, the child cache receives the response from the parent (ChildRecvResp)
+and sets its state while resetting its wait state if the response contains the
 desired upgrade.
@@
 downgraded to a state required by a downgrade request from its parent. In such
-cases, the child simply drops the downgrade request (ChildDropReq).
+a case, the child simply drops the downgrade request (ChildDropReq).
 
@@
 and/or the channel the cache is connected to. This restriction allows such a
-system to be implemented directly into hardware. Bluespec SystemVerilog (BSV),
+system to be implemented directly into hardware. Bluespec System Verilog (BSV),
 for instance, directly converts these transitions into efficient synchronous
@@
 register array can actually take several hardware clock cycles, though
-logically it is a single transition. Karczmarek \etal \cite{Karczmarek} has
+logically it is a single transition. Karczmarek \etal~\cite{Karczmarek} have
 provided a scheme to convert atomic transitions into synchronous hardware in

### 2014-02-06 paper/Intuition.tex
probably fixed one comment

@@
 \section{Proof that the MSI Protocol Preserves Store Atomicity}
-\label{Sec:Safety}
 \label{safety}
@@
 We first reduce the proof of store atomicity to 4 invariants, presented in
-Section~\ref{firstLevel}. These invariants in turn can be reduced to 4 more
-invariants presented in Section~\ref{secondLevel}. The reason for choosing the 8
- invariants is because they are not specific to MSI protocol
-alone; any other protocol which obeys these invariants obeys Theorem
-\fullref{storeAtomicity}. We finally present the key insight which we
+Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
+invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
+ invariants is because they are not specific to the MSI protocol;
+any other protocol that obeys these invariants obeys Theorem
+\fullref{storeAtomicity}. We finally present the key insight that we
 discovered for proving the second-level invariants in the form of 3 invariants.
@@
 portions of our MSI proof.
-We informally show why this is the case using the example of MESI,
-MOSI and MOESI in Section~\ref{sec:Refinements}.
+We informally show why this is the case using the examples of MESI,
+MOSI and MOESI in Section \ref{sec:Refinements}.
 \adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
@@
 
+%ndave: This needs some repointing
 We present what we believe to be the key insight for proving any cache-coherence protocol.
@@
 \begin{enumerate}
-\item A cache downgrades its state for an address if and only if it sends a response for that address
-\item A parent upgrades the directory entry for an address for one of
-  its children if and only if it sends a response for that child and
-  address
+\item A cache downgrades its state for an address iff it sends a response for that address.
+\item A parent upgrades the directory entry for an address for one of its
+children iff it sends a response for that child and address.
 \item The state or directory entry changes for an address only on sending and
@@
 \end{inv}
+
 \vspace{-.1in}
@@
 
-Figure~\ref{resps} shows an example of the effect of Invariants
+Figure \ref{resps} shows an example of the effect of Invariants
 \fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
@@
 
-Using these three invariants (Invariant~\ref{stateChange} to \ref{respFIFO}, it
+Using these three invariants (Invariants \ref{stateChange} to \ref{respFIFO}), it
 is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
-back to Figure~\ref{resps}, we can see that at all times, either there is no
+back to Figure \ref{resps}, we can see that at all times, either there is no
 response in flight, or only responses from parent in flight, or responses from
@@
 parent upgrades the directory. Since the other side simply lags behind,
-Invariant~\fullref{conservative} holds. Since Invariants~\ref{stateChange} to
+Invariant \fullref{conservative} holds. Since Invariants \ref{stateChange} to
 \ref{respFIFO} deals with just one cache's state for a single address and the
@@
 
-Proving the other 3 invariants in Section~\ref{secondLevel} requires using more
-properties of the MSI transistions of Figure~\ref{trans}, though it is
-straightforward. Analysing a parent-child pair will suffice in proving
-Invariants~\fullref{localCompatibility1} and
-\fullref{localCompatibility2}. Proof of Invariant~\fullref{noTransitWrite} is
-similar to the proof of Invariant~\fullref{nonAncestorCompatible}.
+Proving the other 3 invariants in Section \ref{secondLevel} requires using
+more properties of the MSI transistions of Figure \ref{trans}, though the argument is
+straightforward. Analyzing a parent-child pair will suffice in proving Invariants
+\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
+\fullref{noTransitWrite} is similar to the proof of Invariant
+\fullref{nonAncestorCompatible}.
 

### 2014-02-06 paper/Introduction.tex
Fix Citation that was reverted

--- 2014-02-06 paper/Introduction.tex
@@
 Formal verification of cache coherence protocols has traditionally been done
-using model checking. That family of analyses significantly restricts the size as
-well as complexity of the system that has to be verified; most model-checking
-based approaches can not handle more than one level of hierarchy. To the best
-of our knowledge, the largest system that has been verified using
-model checking has only 12 processors, in an era when many observers are
-predicting mainstream systems with hundreds of processor cores.
-\xxx{citation:
-http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
-data from their related work must cite)}.
+using model checking. That family of analyses significantly restricts the size
+as well as complexity of the system that has to be verified; most model-checking
+based approaches can not handle more than one level of hierarchy. To the best of
+our knowledge, the largest system that has been verified using model checking
+has only 12 processors~\cite{Chen:2008:VHC:1559300}, in an era when many
+observers are predicting mainstream systems with hundreds of processor cores.
 

### 2014-02-06 paper/Introduction.tex
merging Nirav's changes to intro

@@
 Formal verification of cache coherence protocols has traditionally been done
-using model checking. That family of analyses significantly restricts the size
-as well as complexity of the system that has to be verified; most model-checking
-based approaches can not handle more than one level of hierarchy. To the best of
-our knowledge, the largest system that has been verified using model checking
-has only 12 processors~\cite{Chen:2008:VHC:1559300}, in an era when many
-observers are predicting mainstream systems with hundreds of processor cores.
+using model checking. That family of analyses significantly restricts the size as
+well as complexity of the system that has to be verified; most model-checking
+based approaches can not handle more than one level of hierarchy. To the best
+of our knowledge, the largest system that has been verified using
+model checking has only 12 processors, in an era when many observers are
+predicting mainstream systems with hundreds of processor cores.
+\xxx{citation:
+http://www.cs.utah.edu/formal\_verification/pdf/xiaofang\_dissertation.pdf (used
+data from their related work must cite)}.
 
@@
 hardware components, including cache-coherent shared memories, starting from a
-description designed to support verification. Our model of the MSI protocol
+description designed to support verification.
+
+Our model of the MSI protocol
 in Coq is described as an inductively defined set of transitions that is

### 2014-02-06 paper/Introduction.tex
probably fixed one comment

--- 2014-02-06 paper/Introduction.tex
@@
 our knowledge, the largest system that has been verified using model checking
-has only 12 processors~\cite{Chen:2008:VHC:1559300}, in an era when many
+has only 12 processors\xxx{the comment was meant to cite other papers from this paper as well. anyways i will take a crack at it}~\cite{Chen:2008:VHC:1559300}, in an era when many
 observers are predicting mainstream systems with hundreds of processor cores.

### 2014-02-06 paper/System.tex
merging Nirav's changes to system (using word spontaneous)
maybe fixed another of adamc's comments?

@@
 $\textit{ch}[c]$ for the \textit{ch} channel associated
-with cache $c$).  Figure~\ref{format} shows the message formats
-transmitted in the 3
-channels between caches. In each of the channels $ch$, a message $m$
-can be enqueued ($\enq(ch, m)$) or dequeued ($\deq(ch)$), and the
+with cache $c$).  Figure~\ref{format} shows the fields of the messages
+transmitted in the 3 channels between caches. In each of the channels $ch$, a
+message $m$ can be enqueued ($\enq(ch, m)$) or dequeued ($\deq(ch)$), and the
 channels can be examined for the first element ($\first(ch)$) or for
-availability of any message in the channel
-($\avail(ch)$). Each of these channels start out empty, \ie{} $\forall ch. \;
-\neg \avail(ch)$, initially.
+availability of any message in the channel ($\avail(ch)$). Each of these
+channels start out empty, \ie{} $\forall ch. \; \neg \avail(ch)$, initially.
 
@@
 \hline
-\multicolumn{2}{|c|}{\Reqcp}\\
+\multicolumn{2}{|c|}{A message in \cpReq}\\
 \hline
@@
 \hline
-\multicolumn{2}{|c|}{\Respcp}\\
+\multicolumn{2}{|c|}{A message in \cpResp}\\
 \hline
@@
 \hline
-\multicolumn{2}{|c|}{\Mpc}\\
+\multicolumn{2}{|c|}{A message in \pc}\\
 \hline
@@
 We first give an intuition on how the transitions work in the common case.
-A cache can decide, unprovoked, to upgrade its state,
+A cache can decide, spontaneously, to upgrade its state,
 in which case it sends an upgrade request to its parent. The parent then makes

### 2014-02-06 paper/Background.tex
adding tildas

@@
 corresponding output, then the memory subsystem is said to be \textbf{store atomic} if it
-obeys Theorem \fullref{storeAtomicity} defined as follows. We use the word
+obeys Theorem~\fullref{storeAtomicity} defined as follows. We use the word
 ``response'' to denote an element in the set \Response{}.

### 2014-02-06 paper/Intuition.tex
adding tildas

@@
 %address (Invariant~\textit{processReq}).
-Leaf caches cannot be ancestors of any other node, so by Invariant
+Leaf caches cannot be ancestors of any other node, so by Invariant~
 \fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
 leaf caches will be in \In{} state, blocking simultaneous
-load and store requests (because of Invariant \ref{processReq}) for the same
+load and store requests (because of Invariant~\ref{processReq}) for the same
 address at the same time at different leaf caches. Thus, we have proved
@@
 
-It is easy to see that Definition \fullref{uniqRespLabels}, Definition
-\fullref{localOrdering}, and Definition \fullref{allPrevious} directly follow
-from Invariant \fullref{noReqAgain}.  As leaf caches have no children, one in
+It is easy to see that Definition~\fullref{uniqRespLabels}, Definition~
+\fullref{localOrdering}, and Definition~\fullref{allPrevious} directly follow
+from Invariant~\fullref{noReqAgain}.  As leaf caches have no children, one in
 at least \Sh{} state for address $a$ is \clean{} for that address. Since
@@
 supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
-conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants
+conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants~
 \ref{latestValue} to \ref{noReqAgain}.
@@
 imply store atomicity.  Our proof may be reused for any protocol (i.e., alternative
-choice of rules like in Figure \ref{trans}) that satisfies the invariants,
+choice of rules like in Figure~\ref{trans}) that satisfies the invariants,
 so that the only new verification effort is in proving the invariants.
@@
 
-While Invariants \fullref{processReq} and \fullref{noReqAgain} can be seen
-directly from the MSI transitions, correctness of Invariants
+While Invariants~\fullref{processReq} and \fullref{noReqAgain} can be seen
+directly from the MSI transitions, correctness of Invariants~
 \fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
@@
 
-Invariants \fullref{conservative} and \fullref{localCompatibility1} defines a
+Invariants~\fullref{conservative} and \fullref{localCompatibility1} defines a
 $\ge$ relation between the states of a parent and its child for any address.
@@
 between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
-are also compatible. This proves Invariant \fullref{nonAncestorCompatible}.
+are also compatible. This proves Invariant~\fullref{nonAncestorCompatible}.
 
-We prove Invariant \fullref{latestValue} by strong induction on time. Consider
+We prove Invariant~\fullref{latestValue} by strong induction on time. Consider
 two cases: (a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
@@
 none of its children's directory entries can be in \Mo{}, all its descendants
-should be no greater than \Sh{} for that address (because of Invariant
+should be no greater than \Sh{} for that address (because of Invariant~
 \fullref{conservative} and the fact that the $\ge$ relation is preserved for the
@@
 If $c$ was not clean at time $t$, it must have received data from either its
-parent or its child, and the sender is \clean{} by Invariant
+parent or its child, and the sender is \clean{} by Invariant~
 \ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
@@
 is in transit, which means $c$ will have the latest value for address $a$ at
-time $t+1$, establishing Invariant \fullref{latestValue}.
+time $t+1$, establishing Invariant~\fullref{latestValue}.
 
@@
 
-Invariant \fullref{stateChange} ensures that a response message acts as a sync
+Invariant~\fullref{stateChange} ensures that a response message acts as a sync
 between a parent's directory entry for a child and the state of the child, as
@@
 
-Figure \ref{resps} shows an example of the effect of Invariants
+Figure~\ref{resps} shows an example of the effect of Invariants~
 \fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
@@
 
-Using these three invariants (Invariants \ref{stateChange} to \ref{respFIFO}), it
-is easy to see that Invariant \fullref{conservative} holds. Intuitively, going
-back to Figure \ref{resps}, we can see that at all times, either there is no
+Using these three invariants (Invariants~\ref{stateChange} to \ref{respFIFO}), it
+is easy to see that Invariant~\fullref{conservative} holds. Intuitively, going
+back to Figure~\ref{resps}, we can see that at all times, either there is no
 response in flight, or only responses from parent in flight, or responses from
@@
 Proving the other 3 invariants in Section \ref{secondLevel} requires using
-more properties of the MSI transistions of Figure \ref{trans}, though the argument is
-straightforward. Analyzing a parent-child pair will suffice in proving Invariants
-\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant
-\fullref{noTransitWrite} is similar to the proof of Invariant
+more properties of the MSI transistions of Figure~\ref{trans}, though the argument is
+straightforward. Analyzing a parent-child pair will suffice in proving Invariants~
+\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant~
+\fullref{noTransitWrite} is similar to the proof of Invariant~
 \fullref{nonAncestorCompatible}.

### 2014-02-06 paper/System.tex
maybe fixed another of adamc's comments?
adding tildas

@@
 \hline
-\multicolumn{2}{|c|}{A message in \cpReq}\\
+\multicolumn{2}{|c|}{A message in \cpReq{} channel}\\
 \hline
@@
 \hline
-\multicolumn{2}{|c|}{A message in \cpResp}\\
+\multicolumn{2}{|c|}{A message in \cpResp{} channel}\\
 \hline
@@
 \hline
-\multicolumn{2}{|c|}{A message in \pc}\\
+\multicolumn{2}{|c|}{A message in \pc{} channel}\\
 \hline
@@
 gives the atomic transitions for the MSI protocol.  The actions in the
-transitions in Figure \ref{trans} show only the specific states that change,
+transitions in Figure~\ref{trans} show only the specific states that change,
 and other state components are assumed to remain the same.  The notation ($x
@@
 changing the appropriate directory state. The notation \compat{} used in
-the ParentRecvReq transition of Figure \ref{trans} is defined as follows:
+the ParentRecvReq transition of Figure~\ref{trans} is defined as follows:
 \begin{multline*}

### 2014-02-06 paper/Intuition.tex
fixed added one line in system

@@
 We first reduce the proof of store atomicity to 4 invariants, presented in
-Section \ref{firstLevel}. These invariants in turn can be reduced to 4 more
-invariants presented in Section \ref{secondLevel}. The reason for choosing the 8
+Section~\ref{firstLevel}. These invariants in turn can be reduced to 4 more
+invariants presented in Section~\ref{secondLevel}. The reason for choosing the 8
  invariants is because they are not specific to the MSI protocol;
@@
 We informally show why this is the case using the examples of MESI,
-MOSI and MOESI in Section \ref{sec:Refinements}.
+MOSI and MOESI in Section~\ref{sec:Refinements}.
 \adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
@@
 
-Proving the other 3 invariants in Section \ref{secondLevel} requires using
+Proving the other 3 invariants in Section~\ref{secondLevel} requires using
 more properties of the MSI transistions of Figure~\ref{trans}, though the argument is
@@
 %$c.\state[a]$ never changed before $t$, then $c.\state[a] = \In$ because of the
-%initialization conditions described in Section \ref{System}. No response would
+%initialization conditions described in Section~\ref{System}. No response would
 %have been sent by $c$ to its parent for address $a$ before $t$, otherwise

### 2014-02-06 paper/Introduction.tex
intro/system redundancy removed

@@
 hardware components, including cache-coherent shared memories, starting from a
-description designed to support verification.
+specification designed to support verification.
 

### 2014-02-06 paper/System.tex
fixed added one line in system
intro/system redundancy removed

@@
 compatible and its own state is higher, then the request can be handled and a
-response is sent. Each of the transitions is explained in greater detail below.
+response is sent. The transitions for sending requests are allowed to happen
+spontaneously without any trigger to model a family of protocols differing in
+the policy for sending requests.
 
@@
 state, \ie{} do not \emph{conflict}~\cite{Hoe:TCAD,HoeArvind:TRSSynthesis1}.
-Sometimes, these transitions (both the guards and the actions) may not be
-amenable to single-cycle implementations in hardware. For example, writing a
-register array can actually take several hardware clock cycles, though
-logically it is a single transition. Karczmarek \etal~\cite{Karczmarek} have
-provided a scheme to convert atomic transitions into synchronous hardware in
-which each transition can span several clock cycles.
+%Sometimes, these transitions (both the guards and the actions) may not be
+%amenable to single-cycle implementations in hardware. For example, writing a
+%register array can actually take several hardware clock cycles, though
+%logically it is a single transition. Karczmarek \etal~\cite{Karczmarek} have
+%provided a scheme to convert atomic transitions into synchronous hardware in
+%which each transition can span several clock cycles.

### 2014-02-06 paper/Conclusion.tex
CAV doesn't appear to be single blind.  Add authors

@@
 
-We illustrated a direct realization of our proposed proof framework for a
-concrete MSI cache coherence protocol implementation codifying the intuition
-behind the system's correctness as simple invariants. While this took
-significant effort, once done, we showed that many key hardware
-modifications,\eg{} replacement policies, could be made with a very small
-addition to the proof. These bits were small enough that an informed hardware
-designer implementing our protocols can reasonably be expected to provide such
-additional proofs during refinement, producing a provably correct realistic
-cache coherence engine implementation.
+%% We illustrated a direct realization of our proposed proof framework for a
+%% concrete MSI cache coherence protocol implementation codifying the intuition
+%% behind the system's correctness as simple invariants. While this took
+%% significant effort, once done, we showed that many key hardware
+%% modifications,\eg{} replacement policies, could be made with a very small
+%% addition to the proof. These bits were small enough that an informed hardware
+%% designer implementing our protocols can reasonably be expected to provide such
+%% additional proofs during refinement, producing a provably correct realistic
+%% cache coherence engine implementation.
 
-The protocol framework we showed was very flexible, and could easily be
-extended for variants that go beyond micro-architectural variants, \eg{}
-extension to MOSI with a relatively small amount of additional reasoning. This
-gives us good confidence that we can effectively produce frameworks for other
-classes of protocols, \eg{} update-based protocols, non-inclusive cache
-hierarchies. Further, nothing about our methodology requires we limit ourselves
-to cache coherence systems; this approach should naturally extend to any
-hardware-based system.
+%% The protocol framework we showed was very flexible, and could easily be
+%% extended for variants that go beyond micro-architectural variants, \eg{}
+%% extension to MOSI with a relatively small amount of additional reasoning. This
+%% gives us good confidence that we can effectively produce frameworks for other
+%% classes of protocols, \eg{} update-based protocols, non-inclusive cache
+%% hierarchies. Further, nothing about our methodology requires we limit ourselves
+%% to cache coherence systems; this approach should naturally extend to any
+%% hardware-based system.
 
-While fundamentally this work meets our goal of verification of synthesizable
-hardware specifications, there is still some important mechanization to make
-this ready for industrial use. Foremost, the simple transliteration between the
-inductively defined transition system in Coq and the Bluespec hardware
-description should be fully mechanized.
-%Further, the proposed extension to provide fairness of transaction scheduling
-%in the Bluespec compiler should be implemented.
-Further, to fully guarantee correctness of the resulting hardware, we should
-also verify the hardware synthesis compiler, possibly using FeSi~\cite{Tomas}.
+%% While fundamentally this work meets our goal of verification of synthesizable
+%% hardware specifications, there is still some important mechanization to make
+%% this ready for industrial use. Foremost, the simple transliteration between the
+%% inductively defined transition system in Coq and the Bluespec hardware
+%% description should be fully mechanized.
+%% %Further, the proposed extension to provide fairness of transaction scheduling
+%% %in the Bluespec compiler should be implemented.
+%% Further, to fully guarantee correctness of the resulting hardware, we should
+%% also verify the hardware synthesis compiler, possibly using FeSi~\cite{Tomas}.

### 2014-02-06 paper/Title.tex
CAV doesn't appear to be single blind.  Add authors

@@
 
-\author{
-%    Muralidaran Vijayaraghavan\\Massachusetts Institute of Technology\\vmurali@csail.mit.edu\and
-%    Nirav Dave\\SRI International\\ndave@csl.sri.com\and
-%    Adam Chlipala\\Massachusetts Institute of Technology\\adamc@csail.mit.edu\and
-%    Arvind\\Massachusetts Institute of Technology\\arvind@csail.mit.edu
-}
+%% \author{
+%% %  \adamc{Fill in authors, including Richard or any other appropriate additions}
+%%    Muralidaran Vijayaraghavan\\Massachusetts Institute of Technology\\vmurali@csail.mit.edu \and
+%%    Nirav Dave\\SRI International\\ndave@csl.sri.com \and
+%%    Adam Chlipala\\Massachusetts Institute of Technology\\adamc@csail.mit.edu \and
+%%    Arvind\\Massachusetts Institute of Technology\\arvind@csail.mit.edu
+%% }
+
+\author{Muralidaran Vijayaraghavan\inst{1} \and Nirav Dave\inst{2} \and Adam Chlipala\inst{1} \and Arvind\inst{1}}
+
+\institute{{Massachusetts Institute of Technology,}\\{Computer Science and
+    Artificial Intelligence Laboratory, Cambridge, MA, USA}\\\email{\{vmurali, adamc, arvind\}@csail.mit.edu} \and
+           {SRI International, Computer Science Laboratory, Menlo Park, CA,
+       USA}\\\email{ndave@csl.sri.com}}
 

### 2014-02-06 paper/Intuition.tex

@@
 %address (Invariant~\textit{processReq}).
-Leaf caches cannot be ancestors of any other node, so by Invariant~
-\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
+Leaf caches cannot be ancestors of any other node, so by Invariant~\fullref{nonAncestorCompatible}, if a leaf cache is in \Mo{} state, all other
 leaf caches will be in \In{} state, blocking simultaneous
@@
 supplies the latest value by Invariant~\fullref{latestValue}. Thus, all the
-conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants~
-\ref{latestValue} to \ref{noReqAgain}.
+conditions for Theorem~\fullref{storeAtomicity} are satisfied given Invariants~\ref{latestValue} to \ref{noReqAgain}.
 
@@
 While Invariants~\fullref{processReq} and \fullref{noReqAgain} can be seen
-directly from the MSI transitions, correctness of Invariants~
-\fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
+directly from the MSI transitions, correctness of Invariants~\fullref{nonAncestorCompatible} and \fullref{latestValue} requires further
 decomposition. We present 4 more invariants that are helpful in the proof.
@@
 none of its children's directory entries can be in \Mo{}, all its descendants
-should be no greater than \Sh{} for that address (because of Invariant~
-\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
+should be no greater than \Sh{} for that address (because of Invariant~\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
 state of a cache and that of its ancestor). Any non-ancestor of $c$ and
@@
 If $c$ was not clean at time $t$, it must have received data from either its
-parent or its child, and the sender is \clean{} by Invariant~
-\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
+parent or its child, and the sender is \clean{} by Invariant~\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
 the latest value for the address. Again, because of
@@
 
-Figure~\ref{resps} shows an example of the effect of Invariants~
-\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
+Figure~\ref{resps} shows an example of the effect of Invariants~\fullref{stateChange}, \fullref{noCross} and \fullref{respFIFO}. The directory
 entry of the parent or the state of the child changes on sending and receiving
@@
 more properties of the MSI transistions of Figure~\ref{trans}, though the argument is
-straightforward. Analyzing a parent-child pair will suffice in proving Invariants~
-\fullref{localCompatibility1} and \fullref{localCompatibility2}. Proof of Invariant~
-\fullref{noTransitWrite} is similar to the proof of Invariant~
-\fullref{nonAncestorCompatible}.
+straightforward.
+Analyzing a parent-child pair will suffice in proving Invariants~\fullref{localCompatibility1} and \fullref{localCompatibility2}.
+Proof of Invariant~\fullref{noTransitWrite} is similar to the proof of Invariant~\fullref{nonAncestorCompatible}.
 

### 2014-02-06 paper/RelatedWork.tex
stuff

\section{Related work}
\label{relatedWork}
These are lifted as is from that guy. Need to figure out how to restate them, and make it more concise.

For verification of high level specifications, modern industrial practice
consists of modeling small instances of the protocols, e.g., three CPUs
handling two addresses and one bit of data, in terms of interleaving atomic
steps, in guard/action languages such as Murphi \cite{tla} or TLA+ \cite{tla}, and
exploring the reachable states through explicit state enumeration.
Symbolic methods, e.g., BDD [31] or SAT [129, 142], are yet to succeed for the
verification of cache coherence protocols.
Monolithic formal verification methods – methods that treat the protocol as a whole – have
been used fairly routinely for verifying cache coherence protocols from the early 1990s
[10 (2 nodes, murphi), 61 (), 141]. However, these monolithic techniques will not be able to handle the very
large state space of hierarchical protocols. Compositional verification techniques are essential.

### 2014-02-06 paper/Conclusion.tex
stuff

@@
 \section{Conclusions and future work}
+
+* Present work
+- Abstract converted into ``have showns''
+- Replacement policies/policies to decide when a request a sent etc (along with
+  associativity/sizing of caches) is parameterized
+* Future work
+- Forward progress/liveness property
+  - closely related to network/buffers
+  - in presence of bounded buffers
+  - in presence of communication fabric
+- Automating Bluespec specification into Coq/vice versa
+  - Verified compiler
 

### 2014-02-06 paper/Introduction.tex
stuff

@@
 
-The main contribution of this paper is \textbf{the first verification of a
+The main contribution of this paper is, \textbf{the first verification of a
 cache coherence protocol that is parametric over an infinite family of

### 2014-02-06 paper/RelatedWork.tex
stuff

@@
 been used fairly routinely for verifying cache coherence protocols from the early 1990s
-[10 (2 nodes, murphi), 61 (), 141]. However, these monolithic techniques will not be able to handle the very
+[10 (2 nodes, murphi)]. However, these monolithic techniques will not be able to handle the very
 large state space of hierarchical protocols. Compositional verification techniques are essential.

### 2014-02-06 paper/Conclusion.tex
Conclusions added back and changed.

--- 2014-02-06 paper/Conclusion.tex
@@
 \section{Conclusions and future work}
+
+In this paper, we provided a formal definition of cache coherence via the store
+atomicity property and provided a concrete hierarchy-parametric directory-based
+MSI protocol whose correct we formally checked via a Coq-based proof. To our
+knowledge this is the first instance of a formal proof of a protocol parametric
+over the memory hierarchy.
+
+Our protocol is described in a style amenable to mechanical transliteration to a
+Bluespec SystemVerilog hardware description allowing us to provide high-quality
+verified cache coherence engine. As the protocol and associated proof are robust
+to many microarchitecture-oriented changes (\eg{} arbitrary hierarchy and
+arbitrary cache size, associativity, and replacement policy choices in each
+node), we should be able to produce a performant formally verified hardware
+cache coherence system.
+
+Moving this work towards generating a fully verified MSI implementation requires
+a number of small steps. First, the simple transliteration of descriptions
+between Bluespec and Coq, must be done in a mechanical way. Second, the
+compilation of Bluespec to hardware is unverified; to fully trust the
+implementation the Bluespec compiler must be verified. Some initial work on this
+has already been done~\cite{TDBLP:conf/cav/BraibantC13}. Finally, we must extend
+our notion of correctness to apply to the hardware substrate and not just the
+level of atomic transactions. Most notably, this requires the some guarantee of
+liveness and fairness in handling of messages in the final design. We have
+initial Coq-based proofs that with finite buffers a local fairness in
+transaction our protocol will have this property, but more work is needed to
+fully resolve this venture.
 
@@
 %% also verify the hardware synthesis compiler, possibly using FeSi~\cite{Tomas}.
+
+%%  LocalWords:  associativity atomicity

### 2014-02-06 paper/Conclusion.tex
changed ending a bit to be less terrible

@@
 initial Coq-based proofs that with finite buffers a local fairness in
-transaction our protocol will have this property, but more work is needed to
-fully resolve this venture.
+transaction our protocol will have this property, but more work is needed.
 

### 2014-02-06 paper/RelatedWork.tex
stuff

@@
 \label{relatedWork}
-These are lifted as is from that guy. Need to figure out how to restate them, and make it more concise.
+Formal verification of cache coherence protocols, among other hardware systems,
+is mainly done using model checking. Explicit state model checking tools like
+Murphi \cite{murphi} or TLC \cite{tlc} have been used in practise only to
+handle small and simplified systems, \eg{} a single level cache hierarchy with
+three CPUs handling two addresses and one bit of data. Symbolic model checking
+techniques have fared slightly better, McMillan \etal{} have verified a 2-level
+MSI protocol based on Gigamax distributed multiprocessor using SMV.
+Optimizations on these techniques (like partial order reduction \cite{part},
+symmetry reduction \cite{sym1, sym2}, compositional reasoning \cite{somethink}, \etc
+scale the model checking approach further, but they still haven't been able to
+verify realistic hierarchical coherence protocols.
 
-For verification of high level specifications, modern industrial practice
-consists of modeling small instances of the protocols, e.g., three CPUs
-handling two addresses and one bit of data, in terms of interleaving atomic
-steps, in guard/action languages such as Murphi \cite{tla} or TLA+ \cite{tla}, and
-exploring the reachable states through explicit state enumeration.
-Symbolic methods, e.g., BDD [31] or SAT [129, 142], are yet to succeed for the
-verification of cache coherence protocols.
-Monolithic formal verification methods – methods that treat the protocol as a whole – have
-been used fairly routinely for verifying cache coherence protocols from the early 1990s
-[10 (2 nodes, murphi)]. However, these monolithic techniques will not be able to handle the very
-large state space of hierarchical protocols. Compositional verification techniques are essential.
+Theorem prover based approaches have also been used to verify cache coherence
+protocols. FLASH coherence protocol \cite{flash} has been verified using an
+approach called Aggregation of Distributed transactions \cite{park}.  Several
+state-transitions in the implementation accessing localized states intuitively
+correspond to a single atomic transition accessing all the states globally. In
+this approach, the user has to provide a proof of a mapping between the
+implemantion transitions to the aggregated atomic transition; once this is
+done, a model checker is used to verify the aggregated transition resulting in
+verification of the implementation. We believe that the difficult part in
+proving a protocol is precisely the proof of such a mapping from an
+implementation-level transition to a specification-level transitions accessing
+states globally, and their work does not present any approach to go about it.
+
+%These are lifted as is from that guy. Need to figure out how to restate them, and make it more concise.
+%
+%For verification of high level specifications, modern industrial practice
+%consists of modeling small instances of the protocols, e.g., three CPUs
+%handling two addresses and one bit of data, in terms of interleaving atomic
+%steps, in guard/action languages such as Murphi \cite{tla} or TLA+ \cite{tla}, and
+%exploring the reachable states through explicit state enumeration.
+%Symbolic methods, e.g., BDD [31] or SAT [129, 142], are yet to succeed for the
+%verification of cache coherence protocols.
+%Monolithic formal verification methods – methods that treat the protocol as a whole – have
+%been used fairly routinely for verifying cache coherence protocols from the early 1990s
+%[10 (2 nodes, murphi)]. However, these monolithic techniques will not be able to handle the very
+%large state space of hierarchical protocols. Compositional verification techniques are essential. 
+%
+%For coherence protocols which only consider one level (or nonhierarchical protocols),
+%various techniques have been proposed for the verification. These techniques basically
+%include explicit state enumeration \cite{sci-cache-coherence} and symbolic model checking \cite{gigamax}
+%Opti-
+%mizations on these techniques include partial order reduction [24], symmetry reduction
+%[25, 72], compositional reasoning [96], predicate abstraction [22, 49, 82], etc. There is
+%also a large body of work on parameterized verification [42, 53, 54, 100, 117]. However,
+%none of them have shown the capability to verify hierarchical coherence protocols with
+%realistic complexity.
+%
+%McMillan et al. [97, 102] modeled a 2-level MSI coherence protocol based on the
+%Gigamax distributed multiprocessor [102]. In the protocol, bus snooping is used in both
+%levels. SMV [97] was used to verify this protocol with two clusters each having six
+%processors, for both safety and liveness properties. In terms of the complexity, many
+%realistic details are abstracted away from the protocol so that the protocol does not have
+%the typical complexity that a hierarchical coherence protocol has.
+%

### 2014-02-06 paper/Conclusion.tex
stuff

--- 2014-02-06 paper/Conclusion.tex
@@
 
-* Present work
-- Abstract converted into ``have showns''
-- Replacement policies/policies to decide when a request a sent etc (along with
-  associativity/sizing of caches) is parameterized
-* Future work
-- Forward progress/liveness property
-  - closely related to network/buffers
-  - in presence of bounded buffers
-  - in presence of communication fabric
-- Automating Bluespec specification into Coq/vice versa
-  - Verified compiler
+In this paper, we provided a formal definition of cache coherence via the store
+atomicity property and provided a concrete hierarchy-parametric directory-based
+MSI protocol whose correctness we formally checked via a Coq-based proof. To our
+knowledge this is the first instance of a formal proof of a protocol parametric
+over the cache hierarchy. We have also provided a set of invariants not
+specific to MSI which imply store atomicity; other protocols that obey these
+invariants are already proven to be correct. Finally, we have identified a set
+of key properties that seem to be shared with many other protocols which ease
+the verification considerably.
+
+Our protocol is described in a style amenable to mechanical transliteration to a
+Bluespec SystemVerilog hardware description allowing us to provide high-quality
+verified cache coherence engines. As the protocol and associated proof are robust
+to many microarchitecture-oriented changes (\eg{} arbitrary hierarchy and
+arbitrary cache size, associativity, replacement policy choices, policies to
+handle incoming requests, \etc{} in each node), we should be able to produce a
+high performance performance formally verified hardware cache coherence system.
+
+Moving this work towards generating a fully verified MSI implementation
+requires a number of small steps. First, the simple transliteration of
+descriptions between Bluespec and Coq, must be done in a mechanical way.
+Second, the compilation of Bluespec to hardware is unverified; to fully trust
+the implementation the Bluespec compiler must be verified. Some initial work on
+this has already been done~\cite{TDBLP:conf/cav/BraibantC13}. Finally, we must
+extend our notion of correctness to provide some guarantee of
+liveness and fairness in handling of requests from processors. We have
+initial Coq-based proofs that with finite buffers a local fairness in
+transaction our protocol will have this property, but more work is needed to
+fully resolve this venture.
 

### 2014-02-06 paper/Intuition.tex
stuff

@@
 
-Invariants~\fullref{conservative} and \fullref{localCompatibility1} defines a
+Invariants~\fullref{conservative} and \fullref{localCompatibility1} define a
 $\ge$ relation between the states of a parent and its child for any address.
@@
 Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
-lowest common ancestor $x$ that is an ancestor of both $c_1$ and $c_2$. $x$
-has two distinct children $x_1$ and $x_2$ such that they are the ancestors of
-$c_1$ and $c_2$ respectively. For any address $a$, $x_1.\state[a]$ and
-$x_2.\state[a]$ are compatible by 
-Invariant~\fullref{localCompatibility2}. It follows that $x_1.\state[a]$ and
-$x_2.\state[a]$ are compatible since these values are smaller than the
-respective directory values. From the previous paragraph about the relation
-between the states of a cache and its ancestor, it follows that $c_1$ and $c_2$
-are also compatible. This proves Invariant~\fullref{nonAncestorCompatible}.
+lowest common ancestor $x$ that is an ancestor of both $c_1$ and $c_2$. $x$ has
+two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
+and $c_2$ respectively. For any address $a$, $x_1.\state[a]$ and
+$x_2.\state[a]$ are compatible by Invariant~\fullref{localCompatibility2}. It
+follows that $x_1.\state[a]$ and $x_2.\state[a]$ are compatible since these
+values are smaller than the respective directory values. From the previous
+paragraph about the relation between the states of a cache and its ancestor, it
+follows that $c_1$ and $c_2$ are also compatible. This proves
+Invariant~\fullref{nonAncestorCompatible}.
 
-We prove Invariant~\fullref{latestValue} by strong induction on time. Consider
-two cases: (a) cache $c$ is already \clean{} for an address $a$ at time $t$ or
-(b) the contrary.  If \clean{}, then it has the latest value at time $t$. Since
-none of its children's directory entries can be in \Mo{}, all its descendants
-should be no greater than \Sh{} for that address (because of Invariant~\fullref{conservative} and the fact that the $\ge$ relation is preserved for the
-state of a cache and that of its ancestor). Any non-ancestor of $c$ and
-non-descendant of $c$ must be compatible (as discussed in the previous
-paragraph).  Since $c$ is \clean, it is at least in the \Sh{} state, so
-compatibility means that non-descendants cannot be in \Mo{} state. All of
-$c$'s ancestors are non-leaves, so they cannot process any store requests.
-Therefore, no cache save $c$ may have processed a store request, which means that
-$c$ contains the latest value for address $a$ at time $t+1$ if it was \clean{}
-at time $t$.
-
-If $c$ was not clean at time $t$, it must have received data from either its
-parent or its child, and the sender is \clean{} by Invariant~\ref{noTransitWrite}. By the induction hypothesis, it means that the sender has
-the latest value for the address. Again, because of
-Invariant~\ref{noTransitWrite}, no store request can be processed when the data
-is in transit, which means $c$ will have the latest value for address $a$ at
-time $t+1$, establishing Invariant~\fullref{latestValue}.
+Using the properties discussed in the last two paragraphs, it is easy to prove
+Invariant~\ref{latestValue}. The proof is by induction on time $t$. We consider
+the cases when a node is either \clean{} in the previous state or not. One can
+easily prove that if it was \clean{} in the previous state, then no cache, save
+itself, can process a store request, proving the invariant.  If the cache was
+not \clean{} in the previous state, it will receive data from a \clean{} cache
+and no store requests can be processed when the data is in transit (by
+Invariant~\ref{noTransitWrite}). Therefore, by the induction hypothesis it has
+the up-to-date data.
 
@@
 
-Proving the other 3 invariants in Section~\ref{secondLevel} requires using
-more properties of the MSI transistions of Figure~\ref{trans}, though the argument is
-straightforward.
-Analyzing a parent-child pair will suffice in proving Invariants~\fullref{localCompatibility1} and \fullref{localCompatibility2}.
-Proof of Invariant~\fullref{noTransitWrite} is similar to the proof of Invariant~\fullref{nonAncestorCompatible}.
+Proving the other 3 invariants in Section~\ref{secondLevel} requires using more
+properties of the MSI transistions of Figure~\ref{trans}, though the argument
+is straightforward. Analyzing a parent-child pair will suffice in proving
+Invariants~\fullref{localCompatibility1} and \fullref{localCompatibility2}.
+Proof of Invariant~\fullref{noTransitWrite} is similar to the proof of
+Invariant~\fullref{nonAncestorCompatible}.
 

### 2014-02-06 paper/RelatedWork.tex

@@
 \label{relatedWork}
-Formal verification of cache coherence protocols, among other hardware systems,
-is mainly done using model checking. Explicit state model checking tools like
-Murphi \cite{murphi} or TLC \cite{tlc} have been used in practise only to
-handle small and simplified systems, \eg{} a single level cache hierarchy with
-three CPUs handling two addresses and one bit of data. Symbolic model checking
-techniques have fared slightly better, McMillan \etal{} have verified a 2-level
-MSI protocol based on Gigamax distributed multiprocessor using SMV.
-Optimizations on these techniques (like partial order reduction \cite{part},
-symmetry reduction \cite{sym1, sym2}, compositional reasoning \cite{somethink}, \etc
-scale the model checking approach further, but they still haven't been able to
-verify realistic hierarchical coherence protocols.
 
-Theorem prover based approaches have also been used to verify cache coherence
-protocols. FLASH coherence protocol \cite{flash} has been verified using an
-approach called Aggregation of Distributed transactions \cite{park}.  Several
-state-transitions in the implementation accessing localized states intuitively
-correspond to a single atomic transition accessing all the states globally. In
-this approach, the user has to provide a proof of a mapping between the
-implemantion transitions to the aggregated atomic transition; once this is
-done, a model checker is used to verify the aggregated transition resulting in
-verification of the implementation. We believe that the difficult part in
-proving a protocol is precisely the proof of such a mapping from an
-implementation-level transition to a specification-level transitions accessing
-states globally, and their work does not present any approach to go about it.
+Formal verification of hardware-oriented cache coherence protocols, is mainly
+done using model checking. In practice, explicit state model checking tools like
+Murphi \cite{murphi} or TLC \cite{tlc} are only able to handle small and
+simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
+addresses, and one bit of data. Symbolic model checking techniques have fared
+slightly better; McMillan \etal{} have verified a 2-level MSI protocol based on
+Gigamax distributed multiprocessor using SMV.  Optimizations on these techniques
+(\eg{} partial order reduction \cite{part}, symmetry reduction \cite{sym1,
+  sym2}, compositional reasoning \cite{somethink}, \etc{} scale the approach
+further, but they are still unable to handle realistic hierarchical protocols.
+
+Theorem provers have also been used in verifying cache coherence protocols. The
+FLASH coherence protocol \cite{flash} was verified using Aggregation of
+Distributed transactions \cite{park}, where several state transitions accessing
+localized state intuitively correspond to an aggregate atomic transition
+accessing all global state. A proof of a mapping between the local transitions
+and the agreggate one must be provided. Once this is done, a model checker can
+be used to verify the aggregate model is correct. However, We believe that the
+verification challenge here is precisely the proof of such a mapping which was
+not discussed.
 

### 2014-02-06 paper/Introduction.tex
Add TRS intro

@@
 our knowledge, the largest system that has been verified using model checking
-has only 12 processors\xxx{the comment was meant to cite other papers from this paper as well. anyways i will take a crack at it}~\cite{Chen:2008:VHC:1559300}, in an era when many
+has only 12 processors~\cite{Chen:2008:VHC:1559300}, in an era when many
 observers are predicting mainstream systems with hundreds of processor cores.
@@
 
-Protocols, post verification, are often described via Term Rewrite Systems
-(TRS) \adamc{citation for TRSes?}, which apply equational logic to describe
-systems at an interesting point on the spectrum between declarative and
-operational methods.  It was shown by Dave
-\etal{}~\cite{DNA:CoherenceImplementation} that such a protocol, after its
-verification, can be transliterated into Bluespec~\cite{Bluespec:TFRG}, a
-hardware description language based on Guarded Atomic Actions, a restriction of
-TRS~\cite{Hoe:TCAD}. Bluespec specifications can be directly implemented into
-efficient hardware. In this way one reduces the correctness of the
-implementation to the correctness of the protocol, with no need for separate
-reasoning about an implementation. While the current Bluespec compiler does not
-support some implementation elements that are common in practice (\eg{}
-fairness guarantees, atomic actions running over multiple clock cycles), recent
-work has provided key insights toward lifting the
-restrictions~\cite{Karczmarek}. We believe that Bluespec and its future
-evolutions represent a very promising platform for productive implementation of
-hardware components, including cache-coherent shared memories, starting from a
-specification designed to support verification.
+Protocols, post verification, are often described via Term Rewrite
+Systems (TRS)~\cite{bezem2003term}, which apply equational logic to
+describe systems at an interesting point on the spectrum between
+declarative and operational methods.  It was shown by Dave
+\etal{}~\cite{DNA:CoherenceImplementation} that such a protocol, after
+its verification, can be transliterated into
+Bluespec~\cite{Bluespec:TFRG}, a hardware description language based
+on Guarded Atomic Actions, a restriction of
+TRS~\cite{Hoe:TCAD}. Bluespec specifications can be directly
+implemented into efficient hardware. In this way one reduces the
+correctness of the implementation to the correctness of the protocol,
+with no need for separate reasoning about an implementation. While the
+current Bluespec compiler does not support some implementation
+elements that are common in practice (\eg{} fairness guarantees,
+atomic actions running over multiple clock cycles), recent work has
+provided key insights toward lifting the
+restrictions~\cite{Karczmarek}. We believe that Bluespec and its
+future evolutions represent a very promising platform for productive
+implementation of hardware components, including cache-coherent shared
+memories, starting from a specification designed to support
+verification.
 

### 2014-02-06 paper/Intuition.tex
stuff

@@
 MOSI and MOESI in Section~\ref{sec:Refinements}.
-\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
+%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
 
@@
 While we have proofs in Coq for each invariant, our discussion will be mainly
-informal for brevity. A full description can be found at~\xxx{MURALI PUT
-REFERENCE HERE}. \adamc{or just remind of the source code URL given earlier}
+informal for brevity. As mentioned earlier, all the proofs can be found at~\xxx{MURALI PUT
+REFERENCE HERE}. %\adamc{or just remind of the source code URL given earlier}
 
@@
 \includegraphics[scale=.5]{resps}
-\caption{An example showing a sequence of responses ping-ponging between a child and its parent
-\adamc{For a camera-ready version, we should replace these bitmap images with vector images, for better printing.}}
+\caption{An example showing a sequence of responses ping-ponging between a child and its parent}
+%\adamc{For a camera-ready version, we should replace these bitmap images with vector images, for better printing.}}
 \label{resps}

### 2014-02-06 paper/System.tex
stuff

@@
 We have chosen as our concrete example a hierarchical MSI
-protocol~\cite{MSI}. Logically, the system is organized in a tree
+protocol. Logically, the system is organized in a tree
 structure, with the root connected to the main memory and each leaf interfaced
@@
 sequence number of the next request to process; it is set to $0$ initially.
-\adamc{Add processors and main memory, labeled explicitly, to the diagram?}
+%\adamc{Add processors and main memory, labeled explicitly, to the diagram?}
 
@@
 \end{subfigure}
-\caption{Data types for messages between caches
-\adamc{Names here don't match those in the preceding text.}}
+\caption{Data types for messages between caches}
+%\adamc{Names here don't match those in the preceding text.}}
 \label{format}

### 2014-02-06 paper/RelatedWork.tex
stuff
Pass through larval RelatedWork

@@
 
-Formal verification of hardware-oriented cache coherence protocols, is mainly
-done using model checking. In practice, explicit state model checking tools like
+Formal verification of hardware-oriented cache coherence protocols is mainly
+done using model checking. In practice, explicit-state model checking tools like
 Murphi \cite{murphi} or TLC \cite{tlc} are only able to handle small and
@@
 slightly better; McMillan \etal{} have verified a 2-level MSI protocol based on
-Gigamax distributed multiprocessor using SMV.  Optimizations on these techniques
+the Gigamax distributed multiprocessor using SMV.  Optimizations on these techniques
 (\eg{} partial order reduction \cite{part}, symmetry reduction \cite{sym1,
-  sym2}, compositional reasoning \cite{somethink}, \etc{} scale the approach
+  sym2}, compositional reasoning \cite{somethink}, \etc{}) scale the approach
 further, but they are still unable to handle realistic hierarchical protocols.
@@
 FLASH coherence protocol \cite{flash} was verified using Aggregation of
-Distributed transactions \cite{park}, where several state transitions accessing
+Distributed \adamc{why are these words capitalized?} transactions \cite{park}, where several state transitions accessing
 localized state intuitively correspond to an aggregate atomic transition
 accessing all global state. A proof of a mapping between the local transitions
-and the agreggate one must be provided. Once this is done, a model checker can
-be used to verify the aggregate model is correct. However, We believe that the
-verification challenge here is precisely the proof of such a mapping which was
+and the agreggate one must be provided, at which point a model checker can
+be used to verify that the aggregate model is correct. However, we believe that the
+verification challenge here is precisely the proof of such a mapping, which was
 not discussed.
+
+Other approaches involve correct-by-construction
 

### 2014-02-06 paper/Conclusion.tex
Prune related work
Pass over Conclusion and some capitalization of other section titles

@@
-\section{Conclusions and future work}
+\section{Conclusions and Future Work}
 
@@
 MSI protocol whose correctness we formally checked via a Coq-based proof. To our
-knowledge this is the first instance of a formal proof of a protocol parametric
+knowledge, this is the first instance of formal verification of a protocol parametric
 over the cache hierarchy. We have also provided a set of invariants not
-specific to MSI which imply store atomicity; other protocols that obey these
+specific to MSI that imply store atomicity; other protocols that obey these
 invariants are already proven to be correct. Finally, we have identified a set
-of key properties that seem to be shared with many other protocols which ease
+of key properties that seem to be shared with many other protocols and which ease
 the verification considerably.
@@
 Our protocol is described in a style amenable to mechanical transliteration to a
-Bluespec SystemVerilog hardware description allowing us to provide high-quality
+Bluespec SystemVerilog hardware description, allowing us to provide high-quality
 verified cache coherence engines. As the protocol and associated proof are robust
@@
 handle incoming requests, \etc{} in each node), we should be able to produce a
-high performance performance formally verified hardware cache coherence system.
+high-performance formally verified hardware cache coherence system.
 
@@
 requires a number of small steps. First, the simple transliteration of
-descriptions between Bluespec and Coq, must be done in a mechanical way.
+descriptions between Bluespec and Coq must be done in a mechanical way.
 Second, the compilation of Bluespec to hardware is unverified; to fully trust
-the implementation the Bluespec compiler must be verified. Some initial work on
-this has already been done~\cite{TDBLP:conf/cav/BraibantC13}. Finally, we must
+the implementation the Bluespec compiler must be verified. Some first steps
+have been taken in that direction~\cite{TDBLP:conf/cav/BraibantC13}. Finally, we must
 extend our notion of correctness to provide some guarantee of
 liveness and fairness in handling of requests from processors. We have
-initial Coq-based proofs that with finite buffers a local fairness in
-transaction our protocol will have this property, but more work is needed.
+initial Coq-based proofs that with finite buffers a local fairness condition
+holds for our protocol, but more work is needed.
 

### 2014-02-06 paper/RelatedWork.tex
Pass over Conclusion and some capitalization of other section titles
stuff

@@
-\section{Related work}
+\section{Related Work}
 \label{relatedWork}
@@
 slightly better; McMillan \etal{} have verified a 2-level MSI protocol based on
-the Gigamax distributed multiprocessor using SMV.  Optimizations on these techniques
+the Gigamax distributed multiprocessor using SMV \cite{gigamax}.  Optimizations on these techniques
 (\eg{} partial order reduction \cite{part}, symmetry reduction \cite{sym1,
-  sym2}, compositional reasoning \cite{somethink}, \etc{}) scale the approach
+  sym2}, compositional reasoning \cite{comp}, \etc{}) scale the approach
 further, but they are still unable to handle realistic hierarchical protocols.
@@
 Theorem provers have also been used in verifying cache coherence protocols. The
-FLASH coherence protocol \cite{flash} was verified using Aggregation of
-Distributed \adamc{why are these words capitalized?} transactions \cite{park}, where several state transitions accessing
-localized state intuitively correspond to an aggregate atomic transition
-accessing all global state. A proof of a mapping between the local transitions
-and the agreggate one must be provided, at which point a model checker can
-be used to verify that the aggregate model is correct. However, we believe that the
-verification challenge here is precisely the proof of such a mapping, which was
-not discussed.
+FLASH coherence protocol was verified using aggregation of
+distributed \adamc{why are these words capitalized?} transactions \cite{park},
+where several state transitions accessing localized state intuitively
+correspond to an aggregate atomic transition accessing all global state. A
+proof of a mapping between the local transitions and the agreggate one must be
+provided, at which point a model checker can be used to verify that the
+aggregate model is correct. However, we believe that the verification challenge
+here is precisely the proof of such a mapping, which was not discussed.
 
-Other approaches involve correct-by-construction
+\xxx{MUST WRITE ABOUT XIAOWEI's CACHET and JOE STOY'S PVS proof}
 
@@
 %handling two addresses and one bit of data, in terms of interleaving atomic
-%steps, in guard/action languages such as Murphi \cite{tla} or TLA+ \cite{tla}, and
+%steps, in guard/action languages such as Murphi [50] or TLA+ [84], and
 %exploring the reachable states through explicit state enumeration.
@@
 %various techniques have been proposed for the verification. These techniques basically
-%include explicit state enumeration \cite{sci-cache-coherence} and symbolic model checking \cite{gigamax}
+%include explicit state enumeration [133] and symbolic model checking [102]
 %Opti-

### 2014-02-07 paper/Conclusion.tex

@@
 Second, the compilation of Bluespec to hardware is unverified; to fully trust
-the implementation the Bluespec compiler must be verified. Some first steps
-have been taken in that direction~\cite{TDBLP:conf/cav/BraibantC13}. Finally, we must
+the implementation, the Bluespec compiler must be verified. Some first steps
+have been taken in that direction~\cite{DBLP:conf/cav/BraibantC13}. Finally, we must
 extend our notion of correctness to provide some guarantee of

### 2014-02-07 paper/Introduction.tex
stuff

@@
 as well as complexity of the system that has to be verified; most model-checking
-based approaches can not handle more than one level of hierarchy. To the best of
+based approaches cannot handle more than one level of hierarchy. To the best of
 our knowledge, the largest system that has been verified using model checking
@@
 
-\adamc{This might be a good place to include a link to the Coq source code.}
+The Coq source code for the definition of store atomicity and the proof of the
+MSI protocol can be found in \url{https://github.com/vmurali/CacheProof}.
+
+%\adamc{This might be a good place to include a link to the Coq source code.}

### 2014-03-21 rebuttal.txt
added rebuttal

This work is the first verification of a directory-based cache coherence
protocol parameterized for any hierarchy (Reviewer 3, please note that any
cache hierarchy is indeed a tree-hierarchy). Directory based protocols have
intermediate states to keep track of waiting states, and these intermediate
states are designed and reasoned about in an ad-hoc manner - and these ad-hoc
reasoning/design is difficult and usually does not scale with more levels in
the hierarchy. We provide a systematic way to reason about (and design) the
intermediate wait states for a general hierarchy, clearly specifying the
semantics of the intermediate states. This by itself is a novel contribution.
Of course, current processors have only 2 to 4 levels of hierarchy (note to
Reviewer 4: Intel Itanium and Haswell has 4 levels of caches), but with the
current trend, it is not unfathomable to imagine the number of cores increasing
to thousands and even millions, at which point multi-level hierarchies will
become very important.

Next, we provide a declarative definition of store atomicity. We believe this
is the first work that spells out all the conditions sufficient to make a cache system
behave like atomic memory without caches (Lamport's model "How to Make a
Multiprocessor Computer That Correctly Executes Multiprocess Programs"). All
previous verification work stop at proving that two cache states are never both
in "M", which is an important invariant but not sufficient.

The insight for the proof presented in the paper has not been presented before,
namely, viewing multiple cache coherence protocols in a unified manner sharing
a common proof substrate, all of which depends on seemingly unremarkable
properties of "no-crossing of messages" and "FIFO delivery". In fact, most of
the human effort in this project was in coming up with the insight for viewing
multiple protocols in a unified manner. Protocols other than MSI have not been
formally proven, but the libraries we provide in the proof of MSI can be used
without any modification for the other protocols (this constitutes more than
80% of the 12000-line Coq proof that we had written). Moreover, the invariants
that remain to be proven for MOSI/MOESI protocols is restricted to a system
containing exactly 3 nodes, and a single address, making it amenable to
model-checking.

We acknowledge Reviewer 4 for pointing out the other works on verifying cache
coherence (especially using model checking techniques while still being
parametric on the number of processors). These works a) support only one-level
hierarchy and b) all except one prove correctness of snoopy protocols. Other
works that we have looked are similar or they verify a specific instance.

This protocol has been implemented in a multicore PowerPC system (Fast and
cycle-accurate modeling of a multicore processor, ISPASS-2012) with 2 levels of
hierarchy, and the performance obtained is similar to other systems. One
obstacle in the implementation is that one atomic transition of Figure 3 can
span several clock cycles, requiring bookkeeping to keep it atomic.

Extending the proof to broadcast-based protocols: The first-level invariants
remain as is, but beyond that, the reasoning becomes different.
